---
title: "Data preprocessing for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

Genomic selection requires phenotypes and markers to represent the same
individuals in the same order. In this module, both data sources are read,
identifiers are organized, SSR missing-value codes are converted to `NA`,
phenotypes are aligned with genotypes, and SSR loci are subjected to quality
control.

The block column (`BL`) is removed because it is not part of the statistical model adopted in this analysis. Family (`FAM`) is preserved because it will be used both in the fixed-effect matrix and to stratify the cross-validation folds in Module 02. No individual is removed in this module.

Preprocessing produces three central objects for the following stages: the aligned phenotype matrix, the SSR matrix after locus quality control, and metadata containing individual and family. The diagnostics verify the consistency of these objects before genomic-matrix construction.

## 2. Packages used

The packages used in this module have distinct roles:

- `readxl`: reads the Excel workbooks containing genotype and phenotype data;
- `dplyr`: selects, transforms, and organizes tabular data;
- `tidyr`: reshapes data between wide and long formats;
- `ggplot2`: builds the quality-control and phenotype-distribution figures;
- `ggthemes`: applies the GDocs theme and color scales used in the figures;
- `knitr`: formats tables in R Markdown documents;
- `here`: builds project-relative file paths that do not depend on the local working directory.

```{r packages, warning=FALSE, message=FALSE}
library(readxl)  # Read Excel workbooks.
library(dplyr)   # Manipulate and transform tables.
library(tidyr)   # Reshape data between wide and long formats.
library(ggplot2) # Build figures.
library(ggthemes) # Apply the GDocs theme and color scales.
library(knitr)   # Format tables and R Markdown output.
library(here)    # Build portable project-relative paths.
```

```{r table-formatting, include=FALSE}
format_table <- function(x, caption = NULL, digits = getOption("digits"), col.names = NA) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    col.names = col.names,
    table.attr = paste0(
      'class="table table-striped table-hover" ',
      'style="width:100%; margin-bottom:0;"'
    )
  )
  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

## 3. Reading the original spreadsheets

### 3.1 Genotype data

Genotype information is concentrated in two sheets of the genotype workbook. The `Genes` sheet contains the numerical SSR codes: its first column is an index and the next 35 columns correspond to the original loci. Because this sheet has no row containing column names, it is read with `col_names = FALSE`.

The `Genealex` sheet provides individual identifiers and locus names. It is also read without column names because it preserves metadata rows and the original GenAlEx layout. The positions containing these identifiers are handled explicitly in the next section.

```{r read-genotypes}
genes <- read_excel(
  here("data", "papaya_ssr_genotypes.xlsx"),
  sheet = "Genes",
  col_names = FALSE,
  na = c("", "NA")
)

genealex <- read_excel(
  here("data", "papaya_ssr_genotypes.xlsx"),
  sheet = "Genealex",
  col_names = FALSE,
  na = c("", "NA")
)
```

The initial structure of `Genes` can be inspected in its first six rows and six columns:

```{r genes-preview}
as.data.frame(genes[1:6, 1:6]) |>
  format_table(caption = "Preview of the Genes sheet")
```

The preview shows the raw genotype layout: the first column is a sequential index, and the following columns contain the SSR codes organized in the next section.

### 3.2 Phenotype data

Phenotype data are stored in the `Médias_IND_artigo` sheet, which already contains column names. This source includes individual identifiers, family membership, and the evaluated traits.

The `BL` column, corresponding to experimental block, is retained at this first stage so that `pheno_raw` faithfully represents the original spreadsheet. Because `BL` is not part of the statistical model used here, it will be left out when the analytical set of columns is defined in Section 5.

```{r read-phenotypes}
pheno_raw <- read_excel(
  here("data", "papaya_phenotypes_2024_2025.xlsx"),
  sheet = "Médias_IND_artigo",
  na = c("", "NA")
)
```

The phenotype table is inspected while it is still in its original form:

```{r phenotype-preview}
head(pheno_raw) |>
  format_table(caption = "Preview of the original phenotype data")
```

The preview shows the phenotype spreadsheet before analytical variables are selected, including `IND`, `FAM`, `BL`, and the ten evaluated traits.

### 3.3 Initial check of the data sources

Before organization and alignment, the dimensions of the three imported structures are compared. This is a structural check: `Genes` contains genotype codes, `Genealex` preserves GenAlEx metadata, and `pheno_raw` preserves the original phenotype spreadsheet.

```{r data-dimensions}
data.frame(
  Source = c("Genes", "Genealex", "Médias_IND_artigo"),
  Rows = c(nrow(genes), nrow(genealex), nrow(pheno_raw)),
  Columns = c(ncol(genes), ncol(genealex), ncol(pheno_raw))
) |>
  format_table(caption = "Dimensions of the spreadsheets used in the analysis")
```

The table confirms `r nrow(genes)` individuals in `Genes` and `r ncol(genes) - 1` columns corresponding to SSR loci, in addition to the first index column. The phenotype table contains `r nrow(pheno_raw)` individuals. `Genealex` has its own metadata layout, so its dimensions should not be interpreted as a marker matrix.

## 4. Organizing the SSR data

### 4.1 Individual and locus identifiers

The first column of `Genes` contains a sequential population-group index and does not represent the phenotype variable `FAM`. The remaining 35 columns correspond to the original SSR loci.

In `Genealex`, individual identifiers begin on the fourth row of the second column, after the three metadata rows. Locus names begin in the third column and occur every two columns. The numbers of individuals and loci to recover are obtained directly from the dimensions of the `Genes` sheet.

```{r ssr-identifiers}
individual_rows <- seq(
  from = 4,
  length.out = nrow(genes)
)

locus_columns <- seq(
  from = 3,
  by = 2,
  length.out = ncol(genes) - 1
)

ids <- trimws(
  as.character(genealex[[2]][individual_rows])
)

locus_names <- trimws(
  as.character(
    unlist(genealex[3, locus_columns], use.names = FALSE)
  )
)
```

Before building the SSR matrix, the identifiers recovered from the GenAlEx structure are checked. Keeping this diagnostic next to the step that creates `ids` and `locus_names` makes any inconsistency easier to interpret.

```{r ssr-identifier-checks}
ssr_identifier_checks <- data.frame(
  Check = c(
    "Missing genotype identifiers",
    "Duplicated genotype identifiers",
    "Missing locus names",
    "Duplicated locus names"
  ),
  Value = c(
    sum(is.na(ids) | ids == ""),
    sum(duplicated(ids)),
    sum(is.na(locus_names) | locus_names == ""),
    sum(duplicated(locus_names))
  )
)

format_table(
  ssr_identifier_checks,
  caption = "Check of genotype identifiers and locus names"
)
```

The checks contain `r sum(ssr_identifier_checks$Value)` problematic occurrences, confirming that individual identifiers and locus names can be used to construct the SSR matrix.

### 4.2 Building the SSR matrix

After identifiers are recovered, the first column of `Genes` is removed and the 35 marker columns form `ssr_raw`. Codes are stored numerically, and `-9`, used in the original file for missing calls, is converted to `NA`.

Thus, `ssr_raw` preserves the original genotypes while representing individuals in rows, loci in columns, and missing calls explicitly.

```{r ssr-matrix}
ssr_raw <- as.matrix(
  genes[, -1, drop = FALSE]
)

storage.mode(ssr_raw) <- "numeric"
ssr_raw[ssr_raw == -9] <- NA_real_

rownames(ssr_raw) <- ids
colnames(ssr_raw) <- locus_names

as.data.frame(ssr_raw[1:6, 1:8]) |>
  format_table(caption = "Preview of the organized SSR matrix")
```

### 4.3 Interpreting SSR codes

SSR codes represent the two alleles observed in each individual. For example, `12` denotes one copy of allele 1 and one copy of allele 2, `22` denotes two copies of allele 2, and `33` denotes two copies of allele 3.

The `33` value observed for `P3K3256CC` is not an error or a special category: this locus is multi-allelic and allele 3 is represented in the panel. In Module 02, these genotype codes are decomposed into dosages of 0, 1, or 2 copies for each observed allele.

## 5. Organizing and aligning phenotypes

This stage transforms the raw phenotype table into objects aligned with the SSR matrix. The process is presented in four parts: trait definition, individual checking and correspondence, effective alignment, and separation of metadata from numerical phenotypes.

### 5.1 Traits used in the analysis

Ten phenotype traits are used, identified by the same codes throughout all modules.

```{r define-traits}
# Define the traits used in the analysis.
TRAITS <- c(
  "MMCOM", "PRODCOM", "PRODEF", "CF", "DF",
  "DCI", "EP", "FFC", "FFP", "SS"
)
```

The dictionary links the codes used in the R objects to biological descriptions and measurement units. This is important because the ten traits are on different scales and the same codes are used throughout the modeling and results modules.

```{r trait-dictionary}
# Associate each trait with its English and Portuguese descriptions and measurement unit.
trait_dictionary <- data.frame(
  Trait = TRAITS,
  Description_en = c(
    "Mean mass of commercial fruit",
    "Production of commercial fruit",
    "Production of defective fruit",
    "Fruit length",
    "Fruit diameter",
    "Internal-cavity diameter",
    "Mean pulp thickness",
    "External fruit firmness",
    "Internal pulp firmness",
    "Soluble solids"
  ),
  Description_pt = c(
    "Massa média de frutos comerciais",
    "Produção de frutos comerciais",
    "Produção de frutos defeituosos",
    "Comprimento do fruto",
    "Diâmetro do fruto",
    "Diâmetro da cavidade interna",
    "Espessura média da polpa",
    "Firmeza externa do fruto",
    "Firmeza interna da polpa",
    "Sólidos solúveis"
  ),
  Unit = c(
    "kg", "kg", "kg", "cm", "cm",
    "cm", "cm", "N", "N", "°Brix"
  )
)

rownames(trait_dictionary) <- NULL

format_table(
  trait_dictionary[, c("Trait", "Description_en", "Unit")],
  caption = "Dictionary of phenotype traits",
  col.names = c("Trait", "Description", "Unit")
)
```

The dictionary shows that production and mass are expressed in kg, fruit dimensions in cm, firmness in N, and soluble solids in °Brix. These different units explain why the numerical scales of the traits should not be compared directly.

### 5.2 Analytical columns and correspondence between individuals

Before any rows are reordered, the columns that form the analytical table are defined. `BL` is excluded because it is not part of the statistical model. Phenotype identifiers are then checked and the sets of plants represented by both sources are compared.

```{r individual-correspondence}
# Define the analytical columns; BL is excluded because it is not part of the model.
required_columns <- c("IND", "FAM", TRAITS)

# Standardize phenotype identifiers before comparing them with genotype identifiers.
pheno_raw$IND <- trimws(as.character(pheno_raw$IND))

phenotype_checks <- data.frame(
  Check = c(
    "Missing phenotype identifiers",
    "Duplicated phenotype identifiers",
    "Missing required phenotype columns"
  ),
  Value = c(
    sum(is.na(pheno_raw$IND) | pheno_raw$IND == ""),
    sum(duplicated(pheno_raw$IND)),
    length(setdiff(required_columns, names(pheno_raw)))
  )
)

format_table(
  phenotype_checks,
  caption = "Check of phenotype identifiers and required columns"
)

alignment <- data.frame(
  Check = c(
    "Genotyped individuals",
    "Phenotyped individuals",
    "Missing from phenotype data",
    "Missing from genotype data"
  ),
  Value = c(
    length(ids),
    nrow(pheno_raw),
    length(setdiff(ids, pheno_raw$IND)),
    length(setdiff(pheno_raw$IND, ids))
  )
)

format_table(
  alignment,
  caption = "Alignment between genotype and phenotype data"
)
```

The phenotype checks contain `r sum(phenotype_checks$Value)` problematic occurrences. The alignment table confirms `r length(ids)` genotyped individuals and `r nrow(pheno_raw)` phenotyped individuals, with no plants present in only one source. Alignment can therefore proceed without discarding individuals.

### 5.3 Effective phenotype alignment

Only after the previous checks are phenotypes reordered to exactly follow the individual sequence in the SSR matrix.

```{r phenotype-alignment}
# Reorder phenotypes to the same sequence as genotyped individuals.
pheno <- pheno_raw[
  match(ids, pheno_raw$IND),
  required_columns
]

```

### 5.4 Metadata and phenotype matrix

After alignment, `metadata` retains `IND` and `FAM`, whereas `phenotypes` contains only the ten numerical traits. Family therefore remains categorical and is not mixed with numerical phenotypes.

```{r phenotype-objects}
# Separate metadata; FAM remains categorical and is not mixed with numeric phenotypes.
metadata <- data.frame(
  IND = pheno$IND,
  FAM = factor(pheno$FAM),
  row.names = pheno$IND
)

# Build the numeric phenotype matrix in the same order as genotyped individuals.
phenotypes <- as.matrix(pheno[, TRAITS])
storage.mode(phenotypes) <- "numeric"
rownames(phenotypes) <- pheno$IND

# Distribution of individuals among families.
data.frame(
  Family = names(table(metadata$FAM)),
  Individuals = as.integer(table(metadata$FAM))
) |>
  format_table(caption = "Number of individuals per family")
```

After reordering, `FAM` contains `r nlevels(metadata$FAM)` families, with family sizes ranging from `r min(table(metadata$FAM))` to `r max(table(metadata$FAM))` individuals. This distribution is used to stratify the five cross-validation folds in Module 02.

## 6. Marker quality control

### 6.1 Locus-retention criteria

Marker quality control considers two properties of each locus: the proportion of individuals with an observed genotype and the number of genotype classes represented. A locus is retained when its call rate is at least 0.90 — equivalent to 135 of 150 individuals — and at least two genotype classes are present.

These criteria distinguish missingness from lack of variation. A locus with high call rate can therefore still be removed if it is monomorphic in the panel. Filtering acts only on loci: `ssr_raw` preserves the original matrix for diagnostics, whereas `ssr` contains the loci passed to allele-dosage coding in Module 02.

```{r marker-quality-control}
# Define the call-rate threshold immediately before applying it in quality control.
CALL_RATE_MIN <- 0.90

# Summarize completeness and genotype diversity for each locus.
locus_audit <- data.frame(
  Locus = colnames(ssr_raw),
  N_called = colSums(!is.na(ssr_raw)),
  Call_rate = colMeans(!is.na(ssr_raw)),
  N_genotype_classes = apply(
    ssr_raw,
    2,
    function(genotypes) length(unique(na.omit(genotypes)))
  )
)

locus_audit$Keep_locus <-
  locus_audit$Call_rate >= CALL_RATE_MIN &
  locus_audit$N_genotype_classes >= 2

# Retain in the SSR matrix only loci that pass the quality-control criteria.
ssr <- ssr_raw[, locus_audit$Keep_locus, drop = FALSE]

format_table(
  locus_audit,
  digits = 3,
  caption = "Quality control of SSR loci",
  col.names = c("Locus", "Called genotypes", "Call rate", "Genotype classes", "Retain locus")
)
```

### 6.2 Quality-control result

Of the `r nrow(locus_audit)` loci evaluated, `r sum(locus_audit$Keep_locus)` satisfy both the call-rate and genotype-variation criteria. The removed locus is `r paste(locus_audit$Locus[!locus_audit$Keep_locus], collapse = ", ")`; its exclusion is due to insufficient genotype variation in the panel rather than removal of any individual.

### 6.3 Call-rate visualization

The figure summarizes locus completeness and the final quality-control decision together. Each bar represents the proportion of observed genotypes at one locus, and the dashed line marks the minimum call rate of 0.90. Colors follow the final retention decision: reaching 0.90 is necessary, but each locus must also contain at least two genotype classes.

```{r marker-quality-plot, fig.height=8, fig.alt="Horizontal bar plot showing call rates for the 35 SSR loci, the 90% reference line, and the final retained or removed decision for each locus."}
ggplot(
  locus_audit,
  aes(
    x = reorder(Locus, Call_rate),
    y = Call_rate,
    fill = factor(
      Keep_locus,
      levels = c(TRUE, FALSE),
      labels = c("Retained", "Removed")
    )
  )
) +
  geom_col(width = 0.8) +
  geom_hline(
    yintercept = CALL_RATE_MIN,
    linetype = 2,
    linewidth = 0.6
  ) +
  coord_flip() +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = paste0(seq(0, 100, 20), "%"),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_gdocs(name = "Decision") +
  labs(
    x = NULL,
    y = "Call rate",
    title = "Completeness and retention of SSR loci",
    subtitle = "The dashed line marks the 90% minimum call rate; retention also requires at least two genotype classes",
    caption = "Color represents the final quality-control decision, not call rate alone."
  ) +
  theme_gdocs() +
  theme(
    legend.position = "top",
    axis.text.y = element_text(size = 8)
  )
```

The figure should be read by combining bar height and color. Height reports locus completeness, whereas color reports the outcome of the two criteria applied together. A locus can therefore reach or exceed the 90% line and still be removed when only one genotype class is represented. The figure does not indicate removal of individuals; filtering acts only on loci.

## 7. Genotype completeness by individual

Genotype completeness is also summarized by individual using the 35 original loci. For each plant, the number and proportion of available SSR calls are recorded. This information is diagnostic only: no completeness threshold is used to exclude individuals, and all 150 aligned individuals remain in subsequent stages.

```{r individual-summary}
# Compute, across the original loci, the number and proportion of available calls per individual.
individual_audit <- data.frame(
  IND = rownames(ssr_raw),
  N_called = rowSums(!is.na(ssr_raw)),
  Call_rate = rowMeans(!is.na(ssr_raw))
)

rownames(individual_audit) <- NULL

data.frame(
  Measure = c("Minimum", "Mean", "Median", "Maximum"),
  # List the call-rate values used in the individual summary.
  Call_rate = c(
    min(individual_audit$Call_rate),
    mean(individual_audit$Call_rate),
    median(individual_audit$Call_rate),
    max(individual_audit$Call_rate)
  )
) |>
  format_table(
    digits = 3,
    caption = "Distribution of SSR call rate by individual",
    col.names = c("Measure", "Call rate")
  )
```

Individual call rate ranges from `r sprintf("%.1f%%", 100 * min(individual_audit$Call_rate))` to `r sprintf("%.1f%%", 100 * max(individual_audit$Call_rate))`, with a mean of `r sprintf("%.1f%%", 100 * mean(individual_audit$Call_rate))`. Because no individual-exclusion rule is applied, these differences are recorded only as genotype-completeness information.

## 8. Phenotype characterization

### 8.1 Descriptive statistics

Descriptive statistics allow direct inspection of the scale, dispersion, and range of the ten traits before prediction models are fitted. `N` also confirms phenotype availability by trait. No permanent transformation or standardization of phenotypes is performed at this stage; original values are preserved for the models.

```{r phenotype-summary}
# Directly summarize availability, location, dispersion, and range for each trait.
phenotype_summary <- data.frame(
  Trait = colnames(phenotypes),
  N = colSums(!is.na(phenotypes)),
  Mean = colMeans(phenotypes),
  SD = apply(phenotypes, 2, sd),
  Minimum = apply(phenotypes, 2, min),
  Maximum = apply(phenotypes, 2, max)
)

rownames(phenotype_summary) <- NULL

format_table(
  phenotype_summary,
  digits = 3,
  caption = "Summary of phenotype traits"
)
```

All ten traits contain `r min(phenotype_summary$N)` available observations, confirming that the analyzed phenotype variables have no missing values. The mean locates the center of each distribution, the standard deviation summarizes dispersion, and minimum and maximum delimit the observed range. Because traits use different units, these measures should be interpreted within each trait rather than compared directly across different scales.

### 8.2 Phenotype observations far from the mean

Standardized values allow the distance of each observation from the mean of its own trait to be compared independently of the original measurement unit. For each trait,

\[
z = \frac{y-\bar{y}}{s},
\]

where $y$ is the observed value, $\bar{y}$ is the mean, and $s$ is the sample standard deviation. The means and standard deviations already calculated in `phenotype_summary` are reused here, avoiding repetition of the same calculations.

Observations whose absolute standardized distance is at least four standard deviations are highlighted for inspection. This threshold is diagnostic only: no observation is removed, capped, or transformed by this criterion.

```{r distant-phenotype-values}
# Define the diagnostic threshold immediately before identifying observations distant from the mean.
PHENOTYPE_DISTANCE_THRESHOLD <- 4

# Standardize each trait using the mean and standard deviation summarized in the preceding section.
phenotype_standardized <- scale(
  phenotypes,
  center = phenotype_summary$Mean,
  scale = phenotype_summary$SD
)

# Locate observations whose standardized distance reaches the defined threshold.
extreme_positions <- which(
  abs(phenotype_standardized) >= PHENOTYPE_DISTANCE_THRESHOLD,
  arr.ind = TRUE
)

phenotype_extremes <- data.frame(
  IND = rownames(phenotypes)[extreme_positions[, "row"]],
  Trait = colnames(phenotypes)[extreme_positions[, "col"]],
  Value = phenotypes[extreme_positions],
  Trait_mean = phenotype_summary$Mean[extreme_positions[, "col"]],
  Trait_SD = phenotype_summary$SD[extreme_positions[, "col"]],
  Standardized_value = phenotype_standardized[extreme_positions],
  Absolute_standardized_distance = abs(
    phenotype_standardized[extreme_positions]
  )
)

# Sort only the diagnostic table; the `phenotypes` matrix remains unchanged.
phenotype_extremes <- phenotype_extremes[
  order(-phenotype_extremes$Absolute_standardized_distance),
]
rownames(phenotype_extremes) <- NULL

format_table(
  phenotype_extremes,
  digits = 3,
  col.names = c(
    "Individual",
    "Trait",
    "Value",
    "Trait mean",
    "Trait SD",
    "Standardized value",
    "Absolute standardized distance"
  ),
  caption = paste0(
    "Phenotype observations at least ",
    PHENOTYPE_DISTANCE_THRESHOLD,
    " standard deviations from the mean"
  )
)
```

`r nrow(phenotype_extremes)` observations are highlighted at a standardized distance of at least `r PHENOTYPE_DISTANCE_THRESHOLD` standard deviations. The table identifies the individual, trait, original value, and distance in standard-deviation units, allowing each case to be inspected without changing the data used later.

The following figure shows the same diagnostic for all 1,500 individual × trait combinations. Values near zero lie close to the mean of their respective trait; negative and positive values indicate observations below or above that mean. Because the axis is expressed in standard-deviation units, all ten traits can be compared on the same diagnostic scale.

```{r phenotype-distance-plot, fig.height=7.5, fig.alt="Point plot showing standardized phenotype values for the ten traits, with vertical lines at minus four and plus four standard deviations and highlighted observations that reach or exceed those limits."}
data.frame(
  IND = rownames(phenotypes),
  as.data.frame(phenotype_standardized, check.names = FALSE),
  check.names = FALSE
) |>
  pivot_longer(
    cols = -IND,
    names_to = "Trait",
    values_to = "Standardized_value"
  ) |>
  mutate(
    Trait = factor(Trait, levels = rev(TRAITS)),
    Diagnostic = factor(
      abs(Standardized_value) >= PHENOTYPE_DISTANCE_THRESHOLD,
      levels = c(FALSE, TRUE),
      labels = c("Within threshold", "Highlighted")
    )
  ) |>
  ggplot(
    aes(
      x = Standardized_value,
      y = Trait,
      colour = Diagnostic
    )
  ) +
  geom_vline(
    xintercept = c(
      -PHENOTYPE_DISTANCE_THRESHOLD,
      PHENOTYPE_DISTANCE_THRESHOLD
    ),
    linetype = 2,
    linewidth = 0.6
  ) +
  geom_point(
    position = position_jitter(
      width = 0,
      height = 0.16,
      seed = 2026
    ),
    alpha = 0.7,
    size = 1.6
  ) +
  scale_color_gdocs(
    name = "Diagnostic",
    drop = FALSE
  ) +
  labs(
    x = "Standardized distance from the mean (standard deviations)",
    y = NULL,
    title = "Standardized distance of phenotype observations",
    subtitle = "Dashed lines mark the diagnostic limits of -4 and +4 standard deviations",
    caption = "Each point represents one individual × trait combination; no observation is removed by this diagnostic."
  ) +
  theme_gdocs() +
  theme(legend.position = "top")
```

Most observations should concentrate around zero, whereas highlighted points occur farther from their respective trait means. The lines at -4 and +4 make the criterion used in the preceding table visually explicit. The plot is intended to locate unusual observations on a standardized scale; it is not a normality test and does not define automatic exclusions.

### 8.3 Phenotype distributions

The histograms complement the preceding table by showing distribution shape. Each panel represents one trait and identifies its measurement unit. Twenty bins are used to keep the same visual resolution across panels. Because units and ranges differ, only the horizontal axis varies freely; the vertical count scale remains common to all ten traits. The dashed line marks the mean reported in `phenotype_summary`.

```{r phenotype-plot, fig.height=10, fig.alt="Set of ten phenotype histograms, with one GDocs-scale color per trait and a dashed line marking the mean in each panel."}
phenotype_long <- data.frame(
  IND = rownames(phenotypes),
  phenotypes,
  check.names = FALSE
) |>
  pivot_longer(
    cols = -IND,
    names_to = "Trait",
    values_to = "Value"
  ) |>
  mutate(Trait = factor(Trait, levels = TRAITS))

ggplot(
  phenotype_long,
  aes(x = Value, fill = Trait)
) +
  geom_histogram(
    bins = 20,
    colour = "white",
    linewidth = 0.25
  ) +
  geom_vline(
    data = phenotype_summary,
    aes(xintercept = Mean),
    inherit.aes = FALSE,
    linetype = 2,
    linewidth = 0.6
  ) +
  facet_wrap(
    ~ Trait,
    scales = "free_x",
    ncol = 2,
    labeller = as_labeller(
      setNames(
        paste0(trait_dictionary$Trait, " (", trait_dictionary$Unit, ")"),
        trait_dictionary$Trait
      )
    )
  ) +
  scale_fill_gdocs(guide = "none") +
  labs(
    x = "Observed value",
    y = "Number of individuals",
    title = "Phenotype distributions of the ten traits",
    subtitle = "Each panel has its own horizontal scale; the dashed line marks the trait mean",
    caption = "Histograms are descriptive diagnostics and are not tests of normality."
  ) +
  theme_gdocs() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text = element_text(size = 9)
  )
```

Interpretation should be made within each panel. The width occupied by values shows observed dispersion, bar concentration indicates where most individuals lie, and the position of the mean relative to the main mass of the distribution helps visualize asymmetry. Because the vertical axis is common, differences in count concentration can also be compared across traits. These shapes are descriptive only: asymmetry or longer tails do not by themselves imply phenotype transformation or removal of observations.

## 9. Saving preprocessing results

At the end of the module, analytical objects are collected in a single RDS file and the main diagnostics are exported separately as CSV files. `preprocessing_object.rds` is the primary interface with Module 02; the CSV files provide direct access to the checks performed during preprocessing.

### 9.1 Main preprocessing object

The output directory is created only now, when results are actually written. Within `preprocessing_object`, components follow the sequence developed throughout the module: identification and aligned phenotypes, SSR matrices, diagnostics, the trait dictionary, and finally the parameters documenting preprocessing decisions.

```{r save-main-object}
dir.create(
  here("output", "preprocessing"),
  recursive = TRUE,
  showWarnings = FALSE
)

preprocessing_object <- list(
  # Identification, metadata, and aligned phenotypes.
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,

  # SSR matrices before and after locus quality control.
  ssr_raw = ssr_raw,
  ssr = ssr,

  # Diagnostics produced throughout the module.
  locus_audit = locus_audit,
  individual_audit = individual_audit,
  phenotype_summary = phenotype_summary,
  phenotype_extremes = phenotype_extremes,

  # Trait reference and parameters required to reproduce this stage.
  trait_dictionary = trait_dictionary,
  settings = list(
    call_rate_min = CALL_RATE_MIN,
    phenotype_distance_threshold = PHENOTYPE_DISTANCE_THRESHOLD,
    fixed_effects = "FAM"
  )
)

saveRDS(
  preprocessing_object,
  here("output", "preprocessing", "preprocessing_object.rds")
)
```

The RDS preserves together the objects reused in Module 02. `settings` remains at the end because it records preprocessing decisions; it documents how the objects were produced but does not represent an additional analytical table.

### 9.2 Tabular diagnostics

The checks most useful for independent inspection are also written as CSV files. Each command explicitly shows the exported object and its destination. These files do not replace the RDS as input to the next module; they make diagnostics accessible without requiring the complete object to be opened.

```{r save-diagnostics}
write.csv(
  alignment,
  here("output", "preprocessing", "alignment_summary.csv"),
  row.names = FALSE
)

write.csv(
  locus_audit,
  here("output", "preprocessing", "locus_audit.csv"),
  row.names = FALSE
)

write.csv(
  individual_audit,
  here("output", "preprocessing", "individual_audit.csv"),
  row.names = FALSE
)

write.csv(
  phenotype_summary,
  here("output", "preprocessing", "phenotype_summary.csv"),
  row.names = FALSE
)

write.csv(
  phenotype_extremes,
  here("output", "preprocessing", "phenotype_extreme_values.csv"),
  row.names = FALSE
)

write.csv(
  trait_dictionary,
  here("output", "preprocessing", "trait_dictionary.csv"),
  row.names = FALSE
)
```

Thus, `alignment_summary.csv` documents individual correspondence, `locus_audit.csv` and `individual_audit.csv` record genotype completeness, `phenotype_summary.csv` and `phenotype_extreme_values.csv` summarize phenotype diagnostics, and `trait_dictionary.csv` preserves the description and unit of the ten traits.

### 9.3 Final inventory

Before proceeding to Module 02, the final inventory collects the dimensions and counts summarizing preprocessing: number of individuals and traits, number of loci before and after quality control, number of families, missing SSR calls, and phenotype observations highlighted only for inspection.

```{r final-inventory}
data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Original SSR loci",
    "Retained SSR loci",
    "Families",
    "Missing SSR calls",
    "Highlighted phenotype observations"
  ),
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    ncol(ssr),
    nlevels(metadata$FAM),
    sum(is.na(ssr_raw)),
    nrow(phenotype_extremes)
  )
) |>
  format_table(caption = "Final preprocessing summary")
```

The summary confirms the conditions required for genomic-matrix construction: all 150 individuals remain aligned with phenotypes, quality control acts on SSR loci rather than individuals, and highlighted phenotype observations remain in the data because their identification is diagnostic only.

## 10. Next step

Module 02 converts the retained SSR loci into allele-dosage columns, constructs
the marker and genomic relationship matrices, creates the family fixed-effect
matrix, and defines the five cross-validation folds.

<div style="display:flex; justify-content:space-between; gap:1rem; margin-top:2rem; padding-top:1rem; border-top:1px solid #ddd;">
<a href="index.html">← Home</a>
<a href="02_matrices_en.html">Module 02 — Matrix construction →</a>
</div>
