---
title: "Data preprocessing for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

Genomic selection requires phenotypes and markers to represent the same
individuals in the same order. In this module, both data sources are read,
identifiers are organized, SSR missing-value codes are converted to `NA`,
phenotypes are aligned with genotypes, and SSR loci are subjected to quality
control.

The block column (`BL`) is removed because it is not part of the statistical model defined for this tutorial. Family (`FAM`) is preserved because it will be used both in the fixed-effect matrix and to stratify the cross-validation folds in Module 02. No individual is removed in this module.

Preprocessing produces three central objects for the following stages: the aligned phenotype matrix, the SSR matrix after locus quality control, and metadata containing individual and family. The diagnostics presented throughout the module verify the consistency of these objects before genomic-matrix construction.

## 2. Packages and file paths

The packages used in this module have distinct roles:

- `readxl`: reads the Excel workbooks containing genotype and phenotype data;
- `dplyr`: selects, transforms, and organizes tabular data;
- `tidyr`: reshapes data between wide and long formats;
- `ggplot2`: builds the quality-control and phenotype-distribution figures;
- `knitr`: formats tables displayed in the tutorial;
- `here`: builds project-relative file paths that do not depend on the local working directory.

```{r packages, warning=FALSE, message=FALSE}
library(readxl)  # Read Excel workbooks.
library(dplyr)   # Manipulate and transform tables.
library(tidyr)   # Reshape data between wide and long formats.
library(ggplot2) # Build figures.
library(knitr)   # Format tables and R Markdown output.
library(here)    # Build portable project-relative paths.
```

All tables in this module use `format_table()` for a consistent presentation. Tables occupy the available page width and use horizontal scrolling when their contents exceed that width. This helper changes presentation only; the underlying data are unchanged.

```{r table-formatting, include=FALSE}
format_table <- function(x, caption = NULL, digits = getOption("digits"), col.names = NA) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    col.names = col.names,
    table.attr = paste0(
      'class="table table-striped table-hover" ',
      'style="width:100%; margin-bottom:0;"'
    )
  )
  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

The paths defined below organize the input files and output directory; they are not statistical parameters.

Two control parameters are fixed:

- `CALL_RATE_MIN = 0.90`: requires at least 90% of the calls at a locus to be available. Because the panel contains 150 individuals, this corresponds to at least 135 observed genotypes per locus. Together with the requirement of at least two genotype classes, the criterion avoids retaining loci with insufficient information to represent variation.
- `PHENOTYPE_DISTANCE_THRESHOLD = 4`: flags, only for inspection, observations at least four standard deviations from the trait mean. This diagnostic rule does not remove phenotypes, cap them at a cutoff, or otherwise alter their values.

```{r paths}
# Define the two input files and the directory where preprocessing outputs are stored.
GENOTYPE_FILE <- here("data", "papaya_ssr_genotypes.xlsx")
PHENOTYPE_FILE <- here("data", "papaya_phenotypes_2024_2025.xlsx")
OUTPUT_DIR <- here("output", "preprocessing")

# Require at least 90% available calls for an SSR locus to be retained.
CALL_RATE_MIN <- 0.90
# Use four standard deviations only as a diagnostic flagging threshold.
PHENOTYPE_DISTANCE_THRESHOLD <- 4

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
```

## 3. Reading the original spreadsheets

The genotype workbook uses the `Genes` sheet for numerical SSR codes and the `Genealex` sheet for individual and locus identifiers. These two sheets are read without column names because they contain metadata rows specific to the source format; the identifier positions are handled explicitly in the next section. Phenotypes are read from the `Médias_IND_artigo` sheet, which already contains column names.

The `BL` column is removed immediately after phenotype reading so that block is not propagated into analytical objects. The first rows make the original data organization visible, and the dimensions table checks the structure of the three sources before alignment.

```{r read-data}
# Read `Genes` without column names because the sheet contains source-format metadata rows.
genes <- read_excel(
  GENOTYPE_FILE,
  sheet = "Genes",
  col_names = FALSE,
  na = c("", "NA")
)

# Read `Genealex` without column names; identifiers are extracted by position in the next section.
genealex <- read_excel(
  GENOTYPE_FILE,
  sheet = "Genealex",
  col_names = FALSE,
  na = c("", "NA")
)

# Read phenotypes with their column names and remove BL before alignment.
pheno_raw <- read_excel(
  PHENOTYPE_FILE,
  sheet = "Médias_IND_artigo",
  na = c("", "NA")
) |>
  select(-BL)

# First rows of the SSR codes.
genes[1:6, 1:6]

# First phenotype rows after removing BL.
head(pheno_raw)

data.frame(
  Source = c("Genes", "Genealex", "Médias_IND_artigo without BL"),
  Rows = c(nrow(genes), nrow(genealex), nrow(pheno_raw)),
  Columns = c(ncol(genes), ncol(genealex), ncol(pheno_raw))
) |>
  format_table(caption = "Dimensions of the spreadsheets used in the analysis")
```

The dimensions table confirms `r nrow(genes)` individuals in the `Genes` sheet and `r ncol(genes) - 1` columns corresponding to SSR loci, in addition to the first index column. The phenotype table contains `r nrow(pheno_raw)` individuals after removing `BL`. The `Genealex` sheet has its own metadata layout, so its dimensions should not be interpreted as a marker matrix.

## 4. Organizing the SSR data

The first column of `Genes` contains a sequential population-group index and does not represent the phenotype variable `FAM`; it is therefore excluded from the marker matrix. The remaining 35 columns correspond to the original SSR loci.

In `Genealex`, individual identifiers begin on row 4 of column 2, whereas locus names occur on row 3 in alternating columns beginning with column 3. The expressions `4:(3 + n_individuals)` and `seq(3, 2 + 2 * n_loci, by = 2)` translate this source-file layout directly. Identifiers are kept as text — which preserves alphanumeric codes — and surrounding spaces are removed with `trimws()`. Finally, missing SSR calls coded as `-9` are converted to `NA` so that R functions recognize them as missing data.

```{r organize-ssr}
# Obtain panel dimensions from `Genes`; its first column is not an SSR locus.
n_individuals <- nrow(genes)
n_loci <- ncol(genes) - 1

# Translate the `Genealex` layout: individuals start on row 4 and loci occupy alternating columns.
individual_rows <- 4:(3 + n_individuals)
locus_columns <- seq(3, 2 + 2 * n_loci, by = 2)

# Keep identifiers as text to preserve alphanumeric codes.
ids <- trimws(
  as.character(genealex[[2]][individual_rows])
)

locus_names <- trimws(
  as.character(
    unlist(genealex[3, locus_columns], use.names = FALSE)
  )
)

ssr_raw <- as.matrix(
  genes[, -1, drop = FALSE]
)

storage.mode(ssr_raw) <- "numeric"
# Convert the original missing-value code to R's standard missing value.
ssr_raw[ssr_raw == -9] <- NA_real_

rownames(ssr_raw) <- ids
colnames(ssr_raw) <- locus_names

# Initial structure of the SSR matrix.
ssr_raw[1:6, 1:8]
```

SSR codes represent the two alleles observed in each individual. For example, `12` denotes one copy of allele 1 and one copy of allele 2, `22` denotes two copies of allele 2, and `33` denotes two copies of allele 3. The `33` value shown for `P3K3256CC` is not an error or a special category: this locus is multi-allelic and allele 3 is represented in the panel. In Module 02, these genotype codes are decomposed into dosages of 0, 1, or 2 copies for each observed allele.

## 5. Organizing and aligning phenotypes

Ten traits are used. Before the final objects are built, the module checks whether required columns are present, searches for missing or duplicated identifiers, and compares the sets of genotyped and phenotyped individuals. For the canonical dataset, these checks should indicate complete correspondence between the two sources.

The phenotype table is then reordered with `match()` according to the SSR identifiers so that every phenotype row corresponds to the same plant in the genotype matrix. `FAM` is converted to a factor to represent the categorical variable used in Module 02, while the ten traits are collected in a numeric matrix.

```{r organize-phenotypes}
# Define the traits, align phenotypes to genotyped individuals, and build the phenotype objects.
TRAITS <- c(
  "MMCOM", "PRODCOM", "PRODEF", "CF", "DF",
  "DCI", "EP", "FFC", "FFP", "SS"
)

# Define the columns that must be present before alignment.
required_columns <- c("IND", "FAM", TRAITS)
missing_columns <- setdiff(required_columns, names(pheno_raw))

pheno_raw$IND <- trimws(as.character(pheno_raw$IND))

# Compare identifier sets in both sources before reordering rows.
missing_in_pheno <- setdiff(ids, pheno_raw$IND)
missing_in_geno <- setdiff(pheno_raw$IND, ids)

alignment <- data.frame(
  Check = c(
    "Genotyped individuals",
    "Phenotyped individuals",
    "Missing from phenotype data",
    "Missing from genotype data"
  ),
  # List the values corresponding to the table entries.
  Value = c(
    length(ids),
    nrow(pheno_raw),
    length(missing_in_pheno),
    length(missing_in_geno)
  )
)

format_table(
  alignment,
  caption = "Alignment between genotype and phenotype data"
)

# Summarize the results in `identifier_summary`.
identifier_summary <- data.frame(
  Check = c(
    "Missing genotype identifiers",
    "Duplicated genotype identifiers",
    "Missing locus names",
    "Duplicated locus names",
    "Missing phenotype identifiers",
    "Duplicated phenotype identifiers",
    "Missing required phenotype columns"
  ),
  # List the values corresponding to the table entries.
  Value = c(
    sum(is.na(ids) | ids == ""),
    sum(duplicated(ids)),
    sum(is.na(locus_names) | locus_names == ""),
    sum(duplicated(locus_names)),
    sum(is.na(pheno_raw$IND) | pheno_raw$IND == ""),
    sum(duplicated(pheno_raw$IND)),
    length(missing_columns)
  )
)

format_table(
  identifier_summary,
  caption = "Summary of identifiers and required columns"
)

# Reorder phenotypes to the same sequence as genotyped individuals.
pheno <- pheno_raw[
  match(ids, pheno_raw$IND),
  required_columns
]

# Explicitly convert the ten traits to numeric values.
for (trait in TRAITS) {
  pheno[[trait]] <- as.numeric(
    as.character(pheno[[trait]])
  )
}

# Separate metadata; FAM remains categorical and is not mixed with numeric phenotypes.
metadata <- data.frame(
  IND = pheno$IND,
  FAM = factor(pheno$FAM),
  row.names = pheno$IND
)

# Build the numeric phenotype matrix in the same order as genotyped individuals.
phenotypes <- as.matrix(pheno[, TRAITS])
storage.mode(phenotypes) <- "numeric"
rownames(phenotypes) <- pheno$IND

# Distribution of individuals among families.
data.frame(
  Family = names(table(metadata$FAM)),
  Individuals = as.integer(table(metadata$FAM))
) |>
  format_table(caption = "Number of individuals per family")
```

The checks confirm `r length(ids)` individuals in both sources, with no individuals missing between genotype and phenotype data. Missing or duplicated identifiers and locus names are also evaluated before alignment. After reordering, `FAM` contains `r nlevels(metadata$FAM)` families, with family sizes ranging from `r min(table(metadata$FAM))` to `r max(table(metadata$FAM))` individuals.

## 6. Marker quality control

For each locus, the call rate — the proportion of the 150 individuals with an observed genotype — and the number of distinct genotype classes are calculated. A locus is retained when its call rate is at least 0.90 (at least 135 observed calls) and at least two genotype classes are present. The second criterion removes loci with no observable variation in the panel even when their call rate is high.

The table makes it possible to identify which criterion determined retention or removal for each locus. The `ssr` matrix created at the end contains only loci passed to allele-dosage coding in Module 02; `ssr_raw` is retained for diagnostics of the original source.

```{r marker-quality-control}
# Compute call rate and genotype-class diversity to decide which SSR loci are retained.
genotype_classes <- integer(ncol(ssr_raw))

for (j in seq_len(ncol(ssr_raw))) {
  # Count the genotype classes observed at each locus.
  genotype_classes[j] <- length(
    unique(na.omit(ssr_raw[, j]))
  )
}

locus_audit <- data.frame(
  Locus = colnames(ssr_raw),
  N_called = colSums(!is.na(ssr_raw)),
  Call_rate = colMeans(!is.na(ssr_raw)),
  N_genotype_classes = genotype_classes
)

rownames(locus_audit) <- NULL

locus_audit$Keep_locus <-
  locus_audit$Call_rate >= CALL_RATE_MIN &
  locus_audit$N_genotype_classes >= 2

# Retain in the SSR matrix only loci that pass the quality-control criteria.
ssr <- ssr_raw[
  ,
  locus_audit$Keep_locus,
  drop = FALSE
]

format_table(
  locus_audit,
  digits = 3,
  caption = "Quality control of SSR loci"
)
```

Of the `r nrow(locus_audit)` loci evaluated, `r sum(locus_audit$Keep_locus)` satisfy both the call-rate and genotype-variation criteria. The removed locus is `r paste(locus_audit$Locus[!locus_audit$Keep_locus], collapse = ", ")`; its exclusion is due to insufficient genotype variation in the panel rather than removal of any individual.

The figure orders loci by call rate and places a horizontal line at 0.90. Color reports the final retention decision, which also depends on the number of genotype classes; therefore, a locus may lie above the call-rate line and still be removed because it lacks genotype variation.

```{r marker-quality-plot, fig.height=7}
# Build the locus call-rate figure and highlight the minimum quality-control threshold.
ggplot(
  locus_audit,
  aes(
    x = reorder(Locus, Call_rate),
    y = Call_rate,
    fill = Keep_locus
  )
) +
  geom_col() +
  geom_hline(
    # Add the horizontal line representing the minimum call-rate threshold.
    yintercept = CALL_RATE_MIN,
    linetype = 2
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Call rate",
    fill = "Retained",
    title = "Call rate of SSR loci"
  ) +
  theme_bw()
```

The figure shows that retention does not depend only on position relative to the 0.90 line. A locus can have high call rate and still be excluded when only one genotype class is represented, as occurs for the locus identified in the preceding table.

## 7. Summary of genotype information

The amount of available SSR information is summarized for every individual using the 35 original loci (`ssr_raw`). This diagnostic describes completeness of the received panel before locus filtering. No individual-exclusion threshold is applied in this tutorial; all 150 aligned individuals are preserved for subsequent stages.

```{r individual-summary}
# Compute, across the original loci, the number and proportion of available calls per individual.
individual_audit <- data.frame(
  IND = rownames(ssr_raw),
  N_called = rowSums(!is.na(ssr_raw)),
  Call_rate = rowMeans(!is.na(ssr_raw))
)

rownames(individual_audit) <- NULL

individual_call_rate_summary <- data.frame(
  Measure = c("Minimum", "Mean", "Median", "Maximum"),
  # List the call-rate values used in the individual summary.
  Call_rate = c(
    min(individual_audit$Call_rate),
    mean(individual_audit$Call_rate),
    median(individual_audit$Call_rate),
    max(individual_audit$Call_rate)
  )
)

format_table(
  individual_call_rate_summary,
  digits = 3,
  caption = "Distribution of SSR call rate by individual"
)
```

Individual call rate ranges from `r sprintf("%.1f%%", 100 * min(individual_audit$Call_rate))` to `r sprintf("%.1f%%", 100 * max(individual_audit$Call_rate))`, with a mean of `r sprintf("%.1f%%", 100 * mean(individual_audit$Call_rate))`. Because no individual-exclusion rule is applied, these differences are recorded only as genotype-completeness information.

## 8. Summary of phenotype traits

Descriptive statistics allow direct inspection of the scale, dispersion, and range of the ten traits before prediction models are fitted. `N` also confirms phenotype availability by trait. No permanent transformation or standardization of phenotypes is performed at this stage; original values are preserved for the models.

```{r phenotype-summary}
# Compute descriptive statistics to characterize phenotype scale and variability.
phenotype_sd <- numeric(ncol(phenotypes))
phenotype_minimum <- numeric(ncol(phenotypes))
phenotype_maximum <- numeric(ncol(phenotypes))

for (j in seq_len(ncol(phenotypes))) {
  phenotype_sd[j] <- sd(phenotypes[, j])
  phenotype_minimum[j] <- min(phenotypes[, j])
  phenotype_maximum[j] <- max(phenotypes[, j])
}

# Summarize the results in `phenotype_summary`.
phenotype_summary <- data.frame(
  Trait = colnames(phenotypes),
  N = colSums(!is.na(phenotypes)),
  Mean = colMeans(phenotypes),
  SD = phenotype_sd,
  Minimum = phenotype_minimum,
  Maximum = phenotype_maximum
)

rownames(phenotype_summary) <- NULL

format_table(
  phenotype_summary,
  digits = 3,
  caption = "Summary of phenotype traits"
)
```

All ten traits contain `r min(phenotype_summary$N)` available observations, confirming that the analyzed phenotype variables have no missing values. Because traits use different units and ranges, standard deviations and intervals should be interpreted within each trait rather than compared directly across different measurement scales.

### 8.1 Phenotype observations far from the mean

Standardized values are used only as a descriptive diagnostic. For each trait, the module calculates $z=(y-\bar{y})/s$, where $\bar{y}$ is the mean and $s$ is the sample standard deviation. An observation is highlighted when $|z|$ is at least `r PHENOTYPE_DISTANCE_THRESHOLD`. This threshold highlights values far from the mean; it does not declare errors or statistical outliers. Values are retained exactly as recorded in the original spreadsheet and remain available to the models.

```{r distant-phenotype-values}
# Standardize phenotypes and identify observations beyond the defined distance threshold.
phenotype_means <- colMeans(phenotypes)
phenotype_standard_deviations <- numeric(ncol(phenotypes))

for (j in seq_len(ncol(phenotypes))) {
  phenotype_standard_deviations[j] <- sd(phenotypes[, j])
}

phenotype_standardized <- scale(
  phenotypes,
  center = phenotype_means,
  scale = phenotype_standard_deviations
)

# Locate observations whose standardized distance reaches the defined threshold.
extreme_positions <- which(
  abs(phenotype_standardized) >= PHENOTYPE_DISTANCE_THRESHOLD,
  arr.ind = TRUE
)

phenotype_extremes <- data.frame(
  IND = rownames(phenotypes)[extreme_positions[, "row"]],
  Trait = colnames(phenotypes)[extreme_positions[, "col"]],
  Value = phenotypes[extreme_positions],
  Trait_mean = phenotype_means[extreme_positions[, "col"]],
  Trait_SD = phenotype_standard_deviations[extreme_positions[, "col"]],
  Standardized_value = phenotype_standardized[extreme_positions],
  Absolute_standardized_distance = abs(
    phenotype_standardized[extreme_positions]
  )
)

# Sort only the diagnostic table; the `phenotypes` matrix remains unchanged.
phenotype_extremes <- phenotype_extremes[
  order(-phenotype_extremes$Absolute_standardized_distance),
]
rownames(phenotype_extremes) <- NULL

format_table(
  phenotype_extremes,
  digits = 3,
  # Define the headers displayed in the table of highlighted phenotype observations.
  col.names = c(
    "Individual",
    "Trait",
    "Value",
    "Trait mean",
    "Trait SD",
    "Standardized value",
    "Absolute standardized distance"
  ),
  caption = paste0(
    "Phenotype observations at least ",
    PHENOTYPE_DISTANCE_THRESHOLD,
    " standard deviations from the mean"
  )
)
```

`r nrow(phenotype_extremes)` observations are highlighted at a standardized distance of at least `r PHENOTYPE_DISTANCE_THRESHOLD` standard deviations. These observations remain in the data and in the models; the table only documents values that merit additional inspection.

### 8.2 Trait dictionary

The dictionary links the codes used in the R objects to biological descriptions and measurement units. This is important because the ten traits are on different scales and the same codes are used throughout the modeling and results modules.

```{r trait-dictionary}
# Associate each trait with its English and Portuguese descriptions and measurement unit.
trait_dictionary <- data.frame(
  Trait = TRAITS,
  Description_en = c(
    "Mean mass of commercial fruit",
    "Production of commercial fruit",
    "Production of defective fruit",
    "Fruit length",
    "Fruit diameter",
    "Internal-cavity diameter",
    "Mean pulp thickness",
    "External fruit firmness",
    "Internal pulp firmness",
    "Soluble solids"
  ),
  # Record the Portuguese description of each trait.
  Description_pt = c(
    "Massa média de frutos comerciais",
    "Produção de frutos comerciais",
    "Produção de frutos defeituosos",
    "Comprimento do fruto",
    "Diâmetro do fruto",
    "Diâmetro da cavidade interna",
    "Espessura média da polpa",
    "Firmeza externa do fruto",
    "Firmeza interna da polpa",
    "Sólidos solúveis"
  ),
  # Record the measurement unit of each trait.
  Unit = c(
    "kg", "kg", "kg", "cm", "cm",
    "cm", "cm", "N", "N", "°Brix"
  )
)

rownames(trait_dictionary) <- NULL

format_table(
  trait_dictionary[, c("Trait", "Description_en", "Unit")],
  caption = "Dictionary of phenotype traits"
)
```

The dictionary shows that production and mass are expressed in kg, fruit dimensions in cm, firmness in N, and soluble solids in °Brix. These different units explain why the numerical scales of the traits should not be compared directly.

The histograms complement the descriptive statistics by showing distribution shape. Twenty bins are used only to provide consistent visual resolution. Because `scales = "free"`, every trait receives its own horizontal scale; this makes within-trait shape easier to see, but axis limits should not be compared directly across panels.

```{r phenotype-plot, fig.height=8}
# Reshape the phenotype matrix to long format before plotting trait distributions.
phenotype_long <- data.frame(
  IND = rownames(phenotypes),
  phenotypes,
  check.names = FALSE
) |>
  pivot_longer(
    cols = -IND,
    names_to = "Trait",
    values_to = "Value"
  )

# Build the figure used for interpretation.
ggplot(phenotype_long, aes(x = Value)) +
  # Use 20 bins only to standardize the visual resolution of the histograms.
  geom_histogram(bins = 20) +
  # Give each trait its own horizontal scale because units and ranges differ.
  facet_wrap(~ Trait, scales = "free", ncol = 2) +
  labs(
    x = "Observed value",
    y = "Number of individuals",
    title = "Phenotype distributions"
  ) +
  theme_bw()
```

The histograms allow concentration, dispersion, and asymmetry to be examined within each trait. They are descriptive diagnostics; they are not normality tests and do not imply automatic phenotype transformation.

## 9. Saving preprocessing results

The aligned phenotype matrix, SSR data, metadata, trait dictionary, and diagnostics are gathered in `preprocessing_object.rds`. This RDS file is the primary interface with Module 02 because it preserves object structures, names, and classes together.

The CSV files written in parallel are inspection-oriented tabular outputs: they allow alignment, locus quality, individual completeness, phenotype summaries, highlighted observations, and the trait dictionary to be examined without opening the RDS object. Parameters defining this stage are also recorded in `settings`.

```{r save-results}
# Collect settings, results, and diagnostics in the final module object.
preprocessing_object <- list(
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,
  ssr_raw = ssr_raw,
  ssr = ssr,
  locus_audit = locus_audit,
  individual_audit = individual_audit,
  phenotype_summary = phenotype_summary,
  phenotype_extremes = phenotype_extremes,
  trait_dictionary = trait_dictionary,
  # Record parameters and thresholds required to reproduce this stage.
  settings = list(
    call_rate_min = CALL_RATE_MIN,
    phenotype_distance_threshold = PHENOTYPE_DISTANCE_THRESHOLD,
    fixed_effects = "FAM"
  )
)

# Save the RDS object that is read directly by Module 02.
saveRDS(
  preprocessing_object,
  file.path(OUTPUT_DIR, "preprocessing_object.rds")
)

# Export tabular diagnostics as CSV files for inspection independently of the RDS object.
write.csv(alignment, file.path(OUTPUT_DIR, "alignment_summary.csv"), row.names = FALSE)
write.csv(locus_audit, file.path(OUTPUT_DIR, "locus_audit.csv"), row.names = FALSE)
write.csv(individual_audit, file.path(OUTPUT_DIR, "individual_audit.csv"), row.names = FALSE)
write.csv(phenotype_summary, file.path(OUTPUT_DIR, "phenotype_summary.csv"), row.names = FALSE)
write.csv(
  phenotype_extremes,
  file.path(OUTPUT_DIR, "phenotype_extreme_values.csv"),
  row.names = FALSE
)
write.csv(trait_dictionary, file.path(OUTPUT_DIR, "trait_dictionary.csv"), row.names = FALSE)
```

The final inventory collects the dimensions that should be checked before continuing: numbers of individuals and traits, loci before and after quality control, families, missing SSR calls, and phenotype observations highlighted only for inspection.

```{r final-summary}
# Summarize final dimensions and counts of missing or highlighted observations from preprocessing.
data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Original SSR loci",
    "Retained SSR loci",
    "Families",
    "Missing SSR calls",
    "Highlighted phenotype observations"
  ),
  # List the values corresponding to the table entries.
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    ncol(ssr),
    nlevels(metadata$FAM),
    sum(is.na(ssr_raw)),
    nrow(phenotype_extremes)
  )
) |>
  format_table(caption = "Final preprocessing summary")
```

This summary should be read as a workflow checkpoint: individuals and phenotypes remain aligned, filtering acts on SSR loci rather than individuals, and the count of highlighted phenotype observations is diagnostic rather than an automatic exclusion rule.

## 10. Next step

Module 02 converts the retained SSR loci into allele-dosage columns, constructs
the marker and genomic relationship matrices, creates the family fixed-effect
matrix, and defines the five cross-validation folds.

<div style="display:flex; justify-content:space-between; gap:1rem; margin-top:2rem; padding-top:1rem; border-top:1px solid #ddd;">
<a href="index.html">← Home</a>
<a href="02_matrices_en.html">Module 02 — Matrix construction →</a>
</div>
