---
title: "1. Data audit and preprocessing"
author: "Weverton Gomes da Costa"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
bibliography: ../REFERENCES.bib
link-citations: true
---

[Português](01_preprocessing_pt.html) | [Home](index.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 5)
source(file.path("code", "00_config.R"))
source(file.path("code", "01_data_functions.R"))
source(file.path("code", "03_reporting_functions.R"))
configuracao <- obter_configuracao()
```

## Purpose

This page establishes the data contract before any model is fitted. It reads
the source workbooks, links genotype and phenotype rows by the `Genealex`
individual identifier, audits SSR calls, expands multi-allelic genotypes into
allele dosages, and fixes the eight cross-validation folds. The object is
written to `output/preprocessed_papaya_ssr.rds` for all downstream pages.

```{r preprocess-data}
objeto <- executar_preprocessamento(configuracao)
```

## Source-data inventory

The canonical phenotype table is `Médias_IND_artigo`. The genotype table has
35 SSR loci; missing calls are encoded as `-9` in the source and converted to
`NA`. IDs are held as character values so that `144B` is never coerced or
discarded.

```{r inventory-table, results='asis'}
inventario <- data.frame(
  Item = c(
    "Individuals", "Unique IDs", "SSR loci", "Missing SSR calls",
    "Missing-call rate", "Phenotypic traits", "Missing phenotypes",
    "Retained allele-dosage columns"
  ),
  Value = c(
    length(objeto$ids),
    length(unique(objeto$ids)),
    ncol(objeto$codigos_ssr),
    sum(is.na(objeto$codigos_ssr)),
    sprintf("%.2f%%", 100 * mean(is.na(objeto$codigos_ssr))),
    ncol(objeto$fenotipos),
    sum(is.na(objeto$fenotipos)),
    ncol(objeto$dosagens_ssr)
  )
)
print(tabela_relatorio(inventario, "Verified source-data inventory", digitos = 0))
```

The ten responses cover production, fruit dimensions, firmness, and soluble
solids. These codes and units are used consistently throughout the website.

```{r trait-dictionary-en, results='asis'}
dicionario_en <- configuracao$dicionario_caracteristicas |>
  dplyr::select(Trait, Description = Description_en, Unit)
print(tabela_relatorio(dicionario_en, "Phenotypic traits"))
```

## SSR encoding and quality control

Each genotype code stores two allele labels. For every locus and observed
allele, the pipeline creates a dosage column with values 0, 1, or 2. This
preserves four triallelic loci and avoids treating SSRs as biallelic SNPs.
Loci must have at least 90% calls and more than one observed genotype class;
allele columns must have frequencies between 0.05 and 0.95 and non-zero
variance. `P6K1129CC` is removed because it contains one genotype class.

```{r marker-qc-table, results='asis'}
qc_locos <- objeto$auditoria$locos |>
  dplyr::mutate(Call_rate = round(Call_rate, 3)) |>
  dplyr::arrange(Keep_locus, Call_rate)
print(tabela_relatorio(qc_locos, "Quality control by SSR locus", rolagem = TRUE))
```

Call rate is inspected across every locus. The horizontal reference line is
the inclusive 0.90 threshold used by the code.

```{r marker-call-rate-plot, echo=FALSE}
ggplot2::ggplot(
  objeto$auditoria$locos,
  ggplot2::aes(x = stats::reorder(Locus, Call_rate), y = Call_rate, fill = Keep_locus)
) +
  ggplot2::geom_col() +
  ggplot2::geom_hline(yintercept = configuracao$taxa_chamada_minima, linetype = 2) +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(values = c("TRUE" = "#2C7FB8", "FALSE" = "#D95F0E")) +
  ggplot2::labs(
    x = NULL,
    y = "Call rate",
    fill = "Retained",
    title = "SSR call rate by locus"
  ) +
  tema_relatorio()
```

## Phenotype overview

The canonical phenotype matrix has no missing observations. The table below
provides a scale check before model fitting; no trait is standardized because
performance metrics are reported in each trait's original unit.

```{r phenotype-summary, results='asis'}
resumo_fenotipos <- data.frame(
  Trait = colnames(objeto$fenotipos),
  N = colSums(is.finite(objeto$fenotipos)),
  Mean = colMeans(objeto$fenotipos),
  SD = apply(objeto$fenotipos, 2L, stats::sd),
  Minimum = apply(objeto$fenotipos, 2L, min),
  Maximum = apply(objeto$fenotipos, 2L, max),
  row.names = NULL
)
print(tabela_relatorio(resumo_fenotipos, "Phenotype summary"))
```

Distributions are displayed separately because their measurement scales
differ substantially.

```{r phenotype-distributions, echo=FALSE, fig.height=8}
fenotipos_longos <- data.frame(IND = rownames(objeto$fenotipos), objeto$fenotipos) |>
  tidyr::pivot_longer(-IND, names_to = "Trait", values_to = "Value")

ggplot2::ggplot(fenotipos_longos, ggplot2::aes(x = Value)) +
  ggplot2::geom_histogram(bins = 20, fill = "#2C7FB8", color = "white") +
  ggplot2::facet_wrap(~ Trait, scales = "free", ncol = 2) +
  ggplot2::labs(x = "Observed value", y = "Individuals", title = "Phenotypic distributions") +
  tema_relatorio()
```

## Validation folds and leakage control

The deterministic folds are stratified by phenotypic family. Every individual
is held out once, fold sizes differ by at most one, and every training split
retains all family levels. Marker means, missing-value imputation, variable-
column checks, centering, and the genomic relationship matrix are recalculated
using training individuals within each fold.

```{r fold-summary, results='asis'}
resumo_folds <- data.frame(
  Fold = seq_len(configuracao$n_folds),
  Validation_individuals = tabulate(objeto$folds, nbins = configuracao$n_folds)
)
print(tabela_relatorio(resumo_folds, "Fixed eight-fold allocation", digitos = 0))
```

The fixed-effect design is `block + family`. Its full-rank condition is checked
inside every training fold before a model is fitted.

## Output contract

Downstream pages receive the aligned phenotypes, fixed-effect matrices,
quality-controlled allele dosages, common fold labels, and fold-specific
training transformations. This makes the Bayesian, RR-BLUP, and GBLUP
comparisons use exactly the same validation observations.

[Next: Bayesian models](02_bayesian_models_en.html)
