---
title: "Data preprocessing for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

Genomic selection requires phenotypes and markers to represent the same
individuals in the same order. In this module, both data sources are read,
identifiers are organized, SSR missing-value codes are converted to `NA`,
phenotypes are aligned with genotypes, and SSR loci are subjected to quality
control.

The block column (`BL`) is removed because it is not part of the statistical model adopted in this analysis. Family (`FAM`) is preserved because it will be used both in the fixed-effect matrix and to stratify the cross-validation folds in Module 02. No individual is removed in this module.

Preprocessing produces three central objects for the following stages: the aligned phenotype matrix, the SSR matrix after locus quality control, and metadata containing individual and family. The diagnostics verify the consistency of these objects before genomic-matrix construction.

## 2. Packages, files, and parameters

The packages used in this module have distinct roles:

- `readxl`: reads the Excel workbooks containing genotype and phenotype data;
- `dplyr`: selects, transforms, and organizes tabular data;
- `tidyr`: reshapes data between wide and long formats;
- `ggplot2`: builds the quality-control and phenotype-distribution figures;
- `knitr`: formats tables in R Markdown documents;
- `here`: builds project-relative file paths that do not depend on the local working directory.

```{r packages, warning=FALSE, message=FALSE}
library(readxl)  # Read Excel workbooks.
library(dplyr)   # Manipulate and transform tables.
library(tidyr)   # Reshape data between wide and long formats.
library(ggplot2) # Build figures.
library(knitr)   # Format tables and R Markdown output.
library(here)    # Build portable project-relative paths.
```

```{r table-formatting, include=FALSE}
format_table <- function(x, caption = NULL, digits = getOption("digits"), col.names = NA) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    col.names = col.names,
    table.attr = paste0(
      'class="table table-striped table-hover" ',
      'style="width:100%; margin-bottom:0;"'
    )
  )
  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

Two control parameters are fixed:

- `CALL_RATE_MIN = 0.90`: requires at least 90% of the calls at a locus to be available. Because the panel contains 150 individuals, this corresponds to at least 135 observed genotypes per locus. Together with the requirement of at least two genotype classes, the criterion avoids retaining loci with insufficient information to represent variation.
- `PHENOTYPE_DISTANCE_THRESHOLD = 4`: flags, only for inspection, observations at least four standard deviations from the trait mean. This diagnostic rule does not remove phenotypes, cap them at a cutoff, or otherwise alter their values.

```{r paths}
# Define the two input files and the directory where preprocessing outputs are stored.
GENOTYPE_FILE <- here("data", "papaya_ssr_genotypes.xlsx")
PHENOTYPE_FILE <- here("data", "papaya_phenotypes_2024_2025.xlsx")
OUTPUT_DIR <- here("output", "preprocessing")

# Require at least 90% available calls for an SSR locus to be retained.
CALL_RATE_MIN <- 0.90
# Use four standard deviations only as a diagnostic flagging threshold.
PHENOTYPE_DISTANCE_THRESHOLD <- 4

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
```

## 3. Reading the original spreadsheets

### 3.1 Genotype data

Genotype information is concentrated in two sheets of the genotype workbook. The `Genes` sheet contains the numerical SSR codes: its first column is an index and the next 35 columns correspond to the original loci. Because this sheet has no row containing column names, it is read with `col_names = FALSE`.

The `Genealex` sheet provides individual identifiers and locus names. It is also read without column names because it preserves metadata rows and the original GenAlEx layout. The positions containing these identifiers are handled explicitly in the next section.

```{r read-genotypes}
genes <- read_excel(
  GENOTYPE_FILE,
  sheet = "Genes",
  col_names = FALSE,
  na = c("", "NA")
)

genealex <- read_excel(
  GENOTYPE_FILE,
  sheet = "Genealex",
  col_names = FALSE,
  na = c("", "NA")
)
```

The initial structure of `Genes` can be inspected in its first six rows and six columns:

```{r genes-preview}
as.data.frame(genes[1:6, 1:6]) |>
  format_table(caption = "Preview of the Genes sheet")
```

The preview shows the raw genotype layout: the first column is a sequential index, and the following columns contain the SSR codes organized in the next section.

### 3.2 Phenotype data

Phenotype data are stored in the `Médias_IND_artigo` sheet, which already contains column names. This source includes individual identifiers, family membership, and the evaluated traits.

The `BL` column, corresponding to experimental block, is removed immediately after reading because it is not part of the statistical model used here. `IND`, `FAM`, and the ten traits remain available for alignment with the genotype data.

```{r read-phenotypes}
pheno_raw <- read_excel(
  PHENOTYPE_FILE,
  sheet = "Médias_IND_artigo",
  na = c("", "NA")
) |>
  select(-BL)
```

The phenotype table is inspected after `BL` is removed:

```{r phenotype-preview}
head(pheno_raw) |>
  format_table(caption = "Preview of phenotype data")
```

The preview contains `IND`, `FAM`, and the ten traits used in the analysis. The absence of `BL` confirms that this column was removed before genotype–phenotype alignment.

### 3.3 Initial check of the data sources

Before organization and alignment, the dimensions of the three imported structures are compared. This is a structural check: `Genes` contains genotype codes, `Genealex` preserves GenAlEx metadata, and `pheno_raw` contains phenotypes.

```{r data-dimensions}
data.frame(
  Source = c("Genes", "Genealex", "Médias_IND_artigo without BL"),
  Rows = c(nrow(genes), nrow(genealex), nrow(pheno_raw)),
  Columns = c(ncol(genes), ncol(genealex), ncol(pheno_raw))
) |>
  format_table(caption = "Dimensions of the spreadsheets used in the analysis")
```

The table confirms `r nrow(genes)` individuals in `Genes` and `r ncol(genes) - 1` columns corresponding to SSR loci, in addition to the first index column. The phenotype table contains `r nrow(pheno_raw)` individuals. `Genealex` has its own metadata layout, so its dimensions should not be interpreted as a marker matrix.

## 4. Organizing the SSR data

### 4.1 Individual and locus identifiers

The first column of `Genes` contains a sequential population-group index and does not represent the phenotype variable `FAM`. The remaining 35 columns correspond to the original SSR loci.

In `Genealex`, individual identifiers and locus names occupy fixed positions in the GenAlEx layout. These positions are recovered first, before the genotype matrix is constructed, so that rows and columns can receive the correct identifiers.

```{r ssr-identifiers}
n_individuals <- nrow(genes)
n_loci <- ncol(genes) - 1

individual_rows <- 4:(3 + n_individuals)
locus_columns <- seq(3, 2 + 2 * n_loci, by = 2)

ids <- trimws(
  as.character(genealex[[2]][individual_rows])
)

locus_names <- trimws(
  as.character(
    unlist(genealex[3, locus_columns], use.names = FALSE)
  )
)
```

### 4.2 Building the SSR matrix

After identifiers are recovered, the first column of `Genes` is removed and the 35 marker columns form `ssr_raw`. Codes are stored numerically, and `-9`, used in the original file for missing calls, is converted to `NA`.

Thus, `ssr_raw` preserves the original genotypes while representing individuals in rows, loci in columns, and missing calls explicitly.

```{r ssr-matrix}
ssr_raw <- as.matrix(
  genes[, -1, drop = FALSE]
)

storage.mode(ssr_raw) <- "numeric"
ssr_raw[ssr_raw == -9] <- NA_real_

rownames(ssr_raw) <- ids
colnames(ssr_raw) <- locus_names

as.data.frame(ssr_raw[1:6, 1:8]) |>
  format_table(caption = "Preview of the organized SSR matrix")
```

### 4.3 Interpreting SSR codes

SSR codes represent the two alleles observed in each individual. For example, `12` denotes one copy of allele 1 and one copy of allele 2, `22` denotes two copies of allele 2, and `33` denotes two copies of allele 3.

The `33` value observed for `P3K3256CC` is not an error or a special category: this locus is multi-allelic and allele 3 is represented in the panel. In Module 02, these genotype codes are decomposed into dosages of 0, 1, or 2 copies for each observed allele.

## 5. Organizing and aligning phenotypes

This stage transforms the raw phenotype table into objects aligned with the SSR matrix. The process is presented in five parts: trait definition, individual correspondence, identifier checks, effective alignment, and separation of metadata from numerical phenotypes.

### 5.1 Traits used in the analysis

Ten phenotype traits are used, identified by the same codes throughout all modules.

```{r define-traits}
# Define the traits used in the analysis.
TRAITS <- c(
  "MMCOM", "PRODCOM", "PRODEF", "CF", "DF",
  "DCI", "EP", "FFC", "FFP", "SS"
)
```

The dictionary links the codes used in the R objects to biological descriptions and measurement units. This is important because the ten traits are on different scales and the same codes are used throughout the modeling and results modules.

```{r trait-dictionary}
# Associate each trait with its English and Portuguese descriptions and measurement unit.
trait_dictionary <- data.frame(
  Trait = TRAITS,
  Description_en = c(
    "Mean mass of commercial fruit",
    "Production of commercial fruit",
    "Production of defective fruit",
    "Fruit length",
    "Fruit diameter",
    "Internal-cavity diameter",
    "Mean pulp thickness",
    "External fruit firmness",
    "Internal pulp firmness",
    "Soluble solids"
  ),
  Description_pt = c(
    "Massa média de frutos comerciais",
    "Produção de frutos comerciais",
    "Produção de frutos defeituosos",
    "Comprimento do fruto",
    "Diâmetro do fruto",
    "Diâmetro da cavidade interna",
    "Espessura média da polpa",
    "Firmeza externa do fruto",
    "Firmeza interna da polpa",
    "Sólidos solúveis"
  ),
  Unit = c(
    "kg", "kg", "kg", "cm", "cm",
    "cm", "cm", "N", "N", "°Brix"
  )
)

rownames(trait_dictionary) <- NULL

format_table(
  trait_dictionary[, c("Trait", "Description_en", "Unit")],
  caption = "Dictionary of phenotype traits",
  col.names = c("Trait", "Description", "Unit")
)
```

The dictionary shows that production and mass are expressed in kg, fruit dimensions in cm, firmness in N, and soluble solids in °Brix. These different units explain why the numerical scales of the traits should not be compared directly.

### 5.2 Correspondence between genotyped and phenotyped individuals

Before any rows are reordered, we verify that required columns are present and that both sources represent the same set of plants.

```{r individual-correspondence}
# Define the columns that must be present before alignment.
required_columns <- c("IND", "FAM", TRAITS)
missing_columns <- setdiff(required_columns, names(pheno_raw))


pheno_raw$IND <- trimws(as.character(pheno_raw$IND))

# Compare identifier sets in both sources before reordering rows.
missing_in_pheno <- setdiff(ids, pheno_raw$IND)
missing_in_geno <- setdiff(pheno_raw$IND, ids)

alignment <- data.frame(
  Check = c(
    "Genotyped individuals",
    "Phenotyped individuals",
    "Missing from phenotype data",
    "Missing from genotype data"
  ),
  Value = c(
    length(ids),
    nrow(pheno_raw),
    length(missing_in_pheno),
    length(missing_in_geno)
  )
)

format_table(
  alignment,
  caption = "Alignment between genotype and phenotype data"
)
```

The table confirms `r length(ids)` genotyped individuals and `r nrow(pheno_raw)` phenotyped individuals, with no individuals present in only one source. Alignment can therefore proceed without discarding plants.

### 5.3 Identifier checks

```{r identifier-checks}
# Summarize the results in `identifier_summary`.
identifier_summary <- data.frame(
  Check = c(
    "Missing genotype identifiers",
    "Duplicated genotype identifiers",
    "Missing locus names",
    "Duplicated locus names",
    "Missing phenotype identifiers",
    "Duplicated phenotype identifiers",
    "Missing required phenotype columns"
  ),
  Value = c(
    sum(is.na(ids) | ids == ""),
    sum(duplicated(ids)),
    sum(is.na(locus_names) | locus_names == ""),
    sum(duplicated(locus_names)),
    sum(is.na(pheno_raw$IND) | pheno_raw$IND == ""),
    sum(duplicated(pheno_raw$IND)),
    length(missing_columns)
  )
)

format_table(
  identifier_summary,
  caption = "Summary of identifiers and required columns"
)

```

The consistency checks contain `r sum(identifier_summary$Value)` problematic occurrences in total. For the analyzed data, this confirms the absence of empty or duplicated identifiers, missing or duplicated locus names, and missing required phenotype columns.


### 5.4 Effective phenotype alignment

Only after the previous checks are phenotypes reordered to exactly follow the individual sequence in the SSR matrix.

```{r phenotype-alignment}
# Store the row correspondence explicitly.
alignment_index <- match(ids, pheno_raw$IND)

# Reorder phenotypes to the same sequence as genotyped individuals.
pheno <- pheno_raw[
  alignment_index,
  required_columns
]


# Explicitly convert the ten traits to numeric values.
for (trait in TRAITS) {
  pheno[[trait]] <- as.numeric(
    as.character(pheno[[trait]])
  )
}
```

### 5.5 Metadata and phenotype matrix

After alignment, `metadata` retains `IND` and `FAM`, whereas `phenotypes` contains only the ten numerical traits. Family therefore remains categorical and is not mixed with numerical phenotypes.

```{r phenotype-objects}
# Separate metadata; FAM remains categorical and is not mixed with numeric phenotypes.
metadata <- data.frame(
  IND = pheno$IND,
  FAM = factor(pheno$FAM),
  row.names = pheno$IND
)

# Build the numeric phenotype matrix in the same order as genotyped individuals.
phenotypes <- as.matrix(pheno[, TRAITS])
storage.mode(phenotypes) <- "numeric"
rownames(phenotypes) <- pheno$IND

# Distribution of individuals among families.
data.frame(
  Family = names(table(metadata$FAM)),
  Individuals = as.integer(table(metadata$FAM))
) |>
  format_table(caption = "Number of individuals per family")
```

After reordering, `FAM` contains `r nlevels(metadata$FAM)` families, with family sizes ranging from `r min(table(metadata$FAM))` to `r max(table(metadata$FAM))` individuals. This distribution is used to stratify the five cross-validation folds in Module 02.

## 6. Marker quality control

### 6.1 Locus-retention criteria

Marker quality control considers two properties of each locus: the proportion of individuals with an observed genotype and the number of genotype classes represented. A locus is retained when its call rate is at least 0.90 — equivalent to 135 of 150 individuals — and at least two genotype classes are present.

These criteria distinguish missingness from lack of variation. A locus with high call rate can therefore still be removed if it is monomorphic in the panel. Filtering acts only on loci: `ssr_raw` preserves the original matrix for diagnostics, whereas `ssr` contains the loci passed to allele-dosage coding in Module 02.

```{r marker-quality-control}
# Compute call rate and genotype-class diversity to decide which SSR loci are retained.
genotype_classes <- integer(ncol(ssr_raw))

for (j in seq_len(ncol(ssr_raw))) {
  # Count the genotype classes observed at each locus.
  genotype_classes[j] <- length(
    unique(na.omit(ssr_raw[, j]))
  )
}

locus_audit <- data.frame(
  Locus = colnames(ssr_raw),
  N_called = colSums(!is.na(ssr_raw)),
  Call_rate = colMeans(!is.na(ssr_raw)),
  N_genotype_classes = genotype_classes
)

rownames(locus_audit) <- NULL

locus_audit$Keep_locus <-
  locus_audit$Call_rate >= CALL_RATE_MIN &
  locus_audit$N_genotype_classes >= 2

# Retain in the SSR matrix only loci that pass the quality-control criteria.
ssr <- ssr_raw[
  ,
  locus_audit$Keep_locus,
  drop = FALSE
]

format_table(
  locus_audit,
  digits = 3,
  caption = "Quality control of SSR loci",
  col.names = c("Locus", "Called genotypes", "Call rate", "Genotype classes", "Retain locus")
)
```

### 6.2 Quality-control result

Of the `r nrow(locus_audit)` loci evaluated, `r sum(locus_audit$Keep_locus)` satisfy both the call-rate and genotype-variation criteria. The removed locus is `r paste(locus_audit$Locus[!locus_audit$Keep_locus], collapse = ", ")`; its exclusion is due to insufficient genotype variation in the panel rather than removal of any individual.

### 6.3 Call-rate visualization

The figure orders loci by call rate and places a horizontal line at 0.90. Color reports the final retention decision, which also depends on the number of genotype classes; therefore, a locus may lie above the call-rate line and still be removed because it lacks genotype variation.

```{r marker-quality-plot, fig.height=7, fig.alt="Marker quality-control plot showing SSR locus call rates, the 0.90 reference threshold, and the final retention decision for each locus."}
# Build the locus call-rate figure and highlight the minimum quality-control threshold.
ggplot(
  locus_audit,
  aes(
    x = reorder(Locus, Call_rate),
    y = Call_rate,
    fill = Keep_locus
  )
) +
  geom_col() +
  geom_hline(
    # Add the horizontal line representing the minimum call-rate threshold.
    yintercept = CALL_RATE_MIN,
    linetype = 2
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Call rate",
    fill = "Retained",
    title = "Call rate of SSR loci"
  ) +
  theme_bw()
```

The figure shows that retention does not depend only on position relative to the 0.90 line. A locus can have high call rate and still be excluded when only one genotype class is represented, as occurs for the locus identified in the preceding table.

## 7. Genotype completeness by individual

Genotype completeness is also summarized by individual using the 35 original loci. For each plant, the number and proportion of available SSR calls are recorded. This information is diagnostic only: no completeness threshold is used to exclude individuals, and all 150 aligned individuals remain in subsequent stages.

```{r individual-summary}
# Compute, across the original loci, the number and proportion of available calls per individual.
individual_audit <- data.frame(
  IND = rownames(ssr_raw),
  N_called = rowSums(!is.na(ssr_raw)),
  Call_rate = rowMeans(!is.na(ssr_raw))
)

rownames(individual_audit) <- NULL

individual_call_rate_summary <- data.frame(
  Measure = c("Minimum", "Mean", "Median", "Maximum"),
  # List the call-rate values used in the individual summary.
  Call_rate = c(
    min(individual_audit$Call_rate),
    mean(individual_audit$Call_rate),
    median(individual_audit$Call_rate),
    max(individual_audit$Call_rate)
  )
)

format_table(
  individual_call_rate_summary,
  digits = 3,
  caption = "Distribution of SSR call rate by individual",
  col.names = c("Measure", "Call rate")
)
```

Individual call rate ranges from `r sprintf("%.1f%%", 100 * min(individual_audit$Call_rate))` to `r sprintf("%.1f%%", 100 * max(individual_audit$Call_rate))`, with a mean of `r sprintf("%.1f%%", 100 * mean(individual_audit$Call_rate))`. Because no individual-exclusion rule is applied, these differences are recorded only as genotype-completeness information.

## 8. Phenotype characterization

### 8.1 Descriptive statistics

Descriptive statistics allow direct inspection of the scale, dispersion, and range of the ten traits before prediction models are fitted. `N` also confirms phenotype availability by trait. No permanent transformation or standardization of phenotypes is performed at this stage; original values are preserved for the models.

```{r phenotype-summary}
# Compute descriptive statistics to characterize phenotype scale and variability.
phenotype_sd <- numeric(ncol(phenotypes))
phenotype_minimum <- numeric(ncol(phenotypes))
phenotype_maximum <- numeric(ncol(phenotypes))

for (j in seq_len(ncol(phenotypes))) {
  phenotype_sd[j] <- sd(phenotypes[, j])
  phenotype_minimum[j] <- min(phenotypes[, j])
  phenotype_maximum[j] <- max(phenotypes[, j])
}

# Summarize the results in `phenotype_summary`.
phenotype_summary <- data.frame(
  Trait = colnames(phenotypes),
  N = colSums(!is.na(phenotypes)),
  Mean = colMeans(phenotypes),
  SD = phenotype_sd,
  Minimum = phenotype_minimum,
  Maximum = phenotype_maximum
)

rownames(phenotype_summary) <- NULL

format_table(
  phenotype_summary,
  digits = 3,
  caption = "Summary of phenotype traits"
)
```

All ten traits contain `r min(phenotype_summary$N)` available observations, confirming that the analyzed phenotype variables have no missing values. Because traits use different units and ranges, standard deviations and intervals should be interpreted within each trait rather than compared directly across different measurement scales.

### 8.2 Phenotype observations far from the mean

Standardized values are used only as a descriptive diagnostic. For each trait,

\[
z = \frac{y-\bar{y}}{s},
\]

where $y$ is the observed value, $\bar{y}$ is the trait mean, and $s$ is the sample standard deviation. Observations whose absolute standardized distance is at least `r PHENOTYPE_DISTANCE_THRESHOLD` standard deviations are highlighted for inspection. This procedure does not remove, cap, or transform the phenotypes used later in the models.

```{r distant-phenotype-values}
# Standardize phenotypes and identify observations beyond the defined distance threshold.
phenotype_means <- colMeans(phenotypes)
phenotype_standard_deviations <- numeric(ncol(phenotypes))

for (j in seq_len(ncol(phenotypes))) {
  phenotype_standard_deviations[j] <- sd(phenotypes[, j])
}

phenotype_standardized <- scale(
  phenotypes,
  center = phenotype_means,
  scale = phenotype_standard_deviations
)

# Locate observations whose standardized distance reaches the defined threshold.
extreme_positions <- which(
  abs(phenotype_standardized) >= PHENOTYPE_DISTANCE_THRESHOLD,
  arr.ind = TRUE
)

phenotype_extremes <- data.frame(
  IND = rownames(phenotypes)[extreme_positions[, "row"]],
  Trait = colnames(phenotypes)[extreme_positions[, "col"]],
  Value = phenotypes[extreme_positions],
  Trait_mean = phenotype_means[extreme_positions[, "col"]],
  Trait_SD = phenotype_standard_deviations[extreme_positions[, "col"]],
  Standardized_value = phenotype_standardized[extreme_positions],
  Absolute_standardized_distance = abs(
    phenotype_standardized[extreme_positions]
  )
)

# Sort only the diagnostic table; the `phenotypes` matrix remains unchanged.
phenotype_extremes <- phenotype_extremes[
  order(-phenotype_extremes$Absolute_standardized_distance),
]
rownames(phenotype_extremes) <- NULL

format_table(
  phenotype_extremes,
  digits = 3,
  col.names = c(
    "Individual",
    "Trait",
    "Value",
    "Trait mean",
    "Trait SD",
    "Standardized value",
    "Absolute standardized distance"
  ),
  caption = paste0(
    "Phenotype observations at least ",
    PHENOTYPE_DISTANCE_THRESHOLD,
    " standard deviations from the mean"
  )
)
```

`r nrow(phenotype_extremes)` observations are highlighted at a standardized distance of at least `r PHENOTYPE_DISTANCE_THRESHOLD` standard deviations. These observations remain in the data and in the models; the table only documents values that merit additional inspection.

### 8.3 Phenotype distributions

The histograms complement the descriptive statistics by showing concentration, dispersion, and asymmetry within each trait. Twenty bins are used only to standardize visual resolution. Because traits have different units and ranges, each panel uses its own horizontal scale; axis limits should therefore not be compared directly across traits.

```{r phenotype-plot, fig.height=8, fig.alt="Faceted phenotype diagnostic plot showing the distributions of the ten production and fruit-quality traits in the analytical panel."}
# Reshape the phenotype matrix to long format before plotting trait distributions.
phenotype_long <- data.frame(
  IND = rownames(phenotypes),
  phenotypes,
  check.names = FALSE
) |>
  pivot_longer(
    cols = -IND,
    names_to = "Trait",
    values_to = "Value"
  )

# Build the figure used for interpretation.
ggplot(phenotype_long, aes(x = Value)) +
  # Use 20 bins only to standardize the visual resolution of the histograms.
  geom_histogram(bins = 20) +
  # Give each trait its own horizontal scale because units and ranges differ.
  facet_wrap(~ Trait, scales = "free", ncol = 2) +
  labs(
    x = "Observed value",
    y = "Number of individuals",
    title = "Phenotype distributions"
  ) +
  theme_bw()
```

The histograms allow concentration, dispersion, and asymmetry to be examined within each trait. They are descriptive diagnostics; they are not normality tests and do not imply automatic phenotype transformation.

## 9. Saving preprocessing results

The objects required by subsequent stages are gathered in `preprocessing_object.rds`, which is the primary interface with Module 02. It preserves the aligned phenotypes, SSR matrices, metadata, trait dictionary, diagnostics, and preprocessing settings together.

The main checks are also exported as CSV files for independent inspection without changing the objects used in the analysis.

```{r save-results}
# Collect settings, results, and diagnostics in the final module object.
preprocessing_object <- list(
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,
  ssr_raw = ssr_raw,
  ssr = ssr,
  locus_audit = locus_audit,
  individual_audit = individual_audit,
  phenotype_summary = phenotype_summary,
  phenotype_extremes = phenotype_extremes,
  trait_dictionary = trait_dictionary,
  # Record parameters and thresholds required to reproduce this stage.
  settings = list(
    call_rate_min = CALL_RATE_MIN,
    phenotype_distance_threshold = PHENOTYPE_DISTANCE_THRESHOLD,
    fixed_effects = "FAM"
  )
)

# Save the RDS object that is read directly by Module 02.
saveRDS(
  preprocessing_object,
  file.path(OUTPUT_DIR, "preprocessing_object.rds")
)

# Export tabular diagnostics as CSV files for inspection independently of the RDS object.
write.csv(alignment, file.path(OUTPUT_DIR, "alignment_summary.csv"), row.names = FALSE)
write.csv(locus_audit, file.path(OUTPUT_DIR, "locus_audit.csv"), row.names = FALSE)
write.csv(individual_audit, file.path(OUTPUT_DIR, "individual_audit.csv"), row.names = FALSE)
write.csv(phenotype_summary, file.path(OUTPUT_DIR, "phenotype_summary.csv"), row.names = FALSE)
write.csv(
  phenotype_extremes,
  file.path(OUTPUT_DIR, "phenotype_extreme_values.csv"),
  row.names = FALSE
)
write.csv(trait_dictionary, file.path(OUTPUT_DIR, "trait_dictionary.csv"), row.names = FALSE)
```

The final inventory collects the dimensions that should be checked before continuing: numbers of individuals and traits, loci before and after quality control, families, missing SSR calls, and phenotype observations highlighted only for inspection.

```{r final-summary}
# Summarize final dimensions and counts of missing or highlighted observations from preprocessing.
data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Original SSR loci",
    "Retained SSR loci",
    "Families",
    "Missing SSR calls",
    "Highlighted phenotype observations"
  ),
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    ncol(ssr),
    nlevels(metadata$FAM),
    sum(is.na(ssr_raw)),
    nrow(phenotype_extremes)
  )
) |>
  format_table(caption = "Final preprocessing summary")
```

The summary confirms the conditions required before genomic-matrix construction: individuals and phenotypes remain aligned, filtering acts on SSR loci rather than individuals, and highlighted phenotype observations remain in the data because their identification is diagnostic only.

## 10. Next step

Module 02 converts the retained SSR loci into allele-dosage columns, constructs
the marker and genomic relationship matrices, creates the family fixed-effect
matrix, and defines the five cross-validation folds.

<div style="display:flex; justify-content:space-between; gap:1rem; margin-top:2rem; padding-top:1rem; border-top:1px solid #ddd;">
<a href="index.html">← Home</a>
<a href="02_matrices_en.html">Module 02 — Matrix construction →</a>
</div>
