---
title: "Data preprocessing for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

Genomic selection requires phenotype records and marker information to refer
to the same individuals and to follow the same order.

This step reads the phenotype and SSR data, organizes individual and locus
labels, converts missing SSR codes to `NA`, aligns the two data sources, and
removes loci with insufficient call rate or no genotypic variation.

The block column (`BL`) is no longer part of the statistical model and is
removed immediately after the phenotype data are read. Family (`FAM`) is kept
as the only categorical covariate for later use as a fixed effect.

The five cross-validation folds are created after this preprocessing step.

## 2. Packages and file paths

The required packages are loaded first, followed by the paths of the two
source workbooks and the output directory.

```{r packages}
suppressPackageStartupMessages({
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
  library(here)
})
```

```{r paths}
GENOTYPE_FILE <- here("data", "papaya_ssr_genotypes.xlsx")
PHENOTYPE_FILE <- here("data", "papaya_phenotypes_2024_2025.xlsx")
OUTPUT_DIR <- here("output", "preprocessing")

CALL_RATE_MIN <- 0.90
PHENOTYPE_DISTANCE_THRESHOLD <- 4

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
```

## 3. Read the original workbooks

The genotype workbook uses the `Genes` sheet for numerical SSR codes and the
`Genealex` sheet for individual and locus labels. Phenotypes are read from the
`Médias_IND_artigo` sheet.

The `BL` column is discarded during reading because block is no longer included
in the models. Simple samples of both data sources are then shown so readers
can directly inspect the structure of the data used in the analysis.

```{r read-data}
genes <- read_excel(
  GENOTYPE_FILE,
  sheet = "Genes",
  col_names = FALSE,
  na = c("", "NA")
)

genealex <- read_excel(
  GENOTYPE_FILE,
  sheet = "Genealex",
  col_names = FALSE,
  na = c("", "NA")
)

pheno_raw <- read_excel(
  PHENOTYPE_FILE,
  sheet = "Médias_IND_artigo",
  na = c("", "NA")
) |>
  select(-BL)

# Sample of the genotype data as read from the workbook.
genes[1:6, 1:6]

# Sample of the phenotype data after removing BL.
head(pheno_raw)

data.frame(
  Source = c("Genes", "Genealex", "Médias_IND_artigo without BL"),
  Rows = c(nrow(genes), nrow(genealex), nrow(pheno_raw)),
  Columns = c(ncol(genes), ncol(genealex), ncol(pheno_raw))
) |>
  kable(caption = "Dimensions of the data after reading")
```

## 4. Organize the SSR data

The first column of `Genes` identifies population groups. The remaining
columns contain the 35 SSR loci.

Individual identifiers are obtained from `Genealex` and kept as character
values so that labels such as `144B` are preserved. Missing SSR calls coded as
`-9` are converted to `NA`.

```{r organize-ssr}
n_individuals <- nrow(genes)
n_loci <- ncol(genes) - 1

individual_rows <- 4:(3 + n_individuals)
locus_columns <- seq(3, 2 + 2 * n_loci, by = 2)

ids <- trimws(
  as.character(genealex[[2]][individual_rows])
)

locus_names <- trimws(
  as.character(
    unlist(genealex[3, locus_columns], use.names = FALSE)
  )
)

ssr_raw <- as.matrix(
  genes[, -1, drop = FALSE]
)

storage.mode(ssr_raw) <- "numeric"
ssr_raw[ssr_raw == -9] <- NA_real_

rownames(ssr_raw) <- ids
colnames(ssr_raw) <- locus_names
```

## 5. Organize and align the phenotype data

Ten traits are used in the analysis. Family (`FAM`) is retained as a
categorical variable and will be the only fixed effect in addition to the model
intercept.

The phenotype table is reordered according to the SSR identifiers. This
guarantees that each phenotype row refers to the same individual as the
corresponding marker row.

```{r organize-phenotypes}
TRAITS <- c(
  "MMCOM", "PRODCOM", "PRODEF", "CF", "DF",
  "DCI", "EP", "FFC", "FFP", "SS"
)

required_columns <- c("IND", "FAM", TRAITS)

missing_columns <- setdiff(
  required_columns,
  names(pheno_raw)
)

if (length(missing_columns) > 0L) {
  stop(
    "Required phenotype columns are missing: ",
    paste(missing_columns, collapse = ", ")
  )
}

pheno_raw$IND <- trimws(as.character(pheno_raw$IND))

missing_in_pheno <- setdiff(ids, pheno_raw$IND)
missing_in_geno <- setdiff(pheno_raw$IND, ids)

alignment <- data.frame(
  Check = c(
    "Genotyped individuals",
    "Phenotyped individuals",
    "Missing in phenotype data",
    "Missing in genotype data"
  ),
  Value = c(
    length(ids),
    nrow(pheno_raw),
    length(missing_in_pheno),
    length(missing_in_geno)
  )
)

kable(
  alignment,
  caption = "Alignment between genotype and phenotype data"
)

identifier_summary <- data.frame(
  Check = c(
    "Missing genotype identifiers",
    "Duplicated genotype identifiers",
    "Missing locus labels",
    "Duplicated locus labels",
    "Missing phenotype identifiers",
    "Duplicated phenotype identifiers",
    "Missing required phenotype columns"
  ),
  Value = c(
    sum(is.na(ids) | ids == ""),
    sum(duplicated(ids)),
    sum(is.na(locus_names) | locus_names == ""),
    sum(duplicated(locus_names)),
    sum(is.na(pheno_raw$IND) | pheno_raw$IND == ""),
    sum(duplicated(pheno_raw$IND)),
    length(missing_columns)
  )
)

identifier_integrity_checks <- c(
  genotype_ids_complete =
    !any(is.na(ids) | ids == ""),
  genotype_ids_unique =
    !anyDuplicated(ids),
  locus_names_complete =
    !any(is.na(locus_names) | locus_names == ""),
  locus_names_unique =
    !anyDuplicated(locus_names),
  phenotype_ids_complete =
    !any(is.na(pheno_raw$IND) | pheno_raw$IND == ""),
  phenotype_ids_unique =
    !anyDuplicated(pheno_raw$IND),
  genotype_phenotype_id_sets_match =
    length(missing_in_pheno) == 0L &&
    length(missing_in_geno) == 0L
)

if (!all(identifier_integrity_checks)) {
  stop(
    "Identifier/alignment contract failed: ",
    paste(
      names(identifier_integrity_checks)[
        !identifier_integrity_checks
      ],
      collapse = ", "
    )
  )
}

kable(
  identifier_summary,
  caption = "Identifier and column summary"
)

pheno <- pheno_raw[
  match(ids, pheno_raw$IND),
  required_columns
]

for (trait in TRAITS) {
  pheno[[trait]] <- as.numeric(
    as.character(
      pheno[[trait]]
    )
  )
}

metadata <- data.frame(
  IND = pheno$IND,
  FAM = factor(pheno$FAM),
  row.names = pheno$IND
)

phenotypes <- as.matrix(pheno[, TRAITS])
storage.mode(phenotypes) <- "numeric"
rownames(phenotypes) <- pheno$IND
```

## 6. Marker quality control

For each locus, we calculate the call rate and the number of observed genotype
classes. A locus is retained when its call rate is at least 0.90 and at least
two genotype classes are present.

```{r marker-quality}
genotype_classes <- integer(
  ncol(ssr_raw)
)

for (j in seq_len(ncol(ssr_raw))) {
  genotype_classes[j] <- length(
    unique(
      na.omit(
        ssr_raw[, j]
      )
    )
  )
}

locus_audit <- data.frame(
  Locus = colnames(ssr_raw),
  N_called = colSums(!is.na(ssr_raw)),
  Call_rate = colMeans(!is.na(ssr_raw)),
  N_genotype_classes = genotype_classes
)

rownames(locus_audit) <- NULL

locus_audit$Keep_locus <-
  locus_audit$Call_rate >= CALL_RATE_MIN &
  locus_audit$N_genotype_classes >= 2

ssr <- ssr_raw[
  ,
  locus_audit$Keep_locus,
  drop = FALSE
]

kable(
  locus_audit,
  digits = 3,
  caption = "Quality control of the SSR loci"
)
```

The figure shows the call rate of each locus and the minimum accepted value.

```{r marker-quality-plot, fig.height=7}
ggplot(
  locus_audit,
  aes(
    x = reorder(Locus, Call_rate),
    y = Call_rate,
    fill = Keep_locus
  )
) +
  geom_col() +
  geom_hline(
    yintercept = CALL_RATE_MIN,
    linetype = 2
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Call rate",
    fill = "Retained",
    title = "SSR call rate by locus"
  ) +
  theme_bw()
```

## 7. Data summaries

The individual audit summarizes the amount of available SSR information for
each plant.

```{r individual-summary}
individual_audit <- data.frame(
  IND = rownames(ssr_raw),
  N_called = rowSums(!is.na(ssr_raw)),
  Call_rate = rowMeans(!is.na(ssr_raw))
)

rownames(individual_audit) <- NULL

individual_call_rate_summary <- data.frame(
  Measure = c("Minimum", "Mean", "Median", "Maximum"),
  Call_rate = c(
    min(individual_audit$Call_rate),
    mean(individual_audit$Call_rate),
    median(individual_audit$Call_rate),
    max(individual_audit$Call_rate)
  )
)

kable(
  individual_call_rate_summary,
  digits = 3,
  caption = "Distribution summary of individual SSR call rates"
)
```

The phenotype summary provides a simple check of the scale and variation of
the ten traits.

```{r phenotype-summary}
phenotype_sd <- numeric(
  ncol(phenotypes)
)

phenotype_minimum <- numeric(
  ncol(phenotypes)
)

phenotype_maximum <- numeric(
  ncol(phenotypes)
)

for (j in seq_len(ncol(phenotypes))) {
  phenotype_sd[j] <- sd(
    phenotypes[, j]
  )

  phenotype_minimum[j] <- min(
    phenotypes[, j]
  )

  phenotype_maximum[j] <- max(
    phenotypes[, j]
  )
}

phenotype_summary <- data.frame(
  Trait = colnames(phenotypes),
  N = colSums(!is.na(phenotypes)),
  Mean = colMeans(phenotypes),
  SD = phenotype_sd,
  Minimum = phenotype_minimum,
  Maximum = phenotype_maximum
)

rownames(phenotype_summary) <- NULL

kable(
  phenotype_summary,
  digits = 3,
  caption = "Summary of the phenotype traits"
)
```

### Phenotypic observations distant from the trait mean

Standardized values are used only as a descriptive diagnostic to identify
observations located far from the mean of their respective trait. An
observation is highlighted when its absolute standardized value is at least
`r PHENOTYPE_DISTANCE_THRESHOLD`.

The highlighted values are retained exactly as recorded in the source
workbook. They are not automatically corrected, winsorized, transformed, or
removed. Their potential influence must be considered when interpreting model
fit, predictive performance, residual diagnostics, and genomic rankings.

```{r phenotype-extreme-values}
phenotype_means <- colMeans(
  phenotypes
)

phenotype_standard_deviations <- numeric(
  ncol(phenotypes)
)

for (j in seq_len(ncol(phenotypes))) {
  phenotype_standard_deviations[j] <- sd(
    phenotypes[, j]
  )
}

phenotype_standardized <- scale(
  phenotypes,
  center = phenotype_means,
  scale = phenotype_standard_deviations
)

extreme_positions <- which(
  abs(phenotype_standardized) >=
    PHENOTYPE_DISTANCE_THRESHOLD,
  arr.ind = TRUE
)

phenotype_extremes <- data.frame(
  IND = rownames(phenotypes)[
    extreme_positions[, "row"]
  ],
  Trait = colnames(phenotypes)[
    extreme_positions[, "col"]
  ],
  Value = phenotypes[
    extreme_positions
  ],
  Trait_mean = phenotype_means[
    extreme_positions[, "col"]
  ],
  Trait_SD = phenotype_standard_deviations[
    extreme_positions[, "col"]
  ],
  Standardized_value = phenotype_standardized[
    extreme_positions
  ],
  Absolute_standardized_distance = abs(
    phenotype_standardized[
      extreme_positions
    ]
  )
)

phenotype_extremes <- phenotype_extremes[
  order(
    -phenotype_extremes$
      Absolute_standardized_distance
  ),
]

rownames(phenotype_extremes) <- NULL

kable(
  phenotype_extremes,
  digits = 3,
  caption = paste0(
    "Phenotypic observations at least ",
    PHENOTYPE_DISTANCE_THRESHOLD,
    " standard deviations from the trait mean"
  )
)
```

The trait dictionary records the biological meaning and measurement unit of
each phenotype code.

```{r trait-dictionary}
trait_dictionary <- data.frame(
  Trait = TRAITS,
  Description_en = c(
    "Mean mass of commercial fruit",
    "Production of commercial fruit",
    "Production of defective fruit",
    "Fruit length",
    "Fruit diameter",
    "Internal-cavity diameter",
    "Mean pulp thickness",
    "External fruit firmness",
    "Internal pulp firmness",
    "Soluble solids"
  ),
  Description_pt = c(
    "Massa média de frutos comerciais",
    "Produção de frutos comerciais",
    "Produção de frutos defeituosos",
    "Comprimento do fruto",
    "Diâmetro do fruto",
    "Diâmetro da cavidade interna",
    "Espessura média da polpa",
    "Firmeza externa do fruto",
    "Firmeza interna da polpa",
    "Sólidos solúveis"
  ),
  Unit = c(
    "kg", "kg", "kg", "cm", "cm",
    "cm", "cm", "N", "N", "°Brix"
  )
)

rownames(trait_dictionary) <- NULL

kable(
  trait_dictionary[, c("Trait", "Description_en", "Unit")],
  caption = "Dictionary of the phenotype traits"
)
```

```{r phenotype-plot, fig.height=8}
phenotype_long <- data.frame(
  IND = rownames(phenotypes),
  phenotypes,
  check.names = FALSE
) |>
  pivot_longer(
    cols = -IND,
    names_to = "Trait",
    values_to = "Value"
  )

ggplot(phenotype_long, aes(x = Value)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ Trait, scales = "free", ncol = 2) +
  labs(
    x = "Observed value",
    y = "Number of individuals",
    title = "Phenotypic distributions"
  ) +
  theme_bw()
```

## 8. Preprocessing outputs

The aligned phenotype matrix, SSR matrices, metadata, summaries, and trait
dictionary are stored in one object for the following stages. Metadata contain
only the individual identifier and family used as the fixed effect.

```{r save-outputs}
EXPECTED_N_INDIVIDUALS <- 150L
EXPECTED_N_ORIGINAL_LOCI <- 35L
EXPECTED_N_FAMILIES <- 10L
EXPECTED_FAMILY_SIZE <- 15L

expected_locus_keep <-
  locus_audit$Call_rate >= CALL_RATE_MIN &
  locus_audit$N_genotype_classes >= 2

preprocessing_integrity_checks <- c(
  individual_count =
    length(ids) == EXPECTED_N_INDIVIDUALS,
  genotype_ids_unique =
    !anyDuplicated(ids),
  ssr_raw_dimensions =
    identical(
      dim(ssr_raw),
      c(EXPECTED_N_INDIVIDUALS, EXPECTED_N_ORIGINAL_LOCI)
    ),
  ssr_raw_ids_aligned =
    identical(rownames(ssr_raw), ids),
  ssr_raw_observed_values_finite =
    all(is.finite(ssr_raw[!is.na(ssr_raw)])),
  locus_audit_count =
    nrow(locus_audit) == EXPECTED_N_ORIGINAL_LOCI,
  locus_filter_rule_exact =
    identical(locus_audit$Keep_locus, expected_locus_keep),
  retained_loci_match_ssr =
    identical(
      colnames(ssr),
      locus_audit$Locus[locus_audit$Keep_locus]
    ),
  retained_ssr_ids_aligned =
    identical(rownames(ssr), ids),
  phenotype_dimensions =
    identical(
      dim(phenotypes),
      c(EXPECTED_N_INDIVIDUALS, length(TRAITS))
    ),
  phenotype_traits_exact =
    identical(colnames(phenotypes), TRAITS),
  phenotype_ids_aligned =
    identical(rownames(phenotypes), ids),
  phenotype_values_finite =
    all(is.finite(phenotypes)),
  metadata_ids_aligned =
    identical(as.character(metadata$IND), ids),
  families_complete =
    !anyNA(metadata$FAM),
  family_levels_exact =
    setequal(
      unique(as.character(metadata$FAM)),
      as.character(seq_len(EXPECTED_N_FAMILIES))
    ),
  family_sizes_exact =
    length(table(metadata$FAM)) == EXPECTED_N_FAMILIES &&
    all(table(metadata$FAM) == EXPECTED_FAMILY_SIZE),
  trait_dictionary_exact =
    identical(as.character(trait_dictionary$Trait), TRAITS)
)

if (!all(preprocessing_integrity_checks)) {
  stop(
    "Final preprocessing integrity gate failed: ",
    paste(
      names(preprocessing_integrity_checks)[
        !preprocessing_integrity_checks
      ],
      collapse = ", "
    )
  )
}

preprocessing_object <- list(
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,
  ssr_raw = ssr_raw,
  ssr = ssr,
  locus_audit = locus_audit,
  individual_audit = individual_audit,
  phenotype_summary = phenotype_summary,
  phenotype_extremes = phenotype_extremes,
  trait_dictionary = trait_dictionary,
  settings = list(
    call_rate_min = CALL_RATE_MIN,
    phenotype_distance_threshold =
      PHENOTYPE_DISTANCE_THRESHOLD,
    fixed_effects = "FAM"
  )
)

saveRDS(
  preprocessing_object,
  file.path(OUTPUT_DIR, "preprocessing_object.rds")
)

write.csv(
  alignment,
  file.path(OUTPUT_DIR, "alignment_summary.csv"),
  row.names = FALSE
)

write.csv(
  locus_audit,
  file.path(OUTPUT_DIR, "locus_audit.csv"),
  row.names = FALSE
)

write.csv(
  individual_audit,
  file.path(OUTPUT_DIR, "individual_audit.csv"),
  row.names = FALSE
)

write.csv(
  phenotype_summary,
  file.path(OUTPUT_DIR, "phenotype_summary.csv"),
  row.names = FALSE
)

write.csv(
  phenotype_extremes,
  file.path(
    OUTPUT_DIR,
    "phenotype_extreme_values.csv"
  ),
  row.names = FALSE
)

write.csv(
  trait_dictionary,
  file.path(OUTPUT_DIR, "trait_dictionary.csv"),
  row.names = FALSE
)
```

```{r final-inventory}
data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Raw SSR loci",
    "Retained SSR loci",
    "Families (fixed effect)",
    "Missing SSR calls",
    "Flagged phenotype observations"
  ),
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    ncol(ssr),
    nlevels(metadata$FAM),
    sum(is.na(ssr_raw)),
    nrow(phenotype_extremes)
  )
) |>
  kable(caption = "Final preprocessing inventory")
```

## 9. Next step

The next module converts retained SSR loci into allele-dosage columns, builds
the marker and genomic relationship matrices, creates the fixed-effect matrix
containing family only, and defines the five cross-validation folds.
