---
title: "Data preprocessing for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

This first module prepares the phenotype and SSR marker data used in the
papaya genomic selection analysis. The aim is to create a clean and aligned
dataset before constructing marker matrices, genomic relationship matrices,
or prediction models.

The workflow is intentionally direct: read the original files, organize the
identifiers, align both datasets, inspect marker quality, and save the cleaned
objects.

The five cross-validation folds will be created in the next module, together
with the marker and genomic matrices.

## 2. Packages and file paths

The packages below are used to read Excel files, organize data, create simple
tables and figures, and define portable paths inside the workflowr project.

```{r packages}
suppressPackageStartupMessages({
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
  library(here)
})
```

```{r paths}
GENOTYPE_FILE <- here("data", "papaya_ssr_genotypes.xlsx")
PHENOTYPE_FILE <- here("data", "papaya_phenotypes_2024_2025.xlsx")
OUTPUT_DIR <- here("output", "preprocessing")

CALL_RATE_MIN <- 0.90

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

if (!file.exists(GENOTYPE_FILE)) {
  stop("Genotype file not found: ", GENOTYPE_FILE)
}

if (!file.exists(PHENOTYPE_FILE)) {
  stop("Phenotype file not found: ", PHENOTYPE_FILE)
}
```

## 3. Read the original workbooks

The genotype workbook uses the `Genes` sheet for the numerical SSR codes and
the `Genealex` sheet for individual and locus labels. Phenotypes are read from
the `Médias_IND_artigo` sheet.

```{r read-data}
genes <- read_excel(
  GENOTYPE_FILE,
  sheet = "Genes",
  col_names = FALSE,
  na = c("", "NA")
)

genealex <- read_excel(
  GENOTYPE_FILE,
  sheet = "Genealex",
  col_names = FALSE,
  na = c("", "NA")
)

pheno_raw <- read_excel(
  PHENOTYPE_FILE,
  sheet = "Médias_IND_artigo",
  na = c("", "NA")
)

data.frame(
  Source = c("Genes", "Genealex", "Médias_IND_artigo"),
  Rows = c(nrow(genes), nrow(genealex), nrow(pheno_raw)),
  Columns = c(ncol(genes), ncol(genealex), ncol(pheno_raw))
) |>
  kable(caption = "Dimensions of the original data")
```

## 4. Organize the SSR data

The first column of `Genes` identifies population groups. The remaining
columns contain the 35 SSR loci.

Individual identifiers are obtained from `Genealex` and kept as character
values so that labels such as `144B` are preserved. Missing SSR calls coded as
`-9` are converted to `NA`.

```{r organize-ssr}
n_individuals <- nrow(genes)
n_loci <- ncol(genes) - 1

individual_rows <- 4:(3 + n_individuals)
locus_columns <- seq(3, 2 + 2 * n_loci, by = 2)

ids <- trimws(
  as.character(genealex[[2]][individual_rows])
)

locus_names <- trimws(
  as.character(
    unlist(genealex[3, locus_columns], use.names = FALSE)
  )
)

if (length(ids) != n_individuals) {
  stop("The number of identifiers does not match the number of genotype rows.")
}

if (length(locus_names) != n_loci ||
    anyNA(locus_names) ||
    any(locus_names == "")) {
  stop("The locus labels are incomplete or do not match the marker columns.")
}

if (anyNA(ids) || any(ids == "") || anyDuplicated(ids)) {
  stop("The genotype identifiers are incomplete or duplicated.")
}

ssr_raw <- as.matrix(
  genes[, -1, drop = FALSE]
)

storage.mode(ssr_raw) <- "numeric"
ssr_raw[ssr_raw == -9] <- NA_real_

rownames(ssr_raw) <- ids
colnames(ssr_raw) <- locus_names

ssr_raw[1:6, 1:6]
```

## 5. Organize and align the phenotype data

Ten traits are used in the analysis. Block (`BL`) and family (`FAM`) are kept
as categorical variables for later use as fixed effects.

The phenotype table is reordered according to the SSR identifiers. This
guarantees that each phenotype row refers to the same individual as the
corresponding marker row.

```{r organize-phenotypes}
TRAITS <- c(
  "MMCOM", "PRODCOM", "PRODEF", "CF", "DF",
  "DCI", "EP", "FFC", "FFP", "SS"
)

required_columns <- c("IND", "BL", "FAM", TRAITS)
missing_columns <- setdiff(required_columns, names(pheno_raw))

if (length(missing_columns) > 0) {
  stop(
    "Missing phenotype column(s): ",
    paste(missing_columns, collapse = ", ")
  )
}

pheno_raw$IND <- trimws(as.character(pheno_raw$IND))

if (
  anyNA(pheno_raw$IND) ||
  any(pheno_raw$IND == "") ||
  anyDuplicated(pheno_raw$IND)
) {
  stop("The phenotype identifiers are incomplete or duplicated.")
}

missing_in_pheno <- setdiff(ids, pheno_raw$IND)
missing_in_geno <- setdiff(pheno_raw$IND, ids)

alignment <- data.frame(
  Check = c(
    "Genotyped individuals",
    "Phenotyped individuals",
    "Missing in phenotype data",
    "Missing in genotype data"
  ),
  Value = c(
    length(ids),
    nrow(pheno_raw),
    length(missing_in_pheno),
    length(missing_in_geno)
  )
)

kable(alignment, caption = "Alignment between genotype and phenotype data")

if (length(missing_in_pheno) > 0 || length(missing_in_geno) > 0) {
  stop("Genotype and phenotype identifiers do not match.")
}

pheno <- pheno_raw[
  match(ids, pheno_raw$IND),
  required_columns
]

pheno[TRAITS] <- lapply(
  pheno[TRAITS],
  function(x) as.numeric(as.character(x))
)

metadata <- data.frame(
  IND = pheno$IND,
  BL = factor(pheno$BL),
  FAM = factor(pheno$FAM),
  row.names = pheno$IND
)

phenotypes <- as.matrix(pheno[, TRAITS])
storage.mode(phenotypes) <- "numeric"
rownames(phenotypes) <- pheno$IND

if (anyNA(phenotypes)) {
  stop("Missing values were found in the ten phenotype traits.")
}
```

## 6. Marker quality control

For each locus, we calculate the call rate and the number of observed genotype
classes. A locus is retained when its call rate is at least 0.90 and at least
two genotype classes are present.

```{r marker-quality}
locus_audit <- data.frame(
  Locus = colnames(ssr_raw),
  N_called = colSums(!is.na(ssr_raw)),
  Call_rate = colMeans(!is.na(ssr_raw)),
  N_genotype_classes = apply(
    ssr_raw,
    2,
    function(x) length(unique(na.omit(x)))
  )
)

rownames(locus_audit) <- NULL

locus_audit$Keep_locus <-
  locus_audit$Call_rate >= CALL_RATE_MIN &
  locus_audit$N_genotype_classes >= 2

ssr <- ssr_raw[
  ,
  locus_audit$Keep_locus,
  drop = FALSE
]

kable(
  locus_audit,
  digits = 3,
  caption = "Quality control of the SSR loci"
)
```

The figure shows the call rate of each locus and the minimum accepted value.

```{r marker-quality-plot, fig.height=7}
ggplot(
  locus_audit,
  aes(
    x = reorder(Locus, Call_rate),
    y = Call_rate,
    fill = Keep_locus
  )
) +
  geom_col() +
  geom_hline(
    yintercept = CALL_RATE_MIN,
    linetype = 2
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Call rate",
    fill = "Retained",
    title = "SSR call rate by locus"
  ) +
  theme_bw()
```

## 7. Data summaries

The individual audit summarizes the amount of available SSR information for
each plant.

```{r individual-summary}
individual_audit <- data.frame(
  IND = rownames(ssr_raw),
  N_called = rowSums(!is.na(ssr_raw)),
  Call_rate = rowMeans(!is.na(ssr_raw))
)

rownames(individual_audit) <- NULL

head(individual_audit, 10) |>
  kable(
    digits = 3,
    caption = "Marker summary for the first ten individuals"
  )
```

The phenotype summary provides a simple check of the scale and variation of
the ten traits.

```{r phenotype-summary}
phenotype_summary <- data.frame(
  Trait = colnames(phenotypes),
  N = colSums(!is.na(phenotypes)),
  Mean = colMeans(phenotypes),
  SD = apply(phenotypes, 2, sd),
  Minimum = apply(phenotypes, 2, min),
  Maximum = apply(phenotypes, 2, max)
)

rownames(phenotype_summary) <- NULL

kable(
  phenotype_summary,
  digits = 3,
  caption = "Summary of the phenotype traits"
)
```

The trait dictionary records the biological meaning and measurement unit of each phenotype code. These labels will be reused in the tables and figures produced by the following modules.

```{r trait-dictionary}
trait_dictionary <- data.frame(
  Trait = TRAITS,
  Description_en = c(
    "Mean mass of commercial fruit",
    "Production of commercial fruit",
    "Production of defective fruit",
    "Fruit length",
    "Fruit diameter",
    "Internal-cavity diameter",
    "Mean pulp thickness",
    "External fruit firmness",
    "Internal pulp firmness",
    "Soluble solids"
  ),
  Description_pt = c(
    "Massa média de frutos comerciais",
    "Produção de frutos comerciais",
    "Produção de frutos defeituosos",
    "Comprimento do fruto",
    "Diâmetro do fruto",
    "Diâmetro da cavidade interna",
    "Espessura média da polpa",
    "Firmeza externa do fruto",
    "Firmeza interna da polpa",
    "Sólidos solúveis"
  ),
  Unit = c(
    "kg", "kg", "kg", "cm", "cm",
    "cm", "cm", "N", "N", "°Brix"
  )
)

rownames(trait_dictionary) <- NULL

kable(
  trait_dictionary[, c("Trait", "Description_en", "Unit")],
  caption = "Dictionary of the phenotype traits"
)
```

```{r phenotype-plot, fig.height=8}
phenotype_long <- data.frame(
  IND = rownames(phenotypes),
  phenotypes,
  check.names = FALSE
) |>
  pivot_longer(
    cols = -IND,
    names_to = "Trait",
    values_to = "Value"
  )

ggplot(phenotype_long, aes(x = Value)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ Trait, scales = "free", ncol = 2) +
  labs(
    x = "Observed value",
    y = "Number of individuals",
    title = "Phenotypic distributions"
  ) +
  theme_bw()
```

## 8. Save the preprocessing outputs

The main objects are saved in one `.rds` file. The audit tables are also saved
as CSV files for direct inspection.

```{r save-outputs}
preprocessing_object <- list(
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,
  ssr_raw = ssr_raw,
  ssr = ssr,
  locus_audit = locus_audit,
  individual_audit = individual_audit,
  phenotype_summary = phenotype_summary,
  trait_dictionary = trait_dictionary,
  settings = list(
    call_rate_min = CALL_RATE_MIN
  )
)

saveRDS(
  preprocessing_object,
  file.path(OUTPUT_DIR, "preprocessing_object.rds")
)

write.csv(
  alignment,
  file.path(OUTPUT_DIR, "alignment_summary.csv"),
  row.names = FALSE
)

write.csv(
  locus_audit,
  file.path(OUTPUT_DIR, "locus_audit.csv"),
  row.names = FALSE
)

write.csv(
  individual_audit,
  file.path(OUTPUT_DIR, "individual_audit.csv"),
  row.names = FALSE
)

write.csv(
  phenotype_summary,
  file.path(OUTPUT_DIR, "phenotype_summary.csv"),
  row.names = FALSE
)

write.csv(
  trait_dictionary,
  file.path(OUTPUT_DIR, "trait_dictionary.csv"),
  row.names = FALSE
)
```

```{r final-inventory}
data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Original SSR loci",
    "Retained SSR loci",
    "Missing SSR calls"
  ),
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    ncol(ssr),
    sum(is.na(ssr_raw))
  )
) |>
  kable(caption = "Final preprocessing inventory")
```

## 9. Next step

The cleaned phenotype and SSR objects are now ready for matrix construction.
The next module will encode the multiallelic SSR dosages, create the marker and
genomic relationship matrices, and define the common five-fold
cross-validation design.
