---
title: "Data preprocessing for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

Genomic selection requires phenotypes and markers to represent the same
individuals in the same order. In this module, both data sources are read,
identifiers are organized, SSR missing-value codes are converted to `NA`,
phenotypes are aligned with genotypes, and SSR loci are subjected to quality
control.

The block column (`BL`) is removed because it is not part of the statistical model adopted in this analysis. Family (`FAM`) is preserved because it will be used both in the fixed-effect matrix and to stratify the cross-validation folds in Module 02. No individual is removed in this module.

Preprocessing produces three central objects for the following stages: the aligned phenotype matrix, the SSR matrix after locus quality control, and metadata containing individual and family. The diagnostics verify the consistency of these objects before genomic-matrix construction.

## 2. Packages, files, and parameters

The packages used in this module have distinct roles:

- `readxl`: reads the Excel workbooks containing genotype and phenotype data;
- `dplyr`: selects, transforms, and organizes tabular data;
- `tidyr`: reshapes data between wide and long formats;
- `ggplot2`: builds the quality-control and phenotype-distribution figures;
- `knitr`: formats tables in R Markdown documents;
- `here`: builds project-relative file paths that do not depend on the local working directory.

```{r packages, warning=FALSE, message=FALSE}
library(readxl)  # Read Excel workbooks.
library(dplyr)   # Manipulate and transform tables.
library(tidyr)   # Reshape data between wide and long formats.
library(ggplot2) # Build figures.
library(knitr)   # Format tables and R Markdown output.
library(here)    # Build portable project-relative paths.
```

```{r table-formatting, include=FALSE}
format_table <- function(x, caption = NULL, digits = getOption("digits"), col.names = NA) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    col.names = col.names,
    table.attr = paste0(
      'class="table table-striped table-hover" ',
      'style="width:100%; margin-bottom:0;"'
    )
  )
  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

Two control parameters are fixed:

- `CALL_RATE_MIN = 0.90`: requires at least 90% of the calls at a locus to be available. Because the panel contains 150 individuals, this corresponds to at least 135 observed genotypes per locus. Together with the requirement of at least two genotype classes, the criterion avoids retaining loci with insufficient information to represent variation.
- `PHENOTYPE_DISTANCE_THRESHOLD = 4`: flags, only for inspection, observations at least four standard deviations from the trait mean. This diagnostic rule does not remove phenotypes, cap them at a cutoff, or otherwise alter their values.

```{r paths}
# Define the two input files and the directory where preprocessing outputs are stored.
GENOTYPE_FILE <- here("data", "papaya_ssr_genotypes.xlsx")
PHENOTYPE_FILE <- here("data", "papaya_phenotypes_2024_2025.xlsx")
OUTPUT_DIR <- here("output", "preprocessing")

# Require at least 90% available calls for an SSR locus to be retained.
CALL_RATE_MIN <- 0.90
# Use four standard deviations only as a diagnostic flagging threshold.
PHENOTYPE_DISTANCE_THRESHOLD <- 4

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
```

## 3. Reading the original spreadsheets

The genotype workbook uses the `Genes` sheet for numerical SSR codes and the `Genealex` sheet for individual and locus identifiers. The `Genes` sheet is read without column names because its 36 columns correspond directly to the group index and the 35 SSR loci, with no column-name row. The `Genealex` sheet is also read without column names, but in this case because it preserves the metadata rows and original GenAlEx layout; the identifier positions are handled explicitly in the next section. Phenotypes are read from the `Médias_IND_artigo` sheet, which already contains column names.

The `BL` column is removed immediately after phenotype reading so that block is not propagated into analytical objects. The first rows make the original data organization visible, and the dimensions table checks the structure of the three sources before alignment.

```{r read-data}
# Read `Genes` without column names because the sheet has no column-name row.
genes <- read_excel(
  GENOTYPE_FILE,
  sheet = "Genes",
  col_names = FALSE,
  na = c("", "NA")
)

# Read `Genealex` without column names; identifiers are extracted by position in the next section.
genealex <- read_excel(
  GENOTYPE_FILE,
  sheet = "Genealex",
  col_names = FALSE,
  na = c("", "NA")
)

# Read phenotypes with their column names and remove BL before alignment.
pheno_raw <- read_excel(
  PHENOTYPE_FILE,
  sheet = "Médias_IND_artigo",
  na = c("", "NA")
) |>
  select(-BL)

```

The initial structure of `Genes` can be inspected in its first six rows and six columns:

```{r genes-preview}
as.data.frame(genes[1:6, 1:6]) |>
  format_table(caption = "Preview of the Genes sheet")
```

The preview shows the raw genotype layout: the first column is a sequential index, and subsequent columns contain the SSR codes organized in the next section.

The phenotype table is also inspected after `BL` is removed:

```{r phenotype-preview}
head(pheno_raw) |>
  format_table(caption = "Preview of phenotype data")
```

The phenotype preview contains `IND`, `FAM`, and the ten traits used in the analysis. The absence of `BL` confirms that this column has already been removed before genotype–phenotype alignment.

Finally, the dimensions of the three sources are compared:

```{r data-dimensions}
data.frame(
  Source = c("Genes", "Genealex", "Médias_IND_artigo without BL"),
  Rows = c(nrow(genes), nrow(genealex), nrow(pheno_raw)),
  Columns = c(ncol(genes), ncol(genealex), ncol(pheno_raw))
) |>
  format_table(caption = "Dimensions of the spreadsheets used in the analysis")
```

The dimensions table confirms `r nrow(genes)` individuals in `Genes` and `r ncol(genes) - 1` columns corresponding to SSR loci, in addition to the first index column. The phenotype table contains `r nrow(pheno_raw)` individuals. `Genealex` has its own metadata layout, so its dimensions should not be interpreted as a marker matrix.

## 4. Organizing the SSR data

The first column of `Genes` contains a sequential population-group index and does not represent the phenotype variable `FAM`; it is therefore excluded from the marker matrix. The remaining 35 columns correspond to the original SSR loci.

In `Genealex`, each piece of information is located explicitly:

1. **Individual rows.** Identifiers begin on row 4 of column 2. Because there are `n_individuals` individuals, their rows are defined by `individual_rows <- 4:(3 + n_individuals)`. The interval starts at 4 and ends at `3 + n_individuals`, including exactly one row per individual.
2. **Locus columns.** Locus names are stored on row 3 in alternating columns beginning with column 3. These positions are obtained with `locus_columns <- seq(3, 2 + 2 * n_loci, by = 2)`. The `seq()` function starts at column 3 and advances by two columns until all `n_loci` names are included.
3. **Individual identifiers.** The expression `genealex[[2]][individual_rows]` selects column 2 only at the rows defined above. Then `as.character()` preserves alphanumeric identifiers and `trimws()` removes spaces before or after the code: `ids <- trimws(as.character(genealex[[2]][individual_rows]))`.
4. **Locus names.** The expression `genealex[3, locus_columns]` selects row 3 only at the locus columns. `unlist(..., use.names = FALSE)` converts this selection to a vector, and `trimws(as.character(...))` ensures text names without surrounding spaces.
5. **SSR matrix.** The expression `genes[, -1, drop = FALSE]` selects all rows and excludes the first column of `Genes`; the index `-1` means “all columns except the first”. `drop = FALSE` keeps the result two-dimensional. `as.matrix()` converts this selection to a matrix, and `storage.mode(ssr_raw) <- "numeric"` ensures that SSR codes are stored numerically.
6. **Missing calls.** After the numerical SSR matrix is formed, `ssr_raw[ssr_raw == -9] <- NA_real_` replaces `-9`, the missing-value code in the source file, with R's standard missing value (`NA`).

```{r organize-ssr}
# Obtain panel dimensions from `Genes`; its first column is not an SSR locus.
n_individuals <- nrow(genes)
n_loci <- ncol(genes) - 1

# Translate the `Genealex` layout: individuals start on row 4 and loci occupy alternating columns.
individual_rows <- 4:(3 + n_individuals)
locus_columns <- seq(3, 2 + 2 * n_loci, by = 2)

# Keep identifiers as text to preserve alphanumeric codes.
ids <- trimws(
  as.character(genealex[[2]][individual_rows])
)

locus_names <- trimws(
  as.character(
    unlist(genealex[3, locus_columns], use.names = FALSE)
  )
)

ssr_raw <- as.matrix(
  genes[, -1, drop = FALSE]
)

storage.mode(ssr_raw) <- "numeric"
# Convert the original missing-value code to R's standard missing value.
ssr_raw[ssr_raw == -9] <- NA_real_

rownames(ssr_raw) <- ids
colnames(ssr_raw) <- locus_names

# Show a preview of the organized SSR matrix.
as.data.frame(ssr_raw[1:6, 1:8]) |>
  format_table(caption = "Preview of the organized SSR matrix")
```

SSR codes represent the two alleles observed in each individual. For example, `12` denotes one copy of allele 1 and one copy of allele 2, `22` denotes two copies of allele 2, and `33` denotes two copies of allele 3. The `33` value shown for `P3K3256CC` is not an error or a special category: this locus is multi-allelic and allele 3 is represented in the panel. In Module 02, these genotype codes are decomposed into dosages of 0, 1, or 2 copies for each observed allele.

## 5. Organizing and aligning phenotypes

Ten traits are used. Before alignment, three checks prevent genotypes and phenotypes from being associated incorrectly:

1. **Required columns.** `required_columns <- c("IND", "FAM", TRAITS)` collects the expected columns. `setdiff(required_columns, names(pheno_raw))` returns any that are absent from the spreadsheet; an empty result means that all required columns are present.
2. **Individuals present in only one source.** `setdiff(ids, pheno_raw$IND)` identifies genotyped individuals without matching phenotypes, whereas `setdiff(pheno_raw$IND, ids)` performs the reverse comparison and identifies phenotyped individuals without genotypes.
3. **Invalid or repeated identifiers.** `is.na()` and comparison with `""` identify missing or empty codes; `duplicated()` identifies repeated identifiers that would prevent unambiguous one-to-one matching.

After these checks, phenotypes are aligned with `match(ids, pheno_raw$IND)`. For every identifier in `ids`, `match()` returns its corresponding position in `pheno_raw$IND`. Those positions are used directly in `pheno_raw[match(ids, pheno_raw$IND), required_columns]`, which reorders phenotype rows to the same sequence as the SSR matrix. Thus, phenotype row *i* and genotype row *i* represent the same individual.

After alignment, `factor(pheno$FAM)` converts `FAM` to a categorical variable while preserving family levels. `as.matrix(pheno[, TRAITS])` selects the ten traits and gathers them in a matrix, and `storage.mode(phenotypes) <- "numeric"` ensures that their values are treated numerically in subsequent calculations.

```{r organize-phenotypes}
# Define the traits, align phenotypes to genotyped individuals, and build the phenotype objects.
TRAITS <- c(
  "MMCOM", "PRODCOM", "PRODEF", "CF", "DF",
  "DCI", "EP", "FFC", "FFP", "SS"
)

# Define the columns that must be present before alignment.
required_columns <- c("IND", "FAM", TRAITS)
missing_columns <- setdiff(required_columns, names(pheno_raw))

pheno_raw$IND <- trimws(as.character(pheno_raw$IND))

# Compare identifier sets in both sources before reordering rows.
missing_in_pheno <- setdiff(ids, pheno_raw$IND)
missing_in_geno <- setdiff(pheno_raw$IND, ids)

alignment <- data.frame(
  Check = c(
    "Genotyped individuals",
    "Phenotyped individuals",
    "Missing from phenotype data",
    "Missing from genotype data"
  ),
  Value = c(
    length(ids),
    nrow(pheno_raw),
    length(missing_in_pheno),
    length(missing_in_geno)
  )
)

format_table(
  alignment,
  caption = "Alignment between genotype and phenotype data"
)
```

The table confirms `r length(ids)` genotyped individuals and `r nrow(pheno_raw)` phenotyped individuals, with no individuals present in only one source. Alignment can therefore proceed without discarding plants.

```{r identifier-checks}
# Summarize the results in `identifier_summary`.
identifier_summary <- data.frame(
  Check = c(
    "Missing genotype identifiers",
    "Duplicated genotype identifiers",
    "Missing locus names",
    "Duplicated locus names",
    "Missing phenotype identifiers",
    "Duplicated phenotype identifiers",
    "Missing required phenotype columns"
  ),
  Value = c(
    sum(is.na(ids) | ids == ""),
    sum(duplicated(ids)),
    sum(is.na(locus_names) | locus_names == ""),
    sum(duplicated(locus_names)),
    sum(is.na(pheno_raw$IND) | pheno_raw$IND == ""),
    sum(duplicated(pheno_raw$IND)),
    length(missing_columns)
  )
)

format_table(
  identifier_summary,
  caption = "Summary of identifiers and required columns"
)
```

The consistency checks contain `r sum(identifier_summary$Value)` problematic occurrences in total. For the analyzed data, this confirms the absence of empty or duplicated identifiers, missing or duplicated locus names, and missing required phenotype columns.

```{r phenotype-alignment}
# Reorder phenotypes to the same sequence as genotyped individuals.
pheno <- pheno_raw[
  match(ids, pheno_raw$IND),
  required_columns
]

# Explicitly convert the ten traits to numeric values.
for (trait in TRAITS) {
  pheno[[trait]] <- as.numeric(
    as.character(pheno[[trait]])
  )
}

# Separate metadata; FAM remains categorical and is not mixed with numeric phenotypes.
metadata <- data.frame(
  IND = pheno$IND,
  FAM = factor(pheno$FAM),
  row.names = pheno$IND
)

# Build the numeric phenotype matrix in the same order as genotyped individuals.
phenotypes <- as.matrix(pheno[, TRAITS])
storage.mode(phenotypes) <- "numeric"
rownames(phenotypes) <- pheno$IND

# Distribution of individuals among families.
data.frame(
  Family = names(table(metadata$FAM)),
  Individuals = as.integer(table(metadata$FAM))
) |>
  format_table(caption = "Number of individuals per family")
```

After reordering, `FAM` contains `r nlevels(metadata$FAM)` families, with family sizes ranging from `r min(table(metadata$FAM))` to `r max(table(metadata$FAM))` individuals. This distribution is used to stratify the five cross-validation folds in Module 02.

## 6. Marker quality control

For each locus, call rate is calculated with `colMeans(!is.na(ssr_raw))`. First, `is.na(ssr_raw)` identifies missing cells; the `!` operator reverses the result, leaving `TRUE` for observed genotypes and `FALSE` for missing calls. Because R treats `TRUE` as 1 and `FALSE` as 0 when computing a mean, `colMeans()` directly returns the proportion of individuals with an observed genotype at each locus.

The number of genotype classes is calculated for each column `j` with `length(unique(na.omit(ssr_raw[, j])))`: `na.omit()` removes missing values, `unique()` retains distinct genotype codes, and `length()` counts the observed classes. A locus is retained when call rate is at least 0.90 — at least 135 of 150 individuals — and at least two genotype classes are present. The second criterion removes loci with no observable variation even when their call rate is high.

The table makes it possible to identify which criterion determined retention or removal for each locus. The `ssr` matrix created at the end contains only loci passed to allele-dosage coding in Module 02; `ssr_raw` is retained for diagnostics of the original source.

```{r marker-quality-control}
# Compute call rate and genotype-class diversity to decide which SSR loci are retained.
genotype_classes <- integer(ncol(ssr_raw))

for (j in seq_len(ncol(ssr_raw))) {
  # Count the genotype classes observed at each locus.
  genotype_classes[j] <- length(
    unique(na.omit(ssr_raw[, j]))
  )
}

locus_audit <- data.frame(
  Locus = colnames(ssr_raw),
  N_called = colSums(!is.na(ssr_raw)),
  Call_rate = colMeans(!is.na(ssr_raw)),
  N_genotype_classes = genotype_classes
)

rownames(locus_audit) <- NULL

locus_audit$Keep_locus <-
  locus_audit$Call_rate >= CALL_RATE_MIN &
  locus_audit$N_genotype_classes >= 2

# Retain in the SSR matrix only loci that pass the quality-control criteria.
ssr <- ssr_raw[
  ,
  locus_audit$Keep_locus,
  drop = FALSE
]

format_table(
  locus_audit,
  digits = 3,
  caption = "Quality control of SSR loci"
)
```

Of the `r nrow(locus_audit)` loci evaluated, `r sum(locus_audit$Keep_locus)` satisfy both the call-rate and genotype-variation criteria. The removed locus is `r paste(locus_audit$Locus[!locus_audit$Keep_locus], collapse = ", ")`; its exclusion is due to insufficient genotype variation in the panel rather than removal of any individual.

The figure orders loci by call rate and places a horizontal line at 0.90. Color reports the final retention decision, which also depends on the number of genotype classes; therefore, a locus may lie above the call-rate line and still be removed because it lacks genotype variation.

```{r marker-quality-plot, fig.height=7}
# Build the locus call-rate figure and highlight the minimum quality-control threshold.
ggplot(
  locus_audit,
  aes(
    x = reorder(Locus, Call_rate),
    y = Call_rate,
    fill = Keep_locus
  )
) +
  geom_col() +
  geom_hline(
    # Add the horizontal line representing the minimum call-rate threshold.
    yintercept = CALL_RATE_MIN,
    linetype = 2
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Call rate",
    fill = "Retained",
    title = "Call rate of SSR loci"
  ) +
  theme_bw()
```

The figure shows that retention does not depend only on position relative to the 0.90 line. A locus can have high call rate and still be excluded when only one genotype class is represented, as occurs for the locus identified in the preceding table.

## 7. Summary of genotype information

Genotype completeness is also calculated for each individual using the 35 original loci (`ssr_raw`). The expression `rowSums(!is.na(ssr_raw))` counts how many loci have an observed genotype for each individual, whereas `rowMeans(!is.na(ssr_raw))` implicitly divides that count by the number of loci and returns the proportion of available calls. No individual-exclusion threshold based on completeness is applied; all 150 aligned individuals are preserved for subsequent stages.

```{r individual-summary}
# Compute, across the original loci, the number and proportion of available calls per individual.
individual_audit <- data.frame(
  IND = rownames(ssr_raw),
  N_called = rowSums(!is.na(ssr_raw)),
  Call_rate = rowMeans(!is.na(ssr_raw))
)

rownames(individual_audit) <- NULL

individual_call_rate_summary <- data.frame(
  Measure = c("Minimum", "Mean", "Median", "Maximum"),
  # List the call-rate values used in the individual summary.
  Call_rate = c(
    min(individual_audit$Call_rate),
    mean(individual_audit$Call_rate),
    median(individual_audit$Call_rate),
    max(individual_audit$Call_rate)
  )
)

format_table(
  individual_call_rate_summary,
  digits = 3,
  caption = "Distribution of SSR call rate by individual"
)
```

Individual call rate ranges from `r sprintf("%.1f%%", 100 * min(individual_audit$Call_rate))` to `r sprintf("%.1f%%", 100 * max(individual_audit$Call_rate))`, with a mean of `r sprintf("%.1f%%", 100 * mean(individual_audit$Call_rate))`. Because no individual-exclusion rule is applied, these differences are recorded only as genotype-completeness information.

## 8. Summary of phenotype traits

Descriptive statistics allow direct inspection of the scale, dispersion, and range of the ten traits before prediction models are fitted. `N` also confirms phenotype availability by trait. No permanent transformation or standardization of phenotypes is performed at this stage; original values are preserved for the models.

```{r phenotype-summary}
# Compute descriptive statistics to characterize phenotype scale and variability.
phenotype_sd <- numeric(ncol(phenotypes))
phenotype_minimum <- numeric(ncol(phenotypes))
phenotype_maximum <- numeric(ncol(phenotypes))

for (j in seq_len(ncol(phenotypes))) {
  phenotype_sd[j] <- sd(phenotypes[, j])
  phenotype_minimum[j] <- min(phenotypes[, j])
  phenotype_maximum[j] <- max(phenotypes[, j])
}

# Summarize the results in `phenotype_summary`.
phenotype_summary <- data.frame(
  Trait = colnames(phenotypes),
  N = colSums(!is.na(phenotypes)),
  Mean = colMeans(phenotypes),
  SD = phenotype_sd,
  Minimum = phenotype_minimum,
  Maximum = phenotype_maximum
)

rownames(phenotype_summary) <- NULL

format_table(
  phenotype_summary,
  digits = 3,
  caption = "Summary of phenotype traits"
)
```

All ten traits contain `r min(phenotype_summary$N)` available observations, confirming that the analyzed phenotype variables have no missing values. Because traits use different units and ranges, standard deviations and intervals should be interpreted within each trait rather than compared directly across different measurement scales.

### 8.1 Phenotype observations far from the mean

Standardized values are used only as a descriptive diagnostic. For each trait, $z=(y-\bar{y})/s$, where $y$ is the observed value, $\bar{y}$ is the trait mean, and $s$ is the sample standard deviation. The call `scale(phenotypes, center = phenotype_means, scale = phenotype_standard_deviations)` applies this transformation to every column of the phenotype matrix.

Next, `abs(phenotype_standardized) >= PHENOTYPE_DISTANCE_THRESHOLD` compares the absolute standardized distance with the defined threshold, regardless of whether an observation is above or below the mean. `which(..., arr.ind = TRUE)` returns the row and column positions that satisfy the criterion; these positions recover the individual, trait, and original value. An observation is highlighted when $|z|$ is at least `r PHENOTYPE_DISTANCE_THRESHOLD`, but no value is removed or changed.

```{r distant-phenotype-values}
# Standardize phenotypes and identify observations beyond the defined distance threshold.
phenotype_means <- colMeans(phenotypes)
phenotype_standard_deviations <- numeric(ncol(phenotypes))

for (j in seq_len(ncol(phenotypes))) {
  phenotype_standard_deviations[j] <- sd(phenotypes[, j])
}

phenotype_standardized <- scale(
  phenotypes,
  center = phenotype_means,
  scale = phenotype_standard_deviations
)

# Locate observations whose standardized distance reaches the defined threshold.
extreme_positions <- which(
  abs(phenotype_standardized) >= PHENOTYPE_DISTANCE_THRESHOLD,
  arr.ind = TRUE
)

phenotype_extremes <- data.frame(
  IND = rownames(phenotypes)[extreme_positions[, "row"]],
  Trait = colnames(phenotypes)[extreme_positions[, "col"]],
  Value = phenotypes[extreme_positions],
  Trait_mean = phenotype_means[extreme_positions[, "col"]],
  Trait_SD = phenotype_standard_deviations[extreme_positions[, "col"]],
  Standardized_value = phenotype_standardized[extreme_positions],
  Absolute_standardized_distance = abs(
    phenotype_standardized[extreme_positions]
  )
)

# Sort only the diagnostic table; the `phenotypes` matrix remains unchanged.
phenotype_extremes <- phenotype_extremes[
  order(-phenotype_extremes$Absolute_standardized_distance),
]
rownames(phenotype_extremes) <- NULL

format_table(
  phenotype_extremes,
  digits = 3,
  col.names = c(
    "Individual",
    "Trait",
    "Value",
    "Trait mean",
    "Trait SD",
    "Standardized value",
    "Absolute standardized distance"
  ),
  caption = paste0(
    "Phenotype observations at least ",
    PHENOTYPE_DISTANCE_THRESHOLD,
    " standard deviations from the mean"
  )
)
```

`r nrow(phenotype_extremes)` observations are highlighted at a standardized distance of at least `r PHENOTYPE_DISTANCE_THRESHOLD` standard deviations. These observations remain in the data and in the models; the table only documents values that merit additional inspection.

### 8.2 Trait dictionary

The dictionary links the codes used in the R objects to biological descriptions and measurement units. This is important because the ten traits are on different scales and the same codes are used throughout the modeling and results modules.

```{r trait-dictionary}
# Associate each trait with its English and Portuguese descriptions and measurement unit.
trait_dictionary <- data.frame(
  Trait = TRAITS,
  Description_en = c(
    "Mean mass of commercial fruit",
    "Production of commercial fruit",
    "Production of defective fruit",
    "Fruit length",
    "Fruit diameter",
    "Internal-cavity diameter",
    "Mean pulp thickness",
    "External fruit firmness",
    "Internal pulp firmness",
    "Soluble solids"
  ),
  Description_pt = c(
    "Massa média de frutos comerciais",
    "Produção de frutos comerciais",
    "Produção de frutos defeituosos",
    "Comprimento do fruto",
    "Diâmetro do fruto",
    "Diâmetro da cavidade interna",
    "Espessura média da polpa",
    "Firmeza externa do fruto",
    "Firmeza interna da polpa",
    "Sólidos solúveis"
  ),
  Unit = c(
    "kg", "kg", "kg", "cm", "cm",
    "cm", "cm", "N", "N", "°Brix"
  )
)

rownames(trait_dictionary) <- NULL

format_table(
  trait_dictionary[, c("Trait", "Description_en", "Unit")],
  caption = "Dictionary of phenotype traits"
)
```

The dictionary shows that production and mass are expressed in kg, fruit dimensions in cm, firmness in N, and soluble solids in °Brix. These different units explain why the numerical scales of the traits should not be compared directly.

The histograms complement the descriptive statistics by showing distribution shape. To build them, `pivot_longer(cols = -IND, names_to = "Trait", values_to = "Value")` changes the table from one column per trait to long format, in which each row records an individual, a trait, and its value. Twenty bins are used only to define histogram resolution. Because `scales = "free"`, every trait receives its own horizontal scale; this makes within-trait shape easier to see, but axis limits should not be compared directly across panels.

```{r phenotype-plot, fig.height=8}
# Reshape the phenotype matrix to long format before plotting trait distributions.
phenotype_long <- data.frame(
  IND = rownames(phenotypes),
  phenotypes,
  check.names = FALSE
) |>
  pivot_longer(
    cols = -IND,
    names_to = "Trait",
    values_to = "Value"
  )

# Build the figure used for interpretation.
ggplot(phenotype_long, aes(x = Value)) +
  # Use 20 bins only to standardize the visual resolution of the histograms.
  geom_histogram(bins = 20) +
  # Give each trait its own horizontal scale because units and ranges differ.
  facet_wrap(~ Trait, scales = "free", ncol = 2) +
  labs(
    x = "Observed value",
    y = "Number of individuals",
    title = "Phenotype distributions"
  ) +
  theme_bw()
```

The histograms allow concentration, dispersion, and asymmetry to be examined within each trait. They are descriptive diagnostics; they are not normality tests and do not imply automatic phenotype transformation.

## 9. Saving preprocessing results

The aligned phenotype matrix, SSR data, metadata, trait dictionary, and diagnostics are gathered in `preprocessing_object.rds`. This RDS file is the primary interface with Module 02 because it preserves object structures, names, and classes together.

The CSV files written in parallel are inspection-oriented tabular outputs: they allow alignment, locus quality, individual completeness, phenotype summaries, highlighted observations, and the trait dictionary to be examined without opening the RDS object. Parameters defining this stage are also recorded in `settings`.

```{r save-results}
# Collect settings, results, and diagnostics in the final module object.
preprocessing_object <- list(
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,
  ssr_raw = ssr_raw,
  ssr = ssr,
  locus_audit = locus_audit,
  individual_audit = individual_audit,
  phenotype_summary = phenotype_summary,
  phenotype_extremes = phenotype_extremes,
  trait_dictionary = trait_dictionary,
  # Record parameters and thresholds required to reproduce this stage.
  settings = list(
    call_rate_min = CALL_RATE_MIN,
    phenotype_distance_threshold = PHENOTYPE_DISTANCE_THRESHOLD,
    fixed_effects = "FAM"
  )
)

# Save the RDS object that is read directly by Module 02.
saveRDS(
  preprocessing_object,
  file.path(OUTPUT_DIR, "preprocessing_object.rds")
)

# Export tabular diagnostics as CSV files for inspection independently of the RDS object.
write.csv(alignment, file.path(OUTPUT_DIR, "alignment_summary.csv"), row.names = FALSE)
write.csv(locus_audit, file.path(OUTPUT_DIR, "locus_audit.csv"), row.names = FALSE)
write.csv(individual_audit, file.path(OUTPUT_DIR, "individual_audit.csv"), row.names = FALSE)
write.csv(phenotype_summary, file.path(OUTPUT_DIR, "phenotype_summary.csv"), row.names = FALSE)
write.csv(
  phenotype_extremes,
  file.path(OUTPUT_DIR, "phenotype_extreme_values.csv"),
  row.names = FALSE
)
write.csv(trait_dictionary, file.path(OUTPUT_DIR, "trait_dictionary.csv"), row.names = FALSE)
```

The final inventory collects the dimensions that should be checked before continuing: numbers of individuals and traits, loci before and after quality control, families, missing SSR calls, and phenotype observations highlighted only for inspection.

```{r final-summary}
# Summarize final dimensions and counts of missing or highlighted observations from preprocessing.
data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Original SSR loci",
    "Retained SSR loci",
    "Families",
    "Missing SSR calls",
    "Highlighted phenotype observations"
  ),
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    ncol(ssr),
    nlevels(metadata$FAM),
    sum(is.na(ssr_raw)),
    nrow(phenotype_extremes)
  )
) |>
  format_table(caption = "Final preprocessing summary")
```

The summary confirms the conditions required before genomic-matrix construction: individuals and phenotypes remain aligned, filtering acts on SSR loci rather than individuals, and highlighted phenotype observations remain in the data because their identification is diagnostic only.

## 10. Next step

Module 02 converts the retained SSR loci into allele-dosage columns, constructs
the marker and genomic relationship matrices, creates the family fixed-effect
matrix, and defines the five cross-validation folds.

<div style="display:flex; justify-content:space-between; gap:1rem; margin-top:2rem; padding-top:1rem; border-top:1px solid #ddd;">
<a href="index.html">← Home</a>
<a href="02_matrices_en.html">Module 02 — Matrix construction →</a>
</div>
