---
title: "Data preprocessing for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

Genomic selection requires phenotypes and markers to represent the same
individuals in the same order. In this module, both data sources are read,
identifiers are organized, SSR missing-value codes are converted to `NA`,
phenotypes are aligned with genotypes, and SSR loci are subjected to quality
control.

The block column (`BL`) is removed because it is not part of the statistical
model. Family (`FAM`) is retained for later use as a fixed effect.

## 2. Packages and file paths

```{r packages}
suppressPackageStartupMessages({
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
  library(here)
})
```

The next step defines the file paths used as inputs and outputs in this module.

```{r paths}
GENOTYPE_FILE <- here("data", "papaya_ssr_genotypes.xlsx")
PHENOTYPE_FILE <- here("data", "papaya_phenotypes_2024_2025.xlsx")
OUTPUT_DIR <- here("output", "preprocessing")

CALL_RATE_MIN <- 0.90
PHENOTYPE_DISTANCE_THRESHOLD <- 4

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
```

## 3. Reading the original spreadsheets

The genotype workbook uses the `Genes` sheet for numerical SSR codes and the
`Genealex` sheet for individual and locus identifiers. Phenotypes are read from
the `Médias_IND_artigo` sheet.

```{r read-data}
# Load the genotype and phenotype workbooks used by the preprocessing module.
genes <- read_excel(
  GENOTYPE_FILE,
  sheet = "Genes",
  col_names = FALSE,
  na = c("", "NA")
)

genealex <- read_excel(
  GENOTYPE_FILE,
  sheet = "Genealex",
  col_names = FALSE,
  na = c("", "NA")
)

pheno_raw <- read_excel(
  PHENOTYPE_FILE,
  sheet = "Médias_IND_artigo",
  na = c("", "NA")
) |>
  select(-BL)

# First rows of the SSR codes.
genes[1:6, 1:6]

# First phenotype rows after removing BL.
head(pheno_raw)

data.frame(
  Source = c("Genes", "Genealex", "Médias_IND_artigo without BL"),
  Rows = c(nrow(genes), nrow(genealex), nrow(pheno_raw)),
  Columns = c(ncol(genes), ncol(genealex), ncol(pheno_raw))
) |>
  kable(caption = "Dimensions of the spreadsheets used in the analysis")
```

## 4. Organizing the SSR data

The first column of `Genes` identifies population groups and the remaining
columns contain the 35 SSR loci. Individual and locus identifiers are obtained
from `Genealex`. Missing SSR calls coded as `-9` are converted to `NA`.

```{r organize-ssr}
# Organize marker information and compute the diagnostics required at this stage.
n_individuals <- nrow(genes)
n_loci <- ncol(genes) - 1

individual_rows <- 4:(3 + n_individuals)
locus_columns <- seq(3, 2 + 2 * n_loci, by = 2)

ids <- trimws(
  as.character(genealex[[2]][individual_rows])
)

locus_names <- trimws(
  as.character(
    unlist(genealex[3, locus_columns], use.names = FALSE)
  )
)

ssr_raw <- as.matrix(
  genes[, -1, drop = FALSE]
)

storage.mode(ssr_raw) <- "numeric"
ssr_raw[ssr_raw == -9] <- NA_real_

rownames(ssr_raw) <- ids
colnames(ssr_raw) <- locus_names

# Initial structure of the SSR matrix.
ssr_raw[1:6, 1:8]
```

## 5. Organizing and aligning phenotypes

Ten traits are used. The phenotype table is reordered according to the SSR
identifiers so that every phenotype row corresponds to the same plant in the
genotype matrix.

```{r organize-phenotypes}
# Organize the phenotype data and apply the preprocessing rules defined for this stage.
TRAITS <- c(
  "MMCOM", "PRODCOM", "PRODEF", "CF", "DF",
  "DCI", "EP", "FFC", "FFP", "SS"
)

required_columns <- c("IND", "FAM", TRAITS)
missing_columns <- setdiff(required_columns, names(pheno_raw))

pheno_raw$IND <- trimws(as.character(pheno_raw$IND))

missing_in_pheno <- setdiff(ids, pheno_raw$IND)
missing_in_geno <- setdiff(pheno_raw$IND, ids)

alignment <- data.frame(
  Check = c(
    "Genotyped individuals",
    "Phenotyped individuals",
    "Missing from phenotype data",
    "Missing from genotype data"
  ),
  # List the values corresponding to the table entries.
  Value = c(
    length(ids),
    nrow(pheno_raw),
    length(missing_in_pheno),
    length(missing_in_geno)
  )
)

kable(
  alignment,
  caption = "Alignment between genotype and phenotype data"
)

# Summarize the results in `identifier_summary`.
identifier_summary <- data.frame(
  Check = c(
    "Missing genotype identifiers",
    "Duplicated genotype identifiers",
    "Missing locus names",
    "Duplicated locus names",
    "Missing phenotype identifiers",
    "Duplicated phenotype identifiers",
    "Missing required phenotype columns"
  ),
  # List the values corresponding to the table entries.
  Value = c(
    sum(is.na(ids) | ids == ""),
    sum(duplicated(ids)),
    sum(is.na(locus_names) | locus_names == ""),
    sum(duplicated(locus_names)),
    sum(is.na(pheno_raw$IND) | pheno_raw$IND == ""),
    sum(duplicated(pheno_raw$IND)),
    length(missing_columns)
  )
)

kable(
  identifier_summary,
  caption = "Summary of identifiers and required columns"
)

# Reorder phenotypes to the same sequence as genotyped individuals.
pheno <- pheno_raw[
  match(ids, pheno_raw$IND),
  required_columns
]

# Explicitly convert the ten traits to numeric values.
for (trait in TRAITS) {
  pheno[[trait]] <- as.numeric(
    as.character(pheno[[trait]])
  )
}

metadata <- data.frame(
  IND = pheno$IND,
  FAM = factor(pheno$FAM),
  row.names = pheno$IND
)

phenotypes <- as.matrix(pheno[, TRAITS])
storage.mode(phenotypes) <- "numeric"
rownames(phenotypes) <- pheno$IND

# Distribution of individuals among families.
data.frame(
  Family = names(table(metadata$FAM)),
  Individuals = as.integer(table(metadata$FAM))
) |>
  kable(caption = "Number of individuals per family")
```

## 6. Marker quality control

For each locus, the call rate and number of observed genotype classes are
calculated. A locus is retained when its call rate is at least 0.90 and at
least two genotype classes are observed.

```{r marker-quality-control}
# Organize marker information and compute the diagnostics required at this stage.
genotype_classes <- integer(ncol(ssr_raw))

for (j in seq_len(ncol(ssr_raw))) {
  # Count the genotype classes observed at each locus.
  genotype_classes[j] <- length(
    unique(na.omit(ssr_raw[, j]))
  )
}

locus_audit <- data.frame(
  Locus = colnames(ssr_raw),
  N_called = colSums(!is.na(ssr_raw)),
  Call_rate = colMeans(!is.na(ssr_raw)),
  N_genotype_classes = genotype_classes
)

rownames(locus_audit) <- NULL

locus_audit$Keep_locus <-
  locus_audit$Call_rate >= CALL_RATE_MIN &
  locus_audit$N_genotype_classes >= 2

# Retain in the SSR matrix only loci that pass the quality-control criteria.
ssr <- ssr_raw[
  ,
  locus_audit$Keep_locus,
  drop = FALSE
]

kable(
  locus_audit,
  digits = 3,
  caption = "Quality control of SSR loci"
)
```

The figure shows the call rate of every locus relative to the minimum threshold.

```{r marker-quality-plot, fig.height=7}
# Organize marker information and compute the diagnostics required at this stage.
ggplot(
  locus_audit,
  aes(
    x = reorder(Locus, Call_rate),
    y = Call_rate,
    fill = Keep_locus
  )
) +
  geom_col() +
  geom_hline(
    # Define the `yintercept` values used in this structure.
    yintercept = CALL_RATE_MIN,
    linetype = 2
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Call rate",
    fill = "Retained",
    title = "Call rate of SSR loci"
  ) +
  theme_bw()
```

## 7. Summary of genotype information

The amount of available SSR information is summarized for every individual.

```{r individual-summary}
# Assemble the summary table presented in this section.
individual_audit <- data.frame(
  IND = rownames(ssr_raw),
  N_called = rowSums(!is.na(ssr_raw)),
  Call_rate = rowMeans(!is.na(ssr_raw))
)

rownames(individual_audit) <- NULL

individual_call_rate_summary <- data.frame(
  Measure = c("Minimum", "Mean", "Median", "Maximum"),
  # List the call-rate values used in the individual summary.
  Call_rate = c(
    min(individual_audit$Call_rate),
    mean(individual_audit$Call_rate),
    median(individual_audit$Call_rate),
    max(individual_audit$Call_rate)
  )
)

kable(
  individual_call_rate_summary,
  digits = 3,
  caption = "Distribution of SSR call rate by individual"
)
```

## 8. Summary of phenotype traits

Descriptive statistics allow direct inspection of the scale and variation of
the ten traits.

```{r phenotype-summary}
# Organize the phenotype data and apply the preprocessing rules defined for this stage.
phenotype_sd <- numeric(ncol(phenotypes))
phenotype_minimum <- numeric(ncol(phenotypes))
phenotype_maximum <- numeric(ncol(phenotypes))

for (j in seq_len(ncol(phenotypes))) {
  phenotype_sd[j] <- sd(phenotypes[, j])
  phenotype_minimum[j] <- min(phenotypes[, j])
  phenotype_maximum[j] <- max(phenotypes[, j])
}

# Summarize the results in `phenotype_summary`.
phenotype_summary <- data.frame(
  Trait = colnames(phenotypes),
  N = colSums(!is.na(phenotypes)),
  Mean = colMeans(phenotypes),
  SD = phenotype_sd,
  Minimum = phenotype_minimum,
  Maximum = phenotype_maximum
)

rownames(phenotype_summary) <- NULL

kable(
  phenotype_summary,
  digits = 3,
  caption = "Summary of phenotype traits"
)
```

### 8.1 Phenotype observations far from the mean

Standardized values are used only as a descriptive diagnostic. An observation
is highlighted when its absolute distance from the trait mean is at least
`r PHENOTYPE_DISTANCE_THRESHOLD` standard deviations. Values are retained
exactly as recorded in the original spreadsheet.

```{r distant-phenotype-values}
# Organize the phenotype data and apply the preprocessing rules defined for this stage.
phenotype_means <- colMeans(phenotypes)
phenotype_standard_deviations <- numeric(ncol(phenotypes))

for (j in seq_len(ncol(phenotypes))) {
  phenotype_standard_deviations[j] <- sd(phenotypes[, j])
}

phenotype_standardized <- scale(
  phenotypes,
  center = phenotype_means,
  scale = phenotype_standard_deviations
)

# Locate observations whose standardized distance reaches the defined threshold.
extreme_positions <- which(
  abs(phenotype_standardized) >= PHENOTYPE_DISTANCE_THRESHOLD,
  arr.ind = TRUE
)

phenotype_extremes <- data.frame(
  IND = rownames(phenotypes)[extreme_positions[, "row"]],
  Trait = colnames(phenotypes)[extreme_positions[, "col"]],
  Value = phenotypes[extreme_positions],
  Trait_mean = phenotype_means[extreme_positions[, "col"]],
  Trait_SD = phenotype_standard_deviations[extreme_positions[, "col"]],
  Standardized_value = phenotype_standardized[extreme_positions],
  Absolute_standardized_distance = abs(
    phenotype_standardized[extreme_positions]
  )
)

# Organize the phenotypes used in the analysis.
phenotype_extremes <- phenotype_extremes[
  order(-phenotype_extremes$Absolute_standardized_distance),
]
rownames(phenotype_extremes) <- NULL

kable(
  phenotype_extremes,
  digits = 3,
  # Define the headers displayed in the table of highlighted phenotype observations.
  col.names = c(
    "Individual",
    "Trait",
    "Value",
    "Trait mean",
    "Trait SD",
    "Standardized value",
    "Absolute standardized distance"
  ),
  caption = paste0(
    "Phenotype observations at least ",
    PHENOTYPE_DISTANCE_THRESHOLD,
    " standard deviations from the mean"
  )
)
```

### 8.2 Trait dictionary

The next step organizes a tabular summary of the objects and results needed to interpret this section.

```{r trait-dictionary}
# Associate each trait with its English and Portuguese descriptions and measurement unit.
trait_dictionary <- data.frame(
  Trait = TRAITS,
  Description_en = c(
    "Mean mass of commercial fruit",
    "Production of commercial fruit",
    "Production of defective fruit",
    "Fruit length",
    "Fruit diameter",
    "Internal-cavity diameter",
    "Mean pulp thickness",
    "External fruit firmness",
    "Internal pulp firmness",
    "Soluble solids"
  ),
  # Record the Portuguese description of each trait.
  Description_pt = c(
    "Massa média de frutos comerciais",
    "Produção de frutos comerciais",
    "Produção de frutos defeituosos",
    "Comprimento do fruto",
    "Diâmetro do fruto",
    "Diâmetro da cavidade interna",
    "Espessura média da polpa",
    "Firmeza externa do fruto",
    "Firmeza interna da polpa",
    "Sólidos solúveis"
  ),
  # Record the measurement unit of each trait.
  Unit = c(
    "kg", "kg", "kg", "cm", "cm",
    "cm", "cm", "N", "N", "°Brix"
  )
)

rownames(trait_dictionary) <- NULL

kable(
  trait_dictionary[, c("Trait", "Description_en", "Unit")],
  caption = "Dictionary of phenotype traits"
)
```

The next step builds the corresponding visualization to support interpretation of the diagnostic.

```{r phenotype-plot, fig.height=8}
# Organize the phenotype data and apply the preprocessing rules defined for this stage.
phenotype_long <- data.frame(
  IND = rownames(phenotypes),
  phenotypes,
  check.names = FALSE
) |>
  pivot_longer(
    cols = -IND,
    names_to = "Trait",
    values_to = "Value"
  )

# Build the figure used for interpretation.
ggplot(phenotype_long, aes(x = Value)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ Trait, scales = "free", ncol = 2) +
  labs(
    x = "Observed value",
    y = "Number of individuals",
    title = "Phenotype distributions"
  ) +
  theme_bw()
```

## 9. Saving preprocessing results

The aligned phenotype matrix, SSR data, metadata, and summaries are gathered
into an object used to construct the genomic matrices in the next module.

```{r save-results}
# Collect settings, results, and diagnostics in the final module object.
preprocessing_object <- list(
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,
  ssr_raw = ssr_raw,
  ssr = ssr,
  locus_audit = locus_audit,
  individual_audit = individual_audit,
  phenotype_summary = phenotype_summary,
  phenotype_extremes = phenotype_extremes,
  trait_dictionary = trait_dictionary,
  # Record parameters and thresholds required to reproduce this stage.
  settings = list(
    call_rate_min = CALL_RATE_MIN,
    phenotype_distance_threshold = PHENOTYPE_DISTANCE_THRESHOLD,
    fixed_effects = "FAM"
  )
)

saveRDS(
  preprocessing_object,
  file.path(OUTPUT_DIR, "preprocessing_object.rds")
)

# Save the consolidated output for downstream use.
write.csv(alignment, file.path(OUTPUT_DIR, "alignment_summary.csv"), row.names = FALSE)
write.csv(locus_audit, file.path(OUTPUT_DIR, "locus_audit.csv"), row.names = FALSE)
write.csv(individual_audit, file.path(OUTPUT_DIR, "individual_audit.csv"), row.names = FALSE)
write.csv(phenotype_summary, file.path(OUTPUT_DIR, "phenotype_summary.csv"), row.names = FALSE)
write.csv(
  phenotype_extremes,
  file.path(OUTPUT_DIR, "phenotype_extreme_values.csv"),
  row.names = FALSE
)
write.csv(trait_dictionary, file.path(OUTPUT_DIR, "trait_dictionary.csv"), row.names = FALSE)
```

The final structure of this step can be inspected directly from the dimensions
and counts below.

```{r final-summary}
# Assemble the summary table presented in this section.
data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Original SSR loci",
    "Retained SSR loci",
    "Families",
    "Missing SSR calls",
    "Highlighted phenotype observations"
  ),
  # List the values corresponding to the table entries.
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    ncol(ssr),
    nlevels(metadata$FAM),
    sum(is.na(ssr_raw)),
    nrow(phenotype_extremes)
  )
) |>
  kable(caption = "Final preprocessing summary")
```

## 10. Next step

Module 02 converts the retained SSR loci into allele-dosage columns, constructs
the marker and genomic relationship matrices, creates the family fixed-effect
matrix, and defines the five cross-validation folds.
