---
title: "2. Bayesian genomic regression"
author: "Weverton Gomes da Costa"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
bibliography: ../REFERENCES.bib
link-citations: true
---

[Português](02_bayesian_models_pt.html) | [Previous](01_preprocessing_en.html) | [Home](index.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 6)
source(file.path("code", "00_config.R"))
source(file.path("code", "01_data_functions.R"))
source(file.path("code", "02_model_functions.R"))
source(file.path("code", "03_reporting_functions.R"))
configuracao <- obter_configuracao()
objeto <- carregar_resultado(configuracao$arquivo_preprocessamento)
if (is.null(objeto)) objeto <- executar_preprocessamento(configuracao)
```

## Model specification

The six Bayesian models follow the whole-genome regression framework used in
the reference study [@silva2021brrguava] and are fitted with BGLR
[@perez2014bglr]. For one trait at a time,

$$
\mathbf{y}=\mathbf{1}\mu+\mathbf{X}\boldsymbol{\beta}+
\mathbf{M}\boldsymbol{g}+\boldsymbol{e},
$$

where $\mathbf{X}$ contains categorical block and family effects,
$\mathbf{M}$ contains centered SSR allele dosages, and $\boldsymbol{g}$ is the
model-specific vector of marker effects. BGLR supplies the intercept, while
the explicit fixed-effect matrix excludes a redundant intercept column.

```{r bayesian-catalog, results='asis'}
catalogo_exibicao <- configuracao$catalogo_bayesiano |>
  dplyr::mutate(
    Article_zero_effect_probability = dplyr::if_else(
      Model == "BayesB_pi1e-5", "1e-5", "default"
    ),
    BGLR_probIn = dplyr::if_else(
      Model == "BayesB_pi1e-5", format(probIn, digits = 6), "default"
    )
  ) |>
  dplyr::select(Model, BGLR_model, Article_zero_effect_probability, BGLR_probIn)
print(tabela_relatorio(catalogo_exibicao, "Bayesian model catalogue"))
```

Bayesian ridge regression uses a common Gaussian shrinkage variance. BayesA
uses marker-specific scaled-$t$ effects; BayesB combines marker-specific
effects with variable selection; BayesC uses a common slab variance with
variable selection; and Bayesian Lasso uses a double-exponential shrinkage
prior. The models differ in their priors, not in their validation folds.

## BayesB `pi = 1e-5` convention

The article describes `pi = 1e-5` as the probability of a zero marker effect.
In BGLR, `probIn` is the probability that an effect is non-zero. The dedicated
model therefore uses `probIn = 1 - 1e-5` and `counts = 1e6`. This explicit
translation prevents reversing the prior meaning.

## MCMC profiles

Website builds use the `site` profile and do not start MCMC. The `smoke`
profile tests the complete integration with short chains. The `full` profile
uses the 1,000,000 iterations, 200,000 burn-in iterations, and thinning of 4
reported in the main methods text of Silva et al. [-@silva2021brrguava].

```{r mcmc-configuration, results='asis'}
perfis <- data.frame(
  Profile = c("site", "smoke", "full"),
  Execute = c(FALSE, TRUE, TRUE),
  Iterations = c(NA, 5000, 1000000),
  Burn_in = c(NA, 1000, 200000),
  Thin = c(NA, 5, 4)
)
print(tabela_relatorio(perfis, "Controlled execution profiles", digitos = 0))
```

## Controlled execution

The English page owns the computational step. It runs only when
`PAPAYA_RUN_MODE` is `smoke` or `full`; otherwise it loads a previously saved
result. Each model uses exactly the training-specific marker transformation
created on the preprocessing page.

```{r fit-bayesian-models}
if (isTRUE(configuracao$mcmc$executar)) {
  resultado_bayesiano <- executar_modelos_bayesianos(objeto, configuracao)
} else {
  resultado_bayesiano <- carregar_resultado(configuracao$arquivo_bayesiano)
}
```

```{r bayesian-status, results='asis', echo=FALSE}
if (is.null(resultado_bayesiano)) {
  cat(mensagem_resultado_ausente(configuracao$arquivo_bayesiano, "en"))
}
```

## Cross-validated prediction

Predictive correlation, RMSE, MAE, bias, and calibration are calculated in
each fold. The table summarizes fold-level performance; final cross-model
ranking is postponed to the common out-of-fold comparison on page 5.

```{r bayesian-prediction-summary, results='asis'}
if (!is.null(resultado_bayesiano)) {
  resumo_bayesiano <- resumir_metricas_folds(resultado_bayesiano$fold_metrics) |>
    dplyr::arrange(Trait, dplyr::desc(Mean_correlation))
  print(tabela_relatorio(
    resumo_bayesiano,
    "Fold-level predictive performance of Bayesian models",
    rolagem = TRUE
  ))
}
```

When results are available, the plot shows mean predictive correlation across
the eight common folds for every trait and Bayesian prior.

```{r bayesian-correlation-plot, echo=FALSE}
if (!is.null(resultado_bayesiano)) {
  resumo_bayesiano <- resumir_metricas_folds(resultado_bayesiano$fold_metrics)
  print(
    ggplot2::ggplot(
      resumo_bayesiano,
      ggplot2::aes(x = Model, y = Mean_correlation, color = Model, group = Model)
    ) +
      ggplot2::geom_hline(yintercept = 0, color = "grey70") +
      ggplot2::geom_point(size = 2) +
      ggplot2::facet_wrap(~ Trait, scales = "free_y", ncol = 2) +
      ggplot2::labs(
        x = NULL,
        y = "Mean fold correlation",
        title = "Bayesian predictive correlation by trait"
      ) +
      tema_relatorio() +
      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1))
  )
}
```

## DIC within the Bayesian family

DIC, effective parameter count, $\Delta$DIC, normalized DIC weights, and
evidence ratios come from the full-data Bayesian fits. These quantities are
reported only among Bayesian models and are not treated as a common criterion
for RR-BLUP or GBLUP.

```{r dic-table, results='asis'}
if (!is.null(resultado_bayesiano)) {
  dic <- resultado_bayesiano$full_model_summary |>
    dplyr::arrange(Trait, DIC) |>
    dplyr::select(
      Trait, Model, DIC, Effective_parameters, Delta_DIC,
      DIC_weight, Evidence_ratio, Residual_variance, H2
    )
  print(tabela_relatorio(dic, "Bayesian DIC diagnostics", rolagem = TRUE))
}
```

BRR trace files from each full-data trait fit are retained in the saved result
for convergence inspection. Short `smoke` chains validate software behavior;
they must not be used for biological inference.

[Next: RR-BLUP](03_rrblup_en.html)
