---
title: "Construction of matrices for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

In this module, the preprocessed SSR genotypes are converted into allele
dosages. These dosages are used to construct the centered marker matrix `M`,
the additive genomic relationship matrix `G`, and the fixed-effect matrix `X`,
which contains family in addition to the intercept. Five family-stratified
cross-validation folds are also defined.

The matrices are constructed once using all 150 individuals and remain
unchanged across folds. In Module 03, only the phenotype vector is masked to
form training and validation sets.

Because candidates in every fold are already genotyped, marker filtering,
allele frequencies, mean imputation, `M`, and `G` are defined from the full
genotype panel and then held fixed. No validation phenotype enters these
steps. This corresponds to a transductive candidate-genotype setting and
should not be interpreted as external validation in a new genotype population.

The module produces four components used directly in Module 03: the fixed-effect matrix `X`, the centered marker matrix `M`, the genomic relationship matrix `G`, and the `cv_folds` table identifying each individual's validation fold. The diagnostics check allele coding, centering, genomic relationships, and family stratification.

## 2. Packages and input object

The packages used in this module have distinct roles:

- `dplyr`: selects, transforms, and summarizes tabular data;
- `tidyr`: reshapes data for diagnostics and figures;
- `ggplot2`: builds allele-frequency, fold-distribution, and genomic-relationship figures;
- `knitr`: formats tables in R Markdown documents;
- `here`: builds project-relative file paths that do not depend on the local working directory.

```{r packages, warning=FALSE, message=FALSE}
library(dplyr)   # Manipulate, transform, and summarize tables.
library(tidyr)   # Reshape data between formats.
library(ggplot2) # Build figures and heatmaps.
library(knitr)   # Format tables and R Markdown output.
library(here)    # Build portable project-relative paths.
```

```{r table-formatting, include=FALSE}
format_table <- function(x, caption = NULL, digits = getOption("digits"), col.names = NA) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    col.names = col.names,
    table.attr = paste0(
      'class="table table-striped table-hover" ',
      'style="width:100%; margin-bottom:0;"'
    )
  )
  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

The `preprocessing_object.rds` file produced in Module 01 supplies the individuals, phenotypes, families, and SSR loci used to construct the matrices.

Three parameters are defined in this module:

- `N_FOLDS = 5`: divides the 150 individuals into five validation groups of 30 individuals, leaving 120 training individuals per fold. Each individual appears in validation exactly once; this is not repeated cross-validation.
- `SEED_CV = 20260723`: reproduces exactly the same family-stratified allocation across folds. The seed controls partition reproducibility only and has no inferential interpretation.
- `ALLELE_FREQUENCY_MIN = 0.05`: retains allele-dosage columns with frequencies between 0.05 and 0.95 provided that dosage variation is positive. Very rare or nearly fixed alleles therefore do not enter `M`.

The call-rate threshold is recovered from Module 01 because locus filtering has already been applied before allele-dosage construction.

```{r input-object}
# Define paths and load the preprocessing object used to construct the genomic matrices.
PREPROCESSING_FILE <- here(
  "output",
  "preprocessing",
  "preprocessing_object.rds"
)

OUTPUT_DIR <- here(
  "output",
  "matrices"
)

# Use five folds: 30 validation and 120 training individuals per fold.
N_FOLDS <- 5L
# Fix only the reproducibility of the family-stratified allocation.
SEED_CV <- 20260723L
# Exclude alleles with frequency below 5% or above 95%.
ALLELE_FREQUENCY_MIN <- 0.05

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

preprocessing <- readRDS(PREPROCESSING_FILE)

ids <- preprocessing$ids
metadata <- preprocessing$metadata
phenotypes <- preprocessing$phenotypes
ssr_raw <- preprocessing$ssr_raw
locus_audit_full <- preprocessing$locus_audit
trait_dictionary <- preprocessing$trait_dictionary
# Recover, for provenance, the call-rate threshold applied in Module 01.
CALL_RATE_MIN <- preprocessing$settings$call_rate_min
```

Before marker coding, the following table checks the structure received from Module 01: numbers of individuals and traits, loci before and after quality control, and families. The comparison confirms that the input object has the expected structure; no new filter is applied at this stage.

```{r input-summary}
# Summarize the dimensions and categories of the preprocessing object received by the module.
data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Original SSR loci",
    "Retained SSR loci",
    "Families"
  ),
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    sum(locus_audit_full$Keep_locus),
    nlevels(metadata$FAM)
  )
) |>
  format_table(caption = "Input data for matrix construction")
```

The check shows `r length(ids)` individuals, `r ncol(phenotypes)` phenotype traits, `r ncol(ssr_raw)` original SSR loci, and `r sum(locus_audit_full$Keep_locus)` loci retained after Module 01 quality control. Family remains represented by `r nlevels(metadata$FAM)` levels of `FAM`.

## 3. Allele-dosage coding

Each SSR code contains two allele labels. For example, `12` represents one copy of allele 1 and one copy of allele 2, whereas `22` represents two copies of allele 2. For every observed allele at each retained locus, an individual's dosage is computed as the sum of two indicator functions, one for each allele position in the genotype. The resulting value is 0, 1, or 2 according to whether the individual carries zero, one, or two copies of that allele.

One column is created for every locus–allele combination, and missing SSR calls remain `NA` until imputation. Before any allele-column filtering, dosages across all alleles at a diploid locus sum to 2 for every observed genotype. This property is later used as a coding diagnostic.

The coding follows a sequence that maps directly to the operations in the code:

1. **Select retained loci.** `locus_audit_full$Locus[locus_audit_full$Keep_locus]` uses the logical vector `Keep_locus` to recover only locus names that passed quality control. Then `ssr_raw[, retained_loci_full, drop = FALSE]` keeps every row and only those columns; `drop = FALSE` preserves a two-dimensional structure even when few columns are selected.
2. **Separate the two alleles in an SSR code.** For a code such as `12`, `floor(codes / 10)` returns the tens digit, the first allele (`1`). The `%%` operator returns the remainder of integer division, so `codes %% 10` returns the units digit, the second allele (`2`).
3. **Count copies of each allele.** The comparisons `allele_1 == allele` and `allele_2 == allele` indicate whether the allele occurs in each position. `as.numeric()` converts `TRUE` to 1 and `FALSE` to 0; summing both positions yields dosage 0, 1, or 2. If the SSR code is missing, dosage remains `NA`.
4. **Assemble the matrix.** Each dosage vector is stored in a list. At the end, `do.call(cbind, dosage_parts_full)` combines these vectors as columns of one matrix, while `bind_rows(allele_map_parts_full)` combines the metadata linking each column to its source locus and allele.

```{r allele-dosage-coding}
# Organize marker information and compute the diagnostics required at this stage.
retained_loci_full <- locus_audit_full$Locus[
  locus_audit_full$Keep_locus
]

ssr_full <- ssr_raw[
  ,
  retained_loci_full,
  drop = FALSE
]

dosage_parts_full <- list()
allele_map_parts_full <- list()
marker_counter_full <- 0L

for (j in seq_len(ncol(ssr_full))) {
  # Separate the two alleles encoded in each SSR genotype.
  codes <- ssr_full[, j]
  allele_1 <- floor(codes / 10)
  allele_2 <- codes %% 10

  # List the alleles observed at the current locus.
  observed_alleles <- sort(
    unique(
      c(
        allele_1[!is.na(codes)],
        allele_2[!is.na(codes)]
      )
    )
  )

  for (allele in observed_alleles) {
    marker_counter_full <- marker_counter_full + 1L

    # Count 0, 1, or 2 copies of the allele for each individual.
    dosage <-
      as.numeric(allele_1 == allele) +
      as.numeric(allele_2 == allele)
    dosage[is.na(codes)] <- NA_real_

    # Create a unique identifier for each locus-allele combination.
    marker_name <- paste0(
      colnames(ssr_full)[j],
      "__A",
      allele
    )

    dosage_parts_full[[marker_counter_full]] <- dosage
    allele_map_parts_full[[marker_counter_full]] <- data.frame(
      Locus = colnames(ssr_full)[j],
      Allele = allele,
      Marker = marker_name
    )
  }
}

dosage_full_raw <- do.call(cbind, dosage_parts_full)
allele_audit_full <- bind_rows(allele_map_parts_full)

rownames(dosage_full_raw) <- ids
colnames(dosage_full_raw) <- allele_audit_full$Marker
rownames(allele_audit_full) <- NULL

# Display a preview of the resulting dosage columns.
as.data.frame(dosage_full_raw[1:6, 1:8]) |>
  format_table(caption = "Preview of allele dosages")
```

The coding produces `r ncol(dosage_full_raw)` dosage columns from the `r ncol(ssr_full)` retained loci. Within each column, values 0, 1, and 2 give the number of copies of the allele identified after `__A`; `NA` denotes a missing SSR call. One locus can produce more than two columns when more than two alleles are represented in the panel.

## 4. Allele frequencies and retention of dosage columns

The frequency of an allele column is computed as the mean observed dosage divided by 2 because each diploid individual can contribute two copies of the allele. `NA` values are ignored here, so the frequency uses available genotype calls only.

Columns are retained when frequency lies between 0.05 and 0.95 and dosage standard deviation is positive. The lower and upper limits are symmetric: they remove very rare alleles and nearly fixed columns, while `Dosage_sd > 0` guarantees observed variation. The filter acts on allele-dosage columns; it does not remove individuals.

In the code, `colMeans(dosage_full_raw, na.rm = TRUE) / 2` computes frequencies for all columns at once: `na.rm = TRUE` excludes missing calls from the mean, and division by 2 converts mean dosage, which ranges from 0 to 2, into allele frequency between 0 and 1. The loop with `sd(..., na.rm = TRUE)` computes the standard deviation of each column. Finally, `is.finite()` prevents non-finite results from being accepted before the frequency limits and `Dosage_sd > 0` condition are applied.

```{r allele-frequency-control}
# Compute dosage frequency and variability to retain informative allele columns.
allele_audit_full$Allele_frequency <-
  colMeans(dosage_full_raw, na.rm = TRUE) / 2

dosage_sd_full <- numeric(ncol(dosage_full_raw))

for (j in seq_len(ncol(dosage_full_raw))) {
  dosage_sd_full[j] <- sd(
    dosage_full_raw[, j],
    na.rm = TRUE
  )
}

allele_audit_full$Dosage_sd <- dosage_sd_full

allele_audit_full$Keep_allele <-
  is.finite(allele_audit_full$Allele_frequency) &
  allele_audit_full$Allele_frequency >= ALLELE_FREQUENCY_MIN &
  allele_audit_full$Allele_frequency <= (1 - ALLELE_FREQUENCY_MIN) &
  is.finite(allele_audit_full$Dosage_sd) &
  allele_audit_full$Dosage_sd > 0

# Select allele columns that pass the frequency and dosage-variation filters.
retained_markers_full <- allele_audit_full$Marker[
  allele_audit_full$Keep_allele
]

dosage_full <- dosage_full_raw[
  ,
  retained_markers_full,
  drop = FALSE
]

# Summarize the results in `allele_summary_full`.
allele_summary_full <- allele_audit_full |>
  group_by(Locus) |>
  summarise(
    Observed_alleles = n(),
    Retained_alleles = sum(Keep_allele),
    .groups = "drop"
  )

format_table(
  allele_summary_full,
  caption = "Allele-dosage columns by SSR locus"
)
```

Of the `r nrow(allele_audit_full)` initially coded allele columns, `r sum(allele_audit_full$Keep_allele)` satisfy both the frequency interval and positive-variation requirement. The table shows how many alleles were observed and retained at each locus.

The histogram shows where allele-column frequencies are concentrated. Vertical lines mark 0.05 and 0.95, and fill indicates the final retention decision, which also requires positive dosage variation. The 20 bins are only a visual-resolution choice and do not participate in filtering.

```{r allele-frequency-plot, fig.height=5}
# Display the allele-frequency distribution and the thresholds used by the filter.
ggplot(
  allele_audit_full,
  aes(
    x = Allele_frequency,
    fill = Keep_allele
  )
) +
  geom_histogram(
    # Use 20 bins only to standardize the visual resolution of the histogram.
    bins = 20,
    position = "identity",
    alpha = 0.7
  ) +
  geom_vline(
    # Mark the lower and upper allele-frequency thresholds in the figure.
    xintercept = c(
      ALLELE_FREQUENCY_MIN,
      1 - ALLELE_FREQUENCY_MIN
    ),
    linetype = 2
  ) +
  labs(
    x = "Allele frequency",
    y = "Allele-dosage columns",
    fill = "Retained",
    title = "Distribution of allele frequencies"
  ) +
  theme_bw()
```

Observed allele frequencies range from `r sprintf("%.3f", min(allele_audit_full$Allele_frequency, na.rm = TRUE))` to `r sprintf("%.3f", max(allele_audit_full$Allele_frequency, na.rm = TRUE))`. Bars classified as not retained fail at least one of the defined criteria; the dashed lines display the frequency limits only.

## 5. Marker matrix

Missing dosages are replaced by the mean of the corresponding column computed from the full genotype panel. `marker_means_full <- colMeans(dosage_full, na.rm = TRUE)` computes one mean per column. Inside the loop, `is.na(dosage_full_imputed[, j])` identifies missing positions in column `j`, which receive `marker_means_full[j]`.

After imputation, `sweep(dosage_full_imputed, 2, marker_means_full, FUN = "-")` centers the dosages. The argument `2` specifies a column-wise operation, and `FUN = "-"` subtracts the corresponding mean from every value. Because the same mean was used for imputation, a mean-imputed dosage becomes exactly zero in `M`: it represents no deviation from the column average rather than a new genotype class.

The resulting matrix is denoted by \(\mathbf M\). Its columns are centered but are not standardized to unit variance. Imputation uses genotype information from all 150 candidates; no validation phenotype enters this calculation.

```{r marker-matrix}
# Impute missing dosages by column means, center the columns, and construct marker matrix M.
marker_means_full <- colMeans(
  dosage_full,
  na.rm = TRUE
)

dosage_full_imputed <- dosage_full

for (j in seq_len(ncol(dosage_full_imputed))) {
  # Locate missing dosages in the current column.
  missing_values <- is.na(dosage_full_imputed[, j])

  # Impute the column mean.
  dosage_full_imputed[
    missing_values,
    j
  ] <- marker_means_full[j]
}

# Center every dosage column.
M_full <- sweep(
  dosage_full_imputed,
  2,
  marker_means_full,
  FUN = "-"
)
```

The summary of `M` checks four basic properties. Its row count should match the 150 candidates; its column count is the number of retained allele-dosage columns; its rank reveals linear dependencies among columns; and the maximum absolute column mean should be close to zero, confirming numerical centering. The expression `qr(M_full)$rank` obtains rank from the QR decomposition: this value is the number of linearly independent directions represented by the columns of `M`.

```{r marker-matrix-summary}
data.frame(
  Rows = nrow(M_full),
  Columns = ncol(M_full),
  Rank = qr(M_full)$rank,
  Maximum_absolute_column_mean = max(abs(colMeans(M_full)))
) |>
  format_table(
    digits = 8,
    caption = "Dimensions and centering of the marker matrix"
  )
```

Matrix `M` contains `r nrow(M_full)` rows and `r ncol(M_full)` retained dosage columns. Its rank is `r qr(M_full)$rank`, below the number of columns because dosage columns from the same locus can be linearly dependent. The largest absolute column mean is `r format(max(abs(colMeans(M_full))), scientific = TRUE, digits = 3)`, confirming numerical centering.

## 6. Genomic relationship matrix

Let \(p_j\) be the allele frequency represented by column \(j\). Using a
VanRaden-type allele-frequency scaling [@vanraden2008efficient], the additive
genomic relationship matrix is calculated as

\[
\mathbf G =
\frac{\mathbf M\mathbf M^{\mathsf T}}
{\sum_j 2p_j(1-p_j)}.
\]

In the code, `allele_frequencies_full <- marker_means_full / 2` recovers \(p_j\) because mean dosage equals \(2p_j\). `tcrossprod(M_full)` directly computes \(\mathbf M\mathbf M^{\mathsf T}\), a product among individuals based on the centered columns. Dividing by `denominator_full` applies the scale defined in the equation. Then `(G_full + t(G_full)) / 2` removes only small numerical differences between symmetric elements, and `eigen(..., symmetric = TRUE, only.values = TRUE)` obtains the eigenvalues used to diagnose positive semidefiniteness.

```{r genomic-relationship-matrix}
# Recover p_j from the dosage mean, which equals 2p_j.
allele_frequencies_full <- marker_means_full / 2

# Compute the VanRaden scaling from allele-column frequencies.
denominator_full <- sum(
  2 *
    allele_frequencies_full *
    (1 - allele_frequencies_full)
)

# Convert cross-products of centered dosages into scaled genomic relationships.
G_full <- tcrossprod(M_full) / denominator_full

# Remove only numerical asymmetry.
G_full <- (G_full + t(G_full)) / 2

dimnames(G_full) <- list(ids, ids)

eigenvalues_G_full <- eigen(
  G_full,
  symmetric = TRUE,
  only.values = TRUE
)$values
```

The denominator reports the scale used to construct `G`. The mean diagonal summarizes each individual's relationship with itself on this scale and is not forced to equal exactly 1. The symmetry difference should be close to zero. Because `G` is built from a cross-product, it is theoretically positive semidefinite; therefore, a slightly negative smallest eigenvalue, when limited to numerical precision, should be interpreted as rounding rather than evidence of negative relatedness.

```{r genomic-relationship-summary}
# Build or check the matrices used for genomic prediction.
data.frame(
  Quantity = c(
    "Denominator sum 2p(1-p)",
    "Mean diagonal",
    "Maximum symmetry difference",
    "Smallest eigenvalue"
  ),
  Value = c(
    denominator_full,
    mean(diag(G_full)),
    max(abs(G_full - t(G_full))),
    min(eigenvalues_G_full)
  )
) |>
  format_table(
    digits = 8,
    caption = "Numerical summary of the genomic relationship matrix"
  )
```

Matrix `G` is numerically symmetric, with a maximum difference of `r format(max(abs(G_full - t(G_full))), scientific = TRUE, digits = 3)`. Its mean diagonal is `r sprintf("%.3f", mean(diag(G_full)))`, and its smallest eigenvalue is `r format(min(eigenvalues_G_full), scientific = TRUE, digits = 3)`. Negative values of very small magnitude are compatible with numerical rounding.

## 7. Fixed-effect matrix

Family is the only categorical fixed effect. With `~ FAM`, `model.matrix()` uses R's default treatment contrasts: it creates an intercept and enough indicator columns to represent differences among family levels relative to a reference category. Thus, `X` incorporates family structure without treating `FAM` as a numeric variable.

In Module 03, the intercept column is removed from `X` because BGLR already fits an intercept automatically; retaining both would duplicate the same constant effect.

```{r fixed-effect-matrix}
# Build the design matrix for the only categorical fixed effect used in the analysis.
X <- model.matrix(
  ~ FAM,
  data = metadata
)

rownames(X) <- ids

data.frame(
  Matrix = c("X", "M", "G"),
  Rows = c(nrow(X), nrow(M_full), nrow(G_full)),
  Columns = c(ncol(X), ncol(M_full), ncol(G_full)),
  Rank = c(qr(X)$rank, qr(M_full)$rank, qr(G_full)$rank)
) |>
  format_table(caption = "Dimensions and ranks of the analytical matrices")
```

`X`, `M`, and `G` all have `r length(ids)` rows, preserving the same individual order. `X` contains the intercept and columns required to represent `FAM`; `M` contains centered dosages; and `G` is the square relationship matrix among all individuals.

## 8. Five family-stratified folds

Family stratification targets prediction among individuals from the represented
breeding population. It is not a leave-family-out design and therefore does
not estimate predictive performance for entirely new families.

The 150 individuals are divided into five validation groups of 30 individuals. The construction proceeds family by family and can be followed directly through the functions used in the code:

1. `which(family == family_level)` locates individuals from the current family, and `sample()` shuffles their positions.
2. `seq_len(N_FOLDS)` generates the five fold numbers. A second `sample()` call creates a random order used only to break ties among folds with equal accumulated load.
3. `order(fold_load, match(seq_len(N_FOLDS), random_tie_order))` prioritizes less-filled folds first and uses the random order only when two loads are equal.
4. `rep(ordered_folds, length.out = length(family_indices))` cycles through this order until one fold label is produced for every family member.
5. `tabulate(family_labels, nbins = N_FOLDS)` counts how many individuals were just assigned to each fold and updates `fold_load`.

This procedure spreads members of each family across folds while keeping total fold sizes balanced. After allocation, `apply(family_fold_table, 1, max)` and `apply(family_fold_table, 1, min)` compute, for each row of the table — that is, for each family — the largest and smallest counts across folds. The argument `1` specifies a row-wise operation; the difference between maximum and minimum forms `family_balance`.

The fixed seed preserves exactly the same partition across executions. The same fold table is used for all ten traits and all methods, so differences among models cannot arise from different validation partitions.

```{r cross-validation-folds}
# Initialize the stratified partition and track each fold's accumulated total load.
set.seed(SEED_CV)

family <- droplevels(metadata$FAM)
folds <- integer(length(ids))
fold_load <- integer(N_FOLDS)

for (family_level in levels(family)) {
  # Randomize the individuals of the current family.
  family_indices <- sample(
    which(family == family_level)
  )

  # Define a random order to break ties between equally loaded folds.
  random_tie_order <- sample(seq_len(N_FOLDS))

  ordered_folds <- order(
    fold_load,
    match(seq_len(N_FOLDS), random_tie_order)
  )

  # Distribute family members across the five folds.
  family_labels <- rep(
    ordered_folds,
    length.out = length(family_indices)
  )

  folds[family_indices] <- family_labels

  fold_load <- fold_load + tabulate(
    family_labels,
    nbins = N_FOLDS
  )
}

fold_sizes <- tabulate(folds, nbins = N_FOLDS)

family_fold_table <- table(
  Family = family,
  Fold = folds
)

# Quantify, within each family, the largest difference in individual counts across folds.
family_balance <- apply(family_fold_table, 1, max) -
  apply(family_fold_table, 1, min)

cv_folds <- data.frame(
  IND = ids,
  FAM = as.character(metadata$FAM),
  Fold = folds
)

fold_summary <- data.frame(
  Fold = seq_len(N_FOLDS),
  Training_n = length(ids) - fold_sizes,
  Validation_n = fold_sizes
)

format_table(
  fold_summary,
  col.names = c("Fold", "Training N", "Validation N"),
  caption = "Sample sizes in cross-validation"
)
```

All five folds contain `r unique(fold_summary$Validation_n)` validation individuals and `r unique(fold_summary$Training_n)` training individuals. Sample size is therefore constant across the five rounds.

The family × fold table shows the resulting stratification directly. Within each family, counts should be as balanced as its sample size allows; `family_balance`, computed previously, measures the difference between the largest and smallest count for that family. Because members are distributed cyclically, this difference is at most one individual for every family.

```{r families-by-fold}
format_table(
  as.data.frame.matrix(family_fold_table),
  caption = "Number of individuals from each family in each fold"
)
```

The largest difference between the maximum and minimum count for the same family across folds is `r max(family_balance)` individual. This confirms that stratification distributes each family as evenly as its size permits.

The heatmap displays the same family × fold table visually: every cell explicitly shows the number of individuals, while fill makes imbalances easier to spot. It is a diagnostic of the partition, not an additional measure of predictive performance.

```{r family-fold-plot, fig.height=7}
# Reshape the family-by-fold table and build the individual-distribution figure.
family_fold_long <- as.data.frame(family_fold_table)

ggplot(
  family_fold_long,
  aes(
    x = factor(Fold),
    y = Family,
    fill = Freq
  )
) +
  geom_tile() +
  geom_text(aes(label = Freq)) +
  labs(
    x = "Fold",
    y = "Family",
    fill = "Individuals",
    title = "Distribution of families across folds"
  ) +
  theme_bw()
```

The heatmap visually confirms the balance shown in the table: counts within a family differ by no more than the amount reported above. The figure describes fold composition and does not modify the partition.

## 9. Using the folds for model fitting

The `cv_folds` table identifies which individuals are masked in each validation round. For fold `k`, the 30 individuals with `Fold == k` have their phenotypes replaced by `NA`, while the other 120 provide training phenotypes. This procedure is repeated for every trait and method.

`X`, `M`, and `G` remain fixed and include all 150 genotyped candidates, consistent with the transductive design defined at the start of the module. Consequently, every individual receives a prediction when its own phenotype is masked, producing an out-of-fold prediction downstream; this is not equivalent to predicting an external genotype population.

The expression `cv_folds |> arrange(Fold, FAM, IND) |> head(15)` sorts the table first by fold, then by family and individual identifier, and retains the first 15 rows. `arrange()` defines this order and `head(15)` limits only the displayed preview; the `cv_folds` object still contains all 150 individuals. The `Fold` column is the information Module 03 uses to decide which phenotypes are masked in each validation round.

```{r fold-example}
cv_folds |>
  arrange(Fold, FAM, IND) |>
  head(15) |>
  format_table(caption = "Example of individual allocation to folds")
```

The 15 displayed rows make the allocation structure visible without modifying `cv_folds`, which retains all 150 individuals and is used directly in Module 03.

## 10. Genomic relationship heatmap

The heatmap provides a full view of `G`: diagonal cells represent each individual's relationship with itself, whereas off-diagonal cells represent pairwise genomic relationships. Individuals remain in their original object order; no hierarchical clustering or reordering is applied to create visual blocks. Apparent patterns can therefore suggest relationship structure, but the figure is not a clustering or population-structure test.

```{r genomic-relationship-plot, fig.height=8}
# Reshape G to long format and build the genomic-relationship heatmap.
G_long <- as.data.frame(as.table(G_full))
names(G_long) <- c("Individual_1", "Individual_2", "Relationship")

ggplot(
  G_long,
  aes(
    x = Individual_1,
    y = Individual_2,
    fill = Relationship
  )
) +
  geom_raster() +
  scale_x_discrete(breaks = NULL) +
  scale_y_discrete(breaks = NULL) +
  labs(
    x = "Individuals",
    y = "Individuals",
    fill = "G",
    title = "Additive genomic relationship matrix"
  ) +
  theme_bw()
```

The heatmap summarizes pairwise relationships in `G`. Because individual order is preserved and no clustering is applied, areas of higher or lower intensity should be interpreted as descriptive relationship patterns rather than groups inferred by the figure.

## 11. Diagnostics of the constructed matrices

All allele columns satisfying the declared filters are retained; no category is artificially discarded merely to force full rank. When all alleles at a locus remain represented, their dosages sum to 2 in each diploid individual and an exact linear dependency exists among those columns. Individual allele coefficients are therefore not interpreted in isolation; later modules focus on predictions and genomic values.

In the analyzed dataset, the 72 dosage columns do not represent 72 independent dimensions. The theoretical rank is computed by `sum(table(...) - 1L)`: `table()` counts retained allele columns at each locus, subtracting 1 gives \(k_l-1\) for each locus, and `sum()` combines those contributions. The observed rank of `M` is then compared with that value.

The diploid property is checked locus by locus with `rowSums(...) - 2`: `rowSums()` adds the retained allele dosages for each individual, and the difference from 2 should be numerically zero. Centering is checked through the column means of `M`, and `tcrossprod(M_full) / denominator_full` reconstructs `G` for comparison with the stored matrix. Symmetry, rank, diagonal, and smallest eigenvalue complete the numerical check of `G`.

```{r matrix-diagnostics}
# Check dimensions, ranks, linear dependencies, and numerical properties of X, M, and G.
theoretical_rank_M <- sum(
  table(
    allele_audit_full$Locus[
      allele_audit_full$Keep_allele
    ]
  ) - 1L
)

maximum_M_column_mean <- max(abs(colMeans(M_full)))

G_reconstructed <- tcrossprod(M_full) / denominator_full
maximum_G_reconstruction_difference <- max(
  abs(G_full - G_reconstructed)
)

locus_names <- unique(
  allele_audit_full$Locus[
    allele_audit_full$Keep_allele
  ]
)

dosage_sum_error <- numeric(length(locus_names))

for (j in seq_along(locus_names)) {
  # Select dosage columns belonging to the same locus.
  locus_columns <- allele_audit_full$Marker[
    allele_audit_full$Keep_allele &
      allele_audit_full$Locus == locus_names[j]
  ]

  # In a diploid individual, dosage sums to 2 within a locus.
  dosage_sum_error[j] <- max(
    abs(
      rowSums(
        dosage_full_imputed[
          ,
          locus_columns,
          drop = FALSE
        ]
      ) - 2
    )
  )
}

maximum_dosage_sum_error <- max(dosage_sum_error)

matrix_inventory <- data.frame(
  Item = c(
    "Individuals",
    "Rows of X",
    "Columns of X",
    "Rank of X",
    "Rows of M",
    "Columns of M",
    "Rank of M",
    "Theoretical rank of M",
    "Maximum absolute column mean of M",
    "Maximum dosage-sum error within locus",
    "Maximum G reconstruction difference",
    "Rows of G",
    "Columns of G",
    "Rank of G",
    "Folds",
    "Training individuals per fold",
    "Validation individuals per fold",
    "Maximum symmetry difference of G",
    "Mean diagonal of G",
    "Smallest eigenvalue of G"
  ),
  Value = c(
    length(ids),
    nrow(X),
    ncol(X),
    qr(X)$rank,
    nrow(M_full),
    ncol(M_full),
    qr(M_full)$rank,
    theoretical_rank_M,
    maximum_M_column_mean,
    maximum_dosage_sum_error,
    maximum_G_reconstruction_difference,
    nrow(G_full),
    ncol(G_full),
    qr(G_full)$rank,
    N_FOLDS,
    length(ids) - unique(fold_sizes),
    unique(fold_sizes),
    max(abs(G_full - t(G_full))),
    mean(diag(G_full)),
    min(eigenvalues_G_full)
  )
)

format_table(
  matrix_inventory,
  digits = 8,
  col.names = c("Item", "Value"),
  caption = "Summary of matrices and cross-validation"
)
```

The final inventory collects the preceding checks in one table. For the analyzed data, `M` contains `r ncol(M_full)` columns and rank `r qr(M_full)$rank`, equal to the theoretical rank `r theoretical_rank_M`; dosage sums within loci and reconstruction of `G` have maximum errors close to zero, and all five folds retain the same training and validation sizes.

## 12. Saving matrices and folds

The `matrices_object.rds` file gathers matrices, aligned phenotypes, metadata, frequencies, folds, and parameters in one object while preserving their classes. `saveRDS()` writes this list to a file that can be recovered intact with `readRDS()` in Module 03. The `write.csv(..., row.names = FALSE)` calls export fold, frequency, and diagnostic tables in tabular form; `row.names = FALSE` prevents an artificial row-number column from being added.

```{r save-matrix-objects}
# Collect matrices, metadata, frequencies, and folds in the object persisted by the module.
matrices_object <- list(
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,
  trait_dictionary = trait_dictionary,
  X = X,
  M = M_full,
  G = G_full,
  cv_folds = cv_folds,
  fold_summary = fold_summary,
  locus_audit = locus_audit_full,
  allele_audit = allele_audit_full,
  marker_means = marker_means_full,
  allele_frequencies = allele_frequencies_full,
  denominator = denominator_full,
  # Record parameters and thresholds required to reproduce this stage.
  settings = list(
    n_folds = N_FOLDS,
    seed_cv = SEED_CV,
    call_rate_min = CALL_RATE_MIN,
    allele_frequency_min = ALLELE_FREQUENCY_MIN,
    fixed_effects = "FAM"
  )
)

# Save the RDS object read directly by Module 03.
saveRDS(
  matrices_object,
  file.path(OUTPUT_DIR, "matrices_object.rds")
)

# Export tabular diagnostics as CSV files for inspection independently of the RDS object.
write.csv(cv_folds, file.path(OUTPUT_DIR, "cv_folds.csv"), row.names = FALSE)
write.csv(fold_summary, file.path(OUTPUT_DIR, "fold_summary.csv"), row.names = FALSE)
write.csv(
  allele_audit_full,
  file.path(OUTPUT_DIR, "allele_audit_full.csv"),
  row.names = FALSE
)
write.csv(
  matrix_inventory,
  file.path(OUTPUT_DIR, "matrix_inventory.csv"),
  row.names = FALSE
)
```

## 13. Next step

Module 03 uses `X`, `M`, `G`, and the five folds to fit RR-BLUP, BayesA,
BayesB, BayesB2, BayesC, Bayesian Lasso, and GBLUP. In every fit, only the
phenotypes of validation individuals are masked.

<div style="display:flex; justify-content:space-between; gap:1rem; margin-top:2rem; padding-top:1rem; border-top:1px solid #ddd;">
<a href="01_preprocessing_en.html">← Module 01 — Preprocessing</a>
<a href="03_models_en.html">Module 03 — Model fitting →</a>
</div>
