---
title: "Matrix construction for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

Genomic prediction models require numerical matrices that describe fixed
effects and genetic similarity among individuals.

In this step, the SSR genotypes are converted into allele dosages, the fixed
effects of block and family are represented in a design matrix, and five
cross-validation folds are defined. Marker processing is repeated using only
the training individuals of each fold.

## 2. Packages and input object

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
  library(here)
})
```

```{r input-object}
PREPROCESSING_FILE <- here(
  "output",
  "preprocessing",
  "preprocessing_object.rds"
)

OUTPUT_DIR <- here(
  "output",
  "matrices"
)

N_FOLDS <- 5L
SEED_CV <- 20260723L
ALLELE_FREQUENCY_MIN <- 0.05

dir.create(
  OUTPUT_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

preprocessing <- readRDS(
  PREPROCESSING_FILE
)

ids <- preprocessing$ids
metadata <- preprocessing$metadata
phenotypes <- preprocessing$phenotypes
ssr_raw <- preprocessing$ssr_raw
locus_audit_full <- preprocessing$locus_audit
trait_dictionary <- preprocessing$trait_dictionary

CALL_RATE_MIN <- preprocessing$settings$call_rate_min

data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Raw SSR loci",
    "Retained SSR loci",
    "Families",
    "Blocks"
  ),
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    sum(locus_audit_full$Keep_locus),
    nlevels(metadata$FAM),
    nlevels(metadata$BL)
  )
) |>
  kable(
    caption = "Input data for matrix construction"
  )
```

## 3. Allele-dosage coding

Each SSR code contains two allele labels. The code `12`, for example,
represents one copy of allele 1 and one copy of allele 2. The code `22`
represents two copies of allele 2.

One column is created for every allele observed at each retained locus. The
dosage is 0, 1, or 2 according to the number of copies carried by an
individual.

```{r complete-dosage-coding}
retained_loci_full <- locus_audit_full$Locus[
  locus_audit_full$Keep_locus
]

ssr_full <- ssr_raw[
  ,
  retained_loci_full,
  drop = FALSE
]

dosage_parts_full <- list()
allele_map_parts_full <- list()
marker_counter_full <- 0L

for (j in seq_len(ncol(ssr_full))) {
  codes <- ssr_full[, j]

  allele_1 <- floor(codes / 10)
  allele_2 <- codes %% 10

  observed_alleles <- sort(
    unique(
      c(
        allele_1[!is.na(codes)],
        allele_2[!is.na(codes)]
      )
    )
  )

  for (allele in observed_alleles) {
    marker_counter_full <- marker_counter_full + 1L

    dosage <-
      as.numeric(allele_1 == allele) +
      as.numeric(allele_2 == allele)

    dosage[is.na(codes)] <- NA_real_

    marker_name <- paste0(
      colnames(ssr_full)[j],
      "__A",
      allele
    )

    dosage_parts_full[[marker_counter_full]] <- dosage

    allele_map_parts_full[[marker_counter_full]] <- data.frame(
      Locus = colnames(ssr_full)[j],
      Allele = allele,
      Marker = marker_name
    )
  }
}

dosage_full_raw <- do.call(
  cbind,
  dosage_parts_full
)

allele_audit_full <- bind_rows(
  allele_map_parts_full
)

rownames(dosage_full_raw) <- ids
colnames(dosage_full_raw) <- allele_audit_full$Marker
rownames(allele_audit_full) <- NULL

dosage_full_raw[1:6, 1:8]
```

## 4. Allele-frequency control

Very rare allele columns contribute little information and can be unstable in
small training samples. Alleles are retained when their frequency is between
0.05 and 0.95 and their dosage has nonzero variation.

```{r complete-allele-control}
allele_audit_full$Allele_frequency <-
  colMeans(
    dosage_full_raw,
    na.rm = TRUE
  ) / 2

dosage_sd_full <- numeric(
  ncol(dosage_full_raw)
)

for (j in seq_len(ncol(dosage_full_raw))) {
  dosage_sd_full[j] <- sd(
    dosage_full_raw[, j],
    na.rm = TRUE
  )
}

allele_audit_full$Dosage_sd <- dosage_sd_full

allele_audit_full$Keep_allele <-
  is.finite(allele_audit_full$Allele_frequency) &
  allele_audit_full$Allele_frequency >=
    ALLELE_FREQUENCY_MIN &
  allele_audit_full$Allele_frequency <=
    (1 - ALLELE_FREQUENCY_MIN) &
  is.finite(allele_audit_full$Dosage_sd) &
  allele_audit_full$Dosage_sd > 0

retained_markers_full <- allele_audit_full$Marker[
  allele_audit_full$Keep_allele
]

dosage_full <- dosage_full_raw[
  ,
  retained_markers_full,
  drop = FALSE
]

allele_summary_full <- allele_audit_full |>
  group_by(Locus) |>
  summarise(
    Observed_alleles = n(),
    Retained_alleles = sum(Keep_allele),
    .groups = "drop"
  )

kable(
  allele_summary_full,
  caption = "Allele-dosage columns by SSR locus"
)
```

```{r allele-frequency-plot, fig.height=5}
ggplot(
  allele_audit_full,
  aes(
    x = Allele_frequency,
    fill = Keep_allele
  )
) +
  geom_histogram(
    bins = 20,
    position = "identity",
    alpha = 0.7
  ) +
  geom_vline(
    xintercept = c(
      ALLELE_FREQUENCY_MIN,
      1 - ALLELE_FREQUENCY_MIN
    ),
    linetype = 2
  ) +
  labs(
    x = "Allele frequency",
    y = "Allele-dosage columns",
    fill = "Retained",
    title = "Allele-frequency control"
  ) +
  theme_bw()
```

## 5. Complete marker matrix

Missing allele dosages are replaced by the mean dosage of their column. The
columns are then centered by subtracting those means.

The resulting matrix \(\mathbf{M}\) has individuals in rows and retained allele
dosages in columns.

```{r complete-marker-matrix}
marker_means_full <- colMeans(
  dosage_full,
  na.rm = TRUE
)

dosage_full_imputed <- dosage_full

for (j in seq_len(ncol(dosage_full_imputed))) {
  missing_values <- is.na(
    dosage_full_imputed[, j]
  )

  dosage_full_imputed[
    missing_values,
    j
  ] <- marker_means_full[j]
}

M_full <- sweep(
  dosage_full_imputed,
  2,
  marker_means_full,
  FUN = "-"
)

```

## 6. Complete genomic relationship matrix

Let \(p_j\) be the frequency of the allele represented by column \(j\). The
additive genomic relationship matrix is calculated as

\[
\mathbf{G} =
\frac{\mathbf{M}\mathbf{M}^{\mathsf{T}}}
{\sum_j 2p_j(1-p_j)}.
\]

```{r complete-genomic-matrix}
allele_frequencies_full <-
  marker_means_full / 2

denominator_full <- sum(
  2 *
    allele_frequencies_full *
    (1 - allele_frequencies_full)
)

G_full <- tcrossprod(
  M_full
) / denominator_full

G_full <- (
  G_full + t(G_full)
) / 2

dimnames(G_full) <- list(
  ids,
  ids
)

eigenvalues_G_full <- eigen(
  G_full,
  symmetric = TRUE,
  only.values = TRUE
)$values

```

## 7. Fixed-effect matrix

Block and family are represented as categorical fixed effects. The intercept
is included automatically.

```{r fixed-effect-matrix}
X <- model.matrix(
  ~ BL + FAM,
  data = metadata
)

rownames(X) <- ids

data.frame(
  Matrix = c("X", "M", "G"),
  Rows = c(
    nrow(X),
    nrow(M_full),
    nrow(G_full)
  ),
  Columns = c(
    ncol(X),
    ncol(M_full),
    ncol(G_full)
  ),
  Rank = c(
    qr(X)$rank,
    qr(M_full)$rank,
    qr(G_full)$rank
  )
) |>
  kable(
    caption = "Dimensions and ranks of the complete matrices"
  )
```

## 8. Five family-stratified folds

The 150 individuals are divided into five validation groups. Individuals from
each family are distributed as evenly as possible among the folds. The fixed
seed preserves the same allocation in every execution.

```{r cross-validation-folds}
set.seed(
  SEED_CV
)

family <- droplevels(
  metadata$FAM
)

folds <- integer(
  length(ids)
)

fold_load <- integer(
  N_FOLDS
)

for (family_level in levels(family)) {
  family_indices <- sample(
    which(family == family_level)
  )

  random_tie_order <- sample(
    seq_len(N_FOLDS)
  )

  ordered_folds <- order(
    fold_load,
    match(
      seq_len(N_FOLDS),
      random_tie_order
    )
  )

  family_labels <- rep(
    ordered_folds,
    length.out = length(family_indices)
  )

  folds[family_indices] <-
    family_labels

  fold_load <- fold_load + tabulate(
    family_labels,
    nbins = N_FOLDS
  )
}

fold_sizes <- tabulate(
  folds,
  nbins = N_FOLDS
)

family_fold_table <- table(
  Family = family,
  Fold = folds
)

family_balance <- apply(
  family_fold_table,
  1,
  max
) - apply(
  family_fold_table,
  1,
  min
)

cv_folds <- data.frame(
  IND = ids,
  BL = as.character(metadata$BL),
  FAM = as.character(metadata$FAM),
  Fold = folds
)

data.frame(
  Fold = seq_len(N_FOLDS),
  Validation_n = fold_sizes,
  Training_n = length(ids) - fold_sizes
) |>
  kable(
    caption = "Size of the cross-validation folds"
  )
```

```{r family-fold-plot, fig.height=7}
family_fold_long <- as.data.frame(
  family_fold_table
)

ggplot(
  family_fold_long,
  aes(
    x = factor(Fold),
    y = Family,
    fill = Freq
  )
) +
  geom_tile() +
  geom_text(
    aes(label = Freq)
  ) +
  labs(
    x = "Fold",
    y = "Family",
    fill = "Individuals",
    title = "Family distribution across folds"
  ) +
  theme_bw()
```

## 9. Training-specific matrices

For each fold, locus filtering, allele coding, allele-frequency control,
imputation, centering, and the genomic relationship matrix are recalculated
using the 120 training individuals.

The 30 validation individuals are transformed with the means and allele
structure obtained from training.

```{r training-specific-matrices}
cv_objects <- vector(
  "list",
  N_FOLDS
)

fold_diagnostics <- vector(
  "list",
  N_FOLDS
)

for (fold in seq_len(N_FOLDS)) {
  validation_index <- which(
    folds == fold
  )

  training_index <- which(
    folds != fold
  )

  ssr_training <- ssr_raw[
    training_index,
    ,
    drop = FALSE
  ]

  genotype_classes <- integer(
    ncol(ssr_training)
  )

  for (j in seq_len(ncol(ssr_training))) {
    genotype_classes[j] <- length(
      unique(
        na.omit(
          ssr_training[, j]
        )
      )
    )
  }

  locus_audit <- data.frame(
    Locus = colnames(ssr_training),
    N_called = colSums(!is.na(ssr_training)),
    Call_rate = colMeans(!is.na(ssr_training)),
    N_genotype_classes = genotype_classes
  )

  locus_audit$Keep_locus <-
    locus_audit$Call_rate >= CALL_RATE_MIN &
    locus_audit$N_genotype_classes >= 2

  retained_loci <- locus_audit$Locus[
    locus_audit$Keep_locus
  ]

  ssr_fold <- ssr_raw[
    ,
    retained_loci,
    drop = FALSE
  ]

  dosage_parts <- list()
  allele_map_parts <- list()
  marker_counter <- 0L

  for (j in seq_len(ncol(ssr_fold))) {
    codes_all <- ssr_fold[, j]
    codes_training <- ssr_fold[
      training_index,
      j
    ]

    allele_1_training <- floor(
      codes_training / 10
    )

    allele_2_training <-
      codes_training %% 10

    observed_alleles <- sort(
      unique(
        c(
          allele_1_training[
            !is.na(codes_training)
          ],
          allele_2_training[
            !is.na(codes_training)
          ]
        )
      )
    )

    allele_1_all <- floor(
      codes_all / 10
    )

    allele_2_all <-
      codes_all %% 10

    for (allele in observed_alleles) {
      marker_counter <- marker_counter + 1L

      dosage <-
        as.numeric(
          allele_1_all == allele
        ) +
        as.numeric(
          allele_2_all == allele
        )

      dosage[
        is.na(codes_all)
      ] <- NA_real_

      marker_name <- paste0(
        colnames(ssr_fold)[j],
        "__A",
        allele
      )

      dosage_parts[[marker_counter]] <-
        dosage

      allele_map_parts[[marker_counter]] <-
        data.frame(
          Locus = colnames(ssr_fold)[j],
          Allele = allele,
          Marker = marker_name
        )
    }
  }

  dosage_raw <- do.call(
    cbind,
    dosage_parts
  )

  allele_audit <- bind_rows(
    allele_map_parts
  )

  rownames(dosage_raw) <- ids
  colnames(dosage_raw) <-
    allele_audit$Marker

  allele_audit$Allele_frequency <-
    colMeans(
      dosage_raw[
        training_index,
        ,
        drop = FALSE
      ],
      na.rm = TRUE
    ) / 2

  dosage_sd <- numeric(
    ncol(dosage_raw)
  )

  for (j in seq_len(ncol(dosage_raw))) {
    dosage_sd[j] <- sd(
      dosage_raw[
        training_index,
        j
      ],
      na.rm = TRUE
    )
  }

  allele_audit$Dosage_sd <-
    dosage_sd

  allele_audit$Keep_allele <-
    is.finite(
      allele_audit$Allele_frequency
    ) &
    allele_audit$Allele_frequency >=
      ALLELE_FREQUENCY_MIN &
    allele_audit$Allele_frequency <=
      (1 - ALLELE_FREQUENCY_MIN) &
    is.finite(
      allele_audit$Dosage_sd
    ) &
    allele_audit$Dosage_sd > 0

  retained_markers <-
    allele_audit$Marker[
      allele_audit$Keep_allele
    ]

  dosage <- dosage_raw[
    ,
    retained_markers,
    drop = FALSE
  ]

  marker_means <- colMeans(
    dosage[
      training_index,
      ,
      drop = FALSE
    ],
    na.rm = TRUE
  )

  dosage_imputed <- dosage

  for (j in seq_len(ncol(dosage_imputed))) {
    missing_values <- is.na(
      dosage_imputed[, j]
    )

    dosage_imputed[
      missing_values,
      j
    ] <- marker_means[j]
  }

  M <- sweep(
    dosage_imputed,
    2,
    marker_means,
    FUN = "-"
  )

  allele_frequencies <-
    marker_means / 2

  denominator <- sum(
    2 *
      allele_frequencies *
      (1 - allele_frequencies)
  )

  G <- tcrossprod(
    M
  ) / denominator

  G <- (
    G + t(G)
  ) / 2

  dimnames(G) <- list(
    ids,
    ids
  )

  eigenvalues_G <- eigen(
    G,
    symmetric = TRUE,
    only.values = TRUE
  )$values

  cv_objects[[fold]] <- list(
    training_index = training_index,
    validation_index = validation_index,
    training_ids = ids[training_index],
    validation_ids = ids[validation_index],
    locus_audit = locus_audit,
    allele_audit = allele_audit,
    marker_means = marker_means,
    allele_frequencies = allele_frequencies,
    denominator = denominator,
    M = M,
    G = G
  )

  fold_diagnostics[[fold]] <- data.frame(
    Fold = fold,
    Training_n = length(training_index),
    Validation_n = length(validation_index),
    Retained_loci = length(retained_loci),
    Retained_alleles = ncol(M),
    Fixed_effect_rank = qr(
      X[
        training_index,
        ,
        drop = FALSE
      ]
    )$rank,
    Mean_diagonal_G = mean(
      diag(G)
    ),
    Minimum_eigenvalue_G = min(
      eigenvalues_G
    )
  )
}

names(cv_objects) <- paste0(
  "Fold_",
  seq_len(N_FOLDS)
)

fold_diagnostics <- bind_rows(
  fold_diagnostics
)

kable(
  fold_diagnostics,
  digits = 5,
  caption = "Diagnostics of the training-specific matrices"
)
```

## 10. Genomic relationship heatmap

The heatmap represents genomic similarity among the 150 individuals.

```{r genomic-relationship-plot, fig.height=8}
G_long <- as.data.frame(
  as.table(G_full)
)

names(G_long) <- c(
  "Individual_1",
  "Individual_2",
  "Relationship"
)

ggplot(
  G_long,
  aes(
    x = Individual_1,
    y = Individual_2,
    fill = Relationship
  )
) +
  geom_raster() +
  scale_x_discrete(
    breaks = NULL
  ) +
  scale_y_discrete(
    breaks = NULL
  ) +
  labs(
    x = "Individuals",
    y = "Individuals",
    fill = "G",
    title = "Additive genomic relationship matrix"
  ) +
  theme_bw()
```

## 11. Saved objects

```{r save-matrix-objects}
matrix_inventory <- data.frame(
  Item = c(
    "Individuals",
    "Fixed-effect columns",
    "Complete marker columns",
    "Genomic-matrix rows",
    "Cross-validation folds",
    "Validation individuals per fold",
    "Mean diagonal of G",
    "Minimum eigenvalue of G"
  ),
  Value = c(
    length(ids),
    ncol(X),
    ncol(M_full),
    nrow(G_full),
    N_FOLDS,
    unique(fold_sizes),
    mean(diag(G_full)),
    min(eigenvalues_G_full)
  )
)

matrices_object <- list(
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,
  trait_dictionary = trait_dictionary,
  X = X,
  cv_folds = cv_folds,
  full = list(
    locus_audit = locus_audit_full,
    allele_audit = allele_audit_full,
    marker_means = marker_means_full,
    allele_frequencies = allele_frequencies_full,
    denominator = denominator_full,
    M = M_full,
    G = G_full
  ),
  cv = cv_objects,
  fold_diagnostics = fold_diagnostics,
  settings = list(
    n_folds = N_FOLDS,
    seed_cv = SEED_CV,
    call_rate_min = CALL_RATE_MIN,
    allele_frequency_min =
      ALLELE_FREQUENCY_MIN
  )
)

saveRDS(
  matrices_object,
  file.path(
    OUTPUT_DIR,
    "matrices_object.rds"
  )
)

write.csv(
  cv_folds,
  file.path(
    OUTPUT_DIR,
    "cv_folds.csv"
  ),
  row.names = FALSE
)

write.csv(
  fold_diagnostics,
  file.path(
    OUTPUT_DIR,
    "fold_diagnostics.csv"
  ),
  row.names = FALSE
)

write.csv(
  allele_audit_full,
  file.path(
    OUTPUT_DIR,
    "allele_audit_full.csv"
  ),
  row.names = FALSE
)

write.csv(
  matrix_inventory,
  file.path(
    OUTPUT_DIR,
    "matrix_inventory.csv"
  ),
  row.names = FALSE
)

kable(
  matrix_inventory,
  digits = 6,
  caption = "Final matrix inventory"
)
```

## 12. Matrix summary

```{r matrix-summary}
matrix_summary <- data.frame(
  Measure = c(
    "Individuals",
    "Complete marker columns",
    "Fixed-effect columns",
    "Fixed-effect rank",
    "Genomic-matrix symmetry difference",
    "Minimum eigenvalue of G",
    "Validation individuals per fold",
    "Training individuals per fold"
  ),
  Value = c(
    length(ids),
    ncol(M_full),
    ncol(X),
    qr(X)$rank,
    max(abs(G_full - t(G_full))),
    min(eigenvalues_G_full),
    unique(fold_sizes),
    length(ids) - unique(fold_sizes)
  )
)

kable(
  matrix_summary,
  digits = 8,
  caption = "Summary of the matrices used in genomic prediction"
)
```

## 13. Next step

The next step fits the Bayesian regression, RR-BLUP, and GBLUP models using the
same five validation folds.
