---
title: "Construction of matrices for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

The hidden chunk below defines only the rendering options shared by this page. It does not load analytical objects, construct matrices, or affect cross-validation.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

In this module, the preprocessed SSR genotypes are converted into allele
dosages. These dosages are used to construct the centered marker matrix `M`,
the additive genomic relationship matrix `G`, and the fixed-effect matrix `X`,
which contains family in addition to the intercept. Five family-stratified
cross-validation folds are also defined.

The matrices are constructed once using all 150 individuals and remain
unchanged across folds. In Module 03, only the phenotype vector is masked to
form training and validation sets.

Because candidates in every fold are already genotyped, marker filtering,
allele frequencies, mean imputation, `M`, and `G` are defined from the full
genotype panel and then held fixed. No validation phenotype enters these
steps. This corresponds to a transductive candidate-genotype setting and
should not be interpreted as external validation in a new genotype population.

At the end of the module, the reader will have four components used directly in Module 03: the fixed-effect matrix `X`, the centered marker matrix `M`, the genomic relationship matrix `G`, and the `cv_folds` table identifying each individual's validation fold. Diagnostics throughout the page check whether allele coding, centering, genomic relationships, and family stratification are coherent with this design.

## 2. Packages and input object

The first chunk loads only the packages required for object manipulation, tabular reshaping, figures, tables, and portable paths. No matrix is created at this stage.

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
  library(here)
})
```

The next step loads the preprocessing object and fixes the parameters used to construct the matrices and cross-validation design.

Five folds are used because, with 150 individuals, 5-fold partitioning gives 30 validation individuals and 120 training individuals per fold while preserving family stratification. Each individual appears exactly once in validation; this is one family-stratified five-fold partition, not repeated cross-validation. `SEED_CV = 20260723` only reproduces the same allocation of individuals and has no inferential interpretation.

`ALLELE_FREQUENCY_MIN = 0.05` retains allele-dosage columns with frequencies between 0.05 and 0.95 provided that dosage variance is also positive. Thus, very rare or nearly fixed alleles do not enter `M`. The call-rate threshold is inherited from Module 01 to record the same quality-control rule; loci reach this module after that decision has already been applied.

```{r input-object}
# Define paths and load the preprocessing object used to construct the genomic matrices.
PREPROCESSING_FILE <- here(
  "output",
  "preprocessing",
  "preprocessing_object.rds"
)

OUTPUT_DIR <- here(
  "output",
  "matrices"
)

# Use five folds: 30 validation and 120 training individuals per fold.
N_FOLDS <- 5L
# Fix only the reproducibility of the family-stratified allocation.
SEED_CV <- 20260723L
# Exclude alleles with frequency below 5% or above 95%.
ALLELE_FREQUENCY_MIN <- 0.05

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

preprocessing <- readRDS(PREPROCESSING_FILE)

ids <- preprocessing$ids
metadata <- preprocessing$metadata
phenotypes <- preprocessing$phenotypes
ssr_raw <- preprocessing$ssr_raw
locus_audit_full <- preprocessing$locus_audit
trait_dictionary <- preprocessing$trait_dictionary
# Recover, for provenance, the call-rate threshold applied in Module 01.
CALL_RATE_MIN <- preprocessing$settings$call_rate_min
```

Before marker coding, the following table checks the structure received from Module 01: numbers of individuals and traits, loci before and after quality control, and families. It acts as a checkpoint between modules; no new filter is applied in this chunk.

```{r input-summary}
# Summarize the dimensions and categories of the preprocessing object received by the module.
data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Original SSR loci",
    "Retained SSR loci",
    "Families"
  ),
  # List the values corresponding to the table entries.
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    sum(locus_audit_full$Keep_locus),
    nlevels(metadata$FAM)
  )
) |>
  kable(caption = "Input data for matrix construction")
```

## 3. Allele-dosage coding

Each SSR code contains two allele labels. For example, `12` represents one copy of allele 1 and one copy of allele 2, whereas `22` represents two copies of allele 2. For every observed allele at each retained locus, an individual's dosage is computed as the sum of two indicator functions, one for each allele position in the genotype. The resulting value is 0, 1, or 2 according to whether the individual carries zero, one, or two copies of that allele.

One column is created for every locus–allele combination, and missing SSR calls remain `NA` until imputation. Before any allele-column filtering, dosages across all alleles at a diploid locus sum to 2 for every observed genotype. This property is later used as a coding diagnostic.

```{r allele-dosage-coding}
# Organize marker information and compute the diagnostics required at this stage.
retained_loci_full <- locus_audit_full$Locus[
  locus_audit_full$Keep_locus
]

ssr_full <- ssr_raw[
  ,
  retained_loci_full,
  drop = FALSE
]

dosage_parts_full <- list()
allele_map_parts_full <- list()
marker_counter_full <- 0L

for (j in seq_len(ncol(ssr_full))) {
  # Separate the two alleles encoded in each SSR genotype.
  codes <- ssr_full[, j]
  allele_1 <- floor(codes / 10)
  allele_2 <- codes %% 10

  # List the alleles observed at the current locus.
  observed_alleles <- sort(
    unique(
      c(
        allele_1[!is.na(codes)],
        allele_2[!is.na(codes)]
      )
    )
  )

  for (allele in observed_alleles) {
    marker_counter_full <- marker_counter_full + 1L

    # Count 0, 1, or 2 copies of the allele for each individual.
    dosage <-
      as.numeric(allele_1 == allele) +
      as.numeric(allele_2 == allele)
    dosage[is.na(codes)] <- NA_real_

    # Create a unique identifier for each locus-allele combination.
    marker_name <- paste0(
      colnames(ssr_full)[j],
      "__A",
      allele
    )

    dosage_parts_full[[marker_counter_full]] <- dosage
    allele_map_parts_full[[marker_counter_full]] <- data.frame(
      Locus = colnames(ssr_full)[j],
      Allele = allele,
      Marker = marker_name
    )
  }
}

dosage_full_raw <- do.call(cbind, dosage_parts_full)
allele_audit_full <- bind_rows(allele_map_parts_full)

rownames(dosage_full_raw) <- ids
colnames(dosage_full_raw) <- allele_audit_full$Marker
rownames(allele_audit_full) <- NULL

# Example of the resulting dosage columns.
dosage_full_raw[1:6, 1:8]
```

## 4. Allele frequencies and retention of dosage columns

The frequency of an allele column is computed as the mean observed dosage divided by 2 because each diploid individual can contribute two copies of the allele. `NA` values are ignored here, so the frequency uses available genotype calls only.

Columns are retained when frequency lies between 0.05 and 0.95 and dosage standard deviation is positive. The lower and upper limits are symmetric: they remove very rare alleles and nearly fixed columns, while `Dosage_sd > 0` guarantees observed variation. The filter acts on allele-dosage columns; it does not remove individuals.

```{r allele-frequency-control}
# Compute dosage frequency and variability to retain informative allele columns.
allele_audit_full$Allele_frequency <-
  colMeans(dosage_full_raw, na.rm = TRUE) / 2

dosage_sd_full <- numeric(ncol(dosage_full_raw))

for (j in seq_len(ncol(dosage_full_raw))) {
  dosage_sd_full[j] <- sd(
    dosage_full_raw[, j],
    na.rm = TRUE
  )
}

allele_audit_full$Dosage_sd <- dosage_sd_full

allele_audit_full$Keep_allele <-
  is.finite(allele_audit_full$Allele_frequency) &
  allele_audit_full$Allele_frequency >= ALLELE_FREQUENCY_MIN &
  allele_audit_full$Allele_frequency <= (1 - ALLELE_FREQUENCY_MIN) &
  is.finite(allele_audit_full$Dosage_sd) &
  allele_audit_full$Dosage_sd > 0

# Select allele columns that pass the frequency and dosage-variation filters.
retained_markers_full <- allele_audit_full$Marker[
  allele_audit_full$Keep_allele
]

dosage_full <- dosage_full_raw[
  ,
  retained_markers_full,
  drop = FALSE
]

# Summarize the results in `allele_summary_full`.
allele_summary_full <- allele_audit_full |>
  group_by(Locus) |>
  summarise(
    Observed_alleles = n(),
    Retained_alleles = sum(Keep_allele),
    .groups = "drop"
  )

kable(
  allele_summary_full,
  caption = "Allele-dosage columns by SSR locus"
)
```

The histogram shows where allele-column frequencies are concentrated. Vertical lines mark 0.05 and 0.95, and fill indicates the final retention decision, which also requires positive dosage variation. The 20 bins are only a visual-resolution choice and do not participate in filtering.

```{r allele-frequency-plot, fig.height=5}
# Display the allele-frequency distribution and the thresholds used by the filter.
ggplot(
  allele_audit_full,
  aes(
    x = Allele_frequency,
    fill = Keep_allele
  )
) +
  geom_histogram(
    # Use 20 bins only to standardize the visual resolution of the histogram.
    bins = 20,
    position = "identity",
    alpha = 0.7
  ) +
  geom_vline(
    # Mark the lower and upper allele-frequency thresholds in the figure.
    xintercept = c(
      ALLELE_FREQUENCY_MIN,
      1 - ALLELE_FREQUENCY_MIN
    ),
    linetype = 2
  ) +
  labs(
    x = "Allele frequency",
    y = "Allele-dosage columns",
    fill = "Retained",
    title = "Distribution of allele frequencies"
  ) +
  theme_bw()
```

## 5. Marker matrix

Missing dosages are replaced by the mean of the corresponding column computed from the full genotype panel. Because the same mean is subtracted during centering, a mean-imputed dosage becomes exactly zero in `M`: it represents no deviation from the column average rather than a new genotype class.

The resulting matrix is denoted by \(\mathbf M\). Its columns are centered but are not standardized to unit variance. As defined in the objective, imputation uses genotype information from all 150 candidates; no validation phenotype enters this calculation.

```{r marker-matrix}
# Impute missing dosages by column means, center the columns, and construct marker matrix M.
marker_means_full <- colMeans(
  dosage_full,
  na.rm = TRUE
)

dosage_full_imputed <- dosage_full

for (j in seq_len(ncol(dosage_full_imputed))) {
  # Locate missing dosages in the current column.
  missing_values <- is.na(dosage_full_imputed[, j])

  # Impute the column mean.
  dosage_full_imputed[
    missing_values,
    j
  ] <- marker_means_full[j]
}

# Center every dosage column.
M_full <- sweep(
  dosage_full_imputed,
  2,
  marker_means_full,
  FUN = "-"
)
```

The summary of `M` checks four basic properties. Its row count should match the 150 candidates; its column count is the number of retained allele-dosage columns; its rank reveals linear dependencies among columns; and the maximum absolute column mean should be close to zero, confirming numerical centering.

```{r marker-matrix-summary}
data.frame(
  Rows = nrow(M_full),
  Columns = ncol(M_full),
  Rank = qr(M_full)$rank,
  Maximum_absolute_column_mean = max(abs(colMeans(M_full)))
) |>
  kable(
    digits = 8,
    caption = "Dimensions and centering of the marker matrix"
  )
```

## 6. Genomic relationship matrix

Let \(p_j\) be the allele frequency represented by column \(j\). Using a
VanRaden-type allele-frequency scaling [@vanraden2008efficient], the additive
genomic relationship matrix is calculated as

\[
\mathbf G =
\frac{\mathbf M\mathbf M^{\mathsf T}}
{\sum_j 2p_j(1-p_j)}.
\]

```{r genomic-relationship-matrix}
# Recover p_j from the dosage mean, which equals 2p_j.
allele_frequencies_full <- marker_means_full / 2

# Compute the VanRaden scaling from allele-column frequencies.
denominator_full <- sum(
  2 *
    allele_frequencies_full *
    (1 - allele_frequencies_full)
)

# Convert cross-products of centered dosages into scaled genomic relationships.
G_full <- tcrossprod(M_full) / denominator_full

# Remove only numerical asymmetry.
G_full <- (G_full + t(G_full)) / 2

dimnames(G_full) <- list(ids, ids)

eigenvalues_G_full <- eigen(
  G_full,
  symmetric = TRUE,
  only.values = TRUE
)$values
```

The denominator reports the scale used to construct `G`. The mean diagonal summarizes each individual's relationship with itself on this scale and is not forced to equal exactly 1. The symmetry difference should be close to zero. Because `G` is built from a cross-product, it is theoretically positive semidefinite; therefore, a slightly negative smallest eigenvalue, when limited to numerical precision, should be interpreted as rounding rather than evidence of negative relatedness.

```{r genomic-relationship-summary}
# Build or check the matrices used for genomic prediction.
data.frame(
  Quantity = c(
    "Denominator sum 2p(1-p)",
    "Mean diagonal",
    "Maximum symmetry difference",
    "Smallest eigenvalue"
  ),
  # List the values corresponding to the table entries.
  Value = c(
    denominator_full,
    mean(diag(G_full)),
    max(abs(G_full - t(G_full))),
    min(eigenvalues_G_full)
  )
) |>
  kable(
    digits = 8,
    caption = "Numerical summary of the genomic relationship matrix"
  )
```

## 7. Fixed-effect matrix

Family is the only categorical fixed effect. With `~ FAM`, `model.matrix()` uses R's default treatment contrasts: it creates an intercept and enough indicator columns to represent differences among family levels relative to a reference category. Thus, `X` incorporates family structure without treating `FAM` as a numeric variable.

In Module 03, the intercept column is removed from `X` because BGLR already fits an intercept automatically; retaining both would duplicate the same constant effect.

```{r fixed-effect-matrix}
# Build the design matrix for the tutorial's only categorical fixed effect.
X <- model.matrix(
  ~ FAM,
  data = metadata
)

rownames(X) <- ids

data.frame(
  Matrix = c("X", "M", "G"),
  Rows = c(nrow(X), nrow(M_full), nrow(G_full)),
  Columns = c(ncol(X), ncol(M_full), ncol(G_full)),
  Rank = c(qr(X)$rank, qr(M_full)$rank, qr(G_full)$rank)
) |>
  kable(caption = "Dimensions and ranks of the analytical matrices")
```

## 8. Five family-stratified folds

Family stratification targets prediction among individuals from the represented
breeding population. It is not a leave-family-out design and therefore does
not estimate predictive performance for entirely new families.

The 150 individuals are divided into five validation groups of 30 individuals. The algorithm processes one family at a time: it shuffles family members, orders folds by their current total load, uses a random order only to break ties between equally loaded folds, cycles family members across those folds, and then updates the global load. This combines two goals: spreading each family across folds and keeping total fold sizes balanced.

The fixed seed preserves exactly the same partition across executions. The same fold table is used for all ten traits and all methods, so differences among models cannot arise from different validation partitions.

```{r cross-validation-folds}
# Initialize the stratified partition and track each fold's accumulated total load.
set.seed(SEED_CV)

family <- droplevels(metadata$FAM)
folds <- integer(length(ids))
fold_load <- integer(N_FOLDS)

for (family_level in levels(family)) {
  # Randomize the individuals of the current family.
  family_indices <- sample(
    which(family == family_level)
  )

  # Define a random order to break ties between equally loaded folds.
  random_tie_order <- sample(seq_len(N_FOLDS))

  ordered_folds <- order(
    fold_load,
    match(seq_len(N_FOLDS), random_tie_order)
  )

  # Distribute family members across the five folds.
  family_labels <- rep(
    ordered_folds,
    length.out = length(family_indices)
  )

  folds[family_indices] <- family_labels

  fold_load <- fold_load + tabulate(
    family_labels,
    nbins = N_FOLDS
  )
}

fold_sizes <- tabulate(folds, nbins = N_FOLDS)

family_fold_table <- table(
  Family = family,
  Fold = folds
)

# Quantify, within each family, the largest difference in individual counts across folds.
family_balance <- apply(family_fold_table, 1, max) -
  apply(family_fold_table, 1, min)

cv_folds <- data.frame(
  IND = ids,
  FAM = as.character(metadata$FAM),
  Fold = folds
)

fold_summary <- data.frame(
  Fold = seq_len(N_FOLDS),
  Training_n = length(ids) - fold_sizes,
  Validation_n = fold_sizes
)

kable(
  fold_summary,
  col.names = c("Fold", "Training N", "Validation N"),
  caption = "Sample sizes in cross-validation"
)
```

The family × fold table shows the resulting stratification directly. Within each family, counts should be as balanced as its sample size allows; `family_balance`, computed in the previous chunk, measures the difference between the largest and smallest count for that family. Because members are distributed cyclically, this difference is at most one individual for every family.

```{r families-by-fold}
kable(
  as.data.frame.matrix(family_fold_table),
  caption = "Number of individuals from each family in each fold"
)
```

The heatmap displays the same family × fold table visually: every cell explicitly shows the number of individuals, while fill makes imbalances easier to spot. It is a diagnostic of the partition, not an additional measure of predictive performance.

```{r family-fold-plot, fig.height=7}
# Reshape the family-by-fold table and build the individual-distribution figure.
family_fold_long <- as.data.frame(family_fold_table)

# Build the figure used for interpretation.
ggplot(
  family_fold_long,
  aes(
    x = factor(Fold),
    y = Family,
    fill = Freq
  )
) +
  geom_tile() +
  geom_text(aes(label = Freq)) +
  labs(
    x = "Fold",
    y = "Family",
    fill = "Individuals",
    title = "Distribution of families across folds"
  ) +
  theme_bw()
```

## 9. Using the folds for model fitting

The `cv_folds` table identifies which individuals are masked in each validation round. For fold `k`, the 30 individuals with `Fold == k` have their phenotypes replaced by `NA`, while the other 120 provide training phenotypes. This procedure is repeated for every trait and method.

`X`, `M`, and `G` remain fixed and include all 150 genotyped candidates, consistent with the transductive design defined at the start of the module. Consequently, every individual receives a prediction when its own phenotype is masked, producing an out-of-fold prediction downstream; this is not equivalent to predicting an external genotype population.

The table below displays the first assignments after sorting by fold, family, and individual. It simply makes the structure of `cv_folds` concrete: the `Fold` column is the only information Module 03 uses to decide which phenotypes are masked in each validation round.

```{r fold-example}
cv_folds |>
  arrange(Fold, FAM, IND) |>
  head(15) |>
  kable(caption = "Example of individual allocation to folds")
```

## 10. Genomic relationship heatmap

The heatmap provides a full view of `G`: diagonal cells represent each individual's relationship with itself, whereas off-diagonal cells represent pairwise genomic relationships. Individuals remain in their original object order; no hierarchical clustering or reordering is applied to create visual blocks. Apparent patterns can therefore suggest relationship structure, but the figure is not a clustering or population-structure test.

```{r genomic-relationship-plot, fig.height=8}
# Reshape G to long format and build the genomic-relationship heatmap.
G_long <- as.data.frame(as.table(G_full))
names(G_long) <- c("Individual_1", "Individual_2", "Relationship")

# Build the figure used for interpretation.
ggplot(
  G_long,
  aes(
    x = Individual_1,
    y = Individual_2,
    fill = Relationship
  )
) +
  geom_raster() +
  scale_x_discrete(breaks = NULL) +
  scale_y_discrete(breaks = NULL) +
  labs(
    x = "Individuals",
    y = "Individuals",
    fill = "G",
    title = "Additive genomic relationship matrix"
  ) +
  theme_bw()
```

## 11. Diagnostics of the constructed matrices

All allele columns satisfying the declared filters are retained; no category is artificially discarded merely to force full rank. When all alleles at a locus remain represented, their dosages sum to 2 in each diploid individual and an exact linear dependency exists among those columns. Individual allele coefficients are therefore not interpreted in isolation; later modules focus on predictions and genomic values.

In the canonical dataset, the 72 dosage columns do not represent 72 independent dimensions. The diagnostic compares the observed rank of `M` with `sum(k_l - 1)`, where `k_l` is the number of retained allele columns at locus `l`; it also checks whether within-locus dosage still sums to 2, whether `M` is numerically centered, and whether `G` can be reconstructed from `M` and the same denominator. Symmetry, rank, diagonal, and smallest eigenvalue complete the numerical check of `G`.

```{r matrix-diagnostics}
# Check dimensions, ranks, linear dependencies, and numerical properties of X, M, and G.
theoretical_rank_M <- sum(
  table(
    allele_audit_full$Locus[
      allele_audit_full$Keep_allele
    ]
  ) - 1L
)

maximum_M_column_mean <- max(abs(colMeans(M_full)))

G_reconstructed <- tcrossprod(M_full) / denominator_full
maximum_G_reconstruction_difference <- max(
  abs(G_full - G_reconstructed)
)

locus_names <- unique(
  allele_audit_full$Locus[
    allele_audit_full$Keep_allele
  ]
)

dosage_sum_error <- numeric(length(locus_names))

for (j in seq_along(locus_names)) {
  # Select dosage columns belonging to the same locus.
  locus_columns <- allele_audit_full$Marker[
    allele_audit_full$Keep_allele &
      allele_audit_full$Locus == locus_names[j]
  ]

  # In a diploid individual, dosage sums to 2 within a locus.
  dosage_sum_error[j] <- max(
    abs(
      rowSums(
        dosage_full_imputed[
          ,
          locus_columns,
          drop = FALSE
        ]
      ) - 2
    )
  )
}

maximum_dosage_sum_error <- max(dosage_sum_error)

matrix_inventory <- data.frame(
  # List the items checked in the matrix inventory.
  Item = c(
    "Individuals",
    "Rows of X",
    "Columns of X",
    "Rank of X",
    "Rows of M",
    "Columns of M",
    "Rank of M",
    "Theoretical rank of M",
    "Maximum absolute column mean of M",
    "Maximum dosage-sum error within locus",
    "Maximum G reconstruction difference",
    "Rows of G",
    "Columns of G",
    "Rank of G",
    "Folds",
    "Training individuals per fold",
    "Validation individuals per fold",
    "Maximum symmetry difference of G",
    "Mean diagonal of G",
    "Smallest eigenvalue of G"
  ),
  # List the values corresponding to the table entries.
  Value = c(
    length(ids),
    nrow(X),
    ncol(X),
    qr(X)$rank,
    nrow(M_full),
    ncol(M_full),
    qr(M_full)$rank,
    theoretical_rank_M,
    maximum_M_column_mean,
    maximum_dosage_sum_error,
    maximum_G_reconstruction_difference,
    nrow(G_full),
    ncol(G_full),
    qr(G_full)$rank,
    N_FOLDS,
    length(ids) - unique(fold_sizes),
    unique(fold_sizes),
    max(abs(G_full - t(G_full))),
    mean(diag(G_full)),
    min(eigenvalues_G_full)
  )
)

# Format the table for display in the tutorial.
kable(
  matrix_inventory,
  digits = 8,
  col.names = c("Item", "Value"),
  caption = "Summary of matrices and cross-validation"
)
```

## 12. Saving matrices and folds

The `matrices_object.rds` file is the primary interface with Module 03: it gathers matrices, aligned phenotypes, metadata, frequencies, folds, and parameters in one object while preserving their classes. The CSV files are inspection-oriented outputs for checking fold allocation, sample-size summaries, allele-column audits, and the numerical inventory without opening the RDS object.

```{r save-matrix-objects}
# Collect matrices, metadata, frequencies, and folds in the object persisted by the module.
matrices_object <- list(
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,
  trait_dictionary = trait_dictionary,
  X = X,
  M = M_full,
  G = G_full,
  cv_folds = cv_folds,
  fold_summary = fold_summary,
  locus_audit = locus_audit_full,
  allele_audit = allele_audit_full,
  marker_means = marker_means_full,
  allele_frequencies = allele_frequencies_full,
  denominator = denominator_full,
  # Record parameters and thresholds required to reproduce this stage.
  settings = list(
    n_folds = N_FOLDS,
    seed_cv = SEED_CV,
    call_rate_min = CALL_RATE_MIN,
    allele_frequency_min = ALLELE_FREQUENCY_MIN,
    fixed_effects = "FAM"
  )
)

# Save the RDS object read directly by Module 03.
saveRDS(
  matrices_object,
  file.path(OUTPUT_DIR, "matrices_object.rds")
)

# Export tabular diagnostics as CSV files for inspection independently of the RDS object.
write.csv(cv_folds, file.path(OUTPUT_DIR, "cv_folds.csv"), row.names = FALSE)
write.csv(fold_summary, file.path(OUTPUT_DIR, "fold_summary.csv"), row.names = FALSE)
write.csv(
  allele_audit_full,
  file.path(OUTPUT_DIR, "allele_audit_full.csv"),
  row.names = FALSE
)
write.csv(
  matrix_inventory,
  file.path(OUTPUT_DIR, "matrix_inventory.csv"),
  row.names = FALSE
)
```

## 13. Next step

Module 03 uses `X`, `M`, `G`, and the five folds to fit RR-BLUP, BayesA,
BayesB, BayesB2, BayesC, Bayesian Lasso, and GBLUP. In every fit, only the
phenotypes of validation individuals are masked.
