---
title: "Construction of matrices for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

In this module, the preprocessed SSR genotypes are converted into allele
dosages. These dosages are used to construct the centered marker matrix `M`,
the additive genomic relationship matrix `G`, and the fixed-effect matrix `X`,
which contains family in addition to the intercept. Five family-stratified
cross-validation folds are also defined.

The matrices are constructed once using all 150 individuals and remain
unchanged across folds. In Module 03, only the phenotype vector is masked to
form training and validation sets.

Because candidates in every fold are already genotyped, marker filtering,
allele frequencies, mean imputation, `M`, and `G` are defined from the full
genotype panel and then held fixed. No validation phenotype enters these
steps. This corresponds to a transductive candidate-genotype setting and
should not be interpreted as external validation in a new genotype population.

The matrix core of the contract delivered to Module 03 consists of four components: the fixed-effect matrix `X`, the centered marker matrix `M`, the genomic relationship matrix `G`, and the `cv_folds` table identifying each individual's validation fold. The persisted object also preserves identifiers, metadata, aligned phenotypes, the trait dictionary, and auxiliary diagnostics. The diagnostics check allele coding, centering, genomic relationships, and family stratification.

The functional workflow of the module follows this sequence:

**receive the preprocessed object → code allele dosages → filter allele columns → impute and center `M` → construct `G` → construct `X` → define the folds → audit the matrices → export the contract to Module 03.**

## 2. Analysis preparation

Preparation brings together the packages, helper functions, and object produced in Module 01. Methodological decisions remain close to the stages where they are applied, while this section makes explicit the contract that must remain stable throughout matrix construction.

### 2.1 Packages and helper functions

The packages used in this module have distinct roles:

- `dplyr`: selects, transforms, and summarizes tabular data;
- `tidyr`: reshapes data for diagnostics and figures;
- `ggplot2`: builds allele-frequency and genomic-relationship figures;
- `ggthemes`: applies the GDocs theme and color scales used in the figures;
- `knitr`: formats tables in R Markdown documents;
- `here`: builds project-relative file paths that do not depend on the local working directory.

```{r packages, warning=FALSE, message=FALSE}
library(dplyr)   # Manipulate, transform, and summarize tables.
library(tidyr)   # Reshape data between formats.
library(ggplot2) # Build figures and heatmaps.
library(ggthemes) # Apply the GDocs theme and color scales.
library(knitr)   # Format tables and R Markdown output.
library(here)    # Build portable project-relative paths.
```

```{r table-formatting, include=FALSE}
format_table <- function(x, caption = NULL, digits = getOption("digits"), col.names = NA) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    col.names = col.names,
    table.attr = paste0(
      'class="table table-striped table-hover" ',
      'style="width:100%; margin-bottom:0;"'
    )
  )
  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

### 2.2 Input object and matrix contract

The `preprocessing_object.rds` file produced in Module 01 contains the individuals, phenotypes, families, and SSR-locus information required to construct the matrices. It is read directly from the preprocessing output directory, and the components reused in this module are then recovered explicitly.

Methodological parameters specific to Module 02 are defined only in the sections where they are applied: the allele-frequency threshold beside the dosage-column filter, and the number of folds and random seed beside cross-validation construction. The call-rate threshold was already applied in Module 01; its value remains recorded in the preprocessing object and is only carried forward as provenance in the final object from this module.


The contract for this stage preserves all 150 individuals and keeps phenotypes and metadata aligned. Phenotypes do not enter allele-column filtering, frequency calculation, imputation, or construction of `M` and `G`; `FAM` is used to construct `X` and stratify the folds. Under the transductive setting adopted here, `X`, `M`, `G`, and `cv_folds` are constructed once and remain fixed for all traits and methods in Module 03.

The persisted interface consists of `matrices_object.rds`, which gathers the objects required by the next module, and CSV files that record the validation partition and matrix diagnostics for independent inspection.

```{r input-object}
preprocessing <- readRDS(
  here(
    "output",
    "preprocessing",
    "preprocessing_object.rds"
  )
)

ids <- preprocessing$ids
metadata <- preprocessing$metadata
phenotypes <- preprocessing$phenotypes
ssr_raw <- preprocessing$ssr_raw
locus_audit_full <- preprocessing$locus_audit
trait_dictionary <- preprocessing$trait_dictionary
```

### 2.3 Check of the received object

Before marker coding, the following table checks the structure received from Module 01: numbers of individuals and traits, loci before and after quality control, and families. The comparison confirms that the input object has the expected structure; no new filter is applied at this stage.

```{r input-summary}
# Summarize the dimensions and categories of the preprocessing object received by the module.
data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Original SSR loci",
    "Retained SSR loci",
    "Families"
  ),
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    sum(locus_audit_full$Keep_locus),
    nlevels(metadata$FAM)
  )
) |>
  format_table(caption = "Input data for matrix construction")
```

The check shows `r length(ids)` individuals, `r ncol(phenotypes)` phenotype traits, `r ncol(ssr_raw)` original SSR loci, and `r sum(locus_audit_full$Keep_locus)` loci retained after Module 01 quality control. Family remains represented by `r nlevels(metadata$FAM)` levels of `FAM`.


## 3. Allele-dosage coding

### 3.1 From SSR code to allele dosage

Each SSR code contains two allele labels. For example, `12` represents one copy of allele 1 and one copy of allele 2, whereas `22` represents two copies of allele 2. For every observed allele at each retained locus, an individual's dosage is computed as the sum of two indicator functions, one for each allele position in the genotype. The resulting value is 0, 1, or 2 according to whether the individual carries zero, one, or two copies of that allele.

One column is created for every locus–allele combination, and missing SSR calls remain `NA` until imputation. Before any allele-column filtering, dosages across all alleles at a diploid locus sum to 2 for every observed genotype. This property is later used as a coding diagnostic.

### 3.2 Expanding loci into allele columns

Coding proceeds across retained loci, identifies the alleles observed at each locus, and creates one dosage column for every locus–allele combination. To keep this transformation traceable, each combination receives an identifier in the form `Locus__Aallele`. The same identifier names the dosage column and appears in `allele_audit_full`, linking the matrix back to its source locus and allele.

The nested traversal below follows that logic directly: it first iterates over loci and, within each locus, over the observed alleles. Missing SSR calls remain `NA`, and no individual is removed.

```{r allele-dosage-coding}
# Retain only loci that passed Module 01 quality control.
ssr_full <- ssr_raw[
  ,
  locus_audit_full$Locus[locus_audit_full$Keep_locus],
  drop = FALSE
]

# Accumulate dosage columns and the identity of each locus-allele combination.
dosage_parts_full <- list()
allele_map_parts_full <- list()

for (locus_name in colnames(ssr_full)) {
  # Separate the two alleles encoded in each SSR genotype at the current locus.
  codes <- ssr_full[, locus_name]
  allele_1 <- floor(codes / 10)
  allele_2 <- codes %% 10

  # List the alleles observed at the current locus.
  observed_alleles <- sort(
    unique(
      c(
        allele_1[!is.na(codes)],
        allele_2[!is.na(codes)]
      )
    )
  )

  for (allele in observed_alleles) {
    # Create a unique identifier for the locus-allele combination.
    marker_name <- paste0(
      locus_name,
      "__A",
      allele
    )

    # Count 0, 1, or 2 copies of the allele for each individual.
    dosage <-
      as.numeric(allele_1 == allele) +
      as.numeric(allele_2 == allele)
    dosage[is.na(codes)] <- NA_real_

    # Use the dosage-column identifier itself as the key in both lists.
    dosage_parts_full[[marker_name]] <- dosage
    allele_map_parts_full[[marker_name]] <- data.frame(
      Locus = locus_name,
      Allele = allele,
      Marker = marker_name
    )
  }
}

dosage_full_raw <- do.call(cbind, dosage_parts_full)
allele_audit_full <- bind_rows(allele_map_parts_full)

rownames(dosage_full_raw) <- ids
colnames(dosage_full_raw) <- allele_audit_full$Marker
rownames(allele_audit_full) <- NULL

# Show how each dosage column maps back to its source locus and allele.
head(allele_audit_full, 10) |>
  format_table(caption = "Mapping between dosage columns and SSR alleles")

# Display a preview of the resulting dosage columns.
as.data.frame(dosage_full_raw[1:6, 1:8]) |>
  format_table(caption = "Preview of allele dosages")
```

The coding produces `r ncol(dosage_full_raw)` dosage columns from the `r ncol(ssr_full)` retained loci. The first table shows how each column maps to the locus and allele it represents; the second shows dosages for the first individuals. Within each column, values 0, 1, and 2 give the number of copies of the allele identified after `__A`, whereas `NA` denotes a missing SSR call. One locus can produce more than two columns when more than two alleles are represented in the panel.

## 4. Quality control of allele-dosage columns

### 4.1 Allele frequencies and retention criteria

The frequency of an allele column is computed as the mean observed dosage divided by 2 because each diploid individual can contribute two copies of the allele. `NA` values are ignored here, so the frequency uses available genotype calls only.


Columns are retained when frequency lies between 0.05 and 0.95 and dosage standard deviation is positive. The lower and upper limits are symmetric: they remove very rare alleles and nearly fixed columns, while `Dosage_sd > 0` guarantees observed variation. The filter acts on allele-dosage columns; it does not remove individuals.

```{r allele-frequency-control}
# Define the frequency threshold immediately before applying it to the allele-column filter.
ALLELE_FREQUENCY_MIN <- 0.05

# Compute dosage frequency and variability to retain informative allele columns.
allele_audit_full$Allele_frequency <-
  colMeans(dosage_full_raw, na.rm = TRUE) / 2

allele_audit_full$Dosage_sd <- apply(
  dosage_full_raw,
  2,
  sd,
  na.rm = TRUE
)

allele_audit_full$Keep_allele <-
  is.finite(allele_audit_full$Allele_frequency) &
  allele_audit_full$Allele_frequency >= ALLELE_FREQUENCY_MIN &
  allele_audit_full$Allele_frequency <= (1 - ALLELE_FREQUENCY_MIN) &
  is.finite(allele_audit_full$Dosage_sd) &
  allele_audit_full$Dosage_sd > 0

# Keep only allele columns that satisfy the frequency and variation criteria.
dosage_full <- dosage_full_raw[
  ,
  allele_audit_full$Marker[allele_audit_full$Keep_allele],
  drop = FALSE
]

# Directly summarize how many allele columns were observed and retained at each locus.
allele_audit_full |>
  group_by(Locus) |>
  summarise(
    Observed_alleles = n(),
    Retained_alleles = sum(Keep_allele),
    .groups = "drop"
  ) |>
  format_table(
    caption = "Allele-dosage columns by SSR locus",
    col.names = c("Locus", "Observed alleles", "Retained alleles")
  )
```

Of the `r nrow(allele_audit_full)` initially coded allele columns, `r sum(allele_audit_full$Keep_allele)` satisfy both the frequency interval and positive-variation requirement. The table shows how many alleles were observed and retained at each locus.

### 4.2 Distribution of allele frequencies

The histogram complements the preceding table by showing where allele-column frequencies are concentrated and how that distribution relates to the final filtering decision. The horizontal axis covers the full possible frequency interval from 0 to 1 in classes of width 0.05. Dashed lines mark the 0.05 and 0.95 frequency limits, whereas GDocs colors represent the final retained or removed decision.

These two pieces of information should be distinguished: the lines show only the frequency criterion, whereas color also incorporates the requirement of positive dosage standard deviation. The visual classification of a column should therefore not be inferred from its position relative to the vertical lines alone.

```{r allele-frequency-plot, fig.height=5, fig.alt="Histogram of SSR allele-column frequencies, with dashed lines at 0.05 and 0.95 and GDocs colors indicating the final retained or removed decision."}
ggplot(
  allele_audit_full,
  aes(
    x = Allele_frequency,
    fill = factor(
      Keep_allele,
      levels = c(TRUE, FALSE),
      labels = c("Retained", "Removed")
    )
  )
) +
  geom_histogram(
    binwidth = 0.05,
    boundary = 0
  ) +
  geom_vline(
    xintercept = c(
      ALLELE_FREQUENCY_MIN,
      1 - ALLELE_FREQUENCY_MIN
    ),
    linetype = 2,
    linewidth = 0.6
  ) +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = paste0(seq(0, 100, 20), "%"),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_gdocs(name = "Decision") +
  labs(
    x = "Allele frequency",
    y = "Number of allele columns",
    title = "Distribution of allele frequencies",
    subtitle = "Dashed lines mark 5% and 95%; color represents the final filtering decision",
    caption = "Retention simultaneously requires frequency between 0.05 and 0.95 and positive dosage variation."
  ) +
  theme_gdocs() +
  theme(legend.position = "top")
```

Observed allele frequencies range from `r sprintf("%.3f", min(allele_audit_full$Allele_frequency, na.rm = TRUE))` to `r sprintf("%.3f", max(allele_audit_full$Allele_frequency, na.rm = TRUE))`. Bar position shows the frequency range of allele columns, whereas color reports the joint outcome of the retention criteria. The dashed lines are only references for the frequency threshold; a column must also have `Dosage_sd > 0` to remain in `dosage_full`.

## 5. Construction of marker matrix M

### 5.1 Imputation of missing dosages

Missing dosages are imputed with the mean of the corresponding column computed from the full genotype panel. Column means are calculated first. Then `which(..., arr.ind = TRUE)` identifies the row and column of every missing cell; the column index directly retrieves the mean of the corresponding marker. All missing values therefore follow the same replacement rule without creating a new genotype category and without using phenotype information.

The table compares the matrix before and after imputation, reporting both the total number of missing dosages and the number of columns that still contain at least one missing value.

```{r dosage-imputation}
# Compute the mean of each retained dosage column in the full genotype panel.
marker_means_full <- colMeans(
  dosage_full,
  na.rm = TRUE
)

# Locate all missing cells and replace each one with the mean of its column.
dosage_full_imputed <- dosage_full
missing_dosage_positions <- which(
  is.na(dosage_full_imputed),
  arr.ind = TRUE
)

dosage_full_imputed[missing_dosage_positions] <-
  marker_means_full[missing_dosage_positions[, "col"]]

# Compare matrix completeness before and after imputation.
data.frame(
  Stage = c("Before imputation", "After imputation"),
  Missing_dosages = c(
    nrow(missing_dosage_positions),
    sum(is.na(dosage_full_imputed))
  ),
  Columns_with_missing = c(
    sum(colSums(is.na(dosage_full)) > 0),
    sum(colSums(is.na(dosage_full_imputed)) > 0)
  )
) |>
  format_table(
    caption = "Dosage completeness before and after imputation",
    col.names = c("Stage", "Missing dosages", "Columns with missing values")
  )
```

Before imputation, `r nrow(missing_dosage_positions)` cells were missing across `r sum(colSums(is.na(dosage_full)) > 0)` dosage columns. After replacement by the corresponding column means, `r sum(is.na(dosage_full_imputed))` missing values remain. The transformation uses only genotype information available in the full panel.

### 5.2 Centering and construction of M

After imputation, each column is centered by subtracting exactly the same mean stored in `marker_means_full`. The resulting matrix is denoted by \(\mathbf M\). Its columns are centered but are not standardized to unit variance.

At positions that were originally missing, the imputed dosage equals its own column mean. Subtracting that same mean therefore makes those positions equal to zero in `M`. Zero represents no deviation from the column mean rather than a new genotype class. Imputation and centering use genotype information from all 150 candidates; no validation phenotype participates in these operations.

```{r marker-matrix}
M_full <- sweep(
  dosage_full_imputed,
  2,
  marker_means_full,
  FUN = "-"
)
```

### 5.3 Dimensions, centering, and rank of M

The summary of `M` checks four basic properties: number of individuals, number of retained allele-dosage columns, matrix rank, and maximum absolute column mean. Rank represents the number of linearly independent directions in `M`, whereas an absolute mean close to zero numerically confirms that centering was applied as intended.

```{r marker-matrix-summary}
data.frame(
  Rows = nrow(M_full),
  Columns = ncol(M_full),
  Rank = qr(M_full)$rank,
  Maximum_absolute_column_mean = max(abs(colMeans(M_full)))
) |>
  format_table(
    digits = 8,
    caption = "Dimensions and centering of the marker matrix",
    col.names = c("Rows", "Columns", "Rank", "Maximum absolute column mean")
  )
```

Matrix `M` contains `r nrow(M_full)` rows and `r ncol(M_full)` retained dosage columns. Its rank is `r qr(M_full)$rank`, below the number of columns because dosage columns from the same locus can be linearly dependent. The largest absolute column mean is `r format(max(abs(colMeans(M_full))), scientific = TRUE, digits = 3)`, confirming numerical centering after imputation.

## 6. Construction of genomic relationship matrix G

### 6.1 Definition and scale of G

Let \(p_j\) be the allele frequency represented by column \(j\). Using a
VanRaden-type allele-frequency scaling [@vanraden2008efficient], the additive
genomic relationship matrix is calculated as

\[
\mathbf G =
\frac{\mathbf M\mathbf M^{\mathsf T}}
{\sum_j 2p_j(1-p_j)}.
\]

Because an SSR locus can generate several allele-dosage columns, these columns should not be interpreted as independent markers. When all alleles from the same diploid locus remain represented, their dosages sum to 2 within each individual. After centering, the deviations across those columns sum to zero, creating a structural linear dependence among them.

This dependence does not prevent construction of `G`. Matrix `G` does not require inversion of `M` or separate estimation of an identifiable coefficient for every allele column. Instead, the cross-product `M M^T` aggregates, for each pair of individuals, the agreement between their allele-dosage deviation profiles. A multiallelic locus therefore contributes to genomic relationship through its set of alleles, while dependence among its columns remains an expected property of the coding.

### 6.2 Construction of G

Allele frequencies used in the scaling are obtained from dosage means because mean dosage equals \(2p_j\). For each allele column, the term \(2p_j(1-p_j)\) provides a frequency-based variability scale; summing these terms gives a global denominator across all columns. Because `M` is centered but not standardized column by column, this denominator converts the raw cross-product to a common genomic-relationship scale.

For two individuals, an off-diagonal cell of \(\mathbf M\mathbf M^{\mathsf T}\) is the inner product of their centered dosage profiles. Deviations in the same direction contribute positively, whereas deviations in opposite directions contribute negatively. After division by the denominator, off-diagonal elements of `G` describe relative genomic relationship between pairs of individuals, using allele frequencies in this panel as the reference. A negative value indicates lower genomic similarity relative to that reference rather than a biologically antagonistic relationship.

On the diagonal, each individual is compared with itself; diagonal elements therefore summarize self-relatedness on the adopted scale. Because the normalization is global and does not force each individual diagonal separately, these values need not equal exactly 1. The resulting matrix is numerically symmetrized before its eigenvalues are evaluated.

```{r genomic-relationship-matrix}
# Recover p_j from the dosage mean, which equals 2p_j.
allele_frequencies_full <- marker_means_full / 2

# Compute the VanRaden scaling from allele-column frequencies.
denominator_full <- sum(
  2 *
    allele_frequencies_full *
    (1 - allele_frequencies_full)
)


# Convert cross-products of centered dosages into scaled genomic relationships.
G_full <- tcrossprod(M_full) / denominator_full

# Record numerical asymmetry of the scaled cross-product before rounding correction.
maximum_G_asymmetry_before_symmetrization <- max(
  abs(G_full - t(G_full))
)

# Remove only numerical asymmetry.
G_full <- (G_full + t(G_full)) / 2

dimnames(G_full) <- list(ids, ids)

# Keep only the smallest eigenvalue, which is the diagnostic used downstream.
minimum_eigenvalue_G_full <- min(
  eigen(
    G_full,
    symmetric = TRUE,
    only.values = TRUE
  )$values
)
```

### 6.3 Numerical diagnostics of G

The denominator reports the scale used to construct `G`. The mean diagonal summarizes each individual's relationship with itself on this scale and is not forced to equal exactly 1. To avoid making the symmetry diagnostic tautological, numerical asymmetry is measured on the scaled cross-product before symmetrization; this value should be close to zero. Because `G` is built from a cross-product, it is theoretically positive semidefinite; therefore, a slightly negative smallest eigenvalue, when limited to numerical precision, should be interpreted as rounding rather than evidence of negative relatedness.

```{r genomic-relationship-summary}
# Build or check the matrices used for genomic prediction.
data.frame(
  Quantity = c(
    "Denominator sum 2p(1-p)",
    "Mean diagonal",
    "Maximum asymmetry before symmetrization",
    "Smallest eigenvalue"
  ),
  Value = c(
    denominator_full,
    mean(diag(G_full)),
    maximum_G_asymmetry_before_symmetrization,
    minimum_eigenvalue_G_full
  )
) |>
  format_table(
    digits = 8,
    caption = "Numerical summary of the genomic relationship matrix"
  )
```

Before symmetrization, the maximum asymmetry of the scaled cross-product is `r format(maximum_G_asymmetry_before_symmetrization, scientific = TRUE, digits = 3)`, showing the magnitude of the numerical error removed by averaging with the transpose. The mean diagonal of `G` is `r sprintf("%.3f", mean(diag(G_full)))`, and its smallest eigenvalue is `r format(minimum_eigenvalue_G_full, scientific = TRUE, digits = 3)`. Negative values of very small magnitude are compatible with numerical rounding.

## 7. Fixed effects and cross-validation design

### 7.1 Fixed-effect matrix X

Family is the only categorical fixed effect. Because `metadata$FAM` is already a factor, `model.matrix(~ FAM)` uses its levels directly to build the design matrix. Under R's default treatment contrasts, `X` contains one intercept column and `nlevels(FAM) - 1` indicator columns, which are sufficient to represent differences among families without treating `FAM` as a numeric variable.

The first diagnostic summarizes only the structure of `X`. A small preview then places `IND` and `FAM` beside the matrix columns so the intercept and family indicators can be seen for the first individuals. The joint comparison of `X`, `M`, and `G` is kept in the consolidated inventory in Section 8, avoiding repetition of diagnostics already presented for `M` and `G`.

In Module 03, the intercept column is removed from `X` because BGLR already fits an intercept automatically; retaining both would duplicate the same constant effect.

```{r fixed-effect-matrix}
# Build the design matrix for the only categorical fixed effect used in the analysis.
X <- model.matrix(
  ~ FAM,
  data = metadata
)

rownames(X) <- ids

# Summarize the structure specific to the fixed-effect matrix.
data.frame(
  Quantity = c(
    "Individuals",
    "FAM levels",
    "Columns of X",
    "Rank of X"
  ),
  Value = c(
    nrow(X),
    nlevels(metadata$FAM),
    ncol(X),
    qr(X)$rank
  )
) |>
  format_table(caption = "Structure of the fixed-effect matrix X")

# Show how family, intercept, and indicators appear for the first individuals.
data.frame(
  IND = ids[1:6],
  FAM = metadata$FAM[1:6],
  X[1:6, , drop = FALSE],
  check.names = FALSE
) |>
  format_table(caption = "Preview of family coding in matrix X")
```

Matrix `X` has `r nrow(X)` rows, `r ncol(X)` columns, and rank `r qr(X)$rank`. Because `FAM` has `r nlevels(metadata$FAM)` levels, the coding contains one intercept and `r ncol(X) - 1` family indicators. The preview makes this representation explicit without repeating the diagnostics for `M` and `G`.

### 7.2 Evaluation scenario

Family stratification targets prediction among individuals from the represented breeding population. It is not a leave-family-out design and therefore does not estimate predictive performance for entirely new families.

### 7.3 Construction of the stratified folds

The 150 individuals are distributed into five validation groups. Allocation proceeds family by family: individuals in the current family are first shuffled; folds are then ordered by their accumulated total load, with a random order used only to break ties between equally loaded folds. Family members are subsequently distributed cyclically along that order. The algorithm therefore combines family representation with balanced total fold sizes.

The fixed seed preserves exactly the same partition across executions. Because the `sample()` calls are part of the tie-breaking and within-family shuffling rules, their sequence is kept explicit in the code. The same `cv_folds` table will be used for all ten traits and all methods.

```{r cross-validation-folds}
# Define cross-validation parameters immediately before constructing the folds.
N_FOLDS <- 5L
SEED_CV <- 20260723L

set.seed(SEED_CV)

# Initialize each individual's family, fold assignment, and the accumulated fold loads.
family <- droplevels(metadata$FAM)
folds <- integer(length(ids))
fold_load <- integer(N_FOLDS)

for (family_level in levels(family)) {
  # Randomize the individuals of the current family.
  family_indices <- sample(
    which(family == family_level)
  )

  # Draw an order used only to break ties between folds with equal accumulated load.
  random_tie_order <- sample(seq_len(N_FOLDS))

  ordered_folds <- order(
    fold_load,
    match(seq_len(N_FOLDS), random_tie_order)
  )

  # Distribute family members cyclically according to the fold order.
  family_labels <- rep(
    ordered_folds,
    length.out = length(family_indices)
  )

  folds[family_indices] <- family_labels

  # Update total fold loads before processing the next family.
  fold_load <- fold_load + tabulate(
    family_labels,
    nbins = N_FOLDS
  )
}

# Record the partition that will be used directly in Module 03.
cv_folds <- data.frame(
  IND = ids,
  FAM = as.character(metadata$FAM),
  Fold = folds
)
```

After allocation, `cv_folds` contains `r nrow(cv_folds)` rows, one for each candidate, and assigns every individual to exactly one of the five folds. The next subsections separately diagnose fold size and family balance.

### 7.4 Training and validation sizes

Once the partition is defined, the number of individuals in each fold is counted directly from `cv_folds$Fold`. For each round, this count is the validation set size; all remaining individuals form the training set.

```{r fold-sizes}
fold_sizes <- tabulate(
  cv_folds$Fold,
  nbins = N_FOLDS
)

fold_summary <- data.frame(
  Fold = seq_len(N_FOLDS),
  Training_n = length(ids) - fold_sizes,
  Validation_n = fold_sizes
)

format_table(
  fold_summary,
  col.names = c("Fold", "Training N", "Validation N"),
  caption = "Sample sizes in cross-validation"
)
```

All five folds contain `r unique(fold_summary$Validation_n)` validation individuals and `r unique(fold_summary$Training_n)` training individuals. The table makes the constant sample size across the five rounds directly visible.

### 7.5 Family balance across folds

The second diagnostic crosses the family used for stratification with the assigned fold. For each family, `family_balance` calculates the difference between its largest and smallest count across the five folds. Because members of each family are distributed cyclically along the fold order, this difference should be at most one individual.

```{r families-by-fold}
family_fold_table <- table(
  Family = family,
  Fold = cv_folds$Fold
)

family_balance <- apply(family_fold_table, 1, max) -
  apply(family_fold_table, 1, min)

# Display both fold counts and the max-minus-min diagnostic for each family.
data.frame(
  Family = rownames(family_fold_table),
  as.data.frame.matrix(family_fold_table),
  Max_min_difference = as.integer(family_balance),
  check.names = FALSE
) |>
  format_table(
    caption = "Distribution of families across folds",
    col.names = c(
      "Family",
      paste0("Fold ", seq_len(N_FOLDS)),
      "Max-min difference"
    )
  )
```

The largest difference between the maximum and minimum count for the same family across folds is `r max(family_balance)` individual. The table therefore shows both family representation across folds and the quality of the achieved balance.

### 7.6 Using the folds for model fitting

The `cv_folds` table identifies which individuals are masked in each validation round. For fold `k`, the 30 individuals with `Fold == k` have their phenotypes replaced by `NA`, while the other 120 provide training phenotypes. This procedure is repeated for every trait and method.

`X`, `M`, and `G` remain fixed and include all 150 genotyped candidates, consistent with the transductive design defined at the start of the module. Consequently, every individual receives a prediction when its own phenotype is masked, producing an out-of-fold prediction downstream; this is not equivalent to predicting an external genotype population.

The table below shows a sample of the individual-to-fold assignments. The complete `cv_folds` object retains all 150 candidates and records the fold used in Module 03 to determine which phenotypes are masked in each validation round.

```{r fold-example}
cv_folds |>
  arrange(Fold, FAM, IND) |>
  head(15) |>
  format_table(caption = "Example of individual allocation to folds")
```

The 15 displayed rows make the allocation structure visible without modifying `cv_folds`, which retains all 150 individuals and is used directly in Module 03.

## 8. Integrated matrix diagnostics

### 8.1 Genomic relationship heatmap

The heatmap displays all 22,500 cells of `G` in a single view. The diagonal contains each individual's relationship with itself; off the diagonal, each cell represents relative genomic relationship between a pair of candidates. Values close to zero indicate relationship near the reference defined by allele frequencies in this panel, whereas negative and positive values indicate lower and higher genomic similarity relative to that reference, respectively.

Individuals remain in the original order of `G`: no hierarchical clustering, seriation, or other reordering is applied to create visual blocks. The fill scale is continuous and uses the GDocs palette; the legend explicitly marks the observed minimum, zero, and observed maximum. Intensity patterns can therefore be used as a descriptive diagnostic of relationship structure, but they are not a clustering or population-structure test.

```{r genomic-relationship-plot, fig.width=8, fig.height=8, fig.alt="Square heatmap of the genomic relationship matrix G for 150 individuals in their original order and without clustering, using a continuous GDocs scale for negative, near-zero, and positive relative relationships."}
# Reshape G directly to long format and pass the data to the plot without creating an intermediate object.
as.data.frame(as.table(G_full)) |>
  setNames(c("Individual_1", "Individual_2", "Relationship")) |>
  ggplot(
    aes(
      x = Individual_1,
      y = Individual_2,
      fill = Relationship
    )
  ) +
  geom_raster() +
  scale_x_discrete(
    breaks = NULL,
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    breaks = NULL,
    expand = c(0, 0)
  ) +
  scale_fill_gradientn(
    colours = ggthemes::gdocs_pal()(7),
    breaks = c(
      min(G_full),
      0,
      max(G_full)
    ),
    labels = function(x) sprintf("%.2f", x),
    name = "Relative
relationship"
  ) +
  coord_fixed() +
  labs(
    x = "Individuals in original order",
    y = "Individuals in original order",
    title = "Additive genomic relationship matrix",
    subtitle = "No clustering or reordering; zero represents the panel-relative reference",
    caption = "The diagonal represents self-relationship; all other cells represent pairwise genomic relationships."
  ) +
  theme_gdocs() +
  theme(
    legend.position = "top",
    axis.ticks = element_blank()
  )
```

The heatmap preserves the complete structure of `G` and the candidates' original order. The diagonal tends to concentrate self-relationship values, while off-diagonal cells show variation in pairwise relationship around the zero reference. Because no reordering is applied, visually similar regions reflect the existing data order and should be interpreted only as descriptive patterns rather than groups inferred by the figure.

### 8.2 Structural dependence and rank of M

All allele columns satisfying the declared filters are retained; no category is artificially discarded merely to force full rank. When all alleles at a locus remain represented, their dosages sum to 2 in each diploid individual and an exact linear dependency exists among those columns. Individual allele coefficients are therefore not interpreted in isolation; later modules focus on predictions and genomic values.

The `r ncol(M_full)` dosage columns do not represent `r ncol(M_full)` independent dimensions. For each locus with \(k_l\) retained allele columns, the diploid constraint allows at most \(k_l-1\) independent directions among those columns. The structural rank bound of `M` sums these contributions across loci and is compared with the observed rank to determine whether additional linear dependencies occur.

### 8.3 Consolidated numerical checks

The diploid coding property is checked before allele-column filtering: for every retained SSR locus from Module 01, dosages across all alleles originally coded for an observed genotype must sum to 2. Missing SSR calls are excluded from this check rather than imputed for it.

Reconstruction of `G` repeats the complete construction from Section 6: the cross-product of `M`, division by the denominator, and the same numerical symmetrization. The difference between `G_full` and `G_reconstructed` therefore checks consistency of the final matrix actually used, whereas `maximum_G_asymmetry_before_symmetrization`, computed earlier, separately quantifies numerical asymmetry before that correction. The inventory also consolidates centering, dimensions, ranks, fold sizes, the diagonal, and the smallest eigenvalue.

```{r matrix-diagnostics}
# Compute the rank bound imposed by dependencies among alleles from the same locus.
structural_rank_bound_M <- sum(
  table(
    allele_audit_full$Locus[
      allele_audit_full$Keep_allele
    ]
  ) - 1L
)

# Reconstruct exactly the G matrix used in the analysis, including numerical symmetrization.
G_reconstructed <- tcrossprod(M_full) / denominator_full
G_reconstructed <- (G_reconstructed + t(G_reconstructed)) / 2

maximum_G_reconstruction_difference <- max(
  abs(G_full - G_reconstructed)
)

# Check the diploid sum using every allele originally coded for each locus.
maximum_dosage_sum_error <- 0

for (locus_name in unique(allele_audit_full$Locus)) {
  locus_columns <- allele_audit_full$Marker[
    allele_audit_full$Locus == locus_name
  ]

  observed_rows <- !is.na(ssr_full[, locus_name])

  locus_dosage_sum_error <- max(
    abs(
      rowSums(
        dosage_full_raw[
          observed_rows,
          locus_columns,
          drop = FALSE
        ]
      ) - 2
    )
  )

  maximum_dosage_sum_error <- max(
    maximum_dosage_sum_error,
    locus_dosage_sum_error
  )
}

# Keep the inventory numeric for export and reuse independently of HTML formatting.
matrix_inventory <- data.frame(
  Item = c(
    "individuals",
    "x_rows",
    "x_columns",
    "x_rank",
    "m_rows",
    "m_columns",
    "m_rank",
    "m_structural_rank_bound",
    "m_max_abs_column_mean",
    "max_diploid_dosage_sum_error",
    "g_max_reconstruction_difference",
    "g_rows",
    "g_columns",
    "g_rank",
    "cv_folds",
    "training_individuals_per_fold",
    "validation_individuals_per_fold",
    "g_max_asymmetry_before_symmetrization",
    "g_mean_diagonal",
    "g_min_eigenvalue"
  ),
  Value = c(
    length(ids),
    nrow(X),
    ncol(X),
    qr(X)$rank,
    nrow(M_full),
    ncol(M_full),
    qr(M_full)$rank,
    structural_rank_bound_M,
    max(abs(colMeans(M_full))),
    maximum_dosage_sum_error,
    maximum_G_reconstruction_difference,
    nrow(G_full),
    ncol(G_full),
    qr(G_full)$rank,
    N_FOLDS,
    length(ids) - unique(fold_sizes),
    unique(fold_sizes),
    maximum_G_asymmetry_before_symmetrization,
    mean(diag(G_full)),
    minimum_eigenvalue_G_full
  )
)

# Format only the copy passed to the table; `matrix_inventory` remains numeric.
matrix_inventory |>
  mutate(
    Item = case_match(
      Item,
      "individuals" ~ "Individuals",
      "x_rows" ~ "Rows of X",
      "x_columns" ~ "Columns of X",
      "x_rank" ~ "Rank of X",
      "m_rows" ~ "Rows of M",
      "m_columns" ~ "Columns of M",
      "m_rank" ~ "Rank of M",
      "m_structural_rank_bound" ~ "Structural rank bound of M",
      "m_max_abs_column_mean" ~ "Maximum absolute column mean of M",
      "max_diploid_dosage_sum_error" ~ "Maximum dosage-sum error within locus",
      "g_max_reconstruction_difference" ~ "Maximum G reconstruction difference",
      "g_rows" ~ "Rows of G",
      "g_columns" ~ "Columns of G",
      "g_rank" ~ "Rank of G",
      "cv_folds" ~ "Folds",
      "training_individuals_per_fold" ~ "Training individuals per fold",
      "validation_individuals_per_fold" ~ "Validation individuals per fold",
      "g_max_asymmetry_before_symmetrization" ~ "Maximum G asymmetry before symmetrization",
      "g_mean_diagonal" ~ "Mean diagonal of G",
      "g_min_eigenvalue" ~ "Smallest eigenvalue of G"
    ),
    Value = case_when(
      row_number() %in% c(1:8, 12:17) ~ sprintf("%.0f", Value),
      row_number() %in% c(9:11, 18, 20) ~ format(
        Value,
        scientific = TRUE,
        digits = 3
      ),
      TRUE ~ sprintf("%.3f", Value)
    )
  ) |>
  format_table(
    col.names = c("Item", "Value"),
    caption = "Summary of matrices and cross-validation"
  )
```

The final inventory consolidates the diagnostics presented throughout the module. For the analyzed data, `M` contains `r ncol(M_full)` columns and rank `r qr(M_full)$rank`, compared with the structural rank bound `r structural_rank_bound_M`; the maximum diploid-sum error is `r format(maximum_dosage_sum_error, scientific = TRUE, digits = 3)`, and the maximum difference in the complete reconstruction of `G` is `r format(maximum_G_reconstruction_difference, scientific = TRUE, digits = 3)`. All five folds retain the same training and validation sizes.

## 9. Exporting the matrix contract

### 9.1 Matrix object and tabular diagnostics

The `matrices_object.rds` file gathers the matrices, aligned phenotypes, metadata, frequencies, folds, and parameters required by Module 03. Because PT and EN execute the same scientific workflow, the shared artifacts in this section are language-independent: `matrices_object.rds`, `cv_folds.csv`, `fold_summary.csv`, and `allele_audit_full.csv` use common structures, and `matrix_inventory.csv` stores stable language-neutral identifiers in its `Item` field. Translated labels exist only in the HTML tables from Section 8. Therefore, PT/EN rendering order does not change the scientific content written to `output/matrices/`.

```{r save-matrix-objects}
dir.create(
  here("output", "matrices"),
  recursive = TRUE,
  showWarnings = FALSE
)

# Collect matrices, metadata, frequencies, and folds in the object persisted by the module.
matrices_object <- list(
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,
  trait_dictionary = trait_dictionary,
  X = X,
  M = M_full,
  G = G_full,
  cv_folds = cv_folds,
  fold_summary = fold_summary,
  locus_audit = locus_audit_full,
  allele_audit = allele_audit_full,
  marker_means = marker_means_full,
  allele_frequencies = allele_frequencies_full,
  denominator = denominator_full,
  # Record parameters and thresholds required to reproduce this stage.
  settings = list(
    n_folds = N_FOLDS,
    seed_cv = SEED_CV,
    call_rate_min = preprocessing$settings$call_rate_min,
    allele_frequency_min = ALLELE_FREQUENCY_MIN,
    fixed_effects = "FAM"
  )
)

# Save the RDS object read directly by Module 03.
saveRDS(
  matrices_object,
  here("output", "matrices", "matrices_object.rds")
)

# Export tabular diagnostics as CSV files for inspection independently of the RDS object.
write.csv(cv_folds, here("output", "matrices", "cv_folds.csv"), row.names = FALSE)
write.csv(fold_summary, here("output", "matrices", "fold_summary.csv"), row.names = FALSE)
write.csv(
  allele_audit_full,
  here("output", "matrices", "allele_audit_full.csv"),
  row.names = FALSE
)
write.csv(
  matrix_inventory,
  here("output", "matrices", "matrix_inventory.csv"),
  row.names = FALSE
)
```

## 10. Synthesis and transition to Module 03

### 10.1 Contract delivered to Module 03

Module 02 delivers a single reproducible contract to Module 03. Its matrix core contains `X`, `M`, `G`, and `cv_folds`, while `matrices_object.rds` also preserves `ids`, `metadata`, `phenotypes`, and `trait_dictionary` to maintain alignment and analytical context. In Module 03, `X`, `M`, `G`, and `cv_folds` remain fixed and only the phenotypes of individuals in the validation fold are masked for each fit.

### 10.2 Next step

From this contract, the next module fits RR-BLUP, BayesA, BayesB, BayesB2, BayesC, Bayesian Lasso, and GBLUP using exactly the same five folds for all ten traits. Comparisons among methods therefore start from the same genomic representation and the same validation partition constructed in this module.

<div style="display:grid; grid-template-columns:1fr auto 1fr; align-items:center; gap:1rem; margin-top:2rem; padding-top:1rem; border-top:1px solid #ddd;">
<a href="01_preprocessing_en.html">← Module 01 — Preprocessing</a>
<a href="02_matrices_pt.html" style="text-align:center;">Português</a>
<a href="03_models_en.html" style="text-align:right;">Module 03 — Model fitting →</a>
</div>
