---
title: "Construction of matrices for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

In this module, the preprocessed SSR genotypes are converted into allele
dosages. These dosages are used to construct the centered marker matrix `M`,
the additive genomic relationship matrix `G`, and the fixed-effect matrix `X`,
which contains family in addition to the intercept. Five family-stratified
cross-validation folds are also defined.

The matrices are constructed once using all 150 individuals and remain
unchanged across folds. In Module 03, only the phenotype vector is masked to
form training and validation sets.

Because candidates in every fold are already genotyped, marker filtering,
allele frequencies, mean imputation, `M`, and `G` are defined from the full
genotype panel and then held fixed. No validation phenotype enters these
steps. This corresponds to a transductive candidate-genotype setting and
should not be interpreted as external validation in a new genotype population.

The module produces four components used directly in Module 03: the fixed-effect matrix `X`, the centered marker matrix `M`, the genomic relationship matrix `G`, and the `cv_folds` table identifying each individual's validation fold. The diagnostics check allele coding, centering, genomic relationships, and family stratification.

## 2. Packages and input object

The packages used in this module have distinct roles:

- `dplyr`: selects, transforms, and summarizes tabular data;
- `tidyr`: reshapes data for diagnostics and figures;
- `ggplot2`: builds allele-frequency and genomic-relationship figures;
- `knitr`: formats tables in R Markdown documents;
- `here`: builds project-relative file paths that do not depend on the local working directory.

```{r packages, warning=FALSE, message=FALSE}
library(dplyr)   # Manipulate, transform, and summarize tables.
library(tidyr)   # Reshape data between formats.
library(ggplot2) # Build figures and heatmaps.
library(knitr)   # Format tables and R Markdown output.
library(here)    # Build portable project-relative paths.
```

```{r table-formatting, include=FALSE}
format_table <- function(x, caption = NULL, digits = getOption("digits"), col.names = NA) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    col.names = col.names,
    table.attr = paste0(
      'class="table table-striped table-hover" ',
      'style="width:100%; margin-bottom:0;"'
    )
  )
  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

### 2.1 Input object and parameters

The `preprocessing_object.rds` file produced in Module 01 supplies the individuals, phenotypes, families, and SSR loci used to construct the matrices.

Three parameters are defined in this module:

- `N_FOLDS = 5`: divides the 150 individuals into five validation groups of 30 individuals, leaving 120 training individuals per fold. Each individual appears in validation exactly once; this is not repeated cross-validation.
- `SEED_CV = 20260723`: reproduces exactly the same family-stratified allocation across folds. The seed controls partition reproducibility only and has no inferential interpretation.
- `ALLELE_FREQUENCY_MIN = 0.05`: retains allele-dosage columns with frequencies between 0.05 and 0.95 provided that dosage variation is positive. Very rare or nearly fixed alleles therefore do not enter `M`.

The call-rate threshold is recovered from Module 01 because locus filtering has already been applied before allele-dosage construction.

```{r input-object}
# Define paths and load the preprocessing object used to construct the genomic matrices.
PREPROCESSING_FILE <- here(
  "output",
  "preprocessing",
  "preprocessing_object.rds"
)

OUTPUT_DIR <- here(
  "output",
  "matrices"
)

# Use five folds: 30 validation and 120 training individuals per fold.
N_FOLDS <- 5L
# Fix only the reproducibility of the family-stratified allocation.
SEED_CV <- 20260723L
# Exclude alleles with frequency below 5% or above 95%.
ALLELE_FREQUENCY_MIN <- 0.05

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

preprocessing <- readRDS(PREPROCESSING_FILE)

ids <- preprocessing$ids
metadata <- preprocessing$metadata
phenotypes <- preprocessing$phenotypes
ssr_raw <- preprocessing$ssr_raw
locus_audit_full <- preprocessing$locus_audit
trait_dictionary <- preprocessing$trait_dictionary
# Recover, for provenance, the call-rate threshold applied in Module 01.
CALL_RATE_MIN <- preprocessing$settings$call_rate_min

# Require every received object to preserve the same individuals and row order.
stopifnot(
  length(ids) == nrow(metadata),
  length(ids) == nrow(phenotypes),
  length(ids) == nrow(ssr_raw),
  identical(as.character(metadata$IND), ids),
  identical(rownames(metadata), ids),
  identical(rownames(phenotypes), ids),
  identical(rownames(ssr_raw), ids)
)
```

### 2.2 Check of the received object

Before marker coding, the following table checks the structure received from Module 01: numbers of individuals and traits, loci before and after quality control, and families. The comparison confirms that the input object has the expected structure; no new filter is applied at this stage.

```{r input-summary}
# Summarize the dimensions and categories of the preprocessing object received by the module.
data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Original SSR loci",
    "Retained SSR loci",
    "Families"
  ),
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    sum(locus_audit_full$Keep_locus),
    nlevels(metadata$FAM)
  )
) |>
  format_table(caption = "Input data for matrix construction")
```

The check shows `r length(ids)` individuals, `r ncol(phenotypes)` phenotype traits, `r ncol(ssr_raw)` original SSR loci, and `r sum(locus_audit_full$Keep_locus)` loci retained after Module 01 quality control. Family remains represented by `r nlevels(metadata$FAM)` levels of `FAM`.

Individual identity and row order are executable contracts at this module boundary: metadata, phenotypes, and SSR genotypes must all correspond exactly to `ids` before allele coding begins. The guards validate this structure without filtering or reordering the received objects.

## 3. Allele-dosage coding

### 3.1 From SSR code to allele dosage

Each SSR code contains two allele labels. For example, `12` represents one copy of allele 1 and one copy of allele 2, whereas `22` represents two copies of allele 2. For every observed allele at each retained locus, an individual's dosage is computed as the sum of two indicator functions, one for each allele position in the genotype. The resulting value is 0, 1, or 2 according to whether the individual carries zero, one, or two copies of that allele.

One column is created for every locus–allele combination, and missing SSR calls remain `NA` until imputation. Before any allele-column filtering, dosages across all alleles at a diploid locus sum to 2 for every observed genotype. This property is later used as a coding diagnostic.

### 3.2 Expanding loci into allele columns

Coding proceeds across retained loci, identifies the alleles observed at each locus, and creates one dosage column for every locus–allele combination. Metadata preserve the correspondence between each column, its source locus, and the represented allele. The code below implements this expansion without removing individuals.

```{r allele-dosage-coding}
# Organize marker information and compute the diagnostics required at this stage.
retained_loci_full <- locus_audit_full$Locus[
  locus_audit_full$Keep_locus
]

ssr_full <- ssr_raw[
  ,
  retained_loci_full,
  drop = FALSE
]

dosage_parts_full <- list()
allele_map_parts_full <- list()
marker_counter_full <- 0L

for (j in seq_len(ncol(ssr_full))) {
  # Separate the two alleles encoded in each SSR genotype.
  codes <- ssr_full[, j]
  allele_1 <- floor(codes / 10)
  allele_2 <- codes %% 10

  # List the alleles observed at the current locus.
  observed_alleles <- sort(
    unique(
      c(
        allele_1[!is.na(codes)],
        allele_2[!is.na(codes)]
      )
    )
  )

  for (allele in observed_alleles) {
    marker_counter_full <- marker_counter_full + 1L

    # Count 0, 1, or 2 copies of the allele for each individual.
    dosage <-
      as.numeric(allele_1 == allele) +
      as.numeric(allele_2 == allele)
    dosage[is.na(codes)] <- NA_real_

    # Create a unique identifier for each locus-allele combination.
    marker_name <- paste0(
      colnames(ssr_full)[j],
      "__A",
      allele
    )

    dosage_parts_full[[marker_counter_full]] <- dosage
    allele_map_parts_full[[marker_counter_full]] <- data.frame(
      Locus = colnames(ssr_full)[j],
      Allele = allele,
      Marker = marker_name
    )
  }
}

dosage_full_raw <- do.call(cbind, dosage_parts_full)
allele_audit_full <- bind_rows(allele_map_parts_full)

rownames(dosage_full_raw) <- ids
colnames(dosage_full_raw) <- allele_audit_full$Marker
rownames(allele_audit_full) <- NULL

# Display a preview of the resulting dosage columns.
as.data.frame(dosage_full_raw[1:6, 1:8]) |>
  format_table(caption = "Preview of allele dosages")
```

The coding produces `r ncol(dosage_full_raw)` dosage columns from the `r ncol(ssr_full)` retained loci. Within each column, values 0, 1, and 2 give the number of copies of the allele identified after `__A`; `NA` denotes a missing SSR call. One locus can produce more than two columns when more than two alleles are represented in the panel.

## 4. Allele frequencies and retention of dosage columns

### 4.1 Computing allele frequencies

The frequency of an allele column is computed as the mean observed dosage divided by 2 because each diploid individual can contribute two copies of the allele. `NA` values are ignored here, so the frequency uses available genotype calls only.

### 4.2 Retention criteria for allele columns

Columns are retained when frequency lies between 0.05 and 0.95 and dosage standard deviation is positive. The lower and upper limits are symmetric: they remove very rare alleles and nearly fixed columns, while `Dosage_sd > 0` guarantees observed variation. The filter acts on allele-dosage columns; it does not remove individuals.

```{r allele-frequency-control}
# Compute dosage frequency and variability to retain informative allele columns.
allele_audit_full$Allele_frequency <-
  colMeans(dosage_full_raw, na.rm = TRUE) / 2

dosage_sd_full <- numeric(ncol(dosage_full_raw))

for (j in seq_len(ncol(dosage_full_raw))) {
  dosage_sd_full[j] <- sd(
    dosage_full_raw[, j],
    na.rm = TRUE
  )
}

allele_audit_full$Dosage_sd <- dosage_sd_full

allele_audit_full$Keep_allele <-
  is.finite(allele_audit_full$Allele_frequency) &
  allele_audit_full$Allele_frequency >= ALLELE_FREQUENCY_MIN &
  allele_audit_full$Allele_frequency <= (1 - ALLELE_FREQUENCY_MIN) &
  is.finite(allele_audit_full$Dosage_sd) &
  allele_audit_full$Dosage_sd > 0

# Select allele columns that pass the frequency and dosage-variation filters.
retained_markers_full <- allele_audit_full$Marker[
  allele_audit_full$Keep_allele
]

dosage_full <- dosage_full_raw[
  ,
  retained_markers_full,
  drop = FALSE
]

# Summarize the results in `allele_summary_full`.
allele_summary_full <- allele_audit_full |>
  group_by(Locus) |>
  summarise(
    Observed_alleles = n(),
    Retained_alleles = sum(Keep_allele),
    .groups = "drop"
  )

format_table(
  allele_summary_full,
  caption = "Allele-dosage columns by SSR locus",
  col.names = c("Locus", "Observed alleles", "Retained alleles")
)
```

Of the `r nrow(allele_audit_full)` initially coded allele columns, `r sum(allele_audit_full$Keep_allele)` satisfy both the frequency interval and positive-variation requirement. The table shows how many alleles were observed and retained at each locus.

### 4.3 Distribution of allele frequencies

The histogram shows where allele-column frequencies are concentrated. Vertical lines mark 0.05 and 0.95, and fill indicates the final retention decision, which also requires positive dosage variation. The 20 bins are only a visual-resolution choice and do not participate in filtering.

```{r allele-frequency-plot, fig.height=5, fig.alt="Histogram of SSR allele-column frequencies, with reference lines at 0.05 and 0.95 and fill indicating the final retention decision."}
# Display the allele-frequency distribution and the thresholds used by the filter.
ggplot(
  allele_audit_full,
  aes(
    x = Allele_frequency,
    fill = Keep_allele
  )
) +
  geom_histogram(
    # Use 20 bins only to standardize the visual resolution of the histogram.
    bins = 20,
    position = "identity",
    alpha = 0.7
  ) +
  geom_vline(
    # Mark the lower and upper allele-frequency thresholds in the figure.
    xintercept = c(
      ALLELE_FREQUENCY_MIN,
      1 - ALLELE_FREQUENCY_MIN
    ),
    linetype = 2
  ) +
  labs(
    x = "Allele frequency",
    y = "Allele-dosage columns",
    fill = "Retained",
    title = "Distribution of allele frequencies"
  ) +
  theme_bw()
```

Observed allele frequencies range from `r sprintf("%.3f", min(allele_audit_full$Allele_frequency, na.rm = TRUE))` to `r sprintf("%.3f", max(allele_audit_full$Allele_frequency, na.rm = TRUE))`. Bars classified as not retained fail at least one of the defined criteria; the dashed lines display the frequency limits only.

## 5. Marker matrix

### 5.1 Imputation of missing dosages

Missing dosages are imputed with the mean of the corresponding column computed from the full genotype panel. This completes the dosage matrix without creating a new genotype category and without using phenotype information.

```{r dosage-imputation}
# Compute column means and impute missing dosages.
marker_means_full <- colMeans(
  dosage_full,
  na.rm = TRUE
)

dosage_full_imputed <- dosage_full

for (j in seq_len(ncol(dosage_full_imputed))) {
  missing_values <- is.na(dosage_full_imputed[, j])

  dosage_full_imputed[
    missing_values,
    j
  ] <- marker_means_full[j]
}
```

### 5.2 Centering and construction of M

After imputation, each column is centered by subtracting the same mean used for imputation. The resulting matrix is denoted by \(\mathbf M\). Its columns are centered but are not standardized to unit variance.

Consequently, an imputed dosage becomes zero in `M`, indicating no deviation from the column mean rather than a new genotype class. Imputation and centering use genotype information from all 150 candidates; no validation phenotype participates in these operations.

```{r marker-matrix}
M_full <- sweep(
  dosage_full_imputed,
  2,
  marker_means_full,
  FUN = "-"
)
```

### 5.3 Dimensions, centering, and rank of M

The summary of `M` checks four basic properties: number of individuals, number of retained allele-dosage columns, matrix rank, and maximum absolute column mean. Rank represents the number of linearly independent directions in `M`, whereas an absolute mean close to zero confirms numerical centering.

```{r marker-matrix-summary}
data.frame(
  Rows = nrow(M_full),
  Columns = ncol(M_full),
  Rank = qr(M_full)$rank,
  Maximum_absolute_column_mean = max(abs(colMeans(M_full)))
) |>
  format_table(
    digits = 8,
    caption = "Dimensions and centering of the marker matrix",
  col.names = c("Rows", "Columns", "Rank", "Maximum absolute column mean")
  )
```

Matrix `M` contains `r nrow(M_full)` rows and `r ncol(M_full)` retained dosage columns. Its rank is `r qr(M_full)$rank`, below the number of columns because dosage columns from the same locus can be linearly dependent. The largest absolute column mean is `r format(max(abs(colMeans(M_full))), scientific = TRUE, digits = 3)`, confirming numerical centering.

## 6. Genomic relationship matrix

### 6.1 Definition and scale of G

Let \(p_j\) be the allele frequency represented by column \(j\). Using a
VanRaden-type allele-frequency scaling [@vanraden2008efficient], the additive
genomic relationship matrix is calculated as

\[
\mathbf G =
\frac{\mathbf M\mathbf M^{\mathsf T}}
{\sum_j 2p_j(1-p_j)}.
\]

Because an SSR locus can generate several allele-dosage columns, these columns should not be interpreted as independent markers. When all alleles from the same diploid locus remain represented, their dosages sum to 2 within each individual. After centering, the deviations across those columns sum to zero, creating a structural linear dependence among them.

This dependence does not prevent construction of `G`. Matrix `G` does not require inversion of `M` or separate estimation of an identifiable coefficient for every allele column. Instead, the cross-product `M M^T` aggregates, for each pair of individuals, the agreement between their allele-dosage deviation profiles. A multiallelic locus therefore contributes to genomic relationship through its set of alleles, while dependence among its columns remains an expected property of the coding.

### 6.2 Construction of G

Allele frequencies used in the scaling are obtained from dosage means because mean dosage equals \(2p_j\). For each allele column, the term \(2p_j(1-p_j)\) provides a frequency-based variability scale; summing these terms gives a global denominator across all columns. Because `M` is centered but not standardized column by column, this denominator converts the raw cross-product to a common genomic-relationship scale.

For two individuals, an off-diagonal cell of \(\mathbf M\mathbf M^{\mathsf T}\) is the inner product of their centered dosage profiles. Deviations in the same direction contribute positively, whereas deviations in opposite directions contribute negatively. After division by the denominator, off-diagonal elements of `G` describe relative genomic relationship between pairs of individuals, using allele frequencies in this panel as the reference. A negative value indicates lower genomic similarity relative to that reference rather than a biologically antagonistic relationship.

On the diagonal, each individual is compared with itself; diagonal elements therefore summarize self-relatedness on the adopted scale. Because the normalization is global and does not force each individual diagonal separately, these values need not equal exactly 1. The resulting matrix is numerically symmetrized before its eigenvalues are evaluated.

```{r genomic-relationship-matrix}
# Recover p_j from the dosage mean, which equals 2p_j.
allele_frequencies_full <- marker_means_full / 2

# Compute the VanRaden scaling from allele-column frequencies.
denominator_full <- sum(
  2 *
    allele_frequencies_full *
    (1 - allele_frequencies_full)
)

# Require a valid positive scaling constant before constructing G.
stopifnot(
  length(denominator_full) == 1L,
  is.finite(denominator_full),
  denominator_full > 0
)

# Convert cross-products of centered dosages into scaled genomic relationships.
G_full <- tcrossprod(M_full) / denominator_full

# Remove only numerical asymmetry.
G_full <- (G_full + t(G_full)) / 2

dimnames(G_full) <- list(ids, ids)

eigenvalues_G_full <- eigen(
  G_full,
  symmetric = TRUE,
  only.values = TRUE
)$values
```

### 6.3 Numerical diagnostics of G

The denominator reports the scale used to construct `G`. The mean diagonal summarizes each individual's relationship with itself on this scale and is not forced to equal exactly 1. The symmetry difference should be close to zero. Because `G` is built from a cross-product, it is theoretically positive semidefinite; therefore, a slightly negative smallest eigenvalue, when limited to numerical precision, should be interpreted as rounding rather than evidence of negative relatedness.

```{r genomic-relationship-summary}
# Build or check the matrices used for genomic prediction.
data.frame(
  Quantity = c(
    "Denominator sum 2p(1-p)",
    "Mean diagonal",
    "Maximum symmetry difference",
    "Smallest eigenvalue"
  ),
  Value = c(
    denominator_full,
    mean(diag(G_full)),
    max(abs(G_full - t(G_full))),
    min(eigenvalues_G_full)
  )
) |>
  format_table(
    digits = 8,
    caption = "Numerical summary of the genomic relationship matrix"
  )
```

Matrix `G` is numerically symmetric, with a maximum difference of `r format(max(abs(G_full - t(G_full))), scientific = TRUE, digits = 3)`. Its mean diagonal is `r sprintf("%.3f", mean(diag(G_full)))`, and its smallest eigenvalue is `r format(min(eigenvalues_G_full), scientific = TRUE, digits = 3)`. Negative values of very small magnitude are compatible with numerical rounding.

## 7. Fixed-effect matrix

Family is the only categorical fixed effect. With `~ FAM`, `model.matrix()` uses R's default treatment contrasts: it creates an intercept and enough indicator columns to represent differences among family levels relative to a reference category. Thus, `X` incorporates family structure without treating `FAM` as a numeric variable.

In Module 03, the intercept column is removed from `X` because BGLR already fits an intercept automatically; retaining both would duplicate the same constant effect.

```{r fixed-effect-matrix}
# Build the design matrix for the only categorical fixed effect used in the analysis.
X <- model.matrix(
  ~ FAM,
  data = metadata
)

rownames(X) <- ids

data.frame(
  Matrix = c("X", "M", "G"),
  Rows = c(nrow(X), nrow(M_full), nrow(G_full)),
  Columns = c(ncol(X), ncol(M_full), ncol(G_full)),
  Rank = c(qr(X)$rank, qr(M_full)$rank, qr(G_full)$rank)
) |>
  format_table(caption = "Dimensions and ranks of the analytical matrices")
```

`X`, `M`, and `G` all have `r length(ids)` rows, preserving the same individual order. `X` contains the intercept and columns required to represent `FAM`; `M` contains centered dosages; and `G` is the square relationship matrix among all individuals.

## 8. Cross-validation design

### 8.1 Evaluation scenario

Family stratification targets prediction among individuals from the represented
breeding population. It is not a leave-family-out design and therefore does
not estimate predictive performance for entirely new families.

### 8.2 Construction of the stratified folds

The 150 individuals are distributed into five validation groups of 30 individuals. Allocation proceeds family by family: individuals within each family are shuffled and distributed among folds, with priority given to folds with the lowest accumulated number of individuals. This procedure maintains both family representation and balanced total fold sizes.

After allocation, within-family balance is summarized by the difference between the largest and smallest family count across the five folds. This diagnostic is stored in `family_balance` and directly measures the quality of the stratification.

The fixed seed preserves exactly the same partition across executions. The same fold table is used for all ten traits and all methods, so differences among models cannot arise from different validation partitions.

```{r cross-validation-folds}
# Initialize the stratified partition and track each fold's accumulated total load.
set.seed(SEED_CV)

family <- droplevels(metadata$FAM)
folds <- integer(length(ids))
fold_load <- integer(N_FOLDS)

for (family_level in levels(family)) {
  # Randomize the individuals of the current family.
  family_indices <- sample(
    which(family == family_level)
  )

  # Define a random order to break ties between equally loaded folds.
  random_tie_order <- sample(seq_len(N_FOLDS))

  ordered_folds <- order(
    fold_load,
    match(seq_len(N_FOLDS), random_tie_order)
  )

  # Distribute family members across the five folds.
  family_labels <- rep(
    ordered_folds,
    length.out = length(family_indices)
  )

  folds[family_indices] <- family_labels

  fold_load <- fold_load + tabulate(
    family_labels,
    nbins = N_FOLDS
  )
}

fold_sizes <- tabulate(folds, nbins = N_FOLDS)

family_fold_table <- table(
  Family = family,
  Fold = folds
)

# Quantify, within each family, the largest difference in individual counts across folds.
family_balance <- apply(family_fold_table, 1, max) -
  apply(family_fold_table, 1, min)

cv_folds <- data.frame(
  IND = ids,
  FAM = as.character(metadata$FAM),
  Fold = folds
)

fold_summary <- data.frame(
  Fold = seq_len(N_FOLDS),
  Training_n = length(ids) - fold_sizes,
  Validation_n = fold_sizes
)

# Require a complete, balanced, one-fold-per-individual CV partition.
stopifnot(
  length(folds) == length(ids),
  !anyNA(folds),
  all(folds %in% seq_len(N_FOLDS)),
  length(fold_sizes) == N_FOLDS,
  sum(fold_sizes) == length(ids),
  length(ids) %% N_FOLDS == 0L,
  length(unique(fold_sizes)) == 1L,
  all(fold_sizes == length(ids) / N_FOLDS),
  nrow(cv_folds) == length(ids),
  identical(as.character(cv_folds$IND), ids),
  anyDuplicated(cv_folds$IND) == 0L,
  all(fold_summary$Validation_n == fold_sizes),
  all(
    fold_summary$Training_n +
      fold_summary$Validation_n ==
      length(ids)
  ),
  all(family_balance <= 1L)
)

format_table(
  fold_summary,
  col.names = c("Fold", "Training N", "Validation N"),
  caption = "Sample sizes in cross-validation"
)
```

### 8.3 Training and validation sizes

All five folds contain `r unique(fold_summary$Validation_n)` validation individuals and `r unique(fold_summary$Training_n)` training individuals. Sample size is therefore constant across the five rounds.

These partition properties are also executable contracts: every individual must receive exactly one valid fold label, all five folds must contain the expected number of candidates, and within-family counts may differ by at most one individual. Failure of any of these conditions stops the module rather than allowing an inconsistent validation design to propagate downstream.

### 8.4 Family balance across folds

The family × fold table shows the resulting stratification directly. Within each family, counts should be as balanced as its sample size allows; `family_balance`, computed previously, measures the difference between the largest and smallest count for that family. Because members are distributed cyclically, this difference is at most one individual for every family.

```{r families-by-fold}
format_table(
  as.data.frame.matrix(family_fold_table),
  caption = "Number of individuals from each family in each fold"
)
```

The largest difference between the maximum and minimum count for the same family across folds is `r max(family_balance)` individual. This confirms that stratification distributes each family as evenly as its size permits.

### 8.5 Using the folds for model fitting

The `cv_folds` table identifies which individuals are masked in each validation round. For fold `k`, the 30 individuals with `Fold == k` have their phenotypes replaced by `NA`, while the other 120 provide training phenotypes. This procedure is repeated for every trait and method.

`X`, `M`, and `G` remain fixed and include all 150 genotyped candidates, consistent with the transductive design defined at the start of the module. Consequently, every individual receives a prediction when its own phenotype is masked, producing an out-of-fold prediction downstream; this is not equivalent to predicting an external genotype population.

The table below shows a sample of the individual-to-fold assignments. The complete `cv_folds` object retains all 150 candidates and records the fold used in Module 03 to determine which phenotypes are masked in each validation round.

```{r fold-example}
cv_folds |>
  arrange(Fold, FAM, IND) |>
  head(15) |>
  format_table(caption = "Example of individual allocation to folds")
```

The 15 displayed rows make the allocation structure visible without modifying `cv_folds`, which retains all 150 individuals and is used directly in Module 03.

## 9. Genomic relationship heatmap

The heatmap provides a full view of `G`: diagonal cells represent each individual's relationship with itself, whereas off-diagonal cells represent pairwise genomic relationships. Individuals remain in their original object order; no hierarchical clustering or reordering is applied to create visual blocks. Apparent patterns can therefore suggest relationship structure, but the figure is not a clustering or population-structure test.

```{r genomic-relationship-plot, fig.height=8, fig.alt="Heatmap of the genomic relationship matrix G for the 150 individuals, displaying relative genomic similarity across pairs of candidates."}
# Reshape G to long format and build the genomic-relationship heatmap.
G_long <- as.data.frame(as.table(G_full))
names(G_long) <- c("Individual_1", "Individual_2", "Relationship")

ggplot(
  G_long,
  aes(
    x = Individual_1,
    y = Individual_2,
    fill = Relationship
  )
) +
  geom_raster() +
  scale_x_discrete(breaks = NULL) +
  scale_y_discrete(breaks = NULL) +
  labs(
    x = "Individuals",
    y = "Individuals",
    fill = "G",
    title = "Additive genomic relationship matrix"
  ) +
  theme_bw()
```

The heatmap summarizes pairwise relationships in `G`. Because individual order is preserved and no clustering is applied, areas of higher or lower intensity should be interpreted as descriptive relationship patterns rather than groups inferred by the figure.

## 10. Diagnostics of the constructed matrices

### 10.1 Structural dependence and rank of M

All allele columns satisfying the declared filters are retained; no category is artificially discarded merely to force full rank. When all alleles at a locus remain represented, their dosages sum to 2 in each diploid individual and an exact linear dependency exists among those columns. Individual allele coefficients are therefore not interpreted in isolation; later modules focus on predictions and genomic values.

In the analyzed data set, the 72 dosage columns do not represent 72 independent dimensions. When all alleles from the same locus are retained, their dosages contain a linear dependency imposed by diploidy. The structural rank bound of `M` sums the contribution \(k_l-1\) for each locus with \(k_l\) retained allele columns. The observed rank is compared with this bound to determine whether additional linear dependencies occur across loci.

### 10.2 Consolidated numerical checks

The diploid coding property is checked before allele-column filtering: for every retained SSR locus from Module 01, dosages across all alleles originally coded for an observed genotype must sum to 2. Missing SSR calls are excluded from this coding check rather than imputed for it. Centering of `M` is evaluated through its column means, and `G` is reconstructed from `M` to confirm numerical consistency. Symmetry, rank, diagonal values, and the smallest eigenvalue complete the diagnostics of the genomic relationship matrix.

```{r matrix-diagnostics}
# Check dimensions, ranks, linear dependencies, and numerical properties of X, M, and G.
structural_rank_bound_M <- sum(
  table(
    allele_audit_full$Locus[
      allele_audit_full$Keep_allele
    ]
  ) - 1L
)

maximum_M_column_mean <- max(abs(colMeans(M_full)))

G_reconstructed <- tcrossprod(M_full) / denominator_full
maximum_G_reconstruction_difference <- max(
  abs(G_full - G_reconstructed)
)

# Check the diploid coding before allele-column filtering.
# Use every allele originally coded for each retained SSR locus.
coded_locus_names <- unique(allele_audit_full$Locus)

dosage_sum_error <- numeric(length(coded_locus_names))

for (j in seq_along(coded_locus_names)) {
  locus_name <- coded_locus_names[j]

  # Select all originally coded allele columns from the current locus.
  locus_columns <- allele_audit_full$Marker[
    allele_audit_full$Locus == locus_name
  ]

  # Evaluate only individuals with an observed SSR genotype at this locus.
  observed_rows <- !is.na(ssr_full[, locus_name])

  # For an observed diploid genotype, allele dosages across the locus sum to 2.
  dosage_sum_error[j] <- max(
    abs(
      rowSums(
        dosage_full_raw[
          observed_rows,
          locus_columns,
          drop = FALSE
        ]
      ) - 2
    )
  )
}

maximum_dosage_sum_error <- max(dosage_sum_error)

matrix_inventory <- data.frame(
  Item = c(
    "Individuals",
    "Rows of X",
    "Columns of X",
    "Rank of X",
    "Rows of M",
    "Columns of M",
    "Rank of M",
    "Structural rank bound of M",
    "Maximum absolute column mean of M",
    "Maximum dosage-sum error within locus",
    "Maximum G reconstruction difference",
    "Rows of G",
    "Columns of G",
    "Rank of G",
    "Folds",
    "Training individuals per fold",
    "Validation individuals per fold",
    "Maximum symmetry difference of G",
    "Mean diagonal of G",
    "Smallest eigenvalue of G"
  ),
  Value = c(
    length(ids),
    nrow(X),
    ncol(X),
    qr(X)$rank,
    nrow(M_full),
    ncol(M_full),
    qr(M_full)$rank,
    structural_rank_bound_M,
    maximum_M_column_mean,
    maximum_dosage_sum_error,
    maximum_G_reconstruction_difference,
    nrow(G_full),
    ncol(G_full),
    qr(G_full)$rank,
    N_FOLDS,
    length(ids) - unique(fold_sizes),
    unique(fold_sizes),
    max(abs(G_full - t(G_full))),
    mean(diag(G_full)),
    min(eigenvalues_G_full)
  )
)

# DISPLAY-ONLY START
# The formatting below changes only the public table.
# `matrix_inventory` remains numeric and is the object exported to CSV.
matrix_inventory_display <- matrix_inventory

integer_inventory_items <- matrix_inventory$Item[c(1:8, 12:17)]
scientific_inventory_items <- matrix_inventory$Item[c(9:11, 18, 20)]

matrix_inventory_display$Value <- vapply(
  seq_len(nrow(matrix_inventory)),
  function(i) {
    item <- matrix_inventory$Item[i]
    value <- matrix_inventory$Value[i]

    if (item %in% integer_inventory_items) {
      sprintf("%.0f", value)
    } else if (item %in% scientific_inventory_items) {
      format(value, scientific = TRUE, digits = 3)
    } else {
      sprintf("%.3f", value)
    }
  },
  character(1)
)

format_table(
  matrix_inventory_display,
  col.names = c("Item", "Value"),
  caption = "Summary of matrices and cross-validation"
)
# DISPLAY-ONLY END
```

The final inventory collects the preceding checks in one table. For the analyzed data, `M` contains `r ncol(M_full)` columns and rank `r qr(M_full)$rank`, equal to the structural rank bound `r structural_rank_bound_M`; the pre-filter diploid dosage coding and reconstruction of `G` have maximum errors close to zero, and all five folds retain the same training and validation sizes.

## 11. Saving matrices and folds

The `matrices_object.rds` file gathers the matrices, aligned phenotypes, metadata, frequencies, folds, and parameters required by Module 03. The main diagnostic tables are also exported as CSV files for independent inspection without modifying the objects used by the models.

```{r save-matrix-objects}
# Collect matrices, metadata, frequencies, and folds in the object persisted by the module.
matrices_object <- list(
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,
  trait_dictionary = trait_dictionary,
  X = X,
  M = M_full,
  G = G_full,
  cv_folds = cv_folds,
  fold_summary = fold_summary,
  locus_audit = locus_audit_full,
  allele_audit = allele_audit_full,
  marker_means = marker_means_full,
  allele_frequencies = allele_frequencies_full,
  denominator = denominator_full,
  # Record parameters and thresholds required to reproduce this stage.
  settings = list(
    n_folds = N_FOLDS,
    seed_cv = SEED_CV,
    call_rate_min = CALL_RATE_MIN,
    allele_frequency_min = ALLELE_FREQUENCY_MIN,
    fixed_effects = "FAM"
  )
)

# Save the RDS object read directly by Module 03.
saveRDS(
  matrices_object,
  file.path(OUTPUT_DIR, "matrices_object.rds")
)

# Export tabular diagnostics as CSV files for inspection independently of the RDS object.
write.csv(cv_folds, file.path(OUTPUT_DIR, "cv_folds.csv"), row.names = FALSE)
write.csv(fold_summary, file.path(OUTPUT_DIR, "fold_summary.csv"), row.names = FALSE)
write.csv(
  allele_audit_full,
  file.path(OUTPUT_DIR, "allele_audit_full.csv"),
  row.names = FALSE
)
write.csv(
  matrix_inventory,
  file.path(OUTPUT_DIR, "matrix_inventory.csv"),
  row.names = FALSE
)
```

## 12. Next step

Module 03 uses `X`, `M`, `G`, and the five folds to fit RR-BLUP, BayesA,
BayesB, BayesB2, BayesC, Bayesian Lasso, and GBLUP. In every fit, only the
phenotypes of validation individuals are masked.

<div style="display:flex; justify-content:space-between; gap:1rem; margin-top:2rem; padding-top:1rem; border-top:1px solid #ddd;">
<a href="01_preprocessing_en.html">← Module 01 — Preprocessing</a>
<a href="03_models_en.html">Module 03 — Model fitting →</a>
</div>
