---
title: "Construction of matrices for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

In this module, the preprocessed SSR genotypes are converted into allele
dosages. These dosages are used to construct the centered marker matrix `M`,
the additive genomic relationship matrix `G`, and the fixed-effect matrix `X`,
which contains family in addition to the intercept. Five family-stratified
cross-validation folds are also defined.

The matrices are constructed once using all 150 individuals and remain
unchanged across folds. In Module 03, only the phenotype vector is masked to
form training and validation sets.

Because candidates in every fold are already genotyped, marker filtering,
allele frequencies, mean imputation, `M`, and `G` are defined from the full
genotype panel and then held fixed. No validation phenotype enters these
steps. This corresponds to a transductive candidate-genotype setting and
should not be interpreted as external validation in a new genotype population.

## 2. Packages and input object

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
  library(here)
})
```

The next step loads the input objects and organizes the structures required for the analysis.

```{r input-object}
# Define paths and load the preprocessing object used to construct the genomic matrices.
PREPROCESSING_FILE <- here(
  "output",
  "preprocessing",
  "preprocessing_object.rds"
)

OUTPUT_DIR <- here(
  "output",
  "matrices"
)

N_FOLDS <- 5L
# Fix the seed used to create family-stratified folds.
SEED_CV <- 20260723L
ALLELE_FREQUENCY_MIN <- 0.05

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

preprocessing <- readRDS(PREPROCESSING_FILE)

ids <- preprocessing$ids
metadata <- preprocessing$metadata
phenotypes <- preprocessing$phenotypes
ssr_raw <- preprocessing$ssr_raw
locus_audit_full <- preprocessing$locus_audit
trait_dictionary <- preprocessing$trait_dictionary
CALL_RATE_MIN <- preprocessing$settings$call_rate_min
```

The structure received from preprocessing is shown before marker coding.

```{r input-summary}
# Assemble the summary table presented in this section.
data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Original SSR loci",
    "Retained SSR loci",
    "Families"
  ),
  # List the values corresponding to the table entries.
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    sum(locus_audit_full$Keep_locus),
    nlevels(metadata$FAM)
  )
) |>
  kable(caption = "Input data for matrix construction")
```

## 3. Allele-dosage coding

Each SSR code contains two allele labels. For example, `12` represents one copy
of allele 1 and one copy of allele 2, whereas `22` represents two copies of
allele 2. One dosage column is created for every observed allele at each
retained locus.

```{r allele-dosage-coding}
# Organize marker information and compute the diagnostics required at this stage.
retained_loci_full <- locus_audit_full$Locus[
  locus_audit_full$Keep_locus
]

ssr_full <- ssr_raw[
  ,
  retained_loci_full,
  drop = FALSE
]

dosage_parts_full <- list()
allele_map_parts_full <- list()
marker_counter_full <- 0L

for (j in seq_len(ncol(ssr_full))) {
  # Separate the two alleles encoded in each SSR genotype.
  codes <- ssr_full[, j]
  allele_1 <- floor(codes / 10)
  allele_2 <- codes %% 10

  # List the alleles observed at the current locus.
  observed_alleles <- sort(
    unique(
      c(
        allele_1[!is.na(codes)],
        allele_2[!is.na(codes)]
      )
    )
  )

  for (allele in observed_alleles) {
    marker_counter_full <- marker_counter_full + 1L

    # Count 0, 1, or 2 copies of the allele for each individual.
    dosage <-
      as.numeric(allele_1 == allele) +
      as.numeric(allele_2 == allele)
    dosage[is.na(codes)] <- NA_real_

    # Create a unique identifier for each locus-allele combination.
    marker_name <- paste0(
      colnames(ssr_full)[j],
      "__A",
      allele
    )

    dosage_parts_full[[marker_counter_full]] <- dosage
    allele_map_parts_full[[marker_counter_full]] <- data.frame(
      Locus = colnames(ssr_full)[j],
      Allele = allele,
      Marker = marker_name
    )
  }
}

dosage_full_raw <- do.call(cbind, dosage_parts_full)
allele_audit_full <- bind_rows(allele_map_parts_full)

rownames(dosage_full_raw) <- ids
colnames(dosage_full_raw) <- allele_audit_full$Marker
rownames(allele_audit_full) <- NULL

# Example of the resulting dosage columns.
dosage_full_raw[1:6, 1:8]
```

## 4. Allele frequencies and retention of dosage columns

Very rare allele columns provide little information in a small sample. Alleles
with frequencies between 0.05 and 0.95 and nonzero dosage variation are
retained.

```{r allele-frequency-control}
# Organize marker information and compute the diagnostics required at this stage.
allele_audit_full$Allele_frequency <-
  colMeans(dosage_full_raw, na.rm = TRUE) / 2

dosage_sd_full <- numeric(ncol(dosage_full_raw))

for (j in seq_len(ncol(dosage_full_raw))) {
  dosage_sd_full[j] <- sd(
    dosage_full_raw[, j],
    na.rm = TRUE
  )
}

allele_audit_full$Dosage_sd <- dosage_sd_full

allele_audit_full$Keep_allele <-
  is.finite(allele_audit_full$Allele_frequency) &
  allele_audit_full$Allele_frequency >= ALLELE_FREQUENCY_MIN &
  allele_audit_full$Allele_frequency <= (1 - ALLELE_FREQUENCY_MIN) &
  is.finite(allele_audit_full$Dosage_sd) &
  allele_audit_full$Dosage_sd > 0

# Select allele columns that pass the frequency and dosage-variation filters.
retained_markers_full <- allele_audit_full$Marker[
  allele_audit_full$Keep_allele
]

dosage_full <- dosage_full_raw[
  ,
  retained_markers_full,
  drop = FALSE
]

# Summarize the results in `allele_summary_full`.
allele_summary_full <- allele_audit_full |>
  group_by(Locus) |>
  summarise(
    Observed_alleles = n(),
    Retained_alleles = sum(Keep_allele),
    .groups = "drop"
  )

kable(
  allele_summary_full,
  caption = "Allele-dosage columns by SSR locus"
)
```

The next step builds the corresponding visualization to support interpretation of the diagnostic.

```{r allele-frequency-plot, fig.height=5}
# Organize marker information and compute the diagnostics required at this stage.
ggplot(
  allele_audit_full,
  aes(
    x = Allele_frequency,
    fill = Keep_allele
  )
) +
  geom_histogram(
    bins = 20,
    position = "identity",
    alpha = 0.7
  ) +
  geom_vline(
    # Mark the lower and upper allele-frequency thresholds in the figure.
    xintercept = c(
      ALLELE_FREQUENCY_MIN,
      1 - ALLELE_FREQUENCY_MIN
    ),
    linetype = 2
  ) +
  labs(
    x = "Allele frequency",
    y = "Allele-dosage columns",
    fill = "Retained",
    title = "Distribution of allele frequencies"
  ) +
  theme_bw()
```

## 5. Marker matrix

Missing dosages are replaced by the mean of the corresponding column. Each
column is then centered by subtracting its mean. The resulting matrix is
denoted by \(\mathbf M\).

```{r marker-matrix}
# Organize marker information and compute the diagnostics required at this stage.
marker_means_full <- colMeans(
  dosage_full,
  na.rm = TRUE
)

dosage_full_imputed <- dosage_full

for (j in seq_len(ncol(dosage_full_imputed))) {
  # Locate missing dosages in the current column.
  missing_values <- is.na(dosage_full_imputed[, j])

  # Impute the column mean.
  dosage_full_imputed[
    missing_values,
    j
  ] <- marker_means_full[j]
}

# Center every dosage column.
M_full <- sweep(
  dosage_full_imputed,
  2,
  marker_means_full,
  FUN = "-"
)
```

Some properties of `M` can be inspected directly.

```{r marker-matrix-summary}
data.frame(
  Rows = nrow(M_full),
  Columns = ncol(M_full),
  Rank = qr(M_full)$rank,
  Maximum_absolute_column_mean = max(abs(colMeans(M_full)))
) |>
  kable(
    digits = 8,
    caption = "Dimensions and centering of the marker matrix"
  )
```

## 6. Genomic relationship matrix

Let \(p_j\) be the allele frequency represented by column \(j\). Using a
VanRaden-type allele-frequency scaling [@vanraden2008efficient], the additive
genomic relationship matrix is calculated as

\[
\mathbf G =
\frac{\mathbf M\mathbf M^{\mathsf T}}
{\sum_j 2p_j(1-p_j)}.
\]

```{r genomic-relationship-matrix}
allele_frequencies_full <- marker_means_full / 2

denominator_full <- sum(
  2 *
    allele_frequencies_full *
    (1 - allele_frequencies_full)
)

G_full <- tcrossprod(M_full) / denominator_full

# Remove only numerical asymmetry.
G_full <- (G_full + t(G_full)) / 2

dimnames(G_full) <- list(ids, ids)

eigenvalues_G_full <- eigen(
  G_full,
  symmetric = TRUE,
  only.values = TRUE
)$values
```

The next step organizes a tabular summary of the objects and results needed to interpret this section.

```{r genomic-relationship-summary}
# Build or check the matrices used for genomic prediction.
data.frame(
  Quantity = c(
    "Denominator sum 2p(1-p)",
    "Mean diagonal",
    "Maximum symmetry difference",
    "Smallest eigenvalue"
  ),
  # List the values corresponding to the table entries.
  Value = c(
    denominator_full,
    mean(diag(G_full)),
    max(abs(G_full - t(G_full))),
    min(eigenvalues_G_full)
  )
) |>
  kable(
    digits = 8,
    caption = "Numerical summary of the genomic relationship matrix"
  )
```

## 7. Fixed-effect matrix

Family is the only categorical fixed effect. `model.matrix()` includes the
intercept; in Module 03 this column is removed from `X` because BGLR already
fits an intercept automatically.

```{r fixed-effect-matrix}
X <- model.matrix(
  ~ FAM,
  data = metadata
)

rownames(X) <- ids

data.frame(
  Matrix = c("X", "M", "G"),
  Rows = c(nrow(X), nrow(M_full), nrow(G_full)),
  Columns = c(ncol(X), ncol(M_full), ncol(G_full)),
  Rank = c(qr(X)$rank, qr(M_full)$rank, qr(G_full)$rank)
) |>
  kable(caption = "Dimensions and ranks of the analytical matrices")
```

## 8. Five family-stratified folds

Family stratification targets prediction among individuals from the represented
breeding population. It is not a leave-family-out design and therefore does
not estimate predictive performance for entirely new families.

The 150 individuals are divided into five validation groups. Individuals from
each family are distributed evenly across folds, and a fixed seed preserves the
same allocation across executions.

```{r cross-validation-folds}
# Build or summarize the cross-validation fold structure.
set.seed(SEED_CV)

family <- droplevels(metadata$FAM)
folds <- integer(length(ids))
fold_load <- integer(N_FOLDS)

for (family_level in levels(family)) {
  # Randomize the individuals of the current family.
  family_indices <- sample(
    which(family == family_level)
  )

  # Define a random order to break ties between equally loaded folds.
  random_tie_order <- sample(seq_len(N_FOLDS))

  ordered_folds <- order(
    fold_load,
    match(seq_len(N_FOLDS), random_tie_order)
  )

  # Distribute family members across the five folds.
  family_labels <- rep(
    ordered_folds,
    length.out = length(family_indices)
  )

  folds[family_indices] <- family_labels

  fold_load <- fold_load + tabulate(
    family_labels,
    nbins = N_FOLDS
  )
}

fold_sizes <- tabulate(folds, nbins = N_FOLDS)

family_fold_table <- table(
  Family = family,
  Fold = folds
)

# Quantify, within each family, the largest difference in individual counts across folds.
family_balance <- apply(family_fold_table, 1, max) -
  apply(family_fold_table, 1, min)

cv_folds <- data.frame(
  IND = ids,
  FAM = as.character(metadata$FAM),
  Fold = folds
)

fold_summary <- data.frame(
  Fold = seq_len(N_FOLDS),
  Training_n = length(ids) - fold_sizes,
  Validation_n = fold_sizes
)

kable(
  fold_summary,
  col.names = c("Fold", "Training N", "Validation N"),
  caption = "Sample sizes in cross-validation"
)
```

The family × fold table shows the resulting stratification directly.

```{r families-by-fold}
kable(
  as.data.frame.matrix(family_fold_table),
  caption = "Number of individuals from each family in each fold"
)
```

The next step builds the corresponding visualization to support interpretation of the diagnostic.

```{r family-fold-plot, fig.height=7}
# Build or summarize the cross-validation fold structure.
family_fold_long <- as.data.frame(family_fold_table)

# Build the figure used for interpretation.
ggplot(
  family_fold_long,
  aes(
    x = factor(Fold),
    y = Family,
    fill = Freq
  )
) +
  geom_tile() +
  geom_text(aes(label = Freq)) +
  labs(
    x = "Fold",
    y = "Family",
    fill = "Individuals",
    title = "Distribution of families across folds"
  ) +
  theme_bw()
```

## 9. Using the folds for model fitting

The `cv_folds` table identifies which individuals will be masked in each
validation round. `X`, `M`, and `G` remain fixed. For each trait and fold,
Module 03 creates a copy of the phenotype vector and replaces the 30 validation
phenotypes with `NA`.

```{r fold-example}
cv_folds |>
  arrange(Fold, FAM, IND) |>
  head(15) |>
  kable(caption = "Example of individual allocation to folds")
```

## 10. Genomic relationship heatmap

The next step builds the corresponding visualization to support interpretation of the diagnostic.

```{r genomic-relationship-plot, fig.height=8}
# Build or check the matrices used for genomic prediction.
G_long <- as.data.frame(as.table(G_full))
names(G_long) <- c("Individual_1", "Individual_2", "Relationship")

# Build the figure used for interpretation.
ggplot(
  G_long,
  aes(
    x = Individual_1,
    y = Individual_2,
    fill = Relationship
  )
) +
  geom_raster() +
  scale_x_discrete(breaks = NULL) +
  scale_y_discrete(breaks = NULL) +
  labs(
    x = "Individuals",
    y = "Individuals",
    fill = "G",
    title = "Additive genomic relationship matrix"
  ) +
  theme_bw()
```

## 11. Diagnostics of the constructed matrices

All observed allele columns are retained to preserve a symmetric multiallelic
representation. Because dosage columns within a locus are linearly dependent,
individual allele coefficients are not interpreted in isolation; later modules
focus on predictions and genomic values.

Dependencies among allele columns from the same locus mean that the 72 dosage
columns do not represent 72 independent dimensions. The summary below shows
dimensions, ranks, and numerical properties of the matrices obtained.

```{r matrix-diagnostics}
# Build or check the matrices used for genomic prediction.
theoretical_rank_M <- sum(
  table(
    allele_audit_full$Locus[
      allele_audit_full$Keep_allele
    ]
  ) - 1L
)

maximum_M_column_mean <- max(abs(colMeans(M_full)))

G_reconstructed <- tcrossprod(M_full) / denominator_full
maximum_G_reconstruction_difference <- max(
  abs(G_full - G_reconstructed)
)

locus_names <- unique(
  allele_audit_full$Locus[
    allele_audit_full$Keep_allele
  ]
)

dosage_sum_error <- numeric(length(locus_names))

for (j in seq_along(locus_names)) {
  # Select dosage columns belonging to the same locus.
  # Identify retained dosage columns that belong to the current locus.
  locus_columns <- allele_audit_full$Marker[
    allele_audit_full$Keep_allele &
      allele_audit_full$Locus == locus_names[j]
  ]

  # In a diploid individual, dosage sums to 2 within a locus.
  dosage_sum_error[j] <- max(
    abs(
      rowSums(
        dosage_full_imputed[
          ,
          locus_columns,
          drop = FALSE
        ]
      ) - 2
    )
  )
}

maximum_dosage_sum_error <- max(dosage_sum_error)

matrix_inventory <- data.frame(
  # List the items checked in the matrix inventory.
  Item = c(
    "Individuals",
    "Rows of X",
    "Columns of X",
    "Rank of X",
    "Rows of M",
    "Columns of M",
    "Rank of M",
    "Theoretical rank of M",
    "Maximum absolute column mean of M",
    "Maximum dosage-sum error within locus",
    "Maximum G reconstruction difference",
    "Rows of G",
    "Columns of G",
    "Rank of G",
    "Folds",
    "Training individuals per fold",
    "Validation individuals per fold",
    "Maximum symmetry difference of G",
    "Mean diagonal of G",
    "Smallest eigenvalue of G"
  ),
  # List the values corresponding to the table entries.
  Value = c(
    length(ids),
    nrow(X),
    ncol(X),
    qr(X)$rank,
    nrow(M_full),
    ncol(M_full),
    qr(M_full)$rank,
    theoretical_rank_M,
    maximum_M_column_mean,
    maximum_dosage_sum_error,
    maximum_G_reconstruction_difference,
    nrow(G_full),
    ncol(G_full),
    qr(G_full)$rank,
    N_FOLDS,
    length(ids) - unique(fold_sizes),
    unique(fold_sizes),
    max(abs(G_full - t(G_full))),
    mean(diag(G_full)),
    min(eigenvalues_G_full)
  )
)

# Format the table for display in the tutorial.
kable(
  matrix_inventory,
  digits = 8,
  col.names = c("Item", "Value"),
  caption = "Summary of matrices and cross-validation"
)
```

## 12. Saving matrices and folds

The next step organizes and saves the generated outputs for use in the downstream modules.

```{r save-matrix-objects}
# Build or check the matrices used for genomic prediction.
matrices_object <- list(
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,
  trait_dictionary = trait_dictionary,
  X = X,
  M = M_full,
  G = G_full,
  cv_folds = cv_folds,
  fold_summary = fold_summary,
  locus_audit = locus_audit_full,
  allele_audit = allele_audit_full,
  marker_means = marker_means_full,
  allele_frequencies = allele_frequencies_full,
  denominator = denominator_full,
  # Record parameters and thresholds required to reproduce this stage.
  settings = list(
    n_folds = N_FOLDS,
    seed_cv = SEED_CV,
    call_rate_min = CALL_RATE_MIN,
    allele_frequency_min = ALLELE_FREQUENCY_MIN,
    fixed_effects = "FAM"
  )
)

saveRDS(
  matrices_object,
  file.path(OUTPUT_DIR, "matrices_object.rds")
)

# Save the consolidated output for downstream use.
write.csv(cv_folds, file.path(OUTPUT_DIR, "cv_folds.csv"), row.names = FALSE)
write.csv(fold_summary, file.path(OUTPUT_DIR, "fold_summary.csv"), row.names = FALSE)
write.csv(
  allele_audit_full,
  file.path(OUTPUT_DIR, "allele_audit_full.csv"),
  row.names = FALSE
)
write.csv(
  matrix_inventory,
  file.path(OUTPUT_DIR, "matrix_inventory.csv"),
  row.names = FALSE
)
```

## 13. Next step

Module 03 uses `X`, `M`, `G`, and the five folds to fit RR-BLUP, BayesA,
BayesB, BayesB2, BayesC, Bayesian Lasso, and GBLUP. In every fit, only the
phenotypes of validation individuals are masked.
