---
title: "Matrix construction for genomic selection in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

Genomic prediction models require numerical matrices that describe fixed
effects and genetic similarity among individuals.

In this step, SSR genotypes are converted into allele dosages, **family is
represented as the only categorical fixed effect**, the additive genomic
relationship matrix is constructed, and five cross-validation folds are
defined. Block does not enter the matrices or the models.

The fixed-effect matrix, marker matrix, and genomic relationship matrix are
constructed once using the complete preprocessed dataset. They remain
unchanged across folds. Cross-validation is implemented in the model-fitting
module by replacing the phenotypes of validation individuals with missing
values.

## 2. Packages and input object

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
  library(here)
})
```

```{r input-object}
PREPROCESSING_FILE <- here(
  "output",
  "preprocessing",
  "preprocessing_object.rds"
)

OUTPUT_DIR <- here(
  "output",
  "matrices"
)

N_FOLDS <- 5L
SEED_CV <- 20260723L
ALLELE_FREQUENCY_MIN <- 0.05

dir.create(
  OUTPUT_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

preprocessing <- readRDS(
  PREPROCESSING_FILE
)

ids <- preprocessing$ids
metadata <- preprocessing$metadata
phenotypes <- preprocessing$phenotypes
ssr_raw <- preprocessing$ssr_raw
locus_audit_full <- preprocessing$locus_audit
trait_dictionary <- preprocessing$trait_dictionary

CALL_RATE_MIN <- preprocessing$settings$call_rate_min

EXPECTED_N_INDIVIDUALS <- 150L
EXPECTED_N_ORIGINAL_LOCI <- 35L
EXPECTED_N_FAMILIES <- 10L
EXPECTED_FAMILY_SIZE <- 15L
EXPECTED_N_MARKERS <- 72L
EXPECTED_TRAITS <- c(
  "MMCOM", "PRODCOM", "PRODEF", "CF", "DF",
  "DCI", "EP", "FFC", "FFP", "SS"
)
NUMERICAL_TOLERANCE <- 1e-12

matrix_input_checks <- c(
  individual_count =
    length(ids) == EXPECTED_N_INDIVIDUALS,
  ids_complete =
    !any(is.na(ids) | ids == ""),
  ids_unique =
    !anyDuplicated(ids),
  metadata_ids_aligned =
    identical(as.character(metadata$IND), ids),
  phenotype_dimensions =
    identical(
      dim(phenotypes),
      c(EXPECTED_N_INDIVIDUALS, length(EXPECTED_TRAITS))
    ),
  phenotype_traits_exact =
    identical(colnames(phenotypes), EXPECTED_TRAITS),
  phenotype_ids_aligned =
    identical(rownames(phenotypes), ids),
  phenotype_values_finite =
    all(is.finite(phenotypes)),
  ssr_raw_dimensions =
    identical(
      dim(ssr_raw),
      c(EXPECTED_N_INDIVIDUALS, EXPECTED_N_ORIGINAL_LOCI)
    ),
  ssr_raw_ids_aligned =
    identical(rownames(ssr_raw), ids),
  locus_audit_count =
    nrow(locus_audit_full) == EXPECTED_N_ORIGINAL_LOCI,
  families_complete =
    !anyNA(metadata$FAM),
  family_levels_exact =
    setequal(
      unique(as.character(metadata$FAM)),
      as.character(seq_len(EXPECTED_N_FAMILIES))
    ),
  family_sizes_exact =
    length(table(metadata$FAM)) == EXPECTED_N_FAMILIES &&
    all(table(metadata$FAM) == EXPECTED_FAMILY_SIZE),
  trait_dictionary_exact =
    identical(
      as.character(trait_dictionary$Trait),
      EXPECTED_TRAITS
    )
)

if (!all(matrix_input_checks)) {
  stop(
    "Module 02 input contract failed: ",
    paste(
      names(matrix_input_checks)[
        !matrix_input_checks
      ],
      collapse = ", "
    )
  )
}

data.frame(
  Item = c(
    "Individuals",
    "Phenotype traits",
    "Raw SSR loci",
    "Retained SSR loci",
    "Families"
  ),
  Value = c(
    length(ids),
    ncol(phenotypes),
    ncol(ssr_raw),
    sum(locus_audit_full$Keep_locus),
    nlevels(metadata$FAM)
  )
) |>
  kable(
    caption = "Input data for matrix construction"
  )
```

## 3. Allele-dosage coding

Each SSR code contains two allele labels. The code `12`, for example,
represents one copy of allele 1 and one copy of allele 2. The code `22`
represents two copies of allele 2.

One column is created for every allele observed at each retained locus. The
dosage is 0, 1, or 2 according to the number of copies carried by an
individual.

```{r complete-dosage-coding}
retained_loci_full <- locus_audit_full$Locus[
  locus_audit_full$Keep_locus
]

ssr_full <- ssr_raw[
  ,
  retained_loci_full,
  drop = FALSE
]

dosage_parts_full <- list()
allele_map_parts_full <- list()
marker_counter_full <- 0L

for (j in seq_len(ncol(ssr_full))) {
  codes <- ssr_full[, j]

  allele_1 <- floor(codes / 10)
  allele_2 <- codes %% 10

  observed_alleles <- sort(
    unique(
      c(
        allele_1[!is.na(codes)],
        allele_2[!is.na(codes)]
      )
    )
  )

  for (allele in observed_alleles) {
    marker_counter_full <- marker_counter_full + 1L

    dosage <-
      as.numeric(allele_1 == allele) +
      as.numeric(allele_2 == allele)

    dosage[is.na(codes)] <- NA_real_

    marker_name <- paste0(
      colnames(ssr_full)[j],
      "__A",
      allele
    )

    dosage_parts_full[[marker_counter_full]] <- dosage

    allele_map_parts_full[[marker_counter_full]] <- data.frame(
      Locus = colnames(ssr_full)[j],
      Allele = allele,
      Marker = marker_name
    )
  }
}

dosage_full_raw <- do.call(
  cbind,
  dosage_parts_full
)

allele_audit_full <- bind_rows(
  allele_map_parts_full
)

rownames(dosage_full_raw) <- ids
colnames(dosage_full_raw) <- allele_audit_full$Marker
rownames(allele_audit_full) <- NULL

dosage_full_raw[1:6, 1:8]
```

## 4. Allele-frequency control

Very rare allele columns contribute little information and can be unstable in
small training samples. Alleles are retained when their frequency is between
0.05 and 0.95 and their dosage has nonzero variation.

```{r complete-allele-control}
allele_audit_full$Allele_frequency <-
  colMeans(
    dosage_full_raw,
    na.rm = TRUE
  ) / 2

dosage_sd_full <- numeric(
  ncol(dosage_full_raw)
)

for (j in seq_len(ncol(dosage_full_raw))) {
  dosage_sd_full[j] <- sd(
    dosage_full_raw[, j],
    na.rm = TRUE
  )
}

allele_audit_full$Dosage_sd <- dosage_sd_full

allele_audit_full$Keep_allele <-
  is.finite(allele_audit_full$Allele_frequency) &
  allele_audit_full$Allele_frequency >=
    ALLELE_FREQUENCY_MIN &
  allele_audit_full$Allele_frequency <=
    (1 - ALLELE_FREQUENCY_MIN) &
  is.finite(allele_audit_full$Dosage_sd) &
  allele_audit_full$Dosage_sd > 0

retained_markers_full <- allele_audit_full$Marker[
  allele_audit_full$Keep_allele
]

retained_marker_checks <- c(
  marker_count =
    length(retained_markers_full) == EXPECTED_N_MARKERS,
  marker_names_complete =
    !any(is.na(retained_markers_full) | retained_markers_full == ""),
  marker_names_unique =
    !anyDuplicated(retained_markers_full),
  retained_frequencies_finite =
    all(
      is.finite(
        allele_audit_full$Allele_frequency[
          allele_audit_full$Keep_allele
        ]
      )
    ),
  retained_frequencies_in_range =
    all(
      allele_audit_full$Allele_frequency[
        allele_audit_full$Keep_allele
      ] >= ALLELE_FREQUENCY_MIN &
      allele_audit_full$Allele_frequency[
        allele_audit_full$Keep_allele
      ] <= (1 - ALLELE_FREQUENCY_MIN)
    ),
  retained_dosage_sd_positive =
    all(
      is.finite(
        allele_audit_full$Dosage_sd[
          allele_audit_full$Keep_allele
        ]
      ) &
      allele_audit_full$Dosage_sd[
        allele_audit_full$Keep_allele
      ] > 0
    )
)

if (!all(retained_marker_checks)) {
  stop(
    "Retained allele-column contract failed: ",
    paste(
      names(retained_marker_checks)[
        !retained_marker_checks
      ],
      collapse = ", "
    )
  )
}

dosage_full <- dosage_full_raw[
  ,
  retained_markers_full,
  drop = FALSE
]

allele_summary_full <- allele_audit_full |>
  group_by(Locus) |>
  summarise(
    Observed_alleles = n(),
    Retained_alleles = sum(Keep_allele),
    .groups = "drop"
  )

kable(
  allele_summary_full,
  caption = "Allele-dosage columns by SSR locus"
)
```

```{r allele-frequency-plot, fig.height=5}
ggplot(
  allele_audit_full,
  aes(
    x = Allele_frequency,
    fill = Keep_allele
  )
) +
  geom_histogram(
    bins = 20,
    position = "identity",
    alpha = 0.7
  ) +
  geom_vline(
    xintercept = c(
      ALLELE_FREQUENCY_MIN,
      1 - ALLELE_FREQUENCY_MIN
    ),
    linetype = 2
  ) +
  labs(
    x = "Allele frequency",
    y = "Allele-dosage columns",
    fill = "Retained",
    title = "Allele-frequency control"
  ) +
  theme_bw()
```

## 5. Complete marker matrix

Missing allele dosages are replaced by the mean dosage of their column. The
columns are then centered by subtracting those means.

The resulting matrix \(\mathbf{M}\) has individuals in rows and retained allele
dosages in columns.

```{r complete-marker-matrix}
marker_means_full <- colMeans(
  dosage_full,
  na.rm = TRUE
)

dosage_full_imputed <- dosage_full

for (j in seq_len(ncol(dosage_full_imputed))) {
  missing_values <- is.na(
    dosage_full_imputed[, j]
  )

  dosage_full_imputed[
    missing_values,
    j
  ] <- marker_means_full[j]
}

M_full <- sweep(
  dosage_full_imputed,
  2,
  marker_means_full,
  FUN = "-"
)
```

## 6. Complete genomic relationship matrix

Let \(p_j\) be the frequency of the allele represented by column \(j\).
Using a VanRaden-type allele-frequency scaling [@vanraden2008efficient],
the additive genomic relationship matrix is calculated as

\[
\mathbf{G} =
\frac{\mathbf{M}\mathbf{M}^{\mathsf{T}}}
{\sum_j 2p_j(1-p_j)}.
\]

```{r complete-genomic-matrix}
allele_frequencies_full <-
  marker_means_full / 2

denominator_full <- sum(
  2 *
    allele_frequencies_full *
    (1 - allele_frequencies_full)
)

if (
  length(denominator_full) != 1L ||
    !is.finite(denominator_full) ||
    denominator_full <= 0
) {
  stop("The denominator of G must be finite and positive.")
}

G_full <- tcrossprod(
  M_full
) / denominator_full

G_full <- (
  G_full + t(G_full)
) / 2

dimnames(G_full) <- list(
  ids,
  ids
)

eigenvalues_G_full <- eigen(
  G_full,
  symmetric = TRUE,
  only.values = TRUE
)$values
```

## 7. Fixed-effect matrix

Family is represented as the only categorical fixed effect. The intercept is
included automatically by `model.matrix()` and will be handled by BGLR in the
next module.

```{r fixed-effect-matrix}
X <- model.matrix(
  ~ FAM,
  data = metadata
)

rownames(X) <- ids

data.frame(
  Matrix = c("X", "M", "G"),
  Rows = c(
    nrow(X),
    nrow(M_full),
    nrow(G_full)
  ),
  Columns = c(
    ncol(X),
    ncol(M_full),
    ncol(G_full)
  ),
  Rank = c(
    qr(X)$rank,
    qr(M_full)$rank,
    qr(G_full)$rank
  )
) |>
  kable(
    caption = "Dimensions and ranks of the complete matrices"
  )
```

## 8. Five family-stratified folds

The 150 individuals are divided into five validation groups. Individuals from
each family are distributed as evenly as possible among the folds. The fixed
seed preserves the same allocation in every execution.

```{r cross-validation-folds}
set.seed(
  SEED_CV
)

family <- droplevels(
  metadata$FAM
)

folds <- integer(
  length(ids)
)

fold_load <- integer(
  N_FOLDS
)

for (family_level in levels(family)) {
  family_indices <- sample(
    which(family == family_level)
  )

  random_tie_order <- sample(
    seq_len(N_FOLDS)
  )

  ordered_folds <- order(
    fold_load,
    match(
      seq_len(N_FOLDS),
      random_tie_order
    )
  )

  family_labels <- rep(
    ordered_folds,
    length.out = length(family_indices)
  )

  folds[family_indices] <-
    family_labels

  fold_load <- fold_load + tabulate(
    family_labels,
    nbins = N_FOLDS
  )
}

fold_sizes <- tabulate(
  folds,
  nbins = N_FOLDS
)

family_fold_table <- table(
  Family = family,
  Fold = folds
)

family_balance <- apply(
  family_fold_table,
  1,
  max
) - apply(
  family_fold_table,
  1,
  min
)

cv_folds <- data.frame(
  IND = ids,
  FAM = as.character(metadata$FAM),
  Fold = folds
)

data.frame(
  Fold = seq_len(N_FOLDS),
  Validation_n = fold_sizes,
  Training_n = length(ids) - fold_sizes
) |>
  kable(
    caption = "Size of the cross-validation folds"
  )
```

```{r family-fold-plot, fig.height=7}
family_fold_long <- as.data.frame(
  family_fold_table
)

ggplot(
  family_fold_long,
  aes(
    x = factor(Fold),
    y = Family,
    fill = Freq
  )
) +
  geom_tile() +
  geom_text(
    aes(label = Freq)
  ) +
  labs(
    x = "Fold",
    y = "Family",
    fill = "Individuals",
    title = "Family distribution across folds"
  ) +
  theme_bw()
```

## 9. Cross-validation contract

The fold table defines which 30 individuals are used for validation in each
round. It does not define fold-specific marker or genomic matrices.

The matrices used by all models remain fixed:

- `X` contains the intercept and family; in Module 03 the intercept is removed
  from `X` because BGLR already fits it by default, leaving family only in the
  `FIXED` component;
- `M` contains the centered allele-dosage columns;
- `G` contains the additive genomic relationships.

In the next module, a copy of the phenotype vector is created for each trait
and fold. The 30 validation phenotypes are replaced by `NA`, while `X`, `M`,
and `G` remain unchanged.

```{r cross-validation-summary}
fold_summary <- data.frame(
  Fold = seq_len(N_FOLDS),
  Training_n = length(ids) - fold_sizes,
  Validation_n = fold_sizes
)

kable(
  fold_summary,
  caption = "Cross-validation sample sizes"
)
```

## 10. Genomic relationship heatmap

The heatmap represents genomic similarity among the 150 individuals.

```{r genomic-relationship-plot, fig.height=8}
G_long <- as.data.frame(
  as.table(G_full)
)

names(G_long) <- c(
  "Individual_1",
  "Individual_2",
  "Relationship"
)

ggplot(
  G_long,
  aes(
    x = Individual_1,
    y = Individual_2,
    fill = Relationship
  )
) +
  geom_raster() +
  scale_x_discrete(
    breaks = NULL
  ) +
  scale_y_discrete(
    breaks = NULL
  ) +
  labs(
    x = "Individuals",
    y = "Individuals",
    fill = "G",
    title = "Additive genomic relationship matrix"
  ) +
  theme_bw()
```

## 11. Saved objects and matrix diagnostics

The output object contains a family-only fixed-effect matrix, a marker matrix,
a genomic relationship matrix, and the fold-assignment table. No fold-specific
matrix object is created.

```{r save-matrix-objects}
theoretical_rank_M <-
  sum(
    table(
      allele_audit_full$Locus[
        allele_audit_full$Keep_allele
      ]
    ) - 1L
  )

maximum_M_column_mean <-
  max(
    abs(
      colMeans(M_full)
    )
  )

G_reconstructed <-
  tcrossprod(M_full) /
  denominator_full

maximum_G_reconstruction_difference <-
  max(
    abs(
      G_full -
        G_reconstructed
    )
  )

locus_names <- unique(
  allele_audit_full$Locus[
    allele_audit_full$Keep_allele
  ]
)

dosage_sum_error <- numeric(
  length(locus_names)
)

for (j in seq_along(locus_names)) {
  locus_columns <-
    allele_audit_full$Marker[
      allele_audit_full$Keep_allele &
        allele_audit_full$Locus ==
          locus_names[j]
    ]

  dosage_sum_error[j] <-
    max(
      abs(
        rowSums(
          dosage_full_imputed[
            ,
            locus_columns,
            drop = FALSE
          ]
        ) - 2
      )
    )
}

maximum_dosage_sum_error <-
  max(dosage_sum_error)

matrix_inventory <- data.frame(
  Item = c(
    "Individuals",
    "Fixed-effect matrix rows",
    "Fixed-effect matrix columns",
    "Fixed-effect matrix rank",
    "Marker-matrix rows",
    "Marker-matrix columns",
    "Marker-matrix rank",
    "Theoretical marker-matrix rank",
    "Maximum absolute column mean of M",
    "Maximum dosage-sum error by locus",
    "Maximum reconstruction difference for G",
    "Genomic-matrix rows",
    "Genomic-matrix columns",
    "Genomic-matrix rank",
    "Cross-validation folds",
    "Training individuals per fold",
    "Validation individuals per fold",
    "Maximum G symmetry difference",
    "Mean diagonal of G",
    "Minimum eigenvalue of G"
  ),
  Value = c(
    length(ids),
    nrow(X),
    ncol(X),
    qr(X)$rank,
    nrow(M_full),
    ncol(M_full),
    qr(M_full)$rank,
    theoretical_rank_M,
    maximum_M_column_mean,
    maximum_dosage_sum_error,
    maximum_G_reconstruction_difference,
    nrow(G_full),
    ncol(G_full),
    qr(G_full)$rank,
    N_FOLDS,
    length(ids) - unique(fold_sizes),
    unique(fold_sizes),
    max(abs(G_full - t(G_full))),
    mean(diag(G_full)),
    min(eigenvalues_G_full)
  )
)

matrix_integrity_checks <- c(
  X_dimensions =
    identical(
      dim(X),
      c(EXPECTED_N_INDIVIDUALS, EXPECTED_N_FAMILIES)
    ),
  X_columns_exact =
    identical(
      colnames(X),
      c("(Intercept)", paste0("FAM", 2:EXPECTED_N_FAMILIES))
    ),
  X_ids_aligned =
    identical(rownames(X), ids),
  X_finite =
    all(is.finite(X)),
  M_dimensions =
    identical(
      dim(M_full),
      c(EXPECTED_N_INDIVIDUALS, EXPECTED_N_MARKERS)
    ),
  M_markers_exact =
    identical(colnames(M_full), retained_markers_full),
  M_ids_aligned =
    identical(rownames(M_full), ids),
  M_finite =
    all(is.finite(M_full)),
  M_centered =
    maximum_M_column_mean <= NUMERICAL_TOLERANCE,
  marker_means_finite =
    all(is.finite(marker_means_full)),
  allele_frequencies_finite =
    all(is.finite(allele_frequencies_full)),
  allele_frequencies_valid =
    all(
      allele_frequencies_full >= 0 &
      allele_frequencies_full <= 1
    ),
  denominator_positive =
    is.finite(denominator_full) && denominator_full > 0,
  G_dimensions =
    identical(
      dim(G_full),
      c(EXPECTED_N_INDIVIDUALS, EXPECTED_N_INDIVIDUALS)
    ),
  G_ids_aligned =
    identical(rownames(G_full), ids) &&
    identical(colnames(G_full), ids),
  G_finite =
    all(is.finite(G_full)),
  G_symmetric =
    max(abs(G_full - t(G_full))) <= NUMERICAL_TOLERANCE,
  G_reconstruction_exact =
    maximum_G_reconstruction_difference <= NUMERICAL_TOLERANCE,
  dosage_sum_diagnostic_finite =
    is.finite(maximum_dosage_sum_error),
  fold_labels_exact =
    setequal(unique(cv_folds$Fold), seq_len(N_FOLDS)),
  fold_ids_aligned =
    identical(as.character(cv_folds$IND), ids),
  fold_families_aligned =
    identical(
      as.character(cv_folds$FAM),
      as.character(metadata$FAM)
    ),
  fold_sizes_exact =
    length(fold_sizes) == N_FOLDS &&
    all(fold_sizes == 30L),
  training_sizes_exact =
    all(fold_summary$Training_n == 120L),
  validation_sizes_exact =
    all(fold_summary$Validation_n == 30L),
  family_fold_balance_exact =
    identical(
      dim(family_fold_table),
      c(EXPECTED_N_FAMILIES, N_FOLDS)
    ) &&
    all(family_fold_table == 3L) &&
    all(family_balance == 0L)
)

if (!all(matrix_integrity_checks)) {
  stop(
    "Final matrix integrity gate failed: ",
    paste(
      names(matrix_integrity_checks)[
        !matrix_integrity_checks
      ],
      collapse = ", "
    )
  )
}

matrices_object <- list(
  ids = ids,
  metadata = metadata,
  phenotypes = phenotypes,
  trait_dictionary = trait_dictionary,
  X = X,
  M = M_full,
  G = G_full,
  cv_folds = cv_folds,
  fold_summary = fold_summary,
  locus_audit = locus_audit_full,
  allele_audit = allele_audit_full,
  marker_means = marker_means_full,
  allele_frequencies = allele_frequencies_full,
  denominator = denominator_full,
  settings = list(
    n_folds = N_FOLDS,
    seed_cv = SEED_CV,
    call_rate_min = CALL_RATE_MIN,
    allele_frequency_min =
      ALLELE_FREQUENCY_MIN,
    fixed_effects = "FAM"
  )
)

saveRDS(
  matrices_object,
  file.path(
    OUTPUT_DIR,
    "matrices_object.rds"
  )
)

write.csv(
  cv_folds,
  file.path(
    OUTPUT_DIR,
    "cv_folds.csv"
  ),
  row.names = FALSE
)

write.csv(
  fold_summary,
  file.path(
    OUTPUT_DIR,
    "fold_summary.csv"
  ),
  row.names = FALSE
)

write.csv(
  allele_audit_full,
  file.path(
    OUTPUT_DIR,
    "allele_audit_full.csv"
  ),
  row.names = FALSE
)

write.csv(
  matrix_inventory,
  file.path(
    OUTPUT_DIR,
    "matrix_inventory.csv"
  ),
  row.names = FALSE
)

kable(
  matrix_inventory,
  digits = 6,
  caption = "Final diagnostics of the analytical matrices"
)
```

The retained 72 allele-dosage columns do not represent 72 independent marker
dimensions. Within a multiallelic SSR locus, allele dosages are linearly
dependent only up to the diploid dosage constraint. Therefore, the marker
matrix can have a rank smaller than its number of columns. Since `G` is built
from `M`, its rank is also bounded by the rank of the marker matrix.

## 12. Next step

The next stage fits RR-BLUP, BayesA, BayesB, BayesB with the specified
near-zero probability of a zero effect, BayesC, Bayesian Lasso, and GBLUP using
the same five validation folds. RR-BLUP is implemented with the BGLR `BRR`
component and `M`, whereas GBLUP is implemented with `G` and the `RKHS`
component. In every fold, only the phenotype vector is masked; family is the
only categorical fixed effect, while `M` and `G` remain unchanged.
