---
title: "Pilot fitting of seven genomic prediction methods in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

This module first validates the model-fitting contract with one phenotype and
one cross-validation fold. The pilot uses `MMCOM` and fold 1.

Seven genomic prediction methods are fitted exclusively with BGLR
[@perez2014bglr]:

1. RR-BLUP, implemented with the BGLR `BRR` component and the complete marker
   matrix `M`;
2. BayesA with `M`;
3. BayesB with `M`;
4. BayesB2 with `M` and the zero-effect probability fixed operationally near
   \(10^{-5}\);
5. BayesC with `M`;
6. Bayesian Lasso with `M`;
7. GBLUP with the genomic relationship matrix `G` and the BGLR `RKHS`
   component.

RR-BLUP and Bayesian ridge regression are not fitted as separate models. The
Gaussian marker-effect model implemented by `model = "BRR"` is adopted as the
RR-BLUP implementation.

The pilot checks code execution, phenotype masking, output structure,
prediction finiteness, predictive metrics, and the BayesB2 probability
parameter. It is not used for biological conclusions.

## 2. BGLR contract

BGLR includes an intercept by default and accepts missing values in the
phenotype vector. Therefore, the intercept column is removed from `X` before
the remaining block and family columns are passed as fixed effects. For each
fold, only the validation phenotypes are replaced by `NA`; `X`, `M`, and `G`
remain unchanged.

The reference article defines BayesB2 with \(\pi=10^{-5}\) as the probability
of a zero marker effect [@silva2021brrguava]. In BGLR, `probIn` is the
probability of a non-zero effect. Thus, BayesB2 uses
`probIn = 1 - 1e-5`. Because BGLR samples this probability, an extremely
concentrated beta prior (`counts = 1e6`) is used to keep it operationally fixed
at the reference value.

## 3. Packages and pilot settings

```{r packages}
suppressPackageStartupMessages({
  library(BGLR)
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(here)
})
```

```{r pilot-settings}
MATRICES_FILE <- here(
  "output",
  "matrices",
  "matrices_object.rds"
)

OUTPUT_DIR <- here(
  "output",
  "models",
  "pilot"
)

PILOT_TRAIT <- "MMCOM"
PILOT_FOLD <- 1L

PILOT_N_ITER <- 12000L
PILOT_BURN_IN <- 2000L
PILOT_THIN <- 10L

BAYESB2_ZERO_PROBABILITY <- 1e-5
BAYESB2_PROB_IN <-
  1 - BAYESB2_ZERO_PROBABILITY
BAYESB2_COUNTS <- 1e6

dir.create(
  OUTPUT_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)
```

The pilot uses fewer iterations than the final analysis. After the model
contract is approved, the complete analysis will use the reference settings:
1,000,000 iterations, 200,000 burn-in iterations, and thinning equal to 4
[@silva2021brrguava].

## 4. Input matrices and phenotype masking

```{r input-object}
matrices_object <- readRDS(
  MATRICES_FILE
)

ids <- matrices_object$ids
X <- matrices_object$X
M <- matrices_object$M
G <- matrices_object$G
phenotypes <- matrices_object$phenotypes
cv_folds <- matrices_object$cv_folds

X_fixed <- X[
  ,
  colnames(X) != "(Intercept)",
  drop = FALSE
]

y_observed <- phenotypes[
  ,
  PILOT_TRAIT
]

validation_index <- which(
  cv_folds$Fold == PILOT_FOLD
)

training_index <- which(
  cv_folds$Fold != PILOT_FOLD
)

y_model <- y_observed
y_model[validation_index] <- NA_real_
```

```{r input-audit}
input_audit <- data.frame(
  Check = c(
    "Individuals",
    "Fixed-effect columns supplied to BGLR",
    "Marker columns",
    "Rows and columns of G",
    "Observed training phenotypes",
    "Masked validation phenotypes",
    "Validation individuals",
    "Training individuals",
    "X row names aligned with IDs",
    "M row names aligned with IDs",
    "G row names aligned with IDs",
    "G column names aligned with IDs"
  ),
  Value = c(
    length(ids),
    ncol(X_fixed),
    ncol(M),
    paste(dim(G), collapse = " x "),
    sum(!is.na(y_model)),
    sum(is.na(y_model)),
    length(validation_index),
    length(training_index),
    identical(rownames(X), ids),
    identical(rownames(M), ids),
    identical(rownames(G), ids),
    identical(colnames(G), ids)
  )
)

kable(
  input_audit,
  caption = "Pilot input audit"
)
```

## 5. Model catalogue

```{r model-catalogue}
model_catalogue <- data.frame(
  Method = c(
    "RR-BLUP",
    "BayesA",
    "BayesB",
    "BayesB2",
    "BayesC",
    "Bayesian Lasso",
    "GBLUP"
  ),
  BGLR_component = c(
    "BRR",
    "BayesA",
    "BayesB",
    "BayesB",
    "BayesC",
    "BL",
    "RKHS"
  ),
  Genomic_input = c(
    "M",
    "M",
    "M",
    "M",
    "M",
    "M",
    "G"
  ),
  Main_assumption = c(
    "Common Gaussian variance for all marker effects",
    "Marker-specific scaled-t variances",
    "Scaled-t slab and a point mass at zero",
    "BayesB with zero-effect probability near 1e-5",
    "Common Gaussian slab and a point mass at zero",
    "Double-exponential marginal prior",
    "Genomic values structured by G"
  )
)

kable(
  model_catalogue,
  caption = "Seven genomic prediction methods in the pilot"
)
```

## 6. ETA specifications

```{r eta-specifications}
ETA_models <- list(
  RRBLUP = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BRR"
    )
  ),
  BayesA = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BayesA"
    )
  ),
  BayesB = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BayesB"
    )
  ),
  BayesB2 = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BayesB",
      probIn = BAYESB2_PROB_IN,
      counts = BAYESB2_COUNTS
    )
  ),
  BayesC = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BayesC"
    )
  ),
  BL = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BL"
    )
  ),
  GBLUP = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      K = G,
      model = "RKHS"
    )
  )
)
```

## 7. Pilot fitting

Each method receives a deterministic seed and a distinct file prefix. The
validation observations remain absent from `y_model` during fitting.

```{r pilot-fitting}
model_names <- names(
  ETA_models
)

method_seeds <- setNames(
  202608051L + seq_along(model_names) - 1L,
  model_names
)

pilot_fits <- vector(
  mode = "list",
  length = length(model_names)
)

names(pilot_fits) <- model_names

prediction_parts <- vector(
  mode = "list",
  length = length(model_names)
)

metric_parts <- vector(
  mode = "list",
  length = length(model_names)
)

for (method_index in seq_along(model_names)) {
  method_name <- model_names[method_index]

  set.seed(
    method_seeds[method_name]
  )

  fit_start <- proc.time()[["elapsed"]]

  pilot_fits[[method_name]] <- BGLR::BGLR(
    y = y_model,
    ETA = ETA_models[[method_name]],
    response_type = "gaussian",
    nIter = PILOT_N_ITER,
    burnIn = PILOT_BURN_IN,
    thin = PILOT_THIN,
    saveAt = file.path(
      OUTPUT_DIR,
      paste0(
        "pilot_",
        method_name,
        "_"
      )
    ),
    verbose = FALSE,
    rmExistingFiles = TRUE
  )

  fit_end <- proc.time()[["elapsed"]]

  predicted_validation <-
    pilot_fits[[method_name]]$yHat[
      validation_index
    ]

  observed_validation <-
    y_observed[
      validation_index
    ]

  prediction_parts[[method_index]] <- data.frame(
    IND = ids[validation_index],
    Trait = PILOT_TRAIT,
    Fold = PILOT_FOLD,
    Method = method_name,
    Observed = observed_validation,
    Predicted = predicted_validation
  )

  metric_parts[[method_index]] <- data.frame(
    Trait = PILOT_TRAIT,
    Fold = PILOT_FOLD,
    Method = method_name,
    Validation_n = length(validation_index),
    Correlation = cor(
      observed_validation,
      predicted_validation
    ),
    RMSE = sqrt(
      mean(
        (
          observed_validation -
            predicted_validation
        )^2
      )
    ),
    MAE = mean(
      abs(
        observed_validation -
          predicted_validation
      )
    ),
    DIC =
      pilot_fits[[method_name]]$fit$DIC,
    Effective_parameters =
      pilot_fits[[method_name]]$fit$pD,
    Log_likelihood_at_posterior_mean =
      pilot_fits[[method_name]]$fit$
        logLikAtPostMean,
    Residual_variance =
      pilot_fits[[method_name]]$varE,
    Elapsed_seconds =
      fit_end - fit_start,
    Finite_predictions = all(
      is.finite(
        predicted_validation
      )
    )
  )
}
```

## 8. Consolidated pilot outputs

```{r consolidate-pilot}
pilot_predictions <- do.call(
  rbind,
  prediction_parts
)

pilot_metrics <- do.call(
  rbind,
  metric_parts
)

rownames(pilot_predictions) <- NULL
rownames(pilot_metrics) <- NULL

bayesb_probability_audit <- data.frame(
  Method = c(
    "BayesB",
    "BayesB2",
    "BayesC"
  ),
  Prior_nonzero_probability = c(
    NA_real_,
    BAYESB2_PROB_IN,
    NA_real_
  ),
  Posterior_nonzero_probability = c(
    pilot_fits[["BayesB"]]$ETA$
      genomic$probIn,
    pilot_fits[["BayesB2"]]$ETA$
      genomic$probIn,
    pilot_fits[["BayesC"]]$ETA$
      genomic$probIn
  )
)

rrblup_predictions <- pilot_predictions$Predicted[
  pilot_predictions$Method == "RRBLUP"
]

gblup_predictions <- pilot_predictions$Predicted[
  pilot_predictions$Method == "GBLUP"
]

rrblup_gblup_audit <- data.frame(
  Comparison = "RR-BLUP versus GBLUP",
  Prediction_correlation = cor(
    rrblup_predictions,
    gblup_predictions
  ),
  Maximum_absolute_difference = max(
    abs(
      rrblup_predictions -
        gblup_predictions
    )
  ),
  Mean_absolute_difference = mean(
    abs(
      rrblup_predictions -
        gblup_predictions
    )
  )
)

pilot_audit <- data.frame(
  Check = c(
    "Methods fitted",
    "Validation individuals per method",
    "Total prediction rows",
    "Finite predictions",
    "Complete metric rows",
    "BayesB2 posterior probIn",
    "Absolute BayesB2 probIn deviation"
  ),
  Expected = c(
    "7",
    "30",
    "210",
    "TRUE",
    "7",
    format(BAYESB2_PROB_IN, digits = 8),
    "close to 0"
  ),
  Observed = c(
    length(model_names),
    unique(pilot_metrics$Validation_n),
    nrow(pilot_predictions),
    all(pilot_metrics$Finite_predictions),
    nrow(pilot_metrics),
    format(
      pilot_fits[["BayesB2"]]$ETA$
        genomic$probIn,
      digits = 8
    ),
    format(
      abs(
        pilot_fits[["BayesB2"]]$ETA$
          genomic$probIn -
          BAYESB2_PROB_IN
      ),
      scientific = TRUE
    )
  )
)
```

```{r save-pilot-outputs}
write.csv(
  pilot_predictions,
  file.path(
    OUTPUT_DIR,
    "pilot_predictions.csv"
  ),
  row.names = FALSE
)

write.csv(
  pilot_metrics,
  file.path(
    OUTPUT_DIR,
    "pilot_metrics.csv"
  ),
  row.names = FALSE
)

write.csv(
  bayesb_probability_audit,
  file.path(
    OUTPUT_DIR,
    "pilot_bayesb_probability_audit.csv"
  ),
  row.names = FALSE
)

write.csv(
  rrblup_gblup_audit,
  file.path(
    OUTPUT_DIR,
    "pilot_rrblup_gblup_audit.csv"
  ),
  row.names = FALSE
)

write.csv(
  pilot_audit,
  file.path(
    OUTPUT_DIR,
    "pilot_execution_audit.csv"
  ),
  row.names = FALSE
)

saveRDS(
  list(
    fits = pilot_fits,
    predictions = pilot_predictions,
    metrics = pilot_metrics,
    bayesb_probability_audit =
      bayesb_probability_audit,
    rrblup_gblup_audit =
      rrblup_gblup_audit,
    execution_audit = pilot_audit,
    settings = list(
      trait = PILOT_TRAIT,
      fold = PILOT_FOLD,
      n_iter = PILOT_N_ITER,
      burn_in = PILOT_BURN_IN,
      thin = PILOT_THIN,
      bayesb2_zero_probability =
        BAYESB2_ZERO_PROBABILITY,
      bayesb2_prob_in =
        BAYESB2_PROB_IN,
      bayesb2_counts =
        BAYESB2_COUNTS,
      seeds = method_seeds
    )
  ),
  file.path(
    OUTPUT_DIR,
    "pilot_models_object.rds"
  )
)
```

## 9. Pilot results

```{r pilot-metric-table}
pilot_metrics |>
  arrange(
    desc(Correlation),
    RMSE
  ) |>
  kable(
    digits = 6,
    caption = "Predictive metrics and fitting diagnostics for the pilot"
  )
```

```{r bayesb-probability-table}
kable(
  bayesb_probability_audit,
  digits = 8,
  caption = "Non-zero-effect probability audit"
)
```

```{r rrblup-gblup-table}
kable(
  rrblup_gblup_audit,
  digits = 8,
  caption = "Diagnostic comparison of RR-BLUP and GBLUP predictions"
)
```

```{r pilot-execution-table}
kable(
  pilot_audit,
  caption = "Pilot execution audit"
)
```

```{r pilot-prediction-plot, fig.height=8}
ggplot(
  pilot_predictions,
  aes(
    x = Observed,
    y = Predicted
  )
) +
  geom_point() +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = 2
  ) +
  facet_wrap(
    ~ Method,
    scales = "free"
  ) +
  labs(
    x = "Observed phenotype",
    y = "Predicted phenotype",
    title = paste(
      "Pilot predictions:",
      PILOT_TRAIT,
      "- fold",
      PILOT_FOLD
    )
  ) +
  theme_bw()
```

## 10. Decision gate

The pilot is approved for expansion only when:

- all seven methods complete without interruption;
- each method returns 30 validation predictions;
- all 210 predictions are finite;
- correlation, RMSE, MAE, DIC, and runtime are available for every method;
- the BayesB2 posterior `probIn` remains extremely close to `0.99999`;
- the generated CSV and RDS files are internally consistent.

After this gate is approved, the same contract will be expanded to 10 traits,
5 folds, and 7 methods, totaling 350 fits. The complete analysis will preserve
the same global `X`, `M`, and `G` matrices and will change only the masked
phenotype vector.
