---
title: "Model fitting and MCMC diagnostics for genomic prediction in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

This module performs the complete cross-validated fitting of seven genomic
prediction methods for ten papaya traits. The same five deterministic,
family-stratified folds are used for every method and trait.

The complete analysis contains:

\[
10\ \text{traits}
\times
5\ \text{folds}
\times
7\ \text{methods}
=
350\ \text{fits}.
\]

All methods are fitted exclusively with BGLR [@perez2014bglr]:

1. RR-BLUP, implemented with the BGLR `BRR` component and the complete marker
   matrix `M`;
2. BayesA with `M`;
3. BayesB with `M`;
4. BayesB2 with `M` and a zero-effect probability operationally fixed near
   \(10^{-5}\);
5. BayesC with `M`;
6. Bayesian Lasso with `M`;
7. GBLUP with the genomic relationship matrix `G` and the BGLR `RKHS`
   component.

RR-BLUP and Bayesian ridge regression are not treated as separate methods. The
Gaussian marker-effect model implemented by `model = "BRR"` is the RR-BLUP
implementation used in this project.

## 2. Analytical contract

The matrices constructed in Module 02 are used without fold-specific
reconstruction:

- `X` contains the intercept and the family fixed-effect coding;
- `M` contains the 72 centered allele-dosage columns;
- `G` contains the additive genomic relationships;
- each fold identifies 30 validation individuals and 120 training individuals.

BGLR includes an intercept by default. Therefore, the intercept column is
removed from `X`, leaving only the family columns to be passed through the
`FIXED` component.

For each trait and fold, only a copy of the phenotype vector is changed. The
30 validation observations are replaced by `NA`; `X`, `M`, and `G` remain
global and unchanged.

The MCMC settings follow the main methods section of the reference article
[@silva2021brrguava]:

- 1,000,000 iterations;
- 200,000 burn-in iterations;
- thinning equal to 4;
- 200,000 retained posterior samples.

The BGLR default prior allocation parameter is made explicit with `R2 = 0.5`.

## 3. Reproducible and resumable execution

Each trait-fold-method combination has:

- a deterministic seed;
- a unique run identifier;
- an individual compact RDS result;
- a persistent directory containing the native files written by BGLR;
- a row in the run manifest.

The compact RDS is the checkpoint that determines whether a statistical fit is
missing during an ordinary rebuild. Completed checkpoints are skipped, so an
ordinary rebuild does not refit completed models.

Native BGLR files are preserved under
`output/models/bglr/Trait/Fold/Method/`. They contain the thinned chains written
by BGLR and are used directly for posterior diagnostics and final graphical
summaries. The compact RDS, rather than the native chain files, remains the
criterion used to resume the 350-fit analysis.

## 4. Packages and complete-analysis settings

```{r packages}
suppressPackageStartupMessages({
  library(BGLR)
  library(coda)
  library(digest)
  library(dplyr)
  library(knitr)
  library(here)
})
```

```{r complete-settings}
MATRICES_FILE <- here(
  "output",
  "matrices",
  "matrices_object.rds"
)

OUTPUT_DIR <- here(
  "output",
  "models"
)

RUNS_DIR <- file.path(
  OUTPUT_DIR,
  "runs"
)

BGLR_RUNS_DIR <- file.path(
  OUTPUT_DIR,
  "bglr"
)

N_ITER <- 1000000L
BURN_IN <- 200000L
THIN <- 4L
R2_PRIOR <- 0.5

RAW_SAVED_SAMPLES <- floor(
  N_ITER / THIN
)

BURN_IN_SAVED_SAMPLES <- floor(
  BURN_IN / THIN
)

RETAINED_SAMPLES <-
  RAW_SAVED_SAMPLES -
  BURN_IN_SAVED_SAMPLES

PIPELINE_VERSION <-
  "module03-full-v3"

MAX_NEW_RUNS_PER_BUILD <- as.integer(
  Sys.getenv(
    "PAPAYA_MAX_NEW_RUNS",
    unset = "350"
  )
)

SEED_BASE <- 202608050L

BAYESB2_ZERO_PROBABILITY <- 1e-5
BAYESB2_PROB_IN <-
  1 - BAYESB2_ZERO_PROBABILITY
BAYESB2_COUNTS <- 1e6

ESS_REVIEW_THRESHOLD <- 1000
GEWEKE_REVIEW_THRESHOLD <- 2

dir.create(
  RUNS_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  BGLR_RUNS_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)
```

### Meaning of the fitting settings

The settings above control MCMC sampling and prior calibration; they are not
model-selection criteria.

- `N_ITER` is the total number of MCMC iterations.
- `BURN_IN` defines the initial iterations discarded before posterior summaries.
- `THIN` is the interval between saved samples. With the adopted settings,
  200,000 posterior samples remain for each fit.
- `R2 = 0.5` is used by BGLR to calibrate prior hyperparameters when they are
  not specified directly. It represents the proportion of variance expected
  **a priori** to be explained by the regression and must not be interpreted as
  an estimate of heritability [@perez2014bglr].
- For `BayesB2`, `BAYESB2_ZERO_PROBABILITY = 1e-5` is the intended probability
  of a zero marker effect. Because BGLR defines `probIn` as the probability of
  a nonzero effect, the implementation uses `probIn = 1 - 1e-5`.
- `BAYESB2_COUNTS = 1e6` makes the beta prior for `probIn` extremely
  concentrated around that value, approximating the nearly fixed probability
  convention used in the reference article.
- `ESS_REVIEW_THRESHOLD` and `GEWEKE_REVIEW_THRESHOLD` are diagnostic limits
  used to flag chains for additional inspection. They are not biological
  parameters and do not automatically exclude a method.

## 5. Input objects and invariant matrices

```{r input-objects}
matrices_object <- readRDS(
  MATRICES_FILE
)

ids <- matrices_object$ids
metadata <- matrices_object$metadata
phenotypes <- matrices_object$phenotypes
trait_dictionary <- matrices_object$trait_dictionary
X <- matrices_object$X
M <- matrices_object$M
G <- matrices_object$G
cv_folds <- matrices_object$cv_folds

traits <- colnames(
  phenotypes
)

X_fixed <- X[
  ,
  colnames(X) != "(Intercept)",
  drop = FALSE
]

method_names <- c(
  "RRBLUP",
  "BayesA",
  "BayesB",
  "BayesB2",
  "BayesC",
  "BL",
  "GBLUP"
)

method_labels <- c(
  RRBLUP = "RR-BLUP",
  BayesA = "BayesA",
  BayesB = "BayesB",
  BayesB2 = "BayesB2",
  BayesC = "BayesC",
  BL = "Bayesian Lasso",
  GBLUP = "GBLUP"
)
```


## 6. Model catalogue and BGLR specifications

```{r model-catalogue}
model_catalogue <- data.frame(
  Method = method_names,
  Label = unname(
    method_labels[
      method_names
    ]
  ),
  BGLR_component = c(
    "BRR",
    "BayesA",
    "BayesB",
    "BayesB",
    "BayesC",
    "BL",
    "RKHS"
  ),
  Genomic_input = c(
    "M",
    "M",
    "M",
    "M",
    "M",
    "M",
    "G"
  ),
  Main_assumption = c(
    "Common Gaussian variance for all marker effects",
    "Marker-specific scaled-t variances",
    "Scaled-t slab and a point mass at zero",
    "BayesB with zero-effect probability near 1e-5",
    "Common Gaussian slab and a point mass at zero",
    "Double-exponential marginal prior",
    "Genomic values structured by G"
  )
)

kable(
  model_catalogue,
  caption = "Seven genomic prediction methods"
)
```

```{r eta-specifications}
ETA_models <- list(
  RRBLUP = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BRR"
    )
  ),
  BayesA = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BayesA"
    )
  ),
  BayesB = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BayesB"
    )
  ),
  BayesB2 = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BayesB",
      probIn = BAYESB2_PROB_IN,
      counts = BAYESB2_COUNTS
    )
  ),
  BayesC = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BayesC"
    )
  ),
  BL = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BL"
    )
  ),
  GBLUP = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      K = G,
      model = "RKHS"
    )
  )
)
```

The official BGLR implementation writes a residual-variance chain for every
model. The genomic scalar used for convergence review differs by method:
marker variance for RR-BLUP, scale for BayesA and BayesB, common marker
variance for BayesC, shrinkage parameter for the Bayesian Lasso, and genomic
variance for GBLUP.

```{r trace-contract}
genomic_trace_suffix <- c(
  RRBLUP = "ETA_genomic_varB.dat",
  BayesA = "ETA_genomic_ScaleBayesA.dat",
  BayesB = "ETA_genomic_parBayesB.dat",
  BayesB2 = "ETA_genomic_parBayesB.dat",
  BayesC = "ETA_genomic_parBayesC.dat",
  BL = "ETA_genomic_lambda.dat",
  GBLUP = "ETA_genomic_varU.dat"
)

genomic_trace_header <- c(
  RRBLUP = FALSE,
  BayesA = FALSE,
  BayesB = TRUE,
  BayesB2 = TRUE,
  BayesC = TRUE,
  BL = FALSE,
  GBLUP = FALSE
)

genomic_trace_column <- c(
  RRBLUP = 1L,
  BayesA = 1L,
  BayesB = 2L,
  BayesB2 = 2L,
  BayesC = 2L,
  BL = 1L,
  GBLUP = 1L
)

genomic_trace_parameter <- c(
  RRBLUP = "Marker_variance",
  BayesA = "Scale",
  BayesB = "Scale",
  BayesB2 = "Scale",
  BayesC = "Marker_variance",
  BL = "Lambda",
  GBLUP = "Genomic_variance"
)
```

## 7. Run design and analytical signature

```{r run-design}
run_design <- expand.grid(
  Trait = traits,
  Fold = seq_len(
    matrices_object$settings$n_folds
  ),
  Method = method_names,
  stringsAsFactors = FALSE
)

run_design <- run_design[
  order(
    match(
      run_design$Trait,
      traits
    ),
    run_design$Fold,
    match(
      run_design$Method,
      method_names
    )
  ),
]

rownames(run_design) <- NULL

run_design$Run_number <- seq_len(
  nrow(run_design)
)

run_design$Seed <-
  SEED_BASE +
  run_design$Run_number

analysis_settings <- list(
  pipeline_version =
    PIPELINE_VERSION,
  n_iter = N_ITER,
  burn_in = BURN_IN,
  thin = THIN,
  R2 = R2_PRIOR,
  seed_base = SEED_BASE,
  bayesb2_zero_probability =
    BAYESB2_ZERO_PROBABILITY,
  bayesb2_prob_in =
    BAYESB2_PROB_IN,
  bayesb2_counts =
    BAYESB2_COUNTS,
  traits = traits,
  folds = seq_len(
    matrices_object$settings$n_folds
  ),
  methods = method_names
)

analysis_signature <- substr(
  digest::digest(
    list(
      ids = ids,
      phenotypes = phenotypes,
      X = X,
      M = M,
      G = G,
      cv_folds = cv_folds,
      settings = analysis_settings
    ),
    algo = "sha256"
  ),
  1,
  16
)

run_design$Signature <-
  analysis_signature

run_design$Run_id <- paste(
  run_design$Trait,
  paste0(
    "F",
    run_design$Fold
  ),
  run_design$Method,
  sep = "_"
)

run_design$Result_relative_file <- file.path(
  "output",
  "models",
  "runs",
  paste0(
    run_design$Run_id,
    "_",
    analysis_signature,
    ".rds"
  )
)

result_files <- file.path(
  here(),
  run_design$Result_relative_file
)

run_design$Completed_before_build <-
  file.exists(
    result_files
  )

write.csv(
  run_design,
  file.path(
    OUTPUT_DIR,
    "run_design.csv"
  ),
  row.names = FALSE
)

kable(
  head(
    run_design,
    14
  ),
  caption = paste(
    "First fourteen planned runs; analytical signature",
    analysis_signature
  )
)
```

## 8. Complete fitting with checkpointing

The loop below fits only runs whose compact result file is absent. A completed
run is never recomputed by an ordinary rebuild of this module. The optional
`PAPAYA_MAX_NEW_RUNS` environment variable limits how many new fits are
performed in one build. Its default value is 350, so an ordinary build fits
all remaining runs. Setting it to 7 validates one complete trait-fold block
containing all methods before the remaining analysis is launched.

The native BGLR files are preserved permanently for every new fit. For the
single-chain diagnostics used in this module, two of those saved chains are read
directly:

- residual variance;
- one method-specific genomic scale or variance parameter.

For each chain, effective sample size and the Geweke diagnostic are calculated
with `coda`. The thresholds below are review flags rather than automatic model
exclusion rules:

- ESS below 1,000;
- absolute Geweke statistic above 2.

```{r complete-fitting, results="hide"}
pending_indices <- head(
  which(
    !file.exists(
      result_files
    )
  ),
  n = MAX_NEW_RUNS_PER_BUILD
)

for (run_index in pending_indices) {
  trait_name <-
    run_design$Trait[
      run_index
    ]

  fold_number <-
    run_design$Fold[
      run_index
    ]

  method_name <-
    run_design$Method[
      run_index
    ]

  run_id <-
    run_design$Run_id[
      run_index
    ]

  run_seed <-
    run_design$Seed[
      run_index
    ]

  result_file <-
    result_files[
      run_index
    ]

  temporary_result_file <- paste0(
    result_file,
    ".tmp"
  )

  bglr_run_dir <- file.path(
    BGLR_RUNS_DIR,
    trait_name,
    paste0(
      "F",
      fold_number
    ),
    method_name
  )

  dir.create(
    bglr_run_dir,
    recursive = TRUE,
    showWarnings = FALSE
  )

  save_prefix <- file.path(
    bglr_run_dir,
    "bglr_"
  )

  unlink(
    Sys.glob(
      paste0(
        save_prefix,
        "*"
      )
    )
  )

  y_observed <- phenotypes[
    ,
    trait_name
  ]

  names(y_observed) <- ids

  validation_index <- which(
    cv_folds$Fold == fold_number
  )

  training_index <- which(
    cv_folds$Fold != fold_number
  )

  y_model <- y_observed
  y_model[validation_index] <- NA_real_

  set.seed(
    run_seed
  )

  fit_start <- proc.time()[["elapsed"]]

  fit <- BGLR::BGLR(
    y = y_model,
    ETA = ETA_models[[method_name]],
    response_type = "gaussian",
    nIter = N_ITER,
    burnIn = BURN_IN,
    thin = THIN,
    R2 = R2_PRIOR,
    saveAt = save_prefix,
    verbose = FALSE,
    rmExistingFiles = TRUE
  )

  fit_end <- proc.time()[["elapsed"]]

  predicted_validation <- fit$yHat[
    validation_index
  ]

  observed_validation <- y_observed[
    validation_index
  ]

  fixed_component <- as.vector(
    X_fixed %*%
      fit$ETA$fixed$b
  )

  genomic_component <- as.vector(
    fit$yHat -
      fit$mu -
      fixed_component
  )

  genomic_value_table <- data.frame(
    Run_id = rep(
      run_id,
      length(ids)
    ),
    Signature = rep(
      analysis_signature,
      length(ids)
    ),
    IND = ids,
    Trait = rep(
      trait_name,
      length(ids)
    ),
    Fold = rep(
      fold_number,
      length(ids)
    ),
    Method = rep(
      method_name,
      length(ids)
    ),
    Genomic_value =
      genomic_component
  )

  prediction_table <- data.frame(
    Run_id = run_id,
    Signature = analysis_signature,
    IND = ids[
      validation_index
    ],
    Trait = trait_name,
    Fold = fold_number,
    Method = method_name,
    Observed = observed_validation,
    Predicted = predicted_validation,
    Fixed_component =
      fixed_component[
        validation_index
      ],
    Genomic_component =
      genomic_component[
        validation_index
      ]
  )

  posterior_prob_in <- c(
    fit$ETA$genomic$probIn,
    NA_real_
  )[1]

  metric_table <- data.frame(
    Run_id = run_id,
    Signature = analysis_signature,
    Trait = trait_name,
    Fold = fold_number,
    Method = method_name,
    Seed = run_seed,
    Training_n = length(
      training_index
    ),
    Validation_n = length(
      validation_index
    ),
    Correlation = cor(
      observed_validation,
      predicted_validation
    ),
    RMSE = sqrt(
      mean(
        (
          observed_validation -
            predicted_validation
        )^2
      )
    ),
    MAE = mean(
      abs(
        observed_validation -
          predicted_validation
      )
    ),
    Prediction_bias = mean(
      predicted_validation -
        observed_validation
    ),
    Regression_slope =
      cov(
        observed_validation,
        predicted_validation
      ) /
      var(
        predicted_validation
      ),
    DIC = fit$fit$DIC,
    Effective_parameters =
      fit$fit$pD,
    Log_likelihood_at_posterior_mean =
      fit$fit$
        logLikAtPostMean,
    Posterior_mean =
      fit$mu,
    Posterior_mean_SD = c(
      fit$SD.mu,
      NA_real_
    )[1],
    Residual_variance =
      fit$varE,
    Residual_variance_SD = c(
      fit$SD.varE,
      NA_real_
    )[1],
    Posterior_probIn =
      posterior_prob_in,
    Elapsed_seconds =
      fit_end - fit_start,
    Finite_predictions = all(
      is.finite(
        predicted_validation
      )
    )
  )

  metric_table$Finite_metrics <- all(
    is.finite(
      as.numeric(
        metric_table[
          ,
          c(
            "Correlation",
            "RMSE",
            "MAE",
            "Prediction_bias",
            "Regression_slope",
            "DIC",
            "Effective_parameters",
            "Log_likelihood_at_posterior_mean",
            "Posterior_mean",
            "Residual_variance",
            "Elapsed_seconds"
          )
        ]
      )
    )
  )

  fixed_effect_table <- data.frame(
    Run_id = rep(
      run_id,
      length(
        fit$ETA$fixed$b
      )
    ),
    Signature = rep(
      analysis_signature,
      length(
        fit$ETA$fixed$b
      )
    ),
    Trait = rep(
      trait_name,
      length(
        fit$ETA$fixed$b
      )
    ),
    Fold = rep(
      fold_number,
      length(
        fit$ETA$fixed$b
      )
    ),
    Method = rep(
      method_name,
      length(
        fit$ETA$fixed$b
      )
    ),
    Effect = names(
      fit$ETA$fixed$b
    ),
    Posterior_mean = as.numeric(
      fit$ETA$fixed$b
    ),
    Posterior_SD = as.numeric(
      fit$ETA$fixed$SD.b
    )
  )

  marker_count <-
    as.integer(
      method_name != "GBLUP"
    ) *
    ncol(M)

  marker_indices <- seq_len(
    marker_count
  )

  marker_effect_values <-
    c(
      fit$ETA$genomic$b
    )[
      marker_indices
    ]

  marker_effect_sds <-
    c(
      fit$ETA$genomic$SD.b
    )[
      marker_indices
    ]

  marker_effect_table <- data.frame(
    Run_id = rep(
      run_id,
      marker_count
    ),
    Signature = rep(
      analysis_signature,
      marker_count
    ),
    Trait = rep(
      trait_name,
      marker_count
    ),
    Fold = rep(
      fold_number,
      marker_count
    ),
    Method = rep(
      method_name,
      marker_count
    ),
    Marker = colnames(M)[
      marker_indices
    ],
    Posterior_mean = as.numeric(
      marker_effect_values
    ),
    Posterior_SD = as.numeric(
      marker_effect_sds
    )
  )

  residual_trace_full <- scan(
    paste0(
      save_prefix,
      "varE.dat"
    ),
    quiet = TRUE
  )

  genomic_trace_table_full <- read.table(
    paste0(
      save_prefix,
      genomic_trace_suffix[
        method_name
      ]
    ),
    header =
      genomic_trace_header[
        method_name
      ],
    check.names = FALSE
  )

  genomic_trace_full <-
    genomic_trace_table_full[
      ,
      genomic_trace_column[
        method_name
      ]
    ]

  posterior_sample_indices <- seq.int(
    from =
      BURN_IN_SAVED_SAMPLES +
      1L,
    to = length(
      residual_trace_full
    )
  )

  residual_trace <-
    residual_trace_full[
      posterior_sample_indices
    ]

  genomic_trace <-
    genomic_trace_full[
      posterior_sample_indices
    ]

  trace_matrix <- cbind(
    Residual_variance =
      residual_trace,
    Genomic_parameter =
      genomic_trace
  )

  trace_object <- coda::mcmc(
    trace_matrix,
    start = BURN_IN + THIN,
    thin = THIN
  )

  trace_ess <- coda::effectiveSize(
    trace_object
  )

  trace_geweke <- coda::geweke.diag(
    trace_object
  )$z

  trace_parameter_names <- c(
    "Residual_variance",
    unname(
      genomic_trace_parameter[
        method_name
      ]
    )
  )

  convergence_table <- data.frame(
    Run_id = rep(
      run_id,
      length(
        trace_parameter_names
      )
    ),
    Signature = rep(
      analysis_signature,
      length(
        trace_parameter_names
      )
    ),
    Trait = rep(
      trait_name,
      length(
        trace_parameter_names
      )
    ),
    Fold = rep(
      fold_number,
      length(
        trace_parameter_names
      )
    ),
    Method = rep(
      method_name,
      length(
        trace_parameter_names
      )
    ),
    Parameter =
      trace_parameter_names,
    Retained_samples = rep(
      nrow(
        trace_matrix
      ),
      length(
        trace_parameter_names
      )
    ),
    Posterior_mean = colMeans(
      trace_matrix
    ),
    Posterior_SD = apply(
      trace_matrix,
      2,
      sd
    ),
    Effective_sample_size = as.numeric(
      trace_ess
    ),
    Geweke_Z = as.numeric(
      trace_geweke
    )
  )

  convergence_table$Review_flag <-
    !is.finite(
      convergence_table$
        Effective_sample_size
    ) |
    convergence_table$
      Effective_sample_size <
      ESS_REVIEW_THRESHOLD |
    !is.finite(
      convergence_table$
        Geweke_Z
    ) |
    abs(
      convergence_table$
        Geweke_Z
    ) >
      GEWEKE_REVIEW_THRESHOLD

  run_audit <- data.frame(
    Run_id = run_id,
    Signature = analysis_signature,
    Trait = trait_name,
    Fold = fold_number,
    Method = method_name,
    Seed = run_seed,
    Expected_raw_saved_samples =
      RAW_SAVED_SAMPLES,
    Observed_raw_saved_samples =
      length(
        residual_trace_full
      ),
    Expected_retained_samples =
      RETAINED_SAMPLES,
    Observed_retained_samples =
      length(
        residual_trace
      ),
    Finite_predictions = all(
      is.finite(
        predicted_validation
      )
    ),
    Minimum_effective_sample_size =
      min(
        convergence_table$
          Effective_sample_size,
        na.rm = TRUE
      ),
    Maximum_absolute_Geweke =
      max(
        abs(
          convergence_table$
            Geweke_Z
        ),
        na.rm = TRUE
      ),
    Parameters_flagged_for_review =
      sum(
        convergence_table$
          Review_flag
      ),
    Elapsed_seconds =
      fit_end - fit_start
  )

  compact_result <- list(
    run = run_design[
      run_index,
      ,
      drop = FALSE
    ],
    metrics = metric_table,
    predictions = prediction_table,
    genomic_values =
      genomic_value_table,
    fixed_effects = fixed_effect_table,
    marker_effects = marker_effect_table,
    convergence = convergence_table,
    run_audit = run_audit,
    settings = analysis_settings
  )

  saveRDS(
    compact_result,
    temporary_result_file
  )

  file.rename(
    temporary_result_file,
    result_file
  )

  progress_table <- run_design
  progress_table$Completed <-
    file.exists(
      result_files
    )

  write.csv(
    progress_table,
    file.path(
      OUTPUT_DIR,
      "run_progress.csv"
    ),
    row.names = FALSE
  )

  progress_summary <- data.frame(
    Signature =
      analysis_signature,
    Completed_runs = sum(
      progress_table$Completed
    ),
    Total_runs = nrow(
      progress_table
    ),
    Remaining_runs = sum(
      !progress_table$Completed
    ),
    Last_completed_run =
      run_id,
    Updated_at = format(
      Sys.time(),
      "%Y-%m-%d %H:%M:%S %Z"
    )
  )

  write.csv(
    progress_summary,
    file.path(
      OUTPUT_DIR,
      "progress_summary.csv"
    ),
    row.names = FALSE
  )

  fit <- NULL
  compact_result <- NULL
  trace_object <- NULL

  gc(
    verbose = FALSE
  )
}
```


## 9. Consolidation of the official results

The compact checkpoints are consolidated without refitting completed models.
The final analytical object retains the fields required by the comparison and
ranking modules: the analytical signature and settings, run design, model
catalogue, OOF predictions, fold-specific metrics, genomic values by fold, and
formal convergence diagnostics.

```{r consolidate-results}
completed_after_build <- file.exists(
  result_files
)

completed_indices <- which(
  completed_after_build
)

metric_parts <- vector("list", length(completed_indices))
prediction_parts <- vector("list", length(completed_indices))
genomic_value_parts <- vector("list", length(completed_indices))
convergence_parts <- vector("list", length(completed_indices))

for (part_index in seq_along(completed_indices)) {
  run_index <- completed_indices[part_index]
  compact_result <- readRDS(result_files[run_index])
  metric_parts[[part_index]] <- compact_result$metrics
  prediction_parts[[part_index]] <- compact_result$predictions
  genomic_value_parts[[part_index]] <- compact_result$genomic_values
  convergence_parts[[part_index]] <- compact_result$convergence
}

cv_metrics_by_fold <- bind_rows(metric_parts)
cv_predictions <- bind_rows(prediction_parts)
genomic_values_by_fold <- bind_rows(genomic_value_parts)
convergence_diagnostics <- bind_rows(convergence_parts)

convergence_summary <- convergence_diagnostics |>
  group_by(Method) |>
  summarise(
    Runs = n_distinct(Run_id),
    Parameters_reviewed = n(),
    Parameters_flagged = sum(Review_flag),
    Runs_with_flags = n_distinct(Run_id[Review_flag]),
    Minimum_ESS = min(Effective_sample_size, na.rm = TRUE),
    Median_ESS = median(Effective_sample_size, na.rm = TRUE),
    Maximum_absolute_Geweke = max(abs(Geweke_Z), na.rm = TRUE),
    .groups = "drop"
  )

convergence_summary$Method_label <-
  unname(method_labels[convergence_summary$Method])

fit_status <- data.frame(
  Planned_fits = nrow(run_design),
  Completed_fits = sum(completed_after_build),
  Remaining_fits = sum(!completed_after_build)
)

kable(
  fit_status,
  caption = "Status of the planned genomic-prediction fits"
)
```

## 10. Analytical outputs

The fold-level files are kept as direct analytical exports. Detailed fixed
effects, marker effects, run-completion audits, native-file inventories, and
duplicated predictive or runtime summaries are not part of the final
consolidated object because they are not required by the downstream scientific
workflow.

```{r save-complete-outputs}
write.csv(cv_predictions, file.path(OUTPUT_DIR, "cv_predictions.csv"), row.names = FALSE)
write.csv(genomic_values_by_fold, file.path(OUTPUT_DIR, "genomic_values_by_fold.csv"), row.names = FALSE)
write.csv(cv_metrics_by_fold, file.path(OUTPUT_DIR, "cv_metrics_by_fold.csv"), row.names = FALSE)
write.csv(convergence_diagnostics, file.path(OUTPUT_DIR, "convergence_diagnostics.csv"), row.names = FALSE)
write.csv(convergence_summary, file.path(OUTPUT_DIR, "convergence_summary.csv"), row.names = FALSE)

saveRDS(
  list(
    signature = analysis_signature,
    settings = analysis_settings,
    run_design = run_design,
    model_catalogue = model_catalogue,
    trait_dictionary = trait_dictionary,
    predictions = cv_predictions,
    genomic_values_by_fold = genomic_values_by_fold,
    metrics_by_fold = cv_metrics_by_fold,
    convergence_diagnostics = convergence_diagnostics
  ),
  file.path(OUTPUT_DIR, "models_object.rds")
)
```

## 11. MCMC convergence diagnostics

ESS and the Geweke statistic are single-chain diagnostics. They do not prove
convergence, but they identify runs and scalar parameters that require closer
inspection. The review thresholds are diagnostic indicators and are not
automatic model-exclusion rules.

```{r convergence-summary-table}
convergence_summary |>
  select(
    Method_label, Runs, Parameters_reviewed, Parameters_flagged,
    Runs_with_flags, Minimum_ESS, Median_ESS, Maximum_absolute_Geweke
  ) |>
  kable(
    digits = 3,
    caption = "Single-chain MCMC diagnostics by method"
  )
```

```{r convergence-flag-table}
convergence_diagnostics |>
  filter(Review_flag) |>
  arrange(Method, Trait, Fold, Parameter) |>
  kable(
    digits = 4,
    caption = "Scalar chains flagged for additional review"
  )
```

The native BGLR chains remain in `output/models/bglr/` for posterior inspection
and for the final graphical summaries. They are not duplicated in this module.

## 12. Next step

Module 04 uses these official outputs, without refitting models, to compare the
seven methods within each trait, integrate OOF predictive performance, DIC and
MCMC diagnostics, derive the selected OOF genomic values, and construct
trait-specific genomic rankings.
