---
title: "Model fitting and MCMC diagnostics for genomic prediction in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

This module documents the complete cross-validated fitting of seven genomic
prediction methods for ten papaya traits. The 350 fits are executed once by
`analysis/03_models_fit.R`; this page reads and presents the resulting canonical
object without refitting models. The same five deterministic,
family-stratified folds are used for every method and trait.

The complete analysis contains:

\[
10\ \text{traits}
\times
5\ \text{folds}
\times
7\ \text{methods}
=
350\ \text{fits}.
\]

All methods are fitted exclusively with BGLR [@perez2014bglr]:

1. RR-BLUP, implemented with the BGLR `BRR` component and the complete marker
   matrix `M` [@endelman2011rrblup];
2. BayesA with `M`;
3. BayesB with `M`;
4. BayesB2 with `M` and a zero-effect probability operationally fixed near
   \(10^{-5}\);
5. BayesC with `M`;
6. Bayesian Lasso with `M` [@park2008bayesianlasso], with the shrinkage
   parameter `lambda` fixed at the global value deterministically calibrated
   by the BGLR default rule;
7. GBLUP with the genomic relationship matrix `G` and the BGLR `RKHS`
   component.

BayesA and BayesB follow the genomic-prediction framework introduced by
Meuwissen et al. and its subsequent Bayesian-alphabet development
[@meuwissen2001prediction; @habier2011bayesianalphabet].

RR-BLUP and Bayesian ridge regression are not treated as separate methods. The
Gaussian marker-effect model implemented by `model = "BRR"` is the RR-BLUP
implementation used in this project.

## 2. Analytical contract

The matrices constructed in Module 02 are used without fold-specific
reconstruction:

- `X` contains the intercept and the family fixed-effect coding;
- `M` contains the 72 centered allele-dosage columns;
- `G` contains the additive genomic relationships;
- each fold identifies 30 validation individuals and 120 training individuals.

BGLR includes an intercept by default. Therefore, the intercept column is
removed from `X`, leaving only the family columns to be passed through the
`FIXED` component.

For each trait and fold, only a copy of the phenotype vector is changed. The
30 validation observations are replaced by `NA`; `X`, `M`, and `G` remain
global and unchanged.

The MCMC settings follow the main methods section of the reference article
[@silva2021brrguava]:

- 1,000,000 iterations;
- 200,000 burn-in iterations;
- thinning equal to 4;
- 200,000 retained posterior samples.

The BGLR default prior allocation parameter is made explicit with `R2 = 0.5`.
For the Bayesian Lasso, `lambda` is held fixed at the deterministic `lambda0`
obtained from BGLR's own calibration. Because the predictor contains two
`ETA` terms (family `FIXED` and genomic `BL`), BGLR allocates `R2/2` to the
BL component. Matrix `M` is global, so `MSx` and `lambda0` are also global
and identical across traits and folds.

## 3. Single canonical execution

Model fitting is separated from presentation. The complete analysis is executed explicitly with:

```bash
Rscript --vanilla analysis/03_models_fit.R
```

The script performs the 350 trait-fold-method fits sequentially in one R session. Deterministic seeds are defined for every fit.

Native BGLR files are written directly to `output/models/bglr/Trait/Fold/Method/`. The compact canonical analytical object is `output/models/models_object.rds`. Rendering this page never fits or refits a genomic model.

## 4. Canonical M03 outputs

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
  library(here)
})
```

```{r load-model-object}
MODELS_FILE <- here("output", "models", "models_object.rds")
BGLR_DIR <- here("output", "models", "bglr")

if (!file.exists(MODELS_FILE)) {
  stop("Canonical M03 object not found. Run `Rscript --vanilla analysis/03_models_fit.R` first.")
}

models_object <- readRDS(MODELS_FILE)
settings <- models_object$settings
run_design <- models_object$run_design
model_catalogue <- models_object$model_catalogue
cv_predictions <- models_object$predictions
genomic_values_by_fold <- models_object$genomic_values_by_fold
cv_metrics_by_fold <- models_object$metrics_by_fold
convergence_diagnostics <- models_object$convergence_diagnostics

expected_fits <-
  length(settings$traits) * length(settings$folds) * length(settings$methods)

stopifnot(
  expected_fits == 350L,
  nrow(run_design) == expected_fits,
  nrow(cv_metrics_by_fold) == expected_fits,
  nrow(cv_predictions) == 150L * length(settings$traits) * length(settings$methods),
  nrow(genomic_values_by_fold) == 150L * expected_fits,
  nrow(convergence_diagnostics) == 2L * expected_fits
)
```

The root of `output/models/` contains the canonical object and direct analytical exports. The only subdirectory belonging to the final M03 pipeline is `bglr/`.

## 5. Fitting settings

```{r settings-table}
settings_table <- data.frame(
  Setting = c(
    "Pipeline version", "Iterations", "Burn-in", "Thinning", "R2 prior",
    "Seed base", "BayesB2 zero probability", "BL lambda type", "BL lambda"
  ),
  Value = c(
    settings$pipeline_version, settings$n_iter, settings$burn_in, settings$thin,
    settings$R2, settings$seed_base, settings$bayesb2_zero_probability,
    settings$bl_lambda_type, settings$bl_lambda
  )
)
kable(settings_table, caption = "Canonical M03 fitting settings")
```

These settings are fitting specifications, not model-selection criteria.

## 6. Model catalogue

```{r model-catalogue-table}
kable(model_catalogue, caption = "Seven genomic prediction methods")
```

The complete `ETA` specifications and the fixed-lambda Bayesian Lasso contract are centralized in `analysis/03_models_fit.R`.

## 7. Run design

```{r run-design-table}
stopifnot(
  length(unique(run_design$Signature)) == 1L,
  identical(as.character(unique(run_design$Signature)), as.character(models_object$signature))
)
kable(
  head(run_design, 14),
  caption = paste("First fourteen planned fits; analytical signature", models_object$signature)
)
```

The design contains 350 deterministic fits: ten traits, five family-stratified folds, and seven methods. `run_design.csv` is a permanent design manifest, not an execution-state file.

## 8. Analytical results

```{r fit-status}
fit_status <- data.frame(
  Planned_fits = expected_fits,
  Completed_fits = nrow(cv_metrics_by_fold),
  Prediction_rows = nrow(cv_predictions),
  Genomic_value_rows = nrow(genomic_values_by_fold)
)
kable(fit_status, caption = "Canonical M03 analytical output dimensions")
```

The direct fold-level exports are `cv_predictions.csv`, `genomic_values_by_fold.csv`, `cv_metrics_by_fold.csv`, `convergence_diagnostics.csv`, and `convergence_summary.csv`.

## 9. MCMC convergence diagnostics

```{r convergence-summary}
convergence_summary <- convergence_diagnostics |>
  group_by(Method) |>
  summarise(
    Runs = n_distinct(Run_id),
    Parameters_reviewed = n(),
    Parameters_flagged = sum(Review_flag),
    Runs_with_flags = n_distinct(Run_id[Review_flag]),
    Minimum_ESS = min(Effective_sample_size, na.rm = TRUE),
    Median_ESS = median(Effective_sample_size, na.rm = TRUE),
    Maximum_absolute_Geweke = max(abs(Geweke_Z), na.rm = TRUE),
    .groups = "drop"
  )
convergence_summary$Method_label <-
  unname(setNames(model_catalogue$Label, model_catalogue$Method)[convergence_summary$Method])

kable(
  convergence_summary |>
    select(Method_label, Runs, Parameters_reviewed, Parameters_flagged,
           Runs_with_flags, Minimum_ESS, Median_ESS, Maximum_absolute_Geweke),
  digits = 3,
  caption = "Single-chain MCMC diagnostics by method"
)
```

```{r convergence-flags}
convergence_diagnostics |>
  filter(Review_flag) |>
  arrange(Method, Trait, Fold, Parameter) |>
  kable(digits = 4, caption = "Scalar chains flagged for additional review")
```

ESS and the Geweke statistic are single-chain diagnostics. They identify chains for additional inspection but do not prove convergence and are not automatic method-exclusion rules.

## 10. Native BGLR outputs

```{r bglr-inventory}
expected_bglr_dirs <- expand.grid(
  Trait = settings$traits,
  Fold = settings$folds,
  Method = settings$methods,
  stringsAsFactors = FALSE
)
expected_bglr_paths <- file.path(
  BGLR_DIR,
  expected_bglr_dirs$Trait,
  paste0("F", expected_bglr_dirs$Fold),
  expected_bglr_dirs$Method
)
if (!all(dir.exists(expected_bglr_paths))) stop("M03 BGLR inventory is incomplete.")

bglr_inventory <- data.frame(
  Trait_directories = length(list.dirs(BGLR_DIR, recursive = FALSE)),
  Fit_directories = sum(dir.exists(expected_bglr_paths)),
  Native_files = length(list.files(BGLR_DIR, recursive = TRUE))
)
kable(bglr_inventory, caption = "Native BGLR output inventory")
```

The native files contain the thinned MCMC chains used for posterior diagnostics and graphical inspection.

## 11. Next step

Module 04 reads `models_object.rds` without refitting models to compare the seven methods within each trait and derive the downstream genomic results.
