---
title: "Model fitting and MCMC diagnostics for genomic prediction in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

This module fits seven genomic-prediction methods to ten papaya traits using
the same five family-stratified cross-validation folds constructed in Module
02.

The workflow proceeds directly from the matrices to model fitting, out-of-fold
prediction, decomposition of genomic values, fold-level predictive metrics,
and MCMC diagnostics.

The complete analysis contains

\[
10\ \text{traits}
\times
5\ \text{folds}
\times
7\ \text{methods}
=
350\ \text{fits}.
\]

The evaluated methods are RR-BLUP, BayesA, BayesB, BayesB2, BayesC, Bayesian
Lasso (BL), and GBLUP. All models are fitted with BGLR [@perez2014bglr].

## 2. Packages, inputs, and output directories

```{r packages}
suppressPackageStartupMessages({
  library(BGLR)
  library(coda)
  library(dplyr)
  library(knitr)
  library(here)
})
```

The matrices constructed in Module 02 are read directly. Outputs from this
module are written to `output/models/`, while the native BGLR chains are
organized under `output/models/bglr/`.

```{r inputs}
MATRICES_FILE <- here("output", "matrices", "matrices_object.rds")
OUTPUT_DIR <- here("output", "models")
BGLR_DIR <- file.path(OUTPUT_DIR, "bglr")

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(BGLR_DIR, recursive = TRUE, showWarnings = FALSE)

matrices_object <- readRDS(MATRICES_FILE)

ids <- matrices_object$ids
metadata <- matrices_object$metadata
phenotypes <- matrices_object$phenotypes
trait_dictionary <- matrices_object$trait_dictionary
X <- matrices_object$X
M <- matrices_object$M
G <- matrices_object$G
cv_folds <- matrices_object$cv_folds
traits <- colnames(phenotypes)
```

The structure received from the previous module can be inspected directly.

```{r input-summary}
data.frame(
  Object = c("Phenotypes", "Matrix X", "Matrix M", "Matrix G", "Folds"),
  Rows = c(
    nrow(phenotypes), nrow(X), nrow(M), nrow(G), nrow(cv_folds)
  ),
  Columns = c(
    ncol(phenotypes), ncol(X), ncol(M), ncol(G), ncol(cv_folds)
  )
) |>
  kable(caption = "Structure of the Module 03 inputs")
```

## 3. Statistical model

For each trait, the common model is

\[
\mathbf y =
\mathbf 1\mu +
\mathbf X_F\boldsymbol\beta_F +
\mathbf g +
\boldsymbol\varepsilon,
\]

where \(\mu\) is the intercept, \(\mathbf X_F\boldsymbol\beta_F\) represents
the family fixed effect, \(\mathbf g\) is the genomic component, and
\(\boldsymbol\varepsilon\) is the residual vector.

### 3.1 Intercept and family fixed effect

BGLR fits an intercept automatically. Therefore, the intercept column created
in Module 02 is removed from `X` before the family term is specified.

```{r family-fixed-effect}
X_fixed <- X[
  ,
  colnames(X) != "(Intercept)",
  drop = FALSE
]

colnames(X_fixed)
```

In BGLR, `model = "FIXED"` means that the coefficients associated with the
family columns are estimated as fixed effects without the shrinkage applied to
the genomic effects.

### 3.2 Genomic component

For the six methods based directly on markers,

\[
\mathbf g = \mathbf M\boldsymbol\alpha,
\]

so that

\[
\mathbf y =
\mathbf 1\mu +
\mathbf X_F\boldsymbol\beta_F +
\mathbf M\boldsymbol\alpha +
\boldsymbol\varepsilon.
\]

These methods mainly differ in the prior distribution assigned to marker
effects \(\boldsymbol\alpha\).

GBLUP instead models genomic values directly as

\[
\mathbf g \sim
N\left(\mathbf 0,\mathbf G\sigma_g^2\right).
\]

## 4. Cross-validation

The five folds defined in Module 02 are used without reconstruction. Each fold
contains 30 validation individuals and 120 training individuals while
preserving family representation.

```{r cv-summary}
kable(
  matrices_object$fold_summary,
  caption = "Distribution of individuals across the five folds"
)
```

During each fit, phenotypes of the individuals in the validation fold are set
to `NA`. BGLR fits the model using the remaining phenotypes and returns
predictions for the masked individuals.

## 5. MCMC settings

The analysis uses 1,000,000 iterations, a burn-in of 200,000 iterations, and a
thinning interval of 4, following the main configuration of the reference
analysis [@silva2021brrguava]. The global BGLR prior-allocation parameter is
set to `R2 = 0.5`.

```{r mcmc-settings}
N_ITER <- 1000000L
BURN_IN <- 200000L
THIN <- 4L
R2_PRIOR <- 0.5
SEED_BASE <- 202608050L

RAW_SAVED_SAMPLES <- floor(N_ITER / THIN)
BURN_IN_SAVED_SAMPLES <- floor(BURN_IN / THIN)
RETAINED_SAMPLES <- RAW_SAVED_SAMPLES - BURN_IN_SAVED_SAMPLES

BAYESB2_ZERO_PROBABILITY <- 1e-5
BAYESB2_PROB_IN <- 1 - BAYESB2_ZERO_PROBABILITY
BAYESB2_COUNTS <- 1e6

ESS_REVIEW_THRESHOLD <- 1000
GEWEKE_REVIEW_THRESHOLD <- 2

BL_LAMBDA_TYPE <- "FIXED"
```

```{r mcmc-settings-table}
data.frame(
  Setting = c(
    "Iterations",
    "Burn-in",
    "Thinning",
    "Retained posterior samples",
    "R2",
    "BayesB2 zero-effect probability"
  ),
  Value = c(
    N_ITER,
    BURN_IN,
    THIN,
    RETAINED_SAMPLES,
    R2_PRIOR,
    BAYESB2_ZERO_PROBABILITY
  )
) |>
  kable(caption = "MCMC settings used for the seven methods")
```

## 6. Genomic prediction methods

```{r model-catalogue}
method_names <- c(
  "RRBLUP", "BayesA", "BayesB", "BayesB2",
  "BayesC", "BL", "GBLUP"
)

method_labels <- c(
  RRBLUP = "RR-BLUP",
  BayesA = "BayesA",
  BayesB = "BayesB",
  BayesB2 = "BayesB2",
  BayesC = "BayesC",
  BL = "Bayesian Lasso",
  GBLUP = "GBLUP"
)

data.frame(
  Method = unname(method_labels[method_names]),
  BGLR_component = c(
    "BRR", "BayesA", "BayesB", "BayesB",
    "BayesC", "BL", "RKHS"
  ),
  Genomic_input = c("M", "M", "M", "M", "M", "M", "G"),
  Structure = c(
    "Common Gaussian variance for all marker effects",
    "Marker-specific variances",
    "Point mass at zero and a continuous component for nonzero effects",
    "BayesB with zero-effect probability near 1e-5",
    "Point mass at zero and common variance for nonzero effects",
    "Double-exponential marginal prior with fixed lambda",
    "Genomic values structured by matrix G"
  )
) |>
  kable(caption = "Genomic prediction methods evaluated")
```

### 6.1 RR-BLUP

RR-BLUP is implemented with the BGLR `BRR` component. All marker effects share
a common Gaussian variance, producing ridge-type shrinkage
[@endelman2011rrblup; @perez2014bglr].

### 6.2 BayesA

BayesA assigns a separate variance to each marker, allowing different amounts
of shrinkage across markers [@meuwissen2001prediction;
@habier2011bayesianalphabet].

### 6.3 BayesB

BayesB uses a mixture between a point mass at zero and a continuous
distribution for nonzero effects. It therefore combines shrinkage and variable
selection.

### 6.4 BayesB2

`BayesB2` is the name used in this project for a specific parameterization of
the BGLR `BayesB` component. The intended probability of a zero marker effect
is

\[
P(\alpha_j=0)=10^{-5}.
\]

Because `probIn` represents the probability that a marker is included with a
nonzero effect,

\[
probIn=1-10^{-5}=0.99999.
\]

The value `counts = 10^6` makes the beta prior for `probIn` highly concentrated
around this value.

### 6.5 BayesC

BayesC also allows a marker effect to be zero, but nonzero effects share a
common variance.

### 6.6 Bayesian Lasso and definition of the fixed lambda

The Bayesian Lasso uses a double-exponential marginal prior for marker effects
[@park2008bayesianlasso]. The shrinkage parameter \(\lambda\) controls the
amount of shrinkage.

Here, `type = "FIXED"` has a different meaning from `model = "FIXED"` used for
family. Within the BL term, `type = "FIXED"` means that \(\lambda\) remains
constant throughout the MCMC chain.

The adopted value corresponds to BGLR's own initial calibration. Because the
predictor contains two linear terms — family and the BL genomic component —
the share of `R2` allocated to BL is

\[
R^2_{BL}=\frac{0.5}{2}=0.25.
\]

For the centered marker matrix,

\[
MS_x=
\frac{1}{n}\sum_j\sum_i m_{ij}^2-
\sum_j\bar m_j^2,
\]

followed by

\[
\lambda_0^2=
2\frac{1-R^2}{R^2_{BL}}MS_x.
\]

```{r bl-lambda}
BL_N_LINEAR_TERMS <- 2L
BL_COMPONENT_R2 <- R2_PRIOR / BL_N_LINEAR_TERMS

BL_MSX <-
  sum(apply(M, 2, function(x) sum(x^2))) / nrow(M) -
  sum(colMeans(M)^2)

BL_LAMBDA_SQUARED <-
  2 * (1 - R2_PRIOR) / BL_COMPONENT_R2 * BL_MSX

BL_LAMBDA <- sqrt(BL_LAMBDA_SQUARED)
```

```{r bl-lambda-table}
data.frame(
  Quantity = c("MSx", "Global R2", "BL R2", "Fixed lambda0"),
  Value = c(BL_MSX, R2_PRIOR, BL_COMPONENT_R2, BL_LAMBDA)
) |>
  kable(
    digits = 6,
    caption = "Definition of the Bayesian-Lasso shrinkage parameter"
  )
```

For these data, \(MS_x\approx25.6050\) and
\(\lambda_0\approx10.1203\). The same value is used for every trait and fold.
It was held fixed after the standard stochastic update showed numerical
instability, without using predictive correlation, RMSE, DIC, or any
trait-specific result to choose the value.

### 6.7 GBLUP

GBLUP uses the genomic relationship matrix `G`. In BGLR, this structure is
specified through an `RKHS` component with `K = G`.

## 7. BGLR model specifications

The family term is common to all seven methods. Each model then receives its
corresponding genomic component.

```{r eta-specifications}
family_term <- list(
  X = X_fixed,
  model = "FIXED"
)

ETA_models <- list(
  RRBLUP = list(
    fixed = family_term,
    genomic = list(X = M, model = "BRR")
  ),
  BayesA = list(
    fixed = family_term,
    genomic = list(X = M, model = "BayesA")
  ),
  BayesB = list(
    fixed = family_term,
    genomic = list(X = M, model = "BayesB")
  ),
  BayesB2 = list(
    fixed = family_term,
    genomic = list(
      X = M,
      model = "BayesB",
      probIn = BAYESB2_PROB_IN,
      counts = BAYESB2_COUNTS
    )
  ),
  BayesC = list(
    fixed = family_term,
    genomic = list(X = M, model = "BayesC")
  ),
  BL = list(
    fixed = family_term,
    genomic = list(
      X = M,
      model = "BL",
      type = BL_LAMBDA_TYPE,
      lambda = BL_LAMBDA
    )
  ),
  GBLUP = list(
    fixed = family_term,
    genomic = list(K = G, model = "RKHS")
  )
)
```

## 8. Trait, fold, and method combinations

The same seven specifications are applied to all ten traits and five folds. A
deterministic seed is assigned to each combination.

```{r run-design}
run_design <- expand.grid(
  Trait = traits,
  Fold = seq_len(matrices_object$settings$n_folds),
  Method = method_names,
  stringsAsFactors = FALSE
)

run_design <- run_design[
  order(
    match(run_design$Trait, traits),
    run_design$Fold,
    match(run_design$Method, method_names)
  ),
]
rownames(run_design) <- NULL

run_design$Run_number <- seq_len(nrow(run_design))
run_design$Seed <- SEED_BASE + run_design$Run_number
run_design$Run_id <- paste(
  run_design$Trait,
  paste0("F", run_design$Fold),
  run_design$Method,
  sep = "_"
)

run_design |>
  count(Method, name = "Number_of_fits") |>
  kable(caption = "Number of fits by method")
```

## 9. Fitting the 350 models

The block below iterates over the trait, fold, and method combinations. Short
comments highlight the main steps of each fit.

Before the loop, the native files containing the scalar chains used in the
MCMC diagnostics are identified.

```{r trace-specifications}
genomic_trace_suffix <- c(
  RRBLUP = "ETA_genomic_varB.dat",
  BayesA = "ETA_genomic_ScaleBayesA.dat",
  BayesB = "ETA_genomic_parBayesB.dat",
  BayesB2 = "ETA_genomic_parBayesB.dat",
  BayesC = "ETA_genomic_parBayesC.dat",
  BL = "mu.dat",
  GBLUP = "ETA_genomic_varU.dat"
)

genomic_trace_header <- c(
  RRBLUP = FALSE,
  BayesA = FALSE,
  BayesB = TRUE,
  BayesB2 = TRUE,
  BayesC = TRUE,
  BL = FALSE,
  GBLUP = FALSE
)

genomic_trace_column <- c(
  RRBLUP = 1L,
  BayesA = 1L,
  BayesB = 2L,
  BayesB2 = 2L,
  BayesC = 2L,
  BL = 1L,
  GBLUP = 1L
)

genomic_trace_parameter <- c(
  RRBLUP = "Marker_variance",
  BayesA = "Scale",
  BayesB = "Scale",
  BayesB2 = "Scale",
  BayesC = "Marker_variance",
  BL = "Intercept",
  GBLUP = "Genomic_variance"
)
```

```{r fit-models, eval=FALSE}
# Lists that collect the outputs from each fit.
metric_parts <- vector("list", nrow(run_design))
prediction_parts <- vector("list", nrow(run_design))
genomic_value_parts <- vector("list", nrow(run_design))
convergence_parts <- vector("list", nrow(run_design))

for (run_index in seq_len(nrow(run_design))) {

  # Identify the current trait-fold-method combination.
  trait_name <- run_design$Trait[run_index]
  fold_number <- run_design$Fold[run_index]
  method_name <- run_design$Method[run_index]
  run_id <- run_design$Run_id[run_index]
  run_seed <- run_design$Seed[run_index]

  cat(sprintf("[%03d/%03d] %s\n", run_index, nrow(run_design), run_id))

  # Create the directory for the native BGLR chains.
  bglr_run_dir <- file.path(
    BGLR_DIR,
    trait_name,
    paste0("F", fold_number),
    method_name
  )
  dir.create(bglr_run_dir, recursive = TRUE, showWarnings = FALSE)

  save_prefix <- file.path(bglr_run_dir, "bglr_")
  unlink(Sys.glob(paste0(save_prefix, "*")))

  # Select the phenotype vector for the current trait.
  y_observed <- phenotypes[, trait_name]
  names(y_observed) <- ids

  # Identify training and validation individuals.
  validation_index <- which(cv_folds$Fold == fold_number)
  training_index <- which(cv_folds$Fold != fold_number)

  # Mask phenotypes from the validation fold.
  y_model <- y_observed
  y_model[validation_index] <- NA_real_

  # Use the deterministic seed assigned to this fit.
  set.seed(run_seed)

  # Store BGLR warnings for later description.
  run_warning_messages <- character(0)
  fit_start <- proc.time()[["elapsed"]]

  fit <- withCallingHandlers(
    BGLR::BGLR(
      y = y_model,
      ETA = ETA_models[[method_name]],
      response_type = "gaussian",
      nIter = N_ITER,
      burnIn = BURN_IN,
      thin = THIN,
      R2 = R2_PRIOR,
      saveAt = save_prefix,
      verbose = FALSE,
      rmExistingFiles = TRUE
    ),
    warning = function(w) {
      run_warning_messages <<- c(
        run_warning_messages,
        conditionMessage(w)
      )
      invokeRestart("muffleWarning")
    }
  )

  fit_end <- proc.time()[["elapsed"]]

  # Recover predictions for validation individuals.
  observed_validation <- y_observed[validation_index]
  predicted_validation <- fit$yHat[validation_index]

  # Recover the fixed family component.
  fixed_component <- as.vector(
    X_fixed %*% fit$ETA$fixed$b
  )

  # Obtain the genomic component by subtraction.
  genomic_component <- as.vector(
    fit$yHat - fit$mu - fixed_component
  )

  # Store genomic values for all individuals.
  genomic_value_parts[[run_index]] <- data.frame(
    Run_id = run_id,
    IND = ids,
    Trait = trait_name,
    Fold = fold_number,
    Method = method_name,
    Genomic_value = genomic_component
  )

  # Store only the out-of-fold predictions for the current fold.
  prediction_parts[[run_index]] <- data.frame(
    Run_id = run_id,
    IND = ids[validation_index],
    Trait = trait_name,
    Fold = fold_number,
    Method = method_name,
    Observed = observed_validation,
    Predicted = predicted_validation,
    Fixed_component = fixed_component[validation_index],
    Genomic_component = genomic_component[validation_index]
  )

  # Method-specific parameters are recorded when available.
  posterior_prob_in <- c(fit$ETA$genomic$probIn, NA_real_)[1]
  bl_lambda <- c(fit$ETA$genomic$lambda, NA_real_)[1]

  # Calculate predictive and model-fit metrics for the fold.
  metric_parts[[run_index]] <- data.frame(
    Run_id = run_id,
    Trait = trait_name,
    Fold = fold_number,
    Method = method_name,
    Seed = run_seed,
    Training_n = length(training_index),
    Validation_n = length(validation_index),
    Correlation = cor(observed_validation, predicted_validation),
    RMSE = sqrt(mean((observed_validation - predicted_validation)^2)),
    MAE = mean(abs(observed_validation - predicted_validation)),
    Prediction_bias = mean(predicted_validation - observed_validation),
    Regression_slope =
      cov(observed_validation, predicted_validation) /
      var(predicted_validation),
    DIC = fit$fit$DIC,
    Effective_parameters = fit$fit$pD,
    Log_likelihood_at_posterior_mean = fit$fit$logLikAtPostMean,
    Posterior_mean = fit$mu,
    Posterior_mean_SD = c(fit$SD.mu, NA_real_)[1],
    Residual_variance = fit$varE,
    Residual_variance_SD = c(fit$SD.varE, NA_real_)[1],
    BL_lambda = bl_lambda,
    BGLR_warning_count = length(run_warning_messages),
    Tau2_warning_count = sum(grepl(
      "tau2 was not updated",
      run_warning_messages,
      fixed = TRUE
    )),
    Lambda_update_warning_count = sum(grepl(
      "lambda was not updated",
      run_warning_messages,
      fixed = TRUE
    )),
    Posterior_probIn = posterior_prob_in,
    Elapsed_seconds = fit_end - fit_start
  )

  # Read the residual-variance chain.
  residual_trace_full <- scan(
    paste0(save_prefix, "varE.dat"),
    quiet = TRUE
  )

  # Read the scalar chain representing the genomic component.
  genomic_trace_table <- read.table(
    paste0(save_prefix, genomic_trace_suffix[method_name]),
    header = genomic_trace_header[method_name],
    check.names = FALSE
  )

  genomic_trace_full <- genomic_trace_table[
    ,
    genomic_trace_column[method_name]
  ]

  # Remove the saved samples corresponding to burn-in.
  posterior_index <- seq.int(
    BURN_IN_SAVED_SAMPLES + 1L,
    length(residual_trace_full)
  )

  trace_matrix <- cbind(
    Residual_variance = residual_trace_full[posterior_index],
    Genomic_parameter = genomic_trace_full[posterior_index]
  )

  # Convert chains to the coda representation.
  trace_object <- coda::mcmc(
    trace_matrix,
    start = BURN_IN + THIN,
    thin = THIN
  )

  # Calculate effective sample size and the Geweke statistic.
  trace_ess <- coda::effectiveSize(trace_object)
  trace_geweke <- coda::geweke.diag(trace_object)$z

  parameter_names <- c(
    "Residual_variance",
    unname(genomic_trace_parameter[method_name])
  )

  convergence_parts[[run_index]] <- data.frame(
    Run_id = run_id,
    Trait = trait_name,
    Fold = fold_number,
    Method = method_name,
    Parameter = parameter_names,
    Retained_samples = nrow(trace_matrix),
    Posterior_mean = colMeans(trace_matrix),
    Posterior_SD = apply(trace_matrix, 2, sd),
    Effective_sample_size = as.numeric(trace_ess),
    Geweke_Z = as.numeric(trace_geweke)
  )

  # Release the fitted object before the next combination.
  fit <- NULL
  trace_object <- NULL
  gc(verbose = FALSE)
}
```

### 9.1 Assembling and writing the primary outputs

At the end of the loop, the populated lists are combined into the four tables
used by the diagnostics and downstream modules. The block below is the direct
continuation of `fit-models`.

```{r assemble-fit-results, eval=FALSE}
cv_metrics_by_fold <- bind_rows(metric_parts)
cv_predictions <- bind_rows(prediction_parts)
genomic_values_by_fold <- bind_rows(genomic_value_parts)
convergence_diagnostics <- bind_rows(convergence_parts)

write.csv(
  run_design,
  file.path(OUTPUT_DIR, "run_design.csv"),
  row.names = FALSE
)

write.csv(
  cv_predictions,
  file.path(OUTPUT_DIR, "cv_predictions.csv"),
  row.names = FALSE
)

write.csv(
  genomic_values_by_fold,
  file.path(OUTPUT_DIR, "genomic_values_by_fold.csv"),
  row.names = FALSE
)

write.csv(
  cv_metrics_by_fold,
  file.path(OUTPUT_DIR, "cv_metrics_by_fold.csv"),
  row.names = FALSE
)

write.csv(
  convergence_diagnostics,
  file.path(OUTPUT_DIR, "convergence_diagnostics.csv"),
  row.names = FALSE
)
```

## 10. Predictions, genomic values, and fold-level metrics

The tables produced by model fitting are read from `output/models/` for the
diagnostics and summaries presented below.

```{r combine-results}
cv_metrics_by_fold <- read.csv(
  file.path(OUTPUT_DIR, "cv_metrics_by_fold.csv"),
  check.names = FALSE
)

cv_predictions <- read.csv(
  file.path(OUTPUT_DIR, "cv_predictions.csv"),
  check.names = FALSE
)

genomic_values_by_fold <- read.csv(
  file.path(OUTPUT_DIR, "genomic_values_by_fold.csv"),
  check.names = FALSE
)

convergence_diagnostics <- read.csv(
  file.path(OUTPUT_DIR, "convergence_diagnostics.csv"),
  check.names = FALSE
)
```

The total prediction and genomic value are related by

\[
\widehat y_i =
\widehat\mu+
\mathbf x_{Fi}'\widehat{\boldsymbol\beta}_F+
\widehat g_i,
\]

so that

\[
\widehat g_i=
\widehat y_i-
\widehat\mu-
\mathbf x_{Fi}'\widehat{\boldsymbol\beta}_F.
\]

A subset of the OOF predictions allows the output structure to be inspected
directly.

```{r prediction-preview}
cv_predictions |>
  select(
    IND, Trait, Fold, Method,
    Observed, Predicted, Fixed_component, Genomic_component
  ) |>
  head(12) |>
  kable(digits = 4, caption = "Example of out-of-fold predictions")
```

Fold-level metrics include correlation, RMSE, MAE, bias, calibration slope,
DIC, and the effective number of parameters.

```{r metrics-preview}
cv_metrics_by_fold |>
  select(
    Trait, Fold, Method,
    Correlation, RMSE, MAE,
    Prediction_bias, Regression_slope, DIC
  ) |>
  head(14) |>
  kable(digits = 4, caption = "Example of fold-level model metrics")
```

The number of results produced can also be inspected directly.

```{r output-size-summary}
data.frame(
  Output = c(
    "Fits",
    "OOF predictions",
    "Fold-level genomic values",
    "Scalar chains reviewed"
  ),
  Rows = c(
    nrow(cv_metrics_by_fold),
    nrow(cv_predictions),
    nrow(genomic_values_by_fold),
    nrow(convergence_diagnostics)
  )
) |>
  kable(caption = "Dimensions of the main model-fitting outputs")
```

## 11. MCMC diagnostics

For every fit, the residual-variance chain and one scalar chain representative
of the model are examined. For BL, because \(\lambda\) is fixed, the second
diagnostic uses the intercept chain.

Effective sample size (ESS) and the Geweke statistic are used as chain
inspection indicators. ESS values below 1,000 or \(|Z_{Geweke}|>2\) are
highlighted for review but are not automatic model-exclusion rules.

```{r convergence-summary}
convergence_diagnostics <- convergence_diagnostics |>
  mutate(
    Review_flag =
      Effective_sample_size < ESS_REVIEW_THRESHOLD |
      abs(Geweke_Z) > GEWEKE_REVIEW_THRESHOLD
  )

convergence_summary <- convergence_diagnostics |>
  group_by(Method) |>
  summarise(
    Runs = n_distinct(Run_id),
    Parameters_reviewed = n(),
    Parameters_flagged = sum(Review_flag),
    Runs_with_flags = n_distinct(Run_id[Review_flag]),
    Minimum_ESS = min(Effective_sample_size, na.rm = TRUE),
    Median_ESS = median(Effective_sample_size, na.rm = TRUE),
    Maximum_absolute_Geweke = max(abs(Geweke_Z), na.rm = TRUE),
    .groups = "drop"
  )

convergence_summary$Method_label <-
  unname(method_labels[convergence_summary$Method])

convergence_summary |>
  select(
    Method_label, Runs, Parameters_reviewed,
    Parameters_flagged, Runs_with_flags,
    Minimum_ESS, Median_ESS, Maximum_absolute_Geweke
  ) |>
  kable(
    digits = 3,
    caption = "Summary of MCMC diagnostics by method"
  )
```

The chains highlighted by these diagnostics can be inspected directly below.

```{r convergence-flags}
convergence_diagnostics |>
  filter(Review_flag) |>
  arrange(Method, Trait, Fold, Parameter) |>
  select(
    Trait, Fold, Method, Parameter,
    Effective_sample_size, Geweke_Z
  ) |>
  kable(
    digits = 4,
    caption = "Chains highlighted by ESS or Geweke diagnostics"
  )
```

## 12. Saving diagnostics and the final object

After the diagnostics, the convergence table is updated with the review
indicators, the method-level summary is written, and the final object is
assembled for use by the following modules.

```{r save-results, eval=FALSE}
analysis_settings <- list(
  n_iter = N_ITER,
  burn_in = BURN_IN,
  thin = THIN,
  R2 = R2_PRIOR,
  seed_base = SEED_BASE,
  bayesb2_zero_probability = BAYESB2_ZERO_PROBABILITY,
  bayesb2_prob_in = BAYESB2_PROB_IN,
  bayesb2_counts = BAYESB2_COUNTS,
  bl_lambda_type = BL_LAMBDA_TYPE,
  bl_n_linear_terms = BL_N_LINEAR_TERMS,
  bl_component_R2 = BL_COMPONENT_R2,
  bl_msx = BL_MSX,
  bl_lambda_squared = BL_LAMBDA_SQUARED,
  bl_lambda = BL_LAMBDA,
  traits = traits,
  folds = seq_len(matrices_object$settings$n_folds),
  methods = method_names
)

model_catalogue <- data.frame(
  Method = method_names,
  Label = c(
    "RR-BLUP", "BayesA", "BayesB", "BayesB2",
    "BayesC", "Bayesian Lasso", "GBLUP"
  ),
  BGLR_component = c(
    "BRR", "BayesA", "BayesB", "BayesB",
    "BayesC", "BL", "RKHS"
  ),
  Genomic_input = c("M", "M", "M", "M", "M", "M", "G"),
  stringsAsFactors = FALSE
)

write.csv(
  convergence_diagnostics,
  file.path(OUTPUT_DIR, "convergence_diagnostics.csv"),
  row.names = FALSE
)

write.csv(
  convergence_summary,
  file.path(OUTPUT_DIR, "convergence_summary.csv"),
  row.names = FALSE
)

saveRDS(
  list(
    settings = analysis_settings,
    run_design = run_design,
    model_catalogue = model_catalogue,
    trait_dictionary = trait_dictionary,
    predictions = cv_predictions,
    genomic_values_by_fold = genomic_values_by_fold,
    metrics_by_fold = cv_metrics_by_fold,
    convergence_diagnostics = convergence_diagnostics
  ),
  file.path(OUTPUT_DIR, "models_object.rds")
)
```

## 13. Next step

Module 04 uses the OOF predictions, fold-level metrics, genomic values, and MCMC
diagnostics obtained here to compare methods and derive downstream genomic
results.
