---
title: "Complete fitting of seven genomic prediction methods in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

This module performs the complete cross-validated fitting of seven genomic
prediction methods for ten papaya traits. The same five deterministic,
family-stratified folds are used for every method and trait.

The complete analysis contains:

\[
10\ \text{traits}
\times
5\ \text{folds}
\times
7\ \text{methods}
=
350\ \text{fits}.
\]

All methods are fitted exclusively with BGLR [@perez2014bglr]:

1. RR-BLUP, implemented with the BGLR `BRR` component and the complete marker
   matrix `M`;
2. BayesA with `M`;
3. BayesB with `M`;
4. BayesB2 with `M` and a zero-effect probability operationally fixed near
   \(10^{-5}\);
5. BayesC with `M`;
6. Bayesian Lasso with `M`;
7. GBLUP with the genomic relationship matrix `G` and the BGLR `RKHS`
   component.

RR-BLUP and Bayesian ridge regression are not treated as separate methods. The
Gaussian marker-effect model implemented by `model = "BRR"` is the RR-BLUP
implementation used in this project.

## 2. Analytical contract

The matrices constructed in Module 02 are used without fold-specific
reconstruction:

- `X` contains intercept, block, and family effects;
- `M` contains the 72 centered allele-dosage columns;
- `G` contains the additive genomic relationships;
- each fold identifies 30 validation individuals and 120 training individuals.

BGLR includes an intercept by default. Therefore, the intercept column is
removed from `X` before the block and family columns are passed through the
`FIXED` component.

For each trait and fold, only a copy of the phenotype vector is changed. The
30 validation observations are replaced by `NA`; `X`, `M`, and `G` remain
global and unchanged.

The MCMC settings follow the main methods section of the reference article
[@silva2021brrguava]:

- 1,000,000 iterations;
- 200,000 burn-in iterations;
- thinning equal to 4;
- 200,000 retained posterior samples.

The BGLR default prior allocation parameter is made explicit with `R2 = 0.5`.

## 3. Reproducible and resumable execution

Each trait-fold-method combination has:

- a deterministic seed;
- a unique run identifier;
- an individual compact RDS result;
- temporary BGLR chain files;
- a row in the run manifest.

A run is considered complete only when its compact RDS result exists. When the
module is rebuilt, completed runs are skipped and only missing runs are fitted.
This provides deterministic checkpointing without rebuilding matrices or
silently changing the model contract.

Temporary BGLR files are read immediately after each fit. BGLR writes
thinned samples during both burn-in and posterior sampling, so the saved burn-in
portion is removed before the single-chain diagnostics are calculated. The raw
files are then deleted. Only compact posterior summaries, predictions, genomic
values, effects, diagnostics, and metrics are retained.

## 4. Packages and complete-analysis settings

```{r packages}
suppressPackageStartupMessages({
  library(BGLR)
  library(coda)
  library(digest)
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(here)
})
```

```{r complete-settings}
MATRICES_FILE <- here(
  "output",
  "matrices",
  "matrices_object.rds"
)

OUTPUT_DIR <- here(
  "output",
  "models",
  "full"
)

RUNS_DIR <- file.path(
  OUTPUT_DIR,
  "runs"
)

TEMP_DIR <- file.path(
  OUTPUT_DIR,
  "temporary"
)

N_ITER <- 1000000L
BURN_IN <- 200000L
THIN <- 4L
R2_PRIOR <- 0.5

RAW_SAVED_SAMPLES <- floor(
  N_ITER / THIN
)

BURN_IN_SAVED_SAMPLES <- floor(
  BURN_IN / THIN
)

RETAINED_SAMPLES <-
  RAW_SAVED_SAMPLES -
  BURN_IN_SAVED_SAMPLES

PIPELINE_VERSION <-
  "module03-full-v3"

MAX_NEW_RUNS_PER_BUILD <- as.integer(
  Sys.getenv(
    "PAPAYA_MAX_NEW_RUNS",
    unset = "350"
  )
)

SEED_BASE <- 202608050L

BAYESB2_ZERO_PROBABILITY <- 1e-5
BAYESB2_PROB_IN <-
  1 - BAYESB2_ZERO_PROBABILITY
BAYESB2_COUNTS <- 1e6

ESS_REVIEW_THRESHOLD <- 1000
GEWEKE_REVIEW_THRESHOLD <- 2

dir.create(
  RUNS_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  TEMP_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)
```

## 5. Input objects and invariant matrices

```{r input-objects}
matrices_object <- readRDS(
  MATRICES_FILE
)

ids <- matrices_object$ids
metadata <- matrices_object$metadata
phenotypes <- matrices_object$phenotypes
trait_dictionary <- matrices_object$trait_dictionary
X <- matrices_object$X
M <- matrices_object$M
G <- matrices_object$G
cv_folds <- matrices_object$cv_folds

traits <- colnames(
  phenotypes
)

X_fixed <- X[
  ,
  colnames(X) != "(Intercept)",
  drop = FALSE
]

method_names <- c(
  "RRBLUP",
  "BayesA",
  "BayesB",
  "BayesB2",
  "BayesC",
  "BL",
  "GBLUP"
)

method_labels <- c(
  RRBLUP = "RR-BLUP",
  BayesA = "BayesA",
  BayesB = "BayesB",
  BayesB2 = "BayesB2",
  BayesC = "BayesC",
  BL = "Bayesian Lasso",
  GBLUP = "GBLUP"
)
```

```{r input-audit}
input_audit <- data.frame(
  Check = c(
    "Individuals",
    "Traits",
    "Folds",
    "Methods",
    "Planned fits",
    "Validation individuals per fold",
    "Training individuals per fold",
    "Fixed-effect columns supplied to BGLR",
    "Marker columns",
    "Rank of M",
    "Rows and columns of G",
    "Rank of G",
    "X row names aligned with IDs",
    "M row names aligned with IDs",
    "G row names aligned with IDs",
    "G column names aligned with IDs"
  ),
  Value = c(
    length(ids),
    length(traits),
    matrices_object$settings$n_folds,
    length(method_names),
    length(traits) *
      matrices_object$settings$n_folds *
      length(method_names),
    unique(
      matrices_object$fold_summary$
        Validation_n
    ),
    unique(
      matrices_object$fold_summary$
        Training_n
    ),
    ncol(X_fixed),
    ncol(M),
    qr(M)$rank,
    paste(dim(G), collapse = " x "),
    qr(G)$rank,
    identical(rownames(X), ids),
    identical(rownames(M), ids),
    identical(rownames(G), ids),
    identical(colnames(G), ids)
  )
)

kable(
  input_audit,
  caption = "Invariant input audit for the complete model fitting"
)
```

## 6. Model catalogue and BGLR specifications

```{r model-catalogue}
model_catalogue <- data.frame(
  Method = method_names,
  Label = unname(
    method_labels[
      method_names
    ]
  ),
  BGLR_component = c(
    "BRR",
    "BayesA",
    "BayesB",
    "BayesB",
    "BayesC",
    "BL",
    "RKHS"
  ),
  Genomic_input = c(
    "M",
    "M",
    "M",
    "M",
    "M",
    "M",
    "G"
  ),
  Main_assumption = c(
    "Common Gaussian variance for all marker effects",
    "Marker-specific scaled-t variances",
    "Scaled-t slab and a point mass at zero",
    "BayesB with zero-effect probability near 1e-5",
    "Common Gaussian slab and a point mass at zero",
    "Double-exponential marginal prior",
    "Genomic values structured by G"
  )
)

kable(
  model_catalogue,
  caption = "Seven genomic prediction methods"
)
```

```{r eta-specifications}
ETA_models <- list(
  RRBLUP = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BRR"
    )
  ),
  BayesA = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BayesA"
    )
  ),
  BayesB = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BayesB"
    )
  ),
  BayesB2 = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BayesB",
      probIn = BAYESB2_PROB_IN,
      counts = BAYESB2_COUNTS
    )
  ),
  BayesC = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BayesC"
    )
  ),
  BL = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      X = M,
      model = "BL"
    )
  ),
  GBLUP = list(
    fixed = list(
      X = X_fixed,
      model = "FIXED"
    ),
    genomic = list(
      K = G,
      model = "RKHS"
    )
  )
)
```

The official BGLR implementation writes a residual-variance chain for every
model. The genomic scalar used for convergence review differs by method:
marker variance for RR-BLUP, scale for BayesA and BayesB, common marker
variance for BayesC, shrinkage parameter for the Bayesian Lasso, and genomic
variance for GBLUP.

```{r trace-contract}
genomic_trace_suffix <- c(
  RRBLUP = "ETA_genomic_varB.dat",
  BayesA = "ETA_genomic_ScaleBayesA.dat",
  BayesB = "ETA_genomic_parBayesB.dat",
  BayesB2 = "ETA_genomic_parBayesB.dat",
  BayesC = "ETA_genomic_parBayesC.dat",
  BL = "ETA_genomic_lambda.dat",
  GBLUP = "ETA_genomic_varU.dat"
)

genomic_trace_header <- c(
  RRBLUP = FALSE,
  BayesA = FALSE,
  BayesB = TRUE,
  BayesB2 = TRUE,
  BayesC = TRUE,
  BL = FALSE,
  GBLUP = FALSE
)

genomic_trace_column <- c(
  RRBLUP = 1L,
  BayesA = 1L,
  BayesB = 2L,
  BayesB2 = 2L,
  BayesC = 2L,
  BL = 1L,
  GBLUP = 1L
)

genomic_trace_parameter <- c(
  RRBLUP = "Marker_variance",
  BayesA = "Scale",
  BayesB = "Scale",
  BayesB2 = "Scale",
  BayesC = "Marker_variance",
  BL = "Lambda",
  GBLUP = "Genomic_variance"
)
```

## 7. Run design and analytical signature

```{r run-design}
run_design <- expand.grid(
  Trait = traits,
  Fold = seq_len(
    matrices_object$settings$n_folds
  ),
  Method = method_names,
  stringsAsFactors = FALSE
)

run_design <- run_design[
  order(
    match(
      run_design$Trait,
      traits
    ),
    run_design$Fold,
    match(
      run_design$Method,
      method_names
    )
  ),
]

rownames(run_design) <- NULL

run_design$Run_number <- seq_len(
  nrow(run_design)
)

run_design$Seed <-
  SEED_BASE +
  run_design$Run_number

analysis_settings <- list(
  pipeline_version =
    PIPELINE_VERSION,
  n_iter = N_ITER,
  burn_in = BURN_IN,
  thin = THIN,
  R2 = R2_PRIOR,
  seed_base = SEED_BASE,
  bayesb2_zero_probability =
    BAYESB2_ZERO_PROBABILITY,
  bayesb2_prob_in =
    BAYESB2_PROB_IN,
  bayesb2_counts =
    BAYESB2_COUNTS,
  traits = traits,
  folds = seq_len(
    matrices_object$settings$n_folds
  ),
  methods = method_names
)

analysis_signature <- substr(
  digest::digest(
    list(
      ids = ids,
      phenotypes = phenotypes,
      X = X,
      M = M,
      G = G,
      cv_folds = cv_folds,
      settings = analysis_settings
    ),
    algo = "sha256"
  ),
  1,
  16
)

run_design$Signature <-
  analysis_signature

run_design$Run_id <- paste(
  run_design$Trait,
  paste0(
    "F",
    run_design$Fold
  ),
  run_design$Method,
  sep = "_"
)

run_design$Result_relative_file <- file.path(
  "output",
  "models",
  "full",
  "runs",
  paste0(
    run_design$Run_id,
    "_",
    analysis_signature,
    ".rds"
  )
)

result_files <- file.path(
  here(),
  run_design$Result_relative_file
)

run_design$Completed_before_build <-
  file.exists(
    result_files
  )

write.csv(
  run_design,
  file.path(
    OUTPUT_DIR,
    "run_design.csv"
  ),
  row.names = FALSE
)

kable(
  head(
    run_design,
    14
  ),
  caption = paste(
    "First fourteen planned runs; analytical signature",
    analysis_signature
  )
)
```

## 8. Complete fitting with checkpointing

The loop below fits only runs whose compact result file is absent. A completed
run is never recomputed by an ordinary rebuild of this module. The optional
`PAPAYA_MAX_NEW_RUNS` environment variable limits how many new fits are
performed in one build. Its default value is 350, so an ordinary build fits
all remaining runs. Setting it to 7 validates one complete trait-fold block
containing all methods before the remaining analysis is launched.

The temporary chains retained for diagnostic calculation are:

- residual variance;
- one method-specific genomic scale or variance parameter.

For each chain, effective sample size and the Geweke diagnostic are calculated
with `coda`. The thresholds below are review flags rather than automatic model
exclusion rules:

- ESS below 1,000;
- absolute Geweke statistic above 2.

```{r complete-fitting, results="hide"}
pending_indices <- head(
  which(
    !file.exists(
      result_files
    )
  ),
  n = MAX_NEW_RUNS_PER_BUILD
)

for (run_index in pending_indices) {
  trait_name <-
    run_design$Trait[
      run_index
    ]

  fold_number <-
    run_design$Fold[
      run_index
    ]

  method_name <-
    run_design$Method[
      run_index
    ]

  run_id <-
    run_design$Run_id[
      run_index
    ]

  run_seed <-
    run_design$Seed[
      run_index
    ]

  result_file <-
    result_files[
      run_index
    ]

  temporary_result_file <- paste0(
    result_file,
    ".tmp"
  )

  save_prefix <- file.path(
    TEMP_DIR,
    paste0(
      run_id,
      "_",
      analysis_signature,
      "_"
    )
  )

  unlink(
    Sys.glob(
      paste0(
        save_prefix,
        "*"
      )
    )
  )

  y_observed <- phenotypes[
    ,
    trait_name
  ]

  names(y_observed) <- ids

  validation_index <- which(
    cv_folds$Fold == fold_number
  )

  training_index <- which(
    cv_folds$Fold != fold_number
  )

  y_model <- y_observed
  y_model[validation_index] <- NA_real_

  set.seed(
    run_seed
  )

  fit_start <- proc.time()[["elapsed"]]

  fit <- BGLR::BGLR(
    y = y_model,
    ETA = ETA_models[[method_name]],
    response_type = "gaussian",
    nIter = N_ITER,
    burnIn = BURN_IN,
    thin = THIN,
    R2 = R2_PRIOR,
    saveAt = save_prefix,
    verbose = FALSE,
    rmExistingFiles = TRUE
  )

  fit_end <- proc.time()[["elapsed"]]

  predicted_validation <- fit$yHat[
    validation_index
  ]

  observed_validation <- y_observed[
    validation_index
  ]

  fixed_component <- as.vector(
    X_fixed %*%
      fit$ETA$fixed$b
  )

  genomic_component <- as.vector(
    fit$yHat -
      fit$mu -
      fixed_component
  )

  genomic_value_table <- data.frame(
    Run_id = rep(
      run_id,
      length(ids)
    ),
    Signature = rep(
      analysis_signature,
      length(ids)
    ),
    IND = ids,
    Trait = rep(
      trait_name,
      length(ids)
    ),
    Fold = rep(
      fold_number,
      length(ids)
    ),
    Method = rep(
      method_name,
      length(ids)
    ),
    Genomic_value =
      genomic_component
  )

  prediction_table <- data.frame(
    Run_id = run_id,
    Signature = analysis_signature,
    IND = ids[
      validation_index
    ],
    Trait = trait_name,
    Fold = fold_number,
    Method = method_name,
    Observed = observed_validation,
    Predicted = predicted_validation,
    Fixed_component =
      fixed_component[
        validation_index
      ],
    Genomic_component =
      genomic_component[
        validation_index
      ]
  )

  posterior_prob_in <- c(
    fit$ETA$genomic$probIn,
    NA_real_
  )[1]

  metric_table <- data.frame(
    Run_id = run_id,
    Signature = analysis_signature,
    Trait = trait_name,
    Fold = fold_number,
    Method = method_name,
    Seed = run_seed,
    Training_n = length(
      training_index
    ),
    Validation_n = length(
      validation_index
    ),
    Correlation = cor(
      observed_validation,
      predicted_validation
    ),
    RMSE = sqrt(
      mean(
        (
          observed_validation -
            predicted_validation
        )^2
      )
    ),
    MAE = mean(
      abs(
        observed_validation -
          predicted_validation
      )
    ),
    Prediction_bias = mean(
      predicted_validation -
        observed_validation
    ),
    Regression_slope =
      cov(
        observed_validation,
        predicted_validation
      ) /
      var(
        predicted_validation
      ),
    DIC = fit$fit$DIC,
    Effective_parameters =
      fit$fit$pD,
    Log_likelihood_at_posterior_mean =
      fit$fit$
        logLikAtPostMean,
    Posterior_mean =
      fit$mu,
    Posterior_mean_SD = c(
      fit$SD.mu,
      NA_real_
    )[1],
    Residual_variance =
      fit$varE,
    Residual_variance_SD = c(
      fit$SD.varE,
      NA_real_
    )[1],
    Posterior_probIn =
      posterior_prob_in,
    Elapsed_seconds =
      fit_end - fit_start,
    Finite_predictions = all(
      is.finite(
        predicted_validation
      )
    )
  )

  metric_table$Finite_metrics <- all(
    is.finite(
      as.numeric(
        metric_table[
          ,
          c(
            "Correlation",
            "RMSE",
            "MAE",
            "Prediction_bias",
            "Regression_slope",
            "DIC",
            "Effective_parameters",
            "Log_likelihood_at_posterior_mean",
            "Posterior_mean",
            "Residual_variance",
            "Elapsed_seconds"
          )
        ]
      )
    )
  )

  fixed_effect_table <- data.frame(
    Run_id = rep(
      run_id,
      length(
        fit$ETA$fixed$b
      )
    ),
    Signature = rep(
      analysis_signature,
      length(
        fit$ETA$fixed$b
      )
    ),
    Trait = rep(
      trait_name,
      length(
        fit$ETA$fixed$b
      )
    ),
    Fold = rep(
      fold_number,
      length(
        fit$ETA$fixed$b
      )
    ),
    Method = rep(
      method_name,
      length(
        fit$ETA$fixed$b
      )
    ),
    Effect = names(
      fit$ETA$fixed$b
    ),
    Posterior_mean = as.numeric(
      fit$ETA$fixed$b
    ),
    Posterior_SD = as.numeric(
      fit$ETA$fixed$SD.b
    )
  )

  marker_count <-
    as.integer(
      method_name != "GBLUP"
    ) *
    ncol(M)

  marker_indices <- seq_len(
    marker_count
  )

  marker_effect_values <-
    c(
      fit$ETA$genomic$b
    )[
      marker_indices
    ]

  marker_effect_sds <-
    c(
      fit$ETA$genomic$SD.b
    )[
      marker_indices
    ]

  marker_effect_table <- data.frame(
    Run_id = rep(
      run_id,
      marker_count
    ),
    Signature = rep(
      analysis_signature,
      marker_count
    ),
    Trait = rep(
      trait_name,
      marker_count
    ),
    Fold = rep(
      fold_number,
      marker_count
    ),
    Method = rep(
      method_name,
      marker_count
    ),
    Marker = colnames(M)[
      marker_indices
    ],
    Posterior_mean = as.numeric(
      marker_effect_values
    ),
    Posterior_SD = as.numeric(
      marker_effect_sds
    )
  )

  residual_trace_full <- scan(
    paste0(
      save_prefix,
      "varE.dat"
    ),
    quiet = TRUE
  )

  genomic_trace_table_full <- read.table(
    paste0(
      save_prefix,
      genomic_trace_suffix[
        method_name
      ]
    ),
    header =
      genomic_trace_header[
        method_name
      ],
    check.names = FALSE
  )

  genomic_trace_full <-
    genomic_trace_table_full[
      ,
      genomic_trace_column[
        method_name
      ]
    ]

  posterior_sample_indices <- seq.int(
    from =
      BURN_IN_SAVED_SAMPLES +
      1L,
    to = length(
      residual_trace_full
    )
  )

  residual_trace <-
    residual_trace_full[
      posterior_sample_indices
    ]

  genomic_trace <-
    genomic_trace_full[
      posterior_sample_indices
    ]

  trace_matrix <- cbind(
    Residual_variance =
      residual_trace,
    Genomic_parameter =
      genomic_trace
  )

  trace_object <- coda::mcmc(
    trace_matrix,
    start = BURN_IN + THIN,
    thin = THIN
  )

  trace_ess <- coda::effectiveSize(
    trace_object
  )

  trace_geweke <- coda::geweke.diag(
    trace_object
  )$z

  trace_parameter_names <- c(
    "Residual_variance",
    unname(
      genomic_trace_parameter[
        method_name
      ]
    )
  )

  convergence_table <- data.frame(
    Run_id = rep(
      run_id,
      length(
        trace_parameter_names
      )
    ),
    Signature = rep(
      analysis_signature,
      length(
        trace_parameter_names
      )
    ),
    Trait = rep(
      trait_name,
      length(
        trace_parameter_names
      )
    ),
    Fold = rep(
      fold_number,
      length(
        trace_parameter_names
      )
    ),
    Method = rep(
      method_name,
      length(
        trace_parameter_names
      )
    ),
    Parameter =
      trace_parameter_names,
    Retained_samples = rep(
      nrow(
        trace_matrix
      ),
      length(
        trace_parameter_names
      )
    ),
    Posterior_mean = colMeans(
      trace_matrix
    ),
    Posterior_SD = apply(
      trace_matrix,
      2,
      sd
    ),
    Effective_sample_size = as.numeric(
      trace_ess
    ),
    Geweke_Z = as.numeric(
      trace_geweke
    )
  )

  convergence_table$Review_flag <-
    !is.finite(
      convergence_table$
        Effective_sample_size
    ) |
    convergence_table$
      Effective_sample_size <
      ESS_REVIEW_THRESHOLD |
    !is.finite(
      convergence_table$
        Geweke_Z
    ) |
    abs(
      convergence_table$
        Geweke_Z
    ) >
      GEWEKE_REVIEW_THRESHOLD

  run_audit <- data.frame(
    Run_id = run_id,
    Signature = analysis_signature,
    Trait = trait_name,
    Fold = fold_number,
    Method = method_name,
    Seed = run_seed,
    Expected_raw_saved_samples =
      RAW_SAVED_SAMPLES,
    Observed_raw_saved_samples =
      length(
        residual_trace_full
      ),
    Expected_retained_samples =
      RETAINED_SAMPLES,
    Observed_retained_samples =
      length(
        residual_trace
      ),
    Finite_predictions = all(
      is.finite(
        predicted_validation
      )
    ),
    Minimum_effective_sample_size =
      min(
        convergence_table$
          Effective_sample_size,
        na.rm = TRUE
      ),
    Maximum_absolute_Geweke =
      max(
        abs(
          convergence_table$
            Geweke_Z
        ),
        na.rm = TRUE
      ),
    Parameters_flagged_for_review =
      sum(
        convergence_table$
          Review_flag
      ),
    Elapsed_seconds =
      fit_end - fit_start
  )

  compact_result <- list(
    run = run_design[
      run_index,
      ,
      drop = FALSE
    ],
    metrics = metric_table,
    predictions = prediction_table,
    genomic_values =
      genomic_value_table,
    fixed_effects = fixed_effect_table,
    marker_effects = marker_effect_table,
    convergence = convergence_table,
    run_audit = run_audit,
    settings = analysis_settings
  )

  saveRDS(
    compact_result,
    temporary_result_file
  )

  file.rename(
    temporary_result_file,
    result_file
  )

  progress_table <- run_design
  progress_table$Completed <-
    file.exists(
      result_files
    )

  write.csv(
    progress_table,
    file.path(
      OUTPUT_DIR,
      "run_progress.csv"
    ),
    row.names = FALSE
  )

  progress_summary <- data.frame(
    Signature =
      analysis_signature,
    Completed_runs = sum(
      progress_table$Completed
    ),
    Total_runs = nrow(
      progress_table
    ),
    Remaining_runs = sum(
      !progress_table$Completed
    ),
    Last_completed_run =
      run_id,
    Updated_at = format(
      Sys.time(),
      "%Y-%m-%d %H:%M:%S %Z"
    )
  )

  write.csv(
    progress_summary,
    file.path(
      OUTPUT_DIR,
      "progress_summary.csv"
    ),
    row.names = FALSE
  )

  unlink(
    Sys.glob(
      paste0(
        save_prefix,
        "*"
      )
    )
  )

  fit <- NULL
  compact_result <- NULL
  trace_object <- NULL

  gc(
    verbose = FALSE
  )
}
```

## 9. Consolidation of the 350 compact results

```{r consolidate-results}
completed_after_build <- file.exists(
  result_files
)

completed_indices <- which(
  completed_after_build
)

result_parts <- vector(
  mode = "list",
  length = length(
    completed_indices
  )
)

metric_parts <- vector(
  mode = "list",
  length = length(
    completed_indices
  )
)

prediction_parts <- vector(
  mode = "list",
  length = length(
    completed_indices
  )
)

genomic_value_parts <- vector(
  mode = "list",
  length = length(
    completed_indices
  )
)

fixed_effect_parts <- vector(
  mode = "list",
  length = length(
    completed_indices
  )
)

marker_effect_parts <- vector(
  mode = "list",
  length = length(
    completed_indices
  )
)

convergence_parts <- vector(
  mode = "list",
  length = length(
    completed_indices
  )
)

run_audit_parts <- vector(
  mode = "list",
  length = length(
    completed_indices
  )
)

for (part_index in seq_along(
  completed_indices
)) {
  run_index <- completed_indices[
    part_index
  ]

  result_parts[[part_index]] <- readRDS(
    result_files[
      run_index
    ]
  )

  metric_parts[[part_index]] <-
    result_parts[[part_index]]$
      metrics

  prediction_parts[[part_index]] <-
    result_parts[[part_index]]$
      predictions

  genomic_value_parts[[part_index]] <-
    result_parts[[part_index]]$
      genomic_values

  fixed_effect_parts[[part_index]] <-
    result_parts[[part_index]]$
      fixed_effects

  marker_effect_parts[[part_index]] <-
    result_parts[[part_index]]$
      marker_effects

  convergence_parts[[part_index]] <-
    result_parts[[part_index]]$
      convergence

  run_audit_parts[[part_index]] <-
    result_parts[[part_index]]$
      run_audit
}

cv_metrics_by_fold <- bind_rows(
  metric_parts
)

cv_predictions <- bind_rows(
  prediction_parts
)

genomic_values_by_fold <- bind_rows(
  genomic_value_parts
)

fixed_effects_by_fold <- bind_rows(
  fixed_effect_parts
)

marker_effects_by_fold <- bind_rows(
  marker_effect_parts
)

convergence_diagnostics <- bind_rows(
  convergence_parts
)

run_audit <- bind_rows(
  run_audit_parts
)

cv_fold_summary <- cv_metrics_by_fold |>
  group_by(
    Trait,
    Method
  ) |>
  summarise(
    Mean_fold_correlation = mean(
      Correlation
    ),
    SD_fold_correlation = sd(
      Correlation
    ),
    Mean_fold_RMSE = mean(
      RMSE
    ),
    SD_fold_RMSE = sd(
      RMSE
    ),
    Mean_fold_MAE = mean(
      MAE
    ),
    SD_fold_MAE = sd(
      MAE
    ),
    Mean_fold_bias = mean(
      Prediction_bias
    ),
    Mean_fold_slope = mean(
      Regression_slope
    ),
    Mean_DIC = mean(
      DIC
    ),
    Mean_elapsed_seconds = mean(
      Elapsed_seconds
    ),
    .groups = "drop"
  )

cv_pooled_summary <- cv_predictions |>
  group_by(
    Trait,
    Method
  ) |>
  summarise(
    Validation_n = n(),
    Correlation = cor(
      Observed,
      Predicted
    ),
    RMSE = sqrt(
      mean(
        (
          Observed -
            Predicted
        )^2
      )
    ),
    MAE = mean(
      abs(
        Observed -
          Predicted
      )
    ),
    Prediction_bias = mean(
      Predicted -
        Observed
    ),
    Regression_slope =
      cov(
        Observed,
        Predicted
      ) /
      var(
        Predicted
      ),
    .groups = "drop"
  )

cv_metrics_summary <- left_join(
  cv_pooled_summary,
  cv_fold_summary,
  by = c(
    "Trait",
    "Method"
  )
)

cv_metrics_summary <- left_join(
  cv_metrics_summary,
  trait_dictionary,
  by = "Trait"
)

cv_metrics_summary$Method_label <-
  unname(
    method_labels[
      cv_metrics_summary$Method
    ]
  )

convergence_summary <- convergence_diagnostics |>
  group_by(
    Method
  ) |>
  summarise(
    Runs = n_distinct(
      Run_id
    ),
    Parameters_reviewed = n(),
    Parameters_flagged = sum(
      Review_flag
    ),
    Runs_with_flags = n_distinct(
      Run_id[
        Review_flag
      ]
    ),
    Minimum_ESS = min(
      Effective_sample_size,
      na.rm = TRUE
    ),
    Median_ESS = median(
      Effective_sample_size,
      na.rm = TRUE
    ),
    Maximum_absolute_Geweke = max(
      abs(
        Geweke_Z
      ),
      na.rm = TRUE
    ),
    .groups = "drop"
  )

convergence_summary$Method_label <-
  unname(
    method_labels[
      convergence_summary$Method
    ]
  )

runtime_summary <- cv_metrics_by_fold |>
  group_by(
    Method
  ) |>
  summarise(
    Fits = n(),
    Total_seconds = sum(
      Elapsed_seconds
    ),
    Mean_seconds = mean(
      Elapsed_seconds
    ),
    Median_seconds = median(
      Elapsed_seconds
    ),
    Maximum_seconds = max(
      Elapsed_seconds
    ),
    .groups = "drop"
  )

runtime_summary$Method_label <-
  unname(
    method_labels[
      runtime_summary$Method
    ]
  )

bayesb2_audit <- cv_metrics_by_fold |>
  filter(
    Method == "BayesB2"
  ) |>
  summarise(
    Runs = n(),
    Target_probIn =
      BAYESB2_PROB_IN,
    Minimum_posterior_probIn = min(
      Posterior_probIn
    ),
    Maximum_posterior_probIn = max(
      Posterior_probIn
    ),
    Mean_posterior_probIn = mean(
      Posterior_probIn
    ),
    Maximum_absolute_deviation = max(
      abs(
        Posterior_probIn -
          BAYESB2_PROB_IN
      )
    )
  )

rrblup_predictions <- cv_predictions |>
  filter(
    Method == "RRBLUP"
  ) |>
  select(
    IND,
    Trait,
    Fold,
    RRBLUP_prediction = Predicted
  )

gblup_predictions <- cv_predictions |>
  filter(
    Method == "GBLUP"
  ) |>
  select(
    IND,
    Trait,
    Fold,
    GBLUP_prediction = Predicted
  )

rrblup_gblup_audit <- inner_join(
  rrblup_predictions,
  gblup_predictions,
  by = c(
    "IND",
    "Trait",
    "Fold"
  )
) |>
  group_by(
    Trait,
    Fold
  ) |>
  summarise(
    Prediction_correlation = cor(
      RRBLUP_prediction,
      GBLUP_prediction
    ),
    Maximum_absolute_difference = max(
      abs(
        RRBLUP_prediction -
          GBLUP_prediction
      )
    ),
    Mean_absolute_difference = mean(
      abs(
        RRBLUP_prediction -
          GBLUP_prediction
      )
    ),
    .groups = "drop"
  )

completion_audit <- data.frame(
  Check = c(
    "Analytical signature",
    "Planned fits",
    "Completed fits",
    "Missing fits",
    "Expected prediction rows",
    "Observed prediction rows",
    "Expected genomic-value rows",
    "Observed genomic-value rows",
    "Expected validation predictions per trait and method",
    "Observed validation predictions per trait and method",
    "All predictions finite",
    "All primary metrics finite",
    "Expected raw saved samples per run",
    "Minimum raw saved samples observed",
    "Expected retained samples per run",
    "Minimum retained samples observed"
  ),
  Expected = c(
    analysis_signature,
    nrow(
      run_design
    ),
    nrow(
      run_design
    ),
    0,
    nrow(
      run_design
    ) * 30,
    nrow(
      run_design
    ) * 30,
    nrow(
      run_design
    ) *
      length(ids),
    nrow(
      run_design
    ) *
      length(ids),
    length(ids),
    length(ids),
    TRUE,
    TRUE,
    RAW_SAVED_SAMPLES,
    RAW_SAVED_SAMPLES,
    RETAINED_SAMPLES,
    RETAINED_SAMPLES
  ),
  Observed = c(
    analysis_signature,
    nrow(
      run_design
    ),
    sum(
      completed_after_build
    ),
    sum(
      !completed_after_build
    ),
    nrow(
      run_design
    ) * 30,
    nrow(
      cv_predictions
    ),
    nrow(
      run_design
    ) *
      length(ids),
    nrow(
      genomic_values_by_fold
    ),
    length(ids),
    paste(
      range(
        table(
          cv_predictions$Trait,
          cv_predictions$Method
        )
      ),
      collapse = " to "
    ),
    all(
      is.finite(
        cv_predictions$Predicted
      )
    ),
    all(
      cv_metrics_by_fold$
        Finite_metrics
    ),
    RAW_SAVED_SAMPLES,
    min(
      run_audit$
        Observed_raw_saved_samples
    ),
    RETAINED_SAMPLES,
    min(
      run_audit$
        Observed_retained_samples
    )
  )
)
```

## 10. Saved complete-analysis outputs

```{r save-complete-outputs}
write.csv(
  cv_predictions,
  file.path(
    OUTPUT_DIR,
    "cv_predictions.csv"
  ),
  row.names = FALSE
)

write.csv(
  genomic_values_by_fold,
  file.path(
    OUTPUT_DIR,
    "genomic_values_by_fold.csv"
  ),
  row.names = FALSE
)

write.csv(
  cv_metrics_by_fold,
  file.path(
    OUTPUT_DIR,
    "cv_metrics_by_fold.csv"
  ),
  row.names = FALSE
)

write.csv(
  cv_metrics_summary,
  file.path(
    OUTPUT_DIR,
    "cv_metrics_summary.csv"
  ),
  row.names = FALSE
)

write.csv(
  fixed_effects_by_fold,
  file.path(
    OUTPUT_DIR,
    "fixed_effects_by_fold.csv"
  ),
  row.names = FALSE
)

write.csv(
  marker_effects_by_fold,
  file.path(
    OUTPUT_DIR,
    "marker_effects_by_fold.csv"
  ),
  row.names = FALSE
)

write.csv(
  convergence_diagnostics,
  file.path(
    OUTPUT_DIR,
    "convergence_diagnostics.csv"
  ),
  row.names = FALSE
)

write.csv(
  convergence_summary,
  file.path(
    OUTPUT_DIR,
    "convergence_summary.csv"
  ),
  row.names = FALSE
)

write.csv(
  runtime_summary,
  file.path(
    OUTPUT_DIR,
    "runtime_summary.csv"
  ),
  row.names = FALSE
)

write.csv(
  bayesb2_audit,
  file.path(
    OUTPUT_DIR,
    "bayesb2_probability_audit.csv"
  ),
  row.names = FALSE
)

write.csv(
  rrblup_gblup_audit,
  file.path(
    OUTPUT_DIR,
    "rrblup_gblup_audit.csv"
  ),
  row.names = FALSE
)

write.csv(
  run_audit,
  file.path(
    OUTPUT_DIR,
    "run_audit.csv"
  ),
  row.names = FALSE
)

write.csv(
  completion_audit,
  file.path(
    OUTPUT_DIR,
    "completion_audit.csv"
  ),
  row.names = FALSE
)

saveRDS(
  list(
    signature =
      analysis_signature,
    settings =
      analysis_settings,
    run_design =
      run_design,
    model_catalogue =
      model_catalogue,
    trait_dictionary =
      trait_dictionary,
    predictions =
      cv_predictions,
    genomic_values_by_fold =
      genomic_values_by_fold,
    metrics_by_fold =
      cv_metrics_by_fold,
    metrics_summary =
      cv_metrics_summary,
    fixed_effects_by_fold =
      fixed_effects_by_fold,
    marker_effects_by_fold =
      marker_effects_by_fold,
    convergence_diagnostics =
      convergence_diagnostics,
    convergence_summary =
      convergence_summary,
    runtime_summary =
      runtime_summary,
    bayesb2_audit =
      bayesb2_audit,
    rrblup_gblup_audit =
      rrblup_gblup_audit,
    run_audit =
      run_audit,
    completion_audit =
      completion_audit
  ),
  file.path(
    OUTPUT_DIR,
    "models_object.rds"
  )
)
```

## 11. Completion audit

```{r completion-audit-table}
kable(
  completion_audit,
  caption = "Completion audit for the 350 planned fits"
)
```

## 12. Cross-validated predictive performance

The pooled metrics use each individual's prediction from the fold in which that
individual was withheld. Consequently, every trait-method combination contains
150 out-of-fold predictions.

Predictive correlation, RMSE, MAE, bias, and regression slope are the primary
comparison criteria. DIC remains a model-fit diagnostic and is summarized
across folds.

```{r predictive-summary-table}
cv_metrics_summary |>
  select(
    Trait,
    Description_en,
    Unit,
    Method_label,
    Validation_n,
    Correlation,
    RMSE,
    MAE,
    Prediction_bias,
    Regression_slope,
    Mean_DIC
  ) |>
  arrange(
    Trait,
    desc(
      Correlation
    )
  ) |>
  kable(
    digits = 6,
    caption = "Pooled out-of-fold predictive performance"
  )
```

```{r predictive-correlation-heatmap, fig.height=7}
ggplot(
  cv_metrics_summary,
  aes(
    x = Method_label,
    y = Trait,
    fill = Correlation
  )
) +
  geom_tile() +
  geom_text(
    aes(
      label = sprintf(
        "%.3f",
        Correlation
      )
    )
  ) +
  labs(
    x = NULL,
    y = "Trait",
    fill = "Correlation",
    title = "Pooled out-of-fold predictive correlation"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

## 13. MCMC convergence review

ESS and Geweke statistics are single-chain diagnostics. They do not prove
convergence, but they identify runs and scalar parameters that require closer
inspection before biological interpretation.

```{r convergence-summary-table}
convergence_summary |>
  select(
    Method_label,
    Runs,
    Parameters_reviewed,
    Parameters_flagged,
    Runs_with_flags,
    Minimum_ESS,
    Median_ESS,
    Maximum_absolute_Geweke
  ) |>
  kable(
    digits = 3,
    caption = "Single-chain convergence review by method"
  )
```

```{r convergence-flag-table}
convergence_diagnostics |>
  filter(
    Review_flag
  ) |>
  arrange(
    Method,
    Trait,
    Fold,
    Parameter
  ) |>
  kable(
    digits = 4,
    caption = "Scalar chains flagged for additional review"
  )
```

### 13.1 Representative posterior-chain inspection

The scalar ESS and Geweke summaries above cover all official fits. To provide
the graphical MCMC inspection used in the reference study, the posterior draws
for one representative official fit were recovered by a deterministic replay
of `PRODCOM_F1_RRBLUP`. The replay used the same data, fold, matrices, withheld
phenotypes, seed, BGLR version, and MCMC settings as the original fit. It was
performed only because the production pipeline intentionally discarded the raw
BGLR chain files after calculating the compact convergence summaries.

The replay is not an additional cross-validation result and does not replace
any of the 350 official fits. Before inclusion in this module, its predictions,
effects, model-fit statistics, and convergence summaries were compared with
the original compact checkpoint, and all compared quantities matched exactly.
Only the compact retained posterior chains required for the final graphical
report are kept as a permanent model output.

```{r mcmc-posterior-load}
MCMC_POSTERIOR_DIR <- file.path(
  OUTPUT_DIR,
  "mcmc",
  paste0(
    "PRODCOM_F1_RRBLUP_",
    analysis_signature
  )
)

mcmc_replay_summary <- read.csv(
  file.path(
    MCMC_POSTERIOR_DIR,
    "posterior_chain_summary.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

mcmc_replay_chains <- readRDS(
  file.path(
    MCMC_POSTERIOR_DIR,
    "posterior_chains.rds"
  )
)

```

For the 200,000 retained posterior draws, the residual-variance chain had a
posterior mean of `r round(mcmc_replay_summary$Posterior_mean[mcmc_replay_summary$Parameter == "Residual_variance"], 4)`,
an effective sample size of
`r round(mcmc_replay_summary$Effective_sample_size[mcmc_replay_summary$Parameter == "Residual_variance"])`,
and a Geweke statistic of
`r round(mcmc_replay_summary$Geweke_Z[mcmc_replay_summary$Parameter == "Residual_variance"], 3)`.
The marker-variance chain had a posterior mean of
`r round(mcmc_replay_summary$Posterior_mean[mcmc_replay_summary$Parameter == "Marker_variance"], 4)`,
an effective sample size of
`r round(mcmc_replay_summary$Effective_sample_size[mcmc_replay_summary$Parameter == "Marker_variance"])`,
and a Geweke statistic of
`r round(mcmc_replay_summary$Geweke_Z[mcmc_replay_summary$Parameter == "Marker_variance"], 3)`.
Neither parameter was flagged for additional convergence review.

```{r mcmc-replay-summary-table}
mcmc_replay_summary |>
  select(
    Parameter,
    Retained_samples,
    Posterior_mean,
    Posterior_SD,
    Effective_sample_size,
    Geweke_Z,
    Review_flag
  ) |>
  kable(
    digits = 4,
    caption = "Posterior-chain diagnostics for the representative official RR-BLUP/BRR fit"
  )
```

```{r mcmc-replay-figure, fig.width=12, fig.height=7.6, fig.cap="Representative MCMC diagnostics for the official PRODCOM fold-1 RR-BLUP/BRR fit. Trace plots show the 200,000 retained post-burn-in draws; running means illustrate stabilization of posterior expectations; posterior densities summarize marginal posterior distributions. The deterministic replay reproduced the original official checkpoint exactly before these chains were used for visualization."}
par(
  mfrow = c(2, 3),
  mar = c(4.2, 4.5, 3.0, 1.0),
  oma = c(0, 0, 0.5, 0)
)

plot(
  mcmc_replay_chains$Iteration,
  mcmc_replay_chains$Residual_variance,
  type = "l",
  xlab = "Iteration",
  ylab = expression(sigma[e]^2),
  main = "Trace plot — residual variance"
)

plot(
  mcmc_replay_chains$Iteration,
  mcmc_replay_chains$Residual_running_mean,
  type = "l",
  xlab = "Iteration",
  ylab = "Running mean",
  main = "Running mean — residual variance"
)
abline(
  h = mean(
    mcmc_replay_chains$Residual_variance
  ),
  lty = 2
)

plot(
  density(
    mcmc_replay_chains$Residual_variance
  ),
  xlab = expression(sigma[e]^2),
  ylab = "Density",
  main = "Posterior distribution — residual variance"
)
abline(
  v = mean(
    mcmc_replay_chains$Residual_variance
  ),
  lty = 2
)

plot(
  mcmc_replay_chains$Iteration,
  mcmc_replay_chains$Marker_variance,
  type = "l",
  xlab = "Iteration",
  ylab = expression(sigma[b]^2),
  main = "Trace plot — marker variance"
)

plot(
  mcmc_replay_chains$Iteration,
  mcmc_replay_chains$Marker_running_mean,
  type = "l",
  xlab = "Iteration",
  ylab = "Running mean",
  main = "Running mean — marker variance"
)
abline(
  h = mean(
    mcmc_replay_chains$Marker_variance
  ),
  lty = 2
)

plot(
  density(
    mcmc_replay_chains$Marker_variance
  ),
  xlab = expression(sigma[b]^2),
  ylab = "Density",
  main = "Posterior distribution — marker variance"
)
abline(
  v = mean(
    mcmc_replay_chains$Marker_variance
  ),
  lty = 2
)
```

The trace plots fluctuate around stable regions without visible long-term
drift, while the running means stabilize rapidly over the retained posterior
sample. The marginal posterior distributions are unimodal and right-skewed,
especially for marker variance. Together with the large effective sample sizes
and Geweke statistics well within the review threshold, the representative
chains support adequate MCMC behavior for this official fit. These graphical
diagnostics complement, rather than replace, the systematic ESS and Geweke
review of all 350 official fits.

## 14. BayesB2 probability audit

```{r bayesb2-audit-table}
kable(
  bayesb2_audit,
  digits = 10,
  caption = "BayesB2 non-zero-effect probability audit"
)
```

## 15. RR-BLUP and GBLUP diagnostic comparison

RR-BLUP uses `M` directly through the `BRR` component, while GBLUP uses `G`
through the `RKHS` component. Because `G` is derived from `M`, their predictions
are expected to be related, but exact equality is not imposed.

```{r rrblup-gblup-table}
kable(
  rrblup_gblup_audit,
  digits = 8,
  caption = "RR-BLUP and GBLUP prediction comparison by trait and fold"
)
```

## 16. Runtime inventory

```{r runtime-summary-table}
runtime_summary |>
  select(
    Method_label,
    Fits,
    Total_seconds,
    Mean_seconds,
    Median_seconds,
    Maximum_seconds
  ) |>
  kable(
    digits = 3,
    caption = "Runtime summary for the complete analysis"
  )
```

## 17. Next step

Module 04 will use the out-of-fold predictions and genomic components saved
here to:

- compare methods within each trait;
- examine fold stability;
- summarize marker effects;
- calculate cross-validated genomic values;
- construct trait-specific rankings;
- document the influence of previously highlighted phenotypic observations;
- define the final selection outputs without imposing an undocumented
  multi-trait economic index.
