---
title: "Model fitting and MCMC diagnostics for genomic prediction in papaya"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

This module fits seven genomic-prediction methods to ten papaya traits using
the same five family-stratified cross-validation folds created in Module 02.
The objective is to show, step by step, how the statistical model is specified,
how each genomic method is represented in BGLR, how out-of-fold (OOF)
predictions and genomic values are obtained, and how MCMC diagnostics are
calculated.

The complete analysis contains

\[
10\ \text{traits}
\times
5\ \text{folds}
\times
7\ \text{methods}
=
350\ \text{fits}.
\]

The seven methods are RR-BLUP, BayesA, BayesB, BayesB2, BayesC, Bayesian
Lasso (BL), and GBLUP. RR-BLUP is implemented through the BGLR `BRR`
component, so RR-BLUP and Bayesian ridge regression are not treated as
separate methods in this project [@endelman2011rrblup; @perez2014bglr].

The model-fitting code is contained entirely in this R Markdown document. A
normal workflowr build presents the existing fitted results without repeating
the 350 fits. A complete refit is performed only when the explicit execution
switch described in Section 8 is enabled.

## 2. Packages, inputs, and output paths

Only packages required for presentation and data handling are attached during a
normal build. BGLR, coda, and digest are called with explicit namespaces in the
fitting code.

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
  library(here)
})
```

Module 03 reads the matrices produced in Module 02 and writes model outputs to
`output/models/`. Native BGLR chain files are kept under the single `bglr/`
subdirectory.

```{r paths}
MATRICES_FILE <- here("output", "matrices", "matrices_object.rds")
MODELS_FILE <- here("output", "models", "models_object.rds")
OUTPUT_DIR <- here("output", "models")
BGLR_DIR <- file.path(OUTPUT_DIR, "bglr")

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(BGLR_DIR, recursive = TRUE, showWarnings = FALSE)
```

The Module 02 object provides the aligned individuals, phenotypes, family
design matrix, marker matrix, genomic relationship matrix, and cross-validation
folds.

```{r load-matrices}
matrices_object <- readRDS(MATRICES_FILE)

ids <- matrices_object$ids
metadata <- matrices_object$metadata
phenotypes <- matrices_object$phenotypes
trait_dictionary <- matrices_object$trait_dictionary
X <- matrices_object$X
M <- matrices_object$M
G <- matrices_object$G
cv_folds <- matrices_object$cv_folds

traits <- colnames(phenotypes)
```

## 3. Statistical model

For every trait, the common model is

\[
\mathbf y =
\mathbf 1\mu +
\mathbf X_F\boldsymbol\beta_F +
\mathbf g +
\boldsymbol\varepsilon,
\]

where \(\mu\) is the intercept, \(\mathbf X_F\boldsymbol\beta_F\) is the
family fixed effect, \(\mathbf g\) is the genomic component, and
\(\boldsymbol\varepsilon\) is the residual vector.

### 3.1 Intercept and family fixed effect

BGLR fits an intercept automatically. Therefore, the intercept column created
in Module 02 is removed before `X` is supplied as a `FIXED` term. The remaining
columns code family effects and are estimated without shrinkage.

```{r family-fixed-effect}
X_fixed <- X[
  ,
  colnames(X) != "(Intercept)",
  drop = FALSE
]

colnames(X_fixed)
```

Here, `model = "FIXED"` refers to the treatment of the **family regression
coefficients** as fixed effects. This use of `FIXED` is different from
`type = "FIXED"` in the Bayesian Lasso, which is explained in Section 6.6.

### 3.2 Genomic component

For the six marker-effect methods,

\[
\mathbf g = \mathbf M\boldsymbol\alpha,
\]

so the model becomes

\[
\mathbf y =
\mathbf 1\mu +
\mathbf X_F\boldsymbol\beta_F +
\mathbf M\boldsymbol\alpha +
\boldsymbol\varepsilon.
\]

The methods differ mainly in the prior assigned to marker effects
\(\boldsymbol\alpha\). GBLUP instead models genomic values directly as

\[
\mathbf g \sim N(\mathbf 0,\mathbf G\sigma_g^2).
\]

## 4. Cross-validation design

The five folds are inherited unchanged from Module 02. Each fold contains 30
validation individuals and 120 training individuals, with three members of
each family in validation. The same folds are used for every trait and method,
which makes method comparisons directly paired on the same validation sets.

```{r cv-summary}
kable(
  matrices_object$fold_summary,
  caption = "Five family-stratified cross-validation folds"
)
```

For a given trait and fold, validation phenotypes are replaced by `NA`. BGLR
then fits the model using the remaining phenotypes and returns predictions for
the masked individuals.

```{r cv-mask-example, eval=FALSE}
validation_index <- which(cv_folds$Fold == fold_number)
training_index <- which(cv_folds$Fold != fold_number)

y_model <- y_observed
y_model[validation_index] <- NA_real_
```

Thus, every reported validation prediction is genuinely OOF with respect to
that fold.

## 5. MCMC settings

The MCMC settings follow the main analysis configuration adopted from the
reference study [@silva2021brrguava]. BGLR's prior-allocation parameter is set
to `R2 = 0.5`.

```{r mcmc-settings}
N_ITER <- 1000000L
BURN_IN <- 200000L
THIN <- 4L
R2_PRIOR <- 0.5
SEED_BASE <- 202608050L

RAW_SAVED_SAMPLES <- floor(N_ITER / THIN)
BURN_IN_SAVED_SAMPLES <- floor(BURN_IN / THIN)
RETAINED_SAMPLES <- RAW_SAVED_SAMPLES - BURN_IN_SAVED_SAMPLES

BAYESB2_ZERO_PROBABILITY <- 1e-5
BAYESB2_PROB_IN <- 1 - BAYESB2_ZERO_PROBABILITY
BAYESB2_COUNTS <- 1e6

ESS_REVIEW_THRESHOLD <- 1000
GEWEKE_REVIEW_THRESHOLD <- 2

BL_LAMBDA_TYPE <- "FIXED"
BL_LAMBDA_TOLERANCE_REL <- 1e-10

settings_table <- data.frame(
  Setting = c(
    "Iterations", "Burn-in", "Thinning", "Retained samples",
    "R2 prior", "BayesB2 zero-effect probability",
    "ESS review threshold", "Absolute Geweke review threshold"
  ),
  Value = c(
    N_ITER, BURN_IN, THIN, RETAINED_SAMPLES,
    R2_PRIOR, BAYESB2_ZERO_PROBABILITY,
    ESS_REVIEW_THRESHOLD, GEWEKE_REVIEW_THRESHOLD
  )
)

kable(settings_table, caption = "MCMC settings used in Module 03")
```

The ESS and Geweke thresholds are review rules only. They do not automatically
exclude a fitted model.

## 6. Genomic prediction methods

The seven methods use the same response, family term, folds, and MCMC settings.
Only the genomic component changes.

```{r model-catalogue}
method_names <- c(
  "RRBLUP", "BayesA", "BayesB", "BayesB2",
  "BayesC", "BL", "GBLUP"
)

method_labels <- c(
  RRBLUP = "RR-BLUP",
  BayesA = "BayesA",
  BayesB = "BayesB",
  BayesB2 = "BayesB2",
  BayesC = "BayesC",
  BL = "Bayesian Lasso",
  GBLUP = "GBLUP"
)

model_catalogue <- data.frame(
  Method = method_names,
  Label = unname(method_labels[method_names]),
  BGLR_component = c(
    "BRR", "BayesA", "BayesB", "BayesB",
    "BayesC", "BL", "RKHS"
  ),
  Genomic_input = c("M", "M", "M", "M", "M", "M", "G"),
  Main_assumption = c(
    "Common Gaussian variance for all marker effects",
    "Marker-specific scaled-t variances",
    "Scaled-t slab and a point mass at zero",
    "BayesB with zero-effect probability near 1e-5",
    "Common Gaussian slab and a point mass at zero",
    "Double-exponential marginal prior; lambda fixed at BGLR-calibrated lambda0",
    "Genomic values structured by G"
  )
)

kable(model_catalogue, caption = "Genomic prediction methods fitted in Module 03")
```

### 6.1 RR-BLUP

RR-BLUP is implemented with BGLR's `BRR` component. Marker effects have a
common Gaussian prior variance, producing ridge-type shrinkage of all marker
effects [@endelman2011rrblup; @perez2014bglr].

### 6.2 BayesA

BayesA assigns marker-specific variances, allowing markers to undergo different
amounts of shrinkage [@meuwissen2001prediction; @habier2011bayesianalphabet].

### 6.3 BayesB

BayesB combines a point mass at zero with a continuous distribution for
nonzero effects. Consequently, the model can simultaneously perform shrinkage
and variable selection.

### 6.4 BayesB2

`BayesB2` is the project name for a specific BGLR `BayesB` parameterization; it
is not a separate BGLR model name. The intended probability of a zero marker
effect is

\[
P(\alpha_j=0)=10^{-5}.
\]

BGLR parameterizes BayesB through `probIn`, the probability that a marker is
included with a nonzero effect. Therefore,

\[
probIn = 1-10^{-5}=0.99999.
\]

`counts = 10^6` makes the beta prior for `probIn` highly concentrated around
that value.

### 6.5 BayesC

BayesC also allows a marker effect to be zero, but nonzero marker effects share
a common variance rather than marker-specific variances.

### 6.6 Bayesian Lasso and the fixed lambda

The Bayesian Lasso uses a double-exponential marginal prior for marker effects
[@park2008bayesianlasso]. In BGLR, the shrinkage parameter \(\lambda\) can be
updated during MCMC or held fixed. In this analysis it is held fixed at the
value obtained from BGLR's own default calibration.

This is the second, distinct use of the word `FIXED` in Module 03:

- `model = "FIXED"` describes the unpenalized **family effects**;
- `type = "FIXED"` inside `model = "BL"` means that **lambda is not updated
  during MCMC**.

BGLR allocates the global `R2` among linear terms. The predictor contains two
terms: the family `FIXED` term and the genomic `BL` term. Hence the BL term
receives

\[
R^2_{BL}=\frac{0.5}{2}=0.25.
\]

For the centered marker matrix, the calibration uses

\[
MS_x=
\frac{1}{n}\sum_j\sum_i m_{ij}^2-
\sum_j\bar m_j^2,
\]

and

\[
\lambda_0^2=
2\frac{1-R^2}{R^2_{BL}}MS_x.
\]

The value is calculated directly from the Module 02 marker matrix.

```{r bl-lambda-calibration}
BL_N_LINEAR_TERMS <- 2L
BL_COMPONENT_R2 <- R2_PRIOR / BL_N_LINEAR_TERMS

BL_MSX <-
  sum(apply(M, 2, function(x) sum(x^2))) / nrow(M) -
  sum(colMeans(M)^2)

BL_LAMBDA_SQUARED <-
  2 * (1 - R2_PRIOR) / BL_COMPONENT_R2 * BL_MSX

BL_LAMBDA <- sqrt(BL_LAMBDA_SQUARED)
BL_LAMBDA_TOLERANCE <- max(
  1e-12,
  abs(BL_LAMBDA) * BL_LAMBDA_TOLERANCE_REL
)

stopifnot(
  is.finite(BL_MSX), BL_MSX > 0,
  is.finite(BL_LAMBDA), BL_LAMBDA > 0
)

kable(
  data.frame(
    Quantity = c("MSx", "Global R2", "BL component R2", "Fixed lambda0"),
    Value = c(BL_MSX, R2_PRIOR, BL_COMPONENT_R2, BL_LAMBDA)
  ),
  digits = 6,
  caption = "Calibration of the fixed Bayesian-Lasso shrinkage parameter"
)
```

For these data, this gives \(MS_x\approx25.6050\) and
\(\lambda_0\approx10.1203\). The fixed value was adopted globally after the
standard stochastic lambda update showed numerical instability during model
validation. It was **not** selected according to predictive correlation, RMSE,
DIC, or any trait-specific result. The same value is used for every trait and
fold because `M`, `R2`, and the model structure are common to all fits.

### 6.7 GBLUP

GBLUP uses the genomic relationship matrix `G` rather than individual marker
effects. In BGLR it is represented by an `RKHS` term with `K = G`, which gives
the genomic values the covariance structure defined by `G`.

## 7. BGLR model specifications

The common family term is combined with one genomic term for each method.
These `ETA` objects are the actual specifications passed to BGLR.

```{r eta-specifications}
family_term <- list(
  X = X_fixed,
  model = "FIXED"
)

ETA_models <- list(
  RRBLUP = list(
    fixed = family_term,
    genomic = list(X = M, model = "BRR")
  ),
  BayesA = list(
    fixed = family_term,
    genomic = list(X = M, model = "BayesA")
  ),
  BayesB = list(
    fixed = family_term,
    genomic = list(X = M, model = "BayesB")
  ),
  BayesB2 = list(
    fixed = family_term,
    genomic = list(
      X = M,
      model = "BayesB",
      probIn = BAYESB2_PROB_IN,
      counts = BAYESB2_COUNTS
    )
  ),
  BayesC = list(
    fixed = family_term,
    genomic = list(X = M, model = "BayesC")
  ),
  BL = list(
    fixed = family_term,
    genomic = list(
      X = M,
      model = "BL",
      type = BL_LAMBDA_TYPE,
      lambda = BL_LAMBDA
    )
  ),
  GBLUP = list(
    fixed = family_term,
    genomic = list(K = G, model = "RKHS")
  )
)
```

## 8. Cross-validated model fitting

### 8.1 Explicit execution switch

A normal workflowr build does not repeat the 350 computationally intensive
fits. To deliberately reproduce the complete analysis from this English page,
run

```bash
PAPAYA_RUN_M03_FIT=en Rscript --vanilla -e 'workflowr::wflow_build("analysis/03_models_en.Rmd")'
```

The Portuguese and English documents contain the same statistical code. A
complete fit should be triggered from only one language version at a time.

```{r m03-fit-control, include=FALSE}
M03_FIT_TARGET <- tolower(
  Sys.getenv("PAPAYA_RUN_M03_FIT", unset = "")
)
RUN_M03_FIT <- identical(M03_FIT_TARGET, "en")
```

### 8.2 Trait-fold-method combinations

The 350 combinations are created deterministically. Each fit has its own seed,
so results do not depend on the order in which combinations are processed.

```{r run-design}
run_design <- expand.grid(
  Trait = traits,
  Fold = seq_len(matrices_object$settings$n_folds),
  Method = method_names,
  stringsAsFactors = FALSE
)

run_design <- run_design[
  order(
    match(run_design$Trait, traits),
    run_design$Fold,
    match(run_design$Method, method_names)
  ),
]
rownames(run_design) <- NULL

run_design$Run_number <- seq_len(nrow(run_design))
run_design$Seed <- SEED_BASE + run_design$Run_number
run_design$Run_id <- paste(
  run_design$Trait,
  paste0("F", run_design$Fold),
  run_design$Method,
  sep = "_"
)

stopifnot(nrow(run_design) == 350L)
```

Metadata required only by downstream integrity checks are generated internally
and are not part of the tutorial output.

```{r internal-integrity-metadata, include=FALSE, eval=RUN_M03_FIT}
PIPELINE_VERSION <- "module03-full-v4-bl-fixed"

analysis_settings <- list(
  pipeline_version = PIPELINE_VERSION,
  n_iter = N_ITER,
  burn_in = BURN_IN,
  thin = THIN,
  R2 = R2_PRIOR,
  seed_base = SEED_BASE,
  bayesb2_zero_probability = BAYESB2_ZERO_PROBABILITY,
  bayesb2_prob_in = BAYESB2_PROB_IN,
  bayesb2_counts = BAYESB2_COUNTS,
  bl_lambda_type = BL_LAMBDA_TYPE,
  bl_n_linear_terms = BL_N_LINEAR_TERMS,
  bl_component_R2 = BL_COMPONENT_R2,
  bl_msx = BL_MSX,
  bl_lambda_squared = BL_LAMBDA_SQUARED,
  bl_lambda = BL_LAMBDA,
  bl_lambda_tolerance = BL_LAMBDA_TOLERANCE,
  bl_tau2_warning_policy = "review_not_structural",
  traits = traits,
  folds = seq_len(matrices_object$settings$n_folds),
  methods = method_names
)

analysis_signature <- substr(
  digest::digest(
    list(
      ids = ids,
      phenotypes = phenotypes,
      X = X,
      M = M,
      G = G,
      cv_folds = cv_folds,
      settings = analysis_settings
    ),
    algo = "sha256"
  ),
  1,
  16
)

run_design$Signature <- analysis_signature
run_design <- run_design |>
  relocate(Signature, .before = Run_id)

write.csv(
  run_design,
  file.path(OUTPUT_DIR, "run_design.csv"),
  row.names = FALSE
)
```

### 8.3 Validate the fixed lambda during fitting

The fixed BL value is checked directly in the fitted BGLR object and against
the native lambda file. This validation is the same for every BL fit and is
kept separate from the general fitting function.

```{r bl-validation-helper}
validate_fixed_bl <- function(
  fit,
  save_prefix,
  run_id,
  method_name,
  lambda_update_warning_count
) {
  result <- list(
    lambda = NA_real_,
    native_value = NA_real_,
    serialization_error = NA_real_,
    native_valid = NA,
    valid = NA
  )

  if (method_name != "BL") {
    return(result)
  }

  result$lambda <- as.numeric(fit$ETA$genomic$lambda)
  lambda_error <- abs(result$lambda - BL_LAMBDA)

  lambda_trace <- scan(
    paste0(save_prefix, "ETA_genomic_lambda.dat"),
    quiet = TRUE
  )

  result$native_value <- if (length(lambda_trace) > 0L) {
    lambda_trace[1]
  } else {
    NA_real_
  }

  result$serialization_error <- if (is.finite(result$native_value)) {
    abs(result$native_value - BL_LAMBDA)
  } else {
    Inf
  }

  result$native_valid <-
    length(lambda_trace) == RAW_SAVED_SAMPLES &&
    all(is.finite(lambda_trace)) &&
    all(lambda_trace > 0) &&
    length(unique(lambda_trace)) == 1L

  result$valid <-
    length(result$lambda) == 1L &&
    is.finite(result$lambda) &&
    result$lambda > 0 &&
    lambda_error <= BL_LAMBDA_TOLERANCE &&
    result$native_valid

  if (!result$valid) {
    stop("BL FIXED-lambda contract failed for ", run_id)
  }
  if (lambda_update_warning_count > 0L) {
    stop(
      "Unexpected lambda-update warning under BL type=FIXED for ",
      run_id
    )
  }

  result
}
```

### 8.4 Fit one combination

The fitting helper prepares the masked phenotype vector, fits the requested
BGLR model, captures warnings, and returns the information needed for the
subsequent calculations.

```{r fit-one-run-helper}
fit_one_run <- function(run_index) {
  trait_name <- run_design$Trait[run_index]
  fold_number <- run_design$Fold[run_index]
  method_name <- run_design$Method[run_index]
  run_id <- run_design$Run_id[run_index]
  run_seed <- run_design$Seed[run_index]

  bglr_run_dir <- file.path(
    BGLR_DIR,
    trait_name,
    paste0("F", fold_number),
    method_name
  )
  dir.create(bglr_run_dir, recursive = TRUE, showWarnings = FALSE)

  save_prefix <- file.path(bglr_run_dir, "bglr_")
  unlink(Sys.glob(paste0(save_prefix, "*")))

  y_observed <- phenotypes[, trait_name]
  names(y_observed) <- ids

  validation_index <- which(cv_folds$Fold == fold_number)
  training_index <- which(cv_folds$Fold != fold_number)

  y_model <- y_observed
  y_model[validation_index] <- NA_real_

  set.seed(run_seed)
  run_warning_messages <- character(0)
  fit_start <- proc.time()[["elapsed"]]

  fit <- withCallingHandlers(
    BGLR::BGLR(
      y = y_model,
      ETA = ETA_models[[method_name]],
      response_type = "gaussian",
      nIter = N_ITER,
      burnIn = BURN_IN,
      thin = THIN,
      R2 = R2_PRIOR,
      saveAt = save_prefix,
      verbose = FALSE,
      rmExistingFiles = TRUE
    ),
    warning = function(w) {
      run_warning_messages <<- c(
        run_warning_messages,
        conditionMessage(w)
      )
      invokeRestart("muffleWarning")
    }
  )

  fit_end <- proc.time()[["elapsed"]]

  tau2_warning_count <- sum(grepl(
    "tau2 was not updated",
    run_warning_messages,
    fixed = TRUE
  ))
  lambda_update_warning_count <- sum(grepl(
    "lambda was not updated",
    run_warning_messages,
    fixed = TRUE
  ))

  bl <- validate_fixed_bl(
    fit = fit,
    save_prefix = save_prefix,
    run_id = run_id,
    method_name = method_name,
    lambda_update_warning_count = lambda_update_warning_count
  )

  list(
    fit = fit,
    trait = trait_name,
    fold = fold_number,
    method = method_name,
    run_id = run_id,
    seed = run_seed,
    save_prefix = save_prefix,
    training_index = training_index,
    validation_index = validation_index,
    y_observed = y_observed,
    warning_messages = run_warning_messages,
    tau2_warning_count = tau2_warning_count,
    lambda_update_warning_count = lambda_update_warning_count,
    bl = bl,
    elapsed_seconds = fit_end - fit_start
  )
}
```

### 8.5 OOF predictions and genomic values

For each fitted model,

\[
\widehat y_i =
\widehat\mu+
\mathbf x_{Fi}'\widehat{\boldsymbol\beta}_F+
\widehat g_i.
\]

Therefore the genomic value is recovered as

\[
\widehat g_i=
\widehat y_i-
\widehat\mu-
\mathbf x_{Fi}'\widehat{\boldsymbol\beta}_F.
\]

```{r prediction-helper}
extract_prediction_outputs <- function(run) {
  fit <- run$fit

  fixed_component <- as.vector(
    X_fixed %*% fit$ETA$fixed$b
  )
  genomic_component <- as.vector(
    fit$yHat - fit$mu - fixed_component
  )

  genomic_values <- data.frame(
    Run_id = run$run_id,
    IND = ids,
    Trait = run$trait,
    Fold = run$fold,
    Method = run$method,
    Genomic_value = genomic_component
  )

  predictions <- data.frame(
    Run_id = run$run_id,
    IND = ids[run$validation_index],
    Trait = run$trait,
    Fold = run$fold,
    Method = run$method,
    Observed = run$y_observed[run$validation_index],
    Predicted = fit$yHat[run$validation_index],
    Fixed_component = fixed_component[run$validation_index],
    Genomic_component = genomic_component[run$validation_index]
  )

  list(
    genomic_values = genomic_values,
    predictions = predictions
  )
}
```

### 8.6 Predictive and fit metrics

Metrics are calculated separately within each trait-fold-method fit. The OOF
correlation, RMSE, MAE, bias, and calibration slope quantify prediction; DIC
and the effective number of parameters summarize Bayesian model fit and
complexity.

```{r metric-helper}
extract_metric_output <- function(run) {
  fit <- run$fit
  observed <- run$y_observed[run$validation_index]
  predicted <- fit$yHat[run$validation_index]

  posterior_prob_in <- c(
    fit$ETA$genomic$probIn,
    NA_real_
  )[1]

  metric <- data.frame(
    Run_id = run$run_id,
    Trait = run$trait,
    Fold = run$fold,
    Method = run$method,
    Seed = run$seed,
    Training_n = length(run$training_index),
    Validation_n = length(run$validation_index),
    Correlation = cor(observed, predicted),
    RMSE = sqrt(mean((observed - predicted)^2)),
    MAE = mean(abs(observed - predicted)),
    Prediction_bias = mean(predicted - observed),
    Regression_slope = cov(observed, predicted) / var(predicted),
    DIC = fit$fit$DIC,
    Effective_parameters = fit$fit$pD,
    Log_likelihood_at_posterior_mean = fit$fit$logLikAtPostMean,
    Posterior_mean = fit$mu,
    Posterior_mean_SD = c(fit$SD.mu, NA_real_)[1],
    Residual_variance = fit$varE,
    Residual_variance_SD = c(fit$SD.varE, NA_real_)[1],
    BL_lambda = run$bl$lambda,
    BL_lambda_native_value = run$bl$native_value,
    BL_lambda_native_serialization_error = run$bl$serialization_error,
    BL_lambda_native_valid = run$bl$native_valid,
    BL_lambda_valid = run$bl$valid,
    BGLR_warning_count = length(run$warning_messages),
    Tau2_warning_count = run$tau2_warning_count,
    Lambda_update_warning_count = run$lambda_update_warning_count,
    Posterior_probIn = posterior_prob_in,
    Elapsed_seconds = run$elapsed_seconds,
    Finite_predictions = all(is.finite(predicted))
  )

  finite_columns <- c(
    "Correlation", "RMSE", "MAE", "Prediction_bias",
    "Regression_slope", "DIC", "Effective_parameters",
    "Log_likelihood_at_posterior_mean", "Posterior_mean",
    "Residual_variance", "Elapsed_seconds"
  )

  metric$Finite_metrics <- all(
    is.finite(as.numeric(metric[, finite_columns]))
  )

  metric
}
```

## 9. MCMC diagnostics

BGLR writes thinned scalar chains to the native files. Two scalar chains are
reviewed for every fit: residual variance and one method-specific parameter.
For BL, lambda is fixed and therefore has no MCMC chain to diagnose; the
intercept chain is used as the second scalar diagnostic instead.

```{r trace-specifications}
genomic_trace_suffix <- c(
  RRBLUP = "ETA_genomic_varB.dat",
  BayesA = "ETA_genomic_ScaleBayesA.dat",
  BayesB = "ETA_genomic_parBayesB.dat",
  BayesB2 = "ETA_genomic_parBayesB.dat",
  BayesC = "ETA_genomic_parBayesC.dat",
  BL = "mu.dat",
  GBLUP = "ETA_genomic_varU.dat"
)

genomic_trace_header <- c(
  RRBLUP = FALSE,
  BayesA = FALSE,
  BayesB = TRUE,
  BayesB2 = TRUE,
  BayesC = TRUE,
  BL = FALSE,
  GBLUP = FALSE
)

genomic_trace_column <- c(
  RRBLUP = 1L,
  BayesA = 1L,
  BayesB = 2L,
  BayesB2 = 2L,
  BayesC = 2L,
  BL = 1L,
  GBLUP = 1L
)

genomic_trace_parameter <- c(
  RRBLUP = "Marker_variance",
  BayesA = "Scale",
  BayesB = "Scale",
  BayesB2 = "Scale",
  BayesC = "Marker_variance",
  BL = "Intercept",
  GBLUP = "Genomic_variance"
)
```

After burn-in is removed from the saved chains, effective sample size (ESS) and
the Geweke statistic are calculated with coda.

```{r convergence-helper}
extract_convergence_output <- function(run) {
  residual_trace_full <- scan(
    paste0(run$save_prefix, "varE.dat"),
    quiet = TRUE
  )

  genomic_trace_table <- read.table(
    paste0(
      run$save_prefix,
      genomic_trace_suffix[run$method]
    ),
    header = genomic_trace_header[run$method],
    check.names = FALSE
  )

  genomic_trace_full <- genomic_trace_table[
    ,
    genomic_trace_column[run$method]
  ]

  posterior_index <- seq.int(
    BURN_IN_SAVED_SAMPLES + 1L,
    length(residual_trace_full)
  )

  trace_matrix <- cbind(
    Residual_variance = residual_trace_full[posterior_index],
    Genomic_parameter = genomic_trace_full[posterior_index]
  )

  trace_object <- coda::mcmc(
    trace_matrix,
    start = BURN_IN + THIN,
    thin = THIN
  )

  trace_ess <- coda::effectiveSize(trace_object)
  trace_geweke <- coda::geweke.diag(trace_object)$z

  parameter_names <- c(
    "Residual_variance",
    unname(genomic_trace_parameter[run$method])
  )

  convergence <- data.frame(
    Run_id = run$run_id,
    Trait = run$trait,
    Fold = run$fold,
    Method = run$method,
    Parameter = parameter_names,
    Retained_samples = nrow(trace_matrix),
    Posterior_mean = colMeans(trace_matrix),
    Posterior_SD = apply(trace_matrix, 2, sd),
    Effective_sample_size = as.numeric(trace_ess),
    Geweke_Z = as.numeric(trace_geweke)
  )

  convergence$Review_flag <-
    !is.finite(convergence$Effective_sample_size) |
    convergence$Effective_sample_size < ESS_REVIEW_THRESHOLD |
    !is.finite(convergence$Geweke_Z) |
    abs(convergence$Geweke_Z) > GEWEKE_REVIEW_THRESHOLD

  convergence
}
```

A `Review_flag` therefore indicates a chain requiring inspection. It is not an
automatic model-rejection rule and does not by itself prove non-convergence.

## 10. Execute the 350 fits

The main loop is now short: each iteration fits one model and delegates
prediction, metric, and convergence calculations to the helper functions shown
above.

```{r fit-all-models, eval=RUN_M03_FIT}
metric_parts <- vector("list", nrow(run_design))
prediction_parts <- vector("list", nrow(run_design))
genomic_value_parts <- vector("list", nrow(run_design))
convergence_parts <- vector("list", nrow(run_design))

for (run_index in seq_len(nrow(run_design))) {
  cat(sprintf(
    "[%03d/%03d] %s\n",
    run_index,
    nrow(run_design),
    run_design$Run_id[run_index]
  ))

  run <- fit_one_run(run_index)
  prediction_output <- extract_prediction_outputs(run)

  metric_parts[[run_index]] <- extract_metric_output(run)
  prediction_parts[[run_index]] <- prediction_output$predictions
  genomic_value_parts[[run_index]] <- prediction_output$genomic_values
  convergence_parts[[run_index]] <- extract_convergence_output(run)

  run$fit <- NULL
  gc(verbose = FALSE)
}
```

The individual run outputs are then combined into the five analytical exports
used by the next module.

```{r combine-model-results, eval=RUN_M03_FIT}
cv_metrics_by_fold <- bind_rows(metric_parts)
cv_predictions <- bind_rows(prediction_parts)
genomic_values_by_fold <- bind_rows(genomic_value_parts)
convergence_diagnostics <- bind_rows(convergence_parts)

convergence_summary <- convergence_diagnostics |>
  group_by(Method) |>
  summarise(
    Runs = n_distinct(Run_id),
    Parameters_reviewed = n(),
    Parameters_flagged = sum(Review_flag),
    Runs_with_flags = n_distinct(Run_id[Review_flag]),
    Minimum_ESS = min(Effective_sample_size, na.rm = TRUE),
    Median_ESS = median(Effective_sample_size, na.rm = TRUE),
    Maximum_absolute_Geweke = max(abs(Geweke_Z), na.rm = TRUE),
    .groups = "drop"
  )

convergence_summary$Method_label <-
  unname(method_labels[convergence_summary$Method])
```

Integrity metadata required by the later modules is added internally after the
scientific tables have been constructed.

```{r add-internal-integrity-fields, include=FALSE, eval=RUN_M03_FIT}
cv_metrics_by_fold$Signature <- analysis_signature
cv_metrics_by_fold <- cv_metrics_by_fold |>
  relocate(Signature, .after = Run_id)

cv_predictions$Signature <- analysis_signature
cv_predictions <- cv_predictions |>
  relocate(Signature, .after = Run_id)

genomic_values_by_fold$Signature <- analysis_signature
genomic_values_by_fold <- genomic_values_by_fold |>
  relocate(Signature, .after = Run_id)

convergence_diagnostics$Signature <- analysis_signature
convergence_diagnostics <- convergence_diagnostics |>
  relocate(Signature, .after = Run_id)
```

## 11. Save Module 03 results

Before saving, the expected dimensions are checked. These checks protect the
pipeline from an incomplete fit but are not part of model selection.

```{r validate-results, eval=RUN_M03_FIT}
expected_fits <- nrow(run_design)
expected_predictions <-
  length(ids) * length(traits) * length(method_names)

stopifnot(
  expected_fits == 350L,
  nrow(cv_metrics_by_fold) == expected_fits,
  nrow(cv_predictions) == expected_predictions,
  nrow(genomic_values_by_fold) == length(ids) * expected_fits,
  nrow(convergence_diagnostics) == 2L * expected_fits
)
```

```{r validate-internal-integrity, include=FALSE, eval=RUN_M03_FIT}
stopifnot(
  length(unique(cv_metrics_by_fold$Signature)) == 1L,
  identical(unique(cv_metrics_by_fold$Signature), analysis_signature)
)

models_object <- list(
  signature = analysis_signature,
  settings = analysis_settings,
  run_design = run_design,
  model_catalogue = model_catalogue,
  trait_dictionary = trait_dictionary,
  predictions = cv_predictions,
  genomic_values_by_fold = genomic_values_by_fold,
  metrics_by_fold = cv_metrics_by_fold,
  convergence_diagnostics = convergence_diagnostics
)
```

The analytical tables and the complete Module 03 object are then written to
`output/models/`.

```{r save-model-results, eval=RUN_M03_FIT}
write.csv(
  cv_predictions,
  file.path(OUTPUT_DIR, "cv_predictions.csv"),
  row.names = FALSE
)
write.csv(
  genomic_values_by_fold,
  file.path(OUTPUT_DIR, "genomic_values_by_fold.csv"),
  row.names = FALSE
)
write.csv(
  cv_metrics_by_fold,
  file.path(OUTPUT_DIR, "cv_metrics_by_fold.csv"),
  row.names = FALSE
)
write.csv(
  convergence_diagnostics,
  file.path(OUTPUT_DIR, "convergence_diagnostics.csv"),
  row.names = FALSE
)
write.csv(
  convergence_summary,
  file.path(OUTPUT_DIR, "convergence_summary.csv"),
  row.names = FALSE
)

saveRDS(models_object, MODELS_FILE)

cat("\nM03 fitting completed.\n")
cat("Fits:", nrow(cv_metrics_by_fold), "\n")
cat("OOF predictions:", nrow(cv_predictions), "\n")
```

## 12. Results from the fitted Module 03 object

A normal build reaches this section without refitting models. The existing
Module 03 object is read and its dimensions are checked before the diagnostic
summary is displayed.

```{r load-model-object}
if (!file.exists(MODELS_FILE)) {
  stop(
    "Module 03 results were not found. Refit this document with ",
    "`PAPAYA_RUN_M03_FIT=en`."
  )
}

models_object <- readRDS(MODELS_FILE)
settings <- models_object$settings
model_catalogue_saved <- models_object$model_catalogue
run_design_saved <- models_object$run_design
cv_predictions <- models_object$predictions
genomic_values_by_fold <- models_object$genomic_values_by_fold
cv_metrics_by_fold <- models_object$metrics_by_fold
convergence_diagnostics <- models_object$convergence_diagnostics

expected_fits <-
  length(settings$traits) *
  length(settings$folds) *
  length(settings$methods)

stopifnot(
  expected_fits == 350L,
  nrow(run_design_saved) == expected_fits,
  nrow(cv_metrics_by_fold) == expected_fits,
  nrow(cv_predictions) ==
    150L * length(settings$traits) * length(settings$methods),
  nrow(genomic_values_by_fold) == 150L * expected_fits,
  nrow(convergence_diagnostics) == 2L * expected_fits
)
```

### 12.1 MCMC diagnostic summary

```{r convergence-summary}
convergence_summary_display <- convergence_diagnostics |>
  group_by(Method) |>
  summarise(
    Runs = n_distinct(Run_id),
    Parameters_reviewed = n(),
    Parameters_flagged = sum(Review_flag),
    Runs_with_flags = n_distinct(Run_id[Review_flag]),
    Minimum_ESS = min(Effective_sample_size, na.rm = TRUE),
    Median_ESS = median(Effective_sample_size, na.rm = TRUE),
    Maximum_absolute_Geweke = max(abs(Geweke_Z), na.rm = TRUE),
    .groups = "drop"
  )

convergence_summary_display$Method <- unname(
  setNames(
    model_catalogue_saved$Label,
    model_catalogue_saved$Method
  )[convergence_summary_display$Method]
)

kable(
  convergence_summary_display,
  digits = 3,
  caption = "Single-chain MCMC diagnostics by genomic method"
)
```

```{r convergence-flags}
convergence_diagnostics |>
  filter(Review_flag) |>
  arrange(Method, Trait, Fold, Parameter) |>
  select(
    Trait, Fold, Method, Parameter,
    Retained_samples, Effective_sample_size,
    Geweke_Z, Review_flag
  ) |>
  kable(
    digits = 4,
    caption = "Scalar chains flagged for additional review"
  )
```

The diagnostics are interpreted as review indicators. They are not used to
change the MCMC settings for individual traits or folds, and they do not by
themselves define the best prediction method.

## 13. Summary and next step

Module 03 fits the same statistical structure across seven genomic methods,
using family as a fixed effect and the same five OOF folds for every comparison.
The distinction between `model = "FIXED"` for family and `type = "FIXED"` for
the Bayesian-Lasso lambda is explicit, and the fixed lambda is derived from the
BGLR calibration rather than from predictive performance.

Module 04 reads `models_object.rds` without refitting the 350 models. It pools
OOF predictions, compares predictive performance and DIC, reviews MCMC
stability, and derives trait-specific genomic results.
