---
title: "3. Ridge-regression BLUP"
author: "Weverton Gomes da Costa"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
bibliography: ../REFERENCES.bib
link-citations: true
---

[Português](03_rrblup_pt.html) | [Previous](02_bayesian_models_en.html) | [Home](index.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 7)
source(file.path("code", "00_config.R"))
source(file.path("code", "01_data_functions.R"))
source(file.path("code", "02_model_functions.R"))
source(file.path("code", "03_reporting_functions.R"))
configuracao <- obter_configuracao()
objeto <- carregar_resultado(configuracao$arquivo_preprocessamento)
if (is.null(objeto)) objeto <- executar_preprocessamento(configuracao)
```

## Model

RR-BLUP models all centered allele-dosage effects as exchangeable random
effects [@endelman2011rrblup]:

$$
\mathbf{y}=\mathbf{X}\boldsymbol{\beta}+\mathbf{M}\boldsymbol{u}+
\boldsymbol{e},\qquad
\boldsymbol{u}\sim N(\mathbf{0},\mathbf{I}\sigma_u^2).
$$

The fixed design $\mathbf{X}$ includes the intercept, block, and family. The
same fold-specific dosage matrices used in the Bayesian analysis are supplied
to `rrBLUP::mixed.solve`, so imputation and centering never use validation
phenotypes or validation marker summaries.

## Controlled execution

RR-BLUP is fitted under `smoke` and `full` profiles. Under `site`, this page
loads the saved object if it exists and otherwise remains a methods page.

```{r fit-rrblup}
if (configuracao$modo %in% c("smoke", "full")) {
  resultado_rrblup <- executar_rrblup(objeto, configuracao)
} else {
  resultado_rrblup <- carregar_resultado(configuracao$arquivo_rrblup)
}
```

```{r rrblup-status, results='asis', echo=FALSE}
if (is.null(resultado_rrblup)) {
  cat(mensagem_resultado_ausente(configuracao$arquivo_rrblup, "en"))
}
```

## Variance components

For the all-individual fit, the marker-effect variance is converted to an
additive genomic variance using the multi-allelic denominator
$\sum_j 2p_j(1-p_j)$. The reported genomic heritability is the ratio of this
additive variance to additive plus residual variance.

```{r rrblup-variance-table, results='asis'}
if (!is.null(resultado_rrblup)) {
  resumo_variancias <- resultado_rrblup$full_model_summary |>
    dplyr::select(
      Trait, Marker_effect_variance, Additive_variance,
      Residual_variance, H2, Log_likelihood, Time_seconds
    )
  print(tabela_relatorio(resumo_variancias, "RR-BLUP full-data variance components"))
}
```

## Cross-validated performance

The table aggregates the eight fold-level fits. Page 5 recomputes each metric
once from all 150 concatenated out-of-fold predictions for fair model ranking.

```{r rrblup-metrics, results='asis'}
if (!is.null(resultado_rrblup)) {
  resumo_rrblup <- resumir_metricas_folds(resultado_rrblup$fold_metrics)
  print(tabela_relatorio(resumo_rrblup, "RR-BLUP fold-level performance"))
}
```

Observed and predicted values are shown in their original trait units. The
dashed line represents perfect agreement, not an estimated calibration line.

```{r rrblup-prediction-plot, echo=FALSE, fig.height=10}
if (!is.null(resultado_rrblup)) {
  print(
    ggplot2::ggplot(
      resultado_rrblup$predictions,
      ggplot2::aes(x = Observed, y = Predicted)
    ) +
      ggplot2::geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey55") +
      ggplot2::geom_point(color = "#2C7FB8", alpha = 0.7, size = 1.5) +
      ggplot2::facet_wrap(~ Trait, scales = "free", ncol = 2) +
      ggplot2::labs(
        x = "Observed",
        y = "Out-of-fold prediction",
        title = "RR-BLUP out-of-fold predictions"
      ) +
      tema_relatorio()
  )
}
```

Marker effects and full-data genomic values are retained in the result object.
Candidate ranking is performed only after the best predictive model has been
selected separately for each trait.

[Next: GBLUP](04_gblup_en.html)
