---
title: "4. Genomic BLUP"
author: "Weverton Gomes da Costa"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
bibliography: ../REFERENCES.bib
link-citations: true
---

[Português](04_gblup_pt.html) | [Previous](03_rrblup_en.html) | [Home](index.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 7)
source(file.path("code", "00_config.R"))
source(file.path("code", "01_data_functions.R"))
source(file.path("code", "02_model_functions.R"))
source(file.path("code", "03_reporting_functions.R"))
configuracao <- obter_configuracao()
objeto <- carregar_resultado(configuracao$arquivo_preprocessamento)
if (is.null(objeto)) objeto <- executar_preprocessamento(configuracao)
```

## Model

GBLUP models individual genomic values directly:

$$
\mathbf{y}=\mathbf{X}\boldsymbol{\beta}+\mathbf{Z}\boldsymbol{a}+
\boldsymbol{e},\qquad
\boldsymbol{a}\sim N(\mathbf{0},\mathbf{G}\sigma_a^2).
$$

The genomic relationship matrix is the multi-allelic extension of the first
VanRaden construction [@vanraden2008efficient],

$$
\mathbf{G}=\frac{\mathbf{M}_c\mathbf{M}_c^{\mathsf T}}
{\sum_j 2p_j(1-p_j)},
$$

where $\mathbf{M}_c$ is the centered allele-dosage matrix. Within cross-
validation, allele frequencies, centering means, and $\mathbf{G}$ are defined
from the training sample. Validation individuals remain in the relationship
matrix so their genomic values can be predicted from relationships with the
training set, but their phenotypes are passed as missing.

## Controlled execution

GBLUP runs under `smoke` and `full` profiles and otherwise loads a saved result.
The same fixed effects and eight folds are used in every competing model.

```{r fit-gblup}
if (configuracao$modo %in% c("smoke", "full")) {
  resultado_gblup <- executar_gblup(objeto, configuracao)
} else {
  resultado_gblup <- carregar_resultado(configuracao$arquivo_gblup)
}
```

```{r gblup-status, results='asis', echo=FALSE}
if (is.null(resultado_gblup)) {
  cat(mensagem_resultado_ausente(configuracao$arquivo_gblup, "en"))
}
```

## Relationship-matrix diagnostics

The complete-data matrix is checked for symmetry and positive semidefiniteness
before modeling. Its eigenvalue summary gives a compact diagnostic of the
available genomic dimensions and numerical tolerance.

```{r relationship-diagnostics, results='asis'}
if (!is.null(resultado_gblup)) {
  autovalores <- eigen(
    resultado_gblup$relationship_matrix,
    symmetric = TRUE,
    only.values = TRUE
  )$values
  diagnosticos_g <- data.frame(
    Diagnostic = c(
      "Individuals", "Maximum asymmetry", "Minimum eigenvalue",
      "Maximum eigenvalue", "Positive eigenvalues (> 1e-8)"
    ),
    Value = c(
      nrow(resultado_gblup$relationship_matrix),
      max(abs(resultado_gblup$relationship_matrix - t(resultado_gblup$relationship_matrix))),
      min(autovalores),
      max(autovalores),
      sum(autovalores > 1e-8)
    )
  )
  print(tabela_relatorio(diagnosticos_g, "Genomic relationship diagnostics"))
}
```

The heatmap displays the complete relationship matrix in phenotype-table
order. It is descriptive; model performance is assessed only through held-out
phenotypes.

```{r relationship-heatmap, echo=FALSE, fig.height=8}
if (!is.null(resultado_gblup)) {
  g_long <- as.data.frame(as.table(resultado_gblup$relationship_matrix))
  names(g_long) <- c("Individual_1", "Individual_2", "Relationship")
  g_long$Individual_1 <- factor(g_long$Individual_1, levels = objeto$ids)
  g_long$Individual_2 <- factor(g_long$Individual_2, levels = rev(objeto$ids))

  print(
    ggplot2::ggplot(
      g_long,
      ggplot2::aes(x = Individual_1, y = Individual_2, fill = Relationship)
    ) +
      ggplot2::geom_raster() +
      ggplot2::scale_fill_gradient2(low = "#B2182B", mid = "white", high = "#2166AC") +
      ggplot2::labs(x = NULL, y = NULL, fill = "G", title = "Genomic relationship matrix") +
      tema_relatorio() +
      ggplot2::theme(axis.text = ggplot2::element_blank(), axis.ticks = ggplot2::element_blank())
  )
}
```

## Variance components and prediction

Because $\mathbf{G}$ is scaled by allele-frequency variance, `Vu` estimates the
additive genomic variance directly. Genomic heritability is
$\sigma_a^2/(\sigma_a^2+\sigma_e^2)$.

```{r gblup-variance-table, results='asis'}
if (!is.null(resultado_gblup)) {
  print(tabela_relatorio(
    resultado_gblup$full_model_summary,
    "GBLUP full-data variance components"
  ))
}
```

Fold-level metrics provide an initial view; the common concatenated out-of-
fold comparison is reported on the next page.

```{r gblup-metrics, results='asis'}
if (!is.null(resultado_gblup)) {
  resumo_gblup <- resumir_metricas_folds(resultado_gblup$fold_metrics)
  print(tabela_relatorio(resumo_gblup, "GBLUP fold-level performance"))
}
```

```{r gblup-prediction-plot, echo=FALSE, fig.height=10}
if (!is.null(resultado_gblup)) {
  print(
    ggplot2::ggplot(
      resultado_gblup$predictions,
      ggplot2::aes(x = Observed, y = Predicted)
    ) +
      ggplot2::geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey55") +
      ggplot2::geom_point(color = "#2166AC", alpha = 0.7, size = 1.5) +
      ggplot2::facet_wrap(~ Trait, scales = "free", ncol = 2) +
      ggplot2::labs(
        x = "Observed",
        y = "Out-of-fold prediction",
        title = "GBLUP out-of-fold predictions"
      ) +
      tema_relatorio()
  )
}
```

[Next: model comparison](05_model_comparison_en.html)
