---
title: "Model comparison, genomic selection, and results"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

Using the OOF predictions, performance metrics, and diagnostics obtained in
Module 03, this final module organizes model comparison and genomic selection
in a more direct sequence: **compare → choose the method → define trait
eligibility → interpret GEBVs → select individuals**.

Method choice is trait-specific and prioritizes predictive performance: highest
pooled OOF correlation, followed by lowest RMSE and lowest MAE; DIC is retained
as complementary evidence and a final tie-breaker. MCMC diagnostics are
summarized as a global check of chain adequacy, but they are not converted into
an exclusion or review rule.

Individuals are selected only for traits that meet a pre-defined minimum level
of predictive evidence. In this version, a trait is eligible when the selected
method has OOF correlation greater than 0.30 and NRMSE lower than 1, that is,
RMSE lower than one phenotypic standard deviation. The selected proportion
remains 20% of the individuals for each eligible trait.

To improve interpretation on the original trait scale, the centered OOF genomic
component is preserved and is also presented after adding the phenotypic trait
mean. This transformation shifts the scale but does not change the ranking of
individuals within a trait.

## 2. Packages and model-fitting results

### 2.1 Packages used

The packages used here have distinct roles:

- `dplyr`: grouping, joining, ordering, and summarizing result tables;
- `tidyr`: reshaping tables between long and wide formats;
- `ggplot2`: building performance and observed-versus-predicted figures;
- `ggthemes`: applying the GDocs theme and color scale;
- `knitr`: formatting tables in R Markdown documents;
- `here`: building project-relative file paths.

```{r packages, warning=FALSE, message=FALSE}
library(dplyr) # Group, join, and summarize tables.
library(tidyr)   # Reshape data between long and wide formats.
library(ggplot2) # Build diagnostic figures.
library(ggthemes) # Apply the GDocs theme and color scale.
library(knitr)   # Format tables in R Markdown documents.
library(here)  # Build project-relative paths.
```

```{r table-formatting, include=FALSE}
format_table <- function(x, caption = NULL, digits = getOption("digits"), col.names = NA) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    col.names = col.names,
    table.attr = paste0(
      'class="table table-striped table-hover" ',
      'style="width:100%; margin-bottom:0;"'
    )
  )
  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

### 2.2 Input objects

The module uses two consolidated objects: `matrices_object.rds`, which supplies phenotypes, metadata, and definitions from Module 02, and `models_object.rds`, which supplies fold-level metrics, OOF predictions, and MCMC diagnostics from Module 03. During derivation, no output directory is created; revised results are persisted only in the final section and do not overwrite previous M04 outputs.

```{r inputs}
matrices_object <- readRDS(
  here("output", "matrices", "matrices_object.rds")
)

models_object <- readRDS(
  here("output", "models", "models_object.rds")
)

metrics <- models_object$metrics_by_fold
predictions <- models_object$predictions
convergence <- models_object$convergence_diagnostics

traits <- colnames(matrices_object$phenotypes)
methods <- models_object$settings$methods
```

### 2.3 Presentation labels and graphical standard

The calculations retain the method codes stored in the analytical objects.
Readable labels and the graphical theme below affect presentation only; they
do not modify any scientific quantity.

```{r presentation-style}
method_labels <- c(
  RRBLUP = "RR-BLUP/BRR",
  BayesA = "BayesA",
  BayesB = "BayesB",
  BayesB2 = "BayesB2",
  BayesC = "BayesC",
  BL = "Bayesian Lasso",
  GBLUP = "GBLUP"
)

selection_objective_labels <- c(
  increase = "Increase",
  decrease = "Decrease"
)

project_theme <- theme_gdocs() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )
```

## 3. OOF predictive performance

### 3.1 Scope of OOF evaluation

The OOF estimates inherit the family-stratified design of Modules 02 and
03. They quantify prediction within the represented breeding population and do
not estimate generalization to entirely new families.

The main metrics are recalculated using the 150 out-of-fold predictions for
each trait × method combination. Each individual contributes one OOF
prediction for every method.

### 3.2 Pooled OOF performance

For each trait × method combination, correlation, RMSE, MAE, bias, and slope are calculated from the 150 OOF predictions. The two-sided Pearson-correlation p-value is retained only as a nominal descriptive measure because training sets overlap across folds.

```{r pooled-oof-metrics}
# Combine OOF predictions and compute aggregate predictive metrics by trait and method.
pooled_metrics <- predictions |>
  group_by(Trait, Method) |>
  summarise(
    N = n(),
    r = cor(Observed, Predicted),
    RMSE = sqrt(mean((Observed - Predicted)^2)),
    MAE = mean(abs(Observed - Predicted)),
    Bias = mean(Predicted - Observed),
    Slope = cov(Observed, Predicted) / var(Predicted),
    .groups = "drop"
  ) |>
  mutate(
    t_value = r * sqrt((N - 2) / (1 - r^2)),
    p_value = 2 * pt(abs(t_value), df = N - 2, lower.tail = FALSE)
  ) |>
  select(-t_value)

# Summarize the results in `fold_summary`.
fold_summary <- metrics |>
  group_by(Trait, Method) |>
  summarise(
    Mean_DIC = mean(DIC),
    SD_DIC = sd(DIC),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_fold_RMSE = mean(RMSE),
    SD_fold_RMSE = sd(RMSE),
    Mean_fold_MAE = mean(MAE),
    SD_fold_MAE = sd(MAE),
    .groups = "drop"
  )
```

`pooled_metrics` combines the 150 OOF predictions and provides overall performance for each trait × method combination. `fold_summary` retains summaries inherited from the five folds to integrate DIC and auxiliary metrics in the comparison table.

Pearson-correlation p-values remain descriptive context only and do not define eligibility for genomic selection.

## 4. DIC and relative support for the methods

### 4.1 DIC-derived relative support

Following the logic used in the reference study [@silva2021brrguava], the
difference from the smallest mean DIC within each trait is converted into
relative weights:

\[
\Delta_j=\overline{DIC}_j-\min_k(\overline{DIC}_k),
\]

\[
Wprob_j=\frac{\exp(-\Delta_j/2)}{\sum_k\exp(-\Delta_k/2)},
\qquad
ER_j=\frac{Wprob_{\max}}{Wprob_j}.
\]

`Wprob` is used here as a DIC-derived relative-support weight. It is not
interpreted as a calibrated posterior probability of a model.

The implementation uses log-weights before normalization to improve numerical stability without changing relative ratios. `Wprob` sums to 1 within each trait, and `ER` expresses support relative to the method with the largest weight.

```{r model-comparison}
# Combine fold performance and DIC to build the quantities used for model comparison.
model_comparison <- pooled_metrics |>
  left_join(fold_summary, by = c("Trait", "Method")) |>
  group_by(Trait) |>
  mutate(
    Delta_DIC = Mean_DIC - min(Mean_DIC),
    log_weight = -0.5 * Delta_DIC,
    stable_weight = exp(log_weight - max(log_weight)),
    Wprob = stable_weight / sum(stable_weight),
    # Compute the evidence ratio relative to the method with the largest Wprob in each trait.
    ER = max(Wprob) / Wprob
  ) |>
  ungroup() |>
  select(
    Trait, Method, Mean_DIC, SD_DIC, Delta_DIC, Wprob, ER,
    N, r, p_value, RMSE, MAE, Bias, Slope,
    Mean_fold_r, SD_fold_r, Mean_fold_RMSE, SD_fold_RMSE,
    Mean_fold_MAE, SD_fold_MAE
  ) |>
  arrange(match(Trait, traits), Delta_DIC)
```

### 4.2 Relationship between Bayesian fit and predictive performance

`model_comparison` combines OOF performance and DIC-derived relative support for every trait and method. DIC quantities are comparable only within the same trait and do not replace predictive evidence.

In the joint comparison, methods with similar DIC can have different OOF performance, and the method with the largest `Wprob` need not have the largest predictive correlation. Bayesian fit and predictive ability therefore remain complementary rather than interchangeable criteria.

### 4.3 Visual comparison of OOF predictive performance

The pooled OOF correlation is the primary visual reference for predictive
comparison. The horizontal line at zero separates positive from nonpositive
association. Bars are point estimates, not uncertainty intervals.

```{r predictive-correlation-figure, fig.width=13, fig.height=10, fig.alt="Faceted bar chart comparing pooled out-of-fold predictive correlations of the seven genomic prediction methods across the ten traits, with a horizontal reference at zero."}
# Order methods and build the OOF-correlation figure by trait.
predictive_plot <- model_comparison |>
  mutate(Method = factor(Method, levels = methods)) |>
  ggplot(aes(x = Method, y = r, fill = Method)) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.4) +
  geom_col(width = 0.75) +
  facet_wrap(~ Trait, scales = "fixed", ncol = 2) +
  scale_fill_gdocs(
    breaks = methods,
    labels = method_labels
  ) +
  scale_x_discrete(labels = method_labels) +
  labs(
    x = NULL,
    y = "OOF correlation",
    fill = "Method",
    title = "Predictive ability of the seven methods"
  ) +
  project_theme +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  theme(
    legend.direction = "horizontal",
    legend.box = "horizontal",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

predictive_plot
```


Predictive magnitude and method ordering vary among traits. The workflow
therefore does not define one globally superior method; selection remains
trait-specific and continues with complementary DIC and diagnostic evidence.

### 4.4 Joint predictive and DIC comparison

The table below places the complementary evidence sources side by side.
DIC-derived quantities are compared only within the same trait, while OOF
correlation retains its predictive interpretation. The nominal p-value is
shown for context and is not a ranking-eligibility criterion.

```{r model-comparison-table}
# DISPLAY-ONLY START
model_comparison_display <- model_comparison |>
  mutate(Method_label = unname(method_labels[Method])) |>
  transmute(
    Trait,
    Method,
    Method_label,
    DIC = sprintf("%.2f", Mean_DIC),
    Delta_DIC = sprintf("%.2f", Delta_DIC),
    Wprob = sprintf("%.3f", Wprob),
    ER = sprintf("%.3f", ER),
    r = sprintf("%.3f", r),
    p_value = if_else(
      p_value < 0.001,
      "<0.001",
      sprintf("%.3f", p_value)
    )
  ) |>
  arrange(match(Trait, traits), match(Method, methods))

format_table(
  model_comparison_display |>
    select(Trait, Method_label, DIC, Delta_DIC, Wprob, ER, r, p_value),
  caption = "Comparison of the seven methods by trait",
  col.names = c("Trait", "Method", "DIC", "Delta DIC", "Wprob", "ER", "r", "p-value")
)
# DISPLAY-ONLY END
```

## 5. Method choice and eligibility for selection

### 5.1 Adopted criteria

Method choice and the decision to use a trait for selection are two distinct
steps. First, the seven methods are ordered within each trait by pooled OOF
correlation; ties are resolved, in this order, by lower RMSE, lower MAE, and
lower mean DIC. Thus, no artificial score is created by mixing metrics with
different meanings.

After the method is chosen, the trait is considered eligible for selection only
when it simultaneously satisfies two performance criteria:

\[
r_{OOF} > 0.30
\]

and

\[
NRMSE = \frac{RMSE}{s_y} < 1,
\]

where \(s_y\) is the observed phenotypic standard deviation of the trait. The
second criterion requires model error to be lower than the observed phenotypic
dispersion and complements the correlation threshold.

```{r selection-parameters}
CORRELATION_SELECTION_THRESHOLD <- 0.30
NRMSE_SELECTION_THRESHOLD <- 1.00
SELECTION_PROPORTION <- 0.20

selection_objective_map <- c(
  MMCOM = "increase",
  PRODCOM = "increase",
  PRODEF = "decrease",
  CF = "increase",
  DF = "increase",
  DCI = "decrease",
  EP = "increase",
  FFC = "increase",
  FFP = "increase",
  SS = "increase"
)

selection_objectives <- data.frame(
  Trait = traits,
  Selection_objective = unname(selection_objective_map[traits]),
  stringsAsFactors = FALSE
)

selection_parameters <- data.frame(
  Parameter = c(
    "Method choice",
    "Minimum OOF correlation",
    "Maximum NRMSE",
    "Selected proportion"
  ),
  Rule = c(
    "Highest r; ties by lower RMSE, lower MAE, and lower DIC",
    "> 0.30",
    "< 1.00",
    "20% per eligible trait"
  ),
  stringsAsFactors = FALSE
)

format_table(
  selection_parameters,
  caption = "Parameters adopted for method choice and genomic selection",
  col.names = c("Parameter", "Rule")
)
```

### 5.2 Selected method and trait eligibility

The phenotypic mean and standard deviation are calculated once per trait from
the observed values. Duplicate observations across methods are removed before
this calculation.

```{r final-model-selection}
trait_scale <- predictions |>
  distinct(Trait, IND, Observed) |>
  group_by(Trait) |>
  summarise(
    Phenotypic_mean = mean(Observed, na.rm = TRUE),
    Phenotypic_SD = sd(Observed, na.rm = TRUE),
    .groups = "drop"
  )

model_selection <- model_comparison |>
  group_by(Trait) |>
  arrange(desc(r), RMSE, MAE, Mean_DIC, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    Selected_method = Method,
    r,
    p_value,
    RMSE,
    MAE,
    Bias,
    Slope,
    Mean_DIC,
    Delta_DIC,
    Wprob,
    ER
  ) |>
  left_join(trait_scale, by = "Trait") |>
  left_join(selection_objectives, by = "Trait") |>
  mutate(
    NRMSE = RMSE / Phenotypic_SD,
    Predictive_skill_vs_SD = 1 - NRMSE,
    Eligible_by_correlation = r > CORRELATION_SELECTION_THRESHOLD,
    Eligible_by_error = NRMSE < NRMSE_SELECTION_THRESHOLD,
    Selection_eligible = Eligible_by_correlation & Eligible_by_error
  ) |>
  arrange(match(Trait, traits))

eligible_traits <- model_selection |>
  filter(Selection_eligible) |>
  pull(Trait)
```

The main table concentrates the metrics needed to interpret method choice and
eligibility, avoiding parallel tables for the same decision. The correlation
p-value remains available in the scientific object but is not used as a
selection filter.

```{r model-selection-table}
# DISPLAY-ONLY START
model_selection_display <- model_selection |>
  mutate(
    Method_label = unname(method_labels[Selected_method]),
    Objective = unname(selection_objective_labels[Selection_objective]),
    Eligible = if_else(Selection_eligible, "Yes", "No")
  ) |>
  transmute(
    Trait,
    Method_label,
    Objective,
    r = sprintf("%.3f", r),
    RMSE = sprintf("%.3f", RMSE),
    NRMSE = sprintf("%.3f", NRMSE),
    MAE = sprintf("%.3f", MAE),
    Bias = sprintf("%.3f", Bias),
    Slope = sprintf("%.3f", Slope),
    DIC = sprintf("%.2f", Mean_DIC),
    Eligible
  )

format_table(
  model_selection_display,
  caption = "Selected method and eligibility for selection by trait",
  col.names = c(
    "Trait", "Method", "Objective", "r", "RMSE", "NRMSE",
    "MAE", "Bias", "Slope", "Mean DIC", "Eligible"
  )
)
# DISPLAY-ONLY END
```

### 5.3 Summary of MCMC diagnostics

Diagnostics calculated in Module 03 are summarized here without creating a
review stage. A chain is counted as not flagged when none of the monitored
parameters in that fit received `Review_flag` in the source object. The field
name is preserved for traceability, but its interpretation here is only a
diagnostic summary.

```{r convergence-overview}
fit_convergence <- convergence |>
  group_by(Run_id, Trait, Method) |>
  summarise(
    Diagnostic_flag = any(Review_flag, na.rm = TRUE),
    .groups = "drop"
  )

convergence_overview <- data.frame(
  Level = c("Fits", "Monitored parameters"),
  Total = c(nrow(fit_convergence), nrow(convergence)),
  Without_flags = c(
    sum(!fit_convergence$Diagnostic_flag),
    sum(!convergence$Review_flag, na.rm = TRUE)
  )
) |>
  mutate(
    Proportion_without_flags = Without_flags / Total
  )

format_table(
  convergence_overview,
  digits = 4,
  caption = "Global summary of MCMC diagnostics",
  col.names = c("Level", "Total", "Without flags", "Proportion")
)
```

The predominance of fits and parameters without flags supports the conclusion
that most chains showed behavior compatible with convergence according to the
monitored diagnostics. This summary does not prove convergence of every marker
effect individually and does not participate in method choice.

### 5.4 Relationship between observed and predicted values

In addition to numerical metrics, observed versus predicted values are shown
for the selected method in each trait. The dashed line represents the 1:1
relationship; the fitted line is the linear regression of observed values on
the OOF predictions. Differences between the two help visualize calibration in
addition to the association summarized by \(r\).

```{r observed-predicted-figure, fig.width=12, fig.height=10, fig.alt="Faceted scatterplots of observed values against OOF values predicted by the selected method for each trait, with a 1:1 line and fitted linear regression."}
selected_predictions <- predictions |>
  inner_join(
    model_selection |>
      select(Trait, Selected_method, r, Selection_eligible),
    by = "Trait"
  ) |>
  filter(Method == Selected_method) |>
  mutate(
    Method_label = unname(method_labels[Method]),
    Facet = paste0(
      Trait, " — ", Method_label,
      "\nr = ", sprintf("%.3f", r)
    )
  )

observed_predicted_plot <- selected_predictions |>
  ggplot(aes(x = Predicted, y = Observed)) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, linewidth = 0.5) +
  geom_point(alpha = 0.65, size = 1.6) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.7) +
  facet_wrap(~ Facet, scales = "free", ncol = 2) +
  labs(
    x = "OOF predicted value",
    y = "Observed value",
    title = "Relationship between observed and predicted values"
  ) +
  project_theme +
  theme(legend.position = "none")

observed_predicted_plot
```

## 6. Marker-based heritability analogue and adjusted accuracy in RR-BLUP/BRR

### 6.1 Definition and interpretation limits

Following Silva et al. [@silva2021brrguava], a marker-based additive-variance
analogue is obtained from the common marker variance and the \(\sum 2pq\)
denominator. Fold-level adjusted accuracy is defined as
\(r_f/\sqrt{H_f^2}\). The resulting \(H^2\) is reported as a model-based
heritability analogue rather than as an independent classical estimate of
narrow-sense heritability. The quantity named `Predictive_accuracy` is a derived ratio, `r_f/sqrt(H_f^2)`, rather than an observed correlation between true and predicted genetic values. Because it depends on sample estimates of `r_f` and `H_f^2`, it may be negative or exceed 1; it should therefore be interpreted as an adjusted-accuracy index under this definition, not as a measure necessarily bounded to [0, 1].

### 6.2 Fold-level derivation and across-fold summary

```{r rrblup-heritability}
# Compute and organize the RR-BLUP/BRR model-based heritability summary.
rrblup_variance_components <- convergence |>
  filter(
    Method == "RRBLUP",
    Parameter %in% c("Residual_variance", "Marker_variance")
  ) |>
  select(Run_id, Trait, Fold, Method, Parameter, Posterior_mean) |>
  pivot_wider(names_from = Parameter, values_from = Posterior_mean)

rrblup_fold_correlations <- metrics |>
  filter(Method == "RRBLUP") |>
  select(Trait, Fold, Correlation)

# Organize the cross-validation and fitting structure.
rrblup_heritability_by_fold <- rrblup_variance_components |>
  left_join(rrblup_fold_correlations, by = c("Trait", "Fold")) |>
  mutate(
    Marker_variance_denominator = matrices_object$denominator,
    Additive_variance = Marker_variance * Marker_variance_denominator,
    H2 = Additive_variance / (Additive_variance + Residual_variance),
    Predictive_accuracy = Correlation / sqrt(H2)
  ) |>
  arrange(match(Trait, traits), Fold)

# Summarize the RR-BLUP/BRR model-based heritability analogue and adjusted accuracy across folds.
rrblup_heritability_summary <- rrblup_heritability_by_fold |>
  group_by(Trait) |>
  summarise(
    Mean_H2 = mean(H2),
    SD_H2 = sd(H2),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_predictive_accuracy = mean(Predictive_accuracy),
    SD_predictive_accuracy = sd(Predictive_accuracy),
    .groups = "drop"
  ) |>
  arrange(match(Trait, traits))
```

`rrblup_heritability_summary` retains, by trait, the model-derived RR-BLUP/BRR heritability analogue and the ratio `Correlation / sqrt(H2)`. This ratio is not a conventional correlation and may lie outside [0, 1].

### 6.3 Trait-level summary

The table presents the quantities just derived. `RRBLUP_H2` is a
marker-model-based analogue rather than an independent classical estimate of
narrow-sense heritability, and the adjusted-accuracy ratio is not constrained
to `[0, 1]`.

```{r rrblup-heritability-table}
# DISPLAY-ONLY START
rrblup_table <- rrblup_heritability_summary |>
  transmute(
    Trait,
    RRBLUP_H2 = sprintf("%.3f", Mean_H2),
    SD_RRBLUP_H2 = sprintf("%.3f", SD_H2),
    RRBLUP_predictive_accuracy = sprintf("%.3f", Mean_predictive_accuracy),
    SD_RRBLUP_predictive_accuracy = sprintf("%.3f", SD_predictive_accuracy)
  )

format_table(
  rrblup_table,
  caption = "Marker-based heritability analogue and adjusted predictive accuracy in RR-BLUP/BRR",
  col.names = c("Trait", "RR-BLUP H2", "SD RR-BLUP H2", "RR-BLUP predictive accuracy", "SD RR-BLUP predictive accuracy")
)
# DISPLAY-ONLY END
```

## 7. OOF GEBVs and an interpretable phenotypic scale

### 7.1 Cross-fitted genomic component

The genomic component of each individual continues to be obtained in the fold
in which that individual's phenotype was masked. Therefore, `GEBV_centered` is
a cross-fitted OOF score and not a final genetic value from refitting all 150
phenotypes.

To facilitate interpretation in the trait unit, the following quantity is also
calculated:

\[
GEBV_{scale} = \bar{y} + GEBV_{centered}.
\]

Adding the phenotypic mean shifts the scale. It does not change the ordering of
individuals and should not be confused with the complete phenotypic prediction,
because other model components may contribute to `Predicted`.

```{r selected-oof-genomic-values}
selected_oof_genomic_values <- predictions |>
  inner_join(
    model_selection |>
      select(
        Trait, Selected_method, r, p_value, NRMSE,
        Selection_eligible, Selection_objective
      ),
    by = "Trait"
  ) |>
  filter(Method == Selected_method) |>
  left_join(trait_scale, by = "Trait") |>
  transmute(
    IND,
    Trait,
    Fold,
    Method,
    Observed,
    Predicted,
    Selection_objective,
    Selection_eligible,
    OOF_r = r,
    OOF_p_value = p_value,
    OOF_NRMSE = NRMSE,
    Phenotypic_mean,
    GEBV_centered = Genomic_component,
    GEBV_on_trait_scale = Phenotypic_mean + Genomic_component
  ) |>
  arrange(match(Trait, traits), IND)
```

## 8. Individual selection in eligible traits

### 8.1 Ranking and selection intensity

Only traits that simultaneously satisfy $r > 0.30$ and `NRMSE < 1` enter this
stage. Within each eligible trait, individuals are ordered by the GEBV on the
trait scale while respecting the breeding objective. For traits to be reduced,
the sign is reversed only for ranking. The adopted selection intensity remains
20%, and `IND` is used only as a deterministic tie-breaker for exact ties.

```{r genomic-ranking}
genomic_ranking <- selected_oof_genomic_values |>
  filter(Selection_eligible) |>
  mutate(
    Selection_score = case_when(
      Selection_objective == "increase" ~ GEBV_on_trait_scale,
      Selection_objective == "decrease" ~ -GEBV_on_trait_scale,
      TRUE ~ NA_real_
    )
  ) |>
  group_by(Trait) |>
  arrange(desc(Selection_score), IND, .by_group = TRUE) |>
  mutate(Rank = row_number()) |>
  ungroup() |>
  arrange(match(Trait, traits), Rank)

selection_sizes <- genomic_ranking |>
  count(Trait, name = "N_available") |>
  mutate(
    Selection_proportion = SELECTION_PROPORTION,
    N_selected = pmax(
      1L,
      as.integer(ceiling(Selection_proportion * N_available))
    )
  ) |>
  arrange(match(Trait, traits))

selected_individuals_by_trait <- genomic_ranking |>
  inner_join(selection_sizes, by = "Trait") |>
  filter(Rank <= N_selected) |>
  left_join(
    matrices_object$metadata |>
      select(IND, FAM),
    by = "IND"
  ) |>
  select(
    Trait, Rank, IND, FAM, Method, Selection_objective,
    GEBV_centered, GEBV_on_trait_scale, Selection_score,
    OOF_r, OOF_NRMSE, Selection_proportion,
    N_available, N_selected
  ) |>
  arrange(match(Trait, traits), Rank)
```

### 8.2 Preview of selected individuals

The table shows the first five individuals for each eligible trait. The
`selected_individuals_by_trait` object retains the complete set of the selected
20%.

```{r proportional-selection-preview}
# DISPLAY-ONLY START
selection_preview <- selected_individuals_by_trait |>
  group_by(Trait) |>
  slice_head(n = 5) |>
  ungroup() |>
  mutate(
    Method_label = unname(method_labels[Method]),
    Objective = unname(selection_objective_labels[Selection_objective])
  ) |>
  transmute(
    Trait,
    Rank,
    IND,
    FAM,
    Method_label,
    Objective,
    GEBV = sprintf("%.3f", GEBV_on_trait_scale),
    r = sprintf("%.3f", OOF_r),
    NRMSE = sprintf("%.3f", OOF_NRMSE)
  )

format_table(
  selection_preview,
  caption = "First five selected individuals for each eligible trait",
  col.names = c(
    "Trait", "Rank", "IND", "FAM", "Method", "Objective",
    "GEBV + mean", "OOF r", "NRMSE"
  )
)
# DISPLAY-ONLY END
```

## 9. Recurrence across eligible traits

### 9.1 Definition

Recurrence is descriptive and uses only rankings from traits that passed both
eligibility thresholds. For each individual, it counts how many of these traits
include the individual among the selected 20%. Frequency is calculated relative
to the number of eligible traits; no multi-trait selection index is constructed
and no weights are assigned across traits.

```{r multitrait-recurrence}
recurrence_counts <- selected_individuals_by_trait |>
  group_by(IND, FAM) |>
  summarise(
    N_traits = n(),
    Traits = paste(Trait, collapse = ", "),
    Best_rank = min(Rank),
    Mean_rank_present = mean(Rank),
    .groups = "drop"
  )

n_eligible_traits <- length(eligible_traits)

multitrait_recurrence <- matrices_object$metadata |>
  select(IND, FAM) |>
  left_join(recurrence_counts, by = c("IND", "FAM")) |>
  mutate(
    N_traits = tidyr::replace_na(N_traits, 0L),
    Traits = tidyr::replace_na(Traits, ""),
    Best_rank = if_else(N_traits == 0L, NA_integer_, Best_rank),
    Mean_rank_present = if_else(
      N_traits == 0L,
      NA_real_,
      Mean_rank_present
    ),
    Selection_frequency = if (n_eligible_traits > 0) {
      N_traits / n_eligible_traits
    } else {
      0
    },
    Recurrence_class = case_when(
      N_traits >= 4L ~ "high_4plus",
      N_traits == 3L ~ "intermediate_3",
      N_traits == 2L ~ "broad_2",
      N_traits == 1L ~ "specific_1",
      TRUE ~ "none_0"
    )
  ) |>
  arrange(desc(N_traits), FAM, IND)

recurrence_distribution <- multitrait_recurrence |>
  count(N_traits, Recurrence_class, name = "N_individuals") |>
  mutate(
    Proportion_of_population =
      N_individuals / nrow(matrices_object$metadata)
  ) |>
  arrange(desc(N_traits))

multitrait_candidates <- multitrait_recurrence |>
  filter(N_traits >= 2L) |>
  arrange(desc(N_traits), FAM, IND)
```

### 9.2 Recurrence distribution

```{r recurrence-distribution-table}
# DISPLAY-ONLY START
recurrence_class_labels <- c(
  high_4plus = "High (>=4)",
  intermediate_3 = "Intermediate (3)",
  broad_2 = "Broad (2)",
  specific_1 = "Specific (1)",
  none_0 = "None (0)"
)

recurrence_distribution_display <- recurrence_distribution |>
  mutate(
    Class = unname(recurrence_class_labels[Recurrence_class])
  ) |>
  transmute(
    N_traits,
    Class,
    N_individuals,
    Proportion_of_population = sprintf("%.3f", Proportion_of_population)
  )

format_table(
  recurrence_distribution_display,
  caption = "Distribution of individuals by recurrence across eligible traits",
  col.names = c(
    "Number of traits", "Class", "Individuals",
    "Population proportion"
  )
)
# DISPLAY-ONLY END
```

### 9.3 Recurrent individuals

The table includes individuals selected in at least two eligible traits. An
additional table with one rank column per trait is not displayed; individual
positions remain available in `selected_individuals_by_trait`.

```{r recurrent-candidates-table}
# DISPLAY-ONLY START
recurrent_candidates_display <- multitrait_candidates |>
  mutate(
    Class = unname(recurrence_class_labels[Recurrence_class])
  ) |>
  transmute(
    IND,
    FAM,
    N_traits,
    Selection_frequency = sprintf("%.3f", Selection_frequency),
    Class,
    Traits,
    Best_rank,
    Mean_rank_present = sprintf("%.2f", Mean_rank_present)
  )

format_table(
  recurrent_candidates_display,
  caption = "Individuals recurrent in at least two eligible traits",
  col.names = c(
    "IND", "FAM", "Number of traits", "Frequency",
    "Class", "Traits", "Best rank", "Mean rank"
  )
)
# DISPLAY-ONLY END
```

## 10. Correlations among OOF GEBVs of eligible traits

These correlations describe associations among the OOF genomic values produced by the selected method for each trait. Because different traits may use different methods and the values are cross-fitted, the matrix summarizes associations among predictive scores rather than jointly estimated genetic covariance. It should therefore not be interpreted as a genetic-correlation matrix from a multivariate model.


```{r oof-correlations}
selected_genomic_wide <- selected_oof_genomic_values |>
  filter(Selection_eligible) |>
  select(IND, Trait, GEBV_centered) |>
  pivot_wider(names_from = Trait, values_from = GEBV_centered)

selected_oof_correlations <- cor(
  as.matrix(selected_genomic_wide[, eligible_traits, drop = FALSE])
)
```

`selected_oof_correlations` is derived here and visualized immediately below. It should be interpreted only as association among OOF scores, not as joint genetic correlation.

### 10.1 Exploratory heatmap

The heatmap visualizes the matrix just derived. Colors use the original
unrounded values. These are associations among cross-fitted scores and must
not be interpreted as genetic correlations from a multivariate model.

```{r oof-correlation-figure, fig.width=10, fig.height=9, fig.alt="Heatmap of pairwise correlations among cross-fitted OOF genomic-value scores for the traits, representing exploratory associations rather than multivariate genetic correlations."}
# Reshape the OOF correlation matrix and build the cross-trait heatmap.
correlation_long <- as.data.frame(
  as.table(selected_oof_correlations)
)
names(correlation_long) <- c(
  "Trait_1",
  "Trait_2",
  "Correlation"
)

# Set only the label of |r| < 0.005 to zero; the original correlation still drives analysis and color.
correlation_long <- correlation_long |>
  mutate(
    Correlation_label = replace(
      Correlation,
      abs(Correlation) < 0.005,
      0
    )
  )

# Build the heatmap of correlations among OOF genomic values.
correlation_plot <- ggplot(
  correlation_long,
  aes(x = Trait_1, y = Trait_2, fill = Correlation)
) +
  geom_tile() +
  geom_text(
    aes(label = sprintf("%.2f", Correlation_label)),
    size = 3
  ) +
  scale_fill_distiller(
    type = "div",
    palette = "RdBu",
    direction = 1,
    limits = c(-1, 1)
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "r",
    title = "Correlation among OOF GEBVs of eligible traits"
  ) +
  coord_equal() +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

correlation_plot
```


Positive values indicate that two score sets tend to order candidates in the
same direction, while negative values indicate opposing directions. The matrix
does not create an additional selection criterion.

## 11. Persistence of revised results

The previous results from version `v1.0.5` are not overwritten. This revision
is saved in its own M04 subdirectory, allowing the previous structure to be
compared or recovered when needed.

```{r save-results}
threshold_results_dir <- here(
  "output", "results", "m04", "threshold_selection"
)
dir.create(threshold_results_dir, recursive = TRUE, showWarnings = FALSE)

write.csv(
  model_selection,
  file.path(threshold_results_dir, "model_selection.csv"),
  row.names = FALSE
)
write.csv(
  selected_oof_genomic_values,
  file.path(threshold_results_dir, "selected_oof_genomic_values.csv"),
  row.names = FALSE
)
write.csv(
  selected_individuals_by_trait,
  file.path(threshold_results_dir, "selected_individuals_by_trait.csv"),
  row.names = FALSE
)
write.csv(
  multitrait_recurrence,
  file.path(threshold_results_dir, "multitrait_recurrence.csv"),
  row.names = FALSE
)
write.csv(
  multitrait_candidates,
  file.path(threshold_results_dir, "multitrait_candidates.csv"),
  row.names = FALSE
)
write.csv(
  selected_oof_correlations,
  file.path(threshold_results_dir, "selected_oof_correlations.csv"),
  row.names = TRUE
)

module04_threshold_results <- list(
  model_comparison = model_comparison,
  model_selection = model_selection,
  convergence_overview = convergence_overview,
  rrblup_heritability_by_fold = rrblup_heritability_by_fold,
  rrblup_heritability_summary = rrblup_heritability_summary,
  selected_oof_genomic_values = selected_oof_genomic_values,
  genomic_ranking = genomic_ranking,
  selected_individuals_by_trait = selected_individuals_by_trait,
  multitrait_recurrence = multitrait_recurrence,
  multitrait_candidates = multitrait_candidates,
  selected_oof_correlations = selected_oof_correlations,
  settings = list(
    correlation_threshold = CORRELATION_SELECTION_THRESHOLD,
    nrmse_threshold = NRMSE_SELECTION_THRESHOLD,
    selection_proportion = SELECTION_PROPORTION,
    eligible_traits = eligible_traits
  )
)

saveRDS(
  module04_threshold_results,
  here("output", "results", "m04", "threshold_selection", "module04_threshold_results.rds")
)
```

## 12. Final interpretation

The comparison now clearly separates three decisions. First, the method is
chosen within each trait using OOF predictive metrics, while DIC is retained as
complementary information. Second, only traits with $r > 0.30$ and `NRMSE < 1`
are considered sufficiently predictive for selection. Third, within those
traits, the most favorable 20% of individuals are prioritized according to the
breeding objective.

MCMC diagnostics indicate behavior compatible with convergence for the large
majority of monitored fits and parameters without being used as a decision
filter. The observed-versus-predicted figure adds a visual assessment of
association and calibration to the numerical metrics.

OOF GEBVs remain cross-fitted scores. Adding the phenotypic mean places them on
a more interpretable scale, but does not turn them into complete phenotypic
predictions and does not change rankings. Recurrence across traits remains
descriptive and is not equivalent to a multi-trait selection index.

Conclusions remain conditional on the represented breeding population, the
available SSR panel, and the validation design used.
