---
title: "Derivation and selection of genomic results"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
```

## 1. Objective and analytical contract

This module transforms the official outputs from Module 03 into derived analytical results. No model is refitted and no final-presentation figure is produced here.

Derived outputs include OOF predictive performance; DIC, DIC delta, `Wprob`, and `ER`; trait-specific method decisions; fold stability; computational time; heritability and accuracy in the RR-BLUP/BRR analogue; OOF genomic values; rankings; top 10 individuals; recurrence across rankings; and correlations among genomic values.

Method selection integrates OOF performance, relative fit by DIC, and chain stability. The choices are explicitly stored in `selection_spec` and **must be reviewed whenever Module 03 is rerun with corrected data or specifications**.

Genomic values used for ranking are strictly OOF: each individual receives the genomic component from the fold in which its phenotype was masked. No multi-trait selection is performed.

## 2. Official inputs

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(here)
})
```

```{r inputs}
MATRICES_FILE <- here("output", "matrices", "matrices_object.rds")
MODELS_FILE <- here("output", "models", "models_object.rds")
RESULTS_DIR <- here("output", "results", "m04")

dir.create(RESULTS_DIR, recursive = TRUE, showWarnings = FALSE)

matrices_object <- readRDS(MATRICES_FILE)
models_object <- readRDS(MODELS_FILE)

metrics <- models_object$metrics_by_fold
predictions <- models_object$predictions
convergence <- models_object$convergence_diagnostics
traits <- colnames(matrices_object$phenotypes)
methods <- c("RRBLUP", "BayesA", "BayesB", "BayesB2", "BayesC", "BL", "GBLUP")
```

## 3. OOF predictive performance

Primary metrics are calculated from the 150 out-of-sample predictions for each trait-method combination, preserving exactly one OOF prediction per individual.

```{r pooled-oof-metrics}
pooled_metrics <- predictions |>
  group_by(Trait, Method) |>
  summarise(
    N = n(),
    r = cor(Observed, Predicted),
    RMSE = sqrt(mean((Observed - Predicted)^2)),
    MAE = mean(abs(Observed - Predicted)),
    Bias = mean(Predicted - Observed),
    Slope = cov(Observed, Predicted) / var(Predicted),
    .groups = "drop"
  ) |>
  mutate(
    t_value = r * sqrt((N - 2) / (1 - r^2)),
    p_value = 2 * pt(abs(t_value), df = N - 2, lower.tail = FALSE)
  ) |>
  select(-t_value)

fold_summary <- metrics |>
  group_by(Trait, Method) |>
  summarise(
    Mean_DIC = mean(DIC),
    SD_DIC = sd(DIC),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_fold_RMSE = mean(RMSE),
    SD_fold_RMSE = sd(RMSE),
    Mean_fold_MAE = mean(MAE),
    SD_fold_MAE = sd(MAE),
    .groups = "drop"
  )
```

Pearson-correlation p-values are retained only as nominal descriptive measures because cross-validation training sets overlap across folds.

## 4. Method comparison

Silva et al. [@silva2021brrguava] calculated approximate model probabilities from each model's DIC difference relative to the minimum DIC. The same logic is applied jointly to the seven methods in this project.

\[
\Delta_j=DIC_j-\min(DIC),\qquad
Wprob_j=\frac{\exp(-\Delta_j/2)}{\sum_k\exp(-\Delta_k/2)},\qquad
ER_j=\frac{Wprob_{\max}}{Wprob_j}.
\]

```{r model-comparison}
model_comparison <- pooled_metrics |>
  left_join(fold_summary, by = c("Trait", "Method")) |>
  group_by(Trait) |>
  mutate(
    Delta_DIC = Mean_DIC - min(Mean_DIC),
    log_weight = -0.5 * Delta_DIC,
    stable_weight = exp(log_weight - max(log_weight)),
    Wprob = stable_weight / sum(stable_weight),
    ER = max(Wprob) / Wprob
  ) |>
  ungroup() |>
  select(
    Trait, Method, Mean_DIC, SD_DIC, Delta_DIC, Wprob, ER,
    N, r, p_value, RMSE, MAE, Bias, Slope,
    Mean_fold_r, SD_fold_r, Mean_fold_RMSE, SD_fold_RMSE,
    Mean_fold_MAE, SD_fold_MAE
  ) |>
  arrange(match(Trait, traits), Delta_DIC)
```

## 5. Method selection by trait

The highest OOF `r` is the predictive reference and the minimum DIC is the relative-fit reference. The decision remains explicit, allowing prediction, DIC, and stability to be compared without imposing an arbitrary scalar function.

```{r model-selection}
predictive_winners <- model_comparison |>
  group_by(Trait) |>
  arrange(desc(r), RMSE, MAE, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    Predictive_winner = Method,
    Predictive_winner_r = r,
    Predictive_winner_p_value = p_value,
    Predictive_winner_RMSE = RMSE,
    Predictive_winner_MAE = MAE
  )

dic_winners <- model_comparison |>
  group_by(Trait) |>
  arrange(Mean_DIC, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    DIC_winner = Method,
    Minimum_DIC = Mean_DIC,
    DIC_winner_Wprob = Wprob
  )

convergence_by_trait_method <- convergence |>
  group_by(Trait, Method) |>
  summarise(
    Runs = n_distinct(Run_id),
    Parameters_reviewed = n(),
    Parameters_flagged = sum(Review_flag),
    Runs_with_flags = n_distinct(Run_id[Review_flag]),
    Minimum_ESS = min(Effective_sample_size, na.rm = TRUE),
    Maximum_absolute_Geweke = max(abs(Geweke_Z), na.rm = TRUE),
    .groups = "drop"
  )

# Decisions from the currently published analysis.
# Explicitly review after any new Module 03 run.
selection_spec <- data.frame(
  Trait = c("MMCOM", "PRODCOM", "PRODEF", "CF", "DF", "DCI", "EP", "FFC", "FFP", "SS"),
  Selected_method = c("RRBLUP", "BL", "GBLUP", "BayesB", "RRBLUP", "RRBLUP", "BL", "RRBLUP", "BL", "BayesC"),
  Decision_code = c(
    "prefer_dic_stability_small_predictive_loss",
    "predictive_and_dic_winner_convergence_review",
    "prefer_stable_dic_winner_weak_prediction",
    "predictive_and_dic_winner",
    "predictive_and_dic_winner",
    "best_prediction_similar_dic_stable",
    "predictive_and_dic_winner_no_positive_signal",
    "best_prediction_similar_dic_stable_weak_signal",
    "predictive_and_dic_winner_weak_signal_convergence_review",
    "prefer_stable_dic_winner_negligible_predictive_loss"
  ),
  stringsAsFactors = FALSE
)

selected_performance <- model_comparison |>
  inner_join(selection_spec |> select(Trait, Selected_method), by = "Trait") |>
  filter(Method == Selected_method) |>
  transmute(
    Trait, Selected_method,
    r, p_value, RMSE, MAE, Bias, Slope,
    Mean_DIC, Delta_DIC, Wprob, ER, Mean_fold_r, SD_fold_r
  )

model_selection <- selection_spec |>
  left_join(selected_performance, by = c("Trait", "Selected_method")) |>
  left_join(predictive_winners, by = "Trait") |>
  left_join(dic_winners, by = "Trait") |>
  left_join(
    convergence_by_trait_method,
    by = c("Trait", "Selected_method" = "Method")
  ) |>
  mutate(
    Predictive_loss = Predictive_winner_r - r,
    Predictive_winner_agrees = Selected_method == Predictive_winner,
    DIC_winner_agrees = Selected_method == DIC_winner,
    Convergence_review = Parameters_flagged > 0,
    Predictive_evidence = case_when(
      r > 0 & p_value < 0.05 ~ "positive_p_lt_0.05",
      r > 0 & p_value >= 0.05 ~ "positive_p_ge_0.05",
      TRUE ~ "nonpositive_p_ge_0.05"
    )
  ) |>
  arrange(match(Trait, traits))

predictive_evidence_summary <- model_selection |>
  count(Predictive_evidence, name = "Traits") |>
  arrange(Predictive_evidence)
```

## 6. Fold stability of selected methods

Values are derived here; their graphical presentation belongs exclusively to Module 05.

```{r selected-fold-diagnostics}
selected_fold_diagnostics <- metrics |>
  inner_join(
    selection_spec |> select(Trait, Selected_method),
    by = c("Trait" = "Trait", "Method" = "Selected_method")
  ) |>
  transmute(Trait, Fold, Method, Correlation, RMSE, MAE) |>
  arrange(match(Trait, traits), Fold)
```

## 7. Computational time

Only elapsed times from the 350 official Module 03 fits are used; no additional runs are executed solely for timing.

```{r runtime}
runtime_summary <- metrics |>
  group_by(Method) |>
  summarise(
    Fits = n(),
    Total_seconds = sum(Elapsed_seconds),
    Mean_seconds = mean(Elapsed_seconds),
    SD_seconds = sd(Elapsed_seconds),
    Median_seconds = median(Elapsed_seconds),
    Minimum_seconds = min(Elapsed_seconds),
    Maximum_seconds = max(Elapsed_seconds),
    .groups = "drop"
  ) |>
  arrange(match(Method, methods))
```

## 8. Heritability and accuracy in the RR-BLUP/BRR analogue

Following Silva et al. [@silva2021brrguava], additive genetic variance in the RR-BLUP/BRR analogue is calculated from the common marker variance and the \(\sum 2pq\) denominator. Fold accuracy is \(r_f/\sqrt{H_f^2}\).

```{r rrblup-heritability}
rrblup_variance_components <- convergence |>
  filter(
    Method == "RRBLUP",
    Parameter %in% c("Residual_variance", "Marker_variance")
  ) |>
  select(Run_id, Trait, Fold, Method, Parameter, Posterior_mean) |>
  pivot_wider(names_from = Parameter, values_from = Posterior_mean)

rrblup_fold_correlations <- metrics |>
  filter(Method == "RRBLUP") |>
  select(Trait, Fold, Correlation)

rrblup_heritability_by_fold <- rrblup_variance_components |>
  left_join(rrblup_fold_correlations, by = c("Trait", "Fold")) |>
  mutate(
    Marker_variance_denominator = matrices_object$denominator,
    Additive_variance = Marker_variance * Marker_variance_denominator,
    H2 = Additive_variance / (Additive_variance + Residual_variance),
    Predictive_accuracy = Correlation / sqrt(H2)
  ) |>
  arrange(match(Trait, traits), Fold)

rrblup_heritability_summary <- rrblup_heritability_by_fold |>
  group_by(Trait) |>
  summarise(
    Mean_H2 = mean(H2),
    SD_H2 = sd(H2),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_predictive_accuracy = mean(Predictive_accuracy),
    SD_predictive_accuracy = sd(Predictive_accuracy),
    .groups = "drop"
  ) |>
  arrange(match(Trait, traits))
```

## 9. OOF genomic values

```{r selected-oof-genomic-values}
selected_oof_genomic_values <- predictions |>
  inner_join(
    model_selection |> select(Trait, Selected_method, Predictive_evidence),
    by = "Trait"
  ) |>
  filter(Method == Selected_method) |>
  transmute(
    IND, Trait, Fold, Method, Predictive_evidence,
    Genomic_value = Genomic_component
  ) |>
  arrange(match(Trait, traits), IND)
```

## 10. Ranking, top 10, and recurrence

```{r genomic-ranking}
genomic_ranking <- selected_oof_genomic_values |>
  group_by(Trait) |>
  arrange(desc(Genomic_value), IND, .by_group = TRUE) |>
  mutate(Rank = row_number()) |>
  ungroup() |>
  arrange(match(Trait, traits), Rank)

top10_with_family <- genomic_ranking |>
  filter(Rank <= 10L) |>
  left_join(
    matrices_object$metadata |> select(IND, FAM),
    by = "IND"
  ) |>
  select(
    Trait, Rank, IND, FAM, Method, Genomic_value, Predictive_evidence
  ) |>
  arrange(match(Trait, traits), Rank)

recurring_top10 <- top10_with_family |>
  filter(Predictive_evidence == "positive_p_lt_0.05") |>
  group_by(IND, FAM) |>
  summarise(
    N_traits = n(),
    Traits = paste(Trait, collapse = ", "),
    .groups = "drop"
  ) |>
  filter(N_traits >= 2L) |>
  arrange(desc(N_traits), FAM, IND)
```

Recurrence across rankings is descriptive and does not constitute a multi-trait selection index.

## 11. Correlations among OOF genomic values

These are descriptive correlations among predicted genomic values; they are not multivariate genetic-correlation estimates.

```{r oof-correlations}
selected_genomic_wide <- selected_oof_genomic_values |>
  select(IND, Trait, Genomic_value) |>
  pivot_wider(names_from = Trait, values_from = Genomic_value)

selected_oof_correlations <- cor(
  as.matrix(selected_genomic_wide[, traits, drop = FALSE])
)
```

## 12. Canonical analytical outputs

Module 04 is the canonical source for derived results. The RDS object is language-neutral; presentation labels are added only in Module 05.

```{r save-results}
write.csv(model_comparison, file.path(RESULTS_DIR, "model_comparison.csv"), row.names = FALSE)
write.csv(model_selection, file.path(RESULTS_DIR, "model_selection.csv"), row.names = FALSE)
write.csv(selection_spec, file.path(RESULTS_DIR, "selection_decision_spec.csv"), row.names = FALSE)
write.csv(predictive_evidence_summary, file.path(RESULTS_DIR, "predictive_evidence_summary.csv"), row.names = FALSE)
write.csv(convergence_by_trait_method, file.path(RESULTS_DIR, "convergence_by_trait_method.csv"), row.names = FALSE)
write.csv(selected_fold_diagnostics, file.path(RESULTS_DIR, "selected_fold_diagnostics.csv"), row.names = FALSE)
write.csv(runtime_summary, file.path(RESULTS_DIR, "runtime_summary.csv"), row.names = FALSE)
write.csv(rrblup_heritability_by_fold, file.path(RESULTS_DIR, "rrblup_heritability_by_fold.csv"), row.names = FALSE)
write.csv(rrblup_heritability_summary, file.path(RESULTS_DIR, "rrblup_heritability_summary.csv"), row.names = FALSE)
write.csv(selected_oof_genomic_values, file.path(RESULTS_DIR, "selected_oof_genomic_values.csv"), row.names = FALSE)
write.csv(genomic_ranking, file.path(RESULTS_DIR, "genomic_ranking.csv"), row.names = FALSE)
write.csv(top10_with_family, file.path(RESULTS_DIR, "top10_genomic_values_with_family.csv"), row.names = FALSE)
write.csv(recurring_top10, file.path(RESULTS_DIR, "recurring_top10_individuals.csv"), row.names = FALSE)
write.csv(as.data.frame(selected_oof_correlations), file.path(RESULTS_DIR, "selected_oof_correlations.csv"), row.names = TRUE)

saveRDS(
  list(
    signature = models_object$signature,
    traits = traits,
    methods = methods,
    model_comparison = model_comparison,
    selection_spec = selection_spec,
    model_selection = model_selection,
    predictive_evidence_summary = predictive_evidence_summary,
    convergence_by_trait_method = convergence_by_trait_method,
    selected_fold_diagnostics = selected_fold_diagnostics,
    runtime = runtime_summary,
    rrblup_heritability_by_fold = rrblup_heritability_by_fold,
    rrblup_heritability_summary = rrblup_heritability_summary,
    selected_oof_genomic_values = selected_oof_genomic_values,
    genomic_ranking = genomic_ranking,
    top10_with_family = top10_with_family,
    recurring_top10 = recurring_top10,
    selected_oof_correlations = selected_oof_correlations
  ),
  file.path(RESULTS_DIR, "module04_results.rds")
)
```

## 13. Next step

Module 05 consumes the results derived here to build final tables and figures. No analytical decision is redefined at that stage.
