---
title: "Derivation and selection of genomic results"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
```

## 1. Objective and analytical contract

This module transforms the official outputs from Module 03 into derived analytical results. No model is refitted and no final-presentation figure is produced here.

Derived outputs include OOF predictive performance; DIC, DIC delta, `Wprob`, and `ER`; trait-specific method decisions; fold stability; computational time; heritability and accuracy in the RR-BLUP/BRR analogue; OOF genomic values; rankings; top 10 individuals; recurrence across rankings; and correlations among genomic values.

Method selection follows a reproducible hierarchical rule integrating OOF performance, relative DIC fit, and chain stability. When methods are practically equivalent, RR-BLUP is preferred; decisions involving criterion conflicts or convergence flags remain explicitly marked for review.

Genomic values used for ranking are strictly OOF: each individual receives the genomic component from the fold in which its phenotype was masked. No multi-trait selection is performed.

## 2. Official inputs

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(here)
})
```

```{r inputs}
MATRICES_FILE <- here("output", "matrices", "matrices_object.rds")
MODELS_FILE <- here("output", "models", "models_object.rds")
RESULTS_DIR <- here("output", "results", "m04")

dir.create(RESULTS_DIR, recursive = TRUE, showWarnings = FALSE)

matrices_object <- readRDS(MATRICES_FILE)
models_object <- readRDS(MODELS_FILE)

# Integrity gate: M04 only accepts a complete M03 object generated from the
# current M02 matrices and analytical settings.
expected_traits <- colnames(matrices_object$phenotypes)
expected_methods <- c(
  "RRBLUP", "BayesA", "BayesB", "BayesB2",
  "BayesC", "BL", "GBLUP"
)
expected_folds <- seq_len(
  matrices_object$settings$n_folds
)

expected_signature <- substr(
  digest::digest(
    list(
      ids = matrices_object$ids,
      phenotypes = matrices_object$phenotypes,
      X = matrices_object$X,
      M = matrices_object$M,
      G = matrices_object$G,
      cv_folds = matrices_object$cv_folds,
      settings = models_object$settings
    ),
    algo = "sha256"
  ),
  1,
  16
)

if (
  length(models_object$signature) != 1L ||
  !identical(
    as.character(models_object$signature),
    expected_signature
  )
) {
  stop(
    paste0(
      "M04 integrity gate failed: models_object.rds does not match ",
      "the current matrices_object.rds analytical signature. ",
      "Rebuild M03 from the current M02 outputs before running M04."
    )
  )
}

settings_contract_ok <-
  identical(
    as.character(models_object$settings$traits),
    expected_traits
  ) &&
  identical(
    as.integer(models_object$settings$folds),
    as.integer(expected_folds)
  ) &&
  identical(
    as.character(models_object$settings$methods),
    expected_methods
  )

if (!settings_contract_ok) {
  stop(
    paste0(
      "M04 integrity gate failed: the M03 trait/fold/method settings ",
      "do not match the current project contract."
    )
  )
}

metrics <- models_object$metrics_by_fold
predictions <- models_object$predictions
convergence <- models_object$convergence_diagnostics

expected_runs <-
  length(expected_traits) *
  length(expected_folds) *
  length(expected_methods)

expected_predictions <-
  length(matrices_object$ids) *
  length(expected_traits) *
  length(expected_methods)

expected_convergence_rows <-
  expected_runs * 2L

if (nrow(metrics) != expected_runs) {
  stop(
    sprintf(
      paste0(
        "M04 integrity gate failed: M03 is incomplete. ",
        "Expected %d trait-fold-method fits, found %d."
      ),
      expected_runs,
      nrow(metrics)
    )
  )
}

if (nrow(predictions) != expected_predictions) {
  stop(
    sprintf(
      paste0(
        "M04 integrity gate failed: expected %d OOF predictions, found %d."
      ),
      expected_predictions,
      nrow(predictions)
    )
  )
}

if (nrow(convergence) != expected_convergence_rows) {
  stop(
    sprintf(
      paste0(
        "M04 integrity gate failed: expected %d convergence records, found %d."
      ),
      expected_convergence_rows,
      nrow(convergence)
    )
  )
}

signature_columns_ok <-
  all(metrics$Signature == expected_signature) &&
  all(predictions$Signature == expected_signature) &&
  all(convergence$Signature == expected_signature)

if (!signature_columns_ok) {
  stop(
    "M04 integrity gate failed: mixed analytical signatures were found in M03 outputs."
  )
}

metric_keys <- metrics |>
  distinct(Trait, Fold, Method)

prediction_keys <- predictions |>
  distinct(IND, Trait, Method)

convergence_keys <- convergence |>
  count(Trait, Fold, Method, name = "N_parameters")

if (
  nrow(metric_keys) != expected_runs ||
  anyDuplicated(metric_keys) > 0L
) {
  stop(
    "M04 integrity gate failed: duplicated or missing trait-fold-method metric keys."
  )
}

if (
  nrow(prediction_keys) != expected_predictions ||
  anyDuplicated(prediction_keys) > 0L
) {
  stop(
    "M04 integrity gate failed: OOF predictions are not unique by individual, trait, and method."
  )
}

if (
  nrow(convergence_keys) != expected_runs ||
  any(convergence_keys$N_parameters != 2L)
) {
  stop(
    "M04 integrity gate failed: each M03 fit must contribute exactly two convergence records."
  )
}

metric_coverage_ok <-
  setequal(unique(metrics$Trait), expected_traits) &&
  setequal(unique(metrics$Fold), expected_folds) &&
  setequal(unique(metrics$Method), expected_methods)

if (!metric_coverage_ok) {
  stop(
    "M04 integrity gate failed: trait, fold, or method coverage is incomplete."
  )
}

metrics_per_trait_method <- metrics |>
  count(Trait, Method, name = "N_folds")

predictions_per_trait_method <- predictions |>
  count(Trait, Method, name = "N_predictions")

if (
  nrow(metrics_per_trait_method) !=
    length(expected_traits) * length(expected_methods) ||
  any(metrics_per_trait_method$N_folds != length(expected_folds))
) {
  stop(
    "M04 integrity gate failed: every trait-method combination must contain all five folds."
  )
}

if (
  nrow(predictions_per_trait_method) !=
    length(expected_traits) * length(expected_methods) ||
  any(
    predictions_per_trait_method$N_predictions !=
      length(matrices_object$ids)
  )
) {
  stop(
    "M04 integrity gate failed: every trait-method combination must contain one OOF prediction per individual."
  )
}

expected_training_n <- setNames(
  matrices_object$fold_summary$Training_n,
  matrices_object$fold_summary$Fold
)

expected_validation_n <- setNames(
  matrices_object$fold_summary$Validation_n,
  matrices_object$fold_summary$Fold
)

observed_training_expected <- unname(
  expected_training_n[
    as.character(metrics$Fold)
  ]
)

observed_validation_expected <- unname(
  expected_validation_n[
    as.character(metrics$Fold)
  ]
)

if (
  any(metrics$Training_n != observed_training_expected) ||
  any(metrics$Validation_n != observed_validation_expected)
) {
  stop(
    "M04 integrity gate failed: training/validation sample sizes do not match the M02 fold contract."
  )
}

if (
  !all(metrics$Finite_predictions) ||
  !all(metrics$Finite_metrics) ||
  !all(is.finite(predictions$Observed)) ||
  !all(is.finite(predictions$Predicted))
) {
  stop(
    "M04 integrity gate failed: non-finite predictions or fit metrics require review before derivation."
  )
}

traits <- expected_traits
methods <- expected_methods
```

Before any derivation, the module verifies the analytical signature, completion
of all 350 fits, coverage of ten traits, five folds, and seven methods,
uniqueness of OOF predictions, and training/validation sample sizes. Therefore,
a partial Module 03 test cannot accidentally be used as a final analytical
result.

## 3. OOF predictive performance

Primary metrics are calculated from the 150 out-of-sample predictions for each trait-method combination, preserving exactly one OOF prediction per individual.

```{r pooled-oof-metrics}
pooled_metrics <- predictions |>
  group_by(Trait, Method) |>
  summarise(
    N = n(),
    r = cor(Observed, Predicted),
    RMSE = sqrt(mean((Observed - Predicted)^2)),
    MAE = mean(abs(Observed - Predicted)),
    Bias = mean(Predicted - Observed),
    Slope = cov(Observed, Predicted) / var(Predicted),
    .groups = "drop"
  ) |>
  mutate(
    t_value = r * sqrt((N - 2) / (1 - r^2)),
    p_value = 2 * pt(abs(t_value), df = N - 2, lower.tail = FALSE)
  ) |>
  select(-t_value)

fold_summary <- metrics |>
  group_by(Trait, Method) |>
  summarise(
    Mean_DIC = mean(DIC),
    SD_DIC = sd(DIC),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_fold_RMSE = mean(RMSE),
    SD_fold_RMSE = sd(RMSE),
    Mean_fold_MAE = mean(MAE),
    SD_fold_MAE = sd(MAE),
    .groups = "drop"
  )
```

Pearson-correlation p-values are retained only as nominal descriptive measures because cross-validation training sets overlap across folds.

## 4. Method comparison

Silva et al. [@silva2021brrguava] calculated approximate model probabilities from each model's DIC difference relative to the minimum DIC. The same logic is applied jointly to the seven methods in this project.

\[
\Delta_j=DIC_j-\min(DIC),\qquad
Wprob_j=\frac{\exp(-\Delta_j/2)}{\sum_k\exp(-\Delta_k/2)},\qquad
ER_j=\frac{Wprob_{\max}}{Wprob_j}.
\]

```{r model-comparison}
model_comparison <- pooled_metrics |>
  left_join(fold_summary, by = c("Trait", "Method")) |>
  group_by(Trait) |>
  mutate(
    Delta_DIC = Mean_DIC - min(Mean_DIC),
    log_weight = -0.5 * Delta_DIC,
    stable_weight = exp(log_weight - max(log_weight)),
    Wprob = stable_weight / sum(stable_weight),
    ER = max(Wprob) / Wprob
  ) |>
  ungroup() |>
  select(
    Trait, Method, Mean_DIC, SD_DIC, Delta_DIC, Wprob, ER,
    N, r, p_value, RMSE, MAE, Bias, Slope,
    Mean_fold_r, SD_fold_r, Mean_fold_RMSE, SD_fold_RMSE,
    Mean_fold_MAE, SD_fold_MAE
  ) |>
  arrange(match(Trait, traits), Delta_DIC)
```

## 5. Method selection by trait

Selection follows an **explicit hierarchy** rather than combining the criteria
into a single score. Out-of-sample predictive ability is the primary
criterion; DIC provides complementary evidence about fit and complexity; fold
stability and MCMC diagnostics indicate the robustness of the decision.

### Interpretation of the model-selection parameters

- `r`: Pearson correlation between observed and OOF-predicted values. Higher is
  better and this is the primary predictive reference.
- `p_value`: nominal test of \(r=0\). It is descriptive only because
  cross-validation training sets overlap across folds.
- `RMSE`: root mean squared error in the trait's original unit. Lower is better
  and large errors receive greater penalty.
- `MAE`: mean absolute error in the trait's original unit. Lower is better and
  it is less sensitive to extreme errors than RMSE.
- `Bias`: mean of `Predicted - Observed`. Values close to zero are preferred.
- `Slope`: slope from regressing observed on predicted values. Values close to
  1 indicate better calibration.
- `Mean_DIC`: mean DIC across folds. Lower values indicate a better compromise
  between fit and complexity; DIC is not a direct measure of prediction.
- `Delta_DIC`: difference from the minimum mean DIC for the trait. Zero
  identifies the best mean DIC.
- `Wprob`: normalized approximate weight derived from `Delta_DIC`. Larger
  values indicate stronger relative DIC support; it is not interpreted as an
  exact posterior probability of the model.
- `ER`: ratio of the largest `Wprob` to the method's `Wprob`. The best value is
  1; larger values indicate weaker relative support.
- `Mean_fold_r` and `SD_fold_r`: mean and variability of fold correlations.
- `Mean_fold_RMSE` and `SD_fold_RMSE`: mean and variability of fold RMSE.
- `Minimum_ESS`, `Maximum_absolute_Geweke`, `Parameters_flagged`, and
  `Runs_with_flags`: chain diagnostics. They indicate when review is required
  and are not biological criteria.
- computational time is a secondary operational criterion and should decide
  only between scientifically equivalent methods.

### Operational equivalence criterion

To avoid selecting a different method because of a negligible numerical
advantage, the project uses an **operational equivalence criterion based on one
standard error across folds**. Because cross-validation training sets overlap,
this criterion is not a formal equivalence test between models.

A method is predictively competitive when:

1. its mean `r` is no more than one standard error below the best mean `r`; and
2. its mean RMSE is no more than one standard error above the best mean RMSE.

DIC fit is considered similar under the operational threshold
`Delta_DIC <= 2`. This threshold is also not a formal equivalence test.

The competitive set for selection is the intersection of predictive
equivalence and DIC similarity.

### RR-BLUP preference and convergence review

RR-BLUP preference is applied only when RR-BLUP belongs to the competitive set
**and has no convergence flags**.

- If RR-BLUP is also the predictive winner, it is selected on predictive merit
  and `RRBLUP_preference_applied = FALSE`.
- If another method is the predictive winner but RR-BLUP belongs to the
  competitive set without flags, RR-BLUP is preferred and
  `RRBLUP_preference_applied = TRUE`.
- If RR-BLUP is not an automatic candidate, the best flag-free competitive
  method is selected by `r`, followed by RMSE, MAE, and DIC.
- If no method simultaneously satisfies predictive equivalence, DIC similarity,
  and absence of convergence flags, the predictive winner is retained only
  provisionally and `Selection_requires_review = TRUE`.

The preference for RR-BLUP in equivalent situations is deliberate. In the
current workflow it uses a common marker-effect variance and a residual
variance, allowing a direct decomposition of additive variance and
heritability. This does not mean that the other Bayesian models cannot provide
variance information; the advantage here is the simpler and more directly
interpretable parameterization used by this project.

```{r model-selection}
DIC_EQUIVALENCE_THRESHOLD <- 2
N_FOLDS <- matrices_object$settings$n_folds

predictive_winners <- model_comparison |>
  group_by(Trait) |>
  arrange(desc(r), RMSE, MAE, Mean_DIC, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    Predictive_winner = Method,
    Predictive_winner_r = r,
    Predictive_winner_p_value = p_value,
    Predictive_winner_RMSE = RMSE,
    Predictive_winner_MAE = MAE
  )

dic_winners <- model_comparison |>
  group_by(Trait) |>
  arrange(Mean_DIC, desc(r), RMSE, MAE, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    DIC_winner = Method,
    Minimum_DIC = Mean_DIC,
    DIC_winner_Wprob = Wprob
  )

convergence_by_trait_method <- convergence |>
  group_by(Trait, Method) |>
  summarise(
    Runs = n_distinct(Run_id),
    Parameters_reviewed = n(),
    Parameters_flagged = sum(Review_flag),
    Runs_with_flags = n_distinct(Run_id[Review_flag]),
    Minimum_ESS = min(Effective_sample_size, na.rm = TRUE),
    Maximum_absolute_Geweke = max(abs(Geweke_Z), na.rm = TRUE),
    .groups = "drop"
  )

model_selection_support <- model_comparison |>
  left_join(
    convergence_by_trait_method,
    by = c("Trait", "Method")
  ) |>
  left_join(
    predictive_winners |>
      select(Trait, Predictive_winner),
    by = "Trait"
  ) |>
  group_by(Trait) |>
  mutate(
    Best_mean_fold_r = max(Mean_fold_r, na.rm = TRUE),
    Best_mean_fold_r_SE =
      SD_fold_r[which.max(Mean_fold_r)] / sqrt(N_FOLDS),
    Best_mean_fold_RMSE = min(Mean_fold_RMSE, na.rm = TRUE),
    Best_mean_fold_RMSE_SE =
      SD_fold_RMSE[which.min(Mean_fold_RMSE)] / sqrt(N_FOLDS),
    Within_one_SE_r =
      Mean_fold_r >=
      Best_mean_fold_r - Best_mean_fold_r_SE,
    Within_one_SE_RMSE =
      Mean_fold_RMSE <=
      Best_mean_fold_RMSE + Best_mean_fold_RMSE_SE,
    Predictively_competitive =
      Within_one_SE_r & Within_one_SE_RMSE,
    DIC_equivalence_threshold =
      DIC_EQUIVALENCE_THRESHOLD,
    DIC_competitive =
      Delta_DIC <= DIC_EQUIVALENCE_THRESHOLD,
    Competitive_for_selection =
      Predictively_competitive & DIC_competitive,
    Convergence_review =
      Parameters_flagged > 0,
    Automatic_candidate =
      Competitive_for_selection & !Convergence_review,
    Competitive_set_n =
      sum(Competitive_for_selection),
    Automatic_candidate_n =
      sum(Automatic_candidate),
    RRBLUP_competitive =
      any(
        Method == "RRBLUP" &
          Competitive_for_selection
      ),
    RRBLUP_automatic_candidate =
      any(
        Method == "RRBLUP" &
          Automatic_candidate
      ),
    Automatic_selection_available =
      any(Automatic_candidate),
    Selection_priority = case_when(
      Automatic_candidate &
        Method == "RRBLUP" ~ 1L,
      Automatic_candidate ~ 2L,
      TRUE ~ 3L
    )
  ) |>
  ungroup() |>
  arrange(
    match(Trait, traits),
    Selection_priority,
    desc(r),
    RMSE,
    MAE,
    Mean_DIC
  )

automatic_selections <- model_selection_support |>
  filter(Automatic_candidate) |>
  group_by(Trait) |>
  arrange(
    Selection_priority,
    desc(r),
    RMSE,
    MAE,
    Mean_DIC,
    .by_group = TRUE
  ) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    Automatic_selected_method = Method
  )

selection_spec <- predictive_winners |>
  select(Trait, Predictive_winner) |>
  left_join(
    automatic_selections,
    by = "Trait"
  ) |>
  mutate(
    Selected_method = coalesce(
      Automatic_selected_method,
      Predictive_winner
    )
  ) |>
  select(Trait, Selected_method, Predictive_winner) |>
  left_join(
    model_selection_support |>
      select(
        Trait,
        Method,
        Automatic_selection_available,
        Automatic_candidate_n,
        Competitive_set_n,
        RRBLUP_competitive,
        RRBLUP_automatic_candidate,
        Convergence_review
      ),
    by = c(
      "Trait",
      "Selected_method" = "Method"
    )
  ) |>
  mutate(
    RRBLUP_preference_applied =
      Selected_method == "RRBLUP" &
      Predictive_winner != "RRBLUP" &
      RRBLUP_automatic_candidate,
    Selection_requires_review =
      !Automatic_selection_available |
      Convergence_review,
    Selection_status = if_else(
      Selection_requires_review,
      "provisional_review_required",
      "automatic_supported"
    ),
    Decision_code = case_when(
      Selection_requires_review ~
        "predictive_winner_provisional_review_required",
      Selected_method == "RRBLUP" &
        RRBLUP_preference_applied ~
        "rrblup_preferred_among_equivalent_models",
      Selected_method == "RRBLUP" &
        Predictive_winner == "RRBLUP" ~
        "rrblup_best_predictive_model",
      TRUE ~
        "best_non_rrblup_equivalent_model"
    ),
    Operational_equivalence_rule =
      "one_se_r_and_rmse_plus_delta_dic_le_2"
  ) |>
  select(
    -Predictive_winner,
    -Convergence_review
  ) |>
  arrange(match(Trait, traits))

selected_performance <- model_selection_support |>
  inner_join(
    selection_spec |>
      select(Trait, Selected_method),
    by = "Trait"
  ) |>
  filter(Method == Selected_method) |>
  transmute(
    Trait,
    Selected_method,
    r,
    p_value,
    RMSE,
    MAE,
    Bias,
    Slope,
    Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    Mean_fold_r,
    SD_fold_r,
    Mean_fold_RMSE,
    SD_fold_RMSE,
    Within_one_SE_r,
    Within_one_SE_RMSE,
    Predictively_competitive,
    DIC_competitive,
    Competitive_for_selection,
    Automatic_candidate,
    Competitive_set_n,
    Automatic_candidate_n,
    RRBLUP_competitive,
    RRBLUP_automatic_candidate,
    Parameters_flagged,
    Runs_with_flags,
    Minimum_ESS,
    Maximum_absolute_Geweke,
    Convergence_review
  )

model_selection <- selection_spec |>
  left_join(
    selected_performance,
    by = c("Trait", "Selected_method")
  ) |>
  left_join(
    predictive_winners,
    by = "Trait"
  ) |>
  left_join(
    dic_winners,
    by = "Trait"
  ) |>
  mutate(
    Predictive_loss =
      Predictive_winner_r - r,
    Predictive_winner_agrees =
      Selected_method == Predictive_winner,
    DIC_winner_agrees =
      Selected_method == DIC_winner,
    Predictive_evidence = case_when(
      r > 0 & p_value < 0.05 ~
        "positive_p_lt_0.05",
      r > 0 & p_value >= 0.05 ~
        "positive_p_ge_0.05",
      TRUE ~
        "nonpositive_p_ge_0.05"
    )
  ) |>
  arrange(match(Trait, traits))

predictive_evidence_summary <- model_selection |>
  count(Predictive_evidence, name = "Traits") |>
  arrange(Predictive_evidence)
```

## 6. Fold stability of selected methods

Values are derived here; their graphical presentation belongs exclusively to Module 05.

```{r selected-fold-diagnostics}
selected_fold_diagnostics <- metrics |>
  inner_join(
    selection_spec |> select(Trait, Selected_method),
    by = c("Trait" = "Trait", "Method" = "Selected_method")
  ) |>
  transmute(Trait, Fold, Method, Correlation, RMSE, MAE) |>
  arrange(match(Trait, traits), Fold)
```

## 7. Computational time

Only elapsed times from the 350 official Module 03 fits are used; no additional runs are executed solely for timing.

```{r runtime}
runtime_summary <- metrics |>
  group_by(Method) |>
  summarise(
    Fits = n(),
    Total_seconds = sum(Elapsed_seconds),
    Mean_seconds = mean(Elapsed_seconds),
    SD_seconds = sd(Elapsed_seconds),
    Median_seconds = median(Elapsed_seconds),
    Minimum_seconds = min(Elapsed_seconds),
    Maximum_seconds = max(Elapsed_seconds),
    .groups = "drop"
  ) |>
  arrange(match(Method, methods))
```

## 8. Heritability and accuracy in the RR-BLUP/BRR analogue

Following Silva et al. [@silva2021brrguava], additive genetic variance in the RR-BLUP/BRR analogue is calculated from the common marker variance and the \(\sum 2pq\) denominator. Fold accuracy is \(r_f/\sqrt{H_f^2}\).

```{r rrblup-heritability}
rrblup_variance_components <- convergence |>
  filter(
    Method == "RRBLUP",
    Parameter %in% c("Residual_variance", "Marker_variance")
  ) |>
  select(Run_id, Trait, Fold, Method, Parameter, Posterior_mean) |>
  pivot_wider(names_from = Parameter, values_from = Posterior_mean)

rrblup_fold_correlations <- metrics |>
  filter(Method == "RRBLUP") |>
  select(Trait, Fold, Correlation)

rrblup_heritability_by_fold <- rrblup_variance_components |>
  left_join(rrblup_fold_correlations, by = c("Trait", "Fold")) |>
  mutate(
    Marker_variance_denominator = matrices_object$denominator,
    Additive_variance = Marker_variance * Marker_variance_denominator,
    H2 = Additive_variance / (Additive_variance + Residual_variance),
    Predictive_accuracy = Correlation / sqrt(H2)
  ) |>
  arrange(match(Trait, traits), Fold)

rrblup_heritability_summary <- rrblup_heritability_by_fold |>
  group_by(Trait) |>
  summarise(
    Mean_H2 = mean(H2),
    SD_H2 = sd(H2),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_predictive_accuracy = mean(Predictive_accuracy),
    SD_predictive_accuracy = sd(Predictive_accuracy),
    .groups = "drop"
  ) |>
  arrange(match(Trait, traits))
```

## 9. OOF genomic values

```{r selected-oof-genomic-values}
selected_oof_genomic_values <- predictions |>
  inner_join(
    model_selection |> select(Trait, Selected_method, Predictive_evidence),
    by = "Trait"
  ) |>
  filter(Method == Selected_method) |>
  transmute(
    IND, Trait, Fold, Method, Predictive_evidence,
    Genomic_value = Genomic_component
  ) |>
  arrange(match(Trait, traits), IND)
```

## 10. Ranking, top 10, and recurrence

```{r genomic-ranking}
genomic_ranking <- selected_oof_genomic_values |>
  group_by(Trait) |>
  arrange(desc(Genomic_value), IND, .by_group = TRUE) |>
  mutate(Rank = row_number()) |>
  ungroup() |>
  arrange(match(Trait, traits), Rank)

top10_with_family <- genomic_ranking |>
  filter(Rank <= 10L) |>
  left_join(
    matrices_object$metadata |> select(IND, FAM),
    by = "IND"
  ) |>
  select(
    Trait, Rank, IND, FAM, Method, Genomic_value, Predictive_evidence
  ) |>
  arrange(match(Trait, traits), Rank)

recurring_top10 <- top10_with_family |>
  filter(Predictive_evidence == "positive_p_lt_0.05") |>
  group_by(IND, FAM) |>
  summarise(
    N_traits = n(),
    Traits = paste(Trait, collapse = ", "),
    .groups = "drop"
  ) |>
  filter(N_traits >= 2L) |>
  arrange(desc(N_traits), FAM, IND)
```

Recurrence across rankings is descriptive and does not constitute a multi-trait selection index.

## 11. Correlations among OOF genomic values

These are descriptive correlations among predicted genomic values; they are not multivariate genetic-correlation estimates.

```{r oof-correlations}
selected_genomic_wide <- selected_oof_genomic_values |>
  select(IND, Trait, Genomic_value) |>
  pivot_wider(names_from = Trait, values_from = Genomic_value)

selected_oof_correlations <- cor(
  as.matrix(selected_genomic_wide[, traits, drop = FALSE])
)
```

## 12. Canonical analytical outputs

Module 04 is the canonical source for derived results. The RDS object is language-neutral; presentation labels are added only in Module 05.

```{r save-results}
write.csv(model_comparison, file.path(RESULTS_DIR, "model_comparison.csv"), row.names = FALSE)
write.csv(model_selection_support, file.path(RESULTS_DIR, "model_selection_support.csv"), row.names = FALSE)
write.csv(model_selection, file.path(RESULTS_DIR, "model_selection.csv"), row.names = FALSE)
write.csv(selection_spec, file.path(RESULTS_DIR, "selection_decision_spec.csv"), row.names = FALSE)
write.csv(predictive_evidence_summary, file.path(RESULTS_DIR, "predictive_evidence_summary.csv"), row.names = FALSE)
write.csv(convergence_by_trait_method, file.path(RESULTS_DIR, "convergence_by_trait_method.csv"), row.names = FALSE)
write.csv(selected_fold_diagnostics, file.path(RESULTS_DIR, "selected_fold_diagnostics.csv"), row.names = FALSE)
write.csv(runtime_summary, file.path(RESULTS_DIR, "runtime_summary.csv"), row.names = FALSE)
write.csv(rrblup_heritability_by_fold, file.path(RESULTS_DIR, "rrblup_heritability_by_fold.csv"), row.names = FALSE)
write.csv(rrblup_heritability_summary, file.path(RESULTS_DIR, "rrblup_heritability_summary.csv"), row.names = FALSE)
write.csv(selected_oof_genomic_values, file.path(RESULTS_DIR, "selected_oof_genomic_values.csv"), row.names = FALSE)
write.csv(genomic_ranking, file.path(RESULTS_DIR, "genomic_ranking.csv"), row.names = FALSE)
write.csv(top10_with_family, file.path(RESULTS_DIR, "top10_genomic_values_with_family.csv"), row.names = FALSE)
write.csv(recurring_top10, file.path(RESULTS_DIR, "recurring_top10_individuals.csv"), row.names = FALSE)
write.csv(as.data.frame(selected_oof_correlations), file.path(RESULTS_DIR, "selected_oof_correlations.csv"), row.names = TRUE)

saveRDS(
  list(
    signature = models_object$signature,
    traits = traits,
    methods = methods,
    model_comparison = model_comparison,
    model_selection_support = model_selection_support,
    selection_spec = selection_spec,
    model_selection = model_selection,
    predictive_evidence_summary = predictive_evidence_summary,
    convergence_by_trait_method = convergence_by_trait_method,
    selected_fold_diagnostics = selected_fold_diagnostics,
    runtime = runtime_summary,
    rrblup_heritability_by_fold = rrblup_heritability_by_fold,
    rrblup_heritability_summary = rrblup_heritability_summary,
    selected_oof_genomic_values = selected_oof_genomic_values,
    genomic_ranking = genomic_ranking,
    top10_with_family = top10_with_family,
    recurring_top10 = recurring_top10,
    selected_oof_correlations = selected_oof_correlations
  ),
  file.path(RESULTS_DIR, "module04_results.rds")
)
```

## 13. Next step

Module 05 consumes the results derived here to build final tables and figures. No analytical decision is redefined at that stage.
