---
title: "Model comparison and derivation of genomic results"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective and analytical contract

This module uses only the official results already produced in Module 03.
No model is refitted.

The analysis follows the structure of Silva et al. [@silva2021brrguava] and
adapts it to the methods actually fitted in this project. The article compared
six Bayesian models; here all seven fitted methods are compared jointly:

1. RR-BLUP, implemented with the BGLR `BRR` component;
2. BayesA;
3. BayesB;
4. BayesB2;
5. BayesC;
6. Bayesian Lasso;
7. GBLUP.

The comparison includes DIC, DIC delta, approximate model probability
(`Wprob`), evidence ratio (`ER`), and out-of-sample predictive performance.

The validation design is not changed to imitate the fold size used in the
article. The project retains its validated five deterministic
family-stratified folds, with 120 training and 30 validation individuals in
each fold.

Because the article does not provide a single scalar rule combining DIC and
predictive ability, this project explicitly integrates three sources of
evidence: OOF performance, relative fit by DIC, and chain stability. The
highest OOF `r` remains the predictive reference, but small predictive
differences are not used automatically to override a model with better DIC or
clearer chain stability. The final decision for each trait is stored in an
explicit, auditable table, without selective model refitting.

Genomic values used for ranking follow the cross-validation logic: each
individual receives the genomic value from the fold in which its phenotype was
masked. The five fits are not averaged and no additional full-phenotype fit is
performed.

Multi-trait selection is not performed in this stage.

## 2. Official inputs

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
  library(here)
})
```

```{r paths-and-inputs}
MATRICES_FILE <- here(
  "output",
  "matrices",
  "matrices_object.rds"
)

MODELS_DIR <- here(
  "output",
  "models"
)

MODELS_FILE <- file.path(
  MODELS_DIR,
  "models_object.rds"
)

RESULTS_DIR <- here(
  "output",
  "results",
  "m04"
)

FIGURE_DIR <- file.path(
  RESULTS_DIR,
  "figures"
)

dir.create(
  RESULTS_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  FIGURE_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

matrices_object <- readRDS(
  MATRICES_FILE
)

models_object <- readRDS(
  MODELS_FILE
)

metrics <- models_object$metrics_by_fold
predictions <- models_object$predictions
convergence <- models_object$convergence_diagnostics

traits <- colnames(
  matrices_object$phenotypes
)

methods <- c(
  "RRBLUP",
  "BayesA",
  "BayesB",
  "BayesB2",
  "BayesC",
  "BL",
  "GBLUP"
)

method_labels <- c(
  RRBLUP = "RR-BLUP/BRR",
  BayesA = "BayesA",
  BayesB = "BayesB",
  BayesB2 = "BayesB2",
  BayesC = "BayesC",
  BL = "Bayesian Lasso",
  GBLUP = "GBLUP"
)

```

## 3. OOF predictive performance

Primary metrics are calculated from the 150 out-of-sample predictions for
each trait-method combination. This consolidation preserves exactly one OOF
prediction per individual.

```{r pooled-oof-metrics}
pooled_metrics <- predictions |>
  group_by(
    Trait,
    Method
  ) |>
  summarise(
    N = n(),
    r = cor(
      Observed,
      Predicted
    ),
    RMSE = sqrt(
      mean(
        (
          Observed -
            Predicted
        )^2
      )
    ),
    MAE = mean(
      abs(
        Observed -
          Predicted
      )
    ),
    Bias = mean(
      Predicted -
        Observed
    ),
    Slope =
      cov(
        Observed,
        Predicted
      ) /
      var(
        Predicted
      ),
    .groups = "drop"
  )

pooled_metrics <- pooled_metrics |>
  mutate(
    t_value =
      r *
      sqrt(
        (N - 2) /
          (1 - r^2)
      ),
    p_value =
      2 *
      pt(
        abs(t_value),
        df = N - 2,
        lower.tail = FALSE
      )
  ) |>
  select(
    -t_value
  )
```


The Pearson-correlation p-values are retained to preserve comparability with
the reference analysis. Because cross-validation training sets overlap across
folds, these values are treated as nominal and descriptive rather than as
fully cross-validation-aware inferential tests.

```{r fold-summary}
fold_summary <- metrics |>
  group_by(
    Trait,
    Method
  ) |>
  summarise(
    Mean_DIC = mean(
      DIC
    ),
    SD_DIC = sd(
      DIC
    ),
    Mean_fold_r = mean(
      Correlation
    ),
    SD_fold_r = sd(
      Correlation
    ),
    Mean_fold_RMSE = mean(
      RMSE
    ),
    SD_fold_RMSE = sd(
      RMSE
    ),
    Mean_fold_MAE = mean(
      MAE
    ),
    SD_fold_MAE = sd(
      MAE
    ),
    .groups = "drop"
  )
```

## 4. DIC, delta, Wprob, and evidence ratio

Silva et al. [@silva2021brrguava] calculated the approximate probability of
each model from the difference between its DIC and the minimum DIC. In this
project, normalization is extended from the six models in the article to all
seven methods actually fitted.

\[
\Delta_j = DIC_j - \min(DIC)
\]

\[
Wprob_j =
\frac{\exp(-\Delta_j/2)}
{\sum_k \exp(-\Delta_k/2)}
\]

\[
ER_j =
\frac{Wprob_{\max}}
{Wprob_j}
\]

```{r model-comparison}
model_comparison <- pooled_metrics |>
  left_join(
    fold_summary,
    by = c(
      "Trait",
      "Method"
    )
  ) |>
  mutate(
    Method_label =
      unname(
        method_labels[
          Method
        ]
      )
  ) |>
  group_by(
    Trait
  ) |>
  mutate(
    Delta_DIC =
      Mean_DIC -
      min(
        Mean_DIC
      ),
    log_weight =
      -0.5 *
      Delta_DIC,
    stable_weight =
      exp(
        log_weight -
          max(
            log_weight
          )
      ),
    Wprob =
      stable_weight /
      sum(
        stable_weight
      ),
    ER =
      max(
        Wprob
      ) /
      Wprob
  ) |>
  ungroup() |>
  select(
    Trait,
    Method,
    Method_label,
    Mean_DIC,
    SD_DIC,
    Delta_DIC,
    Wprob,
    ER,
    N,
    r,
    p_value,
    RMSE,
    MAE,
    Bias,
    Slope,
    Mean_fold_r,
    SD_fold_r,
    Mean_fold_RMSE,
    SD_fold_RMSE,
    Mean_fold_MAE,
    SD_fold_MAE
  ) |>
  arrange(
    match(
      Trait,
      traits
    ),
    Delta_DIC
  )

kable(
  model_comparison,
  digits = 5,
  caption = "Joint comparison of the seven methods by trait"
)
```



## 5. Final method selection by trait

The highest OOF correlation is retained as the predictive reference, whereas
the minimum DIC identifies the best relative fit. The final decision does not
apply an arbitrary numerical penalty that combines these two quantities.
Instead, it explicitly records cases in which a very small predictive
difference does not compensate for poorer relative fit or clear evidence of
chain instability.

The integrated comparison of OOF performance, DIC, and convergence led to the following final decisions:

- `MMCOM`: RR-BLUP, because the BL gain in `r` was small, whereas RR-BLUP had
  the minimum DIC and no flags;
- `PRODCOM`: BL, the winner for both prediction and DIC, while retaining its
  convergence warning;
- `PRODEF`: GBLUP, because the BL predictive advantage was weak and accompanied
  by extreme instability, whereas GBLUP had the minimum DIC and no flags;
- `CF`: BayesB, the prediction and DIC winner with no flags;
- `DF`: RR-BLUP, the prediction and DIC winner with no flags;
- `DCI`: RR-BLUP, with the highest `r`, DIC close to BayesC, and no flags;
- `EP`: BL, the relative predictive and DIC winner, although without a positive
  predictive signal;
- `FFC`: RR-BLUP, with the highest `r`, DIC very close to the minimum, and no
  flags;
- `FFP`: BL, the prediction and DIC winner, but with weak predictive evidence
  and convergence flags;
- `SS`: BayesC, because its correlation was essentially the same as BayesB,
  with lower DIC and no flags.

```{r model-selection}
predictive_ranking <- model_comparison |>
  group_by(
    Trait
  ) |>
  arrange(
    desc(r),
    RMSE,
    MAE,
    .by_group = TRUE
  ) |>
  mutate(
    Predictive_rank =
      row_number()
  ) |>
  ungroup()

predictive_winners <- predictive_ranking |>
  filter(
    Predictive_rank == 1L
  ) |>
  transmute(
    Trait,
    Predictive_winner = Method,
    Predictive_winner_r = r,
    Predictive_winner_p_value = p_value,
    Predictive_winner_RMSE = RMSE,
    Predictive_winner_MAE = MAE
  )

dic_winners <- model_comparison |>
  group_by(
    Trait
  ) |>
  arrange(
    Mean_DIC,
    .by_group = TRUE
  ) |>
  slice(
    1L
  ) |>
  ungroup() |>
  transmute(
    Trait,
    DIC_winner = Method,
    Minimum_DIC =
      Mean_DIC,
    DIC_winner_Wprob =
      Wprob
  )

convergence_by_trait_method <- convergence |>
  group_by(
    Trait,
    Method
  ) |>
  summarise(
    Runs =
      n_distinct(
        Run_id
      ),
    Parameters_reviewed =
      n(),
    Parameters_flagged =
      sum(
        Review_flag
      ),
    Runs_with_flags =
      n_distinct(
        Run_id[
          Review_flag
        ]
      ),
    Minimum_ESS =
      min(
        Effective_sample_size,
        na.rm = TRUE
      ),
    Maximum_absolute_Geweke =
      max(
        abs(
          Geweke_Z
        ),
        na.rm = TRUE
      ),
    .groups = "drop"
  )

selection_spec <- data.frame(
  Trait = c(
    "MMCOM",
    "PRODCOM",
    "PRODEF",
    "CF",
    "DF",
    "DCI",
    "EP",
    "FFC",
    "FFP",
    "SS"
  ),
  Selected_method = c(
    "RRBLUP",
    "BL",
    "GBLUP",
    "BayesB",
    "RRBLUP",
    "RRBLUP",
    "BL",
    "RRBLUP",
    "BL",
    "BayesC"
  ),
  Decision_code = c(
    "prefer_dic_stability_small_predictive_loss",
    "predictive_and_dic_winner_convergence_review",
    "prefer_stable_dic_winner_weak_prediction",
    "predictive_and_dic_winner",
    "predictive_and_dic_winner",
    "best_prediction_similar_dic_stable",
    "predictive_and_dic_winner_no_positive_signal",
    "best_prediction_similar_dic_stable_weak_signal",
    "predictive_and_dic_winner_weak_signal_convergence_review",
    "prefer_stable_dic_winner_negligible_predictive_loss"
  ),
  stringsAsFactors = FALSE
)

selected_performance <- model_comparison |>
  inner_join(
    selection_spec |>
      select(
        Trait,
        Selected_method
      ),
    by = "Trait"
  ) |>
  filter(
    Method ==
      Selected_method
  ) |>
  transmute(
    Trait,
    Selected_method,
    Selected_method_label =
      Method_label,
    r,
    p_value,
    RMSE,
    MAE,
    Bias,
    Slope,
    Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    Mean_fold_r,
    SD_fold_r
  )

model_selection <- selection_spec |>
  left_join(
    selected_performance,
    by = c(
      "Trait",
      "Selected_method"
    )
  ) |>
  left_join(
    predictive_winners,
    by = "Trait"
  ) |>
  left_join(
    dic_winners,
    by = "Trait"
  ) |>
  left_join(
    convergence_by_trait_method,
    by = c(
      "Trait",
      "Selected_method" = "Method"
    )
  ) |>
  mutate(
    Predictive_loss =
      Predictive_winner_r -
      r,
    Predictive_winner_agrees =
      Selected_method ==
      Predictive_winner,
    DIC_winner_agrees =
      Selected_method ==
      DIC_winner,
    Convergence_review =
      Parameters_flagged >
      0,
    Predictive_evidence =
      case_when(
        r > 0 &
          p_value < 0.05 ~
          "positive_p_lt_0.05",
        r > 0 &
          p_value >= 0.05 ~
          "positive_p_ge_0.05",
        TRUE ~
          "nonpositive_p_ge_0.05"
      )
  ) |>
  arrange(
    match(
      Trait,
      traits
    )
  )

decision_labels <- c(
  prefer_dic_stability_small_predictive_loss =
    "Lower DIC and greater stability with a small predictive loss",
  predictive_and_dic_winner_convergence_review =
    "Prediction and DIC winner; convergence requires review",
  prefer_stable_dic_winner_weak_prediction =
    "Lower DIC and stable chain under weak predictive ability",
  predictive_and_dic_winner =
    "Prediction and DIC winner with no flags",
  best_prediction_similar_dic_stable =
    "Highest prediction, DIC close to the minimum, and stable chain",
  predictive_and_dic_winner_no_positive_signal =
    "Relative prediction and DIC winner, but no positive predictive signal",
  best_prediction_similar_dic_stable_weak_signal =
    "Highest prediction and stability, but weak predictive evidence",
  predictive_and_dic_winner_weak_signal_convergence_review =
    "Prediction and DIC winner with weak evidence and convergence flags",
  prefer_stable_dic_winner_negligible_predictive_loss =
    "Lower DIC and greater stability with a negligible predictive loss"
)

predictive_evidence_labels <- c(
  positive_p_lt_0.05 =
    "Positive correlation with nominal p < 0.05",
  positive_p_ge_0.05 =
    "Positive correlation, but nominal p >= 0.05",
  nonpositive_p_ge_0.05 =
    "Non-positive correlation and nominal p >= 0.05"
)

model_selection_display <- model_selection |>
  mutate(
    Decision =
      unname(
        decision_labels[
          Decision_code
        ]
      ),
    Predictive_evidence_label =
      unname(
        predictive_evidence_labels[
          Predictive_evidence
        ]
      )
  ) |>
  select(
    Trait,
    Selected_method,
    Predictive_winner,
    DIC_winner,
    r,
    p_value,
    Predictive_loss,
    Delta_DIC,
    Wprob,
    ER,
    Parameters_flagged,
    Runs_with_flags,
    Minimum_ESS,
    Maximum_absolute_Geweke,
    Predictive_evidence_label,
    Decision
  )

kable(
  model_selection_display,
  digits = 5,
  caption = "Final decision by trait integrating prediction, DIC, and convergence"
)
```

### 5.1. Predictive evidence for the final methods

Selecting a final method does not imply that a trait has useful predictive
ability. To separate these two questions, the result is classified only by the
direction of `r` and the p-value of the OOF correlation, without converting
this classification into a new model-fitting criterion.

```{r predictive-evidence-summary}
predictive_evidence_summary <- model_selection |>
  count(
    Predictive_evidence,
    name = "Traits"
  )

predictive_evidence_summary$Evidence <- unname(
  predictive_evidence_labels[
    predictive_evidence_summary$
      Predictive_evidence
  ]
)

kable(
  predictive_evidence_summary |>
    select(
      Evidence,
      Traits
    ),
  caption = "Summary of predictive evidence for the final methods"
)
```

Rankings are calculated for every trait to preserve analytical completeness.
However, `PRODEF`, `FFC`, and `FFP` have positive correlations with
`p >= 0.05`, whereas `EP` has a non-positive correlation. These rankings should
therefore be interpreted cautiously and not with the same strength of evidence
observed for `MMCOM`, `PRODCOM`, `CF`, `DF`, `DCI`, and `SS`.

## 6. Fold stability of the selected methods

As a compact diagnostic of result stability, the validation correlation is
shown separately for the five folds of the method selected for each trait.
This diagnostic does not alter model selection; it shows how the pooled OOF
result is distributed across the predefined family-stratified folds.

```{r selected-fold-diagnostics}
selected_fold_diagnostics <- metrics |>
  inner_join(
    selection_spec |>
      select(
        Trait,
        Selected_method
      ),
    by = c(
      "Trait" = "Trait",
      "Method" = "Selected_method"
    )
  ) |>
  transmute(
    Trait,
    Fold,
    Method,
    Method_label =
      unname(
        method_labels[
          Method
        ]
      ),
    Correlation,
    RMSE,
    MAE
  ) |>
  arrange(
    match(
      Trait,
      traits
    ),
    Fold
  )

selected_fold_diagnostics$Facet <- paste0(
  selected_fold_diagnostics$Trait,
  " — ",
  selected_fold_diagnostics$Method_label
)
```

```{r selected-fold-figure, fig.width=12, fig.height=10}
fold_stability_plot <- ggplot(
  selected_fold_diagnostics,
  aes(
    x = Fold,
    y = Correlation,
    group = Trait
  )
) +
  geom_hline(
    yintercept = 0,
    linetype = 2,
    linewidth = 0.4
  ) +
  geom_line(
    linewidth = 0.6
  ) +
  geom_point(
    size = 2
  ) +
  facet_wrap(
    ~ Facet,
    ncol = 2
  ) +
  scale_x_continuous(
    breaks = 1:5
  ) +
  coord_cartesian(
    ylim = c(-1, 1)
  ) +
  labs(
    x = "Fold",
    y = "Validation correlation",
    title = "Fold-specific predictive correlation of selected methods"
  ) +
  theme_bw(
    base_size = 12
  ) +
  theme(
    strip.text = element_text(
      face = "bold"
    ),
    plot.title = element_text(
      face = "bold"
    )
  )

fold_stability_plot

ggsave(
  filename = file.path(
    FIGURE_DIR,
    "selected_method_fold_stability_en.png"
  ),
  plot = fold_stability_plot,
  width = 12,
  height = 10,
  dpi = 300
)
```

## 7. Computational time

The article measured computation time by repeating chains specifically for
that purpose. No additional repetitions are run here. This adaptation uses the
observed time from the 50 official fits already completed for each method
(\(10\) traits \(\times 5\) folds).

```{r runtime}
runtime_summary_module04 <- metrics |>
  group_by(
    Method
  ) |>
  summarise(
    Fits = n(),
    Total_seconds =
      sum(
        Elapsed_seconds
      ),
    Mean_seconds =
      mean(
        Elapsed_seconds
      ),
    SD_seconds =
      sd(
        Elapsed_seconds
      ),
    Median_seconds =
      median(
        Elapsed_seconds
      ),
    Minimum_seconds =
      min(
        Elapsed_seconds
      ),
    Maximum_seconds =
      max(
        Elapsed_seconds
      ),
    .groups = "drop"
  ) |>
  mutate(
    Method_label =
      unname(
        method_labels[
          Method
        ]
      )
  ) |>
  arrange(
    match(
      Method,
      methods
    )
  )

```

## 8. Heritability and accuracy in the RR-BLUP/BRR analogue

Silva et al. [@silva2021brrguava] estimated additive genetic variance as

\[
\sigma_a^2 =
\sum_j 2p_jq_j
\operatorname{Var}(m_j)
\]

and heritability as

\[
H^2 =
\frac{\sigma_a^2}
{\sigma_a^2+\sigma_e^2}.
\]

In the article, BRR was selected for all traits. In this project, different
methods may win the predictive rule. The preserved native files for BL and
BayesB contain their shrinkage/scale hyperparameters, but not the same direct
marker-variance chain available for BRR. Therefore, to avoid a new fit or a
change in the estimand, the direct reproduction of Table 3 is presented as an
**RR-BLUP/BRR analogue** for all ten traits.

For RR-BLUP/BRR, the common marker variance is stored directly as
`Marker_variance`. Because the marker matrix is global, the
\(\sum 2pq\) denominator is also the same across the five folds.

Accuracy is calculated by fold as in the article:

\[
Accuracy_f =
\frac{r_f}
{\sqrt{H_f^2}}.
\]

```{r rrblup-heritability}
rrblup_variance_components <- convergence |>
  filter(
    Method == "RRBLUP",
    Parameter %in%
      c(
        "Residual_variance",
        "Marker_variance"
      )
  ) |>
  select(
    Run_id,
    Trait,
    Fold,
    Method,
    Parameter,
    Posterior_mean
  ) |>
  pivot_wider(
    names_from = Parameter,
    values_from = Posterior_mean
  )

rrblup_fold_correlations <- metrics |>
  filter(
    Method == "RRBLUP"
  ) |>
  select(
    Trait,
    Fold,
    Correlation
  )

rrblup_heritability_by_fold <-
  rrblup_variance_components |>
  left_join(
    rrblup_fold_correlations,
    by = c(
      "Trait",
      "Fold"
    )
  ) |>
  mutate(
    Marker_variance_denominator =
      matrices_object$denominator,
    Additive_variance =
      Marker_variance *
      Marker_variance_denominator,
    H2 =
      Additive_variance /
      (
        Additive_variance +
          Residual_variance
      ),
    Predictive_accuracy =
      Correlation /
      sqrt(
        H2
      )
  ) |>
  arrange(
    match(
      Trait,
      traits
    ),
    Fold
  )

rrblup_heritability_summary <-
  rrblup_heritability_by_fold |>
  group_by(
    Trait
  ) |>
  summarise(
    Mean_H2 =
      mean(
        H2
      ),
    SD_H2 =
      sd(
        H2
      ),
    Mean_fold_r =
      mean(
        Correlation
      ),
    SD_fold_r =
      sd(
        Correlation
      ),
    Mean_predictive_accuracy =
      mean(
        Predictive_accuracy
      ),
    SD_predictive_accuracy =
      sd(
        Predictive_accuracy
      ),
    .groups = "drop"
  ) |>
  arrange(
    match(
      Trait,
      traits
    )
  )

```

## 9. OOF genomic values from selected methods

For each trait, the selected method is joined to its 150 OOF rows. The value
used is `Genomic_component`, i.e. the genomic component from the fit in which
that individual's phenotype was masked.

```{r selected-oof-genomic-values}
selected_oof_genomic_values <- predictions |>
  inner_join(
    model_selection |>
      select(
        Trait,
        Selected_method,
        Predictive_evidence
      ),
    by = "Trait"
  ) |>
  filter(
    Method ==
      Selected_method
  ) |>
  transmute(
    IND,
    Trait,
    Fold,
    Method,
    Predictive_evidence,
    Genomic_value =
      Genomic_component
  ) |>
  arrange(
    match(
      Trait,
      traits
    ),
    IND
  )

```

## 10. Genomic ranking by trait

Ranking is descending by genomic value. It represents the statistical ranking
rule and does not impose economic weights, trait-specific breeding directions,
or a multi-trait selection index at this stage. The `Predictive_evidence`
column should remain part of the ranking interpretation, especially for traits
without a positive OOF correlation with `p < 0.05`.

```{r genomic-ranking}
genomic_ranking <- selected_oof_genomic_values |>
  group_by(
    Trait
  ) |>
  arrange(
    desc(
      Genomic_value
    ),
    IND,
    .by_group = TRUE
  ) |>
  mutate(
    Rank =
      row_number()
  ) |>
  ungroup() |>
  arrange(
    match(
      Trait,
      traits
    ),
    Rank
  )

```

## 11. Correlations among OOF genomic values

The matrix below is the Pearson correlation among the OOF genomic values
obtained from the selected method for each trait. These are descriptive
correlations among predicted genomic values; they are not multivariate
genetic-correlation estimates.

```{r oof-correlations}
selected_genomic_wide <-
  selected_oof_genomic_values |>
  select(
    IND,
    Trait,
    Genomic_value
  ) |>
  pivot_wider(
    names_from = Trait,
    values_from = Genomic_value
  )

selected_oof_correlations <- cor(
  as.matrix(
    selected_genomic_wide[
      ,
      traits,
      drop = FALSE
    ]
  )
)
```

## 12. Analytical outputs

```{r save-results}
write.csv(
  model_comparison,
  file.path(RESULTS_DIR, "model_comparison.csv"),
  row.names = FALSE
)

write.csv(
  model_selection,
  file.path(RESULTS_DIR, "model_selection.csv"),
  row.names = FALSE
)

write.csv(
  selection_spec,
  file.path(RESULTS_DIR, "selection_decision_spec.csv"),
  row.names = FALSE
)

write.csv(
  convergence_by_trait_method,
  file.path(RESULTS_DIR, "convergence_by_trait_method.csv"),
  row.names = FALSE
)

write.csv(
  selected_fold_diagnostics,
  file.path(RESULTS_DIR, "selected_fold_diagnostics.csv"),
  row.names = FALSE
)

write.csv(
  runtime_summary_module04,
  file.path(RESULTS_DIR, "runtime_summary.csv"),
  row.names = FALSE
)

write.csv(
  rrblup_heritability_by_fold,
  file.path(RESULTS_DIR, "rrblup_heritability_by_fold.csv"),
  row.names = FALSE
)

write.csv(
  rrblup_heritability_summary,
  file.path(RESULTS_DIR, "rrblup_heritability_summary.csv"),
  row.names = FALSE
)

write.csv(
  selected_oof_genomic_values,
  file.path(RESULTS_DIR, "selected_oof_genomic_values.csv"),
  row.names = FALSE
)

write.csv(
  genomic_ranking,
  file.path(RESULTS_DIR, "genomic_ranking.csv"),
  row.names = FALSE
)

write.csv(
  as.data.frame(
    selected_oof_correlations
  ),
  file.path(RESULTS_DIR, "selected_oof_correlations.csv"),
  row.names = TRUE
)

saveRDS(
  list(
    signature =
      models_object$signature,
    model_comparison =
      model_comparison,
    selection_spec =
      selection_spec,
    model_selection =
      model_selection,
    convergence_by_trait_method =
      convergence_by_trait_method,
    selected_fold_diagnostics =
      selected_fold_diagnostics,
    runtime =
      runtime_summary_module04,
    rrblup_heritability_by_fold =
      rrblup_heritability_by_fold,
    rrblup_heritability_summary =
      rrblup_heritability_summary,
    selected_oof_genomic_values =
      selected_oof_genomic_values,
    genomic_ranking =
      genomic_ranking,
    selected_oof_correlations =
      selected_oof_correlations
  ),
  file.path(
    RESULTS_DIR,
    "module04_results.rds"
  )
)
```

## 13. Next step

Module 05 presents the results derived here using the final graphical
standard. It does not refit models or redefine the model-selection
decisions.
