---
title: "Result derivation and model selection"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

Using the predictions, metrics, and diagnostics obtained in Module 03, this
module compares the seven methods within each trait and derives the results
used for genomic selection.

The change from the previous module is therefore from **fitting to decision**: Module 03 produced evidence for each trait, fold, and method; here that evidence is aggregated and organized to determine which method represents each trait and which exploratory rankings can be derived from that choice.

The analysis calculates OOF predictive performance, DIC and relative support
measures, the selected method for each trait, fold stability, computational
time, heritability and accuracy in the RR-BLUP/BRR analogue, OOF genomic
values, oriented rankings, proportional selection, recurrence across traits,
and correlations among genomic values.

The breeding direction is defined before ranking: increase MMCOM, PRODCOM, CF,
DF, EP, FFC, FFP, and SS; decrease PRODEF and DCI.

## 2. Packages and model-fitting results

### 2.1 Packages used

The packages used here have distinct roles:

- `dplyr`: grouping, joining, ordering, and summarizing result tables;
- `tidyr`: reshaping tables between long and wide formats;
- `knitr`: formatting tables in R Markdown documents;
- `here`: building project-relative file paths.

```{r packages, warning=FALSE, message=FALSE}
library(dplyr) # Group, join, and summarize tables.
library(tidyr) # Reshape data between long and wide formats.
library(knitr) # Format tables in R Markdown documents.
library(here)  # Build project-relative paths.
```

```{r table-formatting, include=FALSE}
format_table <- function(x, caption = NULL, digits = getOption("digits"), col.names = NA) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    col.names = col.names,
    table.attr = paste0(
      'class="table table-striped table-hover" ',
      'style="width:100%; margin-bottom:0;"'
    )
  )
  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

### 2.2 Input objects and results directory

The module uses two consolidated objects: `matrices_object.rds`, which supplies phenotypes, metadata, and definitions from Module 02, and `models_object.rds`, which supplies fold-level metrics, OOF predictions, and MCMC diagnostics from Module 03.

```{r inputs}
MATRICES_FILE <- here("output", "matrices", "matrices_object.rds")
MODELS_FILE <- here("output", "models", "models_object.rds")
RESULTS_DIR <- here("output", "results", "m04")

dir.create(RESULTS_DIR, recursive = TRUE, showWarnings = FALSE)

matrices_object <- readRDS(MATRICES_FILE)
models_object <- readRDS(MODELS_FILE)

metrics <- models_object$metrics_by_fold
predictions <- models_object$predictions
convergence <- models_object$convergence_diagnostics

traits <- colnames(matrices_object$phenotypes)
methods <- models_object$settings$methods
ids <- matrices_object$ids

```

## 3. OOF predictive performance

### 3.1 Scope of OOF evaluation

The OOF estimates inherit the family-stratified design of Modules 02 and
03. They quantify prediction within the represented breeding population and do
not estimate generalization to entirely new families.

The main metrics are recalculated using the 150 out-of-fold predictions for
each trait × method combination. Each individual contributes one OOF
prediction for every method.

### 3.2 Pooled performance and variation across folds

For each trait × method combination, correlation, RMSE, MAE, bias, and slope are calculated from the 150 OOF predictions. The two-sided Pearson-correlation p-value is retained only as a nominal descriptive measure because training sets overlap across folds.

```{r pooled-oof-metrics}
# Combine OOF predictions and compute aggregate predictive metrics by trait and method.
pooled_metrics <- predictions |>
  group_by(Trait, Method) |>
  summarise(
    N = n(),
    r = cor(Observed, Predicted),
    RMSE = sqrt(mean((Observed - Predicted)^2)),
    MAE = mean(abs(Observed - Predicted)),
    Bias = mean(Predicted - Observed),
    Slope = cov(Observed, Predicted) / var(Predicted),
    .groups = "drop"
  ) |>
  mutate(
    t_value = r * sqrt((N - 2) / (1 - r^2)),
    p_value = 2 * pt(abs(t_value), df = N - 2, lower.tail = FALSE)
  ) |>
  select(-t_value)

# Summarize the results in `fold_summary`.
fold_summary <- metrics |>
  group_by(Trait, Method) |>
  summarise(
    Mean_DIC = mean(DIC),
    SD_DIC = sd(DIC),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_fold_RMSE = mean(RMSE),
    SD_fold_RMSE = sd(RMSE),
    Mean_fold_MAE = mean(MAE),
    SD_fold_MAE = sd(MAE),
    .groups = "drop"
  )
```

There are two summary scales in this section. `pooled_metrics` combines the 150 OOF predictions and provides the global performance of each trait × method combination; this pooled correlation is the primary reference for ranking predictive winners. `fold_summary` preserves variation across the five folds and is used later to assess stability and construct the descriptive one-standard-error tolerances.

Pearson-correlation p-values are retained as nominal descriptive measures because training sets overlap across folds. They are reported for context and are not used to determine eligibility for genomic selection.

## 4. DIC and relative support for the methods

### 4.1 DIC-derived relative support

Following the logic used in the reference study [@silva2021brrguava], the
difference from the smallest mean DIC within each trait is converted into
relative weights:

\[
\Delta_j=\overline{DIC}_j-\min_k(\overline{DIC}_k),
\]

\[
Wprob_j=\frac{\exp(-\Delta_j/2)}{\sum_k\exp(-\Delta_k/2)},
\qquad
ER_j=\frac{Wprob_{\max}}{Wprob_j}.
\]

`Wprob` is used here as a DIC-derived relative-support weight. It is not
interpreted as a calibrated posterior probability of a model.

The implementation uses log-weights before normalization to improve numerical stability without changing relative ratios. `Wprob` sums to 1 within each trait, and `ER` expresses support relative to the method with the largest weight.

```{r model-comparison}
# Combine fold performance and DIC to build the quantities used for model comparison.
model_comparison <- pooled_metrics |>
  left_join(fold_summary, by = c("Trait", "Method")) |>
  group_by(Trait) |>
  mutate(
    Delta_DIC = Mean_DIC - min(Mean_DIC),
    log_weight = -0.5 * Delta_DIC,
    stable_weight = exp(log_weight - max(log_weight)),
    Wprob = stable_weight / sum(stable_weight),
    # Compute the evidence ratio relative to the method with the largest Wprob in each trait.
    ER = max(Wprob) / Wprob
  ) |>
  ungroup() |>
  select(
    Trait, Method, Mean_DIC, SD_DIC, Delta_DIC, Wprob, ER,
    N, r, p_value, RMSE, MAE, Bias, Slope,
    Mean_fold_r, SD_fold_r, Mean_fold_RMSE, SD_fold_RMSE,
    Mean_fold_MAE, SD_fold_MAE
  ) |>
  arrange(match(Trait, traits), Delta_DIC)
```

### 4.2 Relationship between Bayesian fit and predictive performance

`model_comparison` combines OOF performance and DIC-derived relative support for every trait and method. DIC quantities are comparable only within the same trait and do not replace predictive evidence.

In the joint comparison, methods with similar DIC can have different OOF performance, and the method with the largest `Wprob` need not have the largest predictive correlation. Bayesian fit and predictive ability therefore remain complementary rather than interchangeable criteria.

## 5. Selecting a method for each trait

Selection is trait-specific and hierarchical: predictive evidence, DIC support, and the monitored MCMC diagnostics remain separate rather than being collapsed into a single score.

The selection hierarchy is applied identically to every trait and can be read as one five-step algorithm:

1. identify the OOF predictive winner and the DIC winner separately;
2. define the predictively competitive set by simultaneously requiring the descriptive one-standard-error tolerances for `Mean_fold_r` and `Mean_fold_RMSE`;
3. classify a method as an automatic candidate only if it also satisfies `Delta_DIC <= 2` and has no flags in the monitored MCMC diagnostics;
4. among automatic candidates, prioritize RR-BLUP; if RR-BLUP is not in that set, order the remaining candidates by pooled OOF correlation, RMSE, MAE, and DIC;
5. if no automatic candidate exists, retain the predictive winner as a provisional choice requiring review; then attach the selected method's metrics and calculate `Predictive_loss` only as a descriptive diagnostic.

The blocks below implement these five steps while keeping predictive evidence, DIC fit evidence, and MCMC diagnostics conceptually separate.

### 5.1 Predictive and DIC winners

First, the method with the best OOF predictive performance and the method with the lowest mean DIC are identified separately for each trait.

The `Delta_DIC <= 2` cutoff is the operational rule used to regard fits as having similar DIC; it is not a formal test. The number of folds is not redefined in this module: `N_FOLDS` is recovered from the Module 02 object and is used only to calculate the descriptive one-standard-error tolerances.

```{r selection-winners}
# Use Delta_DIC <= 2 as the operational cutoff for similar fit.
DIC_COMPETITIVE_THRESHOLD <- 2
# Recover the five Module 02 folds for the descriptive one-standard-error tolerances.
N_FOLDS <- matrices_object$settings$n_folds

# Best combination of correlation, RMSE, MAE, and DIC for each trait.
predictive_winners <- model_comparison |>
  group_by(Trait) |>
  arrange(desc(r), RMSE, MAE, Mean_DIC, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    Predictive_winner = Method,
    Predictive_winner_r = r,
    Predictive_winner_p_value = p_value,
    Predictive_winner_RMSE = RMSE,
    Predictive_winner_MAE = MAE
  )

# Method with the smallest mean DIC for each trait.
dic_winners <- model_comparison |>
  group_by(Trait) |>
  arrange(Mean_DIC, desc(r), RMSE, MAE, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    DIC_winner = Method,
    Minimum_DIC = Mean_DIC,
    DIC_winner_Wprob = Wprob
  )
```

Predictive and DIC winners are retained separately. Disagreement between them is not treated as an error because predictive performance and fit–complexity address different criteria.

### 5.2 MCMC diagnostics by trait and method

Chain diagnostics are summarized before defining the set of methods eligible
for selection.

```{r selection-convergence-summary}
convergence_by_trait_method <- convergence |>
  group_by(Trait, Method) |>
  summarise(
    Runs = n_distinct(Run_id),
    Parameters_reviewed = n(),
    Parameters_flagged = sum(Review_flag),
    Runs_with_flags = n_distinct(Run_id[Review_flag]),
    Minimum_ESS = min(Effective_sample_size, na.rm = TRUE),
    Maximum_absolute_Geweke = max(abs(Geweke_Z), na.rm = TRUE),
    .groups = "drop"
  )
```

`convergence_by_trait_method` summarizes the two scalar diagnostics monitored per fit in Module 03. `Parameters_flagged = 0` means only that no flags occurred in those quantities; it is not proof that every marker effect has converged.

### 5.3 Competitive set

Steps 2 and 3 are implemented here. Predictive competitiveness simultaneously requires the one-standard-error tolerances for correlation and RMSE: the former uses the across-fold standard deviation of the method with the largest `Mean_fold_r`, and the latter uses the method with the smallest `Mean_fold_RMSE`, both divided by `sqrt(5)`. `Delta_DIC <= 2` and the absence of flags in the monitored diagnostics then define an automatic candidate. Because training sets overlap across folds, these tolerances are descriptive rather than independent-sample inferential standard errors.

```{r selection-competitive-set}
# Combine performance, DIC, and MCMC diagnostics to define the competitive set for each trait.
model_selection_support <- model_comparison |>
  left_join(
    convergence_by_trait_method,
    by = c("Trait", "Method")
  ) |>
  left_join(
    predictive_winners |>
      select(Trait, Predictive_winner),
    by = "Trait"
  ) |>
  group_by(Trait) |>
  mutate(
    # References for the one-standard-error rule.
    Best_mean_fold_r = max(Mean_fold_r, na.rm = TRUE),
    Best_mean_fold_r_SE =
      SD_fold_r[which.max(Mean_fold_r)] / sqrt(N_FOLDS),
    Best_mean_fold_RMSE = min(Mean_fold_RMSE, na.rm = TRUE),
    Best_mean_fold_RMSE_SE =
      SD_fold_RMSE[which.min(Mean_fold_RMSE)] / sqrt(N_FOLDS),

    # Predictive competitiveness.
    Within_one_SE_r =
      Mean_fold_r >= Best_mean_fold_r - Best_mean_fold_r_SE,
    Within_one_SE_RMSE =
      Mean_fold_RMSE <= Best_mean_fold_RMSE + Best_mean_fold_RMSE_SE,
    Predictively_competitive =
      Within_one_SE_r & Within_one_SE_RMSE,

    # Complementary support from DIC and convergence.
    DIC_competitive_threshold = DIC_COMPETITIVE_THRESHOLD,
    DIC_competitive = Delta_DIC <= DIC_COMPETITIVE_THRESHOLD,
    Competitive_for_selection =
      Predictively_competitive & DIC_competitive,
    Convergence_review = Parameters_flagged > 0,
    Automatic_candidate =
      Competitive_for_selection & !Convergence_review,

    # Candidate-set summary within each trait.
    Competitive_set_n = sum(Competitive_for_selection),
    Automatic_candidate_n = sum(Automatic_candidate),
    RRBLUP_competitive = any(
      Method == "RRBLUP" & Competitive_for_selection
    ),
    RRBLUP_automatic_candidate = any(
      Method == "RRBLUP" & Automatic_candidate
    ),
    Automatic_selection_available = any(Automatic_candidate),

    # RR-BLUP receives priority only within the automatic-candidate set.
    Selection_priority = case_when(
      Automatic_candidate & Method == "RRBLUP" ~ 1L,
      Automatic_candidate ~ 2L,
      TRUE ~ 3L
    )
  ) |>
  ungroup() |>
  arrange(
    match(Trait, traits),
    Selection_priority,
    desc(r),
    RMSE,
    MAE,
    Mean_DIC
  )
```

`Automatic_candidate = TRUE` simultaneously requires predictive competitiveness, `Delta_DIC <= 2`, and no flags in the monitored diagnostics.

### 5.4 Method choice

Steps 4 and 5 are implemented here. RR-BLUP receives priority only within the automatic-candidate set as an explicit parsimony choice and homogeneous-shrinkage benchmark, without implying universal statistical superiority. If RR-BLUP is absent from that set, the remaining candidates follow the order `r`, RMSE, MAE, and DIC; if no automatic candidate exists, the predictive winner is retained provisionally for review.


```{r selection-choice}
# Best automatic candidate for each trait.
automatic_selections <- model_selection_support |>
  filter(Automatic_candidate) |>
  group_by(Trait) |>
  arrange(
    Selection_priority,
    desc(r),
    RMSE,
    MAE,
    Mean_DIC,
    .by_group = TRUE
  ) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    Automatic_selected_method = Method
  )

# Final choice before attaching the selected method's metrics.
# Apply RR-BLUP priority within the automatic set and define the selected method.
selection_spec <- predictive_winners |>
  select(Trait, Predictive_winner) |>
  left_join(automatic_selections, by = "Trait") |>
  mutate(
    Selected_method = coalesce(
      Automatic_selected_method,
      Predictive_winner
    )
  ) |>
  select(Trait, Selected_method, Predictive_winner) |>
  left_join(
    model_selection_support |>
      select(
        Trait,
        Method,
        Automatic_selection_available,
        Automatic_candidate_n,
        Competitive_set_n,
        RRBLUP_competitive,
        RRBLUP_automatic_candidate,
        Convergence_review
      ),
    # Match the selected method to the support table to append the indicators used in the decision.
    by = c("Trait", "Selected_method" = "Method")
  ) |>
  mutate(
    RRBLUP_preference_applied =
      Selected_method == "RRBLUP" &
      Predictive_winner != "RRBLUP" &
      RRBLUP_automatic_candidate,
    Selection_requires_review =
      !Automatic_selection_available | Convergence_review,
    Selection_status = case_when(
      Selection_requires_review ~ "provisional_review_required",
      TRUE ~ "rule_based_selected"
    ),
    Decision_code = case_when(
      Selection_requires_review ~
        "predictive_winner_provisional_review_required",
      Selected_method == "RRBLUP" &
        RRBLUP_preference_applied ~
        "rrblup_preferred_among_competitive_models",
      Selected_method == "RRBLUP" &
        Predictive_winner == "RRBLUP" ~
        "rrblup_best_predictive_model",
      TRUE ~ "best_non_rrblup_competitive_model"
    ),
    Operational_competitive_set_rule =
      "one_se_r_and_rmse_plus_delta_dic_le_2"
  ) |>
  select(-Predictive_winner, -Convergence_review) |>
  arrange(match(Trait, traits))

```

`selection_spec` records the result of this hierarchy. `rule_based_selected` means that the rule could be applied, whereas `provisional_review_required` identifies a choice retained only for review.

### 5.5 Performance of the selected methods

Metrics and diagnostics for the method actually selected are then extracted
from the support table.

```{r selected-performance}
# Extract predictive metrics, DIC, and diagnostics for the selected method in each trait.
selected_performance <- model_selection_support |>
  inner_join(
    selection_spec |>
      select(Trait, Selected_method),
    by = "Trait"
  ) |>
  # Restrict the data to observations relevant at this stage.
  filter(Method == Selected_method) |>
  # Build the columns that form the output of this block.
  transmute(
    Trait,
    Selected_method,
    r,
    p_value,
    RMSE,
    MAE,
    Bias,
    Slope,
    Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    Mean_fold_r,
    SD_fold_r,
    Mean_fold_RMSE,
    SD_fold_RMSE,
    Within_one_SE_r,
    Within_one_SE_RMSE,
    Predictively_competitive,
    DIC_competitive,
    Competitive_for_selection,
    Automatic_candidate,
    Competitive_set_n,
    Automatic_candidate_n,
    RRBLUP_competitive,
    RRBLUP_automatic_candidate,
    Parameters_flagged,
    Runs_with_flags,
    Minimum_ESS,
    Maximum_absolute_Geweke,
    Convergence_review
  )
```

`selected_performance` retains, for every trait, the OOF metrics, DIC, and diagnostics corresponding to the method actually selected.

### 5.6 Final selection result and predictive evidence

Finally, the selected method, predictive and DIC winners, OOF direction, and a
nominal signal class are brought together for each trait. The nominal class is shown only to make weak positive correlations explicit; it is not used as a selection filter. The $r > 0$ boundary separates only positive from nonpositive predictive direction; it is neither a significance test nor a minimum predictive-utility threshold.

`Predictive_loss` records the difference between the pooled OOF correlation of the predictive winner and the pooled OOF correlation of the method actually selected:

\[
\mathrm{Predictive\_loss}
=
r_{\mathrm{predictive\ winner}}
-
r_{\mathrm{selected\ method}}.
\]

Thus, `Predictive_loss = 0` means that the selected method is also the predictive winner, whereas a positive value quantifies how much pooled OOF correlation was given up by the hierarchical selection rule. This quantity is not a loss function used to fit the models, does not combine RMSE or DIC, and is not an additional selection criterion; it is only a descriptive diagnostic of the predictive-performance difference.

```{r final-model-selection}
# Combine the selected method, predictive and DIC winners, and OOF evidence in the final trait-level table.
model_selection <- selection_spec |>
  left_join(
    selected_performance,
    # Join selected-method performance using trait and method as keys.
    by = c("Trait", "Selected_method")
  ) |>
  left_join(predictive_winners, by = "Trait") |>
  left_join(dic_winners, by = "Trait") |>
  mutate(
    # Quantify the OOF-correlation difference between the predictive winner and selected method.
    Predictive_loss = Predictive_winner_r - r,
    Predictive_winner_agrees =
      Selected_method == Predictive_winner,
    DIC_winner_agrees =
      Selected_method == DIC_winner,
    # Use zero only as a direction boundary; r > 0 does not imply significance or minimum utility.
    OOF_predictive_direction = if_else(
      r > 0,
      "positive",
      "nonpositive"
    ),
    # Keep p < 0.05 only as a descriptive nominal class, without filtering rankings.
    OOF_nominal_signal = case_when(
      r > 0 & p_value < 0.05 ~ "positive_nominal_p_lt_0.05",
      r > 0 ~ "positive_nominal_p_ge_0.05",
      TRUE ~ "nonpositive"
    )
  ) |>
  arrange(match(Trait, traits))


# Count how many traits have positive or nonpositive OOF correlation.
oof_direction_summary <- model_selection |>
  count(OOF_predictive_direction, name = "Traits") |>
  arrange(OOF_predictive_direction)
```

This is the main decision table by trait. In addition to the selected method, it preserves the predictive and DIC winners, displays `Predictive_loss` as a descriptive diagnostic, records whether the RR-BLUP preference was applied, and makes the direction and nominal strength of the OOF signal explicit.

```{r model-selection-table}
# DISPLAY-ONLY START
model_selection_display <- model_selection |>
  transmute(
    Trait,
    Selected_method,
    Predictive_winner,
    DIC_winner,
    r = sprintf("%.3f", r),
    p_value = if_else(
      p_value < 0.001,
      "<0.001",
      sprintf("%.3f", p_value)
    ),
    RMSE = sprintf("%.3f", RMSE),
    Mean_DIC = sprintf("%.2f", Mean_DIC),
    Delta_DIC = sprintf("%.2f", Delta_DIC),
    Predictive_loss = sprintf("%.3f", Predictive_loss),
    RRBLUP_preference_applied,
    Selection_status,
    OOF_predictive_direction,
    OOF_nominal_signal
  )

format_table(
  model_selection_display,
  caption = "Method selected for each trait",
  col.names = c("Trait", "Selected method", "Predictive winner", "DIC winner", "r", "p-value", "RMSE", "Mean DIC", "Delta DIC", "Predictive loss", "RR-BLUP preference applied", "Selection status", "OOF predictive direction", "OOF nominal signal")
)
# DISPLAY-ONLY END
```

The direction summary counts how many traits end with positive or nonpositive OOF correlation for the selected method. It does not classify magnitude, practical utility, or significance.

```{r oof-direction-table}
format_table(
  oof_direction_summary,
  caption = "OOF predictive direction of the traits",
  col.names = c("OOF predictive direction", "Traits")
)
```

## 6. Stability across folds for the selected methods

Fold-level metrics for the selected methods are derived here for the Module 05 synthesis. Sign changes and dispersion across folds are descriptive diagnostics; with only five overlapping folds, they do not constitute a formal heterogeneity test.

```{r selected-fold-diagnostics}
selected_fold_diagnostics <- metrics |>
  inner_join(
    selection_spec |> select(Trait, Selected_method),
    by = c("Trait" = "Trait", "Method" = "Selected_method")
  ) |>
  transmute(Trait, Fold, Method, Correlation, RMSE, MAE) |>
  arrange(match(Trait, traits), Fold)
```


## 7. Computational time

Runtime is summarized only for operational documentation and presentation in Module 05. It is not used for model selection and should not be interpreted as a universal benchmark.

```{r runtime}
# Summarize the computational cost of the evaluated methods.
runtime_summary <- metrics |>
  group_by(Method) |>
  summarise(
    Fits = n(),
    Total_seconds = sum(Elapsed_seconds),
    Mean_seconds = mean(Elapsed_seconds),
    SD_seconds = sd(Elapsed_seconds),
    Median_seconds = median(Elapsed_seconds),
    Minimum_seconds = min(Elapsed_seconds),
    Maximum_seconds = max(Elapsed_seconds),
    .groups = "drop"
  ) |>
  arrange(match(Method, methods))
```

## 8. Marker-based heritability analogue and adjusted accuracy in RR-BLUP/BRR

### 8.1 Definition and interpretation limits

Following Silva et al. [@silva2021brrguava], a marker-based additive-variance
analogue is obtained from the common marker variance and the \(\sum 2pq\)
denominator. Fold-level adjusted accuracy is defined as
\(r_f/\sqrt{H_f^2}\). The resulting \(H^2\) is reported as a model-based
heritability analogue rather than as an independent classical estimate of
narrow-sense heritability. The quantity named `Predictive_accuracy` is a derived ratio, `r_f/sqrt(H_f^2)`, rather than an observed correlation between true and predicted genetic values. Because it depends on sample estimates of `r_f` and `H_f^2`, it may be negative or exceed 1; it should therefore be interpreted as an adjusted-accuracy index under this definition, not as a measure necessarily bounded to [0, 1].

### 8.2 Fold-level derivation and across-fold summary

```{r rrblup-heritability}
# Compute and organize the RR-BLUP/BRR model-based heritability summary.
rrblup_variance_components <- convergence |>
  filter(
    Method == "RRBLUP",
    Parameter %in% c("Residual_variance", "Marker_variance")
  ) |>
  select(Run_id, Trait, Fold, Method, Parameter, Posterior_mean) |>
  pivot_wider(names_from = Parameter, values_from = Posterior_mean)

rrblup_fold_correlations <- metrics |>
  filter(Method == "RRBLUP") |>
  select(Trait, Fold, Correlation)

# Organize the cross-validation and fitting structure.
rrblup_heritability_by_fold <- rrblup_variance_components |>
  left_join(rrblup_fold_correlations, by = c("Trait", "Fold")) |>
  mutate(
    Marker_variance_denominator = matrices_object$denominator,
    Additive_variance = Marker_variance * Marker_variance_denominator,
    H2 = Additive_variance / (Additive_variance + Residual_variance),
    Predictive_accuracy = Correlation / sqrt(H2)
  ) |>
  arrange(match(Trait, traits), Fold)

# Summarize the RR-BLUP/BRR model-based heritability analogue and adjusted accuracy across folds.
rrblup_heritability_summary <- rrblup_heritability_by_fold |>
  group_by(Trait) |>
  summarise(
    Mean_H2 = mean(H2),
    SD_H2 = sd(H2),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_predictive_accuracy = mean(Predictive_accuracy),
    SD_predictive_accuracy = sd(Predictive_accuracy),
    .groups = "drop"
  ) |>
  arrange(match(Trait, traits))
```

`rrblup_heritability_summary` retains, by trait, the model-derived RR-BLUP/BRR heritability analogue and the ratio `Correlation / sqrt(H2)`. This ratio is not a conventional correlation and may lie outside [0, 1].

## 9. OOF genomic values and selection direction

### 9.1 Cross-fitted genomic values

Each individual's genomic value is cross-fitted: it comes from the fold in
which that individual's phenotype was masked. These OOF genomic values provide
a validation-consistent ranking because an individual's own phenotype does not
enter its score. They are not final breeding values from a model refitted to
all 150 phenotypes.

### 9.2 Selection direction and score construction

Selection direction is applied to the genomic value only when the ranking is constructed. The original genomic value remains unchanged. For each trait, `Genomic_value` is exactly the OOF `Genomic_component` produced by the selected method; different traits may therefore use genomic values supplied by different methods.

```{r selected-oof-genomic-values}
# Define the desired selection direction and link OOF genomic values to the selected method for each trait.
selection_objective_map <- c(
  MMCOM = "increase",
  PRODCOM = "increase",
  PRODEF = "decrease",
  CF = "increase",
  DF = "increase",
  DCI = "decrease",
  EP = "increase",
  FFC = "increase",
  FFP = "increase",
  SS = "increase"
)

# Organize the desired selection direction and ranking rule for each trait.
selection_objectives <- data.frame(
  Trait = traits,
  Selection_objective = unname(selection_objective_map[traits]),
  stringsAsFactors = FALSE
) |>
  mutate(
    Ranking_rule = case_when(
      Selection_objective == "increase" ~ "maximize_genomic_value",
      Selection_objective == "decrease" ~ "minimize_genomic_value",
      TRUE ~ NA_character_
    )
  )

# Organize the estimated genomic component for each individual.
selected_oof_genomic_values <- predictions |>
  inner_join(
    model_selection |>
      select(
        Trait, Selected_method, OOF_predictive_direction,
        OOF_nominal_signal, r, p_value
      ),
    by = "Trait"
  ) |>
  inner_join(selection_objectives, by = "Trait") |>
  filter(Method == Selected_method) |>
  transmute(
    IND, Trait, Fold, Method, OOF_predictive_direction,
    OOF_nominal_signal, OOF_r = r, OOF_nominal_p_value = p_value,
    Selection_objective, Ranking_rule,
    Genomic_value = Genomic_component
  ) |>
  arrange(match(Trait, traits), IND)

format_table(
  selection_objectives,
  caption = "Selection direction defined for each trait",
  col.names = c("Trait", "Selection objective", "Ranking rule")
)
```

## 10. Exploratory proportional candidate ranking by trait

### 10.1 Eligibility and prioritized proportion

Candidate rankings are shown for traits whose selected method has positive OOF
correlation. Nominal p-values remain descriptive and do not determine whether a
ranking is produced. Positive correlations with nominal `p >= 0.05` are kept
explicitly as weak/exploratory signals and should not be interpreted as
demonstrated predictive utility. Within each positive-OOF trait, the 20% of
individuals with the largest oriented score are listed; this 20% is a
configurable rule in this analysis, not an estimated optimal selection intensity. With 150 available candidates, 20% corresponds to 30 individuals. For traits whose objective is to decrease, the score is the negative genomic value.

### 10.2 Ordering and boundary ties

`IND` is used only as a deterministic secondary ordering key when two scores are exactly tied. The selected proportion remains fixed even if a tie occurs at the boundary; `Exact_boundary_tie = TRUE` only reports that the last included and first excluded candidates have the same score, so there is no genomic-score evidence separating them at that cutoff.


```{r genomic-ranking}
# Orient genomic values by the selection objective and rank individuals within each trait.
genomic_ranking <- selected_oof_genomic_values |>
  mutate(
    Selection_score = case_when(
      Selection_objective == "increase" ~ Genomic_value,
      Selection_objective == "decrease" ~ -Genomic_value,
      TRUE ~ NA_real_
    )
  ) |>
  group_by(Trait) |>
  arrange(desc(Selection_score), IND, .by_group = TRUE) |>
  mutate(Rank = row_number()) |>
  ungroup() |>
  arrange(match(Trait, traits), Rank)

# Define the candidate proportion displayed in each exploratory ranking.
SELECTION_PROPORTION <- 0.20

positive_oof_traits <- model_selection |>
  filter(OOF_predictive_direction == "positive") |>
  pull(Trait)

# Compute how many individuals compose the prioritized proportion in each positive-OOF trait.
selection_sizes <- genomic_ranking |>
  filter(Trait %in% positive_oof_traits) |>
  count(Trait, name = "N_available") |>
  mutate(
    Selection_proportion = SELECTION_PROPORTION,
    N_selected = pmax(
      1L,
      as.integer(ceiling(SELECTION_PROPORTION * N_available))
    )
  ) |>
  arrange(match(Trait, traits))

# Record the cutoff score and genomic-value interval corresponding to each ranking.
selection_cutoff_summary <- genomic_ranking |>
  filter(Trait %in% positive_oof_traits) |>
  inner_join(
    selection_sizes |>
      select(Trait, N_available, N_selected),
    by = "Trait"
  ) |>
  arrange(match(Trait, traits), Rank) |>
  group_by(Trait) |>
  mutate(Next_score = lead(Selection_score)) |>
  summarise(
    N_available = first(N_available),
    N_selected = first(N_selected),
    Score_last_selected = Selection_score[first(N_selected)],
    Score_first_not_selected = Next_score[first(N_selected)],
    Exact_boundary_tie =
      !is.na(Score_first_not_selected) &
      Score_last_selected == Score_first_not_selected,
    .groups = "drop"
  ) |>
  arrange(match(Trait, traits))

# Retain candidates included in the defined ranking proportion for each positive-OOF trait.
selected_individuals_by_trait <- genomic_ranking |>
  inner_join(selection_sizes, by = "Trait") |>
  filter(Rank <= N_selected) |>
  left_join(
    matrices_object$metadata |>
      select(IND, FAM),
    by = "IND"
  ) |>
  select(
    Trait, Rank, IND, FAM, Method,
    Selection_objective, Ranking_rule,
    Genomic_value, Selection_score, OOF_predictive_direction,
    OOF_nominal_signal, OOF_r, OOF_nominal_p_value,
    Selection_proportion, N_available, N_selected
  ) |>
  arrange(match(Trait, traits), Rank)

```

### 10.3 Cutoff diagnostic

The cutoff table shows the score of the last included candidate and the first candidate not included. `Exact_boundary_tie` should be checked before interpreting the cutoff as a substantive separation between candidates.

```{r selection-cutoff-table}
# DISPLAY-ONLY START
selection_cutoff_display <- selection_cutoff_summary |>
  transmute(
    Trait,
    N_available = as.character(N_available),
    N_selected = as.character(N_selected),
    Score_last_selected = sprintf("%.3f", Score_last_selected),
    Score_first_not_selected = if_else(
      is.na(Score_first_not_selected),
      NA_character_,
      sprintf("%.3f", Score_first_not_selected)
    ),
    Exact_boundary_tie
  )

format_table(
  selection_cutoff_display,
  caption = "Scores at the proportional-selection boundary",
  col.names = c("Trait", "Available candidates", "Selected candidates", "Last selected score", "First non-selected score", "Exact boundary tie")
)
# DISPLAY-ONLY END
```

The complete `selected_individuals_by_trait` object retains every candidate in the prioritized 20% for each eligible trait; named candidate presentation is deferred to Module 05.

## 11. Recurrence across traits

### 11.1 Recurrence definition and denominator

Recurrence is descriptive: among traits with positive OOF correlation, it counts how many times each individual appears among the most favorable 20%. `Selection_frequency` divides this count by the number of positive-OOF traits, not necessarily by all ten original traits. Weak positive OOF signals remain exploratory. No weights are used and no multi-trait selection index is constructed; appearing in several lists does not demonstrate joint multi-trait superiority.

### 11.2 Recurrence classes and recurrent candidates

Recurrence classes are descriptive conventions: high for four or more traits, intermediate for three, broad for two, specific for one, and none for zero. The `multitrait_candidates` object retains individuals recurring in at least two traits so potentially useful information is not discarded before downstream presentation. These cutoffs organize the summary and are not statistical-significance thresholds.


```{r multitrait-recurrence}
# Count individual recurrence across exploratory rankings for traits with positive OOF correlation.
recurrence_counts <- selected_individuals_by_trait |>
  arrange(match(Trait, traits), Rank) |>
  group_by(IND, FAM) |>
  summarise(
    N_traits = n(),
    Traits = paste(Trait, collapse = ", "),
    Best_rank = min(Rank),
    Mean_rank_present = mean(Rank),
    .groups = "drop"
  )

# Join recurrence counts to metadata and compute recurrence frequency and class.
multitrait_recurrence <- matrices_object$metadata |>
  select(IND, FAM) |>
  left_join(recurrence_counts, by = c("IND", "FAM")) |>
  mutate(
    N_traits = tidyr::replace_na(N_traits, 0L),
    Traits = tidyr::replace_na(Traits, ""),
    Best_rank = case_when(
      N_traits == 0L ~ NA_integer_,
      TRUE ~ Best_rank
    ),
    Mean_rank_present = case_when(
      N_traits == 0L ~ NA_real_,
      TRUE ~ Mean_rank_present
    ),
    Selection_frequency = N_traits / length(positive_oof_traits),
    # Classify recurrence by number of traits; these cutoffs are descriptive.
    Recurrence_class = case_when(
      N_traits >= 4L ~ "high_4plus",
      N_traits == 3L ~ "intermediate_3",
      N_traits == 2L ~ "broad_2",
      N_traits == 1L ~ "specific_1",
      TRUE ~ "none_0"
    )
  ) |>
  arrange(desc(N_traits), FAM, IND)

# Summarize how many individuals belong to each recurrence class.
recurrence_distribution <- multitrait_recurrence |>
  count(N_traits, Recurrence_class, name = "N_individuals") |>
  mutate(
    Proportion_of_population =
      N_individuals / nrow(matrices_object$metadata)
  ) |>
  arrange(desc(N_traits))

# Reshape trait-specific ranks to one row per recurrent candidate.
candidate_rank_wide <- selected_individuals_by_trait |>
  semi_join(
    multitrait_recurrence |>
      filter(N_traits >= 2L),
    by = c("IND", "FAM")
  ) |>
  select(IND, FAM, Trait, Rank) |>
  pivot_wider(
    names_from = Trait,
    values_from = Rank,
    names_glue = "{Trait}_rank"
  )

# Retain candidates recurring in at least two traits and attach their trait-specific ranks.
multitrait_candidates <- multitrait_recurrence |>
  # Retain in the complete object candidates present in at least two traits.
  filter(N_traits >= 2L) |>
  left_join(candidate_rank_wide, by = c("IND", "FAM")) |>
  arrange(desc(N_traits), FAM, IND)
```

`multitrait_recurrence`, `recurrence_distribution`, and `multitrait_candidates` are derived here and presented in Module 05. The complete scientific object retains candidates recurring in at least two traits; recurrence is not a multi-trait selection index.

## 12. Correlations among OOF genomic values

These correlations describe associations among the OOF genomic values produced by the selected method for each trait. Because different traits may use different methods and the values are cross-fitted, the matrix summarizes associations among predictive scores rather than jointly estimated genetic covariance. It should therefore not be interpreted as a genetic-correlation matrix from a multivariate model.


```{r oof-correlations}
selected_genomic_wide <- selected_oof_genomic_values |>
  select(IND, Trait, Genomic_value) |>
  pivot_wider(names_from = Trait, values_from = Genomic_value)

selected_oof_correlations <- cor(
  as.matrix(selected_genomic_wide[, traits, drop = FALSE])
)
```

`selected_oof_correlations` is derived here for visualization in Module 05. It should be interpreted only as association among OOF scores, not as joint genetic correlation.

## 13. Saving the results

The derived tables and `module04_results.rds` object are persisted for direct use in Module 05. No genomic model is refitted at this stage.

```{r save-results}
# Save the tables derived by the module before assembling the consolidated RDS object.
write.csv(model_comparison, file.path(RESULTS_DIR, "model_comparison.csv"), row.names = FALSE)
write.csv(model_selection_support, file.path(RESULTS_DIR, "model_selection_support.csv"), row.names = FALSE)
write.csv(model_selection, file.path(RESULTS_DIR, "model_selection.csv"), row.names = FALSE)
write.csv(selection_spec, file.path(RESULTS_DIR, "selection_decision_spec.csv"), row.names = FALSE)
write.csv(oof_direction_summary, file.path(RESULTS_DIR, "oof_direction_summary.csv"), row.names = FALSE)
write.csv(convergence_by_trait_method, file.path(RESULTS_DIR, "convergence_by_trait_method.csv"), row.names = FALSE)
write.csv(selected_fold_diagnostics, file.path(RESULTS_DIR, "selected_fold_diagnostics.csv"), row.names = FALSE)
write.csv(runtime_summary, file.path(RESULTS_DIR, "runtime_summary.csv"), row.names = FALSE)
write.csv(rrblup_heritability_by_fold, file.path(RESULTS_DIR, "rrblup_heritability_by_fold.csv"), row.names = FALSE)
write.csv(rrblup_heritability_summary, file.path(RESULTS_DIR, "rrblup_heritability_summary.csv"), row.names = FALSE)
write.csv(selection_objectives, file.path(RESULTS_DIR, "selection_objectives.csv"), row.names = FALSE)
write.csv(selected_oof_genomic_values, file.path(RESULTS_DIR, "selected_oof_genomic_values.csv"), row.names = FALSE)
# Save the consolidated output for downstream use.
write.csv(genomic_ranking, file.path(RESULTS_DIR, "genomic_ranking.csv"), row.names = FALSE)
write.csv(selection_sizes, file.path(RESULTS_DIR, "selection_sizes.csv"), row.names = FALSE)
write.csv(selection_cutoff_summary, file.path(RESULTS_DIR, "selection_cutoff_summary.csv"), row.names = FALSE)
write.csv(selected_individuals_by_trait, file.path(RESULTS_DIR, "selected_individuals_by_trait.csv"), row.names = FALSE)
write.csv(multitrait_recurrence, file.path(RESULTS_DIR, "multitrait_recurrence.csv"), row.names = FALSE)
write.csv(recurrence_distribution, file.path(RESULTS_DIR, "recurrence_distribution.csv"), row.names = FALSE)
write.csv(multitrait_candidates, file.path(RESULTS_DIR, "multitrait_candidates.csv"), row.names = FALSE)
# Save the consolidated output for downstream use.
write.csv(as.data.frame(selected_oof_correlations), file.path(RESULTS_DIR, "selected_oof_correlations.csv"), row.names = TRUE)

# Save the consolidated output for downstream use.
saveRDS(
  list(
    traits = traits,
    methods = methods,
    model_comparison = model_comparison,
    model_selection_support = model_selection_support,
    selection_spec = selection_spec,
    model_selection = model_selection,
    oof_direction_summary = oof_direction_summary,
    convergence_by_trait_method = convergence_by_trait_method,
    selected_fold_diagnostics = selected_fold_diagnostics,
    runtime = runtime_summary,
    rrblup_heritability_by_fold = rrblup_heritability_by_fold,
    rrblup_heritability_summary = rrblup_heritability_summary,
    selection_objectives = selection_objectives,
    selected_oof_genomic_values = selected_oof_genomic_values,
    genomic_ranking = genomic_ranking,
    selection_proportion = SELECTION_PROPORTION,
    positive_oof_traits = positive_oof_traits,
    selection_sizes = selection_sizes,
    selection_cutoff_summary = selection_cutoff_summary,
    selected_individuals_by_trait = selected_individuals_by_trait,
    multitrait_recurrence = multitrait_recurrence,
    recurrence_distribution = recurrence_distribution,
    multitrait_candidates = multitrait_candidates,
    selected_oof_correlations = selected_oof_correlations
  ),
  file.path(RESULTS_DIR, "module04_results.rds")
)
```

## 14. Next step

By the end of this module, the derivation stage is complete: method comparison, trait-specific decisions, associated diagnostics, exploratory rankings, recurrence, and correlations among OOF scores have already been defined in the persisted objects.

Module 05 changes from **derivation to presentation and interpretation**. It uses only these derivations to build summary tables and figures, without reapplying the selection rule, rebuilding rankings, or refitting models. OOF genomic values remain cross-fitted validation scores; this workflow does not transform them into final breeding values from a refit using all 150 phenotypes.

<div style="display:flex; justify-content:space-between; gap:1rem; margin-top:2rem; padding-top:1rem; border-top:1px solid #ddd;">
<a href="03_models_en.html">← Module 03 — Model fitting</a>
<a href="05_results_en.html">Module 05 — Results synthesis →</a>
</div>
