---
title: "Result derivation and model selection"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

Using the predictions, metrics, and diagnostics obtained in Module 03, this
module compares the seven methods within each trait and derives the results
used for genomic selection.

The analysis calculates OOF predictive performance, DIC and relative support
measures, the selected method for each trait, fold stability, computational
time, heritability and accuracy in the RR-BLUP/BRR analogue, OOF genomic
values, oriented rankings, proportional selection, recurrence across traits,
and correlations among genomic values.

The breeding direction is defined before ranking: increase MMCOM, PRODCOM, CF,
DF, EP, FFC, FFP, and SS; decrease PRODEF and DCI.

## 2. Packages and model-fitting results

The packages used here have distinct roles:

- `dplyr`: grouping, joining, ordering, and summarizing result tables;
- `tidyr`: reshaping tables between long and wide formats;
- `knitr`: formatting tables in R Markdown documents;
- `here`: building project-relative file paths.

```{r packages, warning=FALSE, message=FALSE}
library(dplyr) # Group, join, and summarize tables.
library(tidyr) # Reshape data between long and wide formats.
library(knitr) # Format tables in R Markdown documents.
library(here)  # Build project-relative paths.
```

```{r table-formatting, include=FALSE}
format_table <- function(x, caption = NULL, digits = getOption("digits"), col.names = NA) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    col.names = col.names,
    table.attr = paste0(
      'class="table table-striped table-hover" ',
      'style="width:100%; margin-bottom:0;"'
    )
  )
  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

The module uses two consolidated objects: `matrices_object.rds`, which supplies phenotypes, metadata, and definitions inherited from Module 02, and `models_object.rds`, which supplies fold-level metrics, OOF predictions, and MCMC diagnostics produced in Module 03. `readRDS()` restores these objects while preserving their internal structures, and `dir.create(..., recursive = TRUE)` ensures that the derived-output directory exists before results are written.

```{r inputs}
MATRICES_FILE <- here("output", "matrices", "matrices_object.rds")
MODELS_FILE <- here("output", "models", "models_object.rds")
RESULTS_DIR <- here("output", "results", "m04")

dir.create(RESULTS_DIR, recursive = TRUE, showWarnings = FALSE)

matrices_object <- readRDS(MATRICES_FILE)
models_object <- readRDS(MODELS_FILE)

metrics <- models_object$metrics_by_fold
predictions <- models_object$predictions
convergence <- models_object$convergence_diagnostics

traits <- colnames(matrices_object$phenotypes)
methods <- models_object$settings$methods
```

The dimensions of the three input tables show the coverage of the fitted
analysis before the following derivations.

```{r input-summary}
data.frame(
  Output = c(
    "Trait × fold × method metrics",
    "OOF predictions",
    "MCMC diagnostics"
  ),
  Rows = c(
    nrow(metrics),
    nrow(predictions),
    nrow(convergence)
  ),
  # Count how many traits are represented in each input table.
  Traits = c(
    n_distinct(metrics$Trait),
    n_distinct(predictions$Trait),
    n_distinct(convergence$Trait)
  ),
  Methods = c(
    n_distinct(metrics$Method),
    n_distinct(predictions$Method),
    n_distinct(convergence$Method)
  )
) |>
  format_table(caption = "Results received from model fitting")
```

With complete Module 03 results, the table should represent 350 trait × fold × method combinations in the metrics, 10,500 OOF predictions, and 700 scalar MCMC diagnostics. These totals confirm that the following derivations start from the same coverage used for model fitting.

## 3. OOF predictive performance

The OOF estimates inherit the family-stratified design of Modules 02 and
03. They quantify prediction within the represented breeding population and do
not estimate generalization to entirely new families.

The main metrics are recalculated using the 150 out-of-fold predictions for
each trait × method combination. Each individual contributes one OOF
prediction for every method.

`group_by(Trait, Method)` separates predictions by trait and method, and `summarise()` reduces the 150 OOF rows in each group to aggregate metrics. Correlation `r`, RMSE, MAE, bias, and slope use exactly those 150 predictions. For Pearson correlation, `t_value = r * sqrt((N - 2) / (1 - r^2))` transforms `r` into a t statistic with `N - 2` degrees of freedom, and `2 * pt(..., lower.tail = FALSE)` gives the two-sided p-value. This p-value remains nominal and descriptive because training sets overlap across folds.

```{r pooled-oof-metrics}
# Combine OOF predictions and compute aggregate predictive metrics by trait and method.
pooled_metrics <- predictions |>
  group_by(Trait, Method) |>
  summarise(
    N = n(),
    r = cor(Observed, Predicted),
    RMSE = sqrt(mean((Observed - Predicted)^2)),
    MAE = mean(abs(Observed - Predicted)),
    Bias = mean(Predicted - Observed),
    Slope = cov(Observed, Predicted) / var(Predicted),
    .groups = "drop"
  ) |>
  mutate(
    t_value = r * sqrt((N - 2) / (1 - r^2)),
    p_value = 2 * pt(abs(t_value), df = N - 2, lower.tail = FALSE)
  ) |>
  select(-t_value)

# Summarize the results in `fold_summary`.
fold_summary <- metrics |>
  group_by(Trait, Method) |>
  summarise(
    Mean_DIC = mean(DIC),
    SD_DIC = sd(DIC),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_fold_RMSE = mean(RMSE),
    SD_fold_RMSE = sd(RMSE),
    Mean_fold_MAE = mean(MAE),
    SD_fold_MAE = sd(MAE),
    .groups = "drop"
  )
```

There are two summary scales in this section. `pooled_metrics` combines the 150 OOF predictions and provides the global performance of each trait × method combination; this pooled correlation is the primary reference for ranking predictive winners. `fold_summary` preserves variation across the five folds and is used later to assess stability and construct the descriptive one-standard-error tolerances.

Pearson-correlation p-values are retained as nominal descriptive measures because training sets overlap across folds. They are reported for context and are not used to determine eligibility for genomic selection.

```{r pooled-oof-preview}
pooled_metrics |>
  arrange(match(Trait, traits), desc(r), RMSE) |>
  head(14) |>
  format_table(
    digits = 5,
    caption = "Example of OOF predictive performance"
  )
```

The preview orders traits first and then prioritizes larger `r` and smaller RMSE within each trait. It verifies directly that the primary predictive comparison uses all 150 pooled OOF predictions rather than a simple average of the five fold correlations.

## 4. DIC and relative support for the methods

Following the logic used in the reference study [@silva2021brrguava], the
difference from the smallest mean DIC within each trait is converted into
relative weights:

\[
\Delta_j=\overline{DIC}_j-\min_k(\overline{DIC}_k),
\]

\[
Wprob_j=\frac{\exp(-\Delta_j/2)}{\sum_k\exp(-\Delta_k/2)},
\qquad
ER_j=\frac{Wprob_{\max}}{Wprob_j}.
\]

`Wprob` is used here as a DIC-derived relative-support weight. It is not
interpreted as a calibrated posterior probability of a model.

In the code, `Delta_DIC = Mean_DIC - min(Mean_DIC)` sets the best mean DIC within the trait to zero. `log_weight = -0.5 * Delta_DIC` implements the exponent in the formula. Before exponentiation, `max(log_weight)` is subtracted from every log-weight; this translation improves numerical stability without changing relative ratios. `Wprob = stable_weight / sum(stable_weight)` then normalizes the weights to sum to 1, and `ER = max(Wprob) / Wprob` expresses how many times the largest relative weight exceeds each method's weight.

```{r model-comparison}
# Combine fold performance and DIC to build the quantities used for model comparison.
model_comparison <- pooled_metrics |>
  left_join(fold_summary, by = c("Trait", "Method")) |>
  group_by(Trait) |>
  mutate(
    Delta_DIC = Mean_DIC - min(Mean_DIC),
    log_weight = -0.5 * Delta_DIC,
    stable_weight = exp(log_weight - max(log_weight)),
    Wprob = stable_weight / sum(stable_weight),
    # Compute the evidence ratio relative to the method with the largest Wprob in each trait.
    ER = max(Wprob) / Wprob
  ) |>
  ungroup() |>
  select(
    Trait, Method, Mean_DIC, SD_DIC, Delta_DIC, Wprob, ER,
    N, r, p_value, RMSE, MAE, Bias, Slope,
    Mean_fold_r, SD_fold_r, Mean_fold_RMSE, SD_fold_RMSE,
    Mean_fold_MAE, SD_fold_MAE
  ) |>
  arrange(match(Trait, traits), Delta_DIC)
```

The table below places Bayesian fit and OOF performance side by side. `Mean_DIC`, `Delta_DIC`, `Wprob`, and `ER` should be compared only among methods for the same trait; `r`, RMSE, and MAE describe out-of-sample predictive performance. Similarity in DIC does not replace predictive evidence.

```{r model-comparison-preview}
model_comparison |>
  select(
    Trait, Method, Mean_DIC, Delta_DIC,
    Wprob, ER, r, RMSE, MAE
  ) |>
  head(21) |>
  format_table(
    digits = 5,
    caption = "Example of the comparison among methods"
  )
```

In the joint comparison, methods with similar DIC can have different OOF performance, and the method with the largest `Wprob` need not have the largest predictive correlation. Bayesian fit and predictive ability therefore remain complementary rather than interchangeable criteria.

## 5. Selecting a method for each trait

OOF predictive ability is the primary criterion. DIC provides complementary
information about fit and complexity, while variation across folds and MCMC
diagnostics help evaluate the stability of the decision.

A method is considered predictively competitive when its mean fold correlation
is within one standard error below the largest mean and its mean RMSE is within
one standard error above the smallest mean RMSE. DIC fit is considered similar
when `Delta_DIC <= 2`. These thresholds define an operational competitive set,
not a formal equivalence test.

When RR-BLUP/BRR belongs to this competitive set and has no convergence flags,
it is preferred as the homogeneous-shrinkage benchmark with the simpler
parameterization. Otherwise, the best competitive candidate without flags is
used; when no rule-based candidate is available, the predictive winner is
retained as a provisional choice for review.

### 5.1 Predictive and DIC winners

First, the method with the best OOF predictive performance and the method with the lowest mean DIC are identified separately for each trait.

The `Delta_DIC <= 2` cutoff is the operational rule used to regard fits as having similar DIC; it is not a formal test. The number of folds is not redefined in this module: `N_FOLDS` is recovered from the Module 02 object and is used only to calculate the descriptive one-standard-error tolerances.

```{r selection-winners}
# Use Delta_DIC <= 2 as the operational cutoff for similar fit.
DIC_COMPETITIVE_THRESHOLD <- 2
# Recover the five Module 02 folds for the descriptive one-standard-error tolerances.
N_FOLDS <- matrices_object$settings$n_folds

# Best combination of correlation, RMSE, MAE, and DIC for each trait.
predictive_winners <- model_comparison |>
  group_by(Trait) |>
  arrange(desc(r), RMSE, MAE, Mean_DIC, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    Predictive_winner = Method,
    Predictive_winner_r = r,
    Predictive_winner_p_value = p_value,
    Predictive_winner_RMSE = RMSE,
    Predictive_winner_MAE = MAE
  )

# Method with the smallest mean DIC for each trait.
dic_winners <- model_comparison |>
  group_by(Trait) |>
  arrange(Mean_DIC, desc(r), RMSE, MAE, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    DIC_winner = Method,
    Minimum_DIC = Mean_DIC,
    DIC_winner_Wprob = Wprob
  )
```

The following table shows whether the method with the largest OOF correlation is also the method with the smallest mean DIC. Disagreement between the two columns is not treated as an error: predictive performance and fit–complexity address different criteria.

```{r selection-winners-table}
predictive_winners |>
  left_join(dic_winners, by = "Trait") |>
  select(
    Trait, Predictive_winner, DIC_winner,
    Predictive_winner_r, Predictive_winner_RMSE,
    Minimum_DIC
  ) |>
  format_table(
    digits = 5,
    caption = "Predictive and DIC winners for each trait"
  )
```

### 5.2 MCMC diagnostics by trait and method

Chain diagnostics are summarized before defining the set of methods eligible
for selection.

```{r selection-convergence-summary}
convergence_by_trait_method <- convergence |>
  group_by(Trait, Method) |>
  summarise(
    Runs = n_distinct(Run_id),
    Parameters_reviewed = n(),
    Parameters_flagged = sum(Review_flag),
    Runs_with_flags = n_distinct(Run_id[Review_flag]),
    Minimum_ESS = min(Effective_sample_size, na.rm = TRUE),
    Maximum_absolute_Geweke = max(abs(Geweke_Z), na.rm = TRUE),
    .groups = "drop"
  )
```

The summary below should be read as screening of the two scalar diagnostics monitored per fit in Module 03. `Parameters_flagged = 0` means only that no flags occurred in those monitored quantities; it is not proof that every marker effect has converged.

```{r selection-convergence-table}
convergence_by_trait_method |>
  arrange(match(Trait, traits), match(Method, methods)) |>
  head(21) |>
  format_table(
    digits = 4,
    caption = "Example of MCMC diagnostics summarized by trait and method"
  )
```

### 5.3 Competitive set

The one-standard-error rule is applied to mean fold correlation and RMSE. For each trait, the correlation tolerance uses the across-fold standard deviation of the method with the largest `Mean_fold_r`, divided by `sqrt(5)`; analogously, the RMSE tolerance uses the method with the smallest `Mean_fold_RMSE`. A candidate must satisfy both tolerances simultaneously.

`which.max(Mean_fold_r)` locates the method that defines the correlation reference, while `which.min(Mean_fold_RMSE)` locates the RMSE reference. Because `N_FOLDS = 5`, dividing the corresponding standard deviation by `sqrt(N_FOLDS)` produces the descriptive tolerance used by the rule. A method must satisfy both `Within_one_SE_r` and `Within_one_SE_RMSE` to be predictively competitive.

Next, `Delta_DIC <= 2` restricts the set to fits with operationally similar DIC, and any flag among the monitored diagnostics prevents classification as an automatic candidate. Because training sets overlap across folds, these standard errors are descriptive tolerances rather than independent-sample inferential standard errors.

```{r selection-competitive-set}
# Combine performance, DIC, and MCMC diagnostics to define the competitive set for each trait.
model_selection_support <- model_comparison |>
  left_join(
    convergence_by_trait_method,
    by = c("Trait", "Method")
  ) |>
  left_join(
    predictive_winners |>
      select(Trait, Predictive_winner),
    by = "Trait"
  ) |>
  group_by(Trait) |>
  mutate(
    # References for the one-standard-error rule.
    Best_mean_fold_r = max(Mean_fold_r, na.rm = TRUE),
    Best_mean_fold_r_SE =
      SD_fold_r[which.max(Mean_fold_r)] / sqrt(N_FOLDS),
    Best_mean_fold_RMSE = min(Mean_fold_RMSE, na.rm = TRUE),
    Best_mean_fold_RMSE_SE =
      SD_fold_RMSE[which.min(Mean_fold_RMSE)] / sqrt(N_FOLDS),

    # Predictive competitiveness.
    Within_one_SE_r =
      Mean_fold_r >= Best_mean_fold_r - Best_mean_fold_r_SE,
    Within_one_SE_RMSE =
      Mean_fold_RMSE <= Best_mean_fold_RMSE + Best_mean_fold_RMSE_SE,
    Predictively_competitive =
      Within_one_SE_r & Within_one_SE_RMSE,

    # Complementary support from DIC and convergence.
    DIC_competitive_threshold = DIC_COMPETITIVE_THRESHOLD,
    DIC_competitive = Delta_DIC <= DIC_COMPETITIVE_THRESHOLD,
    Competitive_for_selection =
      Predictively_competitive & DIC_competitive,
    Convergence_review = Parameters_flagged > 0,
    Automatic_candidate =
      Competitive_for_selection & !Convergence_review,

    # Candidate-set summary within each trait.
    Competitive_set_n = sum(Competitive_for_selection),
    Automatic_candidate_n = sum(Automatic_candidate),
    RRBLUP_competitive = any(
      Method == "RRBLUP" & Competitive_for_selection
    ),
    RRBLUP_automatic_candidate = any(
      Method == "RRBLUP" & Automatic_candidate
    ),
    Automatic_selection_available = any(Automatic_candidate),

    # RR-BLUP receives priority only within the automatic-candidate set.
    Selection_priority = case_when(
      Automatic_candidate & Method == "RRBLUP" ~ 1L,
      Automatic_candidate ~ 2L,
      TRUE ~ 3L
    )
  ) |>
  ungroup() |>
  arrange(
    match(Trait, traits),
    Selection_priority,
    desc(r),
    RMSE,
    MAE,
    Mean_DIC
  )
```

The example table shows each filter in the rule: competitiveness by correlation/RMSE, DIC proximity, chain-review status, and automatic eligibility. `Automatic_candidate = TRUE` requires all of these operational conditions to be satisfied.

```{r selection-competitive-table}
model_selection_support |>
  select(
    Trait, Method, Predictively_competitive,
    DIC_competitive, Convergence_review,
    Automatic_candidate, r, RMSE, Delta_DIC
  ) |>
  head(21) |>
  format_table(
    digits = 5,
    caption = "Example of the competitive-set definition"
  )
```

### 5.4 Method choice

RR-BLUP receives priority among automatic candidates as an explicit parsimony choice and homogeneous-shrinkage benchmark; this priority does not imply universal statistical superiority of RR-BLUP. If RR-BLUP is not an automatic candidate, the remaining candidates are ordered by pooled OOF correlation, followed by RMSE, MAE, and DIC. If no automatic candidate exists, the predictive winner is retained only as a provisional choice requiring review.

In the code, `filter(Automatic_candidate)` keeps only methods that simultaneously pass the predictive, DIC, and monitored-convergence conditions. `slice(1L)` retains the first method after priority ordering. Then `coalesce(Automatic_selected_method, Predictive_winner)` uses the automatic candidate when it exists and falls back to the provisional predictive winner only when that value is missing.

```{r selection-choice}
# Best automatic candidate for each trait.
automatic_selections <- model_selection_support |>
  filter(Automatic_candidate) |>
  group_by(Trait) |>
  arrange(
    Selection_priority,
    desc(r),
    RMSE,
    MAE,
    Mean_DIC,
    .by_group = TRUE
  ) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    Automatic_selected_method = Method
  )

# Final choice before attaching the selected method's metrics.
# Apply RR-BLUP priority within the automatic set and define the selected method.
selection_spec <- predictive_winners |>
  select(Trait, Predictive_winner) |>
  left_join(automatic_selections, by = "Trait") |>
  mutate(
    Selected_method = coalesce(
      Automatic_selected_method,
      Predictive_winner
    )
  ) |>
  select(Trait, Selected_method, Predictive_winner) |>
  left_join(
    model_selection_support |>
      select(
        Trait,
        Method,
        Automatic_selection_available,
        Automatic_candidate_n,
        Competitive_set_n,
        RRBLUP_competitive,
        RRBLUP_automatic_candidate,
        Convergence_review
      ),
    # Match the selected method to the support table to append the indicators used in the decision.
    by = c("Trait", "Selected_method" = "Method")
  ) |>
  mutate(
    RRBLUP_preference_applied =
      Selected_method == "RRBLUP" &
      Predictive_winner != "RRBLUP" &
      RRBLUP_automatic_candidate,
    Selection_requires_review =
      !Automatic_selection_available | Convergence_review,
    Selection_status = case_when(
      Selection_requires_review ~ "provisional_review_required",
      TRUE ~ "rule_based_selected"
    ),
    Decision_code = case_when(
      Selection_requires_review ~
        "predictive_winner_provisional_review_required",
      Selected_method == "RRBLUP" &
        RRBLUP_preference_applied ~
        "rrblup_preferred_among_competitive_models",
      Selected_method == "RRBLUP" &
        Predictive_winner == "RRBLUP" ~
        "rrblup_best_predictive_model",
      TRUE ~ "best_non_rrblup_competitive_model"
    ),
    Operational_competitive_set_rule =
      "one_se_r_and_rmse_plus_delta_dic_le_2"
  ) |>
  select(-Predictive_winner, -Convergence_review) |>
  arrange(match(Trait, traits))
```

The table summarizes the result of this hierarchy. `rule_based_selected` means that at least one automatic candidate existed and the rule could be applied; `provisional_review_required` means that the choice was retained for inspection rather than automatically validated.

```{r selection-choice-table}
selection_spec |>
  select(
    Trait, Selected_method, Selection_status,
    RRBLUP_preference_applied,
    Competitive_set_n, Automatic_candidate_n
  ) |>
  format_table(caption = "Method chosen after applying the hierarchical rule")
```

### 5.5 Performance of the selected methods

Metrics and diagnostics for the method actually selected are then extracted
from the support table.

```{r selected-performance}
# Extract predictive metrics, DIC, and diagnostics for the selected method in each trait.
selected_performance <- model_selection_support |>
  inner_join(
    selection_spec |>
      select(Trait, Selected_method),
    by = "Trait"
  ) |>
  # Restrict the data to observations relevant at this stage.
  filter(Method == Selected_method) |>
  # Build the columns that form the output of this block.
  transmute(
    Trait,
    Selected_method,
    r,
    p_value,
    RMSE,
    MAE,
    Bias,
    Slope,
    Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    Mean_fold_r,
    SD_fold_r,
    Mean_fold_RMSE,
    SD_fold_RMSE,
    Within_one_SE_r,
    Within_one_SE_RMSE,
    Predictively_competitive,
    DIC_competitive,
    Competitive_for_selection,
    Automatic_candidate,
    Competitive_set_n,
    Automatic_candidate_n,
    RRBLUP_competitive,
    RRBLUP_automatic_candidate,
    Parameters_flagged,
    Runs_with_flags,
    Minimum_ESS,
    Maximum_absolute_Geweke,
    Convergence_review
  )
```

The following table contains only the method actually selected for each trait, combining its OOF evidence, DIC, and number of monitored parameters with a flag. It should be read together with `Selection_status`, because a provisional choice can appear here even when review is required.

```{r selected-performance-table}
selected_performance |>
  select(
    Trait, Selected_method, r, p_value, RMSE,
    Mean_DIC, Delta_DIC, Parameters_flagged
  ) |>
  format_table(
    digits = 5,
    caption = "Performance and diagnostics of the selected methods"
  )
```

### 5.6 Final selection result and predictive evidence

Finally, the selected method, predictive and DIC winners, OOF direction, and a
nominal signal class are brought together for each trait. The nominal class is shown only to make weak positive correlations explicit; it is not used as a selection filter. The $r > 0$ boundary separates only positive from nonpositive predictive direction; it is neither a significance test nor a minimum predictive-utility threshold.

```{r final-model-selection}
# Combine the selected method, predictive and DIC winners, and OOF evidence in the final trait-level table.
model_selection <- selection_spec |>
  left_join(
    selected_performance,
    # Join selected-method performance using trait and method as keys.
    by = c("Trait", "Selected_method")
  ) |>
  left_join(predictive_winners, by = "Trait") |>
  left_join(dic_winners, by = "Trait") |>
  mutate(
    Predictive_loss = Predictive_winner_r - r,
    Predictive_winner_agrees =
      Selected_method == Predictive_winner,
    DIC_winner_agrees =
      Selected_method == DIC_winner,
    # Use zero only as a direction boundary; r > 0 does not imply significance or minimum utility.
    OOF_predictive_direction = if_else(
      r > 0,
      "positive",
      "nonpositive"
    ),
    # Keep p < 0.05 only as a descriptive nominal class, without filtering rankings.
    OOF_nominal_signal = case_when(
      r > 0 & p_value < 0.05 ~ "positive_nominal_p_lt_0.05",
      r > 0 ~ "positive_nominal_p_ge_0.05",
      TRUE ~ "nonpositive"
    )
  ) |>
  arrange(match(Trait, traits))

# Count how many traits have positive or nonpositive OOF correlation.
oof_direction_summary <- model_selection |>
  count(OOF_predictive_direction, name = "Traits") |>
  arrange(OOF_predictive_direction)
```

This is the main decision table by trait. In addition to the selected method, it preserves the predictive and DIC winners, records whether the RR-BLUP preference was applied, and makes the direction and nominal strength of the OOF signal explicit.

```{r model-selection-table}
model_selection |>
  select(
    Trait, Selected_method, Predictive_winner, DIC_winner,
    r, p_value, RMSE, Mean_DIC, Delta_DIC,
    RRBLUP_preference_applied,
    Selection_status, OOF_predictive_direction, OOF_nominal_signal
  ) |>
  format_table(
    digits = 5,
    caption = "Selected method for each trait"
  )
```

The direction summary counts how many traits end with positive or nonpositive OOF correlation for the selected method. It does not classify magnitude, practical utility, or significance.

```{r oof-direction-table}
format_table(
  oof_direction_summary,
  caption = "OOF predictive direction of the traits"
)
```

## 6. Stability across folds for the selected methods

Fold-level metrics show whether pooled performance depends strongly on a particular part of the partition. Sign changes in correlation, large RMSE dispersion, or clearly discrepant folds should be interpreted as descriptive instability; with only five overlapping folds, this table is not a formal heterogeneity test.

```{r selected-fold-diagnostics}
selected_fold_diagnostics <- metrics |>
  inner_join(
    selection_spec |> select(Trait, Selected_method),
    by = c("Trait" = "Trait", "Method" = "Selected_method")
  ) |>
  transmute(Trait, Fold, Method, Correlation, RMSE, MAE) |>
  arrange(match(Trait, traits), Fold)

selected_fold_diagnostics |>
  head(15) |>
  format_table(
    digits = 5,
    caption = "Example of fold performance for the selected methods"
  )
```

The fold-level preview should be read for changes in correlation sign and marked RMSE/MAE dispersion. Because folds share most of their training data, these differences are descriptive stability diagnostics rather than independent replications.

## 7. Computational time

The table summarizes times recorded across the 50 fits of each method. These values support operational comparison within this execution, but depend on hardware, system load, implementation, and I/O; they should not be treated as universal speed benchmarks.

```{r runtime}
# Summarize the computational cost of the evaluated methods.
runtime_summary <- metrics |>
  group_by(Method) |>
  summarise(
    Fits = n(),
    Total_seconds = sum(Elapsed_seconds),
    Mean_seconds = mean(Elapsed_seconds),
    SD_seconds = sd(Elapsed_seconds),
    Median_seconds = median(Elapsed_seconds),
    Minimum_seconds = min(Elapsed_seconds),
    Maximum_seconds = max(Elapsed_seconds),
    .groups = "drop"
  ) |>
  arrange(match(Method, methods))

format_table(
  runtime_summary,
  digits = 3,
  caption = "Computational time of the seven methods"
)
```

## 8. Marker-based heritability analogue and adjusted accuracy in RR-BLUP/BRR

Following Silva et al. [@silva2021brrguava], a marker-based additive-variance
analogue is obtained from the common marker variance and the \(\sum 2pq\)
denominator. Fold-level adjusted accuracy is defined as
\(r_f/\sqrt{H_f^2}\). The resulting \(H^2\) is reported as a model-based
heritability analogue rather than as an independent classical estimate of
narrow-sense heritability. The quantity named `Predictive_accuracy` is a derived ratio, `r_f/sqrt(H_f^2)`, rather than an observed correlation between true and predicted genetic values. Because it depends on sample estimates of `r_f` and `H_f^2`, it may be negative or exceed 1; it should therefore be interpreted as an adjusted-accuracy index under this definition, not as a measure necessarily bounded to [0, 1].

```{r rrblup-heritability}
# Compute and organize the RR-BLUP/BRR model-based heritability summary.
rrblup_variance_components <- convergence |>
  filter(
    Method == "RRBLUP",
    Parameter %in% c("Residual_variance", "Marker_variance")
  ) |>
  select(Run_id, Trait, Fold, Method, Parameter, Posterior_mean) |>
  pivot_wider(names_from = Parameter, values_from = Posterior_mean)

rrblup_fold_correlations <- metrics |>
  filter(Method == "RRBLUP") |>
  select(Trait, Fold, Correlation)

# Organize the cross-validation and fitting structure.
rrblup_heritability_by_fold <- rrblup_variance_components |>
  left_join(rrblup_fold_correlations, by = c("Trait", "Fold")) |>
  mutate(
    Marker_variance_denominator = matrices_object$denominator,
    Additive_variance = Marker_variance * Marker_variance_denominator,
    H2 = Additive_variance / (Additive_variance + Residual_variance),
    Predictive_accuracy = Correlation / sqrt(H2)
  ) |>
  arrange(match(Trait, traits), Fold)

# Summarize the RR-BLUP/BRR model-based heritability analogue and adjusted accuracy across folds.
rrblup_heritability_summary <- rrblup_heritability_by_fold |>
  group_by(Trait) |>
  summarise(
    Mean_H2 = mean(H2),
    SD_H2 = sd(H2),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_predictive_accuracy = mean(Predictive_accuracy),
    SD_predictive_accuracy = sd(Predictive_accuracy),
    .groups = "drop"
  ) |>
  arrange(match(Trait, traits))

format_table(
  rrblup_heritability_summary,
  digits = 5,
  caption = "Marker-based heritability analogue and adjusted predictive accuracy in RR-BLUP/BRR"
)
```

The RR-BLUP/BRR summary should be interpreted trait by trait: `Mean_H2` describes the model-derived heritability analogue, while `Mean_predictive_accuracy` summarizes `Correlation / sqrt(H2)` across folds. This ratio is not a conventional correlation and may lie outside [0, 1].

## 9. OOF genomic values and selection direction

Each individual's genomic value is cross-fitted: it comes from the fold in
which that individual's phenotype was masked. These OOF genomic values provide
a validation-consistent ranking because an individual's own phenotype does not
enter its score. They are not final breeding values from a model refitted to
all 150 phenotypes.

Selection direction is applied to the genomic value only when the ranking is constructed. The original genomic value remains unchanged. For each trait, `Genomic_value` is exactly the OOF `Genomic_component` produced by the selected method; different traits may therefore use genomic values supplied by different methods.

```{r selected-oof-genomic-values}
# Define the desired selection direction and link OOF genomic values to the selected method for each trait.
selection_objective_map <- c(
  MMCOM = "increase",
  PRODCOM = "increase",
  PRODEF = "decrease",
  CF = "increase",
  DF = "increase",
  DCI = "decrease",
  EP = "increase",
  FFC = "increase",
  FFP = "increase",
  SS = "increase"
)

# Organize the desired selection direction and ranking rule for each trait.
selection_objectives <- data.frame(
  Trait = traits,
  Selection_objective = unname(selection_objective_map[traits]),
  stringsAsFactors = FALSE
) |>
  mutate(
    Ranking_rule = case_when(
      Selection_objective == "increase" ~ "maximize_genomic_value",
      Selection_objective == "decrease" ~ "minimize_genomic_value",
      TRUE ~ NA_character_
    )
  )

# Organize the estimated genomic component for each individual.
selected_oof_genomic_values <- predictions |>
  inner_join(
    model_selection |>
      select(
        Trait, Selected_method, OOF_predictive_direction,
        OOF_nominal_signal, r, p_value
      ),
    by = "Trait"
  ) |>
  inner_join(selection_objectives, by = "Trait") |>
  filter(Method == Selected_method) |>
  transmute(
    IND, Trait, Fold, Method, OOF_predictive_direction,
    OOF_nominal_signal, OOF_r = r, OOF_nominal_p_value = p_value,
    Selection_objective, Ranking_rule,
    Genomic_value = Genomic_component
  ) |>
  arrange(match(Trait, traits), IND)

format_table(
  selection_objectives,
  caption = "Selection direction defined for each trait"
)
```

## 10. Exploratory proportional candidate ranking by trait

Candidate rankings are shown for traits whose selected method has positive OOF
correlation. Nominal p-values remain descriptive and do not determine whether a
ranking is produced. Positive correlations with nominal `p >= 0.05` are kept
explicitly as weak/exploratory signals and should not be interpreted as
demonstrated predictive utility. Within each positive-OOF trait, the 20% of
individuals with the largest oriented score are listed; this 20% is a
configurable rule in this analysis, not an estimated optimal selection intensity. With 150 available candidates, 20% corresponds to 30 individuals. For traits whose objective is to decrease, the score is the negative genomic value.

`IND` is used only as a deterministic secondary ordering key when two scores are exactly tied. The selected proportion remains fixed even if a tie occurs at the boundary; `Exact_boundary_tie = TRUE` only reports that the last included and first excluded candidates have the same score, so there is no genomic-score evidence separating them at that cutoff.

`ceiling(SELECTION_PROPORTION * N_available)` rounds the target count upward, while `pmax(1L, ...)` guarantees at least one candidate when individuals are available. In the cutoff table, `lead(Selection_score)` moves the next score onto the current row; `first(N_selected)` identifies the last selected position so that its score can be compared directly with the first excluded score.

```{r genomic-ranking}
# Orient genomic values by the selection objective and rank individuals within each trait.
genomic_ranking <- selected_oof_genomic_values |>
  mutate(
    Selection_score = case_when(
      Selection_objective == "increase" ~ Genomic_value,
      Selection_objective == "decrease" ~ -Genomic_value,
      TRUE ~ NA_real_
    )
  ) |>
  group_by(Trait) |>
  arrange(desc(Selection_score), IND, .by_group = TRUE) |>
  mutate(Rank = row_number()) |>
  ungroup() |>
  arrange(match(Trait, traits), Rank)

# Define the candidate proportion displayed in each exploratory ranking.
SELECTION_PROPORTION <- 0.20

positive_oof_traits <- model_selection |>
  filter(OOF_predictive_direction == "positive") |>
  pull(Trait)

# Compute how many individuals compose the prioritized proportion in each positive-OOF trait.
selection_sizes <- genomic_ranking |>
  filter(Trait %in% positive_oof_traits) |>
  count(Trait, name = "N_available") |>
  mutate(
    Selection_proportion = SELECTION_PROPORTION,
    N_selected = pmax(
      1L,
      as.integer(ceiling(SELECTION_PROPORTION * N_available))
    )
  ) |>
  arrange(match(Trait, traits))

# Record the cutoff score and genomic-value interval corresponding to each ranking.
selection_cutoff_summary <- genomic_ranking |>
  filter(Trait %in% positive_oof_traits) |>
  inner_join(
    selection_sizes |>
      select(Trait, N_available, N_selected),
    by = "Trait"
  ) |>
  arrange(match(Trait, traits), Rank) |>
  group_by(Trait) |>
  mutate(Next_score = lead(Selection_score)) |>
  summarise(
    N_available = first(N_available),
    N_selected = first(N_selected),
    Score_last_selected = Selection_score[first(N_selected)],
    Score_first_not_selected = Next_score[first(N_selected)],
    Exact_boundary_tie =
      !is.na(Score_first_not_selected) &
      Score_last_selected == Score_first_not_selected,
    .groups = "drop"
  ) |>
  arrange(match(Trait, traits))

# Retain candidates included in the defined ranking proportion for each positive-OOF trait.
selected_individuals_by_trait <- genomic_ranking |>
  inner_join(selection_sizes, by = "Trait") |>
  filter(Rank <= N_selected) |>
  left_join(
    matrices_object$metadata |>
      select(IND, FAM),
    by = "IND"
  ) |>
  select(
    Trait, Rank, IND, FAM, Method,
    Selection_objective, Ranking_rule,
    Genomic_value, Selection_score, OOF_predictive_direction,
    OOF_nominal_signal, OOF_r, OOF_nominal_p_value,
    Selection_proportion, N_available, N_selected
  ) |>
  arrange(match(Trait, traits), Rank)
```

The first table confirms, for each eligible trait, how many candidates were available and how many compose the 20% proportion.

```{r selection-size-table}
format_table(
  selection_sizes,
  digits = 2,
  caption = "Exploratory candidate-ranking size for each positive-OOF trait"
)
```

The cutoff table shows the score of the last included candidate and the first candidate not included. `Exact_boundary_tie` should be checked before interpreting the cutoff as a substantive separation between candidates.

```{r selection-cutoff-table}
format_table(
  selection_cutoff_summary,
  digits = 5,
  caption = "Scores at the proportional-selection boundary"
)
```

The preview shows only the first five names in each ranking for readability. The complete object retains every candidate belonging to the prioritized 20%.

```{r selected-individuals-preview}
selected_individuals_by_trait |>
  group_by(Trait) |>
  slice_head(n = 5) |>
  ungroup() |>
  format_table(
    digits = 5,
    caption = "Top five candidates in each positive-OOF trait ranking"
  )
```

## 11. Recurrence across traits

Recurrence is descriptive: among traits with positive OOF correlation, it counts how many times each individual appears among the most favorable 20%. `Selection_frequency` divides this count by the number of positive-OOF traits, not necessarily by all ten original traits. Weak positive OOF signals remain exploratory. No weights are used and no multi-trait selection index is constructed; appearing in several lists does not demonstrate joint multi-trait superiority.

Recurrence classes are descriptive conventions: high for four or more traits, intermediate for three, broad for two, specific for one, and none for zero. The `multitrait_candidates` object retains individuals recurring in at least two traits so potentially useful information is not discarded before downstream presentation. These cutoffs organize the summary and are not statistical-significance thresholds.

After the `left_join()` with all individuals, `replace_na()` converts missing counts to zero and missing trait lists to empty text. `Selection_frequency = N_traits / length(positive_oof_traits)` uses only traits with positive OOF correlation in the denominator. `pivot_wider()` places trait-specific ranks in separate columns, and `filter(N_traits >= 2L)` defines the complete recurrent-candidate object used downstream.

```{r multitrait-recurrence}
# Count individual recurrence across exploratory rankings for traits with positive OOF correlation.
recurrence_counts <- selected_individuals_by_trait |>
  arrange(match(Trait, traits), Rank) |>
  group_by(IND, FAM) |>
  summarise(
    N_traits = n(),
    Traits = paste(Trait, collapse = ", "),
    Best_rank = min(Rank),
    Mean_rank_present = mean(Rank),
    .groups = "drop"
  )

# Join recurrence counts to metadata and compute recurrence frequency and class.
multitrait_recurrence <- matrices_object$metadata |>
  select(IND, FAM) |>
  left_join(recurrence_counts, by = c("IND", "FAM")) |>
  mutate(
    N_traits = tidyr::replace_na(N_traits, 0L),
    Traits = tidyr::replace_na(Traits, ""),
    Best_rank = case_when(
      N_traits == 0L ~ NA_integer_,
      TRUE ~ Best_rank
    ),
    Mean_rank_present = case_when(
      N_traits == 0L ~ NA_real_,
      TRUE ~ Mean_rank_present
    ),
    Selection_frequency = N_traits / length(positive_oof_traits),
    # Classify recurrence by number of traits; these cutoffs are descriptive.
    Recurrence_class = case_when(
      N_traits >= 4L ~ "high_4plus",
      N_traits == 3L ~ "intermediate_3",
      N_traits == 2L ~ "broad_2",
      N_traits == 1L ~ "specific_1",
      TRUE ~ "none_0"
    )
  ) |>
  arrange(desc(N_traits), FAM, IND)

# Summarize how many individuals belong to each recurrence class.
recurrence_distribution <- multitrait_recurrence |>
  count(N_traits, Recurrence_class, name = "N_individuals") |>
  mutate(
    Proportion_of_population =
      N_individuals / nrow(matrices_object$metadata)
  ) |>
  arrange(desc(N_traits))

# Reshape trait-specific ranks to one row per recurrent candidate.
candidate_rank_wide <- selected_individuals_by_trait |>
  semi_join(
    multitrait_recurrence |>
      filter(N_traits >= 2L),
    by = c("IND", "FAM")
  ) |>
  select(IND, FAM, Trait, Rank) |>
  pivot_wider(
    names_from = Trait,
    values_from = Rank,
    names_glue = "{Trait}_rank"
  )

# Retain candidates recurring in at least two traits and attach their trait-specific ranks.
multitrait_candidates <- multitrait_recurrence |>
  # Retain in the complete object candidates present in at least two traits.
  filter(N_traits >= 2L) |>
  left_join(candidate_rank_wide, by = c("IND", "FAM")) |>
  arrange(desc(N_traits), FAM, IND)

format_table(
  recurrence_distribution,
  digits = 4,
  caption = "Distribution of recurrence across traits with positive OOF correlation"
)
```

The recurrence distribution reports how many individuals occur in zero, one, two, or more selection lists. The `multitrait_candidates` object keeps the full threshold of at least two traits; none of these counts turns recurrence into a multi-trait selection index.

## 12. Correlations among OOF genomic values

These correlations describe associations among the OOF genomic values produced by the selected method for each trait. Because different traits may use different methods and the values are cross-fitted, the matrix summarizes associations among predictive scores rather than jointly estimated genetic covariance. It should therefore not be interpreted as a genetic-correlation matrix from a multivariate model.

`pivot_wider(names_from = Trait, values_from = Genomic_value)` places the ten traits in columns while retaining one row per individual. `as.matrix(...[, traits, drop = FALSE])` selects those columns in the order defined by `traits` and preserves matrix structure; `cor()` then computes Pearson correlations among the OOF-score columns.

```{r oof-correlations}
selected_genomic_wide <- selected_oof_genomic_values |>
  select(IND, Trait, Genomic_value) |>
  pivot_wider(names_from = Trait, values_from = Genomic_value)

selected_oof_correlations <- cor(
  as.matrix(selected_genomic_wide[, traits, drop = FALSE])
)

format_table(
  round(selected_oof_correlations, 3),
  caption = "Correlations among OOF genomic values"
)
```

The resulting matrix should be read as association among OOF scores: positive values indicate individuals tending to receive scores in the same direction for two traits, whereas negative values indicate opposing directions. The interpretation is not joint genetic correlation.

## 13. Saving the results

At the end, `write.csv(..., row.names = FALSE)` writes the derived tables without adding an artificial row-number column. The correlation matrix preserves its row labels with `row.names = TRUE`. `saveRDS()` collects the main tables, rules, and objects in `module04_results.rds`, preserving their classes for direct reading in Module 05. No genomic model is refitted at this stage.

```{r save-results}
# Save the tables derived by the module before assembling the consolidated RDS object.
write.csv(model_comparison, file.path(RESULTS_DIR, "model_comparison.csv"), row.names = FALSE)
write.csv(model_selection_support, file.path(RESULTS_DIR, "model_selection_support.csv"), row.names = FALSE)
write.csv(model_selection, file.path(RESULTS_DIR, "model_selection.csv"), row.names = FALSE)
write.csv(selection_spec, file.path(RESULTS_DIR, "selection_decision_spec.csv"), row.names = FALSE)
write.csv(oof_direction_summary, file.path(RESULTS_DIR, "oof_direction_summary.csv"), row.names = FALSE)
write.csv(convergence_by_trait_method, file.path(RESULTS_DIR, "convergence_by_trait_method.csv"), row.names = FALSE)
write.csv(selected_fold_diagnostics, file.path(RESULTS_DIR, "selected_fold_diagnostics.csv"), row.names = FALSE)
write.csv(runtime_summary, file.path(RESULTS_DIR, "runtime_summary.csv"), row.names = FALSE)
write.csv(rrblup_heritability_by_fold, file.path(RESULTS_DIR, "rrblup_heritability_by_fold.csv"), row.names = FALSE)
write.csv(rrblup_heritability_summary, file.path(RESULTS_DIR, "rrblup_heritability_summary.csv"), row.names = FALSE)
write.csv(selection_objectives, file.path(RESULTS_DIR, "selection_objectives.csv"), row.names = FALSE)
write.csv(selected_oof_genomic_values, file.path(RESULTS_DIR, "selected_oof_genomic_values.csv"), row.names = FALSE)
# Save the consolidated output for downstream use.
write.csv(genomic_ranking, file.path(RESULTS_DIR, "genomic_ranking.csv"), row.names = FALSE)
write.csv(selection_sizes, file.path(RESULTS_DIR, "selection_sizes.csv"), row.names = FALSE)
write.csv(selection_cutoff_summary, file.path(RESULTS_DIR, "selection_cutoff_summary.csv"), row.names = FALSE)
write.csv(selected_individuals_by_trait, file.path(RESULTS_DIR, "selected_individuals_by_trait.csv"), row.names = FALSE)
write.csv(multitrait_recurrence, file.path(RESULTS_DIR, "multitrait_recurrence.csv"), row.names = FALSE)
write.csv(recurrence_distribution, file.path(RESULTS_DIR, "recurrence_distribution.csv"), row.names = FALSE)
write.csv(multitrait_candidates, file.path(RESULTS_DIR, "multitrait_candidates.csv"), row.names = FALSE)
# Save the consolidated output for downstream use.
write.csv(as.data.frame(selected_oof_correlations), file.path(RESULTS_DIR, "selected_oof_correlations.csv"), row.names = TRUE)

# Save the consolidated output for downstream use.
saveRDS(
  list(
    traits = traits,
    methods = methods,
    model_comparison = model_comparison,
    model_selection_support = model_selection_support,
    selection_spec = selection_spec,
    model_selection = model_selection,
    oof_direction_summary = oof_direction_summary,
    convergence_by_trait_method = convergence_by_trait_method,
    selected_fold_diagnostics = selected_fold_diagnostics,
    runtime = runtime_summary,
    rrblup_heritability_by_fold = rrblup_heritability_by_fold,
    rrblup_heritability_summary = rrblup_heritability_summary,
    selection_objectives = selection_objectives,
    selected_oof_genomic_values = selected_oof_genomic_values,
    genomic_ranking = genomic_ranking,
    selection_proportion = SELECTION_PROPORTION,
    positive_oof_traits = positive_oof_traits,
    selection_sizes = selection_sizes,
    selection_cutoff_summary = selection_cutoff_summary,
    selected_individuals_by_trait = selected_individuals_by_trait,
    multitrait_recurrence = multitrait_recurrence,
    recurrence_distribution = recurrence_distribution,
    multitrait_candidates = multitrait_candidates,
    selected_oof_correlations = selected_oof_correlations
  ),
  file.path(RESULTS_DIR, "module04_results.rds")
)
```

## 14. Next step

Module 05 uses only these derivations to build the summary tables and figures. The OOF genomic values remain cross-fitted validation scores; this workflow does not transform them into final breeding values from a refit using all 150 phenotypes.

<div style="display:flex; justify-content:space-between; gap:1rem; margin-top:2rem; padding-top:1rem; border-top:1px solid #ddd;">
<a href="03_models_en.html">← Module 03 — Model fitting</a>
<a href="05_results_en.html">Module 05 — Results synthesis →</a>
</div>
