---
title: "Derivation and selection of genomic results"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

Using the predictions, metrics, and diagnostics obtained in Module 03, this
module compares the seven methods within each trait and derives the results
used for genomic selection.

The analysis calculates OOF predictive performance, DIC and relative support
measures, the selected method for each trait, fold stability, computational
time, heritability and accuracy in the RR-BLUP/BRR analogue, OOF genomic
values, oriented rankings, proportional selection, recurrence across traits,
and correlations among genomic values.

The breeding direction is defined before ranking: increase MMCOM, PRODCOM, CF,
DF, EP, FFC, FFP, and SS; decrease PRODEF and DCI.

## 2. Packages and model-fitting results

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(knitr)
  library(here)
})
```

```{r inputs}
MATRICES_FILE <- here("output", "matrices", "matrices_object.rds")
MODELS_FILE <- here("output", "models", "models_object.rds")
RESULTS_DIR <- here("output", "results", "m04")

dir.create(RESULTS_DIR, recursive = TRUE, showWarnings = FALSE)

matrices_object <- readRDS(MATRICES_FILE)
models_object <- readRDS(MODELS_FILE)

metrics <- models_object$metrics_by_fold
predictions <- models_object$predictions
convergence <- models_object$convergence_diagnostics

traits <- colnames(matrices_object$phenotypes)
methods <- models_object$settings$methods
```

The dimensions of the three input tables show the coverage of the fitted
analysis before the following derivations.

```{r input-summary}
data.frame(
  Output = c(
    "Trait × fold × method metrics",
    "OOF predictions",
    "MCMC diagnostics"
  ),
  Rows = c(
    nrow(metrics),
    nrow(predictions),
    nrow(convergence)
  ),
  Traits = c(
    n_distinct(metrics$Trait),
    n_distinct(predictions$Trait),
    n_distinct(convergence$Trait)
  ),
  Methods = c(
    n_distinct(metrics$Method),
    n_distinct(predictions$Method),
    n_distinct(convergence$Method)
  )
) |>
  kable(caption = "Results received from model fitting")
```

## 3. OOF predictive performance

The main metrics are recalculated using the 150 out-of-fold predictions for
each trait × method combination. Each individual contributes one OOF
prediction for every method.

```{r pooled-oof-metrics}
pooled_metrics <- predictions |>
  group_by(Trait, Method) |>
  summarise(
    N = n(),
    r = cor(Observed, Predicted),
    RMSE = sqrt(mean((Observed - Predicted)^2)),
    MAE = mean(abs(Observed - Predicted)),
    Bias = mean(Predicted - Observed),
    Slope = cov(Observed, Predicted) / var(Predicted),
    .groups = "drop"
  ) |>
  mutate(
    t_value = r * sqrt((N - 2) / (1 - r^2)),
    p_value = 2 * pt(abs(t_value), df = N - 2, lower.tail = FALSE)
  ) |>
  select(-t_value)

fold_summary <- metrics |>
  group_by(Trait, Method) |>
  summarise(
    Mean_DIC = mean(DIC),
    SD_DIC = sd(DIC),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_fold_RMSE = mean(RMSE),
    SD_fold_RMSE = sd(RMSE),
    Mean_fold_MAE = mean(MAE),
    SD_fold_MAE = sd(MAE),
    .groups = "drop"
  )
```

Pearson-correlation p-values are retained as nominal descriptive measures
because training sets overlap across folds.

```{r pooled-oof-preview}
pooled_metrics |>
  arrange(match(Trait, traits), desc(r), RMSE) |>
  head(14) |>
  kable(
    digits = 5,
    caption = "Example of OOF predictive performance"
  )
```

## 4. DIC and relative support for the methods

Following the logic used in the reference study [@silva2021brrguava], the
difference from the smallest mean DIC within each trait is converted into
relative weights:

\[
\Delta_j=DIC_j-\min(DIC),
\]

\[
Wprob_j=\frac{\exp(-\Delta_j/2)}{\sum_k\exp(-\Delta_k/2)},
\qquad
ER_j=\frac{Wprob_{\max}}{Wprob_j}.
\]

```{r model-comparison}
model_comparison <- pooled_metrics |>
  left_join(fold_summary, by = c("Trait", "Method")) |>
  group_by(Trait) |>
  mutate(
    Delta_DIC = Mean_DIC - min(Mean_DIC),
    log_weight = -0.5 * Delta_DIC,
    stable_weight = exp(log_weight - max(log_weight)),
    Wprob = stable_weight / sum(stable_weight),
    ER = max(Wprob) / Wprob
  ) |>
  ungroup() |>
  select(
    Trait, Method, Mean_DIC, SD_DIC, Delta_DIC, Wprob, ER,
    N, r, p_value, RMSE, MAE, Bias, Slope,
    Mean_fold_r, SD_fold_r, Mean_fold_RMSE, SD_fold_RMSE,
    Mean_fold_MAE, SD_fold_MAE
  ) |>
  arrange(match(Trait, traits), Delta_DIC)
```

```{r model-comparison-preview}
model_comparison |>
  select(
    Trait, Method, Mean_DIC, Delta_DIC,
    Wprob, ER, r, RMSE, MAE
  ) |>
  head(21) |>
  kable(
    digits = 5,
    caption = "Example of the comparison among methods"
  )
```

## 5. Selecting a method for each trait

OOF predictive ability is the primary criterion. DIC provides complementary
information about fit and complexity, while variation across folds and MCMC
diagnostics help evaluate the stability of the decision.

A method is considered predictively competitive when its mean fold correlation
is within one standard error below the largest mean and its mean RMSE is within
one standard error above the smallest mean RMSE. DIC fit is considered similar
when `Delta_DIC <= 2`. These thresholds are operational comparison rules, not
formal equivalence tests.

When RR-BLUP belongs to the competitive set and has no convergence flags, it is
preferred because of its simpler parameterization and direct interpretation of
variance components. Otherwise, the best competitive candidate without flags
is used; when no automatic candidate is available, the predictive winner is
retained as a provisional choice for review.

```{r model-selection}
DIC_EQUIVALENCE_THRESHOLD <- 2
N_FOLDS <- matrices_object$settings$n_folds

predictive_winners <- model_comparison |>
  group_by(Trait) |>
  arrange(desc(r), RMSE, MAE, Mean_DIC, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    Predictive_winner = Method,
    Predictive_winner_r = r,
    Predictive_winner_p_value = p_value,
    Predictive_winner_RMSE = RMSE,
    Predictive_winner_MAE = MAE
  )

dic_winners <- model_comparison |>
  group_by(Trait) |>
  arrange(Mean_DIC, desc(r), RMSE, MAE, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    DIC_winner = Method,
    Minimum_DIC = Mean_DIC,
    DIC_winner_Wprob = Wprob
  )

convergence_by_trait_method <- convergence |>
  group_by(Trait, Method) |>
  summarise(
    Runs = n_distinct(Run_id),
    Parameters_reviewed = n(),
    Parameters_flagged = sum(Review_flag),
    Runs_with_flags = n_distinct(Run_id[Review_flag]),
    Minimum_ESS = min(Effective_sample_size, na.rm = TRUE),
    Maximum_absolute_Geweke = max(abs(Geweke_Z), na.rm = TRUE),
    .groups = "drop"
  )

model_selection_support <- model_comparison |>
  left_join(
    convergence_by_trait_method,
    by = c("Trait", "Method")
  ) |>
  left_join(
    predictive_winners |>
      select(Trait, Predictive_winner),
    by = "Trait"
  ) |>
  group_by(Trait) |>
  mutate(
    Best_mean_fold_r = max(Mean_fold_r, na.rm = TRUE),
    Best_mean_fold_r_SE =
      SD_fold_r[which.max(Mean_fold_r)] / sqrt(N_FOLDS),
    Best_mean_fold_RMSE = min(Mean_fold_RMSE, na.rm = TRUE),
    Best_mean_fold_RMSE_SE =
      SD_fold_RMSE[which.min(Mean_fold_RMSE)] / sqrt(N_FOLDS),
    Within_one_SE_r =
      Mean_fold_r >= Best_mean_fold_r - Best_mean_fold_r_SE,
    Within_one_SE_RMSE =
      Mean_fold_RMSE <= Best_mean_fold_RMSE + Best_mean_fold_RMSE_SE,
    Predictively_competitive =
      Within_one_SE_r & Within_one_SE_RMSE,
    DIC_equivalence_threshold = DIC_EQUIVALENCE_THRESHOLD,
    DIC_competitive = Delta_DIC <= DIC_EQUIVALENCE_THRESHOLD,
    Competitive_for_selection =
      Predictively_competitive & DIC_competitive,
    Convergence_review = Parameters_flagged > 0,
    Automatic_candidate =
      Competitive_for_selection & !Convergence_review,
    Competitive_set_n = sum(Competitive_for_selection),
    Automatic_candidate_n = sum(Automatic_candidate),
    RRBLUP_competitive = any(
      Method == "RRBLUP" & Competitive_for_selection
    ),
    RRBLUP_automatic_candidate = any(
      Method == "RRBLUP" & Automatic_candidate
    ),
    Automatic_selection_available = any(Automatic_candidate),
    Selection_priority = case_when(
      Automatic_candidate & Method == "RRBLUP" ~ 1L,
      Automatic_candidate ~ 2L,
      TRUE ~ 3L
    )
  ) |>
  ungroup() |>
  arrange(
    match(Trait, traits),
    Selection_priority,
    desc(r),
    RMSE,
    MAE,
    Mean_DIC
  )

automatic_selections <- model_selection_support |>
  filter(Automatic_candidate) |>
  group_by(Trait) |>
  arrange(
    Selection_priority,
    desc(r),
    RMSE,
    MAE,
    Mean_DIC,
    .by_group = TRUE
  ) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    Automatic_selected_method = Method
  )

selection_spec <- predictive_winners |>
  select(Trait, Predictive_winner) |>
  left_join(automatic_selections, by = "Trait") |>
  mutate(
    Selected_method = coalesce(
      Automatic_selected_method,
      Predictive_winner
    )
  ) |>
  select(Trait, Selected_method, Predictive_winner) |>
  left_join(
    model_selection_support |>
      select(
        Trait,
        Method,
        Automatic_selection_available,
        Automatic_candidate_n,
        Competitive_set_n,
        RRBLUP_competitive,
        RRBLUP_automatic_candidate,
        Convergence_review
      ),
    by = c("Trait", "Selected_method" = "Method")
  ) |>
  mutate(
    RRBLUP_preference_applied =
      Selected_method == "RRBLUP" &
      Predictive_winner != "RRBLUP" &
      RRBLUP_automatic_candidate,
    Selection_requires_review =
      !Automatic_selection_available | Convergence_review,
    Selection_status = case_when(
      Selection_requires_review ~ "provisional_review_required",
      TRUE ~ "automatic_supported"
    ),
    Decision_code = case_when(
      Selection_requires_review ~
        "predictive_winner_provisional_review_required",
      Selected_method == "RRBLUP" &
        RRBLUP_preference_applied ~
        "rrblup_preferred_among_equivalent_models",
      Selected_method == "RRBLUP" &
        Predictive_winner == "RRBLUP" ~
        "rrblup_best_predictive_model",
      TRUE ~ "best_non_rrblup_equivalent_model"
    ),
    Operational_equivalence_rule =
      "one_se_r_and_rmse_plus_delta_dic_le_2"
  ) |>
  select(-Predictive_winner, -Convergence_review) |>
  arrange(match(Trait, traits))

selected_performance <- model_selection_support |>
  inner_join(
    selection_spec |>
      select(Trait, Selected_method),
    by = "Trait"
  ) |>
  filter(Method == Selected_method) |>
  transmute(
    Trait,
    Selected_method,
    r,
    p_value,
    RMSE,
    MAE,
    Bias,
    Slope,
    Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    Mean_fold_r,
    SD_fold_r,
    Mean_fold_RMSE,
    SD_fold_RMSE,
    Within_one_SE_r,
    Within_one_SE_RMSE,
    Predictively_competitive,
    DIC_competitive,
    Competitive_for_selection,
    Automatic_candidate,
    Competitive_set_n,
    Automatic_candidate_n,
    RRBLUP_competitive,
    RRBLUP_automatic_candidate,
    Parameters_flagged,
    Runs_with_flags,
    Minimum_ESS,
    Maximum_absolute_Geweke,
    Convergence_review
  )

model_selection <- selection_spec |>
  left_join(
    selected_performance,
    by = c("Trait", "Selected_method")
  ) |>
  left_join(predictive_winners, by = "Trait") |>
  left_join(dic_winners, by = "Trait") |>
  mutate(
    Predictive_loss = Predictive_winner_r - r,
    Predictive_winner_agrees =
      Selected_method == Predictive_winner,
    DIC_winner_agrees =
      Selected_method == DIC_winner,
    Predictive_evidence = case_when(
      r > 0 & p_value < 0.05 ~ "positive_p_lt_0.05",
      r > 0 & p_value >= 0.05 ~ "positive_p_ge_0.05",
      TRUE ~ "nonpositive_p_ge_0.05"
    )
  ) |>
  arrange(match(Trait, traits))

predictive_evidence_summary <- model_selection |>
  count(Predictive_evidence, name = "Traits") |>
  arrange(Predictive_evidence)
```

The table below makes the decision for each trait explicit.

```{r model-selection-table}
model_selection |>
  select(
    Trait, Selected_method, Predictive_winner, DIC_winner,
    r, p_value, RMSE, Mean_DIC, Delta_DIC,
    RRBLUP_preference_applied,
    Selection_status, Predictive_evidence
  ) |>
  kable(
    digits = 5,
    caption = "Selected method for each trait"
  )
```

```{r predictive-evidence-table}
kable(
  predictive_evidence_summary,
  caption = "Classification of predictive evidence across traits"
)
```

## 6. Fold stability of the selected methods

```{r selected-fold-diagnostics}
selected_fold_diagnostics <- metrics |>
  inner_join(
    selection_spec |> select(Trait, Selected_method),
    by = c("Trait" = "Trait", "Method" = "Selected_method")
  ) |>
  transmute(Trait, Fold, Method, Correlation, RMSE, MAE) |>
  arrange(match(Trait, traits), Fold)

selected_fold_diagnostics |>
  head(15) |>
  kable(
    digits = 5,
    caption = "Example of fold-level performance for selected methods"
  )
```

## 7. Computational time

```{r runtime}
runtime_summary <- metrics |>
  group_by(Method) |>
  summarise(
    Fits = n(),
    Total_seconds = sum(Elapsed_seconds),
    Mean_seconds = mean(Elapsed_seconds),
    SD_seconds = sd(Elapsed_seconds),
    Median_seconds = median(Elapsed_seconds),
    Minimum_seconds = min(Elapsed_seconds),
    Maximum_seconds = max(Elapsed_seconds),
    .groups = "drop"
  ) |>
  arrange(match(Method, methods))

kable(
  runtime_summary,
  digits = 3,
  caption = "Computational time for the seven methods"
)
```

## 8. Heritability and accuracy in the RR-BLUP/BRR analogue

Following Silva et al. [@silva2021brrguava], additive genetic variance in the
RR-BLUP/BRR analogue is obtained from the common marker variance and the
\(\sum 2pq\) denominator. Fold accuracy is defined as
\(r_f/\sqrt{H_f^2}\).

```{r rrblup-heritability}
rrblup_variance_components <- convergence |>
  filter(
    Method == "RRBLUP",
    Parameter %in% c("Residual_variance", "Marker_variance")
  ) |>
  select(Run_id, Trait, Fold, Method, Parameter, Posterior_mean) |>
  pivot_wider(names_from = Parameter, values_from = Posterior_mean)

rrblup_fold_correlations <- metrics |>
  filter(Method == "RRBLUP") |>
  select(Trait, Fold, Correlation)

rrblup_heritability_by_fold <- rrblup_variance_components |>
  left_join(rrblup_fold_correlations, by = c("Trait", "Fold")) |>
  mutate(
    Marker_variance_denominator = matrices_object$denominator,
    Additive_variance = Marker_variance * Marker_variance_denominator,
    H2 = Additive_variance / (Additive_variance + Residual_variance),
    Predictive_accuracy = Correlation / sqrt(H2)
  ) |>
  arrange(match(Trait, traits), Fold)

rrblup_heritability_summary <- rrblup_heritability_by_fold |>
  group_by(Trait) |>
  summarise(
    Mean_H2 = mean(H2),
    SD_H2 = sd(H2),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_predictive_accuracy = mean(Predictive_accuracy),
    SD_predictive_accuracy = sd(Predictive_accuracy),
    .groups = "drop"
  ) |>
  arrange(match(Trait, traits))

kable(
  rrblup_heritability_summary,
  digits = 5,
  caption = "Heritability and accuracy in the RR-BLUP/BRR analogue"
)
```

## 9. OOF genomic values and selection direction

Selection direction is applied to the genomic value only when the ranking is
constructed. The original genomic value is left unchanged.

```{r selected-oof-genomic-values}
selection_objective_map <- c(
  MMCOM = "increase",
  PRODCOM = "increase",
  PRODEF = "decrease",
  CF = "increase",
  DF = "increase",
  DCI = "decrease",
  EP = "increase",
  FFC = "increase",
  FFP = "increase",
  SS = "increase"
)

selection_objectives <- data.frame(
  Trait = traits,
  Selection_objective = unname(selection_objective_map[traits]),
  stringsAsFactors = FALSE
) |>
  mutate(
    Ranking_rule = case_when(
      Selection_objective == "increase" ~ "maximize_genomic_value",
      Selection_objective == "decrease" ~ "minimize_genomic_value",
      TRUE ~ NA_character_
    )
  )

selected_oof_genomic_values <- predictions |>
  inner_join(
    model_selection |>
      select(Trait, Selected_method, Predictive_evidence),
    by = "Trait"
  ) |>
  inner_join(selection_objectives, by = "Trait") |>
  filter(Method == Selected_method) |>
  transmute(
    IND, Trait, Fold, Method, Predictive_evidence,
    Selection_objective, Ranking_rule,
    Genomic_value = Genomic_component
  ) |>
  arrange(match(Trait, traits), IND)

kable(
  selection_objectives,
  caption = "Selection direction defined for each trait"
)
```

## 10. Proportional selection by trait

Selection is restricted to traits with positive OOF correlation and
`p < 0.05`. Within each supported trait, the 20% of individuals with the
largest oriented score are selected. For traits whose objective is to decrease,
the score is the negative genomic value.

```{r genomic-ranking}
genomic_ranking <- selected_oof_genomic_values |>
  mutate(
    Selection_score = case_when(
      Selection_objective == "increase" ~ Genomic_value,
      Selection_objective == "decrease" ~ -Genomic_value,
      TRUE ~ NA_real_
    )
  ) |>
  group_by(Trait) |>
  arrange(desc(Selection_score), IND, .by_group = TRUE) |>
  mutate(Rank = row_number()) |>
  ungroup() |>
  arrange(match(Trait, traits), Rank)

SELECTION_PROPORTION <- 0.20

supported_traits <- model_selection |>
  filter(Predictive_evidence == "positive_p_lt_0.05") |>
  pull(Trait)

selection_sizes <- genomic_ranking |>
  filter(Trait %in% supported_traits) |>
  count(Trait, name = "N_available") |>
  mutate(
    Selection_proportion = SELECTION_PROPORTION,
    N_selected = pmax(
      1L,
      as.integer(ceiling(SELECTION_PROPORTION * N_available))
    )
  ) |>
  arrange(match(Trait, traits))

selection_cutoff_summary <- genomic_ranking |>
  filter(Trait %in% supported_traits) |>
  inner_join(
    selection_sizes |>
      select(Trait, N_available, N_selected),
    by = "Trait"
  ) |>
  arrange(match(Trait, traits), Rank) |>
  group_by(Trait) |>
  mutate(Next_score = lead(Selection_score)) |>
  summarise(
    N_available = first(N_available),
    N_selected = first(N_selected),
    Score_last_selected = Selection_score[first(N_selected)],
    Score_first_not_selected = Next_score[first(N_selected)],
    Exact_boundary_tie =
      !is.na(Score_first_not_selected) &
      Score_last_selected == Score_first_not_selected,
    .groups = "drop"
  ) |>
  arrange(match(Trait, traits))

selected_individuals_by_trait <- genomic_ranking |>
  inner_join(selection_sizes, by = "Trait") |>
  filter(Rank <= N_selected) |>
  left_join(
    matrices_object$metadata |>
      select(IND, FAM),
    by = "IND"
  ) |>
  select(
    Trait, Rank, IND, FAM, Method,
    Selection_objective, Ranking_rule,
    Genomic_value, Selection_score, Predictive_evidence,
    Selection_proportion, N_available, N_selected
  ) |>
  arrange(match(Trait, traits), Rank)
```

```{r selection-size-table}
kable(
  selection_sizes,
  digits = 2,
  caption = "Number of selected individuals for each supported trait"
)
```

```{r selection-cutoff-table}
kable(
  selection_cutoff_summary,
  digits = 5,
  caption = "Scores at the proportional-selection boundary"
)
```

```{r selected-individuals-preview}
selected_individuals_by_trait |>
  group_by(Trait) |>
  slice_head(n = 5) |>
  ungroup() |>
  kable(
    digits = 5,
    caption = "Top five individuals in each supported-trait ranking"
  )
```

## 11. Recurrence across traits

Recurrence is descriptive: it counts how many supported traits place an
individual among the most favorable 20%. No weights are used and no
multi-trait selection index is constructed.

```{r multitrait-recurrence}
recurrence_counts <- selected_individuals_by_trait |>
  arrange(match(Trait, traits), Rank) |>
  group_by(IND, FAM) |>
  summarise(
    N_traits = n(),
    Traits = paste(Trait, collapse = ", "),
    Best_rank = min(Rank),
    Mean_rank_present = mean(Rank),
    .groups = "drop"
  )

multitrait_recurrence <- matrices_object$metadata |>
  select(IND, FAM) |>
  left_join(recurrence_counts, by = c("IND", "FAM")) |>
  mutate(
    N_traits = tidyr::replace_na(N_traits, 0L),
    Traits = tidyr::replace_na(Traits, ""),
    Best_rank = case_when(
      N_traits == 0L ~ NA_integer_,
      TRUE ~ Best_rank
    ),
    Mean_rank_present = case_when(
      N_traits == 0L ~ NA_real_,
      TRUE ~ Mean_rank_present
    ),
    Selection_frequency = N_traits / length(supported_traits),
    Recurrence_class = case_when(
      N_traits >= 4L ~ "high_4plus",
      N_traits == 3L ~ "intermediate_3",
      N_traits == 2L ~ "broad_2",
      N_traits == 1L ~ "specific_1",
      TRUE ~ "none_0"
    )
  ) |>
  arrange(desc(N_traits), FAM, IND)

recurrence_distribution <- multitrait_recurrence |>
  count(N_traits, Recurrence_class, name = "N_individuals") |>
  mutate(
    Proportion_of_population =
      N_individuals / nrow(matrices_object$metadata)
  ) |>
  arrange(desc(N_traits))

candidate_rank_wide <- selected_individuals_by_trait |>
  semi_join(
    multitrait_recurrence |>
      filter(N_traits >= 2L),
    by = c("IND", "FAM")
  ) |>
  select(IND, FAM, Trait, Rank) |>
  pivot_wider(
    names_from = Trait,
    values_from = Rank,
    names_glue = "{Trait}_rank"
  )

multitrait_candidates <- multitrait_recurrence |>
  filter(N_traits >= 2L) |>
  left_join(candidate_rank_wide, by = c("IND", "FAM")) |>
  arrange(desc(N_traits), FAM, IND)

kable(
  recurrence_distribution,
  digits = 4,
  caption = "Distribution of recurrence across supported traits"
)
```

## 12. Correlations among OOF genomic values

These correlations describe associations among predicted genomic values and are
not genetic-correlation estimates from a multivariate model.

```{r oof-correlations}
selected_genomic_wide <- selected_oof_genomic_values |>
  select(IND, Trait, Genomic_value) |>
  pivot_wider(names_from = Trait, values_from = Genomic_value)

selected_oof_correlations <- cor(
  as.matrix(selected_genomic_wide[, traits, drop = FALSE])
)

kable(
  round(selected_oof_correlations, 3),
  caption = "Correlations among OOF genomic values"
)
```

## 13. Saving results

```{r save-results}
write.csv(model_comparison, file.path(RESULTS_DIR, "model_comparison.csv"), row.names = FALSE)
write.csv(model_selection_support, file.path(RESULTS_DIR, "model_selection_support.csv"), row.names = FALSE)
write.csv(model_selection, file.path(RESULTS_DIR, "model_selection.csv"), row.names = FALSE)
write.csv(selection_spec, file.path(RESULTS_DIR, "selection_decision_spec.csv"), row.names = FALSE)
write.csv(predictive_evidence_summary, file.path(RESULTS_DIR, "predictive_evidence_summary.csv"), row.names = FALSE)
write.csv(convergence_by_trait_method, file.path(RESULTS_DIR, "convergence_by_trait_method.csv"), row.names = FALSE)
write.csv(selected_fold_diagnostics, file.path(RESULTS_DIR, "selected_fold_diagnostics.csv"), row.names = FALSE)
write.csv(runtime_summary, file.path(RESULTS_DIR, "runtime_summary.csv"), row.names = FALSE)
write.csv(rrblup_heritability_by_fold, file.path(RESULTS_DIR, "rrblup_heritability_by_fold.csv"), row.names = FALSE)
write.csv(rrblup_heritability_summary, file.path(RESULTS_DIR, "rrblup_heritability_summary.csv"), row.names = FALSE)
write.csv(selection_objectives, file.path(RESULTS_DIR, "selection_objectives.csv"), row.names = FALSE)
write.csv(selected_oof_genomic_values, file.path(RESULTS_DIR, "selected_oof_genomic_values.csv"), row.names = FALSE)
write.csv(genomic_ranking, file.path(RESULTS_DIR, "genomic_ranking.csv"), row.names = FALSE)
write.csv(selection_sizes, file.path(RESULTS_DIR, "selection_sizes.csv"), row.names = FALSE)
write.csv(selection_cutoff_summary, file.path(RESULTS_DIR, "selection_cutoff_summary.csv"), row.names = FALSE)
write.csv(selected_individuals_by_trait, file.path(RESULTS_DIR, "selected_individuals_by_trait.csv"), row.names = FALSE)
write.csv(multitrait_recurrence, file.path(RESULTS_DIR, "multitrait_recurrence.csv"), row.names = FALSE)
write.csv(recurrence_distribution, file.path(RESULTS_DIR, "recurrence_distribution.csv"), row.names = FALSE)
write.csv(multitrait_candidates, file.path(RESULTS_DIR, "multitrait_candidates.csv"), row.names = FALSE)
write.csv(as.data.frame(selected_oof_correlations), file.path(RESULTS_DIR, "selected_oof_correlations.csv"), row.names = TRUE)

saveRDS(
  list(
    traits = traits,
    methods = methods,
    model_comparison = model_comparison,
    model_selection_support = model_selection_support,
    selection_spec = selection_spec,
    model_selection = model_selection,
    predictive_evidence_summary = predictive_evidence_summary,
    convergence_by_trait_method = convergence_by_trait_method,
    selected_fold_diagnostics = selected_fold_diagnostics,
    runtime = runtime_summary,
    rrblup_heritability_by_fold = rrblup_heritability_by_fold,
    rrblup_heritability_summary = rrblup_heritability_summary,
    selection_objectives = selection_objectives,
    selected_oof_genomic_values = selected_oof_genomic_values,
    genomic_ranking = genomic_ranking,
    selection_proportion = SELECTION_PROPORTION,
    supported_traits = supported_traits,
    selection_sizes = selection_sizes,
    selection_cutoff_summary = selection_cutoff_summary,
    selected_individuals_by_trait = selected_individuals_by_trait,
    multitrait_recurrence = multitrait_recurrence,
    recurrence_distribution = recurrence_distribution,
    multitrait_candidates = multitrait_candidates,
    selected_oof_correlations = selected_oof_correlations
  ),
  file.path(RESULTS_DIR, "module04_results.rds")
)
```

## 14. Next step

Module 05 uses these results to construct summary tables and figures for method
comparison and genomic selection.
