---
title: "Result derivation and model selection"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

Using the predictions, metrics, and diagnostics obtained in Module 03, this
module compares the seven methods within each trait and derives the results
used for genomic selection.

The analysis calculates OOF predictive performance, DIC and relative support
measures, the selected method for each trait, fold stability, computational
time, heritability and accuracy in the RR-BLUP/BRR analogue, OOF genomic
values, oriented rankings, proportional selection, recurrence across traits,
and correlations among genomic values.

The breeding direction is defined before ranking: increase MMCOM, PRODCOM, CF,
DF, EP, FFC, FFP, and SS; decrease PRODEF and DCI.

## 2. Packages and model-fitting results

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(knitr)
  library(here)
})
```

The next step loads the input objects and organizes the structures required for the analysis.

```{r inputs}
MATRICES_FILE <- here("output", "matrices", "matrices_object.rds")
MODELS_FILE <- here("output", "models", "models_object.rds")
RESULTS_DIR <- here("output", "results", "m04")

dir.create(RESULTS_DIR, recursive = TRUE, showWarnings = FALSE)

matrices_object <- readRDS(MATRICES_FILE)
models_object <- readRDS(MODELS_FILE)

metrics <- models_object$metrics_by_fold
predictions <- models_object$predictions
convergence <- models_object$convergence_diagnostics

traits <- colnames(matrices_object$phenotypes)
methods <- models_object$settings$methods
```

The dimensions of the three input tables show the coverage of the fitted
analysis before the following derivations.

```{r input-summary}
# Summarize the dimensions and categories of the preprocessing object received by the module.
data.frame(
  Output = c(
    "Trait × fold × method metrics",
    "OOF predictions",
    "MCMC diagnostics"
  ),
  Rows = c(
    nrow(metrics),
    nrow(predictions),
    nrow(convergence)
  ),
  # Count how many traits are represented in each input table.
  Traits = c(
    n_distinct(metrics$Trait),
    n_distinct(predictions$Trait),
    n_distinct(convergence$Trait)
  ),
  Methods = c(
    n_distinct(metrics$Method),
    n_distinct(predictions$Method),
    n_distinct(convergence$Method)
  )
) |>
  kable(caption = "Results received from model fitting")
```

## 3. OOF predictive performance

The OOF estimates inherit the family-stratified design of Modules 02 and
03. They quantify prediction within the represented breeding population and do
not estimate generalization to entirely new families.

The main metrics are recalculated using the 150 out-of-fold predictions for
each trait × method combination. Each individual contributes one OOF
prediction for every method.

```{r pooled-oof-metrics}
# Combine OOF predictions and compute aggregate predictive metrics by trait and method.
pooled_metrics <- predictions |>
  group_by(Trait, Method) |>
  summarise(
    N = n(),
    r = cor(Observed, Predicted),
    RMSE = sqrt(mean((Observed - Predicted)^2)),
    MAE = mean(abs(Observed - Predicted)),
    Bias = mean(Predicted - Observed),
    Slope = cov(Observed, Predicted) / var(Predicted),
    .groups = "drop"
  ) |>
  mutate(
    t_value = r * sqrt((N - 2) / (1 - r^2)),
    p_value = 2 * pt(abs(t_value), df = N - 2, lower.tail = FALSE)
  ) |>
  select(-t_value)

# Summarize the results in `fold_summary`.
fold_summary <- metrics |>
  group_by(Trait, Method) |>
  summarise(
    Mean_DIC = mean(DIC),
    SD_DIC = sd(DIC),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_fold_RMSE = mean(RMSE),
    SD_fold_RMSE = sd(RMSE),
    Mean_fold_MAE = mean(MAE),
    SD_fold_MAE = sd(MAE),
    .groups = "drop"
  )
```

Pearson-correlation p-values are retained as nominal descriptive measures
because training sets overlap across folds. They are reported for context and
are not used to determine eligibility for genomic selection.

```{r pooled-oof-preview}
pooled_metrics |>
  arrange(match(Trait, traits), desc(r), RMSE) |>
  head(14) |>
  kable(
    digits = 5,
    caption = "Example of OOF predictive performance"
  )
```

## 4. DIC and relative support for the methods

Following the logic used in the reference study [@silva2021brrguava], the
difference from the smallest mean DIC within each trait is converted into
relative weights:

\[
\Delta_j=DIC_j-\min(DIC),
\]

\[
Wprob_j=\frac{\exp(-\Delta_j/2)}{\sum_k\exp(-\Delta_k/2)},
\qquad
ER_j=\frac{Wprob_{\max}}{Wprob_j}.
\]

`Wprob` is used here as a DIC-derived relative-support weight. It is not
interpreted as a calibrated posterior probability of a model.

```{r model-comparison}
# Combine fold performance and DIC to build the quantities used for model comparison.
model_comparison <- pooled_metrics |>
  left_join(fold_summary, by = c("Trait", "Method")) |>
  group_by(Trait) |>
  mutate(
    Delta_DIC = Mean_DIC - min(Mean_DIC),
    log_weight = -0.5 * Delta_DIC,
    stable_weight = exp(log_weight - max(log_weight)),
    Wprob = stable_weight / sum(stable_weight),
    # Compute the evidence ratio relative to the method with the largest Wprob in each trait.
    ER = max(Wprob) / Wprob
  ) |>
  ungroup() |>
  select(
    Trait, Method, Mean_DIC, SD_DIC, Delta_DIC, Wprob, ER,
    N, r, p_value, RMSE, MAE, Bias, Slope,
    Mean_fold_r, SD_fold_r, Mean_fold_RMSE, SD_fold_RMSE,
    Mean_fold_MAE, SD_fold_MAE
  ) |>
  arrange(match(Trait, traits), Delta_DIC)
```

The next step organizes a tabular summary of the objects and results needed to interpret this section.

```{r model-comparison-preview}
model_comparison |>
  select(
    Trait, Method, Mean_DIC, Delta_DIC,
    Wprob, ER, r, RMSE, MAE
  ) |>
  head(21) |>
  kable(
    digits = 5,
    caption = "Example of the comparison among methods"
  )
```

## 5. Selecting a method for each trait

OOF predictive ability is the primary criterion. DIC provides complementary
information about fit and complexity, while variation across folds and MCMC
diagnostics help evaluate the stability of the decision.

A method is considered predictively competitive when its mean fold correlation
is within one standard error below the largest mean and its mean RMSE is within
one standard error above the smallest mean RMSE. DIC fit is considered similar
when `Delta_DIC <= 2`. These thresholds define an operational competitive set,
not a formal equivalence test.

When RR-BLUP/BRR belongs to this competitive set and has no convergence flags,
it is preferred as the homogeneous-shrinkage benchmark with the simpler
parameterization. Otherwise, the best competitive candidate without flags is
used; when no rule-based candidate is available, the predictive winner is
retained as a provisional choice for review.

### 5.1 Predictive and DIC winners

First, the method with the best OOF predictive performance and the method with the lowest mean DIC are identified separately for each trait.

The `Delta_DIC <= 2` cutoff is the operational rule used to regard fits as having similar DIC; it is not a formal test. The number of folds is not redefined in this module: `N_FOLDS` is recovered from the Module 02 object and is used only to calculate the descriptive one-standard-error tolerances.

```{r selection-winners}
# Use Delta_DIC <= 2 as the operational cutoff for similar fit.
DIC_COMPETITIVE_THRESHOLD <- 2
# Recover the five Module 02 folds for the descriptive one-standard-error tolerances.
N_FOLDS <- matrices_object$settings$n_folds

# Best combination of correlation, RMSE, MAE, and DIC for each trait.
predictive_winners <- model_comparison |>
  group_by(Trait) |>
  arrange(desc(r), RMSE, MAE, Mean_DIC, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    Predictive_winner = Method,
    Predictive_winner_r = r,
    Predictive_winner_p_value = p_value,
    Predictive_winner_RMSE = RMSE,
    Predictive_winner_MAE = MAE
  )

# Method with the smallest mean DIC for each trait.
dic_winners <- model_comparison |>
  group_by(Trait) |>
  arrange(Mean_DIC, desc(r), RMSE, MAE, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    DIC_winner = Method,
    Minimum_DIC = Mean_DIC,
    DIC_winner_Wprob = Wprob
  )
```

The next step derives and presents the corresponding selection or prioritization rule from the calculated results.

```{r selection-winners-table}
predictive_winners |>
  left_join(dic_winners, by = "Trait") |>
  select(
    Trait, Predictive_winner, DIC_winner,
    Predictive_winner_r, Predictive_winner_RMSE,
    Minimum_DIC
  ) |>
  kable(
    digits = 5,
    caption = "Predictive and DIC winners for each trait"
  )
```

### 5.2 MCMC diagnostics by trait and method

Chain diagnostics are summarized before defining the set of methods eligible
for selection.

```{r selection-convergence-summary}
convergence_by_trait_method <- convergence |>
  group_by(Trait, Method) |>
  summarise(
    Runs = n_distinct(Run_id),
    Parameters_reviewed = n(),
    Parameters_flagged = sum(Review_flag),
    Runs_with_flags = n_distinct(Run_id[Review_flag]),
    Minimum_ESS = min(Effective_sample_size, na.rm = TRUE),
    Maximum_absolute_Geweke = max(abs(Geweke_Z), na.rm = TRUE),
    .groups = "drop"
  )
```

The next step derives and presents the corresponding selection or prioritization rule from the calculated results.

```{r selection-convergence-table}
convergence_by_trait_method |>
  arrange(match(Trait, traits), match(Method, methods)) |>
  head(21) |>
  kable(
    digits = 4,
    caption = "Example of MCMC diagnostics summarized by trait and method"
  )
```

### 5.3 Competitive set

The one-standard-error rule is applied to mean fold correlation and RMSE. The
`Delta_DIC <= 2` threshold and MCMC diagnostics are then used to determine the
candidates eligible under the rule. Because the training sets overlap across
folds, these standard errors are descriptive tolerances and are not interpreted
as independent-sample inferential standard errors.

```{r selection-competitive-set}
# Combine performance, DIC, and MCMC diagnostics to define the competitive set for each trait.
model_selection_support <- model_comparison |>
  left_join(
    convergence_by_trait_method,
    by = c("Trait", "Method")
  ) |>
  left_join(
    predictive_winners |>
      select(Trait, Predictive_winner),
    by = "Trait"
  ) |>
  group_by(Trait) |>
  mutate(
    # References for the one-standard-error rule.
    Best_mean_fold_r = max(Mean_fold_r, na.rm = TRUE),
    Best_mean_fold_r_SE =
      SD_fold_r[which.max(Mean_fold_r)] / sqrt(N_FOLDS),
    Best_mean_fold_RMSE = min(Mean_fold_RMSE, na.rm = TRUE),
    Best_mean_fold_RMSE_SE =
      SD_fold_RMSE[which.min(Mean_fold_RMSE)] / sqrt(N_FOLDS),

    # Predictive competitiveness.
    Within_one_SE_r =
      Mean_fold_r >= Best_mean_fold_r - Best_mean_fold_r_SE,
    Within_one_SE_RMSE =
      Mean_fold_RMSE <= Best_mean_fold_RMSE + Best_mean_fold_RMSE_SE,
    Predictively_competitive =
      Within_one_SE_r & Within_one_SE_RMSE,

    # Complementary support from DIC and convergence.
    DIC_competitive_threshold = DIC_COMPETITIVE_THRESHOLD,
    DIC_competitive = Delta_DIC <= DIC_COMPETITIVE_THRESHOLD,
    Competitive_for_selection =
      Predictively_competitive & DIC_competitive,
    Convergence_review = Parameters_flagged > 0,
    Automatic_candidate =
      Competitive_for_selection & !Convergence_review,

    # Candidate-set summary within each trait.
    Competitive_set_n = sum(Competitive_for_selection),
    Automatic_candidate_n = sum(Automatic_candidate),
    RRBLUP_competitive = any(
      Method == "RRBLUP" & Competitive_for_selection
    ),
    RRBLUP_automatic_candidate = any(
      Method == "RRBLUP" & Automatic_candidate
    ),
    Automatic_selection_available = any(Automatic_candidate),

    # RR-BLUP receives priority only within the automatic-candidate set.
    Selection_priority = case_when(
      Automatic_candidate & Method == "RRBLUP" ~ 1L,
      Automatic_candidate ~ 2L,
      TRUE ~ 3L
    )
  ) |>
  ungroup() |>
  arrange(
    match(Trait, traits),
    Selection_priority,
    desc(r),
    RMSE,
    MAE,
    Mean_DIC
  )
```

The next step derives and presents the corresponding selection or prioritization rule from the calculated results.

```{r selection-competitive-table}
model_selection_support |>
  select(
    Trait, Method, Predictively_competitive,
    DIC_competitive, Convergence_review,
    Automatic_candidate, r, RMSE, Delta_DIC
  ) |>
  head(21) |>
  kable(
    digits = 5,
    caption = "Example of the competitive-set definition"
  )
```

### 5.4 Method choice

RR-BLUP receives priority among automatic candidates. If no automatic candidate
is available, the predictive winner is retained as a provisional choice.

```{r selection-choice}
# Best automatic candidate for each trait.
automatic_selections <- model_selection_support |>
  filter(Automatic_candidate) |>
  group_by(Trait) |>
  arrange(
    Selection_priority,
    desc(r),
    RMSE,
    MAE,
    Mean_DIC,
    .by_group = TRUE
  ) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    Automatic_selected_method = Method
  )

# Final choice before attaching the selected method's metrics.
# Apply RR-BLUP priority within the automatic set and define the selected method.
selection_spec <- predictive_winners |>
  select(Trait, Predictive_winner) |>
  left_join(automatic_selections, by = "Trait") |>
  mutate(
    Selected_method = coalesce(
      Automatic_selected_method,
      Predictive_winner
    )
  ) |>
  select(Trait, Selected_method, Predictive_winner) |>
  left_join(
    model_selection_support |>
      # Keep and organize the variables used in the result.
      select(
        Trait,
        Method,
        Automatic_selection_available,
        Automatic_candidate_n,
        Competitive_set_n,
        RRBLUP_competitive,
        RRBLUP_automatic_candidate,
        Convergence_review
      ),
    # Match the selected method to the support table to append the indicators used in the decision.
    by = c("Trait", "Selected_method" = "Method")
  ) |>
  mutate(
    RRBLUP_preference_applied =
      Selected_method == "RRBLUP" &
      Predictive_winner != "RRBLUP" &
      RRBLUP_automatic_candidate,
    Selection_requires_review =
      !Automatic_selection_available | Convergence_review,
    Selection_status = case_when(
      Selection_requires_review ~ "provisional_review_required",
      TRUE ~ "rule_based_selected"
    ),
    Decision_code = case_when(
      Selection_requires_review ~
        "predictive_winner_provisional_review_required",
      Selected_method == "RRBLUP" &
        RRBLUP_preference_applied ~
        "rrblup_preferred_among_competitive_models",
      Selected_method == "RRBLUP" &
        Predictive_winner == "RRBLUP" ~
        "rrblup_best_predictive_model",
      TRUE ~ "best_non_rrblup_competitive_model"
    ),
    Operational_competitive_set_rule =
      "one_se_r_and_rmse_plus_delta_dic_le_2"
  ) |>
  select(-Predictive_winner, -Convergence_review) |>
  arrange(match(Trait, traits))
```

The next step derives and presents the corresponding selection or prioritization rule from the calculated results.

```{r selection-choice-table}
selection_spec |>
  select(
    Trait, Selected_method, Selection_status,
    RRBLUP_preference_applied,
    Competitive_set_n, Automatic_candidate_n
  ) |>
  kable(caption = "Method chosen after applying the hierarchical rule")
```

### 5.5 Performance of the selected methods

Metrics and diagnostics for the method actually selected are then extracted
from the support table.

```{r selected-performance}
# Extract predictive metrics, DIC, and diagnostics for the selected method in each trait.
selected_performance <- model_selection_support |>
  inner_join(
    selection_spec |>
      select(Trait, Selected_method),
    by = "Trait"
  ) |>
  # Restrict the data to observations relevant at this stage.
  filter(Method == Selected_method) |>
  # Build the columns that form the output of this block.
  transmute(
    Trait,
    Selected_method,
    r,
    p_value,
    RMSE,
    MAE,
    Bias,
    Slope,
    Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    Mean_fold_r,
    SD_fold_r,
    Mean_fold_RMSE,
    SD_fold_RMSE,
    Within_one_SE_r,
    Within_one_SE_RMSE,
    Predictively_competitive,
    DIC_competitive,
    Competitive_for_selection,
    Automatic_candidate,
    Competitive_set_n,
    Automatic_candidate_n,
    RRBLUP_competitive,
    RRBLUP_automatic_candidate,
    Parameters_flagged,
    Runs_with_flags,
    Minimum_ESS,
    Maximum_absolute_Geweke,
    Convergence_review
  )
```

The next step derives and presents the corresponding selection or prioritization rule from the calculated results.

```{r selected-performance-table}
selected_performance |>
  select(
    Trait, Selected_method, r, p_value, RMSE,
    Mean_DIC, Delta_DIC, Parameters_flagged
  ) |>
  kable(
    digits = 5,
    caption = "Performance and diagnostics of the selected methods"
  )
```

### 5.6 Final selection result and predictive evidence

Finally, the selected method, predictive and DIC winners, OOF direction, and a
nominal signal class are brought together for each trait. The nominal class is shown only to make weak positive correlations explicit; it is not used as a selection filter. The $r > 0$ boundary separates only positive from nonpositive predictive direction; it is neither a significance test nor a minimum predictive-utility threshold.

```{r final-model-selection}
# Combine the selected method, predictive and DIC winners, and OOF evidence in the final trait-level table.
model_selection <- selection_spec |>
  left_join(
    selected_performance,
    # Join selected-method performance using trait and method as keys.
    by = c("Trait", "Selected_method")
  ) |>
  left_join(predictive_winners, by = "Trait") |>
  left_join(dic_winners, by = "Trait") |>
  mutate(
    Predictive_loss = Predictive_winner_r - r,
    Predictive_winner_agrees =
      Selected_method == Predictive_winner,
    DIC_winner_agrees =
      Selected_method == DIC_winner,
    # Use zero only as a direction boundary; r > 0 does not imply significance or minimum utility.
    OOF_predictive_direction = if_else(
      r > 0,
      "positive",
      "nonpositive"
    ),
    # Keep p < 0.05 only as a descriptive nominal class, without filtering rankings.
    OOF_nominal_signal = case_when(
      r > 0 & p_value < 0.05 ~ "positive_nominal_p_lt_0.05",
      r > 0 ~ "positive_nominal_p_ge_0.05",
      TRUE ~ "nonpositive"
    )
  ) |>
  arrange(match(Trait, traits))

# Count how many traits have positive or nonpositive OOF correlation.
oof_direction_summary <- model_selection |>
  count(OOF_predictive_direction, name = "Traits") |>
  arrange(OOF_predictive_direction)
```

The next step derives and presents the corresponding selection or prioritization rule from the calculated results.

```{r model-selection-table}
model_selection |>
  select(
    Trait, Selected_method, Predictive_winner, DIC_winner,
    r, p_value, RMSE, Mean_DIC, Delta_DIC,
    RRBLUP_preference_applied,
    Selection_status, OOF_predictive_direction, OOF_nominal_signal
  ) |>
  kable(
    digits = 5,
    caption = "Selected method for each trait"
  )
```

The next step organizes a tabular summary of the objects and results needed to interpret this section.

```{r oof-direction-table}
kable(
  oof_direction_summary,
  caption = "OOF predictive direction of the traits"
)
```

## 6. Stability across folds for the selected methods

The next step derives and presents the corresponding selection or prioritization rule from the calculated results.

```{r selected-fold-diagnostics}
selected_fold_diagnostics <- metrics |>
  inner_join(
    selection_spec |> select(Trait, Selected_method),
    by = c("Trait" = "Trait", "Method" = "Selected_method")
  ) |>
  transmute(Trait, Fold, Method, Correlation, RMSE, MAE) |>
  arrange(match(Trait, traits), Fold)

selected_fold_diagnostics |>
  head(15) |>
  kable(
    digits = 5,
    caption = "Example of fold performance for the selected methods"
  )
```

## 7. Computational time

The next step summarizes execution times to compare the computational cost of the methods.

```{r runtime}
# Summarize the computational cost of the evaluated methods.
runtime_summary <- metrics |>
  group_by(Method) |>
  summarise(
    Fits = n(),
    Total_seconds = sum(Elapsed_seconds),
    Mean_seconds = mean(Elapsed_seconds),
    SD_seconds = sd(Elapsed_seconds),
    Median_seconds = median(Elapsed_seconds),
    Minimum_seconds = min(Elapsed_seconds),
    Maximum_seconds = max(Elapsed_seconds),
    .groups = "drop"
  ) |>
  arrange(match(Method, methods))

# Format the table for display in the tutorial.
kable(
  runtime_summary,
  digits = 3,
  caption = "Computational time of the seven methods"
)
```

## 8. Marker-based heritability analogue and adjusted accuracy in RR-BLUP/BRR

Following Silva et al. [@silva2021brrguava], a marker-based additive-variance
analogue is obtained from the common marker variance and the \(\sum 2pq\)
denominator. Fold-level adjusted accuracy is defined as
\(r_f/\sqrt{H_f^2}\). The resulting \(H^2\) is reported as a model-based
heritability analogue rather than as an independent classical estimate of
narrow-sense heritability.

```{r rrblup-heritability}
# Compute and organize the RR-BLUP/BRR model-based heritability summary.
rrblup_variance_components <- convergence |>
  filter(
    Method == "RRBLUP",
    Parameter %in% c("Residual_variance", "Marker_variance")
  ) |>
  select(Run_id, Trait, Fold, Method, Parameter, Posterior_mean) |>
  pivot_wider(names_from = Parameter, values_from = Posterior_mean)

rrblup_fold_correlations <- metrics |>
  filter(Method == "RRBLUP") |>
  select(Trait, Fold, Correlation)

# Organize the cross-validation and fitting structure.
rrblup_heritability_by_fold <- rrblup_variance_components |>
  left_join(rrblup_fold_correlations, by = c("Trait", "Fold")) |>
  mutate(
    Marker_variance_denominator = matrices_object$denominator,
    Additive_variance = Marker_variance * Marker_variance_denominator,
    H2 = Additive_variance / (Additive_variance + Residual_variance),
    Predictive_accuracy = Correlation / sqrt(H2)
  ) |>
  arrange(match(Trait, traits), Fold)

# Summarize the RR-BLUP/BRR model-based heritability analogue and adjusted accuracy across folds.
rrblup_heritability_summary <- rrblup_heritability_by_fold |>
  group_by(Trait) |>
  summarise(
    Mean_H2 = mean(H2),
    SD_H2 = sd(H2),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_predictive_accuracy = mean(Predictive_accuracy),
    SD_predictive_accuracy = sd(Predictive_accuracy),
    .groups = "drop"
  ) |>
  arrange(match(Trait, traits))

kable(
  rrblup_heritability_summary,
  digits = 5,
  caption = "Marker-based heritability analogue and adjusted predictive accuracy in RR-BLUP/BRR"
)
```

## 9. OOF genomic values and selection direction

Each individual's genomic value is cross-fitted: it comes from the fold in
which that individual's phenotype was masked. These OOF genomic values provide
a validation-consistent ranking because an individual's own phenotype does not
enter its score. They are not final breeding values from a model refitted to
all 150 phenotypes.

Selection direction is applied to the genomic value only when the ranking is
constructed. The original genomic value remains unchanged.

```{r selected-oof-genomic-values}
# Define the desired selection direction and link OOF genomic values to the selected method for each trait.
selection_objective_map <- c(
  MMCOM = "increase",
  PRODCOM = "increase",
  PRODEF = "decrease",
  CF = "increase",
  DF = "increase",
  DCI = "decrease",
  EP = "increase",
  FFC = "increase",
  FFP = "increase",
  SS = "increase"
)

# Organize the desired selection direction and ranking rule for each trait.
selection_objectives <- data.frame(
  Trait = traits,
  Selection_objective = unname(selection_objective_map[traits]),
  stringsAsFactors = FALSE
) |>
  mutate(
    Ranking_rule = case_when(
      Selection_objective == "increase" ~ "maximize_genomic_value",
      Selection_objective == "decrease" ~ "minimize_genomic_value",
      TRUE ~ NA_character_
    )
  )

# Organize the estimated genomic component for each individual.
selected_oof_genomic_values <- predictions |>
  inner_join(
    model_selection |>
      select(
        Trait, Selected_method, OOF_predictive_direction,
        OOF_nominal_signal, r, p_value
      ),
    by = "Trait"
  ) |>
  inner_join(selection_objectives, by = "Trait") |>
  filter(Method == Selected_method) |>
  transmute(
    IND, Trait, Fold, Method, OOF_predictive_direction,
    OOF_nominal_signal, OOF_r = r, OOF_nominal_p_value = p_value,
    Selection_objective, Ranking_rule,
    Genomic_value = Genomic_component
  ) |>
  arrange(match(Trait, traits), IND)

# Format the table for display in the tutorial.
kable(
  selection_objectives,
  caption = "Selection direction defined for each trait"
)
```

## 10. Exploratory proportional candidate ranking by trait

Candidate rankings are shown for traits whose selected method has positive OOF
correlation. Nominal p-values remain descriptive and do not determine whether a
ranking is produced. Positive correlations with nominal `p >= 0.05` are kept
explicitly as weak/exploratory signals and should not be interpreted as
demonstrated predictive utility. Within each positive-OOF trait, the 20% of
individuals with the largest oriented score are listed; this 20% is a
configurable tutorial rule, not an estimated optimal selection intensity. For
traits whose objective is to decrease, the score is the negative genomic value.

```{r genomic-ranking}
# Orient genomic values by the selection objective and rank individuals within each trait.
genomic_ranking <- selected_oof_genomic_values |>
  mutate(
    Selection_score = case_when(
      Selection_objective == "increase" ~ Genomic_value,
      Selection_objective == "decrease" ~ -Genomic_value,
      TRUE ~ NA_real_
    )
  ) |>
  group_by(Trait) |>
  arrange(desc(Selection_score), IND, .by_group = TRUE) |>
  mutate(Rank = row_number()) |>
  ungroup() |>
  arrange(match(Trait, traits), Rank)

# Define the candidate proportion displayed in each exploratory ranking.
SELECTION_PROPORTION <- 0.20

positive_oof_traits <- model_selection |>
  filter(OOF_predictive_direction == "positive") |>
  pull(Trait)

# Compute how many individuals compose the prioritized proportion in each positive-OOF trait.
selection_sizes <- genomic_ranking |>
  filter(Trait %in% positive_oof_traits) |>
  count(Trait, name = "N_available") |>
  mutate(
    Selection_proportion = SELECTION_PROPORTION,
    N_selected = pmax(
      1L,
      as.integer(ceiling(SELECTION_PROPORTION * N_available))
    )
  ) |>
  arrange(match(Trait, traits))

# Record the cutoff score and genomic-value interval corresponding to each ranking.
selection_cutoff_summary <- genomic_ranking |>
  filter(Trait %in% positive_oof_traits) |>
  inner_join(
    selection_sizes |>
      select(Trait, N_available, N_selected),
    by = "Trait"
  ) |>
  arrange(match(Trait, traits), Rank) |>
  group_by(Trait) |>
  mutate(Next_score = lead(Selection_score)) |>
  summarise(
    N_available = first(N_available),
    N_selected = first(N_selected),
    Score_last_selected = Selection_score[first(N_selected)],
    Score_first_not_selected = Next_score[first(N_selected)],
    Exact_boundary_tie =
      !is.na(Score_first_not_selected) &
      Score_last_selected == Score_first_not_selected,
    .groups = "drop"
  ) |>
  arrange(match(Trait, traits))

# Retain candidates included in the defined ranking proportion for each positive-OOF trait.
selected_individuals_by_trait <- genomic_ranking |>
  inner_join(selection_sizes, by = "Trait") |>
  filter(Rank <= N_selected) |>
  left_join(
    matrices_object$metadata |>
      select(IND, FAM),
    by = "IND"
  ) |>
  select(
    Trait, Rank, IND, FAM, Method,
    Selection_objective, Ranking_rule,
    Genomic_value, Selection_score, OOF_predictive_direction,
    OOF_nominal_signal, OOF_r, OOF_nominal_p_value,
    Selection_proportion, N_available, N_selected
  ) |>
  arrange(match(Trait, traits), Rank)
```

The next step derives and presents the corresponding selection or prioritization rule from the calculated results.

```{r selection-size-table}
kable(
  selection_sizes,
  digits = 2,
  caption = "Exploratory candidate-ranking size for each positive-OOF trait"
)
```

The next step derives and presents the corresponding selection or prioritization rule from the calculated results.

```{r selection-cutoff-table}
kable(
  selection_cutoff_summary,
  digits = 5,
  caption = "Scores at the proportional-selection boundary"
)
```

The next step derives and presents the corresponding selection or prioritization rule from the calculated results.

```{r selected-individuals-preview}
selected_individuals_by_trait |>
  group_by(Trait) |>
  slice_head(n = 5) |>
  ungroup() |>
  kable(
    digits = 5,
    caption = "Top five candidates in each positive-OOF trait ranking"
  )
```

## 11. Recurrence across traits

Recurrence is descriptive: among traits with positive OOF correlation, it counts how many times each individual appears among the most favorable 20%. Weak positive OOF signals remain exploratory. No weights are used and no multi-trait selection index is constructed.

Recurrence classes are descriptive conventions: high for four or more traits, intermediate for three, broad for two, specific for one, and none for zero. The `multitrait_candidates` object retains individuals recurring in at least two traits so potentially useful information is not discarded before downstream presentation. These cutoffs organize the summary and are not statistical-significance thresholds.

```{r multitrait-recurrence}
# Count individual recurrence across exploratory rankings for traits with positive OOF correlation.
recurrence_counts <- selected_individuals_by_trait |>
  arrange(match(Trait, traits), Rank) |>
  group_by(IND, FAM) |>
  summarise(
    N_traits = n(),
    Traits = paste(Trait, collapse = ", "),
    Best_rank = min(Rank),
    Mean_rank_present = mean(Rank),
    .groups = "drop"
  )

# Join recurrence counts to metadata and compute recurrence frequency and class.
multitrait_recurrence <- matrices_object$metadata |>
  select(IND, FAM) |>
  left_join(recurrence_counts, by = c("IND", "FAM")) |>
  mutate(
    N_traits = tidyr::replace_na(N_traits, 0L),
    Traits = tidyr::replace_na(Traits, ""),
    Best_rank = case_when(
      N_traits == 0L ~ NA_integer_,
      TRUE ~ Best_rank
    ),
    Mean_rank_present = case_when(
      N_traits == 0L ~ NA_real_,
      TRUE ~ Mean_rank_present
    ),
    Selection_frequency = N_traits / length(positive_oof_traits),
    # Classify recurrence by number of traits; these cutoffs are descriptive.
    Recurrence_class = case_when(
      N_traits >= 4L ~ "high_4plus",
      N_traits == 3L ~ "intermediate_3",
      N_traits == 2L ~ "broad_2",
      N_traits == 1L ~ "specific_1",
      TRUE ~ "none_0"
    )
  ) |>
  arrange(desc(N_traits), FAM, IND)

# Summarize how many individuals belong to each recurrence class.
recurrence_distribution <- multitrait_recurrence |>
  count(N_traits, Recurrence_class, name = "N_individuals") |>
  mutate(
    Proportion_of_population =
      N_individuals / nrow(matrices_object$metadata)
  ) |>
  arrange(desc(N_traits))

# Reshape trait-specific ranks to one row per recurrent candidate.
candidate_rank_wide <- selected_individuals_by_trait |>
  semi_join(
    multitrait_recurrence |>
      filter(N_traits >= 2L),
    by = c("IND", "FAM")
  ) |>
  select(IND, FAM, Trait, Rank) |>
  pivot_wider(
    names_from = Trait,
    values_from = Rank,
    names_glue = "{Trait}_rank"
  )

# Retain candidates recurring in at least two traits and attach their trait-specific ranks.
multitrait_candidates <- multitrait_recurrence |>
  # Retain in the complete object candidates present in at least two traits.
  filter(N_traits >= 2L) |>
  left_join(candidate_rank_wide, by = c("IND", "FAM")) |>
  arrange(desc(N_traits), FAM, IND)

kable(
  recurrence_distribution,
  digits = 4,
  caption = "Distribution of recurrence across traits with positive OOF correlation"
)
```

## 12. Correlations among OOF genomic values

These correlations describe associations among predicted genomic values and are
not genetic-correlation estimates from a multivariate model.

```{r oof-correlations}
selected_genomic_wide <- selected_oof_genomic_values |>
  select(IND, Trait, Genomic_value) |>
  pivot_wider(names_from = Trait, values_from = Genomic_value)

selected_oof_correlations <- cor(
  as.matrix(selected_genomic_wide[, traits, drop = FALSE])
)

kable(
  round(selected_oof_correlations, 3),
  caption = "Correlations among OOF genomic values"
)
```

## 13. Saving the results

The next step organizes and saves the generated outputs for use in the downstream modules.

```{r save-results}
# Save the tables derived by the module before assembling the consolidated RDS object.
write.csv(model_comparison, file.path(RESULTS_DIR, "model_comparison.csv"), row.names = FALSE)
write.csv(model_selection_support, file.path(RESULTS_DIR, "model_selection_support.csv"), row.names = FALSE)
write.csv(model_selection, file.path(RESULTS_DIR, "model_selection.csv"), row.names = FALSE)
write.csv(selection_spec, file.path(RESULTS_DIR, "selection_decision_spec.csv"), row.names = FALSE)
write.csv(oof_direction_summary, file.path(RESULTS_DIR, "oof_direction_summary.csv"), row.names = FALSE)
write.csv(convergence_by_trait_method, file.path(RESULTS_DIR, "convergence_by_trait_method.csv"), row.names = FALSE)
write.csv(selected_fold_diagnostics, file.path(RESULTS_DIR, "selected_fold_diagnostics.csv"), row.names = FALSE)
write.csv(runtime_summary, file.path(RESULTS_DIR, "runtime_summary.csv"), row.names = FALSE)
write.csv(rrblup_heritability_by_fold, file.path(RESULTS_DIR, "rrblup_heritability_by_fold.csv"), row.names = FALSE)
write.csv(rrblup_heritability_summary, file.path(RESULTS_DIR, "rrblup_heritability_summary.csv"), row.names = FALSE)
write.csv(selection_objectives, file.path(RESULTS_DIR, "selection_objectives.csv"), row.names = FALSE)
write.csv(selected_oof_genomic_values, file.path(RESULTS_DIR, "selected_oof_genomic_values.csv"), row.names = FALSE)
# Save the consolidated output for downstream use.
write.csv(genomic_ranking, file.path(RESULTS_DIR, "genomic_ranking.csv"), row.names = FALSE)
write.csv(selection_sizes, file.path(RESULTS_DIR, "selection_sizes.csv"), row.names = FALSE)
write.csv(selection_cutoff_summary, file.path(RESULTS_DIR, "selection_cutoff_summary.csv"), row.names = FALSE)
write.csv(selected_individuals_by_trait, file.path(RESULTS_DIR, "selected_individuals_by_trait.csv"), row.names = FALSE)
write.csv(multitrait_recurrence, file.path(RESULTS_DIR, "multitrait_recurrence.csv"), row.names = FALSE)
write.csv(recurrence_distribution, file.path(RESULTS_DIR, "recurrence_distribution.csv"), row.names = FALSE)
write.csv(multitrait_candidates, file.path(RESULTS_DIR, "multitrait_candidates.csv"), row.names = FALSE)
# Save the consolidated output for downstream use.
write.csv(as.data.frame(selected_oof_correlations), file.path(RESULTS_DIR, "selected_oof_correlations.csv"), row.names = TRUE)

# Save the consolidated output for downstream use.
saveRDS(
  list(
    traits = traits,
    methods = methods,
    model_comparison = model_comparison,
    model_selection_support = model_selection_support,
    selection_spec = selection_spec,
    model_selection = model_selection,
    oof_direction_summary = oof_direction_summary,
    convergence_by_trait_method = convergence_by_trait_method,
    selected_fold_diagnostics = selected_fold_diagnostics,
    runtime = runtime_summary,
    rrblup_heritability_by_fold = rrblup_heritability_by_fold,
    rrblup_heritability_summary = rrblup_heritability_summary,
    selection_objectives = selection_objectives,
    selected_oof_genomic_values = selected_oof_genomic_values,
    genomic_ranking = genomic_ranking,
    selection_proportion = SELECTION_PROPORTION,
    positive_oof_traits = positive_oof_traits,
    selection_sizes = selection_sizes,
    selection_cutoff_summary = selection_cutoff_summary,
    selected_individuals_by_trait = selected_individuals_by_trait,
    multitrait_recurrence = multitrait_recurrence,
    recurrence_distribution = recurrence_distribution,
    multitrait_candidates = multitrait_candidates,
    selected_oof_correlations = selected_oof_correlations
  ),
  file.path(RESULTS_DIR, "module04_results.rds")
)
```

## 14. Next step

Module 05 uses these results to construct summary tables and figures for method
comparison and genomic selection.
