---
title: "Model comparison, genomic selection, and results"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

This module brings together the results produced in Module 03 and organizes them into a sequence focused on interpretation and genomic selection:

**compare methods → choose a method for each trait → interpret genomic values → examine relationships among traits → define eligibility → select individuals → evaluate recurrence.**

Method choice is performed separately for each trait and prioritizes OOF predictive performance. DIC remains complementary information. MCMC diagnostics summarize overall chain behavior but do not act as a selection filter.

After method choice, OOF GEBVs are obtained for all traits and are also examined jointly through their correlations. Eligibility criteria for genomic selection are applied only after this step.

## 2. Preparing the analysis

### 2.1 Packages and helper functions

The packages below are used for table manipulation, figures, formatting, and project-relative paths.

```{r packages, warning=FALSE, message=FALSE}
library(dplyr) # Group, join, and summarize tables.
library(tidyr)   # Reshape data between long and wide formats.
library(ggplot2) # Build diagnostic figures.
library(ggthemes) # Apply the GDocs theme and color scale.
library(knitr)   # Format tables in R Markdown documents.
library(here)  # Build project-relative paths.
```

The helper function below standardizes table presentation throughout the tutorial.

```{r table-formatting, include=FALSE}
format_table <- function(x, caption = NULL, digits = getOption("digits"), col.names = NA) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    col.names = col.names,
    table.attr = paste0(
      'class="table table-striped table-hover" ',
      'style="width:100%; margin-bottom:0;"'
    )
  )
  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

### 2.2 Input objects and configuration

The module uses `matrices_object.rds`, containing phenotypes and metadata, and `models_object.rds`, containing fold-level metrics, OOF predictions, and MCMC diagnostics. Results are written only at the end of the analysis.

```{r inputs}
matrices_object <- readRDS(
  here("output", "matrices", "matrices_object.rds")
)

models_object <- readRDS(
  here("output", "models", "models_object.rds")
)

metrics <- models_object$metrics_by_fold
predictions <- models_object$predictions
convergence <- models_object$convergence_diagnostics

traits <- colnames(matrices_object$phenotypes)
methods <- models_object$settings$methods
```

Internal method codes are retained in analytical objects. The labels and theme below affect presentation only.

```{r presentation-style}
method_labels <- c(
  RRBLUP = "RR-BLUP/BRR",
  BayesA = "BayesA",
  BayesB = "BayesB",
  BayesB2 = "BayesB2",
  BayesC = "BayesC",
  BL = "Bayesian Lasso",
  GBLUP = "GBLUP"
)

selection_objective_labels <- c(
  increase = "Increase",
  decrease = "Decrease"
)

project_theme <- theme_gdocs() +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )
```

## 3. Method comparison and choice

### 3.1 OOF predictive performance

OOF estimates inherit the validation design defined in previous modules. Each individual contributes one OOF prediction per method and trait. Pooled metrics describe predictive performance in the population represented by the dataset.

```{r pooled-oof-metrics}
# Combine OOF predictions and compute aggregate predictive metrics by trait and method.
pooled_metrics <- predictions |>
  group_by(Trait, Method) |>
  summarise(
    N = n(),
    r = cor(Observed, Predicted),
    RMSE = sqrt(mean((Observed - Predicted)^2)),
    MAE = mean(abs(Observed - Predicted)),
    Bias = mean(Predicted - Observed),
    Slope = cov(Observed, Predicted) / var(Predicted),
    .groups = "drop"
  ) |>
  mutate(
    t_value = r * sqrt((N - 2) / (1 - r^2)),
    p_value = 2 * pt(abs(t_value), df = N - 2, lower.tail = FALSE)
  ) |>
  select(-t_value)

# Summarize the results in `fold_summary`.
fold_summary <- metrics |>
  group_by(Trait, Method) |>
  summarise(
    Mean_DIC = mean(DIC),
    SD_DIC = sd(DIC),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_fold_RMSE = mean(RMSE),
    SD_fold_RMSE = sd(RMSE),
    Mean_fold_MAE = mean(MAE),
    SD_fold_MAE = sd(MAE),
    .groups = "drop"
  )
```

Pearson correlation, RMSE, MAE, bias, and slope are calculated from the 150 OOF predictions. Correlation p-values are retained only as descriptive information.

### 3.2 DIC and relative support

DIC is summarized within each trait as complementary information. Following the logic of the reference study [@silva2021brrguava]:

\[
\Delta_j=\overline{DIC}_j-\min_k(\overline{DIC}_k),
\]

\[
Wprob_j=\frac{\exp(-\Delta_j/2)}{\sum_k\exp(-\Delta_k/2)},
\qquad
ER_j=\frac{Wprob_{\max}}{Wprob_j}.
\]

`Wprob` represents relative support derived from DIC and is not interpreted as a calibrated posterior model probability.

```{r model-comparison}
# Combine fold performance and DIC to build the quantities used for model comparison.
model_comparison <- pooled_metrics |>
  left_join(fold_summary, by = c("Trait", "Method")) |>
  group_by(Trait) |>
  mutate(
    Delta_DIC = Mean_DIC - min(Mean_DIC),
    log_weight = -0.5 * Delta_DIC,
    stable_weight = exp(log_weight - max(log_weight)),
    Wprob = stable_weight / sum(stable_weight),
    # Compute the evidence ratio relative to the method with the largest Wprob in each trait.
    ER = max(Wprob) / Wprob
  ) |>
  ungroup() |>
  select(
    Trait, Method, Mean_DIC, SD_DIC, Delta_DIC, Wprob, ER,
    N, r, p_value, RMSE, MAE, Bias, Slope,
    Mean_fold_r, SD_fold_r, Mean_fold_RMSE, SD_fold_RMSE,
    Mean_fold_MAE, SD_fold_MAE
  ) |>
  arrange(match(Trait, traits), Delta_DIC)
```

### 3.3 Integrated comparison and method choice

OOF predictive performance and DIC address different questions. OOF correlation is the main reference for predictive ability, whereas DIC provides complementary information on fit and complexity.

```{r predictive-correlation-figure, fig.width=13, fig.height=10, fig.alt="Faceted bar chart comparing pooled out-of-fold predictive correlations of the seven genomic prediction methods across the ten traits, with a horizontal reference at zero."}
# Order methods and build the OOF-correlation figure by trait.
predictive_plot <- model_comparison |>
  mutate(Method = factor(Method, levels = methods)) |>
  ggplot(aes(x = Method, y = r, fill = Method)) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.4) +
  geom_col(width = 0.75) +
  facet_wrap(~ Trait, scales = "fixed", ncol = 2) +
  scale_fill_gdocs(
    breaks = methods,
    labels = method_labels
  ) +
  scale_x_discrete(labels = method_labels) +
  labs(
    x = NULL,
    y = "OOF correlation",
    fill = "Method",
    title = "Predictive ability of the seven methods"
  ) +
  project_theme +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  theme(
    legend.direction = "horizontal",
    legend.box = "horizontal",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

predictive_plot
```

The joint table allows the two evidence sources to be compared without combining them into a single artificial score.

```{r model-comparison-table}
model_comparison_display <- model_comparison |>
  mutate(Method_label = unname(method_labels[Method])) |>
  transmute(
    Trait,
    Method,
    Method_label,
    DIC = sprintf("%.2f", Mean_DIC),
    Delta_DIC = sprintf("%.2f", Delta_DIC),
    Wprob = sprintf("%.3f", Wprob),
    ER = sprintf("%.3f", ER),
    r = sprintf("%.3f", r),
    p_value = if_else(
      p_value < 0.001,
      "<0.001",
      sprintf("%.3f", p_value)
    )
  ) |>
  arrange(match(Trait, traits), match(Method, methods))

format_table(
  model_comparison_display |>
    select(Trait, Method_label, DIC, Delta_DIC, Wprob, ER, r, p_value),
  caption = "Comparison of the seven methods by trait",
  col.names = c("Trait", "Method", "DIC", "Delta DIC", "Wprob", "ER", "r", "p-value")
)
```

The method used in subsequent steps is selected within each trait by the highest OOF correlation. Ties are resolved, in order, by lower RMSE, lower MAE, and lower mean DIC.

```{r final-model-selection}
trait_scale <- predictions |>
  distinct(Trait, IND, Observed) |>
  group_by(Trait) |>
  summarise(
    Phenotypic_mean = mean(Observed, na.rm = TRUE),
    Phenotypic_SD = sd(Observed, na.rm = TRUE),
    .groups = "drop"
  )

selected_methods <- model_comparison |>
  group_by(Trait) |>
  arrange(desc(r), RMSE, MAE, Mean_DIC, .by_group = TRUE) |>
  slice(1L) |>
  ungroup() |>
  transmute(
    Trait,
    Selected_method = Method,
    r,
    p_value,
    RMSE,
    MAE,
    Bias,
    Slope,
    Mean_DIC,
    Delta_DIC,
    Wprob,
    ER
  ) |>
  left_join(trait_scale, by = "Trait") |>
  arrange(match(Trait, traits))
```

The table below summarizes the method selected for each trait and the main
metrics used in that choice.

```{r selected-methods-table}
selected_methods_display <- selected_methods |>
  mutate(
    Method_label = unname(method_labels[Selected_method])
  ) |>
  transmute(
    Trait,
    Method_label,
    r = sprintf("%.3f", r),
    RMSE = sprintf("%.3f", RMSE),
    MAE = sprintf("%.3f", MAE),
    DIC = sprintf("%.2f", Mean_DIC)
  )

format_table(
  selected_methods_display,
  caption = "Method selected for each trait",
  col.names = c(
    "Trait",
    "Method",
    "OOF r",
    "RMSE",
    "MAE",
    "Mean DIC"
  )
)
```


### 3.4 MCMC diagnostics and observed-versus-predicted values

Diagnostics calculated in Module 03 are summarized globally. A fit is counted as unflagged when none of its monitored parameters received `Review_flag`.

```{r convergence-overview}
fit_convergence <- convergence |>
  group_by(Run_id, Trait, Method) |>
  summarise(
    Diagnostic_flag = any(Review_flag, na.rm = TRUE),
    .groups = "drop"
  )

convergence_overview <- data.frame(
  Level = c("Fits", "Monitored parameters"),
  Total = c(nrow(fit_convergence), nrow(convergence)),
  Without_flags = c(
    sum(!fit_convergence$Diagnostic_flag),
    sum(!convergence$Review_flag, na.rm = TRUE)
  )
) |>
  mutate(
    Proportion_without_flags = Without_flags / Total
  )

format_table(
  convergence_overview,
  digits = 4,
  caption = "Global summary of MCMC diagnostics",
  col.names = c("Level", "Total", "Without flags", "Proportion")
)
```

The predominance of fits and parameters without flags indicates that most chains showed behavior compatible with convergence according to the monitored diagnostics. This summary does not prove convergence of every individual marker effect and does not participate in method choice.

Observed-versus-predicted plots complement the numerical metrics. The dashed line represents the 1:1 relationship and the continuous line is the regression of observed values on OOF predictions.

```{r observed-predicted-figure, fig.width=12, fig.height=10, fig.alt="Faceted scatterplots of observed values against OOF values predicted by the selected method for each trait, with a 1:1 line and fitted linear regression."}
selected_predictions <- predictions |>
  inner_join(
    selected_methods |>
      select(Trait, Selected_method, r),
    by = "Trait"
  ) |>
  filter(Method == Selected_method) |>
  mutate(
    Method_label = unname(method_labels[Method]),
    Facet = paste0(
      Trait, " — ", Method_label,
      "\nr = ", sprintf("%.3f", r)
    )
  )

observed_predicted_plot <- selected_predictions |>
  ggplot(aes(x = Predicted, y = Observed)) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, linewidth = 0.5) +
  geom_point(alpha = 0.65, size = 1.6) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.7) +
  facet_wrap(~ Facet, scales = "free", ncol = 2) +
  labs(
    x = "OOF predicted value",
    y = "Observed value",
    title = "Relationship between observed and predicted values"
  ) +
  project_theme +
  theme(legend.position = "none")

observed_predicted_plot
```

## 4. OOF genomic values and relationships among traits

### 4.1 OOF GEBVs on the trait scale

For each trait, the genomic component comes from the selected method and remains cross-fitted: the individual's own phenotype was not used in the fit that produced its OOF score.

A version in the original trait units is also presented:

\[
GEBV_{scale} = \bar{y} + GEBV_{centered}.
\]

Adding the phenotypic mean only shifts the scale. It does not change the ranking and does not turn the GEBV into a complete phenotypic prediction.

```{r selected-oof-genomic-values}
oof_genomic_values <- predictions |>
  inner_join(
    selected_methods |>
      select(
        Trait, Selected_method, r, p_value
      ),
    by = "Trait"
  ) |>
  filter(Method == Selected_method) |>
  left_join(trait_scale, by = "Trait") |>
  transmute(
    IND,
    Trait,
    Fold,
    Method,
    Observed,
    Predicted,
    OOF_r = r,
    OOF_p_value = p_value,
    Phenotypic_mean,
    GEBV_centered = Genomic_component,
    GEBV_on_trait_scale = Phenotypic_mean + Genomic_component
  ) |>
  arrange(match(Trait, traits), IND)
```

### 4.2 Correlation among OOF GEBVs for all ten traits

Before any eligibility filter is applied, OOF GEBVs from the selected methods are organized for **all ten traits**. The matrix below describes associations among these cross-fitted scores.

```{r oof-correlations}
oof_genomic_wide <- oof_genomic_values |>
  select(IND, Trait, GEBV_centered) |>
  pivot_wider(names_from = Trait, values_from = GEBV_centered)

oof_genomic_correlations <- cor(
  as.matrix(oof_genomic_wide[, traits, drop = FALSE])
)
```

These correlations are not estimates of genetic correlation from a multivariate model. They describe only the association among OOF GEBVs produced separately for each trait.

```{r oof-correlation-figure, fig.width=10, fig.height=9, fig.alt="Heatmap of pairwise correlations among cross-fitted OOF genomic-value scores for the traits, representing exploratory associations rather than multivariate genetic correlations."}
# Reshape the OOF correlation matrix and build the cross-trait heatmap.
correlation_long <- as.data.frame(
  as.table(oof_genomic_correlations)
)
names(correlation_long) <- c(
  "Trait_1",
  "Trait_2",
  "Correlation"
)

# Set only the label of |r| < 0.005 to zero; the original correlation still drives analysis and color.
correlation_long <- correlation_long |>
  mutate(
    Correlation_label = replace(
      Correlation,
      abs(Correlation) < 0.005,
      0
    )
  )

# Build the heatmap of correlations among OOF genomic values.
correlation_plot <- ggplot(
  correlation_long,
  aes(x = Trait_1, y = Trait_2, fill = Correlation)
) +
  geom_tile() +
  geom_text(
    aes(label = sprintf("%.2f", Correlation_label)),
    size = 3
  ) +
  scale_fill_distiller(
    type = "div",
    palette = "RdBu",
    direction = 1,
    limits = c(-1, 1)
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "r",
    title = "Correlation among OOF GEBVs of eligible traits"
  ) +
  coord_equal() +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

correlation_plot
```

Positive values indicate a tendency to order individuals similarly, whereas negative values indicate opposite ordering directions. The matrix is descriptive and does not create an additional selection criterion.

## 5. Genomic selection in eligible traits

### 5.1 Eligibility criteria

Selection is performed only when the selected method for a trait simultaneously satisfies:

\[
r_{OOF} > 0.30
\]

and

\[
NRMSE = \frac{RMSE}{s_y} < 1,
\]

where \(s_y\) is the observed phenotypic standard deviation. The first criterion requires a minimum positive predictive association; the second requires prediction error below one phenotypic standard deviation.

```{r selection-parameters}
CORRELATION_SELECTION_THRESHOLD <- 0.30
NRMSE_SELECTION_THRESHOLD <- 1.00
SELECTION_PROPORTION <- 0.20

selection_objective_map <- c(
  MMCOM = "increase",
  PRODCOM = "increase",
  PRODEF = "decrease",
  CF = "increase",
  DF = "increase",
  DCI = "decrease",
  EP = "increase",
  FFC = "increase",
  FFP = "increase",
  SS = "increase"
)

selection_objectives <- data.frame(
  Trait = traits,
  Selection_objective = unname(selection_objective_map[traits]),
  stringsAsFactors = FALSE
)

model_selection <- selected_methods |>
  left_join(selection_objectives, by = "Trait") |>
  mutate(
    NRMSE = RMSE / Phenotypic_SD,
    Predictive_skill_vs_SD = 1 - NRMSE,
    Eligible_by_correlation = r > CORRELATION_SELECTION_THRESHOLD,
    Eligible_by_error = NRMSE < NRMSE_SELECTION_THRESHOLD,
    Selection_eligible =
      Eligible_by_correlation & Eligible_by_error
  ) |>
  arrange(match(Trait, traits))

eligible_traits <- model_selection |>
  filter(Selection_eligible) |>
  pull(Trait)

selection_parameters <- data.frame(
  Parameter = c(
    "Method choice",
    "Minimum OOF correlation",
    "Maximum NRMSE",
    "Selected proportion"
  ),
  Rule = c(
    "Highest r; ties by lower RMSE, lower MAE, and lower DIC",
    "> 0.30",
    "< 1.00",
    "20% per eligible trait"
  ),
  stringsAsFactors = FALSE
)

format_table(
  selection_parameters,
  caption = "Criteria used for method choice and genomic selection",
  col.names = c("Parameter", "Rule")
)
```

The table below combines the selected method, relevant performance metrics, and the eligibility decision. The correlation p-value remains available in the analytical object but is not used as a filter.

```{r model-selection-table}
model_selection_display <- model_selection |>
  mutate(
    Method_label = unname(method_labels[Selected_method]),
    Objective = unname(selection_objective_labels[Selection_objective]),
    Eligible = if_else(Selection_eligible, "Yes", "No")
  ) |>
  transmute(
    Trait,
    Method_label,
    Objective,
    r = sprintf("%.3f", r),
    RMSE = sprintf("%.3f", RMSE),
    NRMSE = sprintf("%.3f", NRMSE),
    MAE = sprintf("%.3f", MAE),
    Bias = sprintf("%.3f", Bias),
    Slope = sprintf("%.3f", Slope),
    DIC = sprintf("%.2f", Mean_DIC),
    Eligible
  )

format_table(
  model_selection_display,
  caption = "Selected method and eligibility for selection by trait",
  col.names = c(
    "Trait", "Method", "Objective", "r", "RMSE", "NRMSE",
    "MAE", "Bias", "Slope", "Mean DIC", "Eligible"
  )
)
```

### 5.2 Ranking and selected individuals

For eligible traits, individuals are ordered by the GEBV on the trait scale while respecting the breeding objective. For traits to be reduced, the sign is reversed only when constructing the ranking score. The selected proportion is 20%.

```{r genomic-ranking}
selection_genomic_values <- oof_genomic_values |>
  inner_join(
    model_selection |>
      select(
        Trait,
        Selection_objective,
        Selection_eligible,
        NRMSE
      ),
    by = "Trait"
  ) |>
  mutate(OOF_NRMSE = NRMSE)

genomic_ranking <- selection_genomic_values |>
  filter(Selection_eligible) |>
  mutate(
    Selection_score = case_when(
      Selection_objective == "increase" ~ GEBV_on_trait_scale,
      Selection_objective == "decrease" ~ -GEBV_on_trait_scale,
      TRUE ~ NA_real_
    )
  ) |>
  group_by(Trait) |>
  arrange(desc(Selection_score), IND, .by_group = TRUE) |>
  mutate(Rank = row_number()) |>
  ungroup() |>
  arrange(match(Trait, traits), Rank)

selection_sizes <- genomic_ranking |>
  count(Trait, name = "N_available") |>
  mutate(
    Selection_proportion = SELECTION_PROPORTION,
    N_selected = pmax(
      1L,
      as.integer(ceiling(Selection_proportion * N_available))
    )
  ) |>
  arrange(match(Trait, traits))

selected_individuals_by_trait <- genomic_ranking |>
  inner_join(selection_sizes, by = "Trait") |>
  filter(Rank <= N_selected) |>
  left_join(
    matrices_object$metadata |>
      select(IND, FAM),
    by = "IND"
  ) |>
  select(
    Trait, Rank, IND, FAM, Method, Selection_objective,
    GEBV_centered, GEBV_on_trait_scale, Selection_score,
    OOF_r, OOF_NRMSE, Selection_proportion,
    N_available, N_selected
  ) |>
  arrange(match(Trait, traits), Rank)
```

The table presents the top five individuals for each eligible trait; the complete object retains all selected individuals.

```{r proportional-selection-preview}
selection_preview <- selected_individuals_by_trait |>
  group_by(Trait) |>
  slice_head(n = 5) |>
  ungroup() |>
  mutate(
    Method_label = unname(method_labels[Method]),
    Objective = unname(selection_objective_labels[Selection_objective])
  ) |>
  transmute(
    Trait,
    Rank,
    IND,
    FAM,
    Method_label,
    Objective,
    GEBV = sprintf("%.3f", GEBV_on_trait_scale),
    r = sprintf("%.3f", OOF_r),
    NRMSE = sprintf("%.3f", OOF_NRMSE)
  )

format_table(
  selection_preview,
  caption = "First five selected individuals for each eligible trait",
  col.names = c(
    "Trait", "Rank", "IND", "FAM", "Method", "Objective",
    "GEBV + mean", "OOF r", "NRMSE"
  )
)
```

### 5.3 Recurrence among eligible traits

Recurrence counts the number of eligible traits in which each individual appears among the selected 20%. Frequency is calculated relative to the number of eligible traits and is not a multi-trait selection index.

```{r multitrait-recurrence}
recurrence_counts <- selected_individuals_by_trait |>
  group_by(IND, FAM) |>
  summarise(
    N_traits = n(),
    Traits = paste(Trait, collapse = ", "),
    Best_rank = min(Rank),
    Mean_rank_present = mean(Rank),
    .groups = "drop"
  )

n_eligible_traits <- length(eligible_traits)

multitrait_recurrence <- matrices_object$metadata |>
  select(IND, FAM) |>
  left_join(recurrence_counts, by = c("IND", "FAM")) |>
  mutate(
    N_traits = tidyr::replace_na(N_traits, 0L),
    Traits = tidyr::replace_na(Traits, ""),
    Best_rank = if_else(N_traits == 0L, NA_integer_, Best_rank),
    Mean_rank_present = if_else(
      N_traits == 0L,
      NA_real_,
      Mean_rank_present
    ),
    Selection_frequency = if (n_eligible_traits > 0) {
      N_traits / n_eligible_traits
    } else {
      0
    },
    Recurrence_class = case_when(
      N_traits >= 4L ~ "high_4plus",
      N_traits == 3L ~ "intermediate_3",
      N_traits == 2L ~ "broad_2",
      N_traits == 1L ~ "specific_1",
      TRUE ~ "none_0"
    )
  ) |>
  arrange(desc(N_traits), FAM, IND)

recurrence_distribution <- multitrait_recurrence |>
  count(N_traits, Recurrence_class, name = "N_individuals") |>
  mutate(
    Proportion_of_population =
      N_individuals / nrow(matrices_object$metadata)
  ) |>
  arrange(desc(N_traits))

multitrait_candidates <- multitrait_recurrence |>
  filter(N_traits >= 2L) |>
  arrange(desc(N_traits), FAM, IND)
```

The distribution summarizes how many individuals were selected in zero, one, two, three, or more traits.

```{r recurrence-distribution-table}
recurrence_class_labels <- c(
  high_4plus = "High (>=4)",
  intermediate_3 = "Intermediate (3)",
  broad_2 = "Broad (2)",
  specific_1 = "Specific (1)",
  none_0 = "None (0)"
)

recurrence_distribution_display <- recurrence_distribution |>
  mutate(
    Class = unname(recurrence_class_labels[Recurrence_class])
  ) |>
  transmute(
    N_traits,
    Class,
    N_individuals,
    Proportion_of_population = sprintf("%.3f", Proportion_of_population)
  )

format_table(
  recurrence_distribution_display,
  caption = "Distribution of individuals by recurrence across eligible traits",
  col.names = c(
    "Number of traits", "Class", "Individuals",
    "Population proportion"
  )
)
```

Individuals selected in at least two traits are shown in a single table.

```{r recurrent-candidates-table}
recurrent_candidates_display <- multitrait_candidates |>
  mutate(
    Class = unname(recurrence_class_labels[Recurrence_class])
  ) |>
  transmute(
    IND,
    FAM,
    N_traits,
    Selection_frequency = sprintf("%.3f", Selection_frequency),
    Class,
    Traits,
    Best_rank,
    Mean_rank_present = sprintf("%.2f", Mean_rank_present)
  )

format_table(
  recurrent_candidates_display,
  caption = "Individuals recurrent in at least two eligible traits",
  col.names = c(
    "IND", "FAM", "Number of traits", "Frequency",
    "Class", "Traits", "Best rank", "Mean rank"
  )
)
```

## 6. Marker-based heritability analogue and adjusted accuracy in RR-BLUP/BRR

Following Silva et al. [@silva2021brrguava], RR-BLUP/BRR allows a marker-based heritability analogue to be derived from marker variance. Fold-level adjusted accuracy is calculated as `Correlation / sqrt(H2)`.

This ratio is a derived quantity rather than a direct correlation between true and predicted genetic values. It may therefore be negative or exceed 1.

```{r rrblup-heritability}
# Compute and organize the RR-BLUP/BRR model-based heritability summary.
rrblup_variance_components <- convergence |>
  filter(
    Method == "RRBLUP",
    Parameter %in% c("Residual_variance", "Marker_variance")
  ) |>
  select(Run_id, Trait, Fold, Method, Parameter, Posterior_mean) |>
  pivot_wider(names_from = Parameter, values_from = Posterior_mean)

rrblup_fold_correlations <- metrics |>
  filter(Method == "RRBLUP") |>
  select(Trait, Fold, Correlation)

# Organize the cross-validation and fitting structure.
rrblup_heritability_by_fold <- rrblup_variance_components |>
  left_join(rrblup_fold_correlations, by = c("Trait", "Fold")) |>
  mutate(
    Marker_variance_denominator = matrices_object$denominator,
    Additive_variance = Marker_variance * Marker_variance_denominator,
    H2 = Additive_variance / (Additive_variance + Residual_variance),
    Predictive_accuracy = Correlation / sqrt(H2)
  ) |>
  arrange(match(Trait, traits), Fold)

# Summarize the RR-BLUP/BRR model-based heritability analogue and adjusted accuracy across folds.
rrblup_heritability_summary <- rrblup_heritability_by_fold |>
  group_by(Trait) |>
  summarise(
    Mean_H2 = mean(H2),
    SD_H2 = sd(H2),
    Mean_fold_r = mean(Correlation),
    SD_fold_r = sd(Correlation),
    Mean_predictive_accuracy = mean(Predictive_accuracy),
    SD_predictive_accuracy = sd(Predictive_accuracy),
    .groups = "drop"
  ) |>
  arrange(match(Trait, traits))
```

The trait summary presents the marker-based heritability analogue and adjusted accuracy across folds.

```{r rrblup-heritability-table}
rrblup_table <- rrblup_heritability_summary |>
  transmute(
    Trait,
    RRBLUP_H2 = sprintf("%.3f", Mean_H2),
    SD_RRBLUP_H2 = sprintf("%.3f", SD_H2),
    RRBLUP_predictive_accuracy = sprintf("%.3f", Mean_predictive_accuracy),
    SD_RRBLUP_predictive_accuracy = sprintf("%.3f", SD_predictive_accuracy)
  )

format_table(
  rrblup_table,
  caption = "Marker-based heritability analogue and adjusted predictive accuracy in RR-BLUP/BRR",
  col.names = c("Trait", "RR-BLUP H2", "SD RR-BLUP H2", "RR-BLUP predictive accuracy", "SD RR-BLUP predictive accuracy")
)
```

## 7. Exporting the analysis results

At the end of the analysis, the main objects and tables are written to a Module 04 results directory. These files allow the results to be inspected later without repeating the derivations shown above.

```{r save-results}
results_dir <- here(
  "output", "results", "m04", "genomic_selection"
)
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

write.csv(
  model_selection,
  file.path(results_dir, "model_selection.csv"),
  row.names = FALSE
)
write.csv(
  oof_genomic_values,
  file.path(results_dir, "oof_genomic_values.csv"),
  row.names = FALSE
)
write.csv(
  selected_individuals_by_trait,
  file.path(results_dir, "selected_individuals_by_trait.csv"),
  row.names = FALSE
)
write.csv(
  multitrait_recurrence,
  file.path(results_dir, "multitrait_recurrence.csv"),
  row.names = FALSE
)
write.csv(
  multitrait_candidates,
  file.path(results_dir, "multitrait_candidates.csv"),
  row.names = FALSE
)
write.csv(
  oof_genomic_correlations,
  file.path(results_dir, "oof_genomic_correlations.csv"),
  row.names = TRUE
)

module04_results <- list(
  model_comparison = model_comparison,
  selected_methods = selected_methods,
  model_selection = model_selection,
  convergence_overview = convergence_overview,
  rrblup_heritability_by_fold = rrblup_heritability_by_fold,
  rrblup_heritability_summary = rrblup_heritability_summary,
  oof_genomic_values = oof_genomic_values,
  oof_genomic_correlations = oof_genomic_correlations,
  genomic_ranking = genomic_ranking,
  selected_individuals_by_trait = selected_individuals_by_trait,
  multitrait_recurrence = multitrait_recurrence,
  multitrait_candidates = multitrait_candidates,
  settings = list(
    correlation_threshold = CORRELATION_SELECTION_THRESHOLD,
    nrmse_threshold = NRMSE_SELECTION_THRESHOLD,
    selection_proportion = SELECTION_PROPORTION,
    eligible_traits = eligible_traits
  )
)

saveRDS(
  module04_results,
  file.path(results_dir, "module04_results.rds")
)
```

## 8. Summary and interpretation

The seven methods are compared separately for each trait. OOF predictive evidence determines which method is used in subsequent steps, while DIC remains complementary information.

OOF GEBVs from the selected methods are first examined across all ten traits, including their correlations. This step precedes eligibility criteria and allows relationships among traits to be interpreted without restricting the analysis to traits that later enter genomic selection.

Genomic selection is then limited to traits with \(r > 0.30\) and `NRMSE < 1`. Within those traits, the most favorable 20% of individuals are prioritized according to the breeding objective, and recurrence summarizes the presence of the same individuals across traits.

The GEBVs remain cross-fitted scores. Adding the phenotypic mean improves interpretation in trait units but does not alter rankings or represent a complete phenotypic prediction.

Conclusions remain conditional on the represented breeding population, the available SSR panel, and the validation design.
