---
title: "5. Out-of-fold model comparison"
author: "Weverton Gomes da Costa"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
bibliography: ../REFERENCES.bib
link-citations: true
---

[Português](05_model_comparison_pt.html) | [Previous](04_gblup_en.html) | [Home](index.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 7)
source(file.path("code", "00_config.R"))
source(file.path("code", "02_model_functions.R"))
source(file.path("code", "03_reporting_functions.R"))
configuracao <- obter_configuracao()
objeto <- carregar_resultado(configuracao$arquivo_preprocessamento)
resultado_bayesiano <- carregar_resultado(configuracao$arquivo_bayesiano)
resultado_rrblup <- carregar_resultado(configuracao$arquivo_rrblup)
resultado_gblup <- carregar_resultado(configuracao$arquivo_gblup)
```

## Comparison rule

All eight models are compared from the same 150 out-of-fold predictions per
trait. Each individual contributes exactly one held-out observation for each
model. The primary ranking is deterministic: maximize predictive Pearson
correlation, then minimize RMSE, then minimize MAE. Bias, calibration slope,
and fold variability remain visible as safeguards rather than being hidden in
a single score.

DIC is not used in the cross-family ranking because it is available only for
the Bayesian likelihood/prior framework. Its separate table on page 2 answers
a different question from held-out predictive performance.

## Build the common result

The common result is generated only when the Bayesian, RR-BLUP, and GBLUP
objects are all available. The output includes raw predictions, fold metrics,
metrics recomputed on concatenated out-of-fold predictions, and the selected
model for every trait.

```{r consolidate-models}
resultados_disponiveis <- all(vapply(
  list(resultado_bayesiano, resultado_rrblup, resultado_gblup, objeto),
  Negate(is.null),
  logical(1)
))

if (resultados_disponiveis) {
  comparacao <- consolidar_comparacao_modelos(
    objeto,
    resultado_bayesiano,
    resultado_rrblup,
    resultado_gblup,
    configuracao
  )
} else {
  comparacao <- carregar_resultado(configuracao$arquivo_comparacao)
}
```

```{r comparison-status, results='asis', echo=FALSE}
if (is.null(comparacao)) {
  cat(mensagem_resultado_ausente(configuracao$arquivo_comparacao, "en"))
}
```

## Common predictive metrics

These metrics are calculated once per trait and model from all concatenated
held-out predictions. Correlation measures predictive ranking, while RMSE and
MAE retain the original units. Bias is prediction minus observation. A well-
calibrated prediction has intercept near zero and slope near one.

```{r common-metrics-table, results='asis'}
if (!is.null(comparacao)) {
  metricas <- comparacao$out_of_fold_metrics |>
    dplyr::arrange(Trait, dplyr::desc(Correlation), RMSE, MAE)
  print(tabela_relatorio(
    metricas,
    "Metrics from concatenated out-of-fold predictions",
    rolagem = TRUE
  ))
}
```

The correlation plot exposes the model-by-trait pattern; it should be read
together with the error and calibration columns in the table.

```{r correlation-comparison-plot, echo=FALSE, fig.height=10}
if (!is.null(comparacao)) {
  print(
    ggplot2::ggplot(
      comparacao$out_of_fold_metrics,
      ggplot2::aes(x = Model, y = Correlation, fill = Model)
    ) +
      ggplot2::geom_hline(yintercept = 0, color = "grey70") +
      ggplot2::geom_col(show.legend = FALSE) +
      ggplot2::facet_wrap(~ Trait, scales = "free_y", ncol = 2) +
      ggplot2::labs(
        x = NULL,
        y = "Out-of-fold correlation",
        title = "Predictive correlation across eight genomic models"
      ) +
      tema_relatorio() +
      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
  )
}
```

## Selected model by trait

The table records the model selected by the explicit tie-breaking rule. Model
selection is trait-specific; no assumption is made that one prior or BLUP
representation must dominate all biological responses.

```{r best-model-table, results='asis'}
if (!is.null(comparacao)) {
  melhores <- comparacao$best_models |>
    dplyr::select(
      Trait, Model, N, Correlation, Correlation_p_value,
      RMSE, MAE, Bias, Calibration_intercept, Calibration_slope
    )
  print(tabela_relatorio(melhores, "Best predictive model for each trait"))
}
```

Predictions from the selected model are displayed below. This is a diagnostic
of agreement and dispersion, not a replacement for the numerical metrics.

```{r best-model-prediction-plot, echo=FALSE, fig.height=10}
if (!is.null(comparacao)) {
  predicoes_melhores <- comparacao$predictions |>
    dplyr::inner_join(
      comparacao$best_models |> dplyr::select(Trait, Model),
      by = c("Trait", "Model")
    )

  print(
    ggplot2::ggplot(predicoes_melhores, ggplot2::aes(x = Observed, y = Predicted)) +
      ggplot2::geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey55") +
      ggplot2::geom_point(ggplot2::aes(color = Model), alpha = 0.75, size = 1.5) +
      ggplot2::facet_wrap(~ Trait, scales = "free", ncol = 2) +
      ggplot2::labs(
        x = "Observed",
        y = "Out-of-fold prediction",
        color = "Selected model",
        title = "Out-of-fold predictions from the selected model"
      ) +
      tema_relatorio()
  )
}
```

The next page uses full-data genomic values from these selected trait-specific
models to form candidate rankings.

[Next: genomic selection](06_genomic_selection_en.html)
