---
title: "Presentation of results"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

This module organizes the results derived in Module 04 into tables and figures.
The presentation starts with the comparison of the seven methods, proceeds to
the selected methods and chain diagnostics, and finishes with genomic-selection
results and recurrence across traits.

## 2. Packages, results, and graphical style

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(here)
})
```

The next step loads the input objects and organizes the structures required for the analysis.

```{r inputs-and-style}
# Execute the calculation block corresponding to this stage of the module.
MODELS_FILE <- here("output", "models", "models_object.rds")
M04_FILE <- here("output", "results", "m04", "module04_results.rds")
RESULTS_DIR <- here("output", "results", "m05")
FIGURE_DIR <- file.path(RESULTS_DIR, "figures")

dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)

models_object <- readRDS(MODELS_FILE)
m04 <- readRDS(M04_FILE)

traits <- m04$traits
methods <- m04$methods
model_comparison <- m04$model_comparison
model_selection <- m04$model_selection
selected_fold_diagnostics <- m04$selected_fold_diagnostics
runtime_summary <- m04$runtime
# Summarize the results in `rrblup_heritability_summary`.
rrblup_heritability_summary <- m04$rrblup_heritability_summary
selection_objectives <- m04$selection_objectives
selection_proportion <- m04$selection_proportion
positive_oof_traits <- m04$positive_oof_traits
selection_sizes <- m04$selection_sizes
selected_individuals_by_trait <- m04$selected_individuals_by_trait
multitrait_recurrence <- m04$multitrait_recurrence
recurrence_distribution <- m04$recurrence_distribution
multitrait_candidates <- m04$multitrait_candidates
selected_oof_correlations <- m04$selected_oof_correlations

# Create `method_labels` for use in the following steps of this block.
method_labels <- c(
  RRBLUP = "RR-BLUP/BRR",
  BayesA = "BayesA",
  BayesB = "BayesB",
  BayesB2 = "BayesB2",
  BayesC = "BayesC",
  BL = "Bayesian Lasso",
  GBLUP = "GBLUP"
)

selection_objective_labels <- c(
  increase = "Increase",
  decrease = "Decrease"
)

# Create `method_colors` for use in the following steps of this block.
method_colors <- c(
  RRBLUP = "#3366CC",
  BayesA = "#DC3912",
  BayesB = "#FF9900",
  BayesB2 = "#109618",
  BayesC = "#990099",
  BL = "#0099C6",
  GBLUP = "#DD4477"
)

project_theme <- theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )
```

The next step loads the input objects and organizes the structures required for the analysis.

```{r input-summary}
# Assemble the summary table presented in this section.
data.frame(
  Item = c(
    "Traits",
    "Methods",
    "Rows in method comparison",
    "Traits with a selected method",
    "Traits with positive OOF correlation"
  ),
  # List the values corresponding to the table entries.
  Value = c(
    length(traits),
    length(methods),
    nrow(model_comparison),
    nrow(model_selection),
    length(positive_oof_traits)
  )
) |>
  kable(caption = "Structure of the results presented in this module")
```

## 3. Comparison of the seven methods

The table combines DIC, difference from the smallest DIC, `Wprob`, `ER`, OOF
correlation, and nominal p-value for every trait × method combination.

```{r model-comparison-table}
# Assemble the summary table presented in this section.
model_comparison_display <- model_comparison |>
  mutate(Method_label = unname(method_labels[Method])) |>
  transmute(
    Trait,
    Method,
    Method_label,
    DIC = Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    r,
    p_value
  ) |>
  arrange(match(Trait, traits), match(Method, methods))

# Format the table for display in the tutorial.
kable(
  model_comparison_display |>
    select(Trait, Method_label, DIC, Delta_DIC, Wprob, ER, r, p_value),
  digits = 5,
  caption = "Comparison of the seven methods by trait"
)
```

The next step builds the corresponding visualization to support interpretation of the diagnostic.

```{r predictive-correlation-figure, fig.width=13, fig.height=10}
# Compute the correlations used to compare traits or results.
predictive_plot <- model_comparison |>
  mutate(Method = factor(Method, levels = methods)) |>
  # Build the figure used for interpretation.
  ggplot(aes(x = Method, y = r, fill = Method)) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.4) +
  geom_col(width = 0.75) +
  facet_wrap(~ Trait, scales = "fixed", ncol = 2) +
  scale_fill_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  scale_x_discrete(labels = method_labels) +
  labs(
    x = NULL,
    y = "OOF correlation",
    fill = "Method",
    title = "Predictive ability of the seven methods"
  ) +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

predictive_plot

# Save the consolidated output for downstream use.
ggsave(
  file.path(FIGURE_DIR, "predictive_correlation_by_method_en.png"),
  predictive_plot,
  width = 13,
  height = 10,
  dpi = 300
)
```

## 4. Selected methods by trait

The next step derives and presents the corresponding selection or prioritization rule from the calculated results.

```{r selected-methods-table}
# Assemble the summary table presented in this section.
final_results_summary <- model_selection |>
  mutate(
    Method_label = unname(method_labels[Selected_method]),
    Predictive_winner_label = unname(method_labels[Predictive_winner]),
    DIC_winner_label = unname(method_labels[DIC_winner]),
    Decision_label = case_when(
      Decision_code == "rrblup_best_predictive_model" ~
        "RR-BLUP/BRR was the best predictive model",
      Decision_code == "rrblup_preferred_among_competitive_models" ~
        "RR-BLUP/BRR preferred among competitive methods",
      Decision_code == "best_non_rrblup_competitive_model" ~
        "Best competitive non-RR-BLUP method",
      Decision_code == "predictive_winner_provisional_review_required" ~
        "Provisional predictive winner; review required",
      TRUE ~ Decision_code
    ),
    Selection_status_label = case_when(
      # Define the `Selection_status` values used in this structure.
      Selection_status == "rule_based_selected" ~
        "Rule-based selection",
      Selection_status == "provisional_review_required" ~
        "Provisional selection; review required",
      TRUE ~ Selection_status
    )
  ) |>
  transmute(
    Trait,
    Method_label,
    Decision_label,
    Selection_status_label,
    RRBLUP_preference_applied,
    Predictive_winner = Predictive_winner_label,
    # Define the `DIC_winner` values used in this structure.
    DIC_winner = DIC_winner_label,
    Predictive_loss,
    r,
    p_value,
    RMSE,
    DIC = Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    Parameters_flagged,
    Runs_with_flags,
    OOF_predictive_direction
  ) |>
  arrange(match(Trait, traits))

# Format the table for display in the tutorial.
kable(
  final_results_summary,
  digits = 5,
  caption = "Selected methods and diagnostics by trait"
)
```

## 5. MCMC chains of the selected methods

For each trait, the posterior chain of \(\mu\), its running mean, and its
posterior density are shown for the first fold of the selected method.

```{r selected-mu-data}
# Execute the calculation block corresponding to this stage of the module.
burn_in_saved <-
  models_object$settings$burn_in /
  models_object$settings$thin

selected_method_lookup <- setNames(
  model_selection$Selected_method,
  model_selection$Trait
)

trace_parts <- vector("list", length(traits))
density_parts <- vector("list", length(traits))

for (trait_index in seq_along(traits)) {
  # Identify the trait and selected method.
  trait_name <- traits[trait_index]
  method_name <- selected_method_lookup[trait_name]

  # Locate the mu chain from the first fold.
  mu_file <- file.path(
    here("output", "models", "bglr"),
    trait_name,
    "F1",
    method_name,
    "bglr_mu.dat"
  )

  mu_full <- scan(mu_file, quiet = TRUE)

  # Remove samples corresponding to burn-in.
  posterior_mu <- mu_full[
    seq.int(burn_in_saved + 1L, length(mu_full))
  ]

  iteration <- seq.int(
    from = models_object$settings$burn_in + models_object$settings$thin,
    by = models_object$settings$thin,
    length.out = length(posterior_mu)
  )

  # Compute the running mean and reduce point density for plotting.
  running_mean <- cumsum(posterior_mu) / seq_along(posterior_mu)
  keep_index <- seq.int(1L, length(posterior_mu), by = 100L)

  trace_parts[[trait_index]] <- data.frame(
    Trait = trait_name,
    Method = method_name,
    Facet = paste0(trait_name, " — ", method_labels[method_name]),
    Iteration = iteration[keep_index],
    Posterior_mu = posterior_mu[keep_index],
    Running_mean = running_mean[keep_index]
  )

  # Create `density_object` for use in the following steps of this block.
  density_object <- density(posterior_mu, n = 512)

  density_parts[[trait_index]] <- data.frame(
    Trait = trait_name,
    Method = method_name,
    Facet = paste0(trait_name, " — ", method_labels[method_name]),
    x = density_object$x,
    y = density_object$y
  )
}

selected_mu_trace <- bind_rows(trace_parts)
selected_mu_density <- bind_rows(density_parts)
```

The next step builds the corresponding visualization to support interpretation of the diagnostic.

```{r selected-mu-trace-figure, fig.width=13, fig.height=14}
# Build the visualization used to interpret the results in this section.
mu_trace_plot <- ggplot(
  selected_mu_trace,
  aes(x = Iteration, y = Posterior_mu)
) +
  geom_line(linewidth = 0.25, alpha = 0.35) +
  geom_line(
    aes(y = Running_mean, color = Method),
    linewidth = 0.8
  ) +
  facet_wrap(~ Facet, scales = "free_y", ncol = 2) +
  scale_color_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  labs(
    x = "Iteration",
    y = expression(mu),
    color = "Method",
    title = "Posterior chain and running mean of mu — fold 1"
  ) +
  project_theme

mu_trace_plot

# Save the consolidated output for downstream use.
ggsave(
  file.path(FIGURE_DIR, "selected_model_mu_trace_en.png"),
  mu_trace_plot,
  width = 13,
  height = 14,
  dpi = 300
)
```

The next step builds the corresponding visualization to support interpretation of the diagnostic.

```{r selected-mu-density-figure, fig.width=13, fig.height=14}
# Build the visualization used to interpret the results in this section.
mu_density_plot <- ggplot(
  selected_mu_density,
  aes(x = x, y = y, color = Method)
) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ Facet, scales = "free", ncol = 2) +
  scale_color_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  labs(
    x = expression(mu),
    y = "Density",
    color = "Method",
    title = "Posterior density of mu — fold 1"
  ) +
  project_theme

mu_density_plot

# Save the consolidated output for downstream use.
ggsave(
  file.path(FIGURE_DIR, "selected_model_mu_density_en.png"),
  mu_density_plot,
  width = 13,
  height = 14,
  dpi = 300
)
```

## 6. Stability across folds

The next step builds the corresponding visualization to support interpretation of the diagnostic.

```{r selected-fold-figure, fig.width=12, fig.height=10}
# Build or summarize the cross-validation fold structure.
fold_stability_display <- selected_fold_diagnostics |>
  mutate(
    Method_label = unname(method_labels[Method]),
    Facet = paste0(Trait, " — ", Method_label)
  )

# Organize the cross-validation and fitting structure.
fold_stability_plot <- ggplot(
  fold_stability_display,
  aes(x = Fold, y = Correlation, group = Trait)
) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.4) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~ Facet, ncol = 2) +
  scale_x_continuous(breaks = 1:5) +
  coord_cartesian(ylim = c(-1, 1)) +
  labs(
    x = "Fold",
    y = "Validation correlation",
    title = "Predictive correlation across folds for selected methods"
  ) +
  project_theme

fold_stability_plot

# Save the consolidated output for downstream use.
ggsave(
  file.path(FIGURE_DIR, "selected_method_fold_stability_en.png"),
  fold_stability_plot,
  width = 12,
  height = 10,
  dpi = 300
)
```

## 7. Computational time

The next step builds the corresponding visualization to support interpretation of the diagnostic.

```{r runtime-figure, fig.width=11, fig.height=7}
# Summarize the computational cost of the evaluated methods.
runtime_plot <- runtime_summary |>
  mutate(Method = factor(Method, levels = methods)) |>
  # Build the figure used for interpretation.
  ggplot(aes(x = Method, y = Mean_seconds, fill = Method)) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = pmax(0, Mean_seconds - SD_seconds),
      ymax = Mean_seconds + SD_seconds
    ),
    width = 0.2
  ) +
  scale_fill_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  scale_x_discrete(labels = method_labels) +
  labs(
    x = NULL,
    y = "Mean time (seconds)",
    fill = "Method",
    title = "Computational time by method"
  ) +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

runtime_plot

# Save the consolidated output for downstream use.
ggsave(
  file.path(FIGURE_DIR, "runtime_by_method_en.png"),
  runtime_plot,
  width = 11,
  height = 7,
  dpi = 300
)
```

## 8. Marker-based heritability analogue and adjusted accuracy in RR-BLUP/BRR

The next step organizes a tabular summary of the objects and results needed to interpret this section.

```{r rrblup-heritability-table}
rrblup_table <- rrblup_heritability_summary |>
  transmute(
    Trait,
    RRBLUP_H2 = Mean_H2,
    SD_RRBLUP_H2 = SD_H2,
    RRBLUP_predictive_accuracy = Mean_predictive_accuracy,
    SD_RRBLUP_predictive_accuracy = SD_predictive_accuracy
  )

kable(
  rrblup_table,
  digits = 5,
  caption = "Marker-based heritability analogue and adjusted predictive accuracy in RR-BLUP/BRR"
)
```

## 9. Correlations among OOF genomic values

The next step builds the corresponding visualization to support interpretation of the diagnostic.

```{r oof-correlation-figure, fig.width=10, fig.height=9}
# Compute the correlations used to compare traits or results.
correlation_long <- as.data.frame(
  as.table(selected_oof_correlations)
)
names(correlation_long) <- c(
  "Trait_1",
  "Trait_2",
  "Correlation"
)

# Create `correlation_long` for use in the following steps of this block.
correlation_long <- correlation_long |>
  mutate(
    Correlation_label = replace(
      Correlation,
      abs(Correlation) < 0.005,
      0
    )
  )

# Create `correlation_plot` for use in the following steps of this block.
correlation_plot <- ggplot(
  correlation_long,
  aes(x = Trait_1, y = Trait_2, fill = Correlation)
) +
  geom_tile() +
  geom_text(
    aes(label = sprintf("%.2f", Correlation_label)),
    size = 3
  ) +
  scale_fill_gradient2(
    low = "#DC3912",
    mid = "white",
    high = "#3366CC",
    midpoint = 0,
    # Organize the `limits` component of this structure.
    limits = c(-1, 1)
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "r",
    title = "Correlation among OOF genomic values"
  ) +
  coord_equal() +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

correlation_plot

# Save the consolidated output for downstream use.
ggsave(
  file.path(FIGURE_DIR, "selected_oof_correlations_en.png"),
  correlation_plot,
  width = 10,
  height = 9,
  dpi = 300
)
```

## 10. Exploratory proportional candidate ranking by trait

Candidate rankings are presented for traits whose selected method has positive
OOF correlation. Nominal p-values are descriptive and do not determine whether
a ranking is shown. Rankings with nominal `p >= 0.05` are explicitly
exploratory and should not be interpreted as demonstrated predictive utility.
The proportion shown is `r sprintf("%.0f%%", 100 * selection_proportion)` of
the individuals within each trait and is a configurable tutorial rule rather
than an estimated optimal intensity. Rankings use cross-fitted OOF genomic
values, not final breeding values from a model refitted to all phenotypes.

```{r proportional-selection-table}
# Derive selection rankings or summaries from the predictive results.
selection_sizes_display <- selection_sizes |>
  left_join(
    selection_objectives |>
      select(Trait, Selection_objective),
    by = "Trait"
  ) |>
  left_join(
    model_selection |>
      select(Trait, r, p_value, OOF_nominal_signal),
    by = "Trait"
  ) |>
  mutate(
    Objective = unname(
      selection_objective_labels[Selection_objective]
    )
  ) |>
  # Keep and organize the variables used in the result.
  select(
    Trait,
    Objective,
    r,
    p_value,
    OOF_nominal_signal,
    Selection_proportion,
    N_available,
    N_selected
  )

# Format the table for display in the tutorial.
kable(
  selection_sizes_display,
  digits = 2,
  caption = "Exploratory candidate-ranking summary by positive-OOF trait"
)
```

The first individuals in each ranking show the resulting selection directly.

```{r proportional-selection-preview}
selected_individuals_by_trait |>
  group_by(Trait) |>
  slice_head(n = 5) |>
  ungroup() |>
  select(
    Trait, Rank, IND, FAM, Method,
    Selection_objective, Genomic_value, Selection_score,
    OOF_r, OOF_nominal_p_value, OOF_nominal_signal
  ) |>
  kable(
    digits = 5,
    caption = "Top five candidates in each positive-OOF trait ranking"
  )
```

## 11. Recurrence across traits

The next step organizes a tabular summary of the objects and results needed to interpret this section.

```{r recurrence-distribution-table}
# Assemble the summary table presented in this section.
recurrence_class_labels <- c(
  high_4plus = "High (>=4)",
  intermediate_3 = "Intermediate (3)",
  broad_2 = "Broad (2)",
  specific_1 = "Specific (1)",
  none_0 = "None (0)"
)

# Derive the structure used to prioritize or summarize candidates.
recurrence_distribution_display <- recurrence_distribution |>
  mutate(
    Class = unname(
      recurrence_class_labels[Recurrence_class]
    )
  ) |>
  select(
    N_traits,
    Class,
    N_individuals,
    Proportion_of_population
  )

kable(
  recurrence_distribution_display,
  digits = 4,
  caption = "Distribution of individuals by recurrence class"
)
```

The next step derives and presents the corresponding selection or prioritization rule from the calculated results.

```{r recurrent-candidates-table}
# Derive selection rankings or summaries from the predictive results.
main_multitrait_candidates <- multitrait_candidates |>
  filter(N_traits >= 3L) |>
  mutate(
    Class = unname(
      recurrence_class_labels[Recurrence_class]
    )
  ) |>
  select(
    IND,
    FAM,
    N_traits,
    Selection_frequency,
    Class,
    Traits,
    Best_rank,
    Mean_rank_present,
    ends_with("_rank")
  )

# Format the table for display in the tutorial.
kable(
  main_multitrait_candidates,
  digits = 4,
  caption = "Candidates with high or intermediate recurrence"
)
```

Recurrence summarizes, across traits with positive OOF correlation, how many
place each individual among the most favorable 20%. Recurrence that depends on
weak positive OOF signals should remain exploratory. `N_traits`, selection
frequency, and ranks are not combined into a multi-trait index.

## 12. Summary

The tables and figures in this module follow the complete sequence from method
comparison to recurrent candidate identification. Biological interpretation
should jointly consider predictive evidence, breeding direction, and the
limitations of the population and marker set.
