---
title: "Presentation of results"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

This module turns the results derived in Module 04 into a visual and tabular synthesis for scientific reading. The sequence follows the analytical reasoning: first compare the seven methods, then present the trait-specific decisions, illustrate MCMC diagnostics and across-fold stability, and finally show exploratory genomic rankings and recurrence across traits. No table or figure on this page constitutes a new selection or inferential step; they present results already defined upstream.

## 2. Packages, results, and graphical style

The packages used here have distinct roles:

- `dplyr`: organizing, joining, and transforming result tables;
- `ggplot2`: building comparison and diagnostic figures;
- `knitr`: formatting the tables presented on the page;
- `here`: building project-relative paths.

```{r packages, warning=FALSE, message=FALSE}
library(dplyr)  # Organize and transform result tables.
library(ggplot2) # Build the synthesis figures.
library(knitr)  # Format tables in R Markdown.
library(here)   # Build project-relative paths.
```

```{r table-formatting, include=FALSE}
format_table <- function(x, caption = NULL, digits = getOption("digits"), col.names = NA) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    col.names = col.names,
    table.attr = paste0(
      'class="table table-striped table-hover" ',
      'style="width:100%; margin-bottom:0;"'
    )
  )
  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

The module consumes two consolidated objects. `models_object.rds` supplies the settings required to locate and interpret previously saved chains; `module04_results.rds` supplies all tables, decisions, rankings, and correlations derived in Module 04. Labels, colors, base font size, dimensions, and export resolution are presentation choices only and do not enter performance, selection, or inferential calculations. The fixed palette keeps each method visually consistent throughout the page.

```{r inputs-and-style}
# Define output paths and load the objects required by this module.
MODELS_FILE <- here("output", "models", "models_object.rds")
M04_FILE <- here("output", "results", "m04", "module04_results.rds")
RESULTS_DIR <- here("output", "results", "m05")
FIGURE_DIR <- file.path(RESULTS_DIR, "figures")

dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)

models_object <- readRDS(MODELS_FILE)
m04 <- readRDS(M04_FILE)

traits <- m04$traits
methods <- m04$methods
model_comparison <- m04$model_comparison
model_selection <- m04$model_selection
selected_fold_diagnostics <- m04$selected_fold_diagnostics
runtime_summary <- m04$runtime
# Retrieve the heritability summary derived in M04.
rrblup_heritability_summary <- m04$rrblup_heritability_summary
selection_objectives <- m04$selection_objectives
selection_proportion <- m04$selection_proportion
positive_oof_traits <- m04$positive_oof_traits
selection_sizes <- m04$selection_sizes
selected_individuals_by_trait <- m04$selected_individuals_by_trait
multitrait_recurrence <- m04$multitrait_recurrence
recurrence_distribution <- m04$recurrence_distribution
multitrait_candidates <- m04$multitrait_candidates
selected_oof_correlations <- m04$selected_oof_correlations

# Define readable method labels used in tables and figures.
method_labels <- c(
  RRBLUP = "RR-BLUP/BRR",
  BayesA = "BayesA",
  BayesB = "BayesB",
  BayesB2 = "BayesB2",
  BayesC = "BayesC",
  BL = "Bayesian Lasso",
  GBLUP = "GBLUP"
)

selection_objective_labels <- c(
  increase = "Increase",
  decrease = "Decrease"
)

# Derive statistics used by the interpretive narrative and validate
# their consistency with the scientific objects consolidated in Module 04.
n_traits <- length(traits)
n_methods <- length(methods)
n_trait_method_combinations <- nrow(model_comparison)
n_selected_traits <- nrow(model_selection)

positive_oof_from_selection <- model_selection$Trait[
  !is.na(model_selection$r) &
    model_selection$r > 0
]

nonpositive_oof_traits <- setdiff(
  traits,
  positive_oof_from_selection
)

positive_oof_nonsignificant_traits <- model_selection$Trait[
  !is.na(model_selection$r) &
    model_selection$r > 0 &
    !is.na(model_selection$p_value) &
    model_selection$p_value >= 0.05
]

n_positive_oof_traits <- length(positive_oof_from_selection)

n_available_values <- unique(selection_sizes$N_available)
n_selected_values <- unique(selection_sizes$N_selected)

highest_oof_trait <- model_comparison$Trait[
  which.max(model_comparison$r)
]

format_list_en <- function(x) {
  x <- as.character(x)
  n <- length(x)

  if (n == 0L) {
    return("none")
  }

  if (n == 1L) {
    return(x)
  }

  paste0(
    paste(x[-n], collapse = ", "),
    " and ",
    x[n]
  )
}

method_order_present <- methods[
  methods %in% model_selection$Selected_method
]

selected_method_summary <- paste(
  vapply(
    method_order_present,
    function(method_name) {
      paste0(
        unname(method_labels[method_name]),
        " for ",
        format_list_en(
          model_selection$Trait[
            model_selection$Selected_method == method_name
          ]
        )
      )
    },
    character(1)
  ),
  collapse = "; "
)

stopifnot(
  length(n_available_values) == 1L,
  length(n_selected_values) == 1L,
  n_trait_method_combinations == n_traits * n_methods,
  n_selected_traits == n_traits,
  !anyDuplicated(model_selection$Trait),
  setequal(model_selection$Trait, traits),
  all(!is.na(model_selection$r)),
  setequal(
    positive_oof_traits,
    positive_oof_from_selection
  ),
  setequal(
    selection_sizes$Trait,
    positive_oof_traits
  ),
  all(
    abs(
      selection_sizes$Selection_proportion -
        selection_proportion
    ) < sqrt(.Machine$double.eps)
  ),
  all(
    model_selection$Selected_method %in% names(method_labels)
  ),
  all(
    model_selection$Selection_status ==
      "rule_based_selected"
  ),
  nrow(selected_individuals_by_trait) ==
    sum(selection_sizes$N_selected),
  all(multitrait_candidates$N_traits >= 2L),
  nrow(multitrait_candidates) ==
    sum(
      recurrence_distribution$N_individuals[
        recurrence_distribution$N_traits >= 2L
      ]
    )
)

n_available <- n_available_values[[1]]
n_selected_per_trait <- n_selected_values[[1]]
n_trait_candidate_occurrences <-
  nrow(selected_individuals_by_trait)
n_recurrent_candidates <- nrow(multitrait_candidates)

# Define the fixed palette used to distinguish methods in figures.
method_colors <- c(
  RRBLUP = "#3366CC",
  BayesA = "#DC3912",
  BayesB = "#FF9900",
  BayesB2 = "#109618",
  BayesC = "#990099",
  BL = "#0099C6",
  GBLUP = "#DD4477"
)

# Use base font size 12 for readability; this choice does not alter any analytical result.
project_theme <- theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )
```

The opening table acts as a coverage check: it confirms how many traits and methods are presented, how many combinations enter the comparison, and how many traits have a selected method and positive OOF correlation. These counts verify the scope of the synthesis before the following panels are interpreted.

```{r input-summary}
data.frame(
  Item = c(
    "Traits",
    "Methods",
    "Rows in method comparison",
    "Traits with a selected method",
    "Traits with positive OOF correlation"
  ),
  Value = c(
    length(traits),
    length(methods),
    nrow(model_comparison),
    nrow(model_selection),
    length(positive_oof_traits)
  )
) |>
  format_table(caption = "Structure of the results presented in this module")
```

The table confirms complete synthesis coverage: `r n_traits` traits, `r n_methods` methods, and `r n_trait_method_combinations` trait × method combinations. All `r n_selected_traits` traits have a selected method and `r n_positive_oof_traits` have positive OOF correlation. Traits outside this positive set are: `r format_list_en(nonpositive_oof_traits)`.

## 3. Comparison of the seven methods

For every trait × method combination, the table reports mean DIC across folds, `Delta_DIC`, `Wprob`, `ER`, pooled OOF correlation, and the nominal p-value. DIC and its relative weights should be compared only among methods for the same trait; OOF correlation represents predictive performance and remains the primary reference. The p-value is descriptive only, as defined in Module 04.

```{r model-comparison-table}
model_comparison_display <- model_comparison |>
  mutate(Method_label = unname(method_labels[Method])) |>
  transmute(
    Trait,
    Method,
    Method_label,
    DIC = Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    r,
    p_value
  ) |>
  arrange(match(Trait, traits), match(Method, methods))

format_table(
  model_comparison_display |>
    select(Trait, Method_label, DIC, Delta_DIC, Wprob, ER, r, p_value),
  digits = 5,
  caption = "Comparison of the seven methods by trait"
)
```

The figure translates the OOF-correlation table into a direct visual comparison. The horizontal zero line separates positive from nonpositive association, and every panel uses the same vertical scale so magnitudes can be compared across traits. Bars are point estimates, not uncertainty intervals.

```{r predictive-correlation-figure, fig.width=13, fig.height=10}
# Order methods and build the OOF-correlation figure by trait.
predictive_plot <- model_comparison |>
  mutate(Method = factor(Method, levels = methods)) |>
  ggplot(aes(x = Method, y = r, fill = Method)) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.4) +
  geom_col(width = 0.75) +
  facet_wrap(~ Trait, scales = "fixed", ncol = 2) +
  scale_fill_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  scale_x_discrete(labels = method_labels) +
  labs(
    x = NULL,
    y = "OOF correlation",
    fill = "Method",
    title = "Predictive ability of the seven methods"
  ) +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

predictive_plot

ggsave(
  file.path(FIGURE_DIR, "predictive_correlation_by_method_en.png"),
  predictive_plot,
  width = 13,
  height = 10,
  dpi = 300
)
```

The comparison shows that predictive magnitude depends strongly on the trait. The largest OOF correlation observed in the comparison occurs for `r highest_oof_trait`, whereas several traits remain in much lower ranges. Within several panels, methods are relatively close for the same trait; method choice therefore remains trait-specific and follows the hierarchical rule consolidated in Module 04 rather than defining one global method.

## 4. Selected methods by trait

The following table does not rerun method selection. It translates the decision already consolidated in Module 04, placing the selected method, predictive winner, DIC winner, any RR-BLUP preference, decision status, and main OOF/MCMC indicators side by side. A provisional status still requires review even though a method appears in the selected-method column.

```{r selected-methods-table}
final_results_summary <- model_selection |>
  mutate(
    Method_label = unname(method_labels[Selected_method]),
    Predictive_winner_label = unname(method_labels[Predictive_winner]),
    DIC_winner_label = unname(method_labels[DIC_winner]),
    Decision_label = case_when(
      Decision_code == "rrblup_best_predictive_model" ~
        "RR-BLUP/BRR was the best predictive model",
      Decision_code == "rrblup_preferred_among_competitive_models" ~
        "RR-BLUP/BRR preferred among competitive methods",
      Decision_code == "best_non_rrblup_competitive_model" ~
        "Best competitive non-RR-BLUP method",
      Decision_code == "predictive_winner_provisional_review_required" ~
        "Provisional predictive winner; review required",
      TRUE ~ Decision_code
    ),
    Selection_status_label = case_when(
      Selection_status == "rule_based_selected" ~
        "Rule-based selection",
      Selection_status == "provisional_review_required" ~
        "Provisional selection; review required",
      TRUE ~ Selection_status
    )
  ) |>
  transmute(
    Trait,
    Method_label,
    Decision_label,
    Selection_status_label,
    RRBLUP_preference_applied,
    Predictive_winner = Predictive_winner_label,
    DIC_winner = DIC_winner_label,
    Predictive_loss,
    r,
    p_value,
    RMSE,
    DIC = Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    Parameters_flagged,
    Runs_with_flags,
    OOF_predictive_direction
  ) |>
  arrange(match(Trait, traits))

format_table(
  final_results_summary,
  digits = 5,
  caption = "Selected methods and diagnostics by trait"
)
```

The consolidated decision selects `r selected_method_summary`. All choices remain `Selection_status = rule_based_selected`. Traits with nonpositive OOF correlation — `r format_list_en(nonpositive_oof_traits)` — do not enter the exploratory candidate rankings.

## 5. MCMC chains of the selected methods

For each trait, this section shows only the posterior chain of \(\mu\) from fold 1 of the selected method, its running mean, and its posterior density. Fold 1 provides a consistent visual illustration across traits; it does not replace ESS/Geweke diagnostics over all five folds and does not prove convergence of every model parameter.

`models_object$settings$burn_in / models_object$settings$thin` converts burn-in from the iteration scale to the saved-sample scale. `setNames(model_selection$Selected_method, model_selection$Trait)` creates a named lookup that directly retrieves the selected method for each trait. `vector("list", length(traits))` preallocates the result lists, and `seq_along(traits)` traverses the ten trait indices. Inside the loop, `scan()` reads the saved `mu` chain; `seq.int(burn_in_saved + 1L, length(mu_full))` retains only post-burn-in samples. Iteration numbers are reconstructed with `seq.int(..., by = thin)`. `cumsum(posterior_mu) / seq_along(posterior_mu)` computes the running mean from the full posterior chain, whereas `seq.int(..., by = 100L)` reduces only the points drawn. Finally, `density(posterior_mu, n = 512)` estimates the marginal density used in the figure; `n = 512` controls curve resolution, not the number of MCMC samples.

```{r selected-mu-data}
# Convert burn-in to the saved-sample scale and prepare posterior mu traces and densities.
burn_in_saved <-
  models_object$settings$burn_in /
  models_object$settings$thin

selected_method_lookup <- setNames(
  model_selection$Selected_method,
  model_selection$Trait
)

trace_parts <- vector("list", length(traits))
density_parts <- vector("list", length(traits))

for (trait_index in seq_along(traits)) {
  # Identify the trait and selected method.
  trait_name <- traits[trait_index]
  method_name <- selected_method_lookup[trait_name]

  # Locate the mu chain from the first fold.
  mu_file <- file.path(
    here("output", "models", "bglr"),
    trait_name,
    "F1",
    method_name,
    "bglr_mu.dat"
  )

  mu_full <- scan(mu_file, quiet = TRUE)

  # Remove samples corresponding to burn-in.
  posterior_mu <- mu_full[
    seq.int(burn_in_saved + 1L, length(mu_full))
  ]

  iteration <- seq.int(
    from = models_object$settings$burn_in + models_object$settings$thin,
    by = models_object$settings$thin,
    length.out = length(posterior_mu)
  )

  # Compute the running mean and reduce point density for plotting.
  running_mean <- cumsum(posterior_mu) / seq_along(posterior_mu)
  keep_index <- seq.int(1L, length(posterior_mu), by = 100L)

  trace_parts[[trait_index]] <- data.frame(
    Trait = trait_name,
    Method = method_name,
    Facet = paste0(trait_name, " — ", method_labels[method_name]),
    Iteration = iteration[keep_index],
    Posterior_mu = posterior_mu[keep_index],
    Running_mean = running_mean[keep_index]
  )

  # Estimate the posterior density of mu for the selected method in this trait.
  density_object <- density(posterior_mu, n = 512)

  density_parts[[trait_index]] <- data.frame(
    Trait = trait_name,
    Method = method_name,
    Facet = paste0(trait_name, " — ", method_labels[method_name]),
    x = density_object$x,
    y = density_object$y
  )
}

selected_mu_trace <- bind_rows(trace_parts)
selected_mu_density <- bind_rows(density_parts)
```

In the trace plot, the thin line shows posterior samples of \(\mu\) after burn-in and the colored line shows the running mean. For readability, only one of every 100 posterior samples is drawn; this reduction affects plotting only, while the running mean was computed from the full available posterior chain. Visual stabilization of the mean is descriptive evidence, not a standalone convergence criterion.

```{r selected-mu-trace-figure, fig.width=13, fig.height=14}
mu_trace_plot <- ggplot(
  selected_mu_trace,
  aes(x = Iteration, y = Posterior_mu)
) +
  geom_line(linewidth = 0.25, alpha = 0.35) +
  geom_line(
    aes(y = Running_mean, color = Method),
    linewidth = 0.8
  ) +
  facet_wrap(~ Facet, scales = "free_y", ncol = 2) +
  scale_color_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  labs(
    x = "Iteration",
    y = expression(mu),
    color = "Method",
    title = "Posterior chain and running mean of mu — fold 1"
  ) +
  project_theme

mu_trace_plot

ggsave(
  file.path(FIGURE_DIR, "selected_model_mu_trace_en.png"),
  mu_trace_plot,
  width = 13,
  height = 14,
  dpi = 300
)
```

The density complements the trace by showing the marginal posterior shape of \(\mu\) in the same fold. `n = 512` controls only the numerical resolution of the density curve and does not alter the chain or inference. Panels use free scales because the location and dispersion of \(\mu\) differ across traits; visual widths across panels should therefore not be compared as though their axes were common.

```{r selected-mu-density-figure, fig.width=13, fig.height=14}
mu_density_plot <- ggplot(
  selected_mu_density,
  aes(x = x, y = y, color = Method)
) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ Facet, scales = "free", ncol = 2) +
  scale_color_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  labs(
    x = expression(mu),
    y = "Density",
    color = "Method",
    title = "Posterior density of mu — fold 1"
  ) +
  project_theme

mu_density_plot

ggsave(
  file.path(FIGURE_DIR, "selected_model_mu_density_en.png"),
  mu_density_plot,
  width = 13,
  height = 14,
  dpi = 300
)
```

The trace and density panels are therefore illustrations of the `mu` chain for the selected methods in fold 1. No selection or convergence classification is changed by these figures; convergence assessment remains based on the ESS and Geweke diagnostics obtained across all five folds in Module 03 and summarized downstream.

## 6. Stability across folds

This figure returns the focus to cross-validation. Each panel shows the five validation correlations for the selected method; sign changes or large ranges among points suggest descriptive instability. Because training sets overlap and there are only five folds, the plot is neither a formal heterogeneity test nor a confidence interval.

```{r selected-fold-figure, fig.width=12, fig.height=10}
# Add labels to fold-level results and build the stability figure for the selected method.
fold_stability_display <- selected_fold_diagnostics |>
  mutate(
    Method_label = unname(method_labels[Method]),
    Facet = paste0(Trait, " — ", Method_label)
  )

# Build the plot of predictive correlation observed in each fold for the selected method.
fold_stability_plot <- ggplot(
  fold_stability_display,
  aes(x = Fold, y = Correlation, group = Trait)
) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.4) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~ Facet, ncol = 2) +
  scale_x_continuous(breaks = 1:5) +
  coord_cartesian(ylim = c(-1, 1)) +
  labs(
    x = "Fold",
    y = "Validation correlation",
    title = "Predictive correlation across folds for selected methods"
  ) +
  project_theme

fold_stability_plot

ggsave(
  file.path(FIGURE_DIR, "selected_method_fold_stability_en.png"),
  fold_stability_plot,
  width = 12,
  height = 10,
  dpi = 300
)
```

Fold stability should be interpreted together with pooled OOF correlation. Larger ranges or sign changes are descriptively more relevant for traits whose pooled signal is already weak, but they do not redefine method selection. With only five folds and overlapping training sets, this variation remains a descriptive stability measure.

## 7. Computational time

Bars show the mean recorded time across the 50 fits of each method, and error bars show one standard deviation across those fits. The comparison is operational for this execution and environment; hardware, system load, implementation, and I/O differences prevent interpreting the plot as a universal speed benchmark.

In the code, `factor(Method, levels = methods)` fixes only the visual ordering of methods. `pmax(0, Mean_seconds - SD_seconds)` prevents the lower end of the graphical error bar from being drawn below zero seconds; the original mean and standard deviation remain unchanged.

```{r runtime-figure, fig.width=11, fig.height=7}
# Summarize the computational cost of the evaluated methods.
runtime_plot <- runtime_summary |>
  mutate(Method = factor(Method, levels = methods)) |>
  ggplot(aes(x = Method, y = Mean_seconds, fill = Method)) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = pmax(0, Mean_seconds - SD_seconds),
      ymax = Mean_seconds + SD_seconds
    ),
    width = 0.2
  ) +
  scale_fill_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  scale_x_discrete(labels = method_labels) +
  labs(
    x = NULL,
    y = "Mean time (seconds)",
    fill = "Method",
    title = "Computational time by method"
  ) +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

runtime_plot

ggsave(
  file.path(FIGURE_DIR, "runtime_by_method_en.png"),
  runtime_plot,
  width = 11,
  height = 7,
  dpi = 300
)
```

The runtime ordering documents only the computational cost observed in this execution. Fitting time does not enter the method-selection rule and should not replace predictive, DIC, or MCMC-diagnostic criteria.

## 8. Marker-based heritability analogue and adjusted accuracy in RR-BLUP/BRR

The table re-presents the quantities derived in Module 04 for RR-BLUP/BRR. `RRBLUP_H2` is a marker-model-based heritability analogue, not an independent classical estimate of narrow-sense heritability. `RRBLUP_predictive_accuracy` is the mean ratio \(r_f/\sqrt{H_f^2}\); because it is derived from sample estimates, it may be negative or exceed 1 and should not be interpreted as a correlation necessarily bounded to `[0, 1]`.

```{r rrblup-heritability-table}
rrblup_table <- rrblup_heritability_summary |>
  transmute(
    Trait,
    RRBLUP_H2 = Mean_H2,
    SD_RRBLUP_H2 = SD_H2,
    RRBLUP_predictive_accuracy = Mean_predictive_accuracy,
    SD_RRBLUP_predictive_accuracy = SD_predictive_accuracy
  )

format_table(
  rrblup_table,
  digits = 5,
  caption = "Marker-based heritability analogue and adjusted predictive accuracy in RR-BLUP/BRR"
)
```

These quantities contextualize the variability captured by RR-BLUP/BRR and the relationship between predictive correlation and the heritability analogue. They do not rerank the already selected methods and should not be converted into conclusions about classical population heritability outside this model.

## 9. Correlations among OOF genomic values

The heatmap presents associations among OOF genomic values produced by the selected method for each trait. Because different traits may use different methods and the values are cross-fitted, these associations are not genetic correlations estimated by a multivariate model. Correlations are used without rounding in calculations and in the color scale; only labels with `|r| < 0.005` are displayed as zero to avoid `-0.00`, without changing the numeric value or fill color.

`as.table(selected_oof_correlations)` converts the 10 × 10 matrix into row-column-value combinations; `as.data.frame()` converts that structure to long form, and `names()` assigns the three column names. `replace(Correlation, abs(Correlation) < 0.005, 0)` acts only on the label column. The `fill = Correlation` aesthetic continues to use the original correlation, while `sprintf("%.2f", Correlation_label)` controls only the text printed in each cell.

```{r oof-correlation-figure, fig.width=10, fig.height=9}
# Reshape the OOF correlation matrix and build the cross-trait heatmap.
correlation_long <- as.data.frame(
  as.table(selected_oof_correlations)
)
names(correlation_long) <- c(
  "Trait_1",
  "Trait_2",
  "Correlation"
)

# Set only the label of |r| < 0.005 to zero; the original correlation still drives analysis and color.
correlation_long <- correlation_long |>
  mutate(
    Correlation_label = replace(
      Correlation,
      abs(Correlation) < 0.005,
      0
    )
  )

# Build the heatmap of correlations among OOF genomic values.
correlation_plot <- ggplot(
  correlation_long,
  aes(x = Trait_1, y = Trait_2, fill = Correlation)
) +
  geom_tile() +
  geom_text(
    aes(label = sprintf("%.2f", Correlation_label)),
    size = 3
  ) +
  scale_fill_gradient2(
    low = "#DC3912",
    mid = "white",
    high = "#3366CC",
    midpoint = 0,
    # Keep the color scale fixed to the theoretical correlation range [-1, 1].
    limits = c(-1, 1)
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "r",
    title = "Correlation among OOF genomic values"
  ) +
  coord_equal() +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

correlation_plot

ggsave(
  file.path(FIGURE_DIR, "selected_oof_correlations_en.png"),
  correlation_plot,
  width = 10,
  height = 9,
  dpi = 300
)
```

The heatmap should be read as association among cross-fitted OOF scores: positive values indicate a tendency for two traits to rank individuals in the same direction, whereas negative values indicate opposing directions. These associations are not genetic correlations from a multivariate model and do not create an additional selection criterion.

## 10. Exploratory proportional candidate ranking by trait

Rankings are shown only for traits whose selected method has positive OOF correlation. Nominal p-values are descriptive and do not control display; when `p >= 0.05`, the ranking remains explicitly exploratory and does not demonstrate predictive utility. The proportion shown is `r sprintf("%.0f%%", 100 * selection_proportion)` per trait — `r n_selected_per_trait` of the `r n_available` individuals in this dataset — and is a configurable rule in this analysis rather than an estimated optimal intensity. Scores are cross-fitted OOF genomic values, so an individual's own phenotype does not contribute to its score; even so, they are not final breeding values from a model refitted to all phenotypes.

The ranking-size table is built with two joins. The first `left_join()` adds each trait's selection objective; the second adds `r`, nominal p-value, and the OOF-signal classification. Because both are `left_join()`, traits already present in `selection_sizes` remain the reference set for presentation. The p-value is displayed but is not used to remove rankings.

```{r proportional-selection-table}
# Combine the selection objective, OOF evidence, and exploratory ranking size for each trait.
selection_sizes_display <- selection_sizes |>
  left_join(
    selection_objectives |>
      select(Trait, Selection_objective),
    by = "Trait"
  ) |>
  left_join(
    model_selection |>
      select(Trait, r, p_value, OOF_nominal_signal),
    by = "Trait"
  ) |>
  mutate(
    Objective = unname(
      selection_objective_labels[Selection_objective]
    )
  ) |>
  select(
    Trait,
    Objective,
    r,
    p_value,
    OOF_nominal_signal,
    Selection_proportion,
    N_available,
    N_selected
  )

format_table(
  selection_sizes_display,
  digits = 2,
  caption = "Exploratory candidate-ranking summary by positive-OOF trait"
)
```

The next table shows only the first five individuals from each ranking as a reading preview. The value `n = 5` is a presentation choice; the complete ranking remains available in the object and derived files.

In the preview, `group_by(Trait)` separates rankings by trait, `slice_head(n = 5)` keeps only the first five rows of each group, and `ungroup()` ends grouping before the displayed columns are selected. The complete `selected_individuals_by_trait` object is not truncated.

```{r proportional-selection-preview}
selected_individuals_by_trait |>
  group_by(Trait) |>
  slice_head(n = 5) |>
  ungroup() |>
  select(
    Trait, Rank, IND, FAM, Method,
    Selection_objective, Genomic_value, Selection_score,
    OOF_r, OOF_nominal_p_value, OOF_nominal_signal
  ) |>
  format_table(
    digits = 5,
    caption = "Top five candidates in each positive-OOF trait ranking"
  )
```

The `r n_positive_oof_traits` traits with positive OOF correlation produce rankings, each containing `r n_selected_per_trait` of the `r n_available` individuals, for `r n_trait_candidate_occurrences` trait × candidate occurrences before recurrence is considered. Traits outside the positive-OOF set are: `r format_list_en(nonpositive_oof_traits)`. Among traits with positive OOF direction, those with nominal p-value `>= 0.05` are: `r format_list_en(positive_oof_nonsignificant_traits)`; these signals therefore remain explicitly exploratory.

## 11. Recurrence across traits

The recurrence classes shown here are the same ones defined in Module 04. The complete object retains candidates present in at least two traits; to keep the main table compact, this module displays only intermediate or high recurrence, that is, candidates present in three or more traits. The `N_traits >= 3` cutoff is a presentation choice only and does not redefine the stored candidate set.

The first recurrence table describes the whole population according to the number of positive-OOF rankings in which each individual appears. The classes are presentation labels defined in Module 04 rather than significance thresholds.

`recurrence_class_labels[Recurrence_class]` translates the internal class codes without recalculating `N_traits`. In the following table, `filter(N_traits >= 3L)` reduces only the main presentation. The `multitrait_candidates` object received from Module 04 remains defined by recurrence in at least two traits.

```{r recurrence-distribution-table}
# Translate recurrence classes and prepare their distribution for presentation.
recurrence_class_labels <- c(
  high_4plus = "High (>=4)",
  intermediate_3 = "Intermediate (3)",
  broad_2 = "Broad (2)",
  specific_1 = "Specific (1)",
  none_0 = "None (0)"
)

# Translate recurrence classes and compute the distribution displayed in the table.
recurrence_distribution_display <- recurrence_distribution |>
  mutate(
    Class = unname(
      recurrence_class_labels[Recurrence_class]
    )
  ) |>
  select(
    N_traits,
    Class,
    N_individuals,
    Proportion_of_population
  )

format_table(
  recurrence_distribution_display,
  digits = 4,
  caption = "Distribution of individuals by recurrence class"
)
```

The main candidate table is deliberately more compact: it shows only individuals recurring in three or more traits. This filter is for presentation only; the complete `multitrait_candidates` object still retains recurrence in at least two traits.

```{r recurrent-candidates-table}
# Show only intermediate/high recurrence (>=3) in the main table; the complete object retains >=2.
main_multitrait_candidates <- multitrait_candidates |>
  filter(N_traits >= 3L) |>
  mutate(
    Class = unname(
      recurrence_class_labels[Recurrence_class]
    )
  ) |>
  select(
    IND,
    FAM,
    N_traits,
    Selection_frequency,
    Class,
    Traits,
    Best_rank,
    Mean_rank_present,
    ends_with("_rank")
  )

format_table(
  main_multitrait_candidates,
  digits = 4,
  caption = "Candidates with high or intermediate recurrence"
)
```

Recurrence summarizes, across traits with positive OOF correlation, in how many lists each individual appears among the most favorable `r sprintf("%.0f%%", 100 * selection_proportion)`. `Selection_frequency` uses the number of positive-OOF traits as its denominator, not necessarily all `r n_traits` original traits. Recurrence supported by weak positive OOF signals remains exploratory. `N_traits`, frequency, and ranks are not combined into a multi-trait index; therefore, greater recurrence does not demonstrate joint superiority across multiple traits.

In the complete Module 04 object, `r n_recurrent_candidates` individuals recur in at least two traits. The main table on this page shows only the subset recurring in three or more traits; this presentation reduction does not modify the stored scientific set or create a multi-trait index.

## 12. Summary

The final reading should preserve the hierarchy established throughout the analysis: OOF performance drives predictive comparison; DIC, across-fold stability, and MCMC diagnostics provide complementary context; rankings are produced only when OOF direction is positive; and recurrence summarizes overlap among rankings without constructing a multi-trait index. All conclusions remain conditional on the represented breeding population, the available SSR panel, and the transductive validation design. For operational selection, OOF rankings should be treated as exploratory evidence rather than substitutes for final breeding values from a prediction stage designed for decision making.

<div style="display:flex; justify-content:space-between; gap:1rem; margin-top:2rem; padding-top:1rem; border-top:1px solid #ddd;">
<a href="04_results_en.html">← Module 04 — Comparison and selection</a>
<a href="index.html">Home →</a>
</div>
