---
title: "Presentation of final results"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Universidade Federal de Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

This module gathers and presents the final results produced in Modules 03 and
04. No model is refitted.

The organization follows the presentation structure of Silva et al.
[@silva2021brrguava], adapted to the seven methods evaluated in this project
and to the previously defined cross-validation design.

The explanations on this page are descriptive and direct. The scientific
discussion of the results will be developed separately in the manuscript.

## 2. Packages, inputs, and graphical standard

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
  library(here)
})
```

```{r inputs-and-style}
MATRICES_FILE <- here(
  "output",
  "matrices",
  "matrices_object.rds"
)

MODELS_FILE <- here(
  "output",
  "models",
  "models_object.rds"
)

M04_FILE <- here(
  "output",
  "results",
  "m04",
  "module04_results.rds"
)

RESULTS_DIR <- here(
  "output",
  "results",
  "m05"
)

FIGURE_DIR <- file.path(
  RESULTS_DIR,
  "figures"
)

dir.create(
  RESULTS_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  FIGURE_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

matrices_object <- readRDS(
  MATRICES_FILE
)

models_object <- readRDS(
  MODELS_FILE
)

m04 <- readRDS(
  M04_FILE
)

model_comparison <- m04$model_comparison
model_selection <- m04$model_selection
runtime_summary <- m04$runtime
rrblup_heritability_summary <- m04$rrblup_heritability_summary
genomic_ranking <- m04$genomic_ranking
selected_genetic_correlations <- m04$selected_genetic_correlations

metadata <- matrices_object$metadata

traits <- colnames(
  matrices_object$phenotypes
)

methods <- c(
  "RRBLUP",
  "BayesA",
  "BayesB",
  "BayesB2",
  "BayesC",
  "BL",
  "GBLUP"
)

method_labels <- c(
  RRBLUP = "RR-BLUP/BRR",
  BayesA = "BayesA",
  BayesB = "BayesB",
  BayesB2 = "BayesB2",
  BayesC = "BayesC",
  BL = "Bayesian Lasso",
  GBLUP = "GBLUP"
)

method_colors <- c(
  RRBLUP = "#3366CC",
  BayesA = "#DC3912",
  BayesB = "#FF9900",
  BayesB2 = "#109618",
  BayesC = "#990099",
  BL = "#0099C6",
  GBLUP = "#DD4477"
)

project_theme <-
  theme_bw(
    base_size = 12
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(
      face = "bold"
    ),
    plot.title = element_text(
      face = "bold"
    ),
    axis.text.x = element_text(
      angle = 0,
      vjust = 0.5
    )
  )
```

```{r input-audit}
input_audit <- data.frame(
  Check = c(
    "M04 signature matches M03",
    "70 trait × method combinations",
    "10 final trait-specific methods",
    "7 methods in runtime summary",
    "10 traits in RR-BLUP/BRR summary",
    "1,500 genomic-ranking rows",
    "10 × 10 correlation matrix"
  ),
  Passed = c(
    identical(
      m04$signature,
      models_object$signature
    ),
    nrow(
      model_comparison
    ) == 70L,
    nrow(
      model_selection
    ) == 10L,
    nrow(
      runtime_summary
    ) == 7L,
    nrow(
      rrblup_heritability_summary
    ) == 10L,
    nrow(
      genomic_ranking
    ) == 1500L,
    identical(
      dim(
        selected_genetic_correlations
      ),
      c(
        10L,
        10L
      )
    )
  )
)

kable(
  input_audit,
  caption = "Audit of the inputs used in Module 05"
)
```

## 3. Comparison of the seven methods

The table below follows the main comparison structure used by Silva et al.
[@silva2021brrguava]: DIC, difference from the minimum DIC, approximate model
probability (`Wprob`), evidence ratio (`ER`), predictive correlation, and
p-value.

In this project, the comparison is performed jointly across the seven fitted
methods.

```{r article-style-model-table}
article_model_table <- model_comparison |>
  transmute(
    Trait,
    Method,
    Method_label,
    DIC = Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    r,
    p_value
  ) |>
  arrange(
    match(
      Trait,
      traits
    ),
    match(
      Method,
      methods
    )
  )

kable(
  article_model_table |>
    select(
      Trait,
      Method_label,
      DIC,
      Delta_DIC,
      Wprob,
      ER,
      r,
      p_value
    ),
  digits = 5,
  caption = "Comparison of the seven methods by trait"
)
```

### 3.1. Predictive ability

```{r predictive-correlation-figure, fig.width=13, fig.height=10}
predictive_plot <- model_comparison |>
  mutate(
    Method = factor(
      Method,
      levels = methods
    )
  ) |>
  ggplot(
    aes(
      x = Method,
      y = r,
      fill = Method
    )
  ) +
  geom_hline(
    yintercept = 0,
    linetype = 2,
    linewidth = 0.4
  ) +
  geom_col(
    width = 0.75
  ) +
  facet_wrap(
    ~ Trait,
    scales = "free_y",
    ncol = 2
  ) +
  scale_fill_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  scale_x_discrete(
    labels = method_labels
  ) +
  labs(
    x = NULL,
    y = "OOF correlation",
    fill = "Method",
    title = "Predictive ability of the seven methods"
  ) +
  project_theme +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

predictive_plot

ggsave(
  filename = file.path(
    FIGURE_DIR,
    "predictive_correlation_by_method_en.png"
  ),
  plot = predictive_plot,
  width = 13,
  height = 10,
  dpi = 300
)
```

## 4. Selected methods by trait

```{r selected-methods-table}
selected_methods_table <- model_selection |>
  transmute(
    Trait,
    Method = Selected_method,
    Method_label = Selected_method_label,
    r,
    p_value,
    DIC = Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    Parameters_flagged,
    Runs_with_flags,
    Predictive_evidence
  ) |>
  arrange(
    match(
      Trait,
      traits
    )
  )

kable(
  selected_methods_table |>
    select(
      Trait,
      Method_label,
      r,
      p_value,
      DIC,
      Delta_DIC,
      Wprob,
      ER,
      Parameters_flagged,
      Runs_with_flags,
      Predictive_evidence
    ),
  digits = 5,
  caption = "Final method and associated results for each trait"
)
```

RR-BLUP/BRR was selected for
`r sum(model_selection$Selected_method == "RRBLUP")` traits and the Bayesian
Lasso for `r sum(model_selection$Selected_method == "BL")`. BayesB, BayesC,
and GBLUP were each selected for one trait.

Of the ten traits,
`r sum(model_selection$Predictive_evidence == "positive_p_lt_0.05")`
showed positive OOF correlation with p < 0.05.

## 5. MCMC chains of the selected methods

As a visual representation analogous to that presented in the reference
article, the posterior chain of \(\mu\), its running mean, and the posterior
density are shown for the first fold of the selected method for each trait.

The complete numerical convergence diagnostics remain documented in Module 03.

```{r selected-mu-data}
burn_in_saved <-
  models_object$settings$burn_in /
  models_object$settings$thin

selected_method_lookup <- setNames(
  model_selection$Selected_method,
  model_selection$Trait
)

trace_parts <- vector(
  "list",
  length(
    traits
  )
)

density_parts <- vector(
  "list",
  length(
    traits
  )
)

for (trait_index in seq_along(
  traits
)) {
  trait_name <- traits[
    trait_index
  ]

  method_name <- selected_method_lookup[
    trait_name
  ]

  mu_file <- file.path(
    here(
      "output",
      "models",
      "bglr"
    ),
    trait_name,
    "F1",
    method_name,
    "bglr_mu.dat"
  )

  mu_full <- scan(
    mu_file,
    quiet = TRUE
  )

  posterior_mu <- mu_full[
    seq.int(
      burn_in_saved + 1L,
      length(
        mu_full
      )
    )
  ]

  iteration <- seq.int(
    from =
      models_object$settings$burn_in +
      models_object$settings$thin,
    by =
      models_object$settings$thin,
    length.out =
      length(
        posterior_mu
      )
  )

  running_mean <-
    cumsum(
      posterior_mu
    ) /
    seq_along(
      posterior_mu
    )

  keep_index <- seq.int(
    from = 1L,
    to = length(
      posterior_mu
    ),
    by = 100L
  )

  trace_parts[[
    trait_index
  ]] <- data.frame(
    Trait = trait_name,
    Method = method_name,
    Facet = paste0(
      trait_name,
      " — ",
      method_labels[
        method_name
      ]
    ),
    Iteration =
      iteration[
        keep_index
      ],
    Posterior_mu =
      posterior_mu[
        keep_index
      ],
    Running_mean =
      running_mean[
        keep_index
      ]
  )

  density_object <- density(
    posterior_mu,
    n = 512
  )

  density_parts[[
    trait_index
  ]] <- data.frame(
    Trait = trait_name,
    Method = method_name,
    Facet = paste0(
      trait_name,
      " — ",
      method_labels[
        method_name
      ]
    ),
    x = density_object$x,
    y = density_object$y
  )
}

selected_mu_trace <- bind_rows(
  trace_parts
)

selected_mu_density <- bind_rows(
  density_parts
)
```

```{r selected-mu-trace-figure, fig.width=13, fig.height=14}
mu_trace_plot <- ggplot(
  selected_mu_trace,
  aes(
    x = Iteration,
    y = Posterior_mu
  )
) +
  geom_line(
    linewidth = 0.25,
    alpha = 0.35
  ) +
  geom_line(
    aes(
      y = Running_mean,
      color = Method
    ),
    linewidth = 0.8
  ) +
  facet_wrap(
    ~ Facet,
    scales = "free_y",
    ncol = 2
  ) +
  scale_color_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  labs(
    x = "Iteration",
    y = expression(mu),
    color = "Method",
    title = "Posterior chain and running mean of mu — fold 1"
  ) +
  project_theme

mu_trace_plot

ggsave(
  filename = file.path(
    FIGURE_DIR,
    "selected_model_mu_trace_en.png"
  ),
  plot = mu_trace_plot,
  width = 13,
  height = 14,
  dpi = 300
)
```

```{r selected-mu-density-figure, fig.width=13, fig.height=14}
mu_density_plot <- ggplot(
  selected_mu_density,
  aes(
    x = x,
    y = y,
    color = Method
  )
) +
  geom_line(
    linewidth = 0.8
  ) +
  facet_wrap(
    ~ Facet,
    scales = "free",
    ncol = 2
  ) +
  scale_color_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  labs(
    x = expression(mu),
    y = "Density",
    color = "Method",
    title = "Posterior density of mu — fold 1"
  ) +
  project_theme

mu_density_plot

ggsave(
  filename = file.path(
    FIGURE_DIR,
    "selected_model_mu_density_en.png"
  ),
  plot = mu_density_plot,
  width = 13,
  height = 14,
  dpi = 300
)
```

## 6. Computational time

The times correspond to the 350 official fits already performed in Module 03.
No additional repetitions are run solely for timing purposes.

```{r runtime-table}
kable(
  runtime_summary |>
    select(
      Method_label,
      Fits,
      Total_seconds,
      Mean_seconds,
      SD_seconds,
      Median_seconds,
      Minimum_seconds,
      Maximum_seconds
    ),
  digits = 2,
  caption = "Observed computational time for the official fits"
)
```

```{r runtime-figure, fig.width=11, fig.height=7}
runtime_plot <- runtime_summary |>
  mutate(
    Method = factor(
      Method,
      levels = methods
    )
  ) |>
  ggplot(
    aes(
      x = Method,
      y = Mean_seconds,
      fill = Method
    )
  ) +
  geom_col(
    width = 0.7
  ) +
  geom_errorbar(
    aes(
      ymin = pmax(
        0,
        Mean_seconds - SD_seconds
      ),
      ymax =
        Mean_seconds + SD_seconds
    ),
    width = 0.2
  ) +
  scale_fill_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  scale_x_discrete(
    labels = method_labels
  ) +
  labs(
    x = NULL,
    y = "Mean time (seconds)",
    fill = "Method",
    title = "Computational time by method"
  ) +
  project_theme +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

runtime_plot

ggsave(
  filename = file.path(
    FIGURE_DIR,
    "runtime_by_method_en.png"
  ),
  plot = runtime_plot,
  width = 11,
  height = 7,
  dpi = 300
)
```

## 7. Heritability and accuracy in the RR-BLUP/BRR analogue

Following the structure of Table 3 in Silva et al.
[@silva2021brrguava], heritability and predictive accuracy are presented.

In this project, these values specifically constitute the **RR-BLUP/BRR**
analogue calculated in Module 04. They are not estimates produced by the
different trait-specific selected methods.

```{r rrblup-heritability-table}
rrblup_table <- rrblup_heritability_summary |>
  transmute(
    Trait,
    RRBLUP_H2 = Mean_H2,
    SD_RRBLUP_H2 = SD_H2,
    RRBLUP_mean_fold_r = Mean_fold_r,
    SD_RRBLUP_fold_r = SD_fold_r,
    RRBLUP_predictive_accuracy =
      Mean_predictive_accuracy,
    SD_RRBLUP_predictive_accuracy =
      SD_predictive_accuracy
  )

kable(
  rrblup_table,
  digits = 5,
  caption = "Heritability and predictive accuracy in the RR-BLUP/BRR analogue"
)
```

## 8. Correlation among OOF genomic values

The matrix below presents Pearson correlations among the OOF genomic values
obtained from the selected method for each trait.

```{r oof-correlation-data}
correlation_long <- as.data.frame(
  as.table(
    selected_genetic_correlations
  )
)

names(
  correlation_long
) <- c(
  "Trait_1",
  "Trait_2",
  "Correlation"
)
```

```{r oof-correlation-figure, fig.width=10, fig.height=9}
correlation_plot <- ggplot(
  correlation_long,
  aes(
    x = Trait_1,
    y = Trait_2,
    fill = Correlation
  )
) +
  geom_tile() +
  geom_text(
    aes(
      label = sprintf(
        "%.2f",
        Correlation
      )
    ),
    size = 3
  ) +
  scale_fill_gradient2(
    low = "#DC3912",
    mid = "white",
    high = "#3366CC",
    midpoint = 0,
    limits = c(
      -1,
      1
    )
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "r",
    title = "Correlation among OOF genomic values"
  ) +
  coord_equal() +
  project_theme +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

correlation_plot

ggsave(
  filename = file.path(
    FIGURE_DIR,
    "selected_oof_correlations_en.png"
  ),
  plot = correlation_plot,
  width = 10,
  height = 9,
  dpi = 300
)
```

The largest positive off-diagonal correlation was observed between DF and DCI,
with a value of 0.834.

## 9. Genomic values and ranking of individuals

Individuals are presented in decreasing order of OOF genomic value while
retaining the selected method for each trait.

The table shows the ten highest-ranked individuals for each trait and adds the
family (`FAM`) recorded in the dataset.

```{r top10-ranking}
top10_with_family <- genomic_ranking |>
  filter(
    Rank <= 10L
  ) |>
  left_join(
    metadata |>
      select(
        IND,
        FAM
      ),
    by = "IND"
  ) |>
  mutate(
    Method_label =
      unname(
        method_labels[
          Method
        ]
      )
  ) |>
  select(
    Trait,
    Rank,
    IND,
    FAM,
    Method,
    Method_label,
    Genomic_value,
    Predictive_evidence
  ) |>
  arrange(
    match(
      Trait,
      traits
    ),
    Rank
  )

kable(
  top10_with_family |>
    select(
      Trait,
      Rank,
      IND,
      FAM,
      Method_label,
      Genomic_value,
      Predictive_evidence
    ),
  digits = 5,
  caption = "Ten highest OOF genomic values by trait"
)
```

## 10. Individuals recurring across rankings

As an additional descriptive summary, individuals appearing in the top 10 of
at least two traits with positive OOF correlation and p < 0.05 are identified.

This table describes recurrence across rankings and does not constitute a
multi-trait selection index.

```{r recurrent-individuals}
positive_top10 <- top10_with_family |>
  filter(
    Predictive_evidence ==
      "positive_p_lt_0.05"
  )

recurring_top10 <- positive_top10 |>
  group_by(
    IND
  ) |>
  summarise(
    N_traits = n(),
    Traits = paste(
      Trait,
      collapse = ", "
    ),
    .groups = "drop"
  ) |>
  filter(
    N_traits >= 2L
  ) |>
  left_join(
    metadata |>
      select(
        IND,
        FAM
      ),
    by = "IND"
  ) |>
  arrange(
    desc(
      N_traits
    ),
    FAM,
    IND
  ) |>
  select(
    IND,
    FAM,
    N_traits,
    Traits
  )

kable(
  recurring_top10,
  caption = "Individuals recurring in the top 10 of traits with positive OOF correlation and p < 0.05"
)
```

A total of `r nrow(recurring_top10)` recurring individuals were identified.
The largest number of traits observed for the same individual was
`r max(recurring_top10$N_traits)`.

## 11. Final summary

```{r final-summary}
final_summary <- model_selection |>
  transmute(
    Trait,
    Selected_method,
    Selected_method_label,
    r,
    p_value,
    DIC = Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    Predictive_evidence
  ) |>
  arrange(
    match(
      Trait,
      traits
    )
  )

kable(
  final_summary |>
    select(
      Trait,
      Selected_method_label,
      r,
      p_value,
      DIC,
      Delta_DIC,
      Wprob,
      ER,
      Predictive_evidence
    ),
  digits = 5,
  caption = "Summary of final results by trait"
)
```

MMCOM, PRODCOM, CF, DF, DCI, and SS showed positive OOF correlation with
p < 0.05. PRODEF, FFC, and FFP showed positive correlation with p >= 0.05,
whereas EP showed nonpositive correlation.

## 12. Module outputs

```{r save-module05-results}
write.csv(
  article_model_table,
  file.path(
    RESULTS_DIR,
    "model_comparison_article_format.csv"
  ),
  row.names = FALSE
)

write.csv(
  final_summary,
  file.path(
    RESULTS_DIR,
    "final_results_summary.csv"
  ),
  row.names = FALSE
)

write.csv(
  rrblup_table,
  file.path(
    RESULTS_DIR,
    "rrblup_heritability_accuracy.csv"
  ),
  row.names = FALSE
)

write.csv(
  top10_with_family,
  file.path(
    RESULTS_DIR,
    "top10_genomic_values_with_family.csv"
  ),
  row.names = FALSE
)

write.csv(
  recurring_top10,
  file.path(
    RESULTS_DIR,
    "recurring_top10_individuals.csv"
  ),
  row.names = FALSE
)
```

## 13. Final audit

```{r module05-audit}
module05_audit <- data.frame(
  Check = c(
    "All official inputs were validated",
    "Comparison table contains 70 rows",
    "Final summary contains 10 traits",
    "RR-BLUP/BRR table contains 10 traits",
    "Top 10 table contains 100 rows",
    "Top 10 contains 10 individuals per trait",
    "All top-10 genomic values are finite",
    "Fixed colors are defined for all seven methods"
  ),
  Passed = c(
    all(
      input_audit$Passed
    ),
    nrow(
      article_model_table
    ) == 70L,
    nrow(
      final_summary
    ) == 10L,
    nrow(
      rrblup_table
    ) == 10L,
    nrow(
      top10_with_family
    ) == 100L,
    all(
      table(
        top10_with_family$Trait
      ) == 10L
    ),
    all(
      is.finite(
        top10_with_family$Genomic_value
      )
    ),
    length(
      method_colors[
        methods
      ]
    ) == 7L &
      all(
        !is.na(
          method_colors[
            methods
          ]
        )
      )
  )
)

write.csv(
  module05_audit,
  file.path(
    RESULTS_DIR,
    "module05_audit.csv"
  ),
  row.names = FALSE
)

kable(
  module05_audit,
  caption = "Final audit of Module 05"
)
```

Module 05 presents the results produced and validated in the previous stages.
An in-depth scientific interpretation of these results is outside the scope of
this page and will be developed in the manuscript.
