---
title: "Presentation of results"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

This module **presents** the results derived in Module 04. No model is refitted, no scientific metric is recalculated, and no selection decision is redefined.

`module04_results.rds` is the canonical source of analytical results. `models_object.rds` and native BGLR files are used only for graphical presentation of selected MCMC chains.

Detailed scientific interpretation is reserved for the manuscript.

## 2. Inputs and graphical standard

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(here)
})
```

```{r inputs-and-style}
MODELS_FILE <- here("output", "models", "models_object.rds")
M04_FILE <- here("output", "results", "m04", "module04_results.rds")
RESULTS_DIR <- here("output", "results", "m05")
FIGURE_DIR <- file.path(RESULTS_DIR, "figures")

dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)

models_object <- readRDS(MODELS_FILE)
m04 <- readRDS(M04_FILE)

required_m04_fields <- c(
  "signature",
  "traits",
  "methods",
  "model_comparison",
  "model_selection",
  "selected_fold_diagnostics",
  "runtime",
  "rrblup_heritability_summary",
  "selection_objectives",
  "selection_proportion",
  "supported_traits",
  "selection_sizes",
  "selection_cutoff_audit",
  "selected_individuals_by_trait",
  "multitrait_recurrence",
  "recurrence_distribution",
  "multitrait_candidates",
  "selected_oof_correlations"
)

missing_m04_fields <- setdiff(
  required_m04_fields,
  names(m04)
)

if (length(missing_m04_fields) > 0L) {
  stop(
    paste0(
      "M05 integrity gate failed: module04_results.rds is missing: ",
      paste(missing_m04_fields, collapse = ", "),
      ". Rebuild M04 before running M05."
    )
  )
}

if (
  length(models_object$signature) != 1L ||
  length(m04$signature) != 1L ||
  !identical(
    as.character(models_object$signature),
    as.character(m04$signature)
  )
) {
  stop(
    paste0(
      "M05 integrity gate failed: M03 and M04 analytical signatures differ. ",
      "Rebuild M04 from the current complete M03 object."
    )
  )
}

traits <- m04$traits
methods <- m04$methods
model_comparison <- m04$model_comparison
model_selection <- m04$model_selection
selected_fold_diagnostics <- m04$selected_fold_diagnostics
runtime_summary <- m04$runtime
rrblup_heritability_summary <- m04$rrblup_heritability_summary
selection_objectives <- m04$selection_objectives
selection_proportion <- m04$selection_proportion
supported_traits <- m04$supported_traits
selection_sizes <- m04$selection_sizes
selection_cutoff_audit <- m04$selection_cutoff_audit
selected_individuals_by_trait <- m04$selected_individuals_by_trait
multitrait_recurrence <- m04$multitrait_recurrence
recurrence_distribution <- m04$recurrence_distribution
multitrait_candidates <- m04$multitrait_candidates
selected_oof_correlations <- m04$selected_oof_correlations

expected_runs <-
  length(traits) *
  length(models_object$settings$folds) *
  length(methods)

expected_model_comparison_rows <-
  length(traits) *
  length(methods)

if (
  nrow(models_object$metrics_by_fold) != expected_runs ||
  nrow(
    distinct(
      models_object$metrics_by_fold,
      Trait,
      Fold,
      Method
    )
  ) != expected_runs
) {
  stop(
    paste0(
      "M05 integrity gate failed: the current M03 object is incomplete. ",
      "M05 requires the complete set of trait-fold-method fits."
    )
  )
}

if (
  nrow(model_comparison) != expected_model_comparison_rows ||
  nrow(model_selection) != length(traits) ||
  anyDuplicated(model_selection$Trait) > 0L ||
  !setequal(model_selection$Trait, traits)
) {
  stop(
    paste0(
      "M05 integrity gate failed: M04 does not contain one complete model ",
      "comparison and one selection decision for every trait."
    )
  )
}

if (
  any(
    !model_selection$Selected_method %in%
      methods
  )
) {
  stop(
    "M05 integrity gate failed: M04 contains an unknown selected method."
  )
}

# Presentation proceeds only when M03 and M04 belong to the same
# complete analytical execution.
method_labels <- c(
  RRBLUP = "RR-BLUP/BRR",
  BayesA = "BayesA",
  BayesB = "BayesB",
  BayesB2 = "BayesB2",
  BayesC = "BayesC",
  BL = "Bayesian Lasso",
  GBLUP = "GBLUP"
)

selection_objective_labels <- c(
  increase = "Increase",
  decrease = "Decrease"
)

method_colors <- c(
  RRBLUP = "#3366CC",
  BayesA = "#DC3912",
  BayesB = "#FF9900",
  BayesB2 = "#109618",
  BayesC = "#990099",
  BL = "#0099C6",
  GBLUP = "#DD4477"
)

project_theme <- theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )
```

## 3. Comparison of the seven methods

The table presents DIC, difference from the minimum DIC, `Wprob`, `ER`, OOF correlation, and nominal p-value. All quantities were derived in Module 04.

```{r model-comparison-table}
model_comparison_display <- model_comparison |>
  mutate(Method_label = unname(method_labels[Method])) |>
  transmute(
    Trait,
    Method,
    Method_label,
    DIC = Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    r,
    p_value
  ) |>
  arrange(match(Trait, traits), match(Method, methods))

kable(
  model_comparison_display |>
    select(Trait, Method_label, DIC, Delta_DIC, Wprob, ER, r, p_value),
  digits = 5,
  caption = "Comparison of the seven methods by trait"
)
```

```{r predictive-correlation-figure, fig.width=13, fig.height=10}
predictive_plot <- model_comparison |>
  mutate(Method = factor(Method, levels = methods)) |>
  ggplot(aes(x = Method, y = r, fill = Method)) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.4) +
  geom_col(width = 0.75) +
  facet_wrap(~ Trait, scales = "fixed", ncol = 2) +
  scale_fill_manual(values = method_colors, breaks = methods, labels = method_labels) +
  scale_x_discrete(labels = method_labels) +
  labs(
    x = NULL,
    y = "OOF correlation",
    fill = "Method",
    title = "Predictive ability of the seven methods"
  ) +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

predictive_plot

ggsave(
  file.path(FIGURE_DIR, "predictive_correlation_by_method_en.png"),
  predictive_plot,
  width = 13,
  height = 10,
  dpi = 300
)
```

## 4. Selected methods by trait

```{r selected-methods-table}
final_results_summary <- model_selection |>
  mutate(
    Method_label =
      unname(method_labels[Selected_method]),
    Predictive_winner_label =
      unname(method_labels[Predictive_winner]),
    DIC_winner_label =
      unname(method_labels[DIC_winner]),
    Decision_label = case_when(
      Decision_code ==
        "rrblup_best_predictive_model" ~
        "RR-BLUP was the best predictive model",
      Decision_code ==
        "rrblup_preferred_among_equivalent_models" ~
        "RR-BLUP preferred among equivalent methods",
      Decision_code ==
        "best_non_rrblup_equivalent_model" ~
        "Best competitive non-RR-BLUP method",
      Decision_code ==
        "predictive_winner_provisional_review_required" ~
        "Provisional predictive winner; review required",
      TRUE ~ Decision_code
    ),
    Selection_status_label = case_when(
      Selection_status ==
        "automatic_supported" ~
        "Supported automatic selection",
      Selection_status ==
        "provisional_review_required" ~
        "Provisional selection; review required",
      TRUE ~ Selection_status
    )
  ) |>
  transmute(
    Trait,
    Method_label,
    Decision_label,
    Selection_status_label,
    RRBLUP_preference_applied,
    Predictive_winner = Predictive_winner_label,
    DIC_winner = DIC_winner_label,
    Predictive_loss,
    r,
    p_value,
    RMSE,
    DIC = Mean_DIC,
    Delta_DIC,
    Wprob,
    ER,
    Parameters_flagged,
    Runs_with_flags,
    Predictive_evidence
  ) |>
  arrange(match(Trait, traits))

kable(
  final_results_summary,
  digits = 5,
  caption = "Selected methods and evidence by trait"
)
```

This table is the single presentation of the final decision. Repetitive textual winner lists are intentionally avoided so the page cannot become inconsistent after analyses are rerun.

## 5. MCMC chains of selected methods

As a visual representation analogous to the reference article, the posterior chain of \(\mu\), its running mean, and posterior density are shown for fold 1 of the selected method for each trait. Complete numerical convergence diagnostics remain in Module 03.

```{r selected-mu-data}
burn_in_saved <- models_object$settings$burn_in / models_object$settings$thin
selected_method_lookup <- setNames(model_selection$Selected_method, model_selection$Trait)

selected_mu_files <- file.path(
  here("output", "models", "bglr"),
  traits,
  "F1",
  unname(selected_method_lookup[traits]),
  "bglr_mu.dat"
)

if (!all(file.exists(selected_mu_files))) {
  stop(
    paste0(
      "M05 integrity gate failed: native fold-1 MCMC files are missing ",
      "for at least one selected method."
    )
  )
}

trace_parts <- vector("list", length(traits))
density_parts <- vector("list", length(traits))

for (trait_index in seq_along(traits)) {
  trait_name <- traits[trait_index]
  method_name <- selected_method_lookup[trait_name]

  mu_file <- file.path(
    here("output", "models", "bglr"),
    trait_name,
    "F1",
    method_name,
    "bglr_mu.dat"
  )

  mu_full <- scan(mu_file, quiet = TRUE)
  posterior_mu <- mu_full[seq.int(burn_in_saved + 1L, length(mu_full))]

  iteration <- seq.int(
    from = models_object$settings$burn_in + models_object$settings$thin,
    by = models_object$settings$thin,
    length.out = length(posterior_mu)
  )

  running_mean <- cumsum(posterior_mu) / seq_along(posterior_mu)
  keep_index <- seq.int(1L, length(posterior_mu), by = 100L)

  trace_parts[[trait_index]] <- data.frame(
    Trait = trait_name,
    Method = method_name,
    Facet = paste0(trait_name, " — ", method_labels[method_name]),
    Iteration = iteration[keep_index],
    Posterior_mu = posterior_mu[keep_index],
    Running_mean = running_mean[keep_index]
  )

  density_object <- density(posterior_mu, n = 512)
  density_parts[[trait_index]] <- data.frame(
    Trait = trait_name,
    Method = method_name,
    Facet = paste0(trait_name, " — ", method_labels[method_name]),
    x = density_object$x,
    y = density_object$y
  )
}

selected_mu_trace <- bind_rows(trace_parts)
selected_mu_density <- bind_rows(density_parts)
```

```{r selected-mu-trace-figure, fig.width=13, fig.height=14}
mu_trace_plot <- ggplot(selected_mu_trace, aes(x = Iteration, y = Posterior_mu)) +
  geom_line(linewidth = 0.25, alpha = 0.35) +
  geom_line(aes(y = Running_mean, color = Method), linewidth = 0.8) +
  facet_wrap(~ Facet, scales = "free_y", ncol = 2) +
  scale_color_manual(values = method_colors, breaks = methods, labels = method_labels) +
  labs(
    x = "Iteration",
    y = expression(mu),
    color = "Method",
    title = "Posterior chain and running mean of mu — fold 1"
  ) +
  project_theme

mu_trace_plot

ggsave(
  file.path(FIGURE_DIR, "selected_model_mu_trace_en.png"),
  mu_trace_plot,
  width = 13,
  height = 14,
  dpi = 300
)
```

```{r selected-mu-density-figure, fig.width=13, fig.height=14}
mu_density_plot <- ggplot(selected_mu_density, aes(x = x, y = y, color = Method)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ Facet, scales = "free", ncol = 2) +
  scale_color_manual(values = method_colors, breaks = methods, labels = method_labels) +
  labs(
    x = expression(mu),
    y = "Density",
    color = "Method",
    title = "Posterior density of mu — fold 1"
  ) +
  project_theme

mu_density_plot

ggsave(
  file.path(FIGURE_DIR, "selected_model_mu_density_en.png"),
  mu_density_plot,
  width = 13,
  height = 14,
  dpi = 300
)
```

## 6. Fold stability

```{r selected-fold-figure, fig.width=12, fig.height=10}
fold_stability_display <- selected_fold_diagnostics |>
  mutate(
    Method_label = unname(method_labels[Method]),
    Facet = paste0(Trait, " — ", Method_label)
  )

fold_stability_plot <- ggplot(
  fold_stability_display,
  aes(x = Fold, y = Correlation, group = Trait)
) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.4) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~ Facet, ncol = 2) +
  scale_x_continuous(breaks = 1:5) +
  coord_cartesian(ylim = c(-1, 1)) +
  labs(
    x = "Fold",
    y = "Validation correlation",
    title = "Fold-specific predictive correlation of selected methods"
  ) +
  project_theme

fold_stability_plot

ggsave(
  file.path(FIGURE_DIR, "selected_method_fold_stability_en.png"),
  fold_stability_plot,
  width = 12,
  height = 10,
  dpi = 300
)
```

## 7. Computational time

```{r runtime-figure, fig.width=11, fig.height=7}
runtime_plot <- runtime_summary |>
  mutate(Method = factor(Method, levels = methods)) |>
  ggplot(aes(x = Method, y = Mean_seconds, fill = Method)) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = pmax(0, Mean_seconds - SD_seconds),
      ymax = Mean_seconds + SD_seconds
    ),
    width = 0.2
  ) +
  scale_fill_manual(values = method_colors, breaks = methods, labels = method_labels) +
  scale_x_discrete(labels = method_labels) +
  labs(
    x = NULL,
    y = "Mean time (seconds)",
    fill = "Method",
    title = "Computational time by method"
  ) +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

runtime_plot

ggsave(
  file.path(FIGURE_DIR, "runtime_by_method_en.png"),
  runtime_plot,
  width = 11,
  height = 7,
  dpi = 300
)
```

The complete numerical timing summary remains a canonical Module 04 output; only the comparative figure is kept here.

## 8. Heritability and accuracy in the RR-BLUP/BRR analogue

```{r rrblup-heritability-table}
rrblup_table <- rrblup_heritability_summary |>
  transmute(
    Trait,
    RRBLUP_H2 = Mean_H2,
    SD_RRBLUP_H2 = SD_H2,
    RRBLUP_predictive_accuracy = Mean_predictive_accuracy,
    SD_RRBLUP_predictive_accuracy = SD_predictive_accuracy
  )

kable(
  rrblup_table,
  digits = 5,
  caption = "Heritability and predictive accuracy in the RR-BLUP/BRR analogue"
)
```

## 9. Correlations among OOF genomic values

```{r oof-correlation-figure, fig.width=10, fig.height=9}
correlation_long <- as.data.frame(as.table(selected_oof_correlations))
names(correlation_long) <- c("Trait_1", "Trait_2", "Correlation")

correlation_long <- correlation_long |>
  mutate(
    Correlation_label = replace(Correlation, abs(Correlation) < 0.005, 0)
  )

correlation_plot <- ggplot(
  correlation_long,
  aes(x = Trait_1, y = Trait_2, fill = Correlation)
) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", Correlation_label)), size = 3) +
  scale_fill_gradient2(
    low = "#DC3912",
    mid = "white",
    high = "#3366CC",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "r",
    title = "Correlation among OOF genomic values"
  ) +
  coord_equal() +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

correlation_plot

ggsave(
  file.path(FIGURE_DIR, "selected_oof_correlations_en.png"),
  correlation_plot,
  width = 10,
  height = 9,
  dpi = 300
)
```

## 10. Proportional selection by trait

Candidate selection uses only traits with positive OOF correlation and `p < 0.05`. The canonical proportion in this analysis is `r sprintf("%.0f%%", 100 * selection_proportion)` of individuals per trait, respecting the breeding direction defined in Module 04.

```{r proportional-selection-table}
selection_sizes_display <- selection_sizes |>
  left_join(
    selection_objectives |> select(Trait, Selection_objective),
    by = "Trait"
  ) |>
  mutate(
    Objective = unname(selection_objective_labels[Selection_objective])
  ) |>
  select(
    Trait, Objective, Selection_proportion,
    N_available, N_selected
  )

kable(
  selection_sizes_display,
  digits = 2,
  caption = "Selection proportion and number of individuals selected per trait with supported predictive evidence"
)
```

Individuals selected by trait remain available in the canonical `selected_individuals_by_trait.csv` output. They are not displayed in full on this page to avoid an excessively long table.

## 11. Multi-trait recurrence

```{r recurrence-distribution-table}
recurrence_class_labels <- c(
  high_4plus = "High (>=4)",
  intermediate_3 = "Intermediate (3)",
  broad_2 = "Broad (2)",
  specific_1 = "Specific (1)",
  none_0 = "None (0)"
)

recurrence_distribution_display <- recurrence_distribution |>
  mutate(
    Class = unname(recurrence_class_labels[Recurrence_class])
  ) |>
  select(
    N_traits, Class, N_individuals,
    Proportion_of_population
  )

kable(
  recurrence_distribution_display,
  digits = 4,
  caption = "Distribution of individuals across recurrence classes among supported traits"
)
```

```{r recurrent-candidates-table}
main_multitrait_candidates <- multitrait_candidates |>
  filter(N_traits >= 3L) |>
  mutate(
    Class = unname(recurrence_class_labels[Recurrence_class])
  ) |>
  select(
    IND, FAM, N_traits, Selection_frequency,
    Class, Traits, Best_rank, Mean_rank_present,
    ends_with("_rank")
  )

kable(
  main_multitrait_candidates,
  digits = 4,
  caption = "Candidates with high or intermediate recurrence among the most favorable 20% of supported traits"
)
```

Recurrence is a descriptive summary. `N_traits`, selection frequency, and ranks are not combined into a multi-trait index and do not imply overall superiority without explicitly defined economic or biological weights.
