---
title: "Presentation of results"
author:
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Bayesian-ridge-regression-papaya/
bibliography: ../REFERENCES.bib
link-citations: true
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
github-repo: WevertonGomesCosta/Bayesian-ridge-regression-papaya
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## 1. Objective

This module turns the results derived in Module 04 into a visual and tabular synthesis for scientific reading. If Module 04 answered **how the evidence is combined to reach the decisions**, Module 05 answers **how those decisions and their limitations should be read**.

The sequence follows the analytical reasoning: first compare the seven methods, then present the trait-specific decisions, illustrate MCMC diagnostics and across-fold stability, and finally show exploratory genomic rankings and recurrence across traits. No table or figure on this page constitutes a new selection or inferential step; they present results already defined upstream.

## 2. Packages, results, and graphical style

### 2.1 Packages used

The packages used here have distinct roles:

- `dplyr`: organizing, joining, and transforming result tables;
- `ggplot2`: building comparison and diagnostic figures;
- `knitr`: formatting the tables presented on the page;
- `here`: building project-relative paths.

```{r packages, warning=FALSE, message=FALSE}
library(dplyr)  # Organize and transform result tables.
library(ggplot2) # Build the synthesis figures.
library(knitr)  # Format tables in R Markdown.
library(here)   # Build project-relative paths.
```

```{r table-formatting, include=FALSE}
format_table <- function(x, caption = NULL, digits = getOption("digits"), col.names = NA) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    col.names = col.names,
    table.attr = paste0(
      'class="table table-striped table-hover" ',
      'style="width:100%; margin-bottom:0;"'
    )
  )
  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

### 2.2 Input objects and graphical style

This module consumes the persisted outputs from Modules 03 and 04. The labels and graphical settings defined below affect only how the results are displayed.

```{r inputs-and-style}
# Define output paths and load the objects required by this module.
MODELS_FILE <- here("output", "models", "models_object.rds")
M04_FILE <- here("output", "results", "m04", "module04_results.rds")
RESULTS_DIR <- here("output", "results", "m05")
FIGURE_DIR <- file.path(RESULTS_DIR, "figures")

dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)

models_object <- readRDS(MODELS_FILE)
m04 <- readRDS(M04_FILE)

traits <- m04$traits
methods <- m04$methods
model_comparison <- m04$model_comparison
model_selection <- m04$model_selection
selected_fold_diagnostics <- m04$selected_fold_diagnostics
runtime_summary <- m04$runtime
# Retrieve the heritability summary derived in M04.
rrblup_heritability_summary <- m04$rrblup_heritability_summary
selection_objectives <- m04$selection_objectives
selection_proportion <- m04$selection_proportion
positive_oof_traits <- m04$positive_oof_traits
selection_sizes <- m04$selection_sizes
selected_individuals_by_trait <- m04$selected_individuals_by_trait
multitrait_recurrence <- m04$multitrait_recurrence
recurrence_distribution <- m04$recurrence_distribution
multitrait_candidates <- m04$multitrait_candidates
selected_oof_correlations <- m04$selected_oof_correlations

# Define readable method labels used in tables and figures.
method_labels <- c(
  RRBLUP = "RR-BLUP/BRR",
  BayesA = "BayesA",
  BayesB = "BayesB",
  BayesB2 = "BayesB2",
  BayesC = "BayesC",
  BL = "Bayesian Lasso",
  GBLUP = "GBLUP"
)

selection_objective_labels <- c(
  increase = "Increase",
  decrease = "Decrease"
)

# Derive statistics used by the interpretive narrative from the scientific objects consolidated in Module 04.
n_traits <- length(traits)
n_methods <- length(methods)
n_selected_traits <- nrow(model_selection)


nonpositive_oof_traits <- setdiff(
  traits,
  positive_oof_traits
)

positive_oof_nonsignificant_traits <- model_selection$Trait[
  !is.na(model_selection$r) &
    model_selection$r > 0 &
    !is.na(model_selection$p_value) &
    model_selection$p_value >= 0.05
]

n_positive_oof_traits <- length(positive_oof_traits)

highest_oof_trait <- model_comparison$Trait[
  which.max(model_comparison$r)
]

format_list_en <- function(x) {
  x <- as.character(x)
  n <- length(x)

  if (n == 0L) {
    return("none")
  }

  if (n == 1L) {
    return(x)
  }

  paste0(
    paste(x[-n], collapse = ", "),
    " and ",
    x[n]
  )
}

method_order_present <- methods[
  methods %in% model_selection$Selected_method
]

selected_method_summary <- paste(
  vapply(
    method_order_present,
    function(method_name) {
      paste0(
        unname(method_labels[method_name]),
        " for ",
        format_list_en(
          model_selection$Trait[
            model_selection$Selected_method == method_name
          ]
        )
      )
    },
    character(1)
  ),
  collapse = "; "
)


n_available <- selection_sizes$N_available[[1]]
n_selected_per_trait <- selection_sizes$N_selected[[1]]
n_trait_candidate_occurrences <-
  nrow(selected_individuals_by_trait)
n_recurrent_candidates <- nrow(multitrait_candidates)

# Define the fixed palette used to distinguish methods in figures.
method_colors <- c(
  RRBLUP = "#3366CC",
  BayesA = "#DC3912",
  BayesB = "#FF9900",
  BayesB2 = "#109618",
  BayesC = "#990099",
  BL = "#0099C6",
  GBLUP = "#DD4477"
)

# Use base font size 12 for readability; this choice does not alter any analytical result.
project_theme <- theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )
```

### 2.3 Scope of the synthesis

The synthesis covers `r n_traits` traits and `r n_methods` methods. All `r n_selected_traits` traits have a selected method; `r n_positive_oof_traits` have positive OOF correlation and proceed to exploratory rankings. The trait outside this positive set is: `r format_list_en(nonpositive_oof_traits)`.

### 2.4 Figure hierarchy

The figures in this module do not have the same interpretive role. The OOF-correlation comparison in Section 3.1 is the **primary predictive-performance figure**. The `mu` traces and across-fold stability are **descriptive diagnostics** that contextualize the selected methods without redefining the choice. Computational time provides **operational context**. The heatmap of OOF genomic values is an **exploratory summary of associations among cross-fitted scores**, not a genetic-correlation matrix from a multivariate model.

The visual reading should therefore follow this priority: **predictive performance → diagnostics → operational context and exploratory associations**. None of the complementary figures creates an additional selection criterion.

## 3. Comparison of the seven methods

### 3.1 Primary figure: visual comparison of OOF performance

This is the primary figure for comparing predictive performance across the seven methods. The figure translates the OOF-correlation table into a direct visual comparison. The horizontal zero line separates positive from nonpositive association, and every panel uses the same vertical scale so magnitudes can be compared across traits. Bars are point estimates, not uncertainty intervals.

```{r predictive-correlation-figure, fig.width=13, fig.height=10, fig.alt="Faceted bar chart comparing pooled out-of-fold predictive correlations of the seven genomic prediction methods across the ten traits, with a horizontal reference at zero."}
# Order methods and build the OOF-correlation figure by trait.
predictive_plot <- model_comparison |>
  mutate(Method = factor(Method, levels = methods)) |>
  ggplot(aes(x = Method, y = r, fill = Method)) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.4) +
  geom_col(width = 0.75) +
  facet_wrap(~ Trait, scales = "fixed", ncol = 2) +
  scale_fill_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  scale_x_discrete(labels = method_labels) +
  labs(
    x = NULL,
    y = "OOF correlation",
    fill = "Method",
    title = "Predictive ability of the seven methods"
  ) +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

predictive_plot

ggsave(
  file.path(FIGURE_DIR, "predictive_correlation_by_method_en.png"),
  predictive_plot,
  width = 13,
  height = 10,
  dpi = 300
)
```

The comparison shows that predictive magnitude depends strongly on the trait. The largest OOF correlation observed in the comparison occurs for `r highest_oof_trait`, whereas several traits remain in much lower ranges. Within several panels, methods are relatively close for the same trait; method choice therefore remains trait-specific and follows the hierarchical rule consolidated in Module 04 rather than defining one global method.

### 3.2 Detailed tabular comparison

For every trait × method combination, the table reports mean DIC across folds, `Delta_DIC`, `Wprob`, `ER`, pooled OOF correlation, and the nominal p-value. DIC and its relative weights should be compared only among methods for the same trait; OOF correlation represents predictive performance and remains the primary reference. The p-value is descriptive only, as defined in Module 04.

```{r model-comparison-table}
# DISPLAY-ONLY START
model_comparison_display <- model_comparison |>
  mutate(Method_label = unname(method_labels[Method])) |>
  transmute(
    Trait,
    Method,
    Method_label,
    DIC = sprintf("%.2f", Mean_DIC),
    Delta_DIC = sprintf("%.2f", Delta_DIC),
    Wprob = sprintf("%.3f", Wprob),
    ER = sprintf("%.3f", ER),
    r = sprintf("%.3f", r),
    p_value = if_else(
      p_value < 0.001,
      "<0.001",
      sprintf("%.3f", p_value)
    )
  ) |>
  arrange(match(Trait, traits), match(Method, methods))

format_table(
  model_comparison_display |>
    select(Trait, Method_label, DIC, Delta_DIC, Wprob, ER, r, p_value),
  caption = "Comparison of the seven methods by trait",
  col.names = c("Trait", "Method", "DIC", "Delta DIC", "Wprob", "ER", "r", "p-value")
)
# DISPLAY-ONLY END
```

## 4. Selected methods by trait

### 4.1 Consolidated decision tables

The following two tables do not rerun method selection. They reorganize the decision already consolidated in Module 04 into a main table of decision and predictive evidence and a complementary table of fit evidence and diagnostics. A provisional status still requires review even though a method appears as selected.

In the first table, `Predictive_loss` is the difference between the pooled OOF correlation of the predictive winner and that of the selected method. A value of zero means that they coincide; a positive value shows how much pooled OOF correlation was given up by the selection rule. This column is descriptive and is neither a training loss function nor an additional selection criterion.

To reduce presentation width without losing information, the results are organized into two complementary tables. The first concentrates the **decision and primary predictive evidence**: selected method, reason for the choice, status, predictive and DIC winners, `Predictive_loss`, OOF correlation, RMSE, and predictive direction. The second collects the **complementary fit and diagnostic evidence**: explicit RR-BLUP preference, nominal p-value, DIC, `Delta_DIC`, `Wprob`, `ER`, and the MCMC indicators.

The objects created in the block below are presentation-only. Both are derived directly from `model_selection` and only translate labels, select columns, and format numbers; they do not recompute metrics, reapply the selection rule, filter traits, or change the selected method.

```{r selected-methods-table}
# DISPLAY-ONLY START

# Main table: consolidated decision and predictive evidence.
selected_decision_display <- model_selection |>
  mutate(
    Method_label = unname(method_labels[Selected_method]),
    Predictive_winner_label = unname(method_labels[Predictive_winner]),
    DIC_winner_label = unname(method_labels[DIC_winner]),
    Decision_label = case_when(
      Decision_code == "rrblup_best_predictive_model" ~
        "RR-BLUP/BRR was the best predictive model",
      Decision_code == "rrblup_preferred_among_competitive_models" ~
        "RR-BLUP/BRR preferred among competitive methods",
      Decision_code == "best_non_rrblup_competitive_model" ~
        "Best competitive non-RR-BLUP method",
      Decision_code == "predictive_winner_provisional_review_required" ~
        "Provisional predictive winner; review required",
      TRUE ~ Decision_code
    ),
    Selection_status_label = case_when(
      Selection_status == "rule_based_selected" ~
        "Rule-based selection",
      Selection_status == "provisional_review_required" ~
        "Provisional selection; review required",
      TRUE ~ Selection_status
    )
  ) |>
  transmute(
    Trait,
    Selected_method = Method_label,
    Decision = Decision_label,
    Status = Selection_status_label,
    Predictive_winner = Predictive_winner_label,
    DIC_winner = DIC_winner_label,
    Predictive_loss = sprintf("%.3f", Predictive_loss),
    r = sprintf("%.3f", r),
    RMSE = sprintf("%.3f", RMSE),
    OOF_direction = OOF_predictive_direction
  ) |>
  arrange(match(Trait, traits))

format_table(
  selected_decision_display,
  caption = "Consolidated decision and predictive evidence by trait",
  col.names = c("Trait", "Selected method", "Decision", "Status", "Predictive winner", "DIC winner", "Predictive loss", "r", "RMSE", "OOF direction")
)

# Complementary table: Bayesian fit and diagnostics used for review.
selected_support_display <- model_selection |>
  transmute(
    Trait,
    RRBLUP_preference_applied,
    p_value = if_else(
      p_value < 0.001,
      "<0.001",
      sprintf("%.3f", p_value)
    ),
    DIC = sprintf("%.2f", Mean_DIC),
    Delta_DIC = sprintf("%.2f", Delta_DIC),
    Wprob = sprintf("%.3f", Wprob),
    ER = sprintf("%.3f", ER),
    Parameters_flagged = as.integer(Parameters_flagged),
    Runs_with_flags = as.integer(Runs_with_flags)
  ) |>
  arrange(match(Trait, traits))

format_table(
  selected_support_display,
  caption = "Complementary fit evidence and diagnostics by trait",
  col.names = c("Trait", "RR-BLUP preference applied", "p-value", "DIC", "Delta DIC", "Wprob", "ER", "Parameters flagged", "Runs with flags")
)

# DISPLAY-ONLY END
```

### 4.2 Selection summary and ranking eligibility

The consolidated decision selects `r selected_method_summary`. All choices remain `Selection_status = rule_based_selected`. Traits with nonpositive OOF correlation — `r format_list_en(nonpositive_oof_traits)` — do not enter the exploratory candidate rankings.

## 5. Illustrative MCMC trace of mu (fold 1) for selected methods

### 5.1 Illustrative scope and chain preparation

For each trait, this section shows the posterior chain of \(\mu\) from fold 1 of the selected method and its running mean. Fold 1 is only a consistent visual illustration across traits; convergence assessment remains based on ESS and Geweke diagnostics over all five folds.

Module 05 reads only the previously saved `mu` chains for the selected methods. Burn-in and thinning inherited from Module 03 are respected; the running mean uses the full available posterior chain, while point reduction affects only trace readability.

```{r selected-mu-data}
# Convert burn-in to the saved-sample scale and prepare the posterior mu trace.
burn_in_saved <-
  models_object$settings$burn_in /
  models_object$settings$thin

selected_method_lookup <- setNames(
  model_selection$Selected_method,
  model_selection$Trait
)

trace_parts <- vector("list", length(traits))

for (trait_index in seq_along(traits)) {
  # Identify the trait and selected method.
  trait_name <- traits[trait_index]
  method_name <- selected_method_lookup[trait_name]

  # Locate the mu chain from the first fold.
  mu_file <- file.path(
    here("output", "models", "bglr"),
    trait_name,
    "F1",
    method_name,
    "bglr_mu.dat"
  )

  mu_full <- scan(mu_file, quiet = TRUE)

  # Remove samples corresponding to burn-in.
  posterior_mu <- mu_full[
    seq.int(burn_in_saved + 1L, length(mu_full))
  ]

  iteration <- seq.int(
    from = models_object$settings$burn_in + models_object$settings$thin,
    by = models_object$settings$thin,
    length.out = length(posterior_mu)
  )

  # Compute the running mean and reduce point density for plotting.
  running_mean <- cumsum(posterior_mu) / seq_along(posterior_mu)
  keep_index <- seq.int(1L, length(posterior_mu), by = 100L)

  trace_parts[[trait_index]] <- data.frame(
    Trait = trait_name,
    Method = method_name,
    Facet = paste0(trait_name, " — ", method_labels[method_name]),
    Iteration = iteration[keep_index],
    Posterior_mu = posterior_mu[keep_index],
    Running_mean = running_mean[keep_index]
  )
}

selected_mu_trace <- bind_rows(trace_parts)
```

### 5.2 Trace and running mean

In the trace plot, the thin line shows posterior samples of \(\mu\) after burn-in and the colored line shows the running mean. For readability, only one of every 100 posterior samples is drawn; this reduction affects plotting only, while the running mean was computed from the full available posterior chain. Visual stabilization of the mean is descriptive evidence, not a standalone convergence criterion.

```{r selected-mu-trace-figure, fig.width=13, fig.height=14, fig.alt="MCMC trace plots of the intercept parameter mu for illustrative fold-1 chains of the trait-specific selected methods."}
mu_trace_plot <- ggplot(
  selected_mu_trace,
  aes(x = Iteration, y = Posterior_mu)
) +
  geom_line(linewidth = 0.25, alpha = 0.35) +
  geom_line(
    aes(y = Running_mean, color = Method),
    linewidth = 0.8
  ) +
  facet_wrap(~ Facet, scales = "free_y", ncol = 2) +
  scale_color_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  labs(
    x = "Iteration",
    y = expression(mu),
    color = "Method",
    title = "Posterior chain and running mean of mu — fold 1"
  ) +
  project_theme

mu_trace_plot

ggsave(
  file.path(FIGURE_DIR, "selected_model_mu_trace_en.png"),
  mu_trace_plot,
  width = 13,
  height = 14,
  dpi = 300
)
```

The trace is only an illustration of the fold-1 `mu` chain. It does not alter selection or convergence classification, which remain based on ESS and Geweke diagnostics over all five folds.

## 6. Descriptive stability diagnostic across folds

This figure returns the focus to cross-validation. Each panel shows the five validation correlations for the selected method; sign changes or large ranges among points suggest descriptive instability. Because training sets overlap and there are only five folds, the plot is neither a formal heterogeneity test nor a confidence interval.

```{r selected-fold-figure, fig.width=12, fig.height=10, fig.alt="Faceted plot showing the five fold-level validation correlations for the selected method of each trait, used as a descriptive view of across-fold stability."}
# Add labels to fold-level results and build the stability figure for the selected method.
fold_stability_display <- selected_fold_diagnostics |>
  mutate(
    Method_label = unname(method_labels[Method]),
    Facet = paste0(Trait, " — ", Method_label)
  )

# Build the plot of predictive correlation observed in each fold for the selected method.
fold_stability_plot <- ggplot(
  fold_stability_display,
  aes(x = Fold, y = Correlation, group = Trait)
) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.4) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~ Facet, ncol = 2) +
  scale_x_continuous(breaks = 1:5) +
  coord_cartesian(ylim = c(-1, 1)) +
  labs(
    x = "Fold",
    y = "Validation correlation",
    title = "Predictive correlation across folds for selected methods"
  ) +
  project_theme

fold_stability_plot

ggsave(
  file.path(FIGURE_DIR, "selected_method_fold_stability_en.png"),
  fold_stability_plot,
  width = 12,
  height = 10,
  dpi = 300
)
```

Fold stability should be interpreted together with pooled OOF correlation. Larger ranges or sign changes are descriptively more relevant for traits whose pooled signal is already weak, but they do not redefine method selection. With only five folds and overlapping training sets, this variation remains a descriptive stability measure.

## 7. Operational context: computational time

Bars show the mean recorded time across the 50 fits of each method, and error bars show one standard deviation across those fits. The comparison is operational for this execution and environment; hardware, system load, implementation, and I/O differences prevent interpreting the plot as a universal speed benchmark.


```{r runtime-figure, fig.width=11, fig.height=7, fig.alt="Bar chart comparing mean computational time across the seven genomic prediction methods, with error bars representing one standard deviation across fits."}
# Summarize the computational cost of the evaluated methods.
runtime_plot <- runtime_summary |>
  mutate(Method = factor(Method, levels = methods)) |>
  ggplot(aes(x = Method, y = Mean_seconds, fill = Method)) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = pmax(0, Mean_seconds - SD_seconds),
      ymax = Mean_seconds + SD_seconds
    ),
    width = 0.2
  ) +
  scale_fill_manual(
    values = method_colors,
    breaks = methods,
    labels = method_labels
  ) +
  scale_x_discrete(labels = method_labels) +
  labs(
    x = NULL,
    y = "Mean time (seconds)",
    fill = "Method",
    title = "Computational time by method"
  ) +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

runtime_plot

ggsave(
  file.path(FIGURE_DIR, "runtime_by_method_en.png"),
  runtime_plot,
  width = 11,
  height = 7,
  dpi = 300
)
```

The runtime ordering documents only the computational cost observed in this execution. Fitting time does not enter the method-selection rule and should not replace predictive, DIC, or MCMC-diagnostic criteria.

## 8. Marker-based heritability analogue and adjusted accuracy in RR-BLUP/BRR

The table re-presents the quantities derived in Module 04 for RR-BLUP/BRR. `RRBLUP_H2` is a marker-model-based heritability analogue, not an independent classical estimate of narrow-sense heritability. `RRBLUP_predictive_accuracy` is the mean ratio \(r_f/\sqrt{H_f^2}\); because it is derived from sample estimates, it may be negative or exceed 1 and should not be interpreted as a correlation necessarily bounded to `[0, 1]`.

```{r rrblup-heritability-table}
# DISPLAY-ONLY START
rrblup_table <- rrblup_heritability_summary |>
  transmute(
    Trait,
    RRBLUP_H2 = sprintf("%.3f", Mean_H2),
    SD_RRBLUP_H2 = sprintf("%.3f", SD_H2),
    RRBLUP_predictive_accuracy = sprintf("%.3f", Mean_predictive_accuracy),
    SD_RRBLUP_predictive_accuracy = sprintf("%.3f", SD_predictive_accuracy)
  )

format_table(
  rrblup_table,
  caption = "Marker-based heritability analogue and adjusted predictive accuracy in RR-BLUP/BRR",
  col.names = c("Trait", "RR-BLUP H2", "SD RR-BLUP H2", "RR-BLUP predictive accuracy", "SD RR-BLUP predictive accuracy")
)
# DISPLAY-ONLY END
```

These quantities contextualize the variability captured by RR-BLUP/BRR and the relationship between predictive correlation and the heritability analogue. They do not rerank the already selected methods and should not be converted into conclusions about classical population heritability outside this model.

## 9. Exploratory associations among OOF genomic values

### 9.1 What the matrix represents

The heatmap presents associations among OOF genomic values produced by the selected method for each trait. Because different traits may use different methods and the values are cross-fitted, these associations are not genetic correlations estimated by a multivariate model. Correlations are used without rounding in calculations and in the color scale; only labels with `|r| < 0.005` are displayed as zero to avoid `-0.00`, without changing the numeric value or fill color.

### 9.2 Heatmap and interpretation

Cell rounding is display-only; the color scale continues to use the original OOF correlations.

```{r oof-correlation-figure, fig.width=10, fig.height=9, fig.alt="Heatmap of pairwise correlations among cross-fitted OOF genomic-value scores for the traits, representing exploratory associations rather than multivariate genetic correlations."}
# Reshape the OOF correlation matrix and build the cross-trait heatmap.
correlation_long <- as.data.frame(
  as.table(selected_oof_correlations)
)
names(correlation_long) <- c(
  "Trait_1",
  "Trait_2",
  "Correlation"
)

# Set only the label of |r| < 0.005 to zero; the original correlation still drives analysis and color.
correlation_long <- correlation_long |>
  mutate(
    Correlation_label = replace(
      Correlation,
      abs(Correlation) < 0.005,
      0
    )
  )

# Build the heatmap of correlations among OOF genomic values.
correlation_plot <- ggplot(
  correlation_long,
  aes(x = Trait_1, y = Trait_2, fill = Correlation)
) +
  geom_tile() +
  geom_text(
    aes(label = sprintf("%.2f", Correlation_label)),
    size = 3
  ) +
  scale_fill_gradient2(
    low = "#DC3912",
    mid = "white",
    high = "#3366CC",
    midpoint = 0,
    # Keep the color scale fixed to the theoretical correlation range [-1, 1].
    limits = c(-1, 1)
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "r",
    title = "Correlation among OOF genomic values"
  ) +
  coord_equal() +
  project_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

correlation_plot

ggsave(
  file.path(FIGURE_DIR, "selected_oof_correlations_en.png"),
  correlation_plot,
  width = 10,
  height = 9,
  dpi = 300
)
```

The heatmap should be read as association among cross-fitted OOF scores: positive values indicate a tendency for two traits to rank individuals in the same direction, whereas negative values indicate opposing directions. These associations are not genetic correlations from a multivariate model and do not create an additional selection criterion.

## 10. Exploratory proportional candidate ranking by trait

### 10.1 Eligibility and ranking size

Rankings are shown only for traits whose selected method has positive OOF correlation. Nominal p-values are descriptive and do not control display; when `p >= 0.05`, the ranking remains explicitly exploratory and does not demonstrate predictive utility. The proportion shown is `r sprintf("%.0f%%", 100 * selection_proportion)` per trait — `r n_selected_per_trait` of the `r n_available` individuals in this dataset — and is a configurable rule in this analysis rather than an estimated optimal intensity. Scores are cross-fitted OOF genomic values, so an individual's own phenotype does not contribute to its score; even so, they are not final breeding values from a model refitted to all phenotypes.

### 10.2 Tabular ranking summary

The table combines selection objective, OOF evidence, and ranking size. The nominal p-value is displayed for context and is not used to remove rankings.

```{r proportional-selection-table}
# DISPLAY-ONLY START
selection_sizes_display <- selection_sizes |>
  left_join(
    selection_objectives |>
      select(Trait, Selection_objective),
    by = "Trait"
  ) |>
  left_join(
    model_selection |>
      select(Trait, r, p_value, OOF_nominal_signal),
    by = "Trait"
  ) |>
  mutate(
    Objective = unname(
      selection_objective_labels[Selection_objective]
    )
  ) |>
  transmute(
    Trait,
    Objective,
    r = sprintf("%.3f", r),
    p_value = if_else(
      p_value < 0.001,
      "<0.001",
      sprintf("%.3f", p_value)
    ),
    OOF_nominal_signal,
    Selection_proportion = sprintf("%.0f%%", 100 * Selection_proportion),
    N_available = as.integer(N_available),
    N_selected = as.integer(N_selected)
  )

format_table(
  selection_sizes_display,
  caption = "Exploratory candidate-ranking summary by positive-OOF trait",
  col.names = c("Trait", "Objective", "r", "p-value", "OOF nominal signal", "Selection proportion", "Available candidates", "Selected candidates")
)
# DISPLAY-ONLY END
```

### 10.3 Candidate preview

The next table previews only the first five individuals from each ranking (`n = 5`) for readability; the complete `selected_individuals_by_trait` ranking remains available in the object and derived files.

```{r proportional-selection-preview}
# DISPLAY-ONLY START
proportional_selection_preview_display <- selected_individuals_by_trait |>
  group_by(Trait) |>
  slice_head(n = 5) |>
  ungroup() |>
  transmute(
    Trait,
    Rank = as.integer(Rank),
    IND,
    FAM,
    Method,
    Selection_objective,
    Genomic_value = sprintf("%.3f", Genomic_value),
    Selection_score = sprintf("%.3f", Selection_score),
    OOF_r = sprintf("%.3f", OOF_r),
    OOF_nominal_p_value = if_else(
      OOF_nominal_p_value < 0.001,
      "<0.001",
      sprintf("%.3f", OOF_nominal_p_value)
    ),
    OOF_nominal_signal
  )

format_table(
  proportional_selection_preview_display,
  caption = "Top five candidates in each positive-OOF trait ranking",
  col.names = c("Trait", "Rank", "IND", "FAM", "Method", "Selection objective", "Genomic value", "Selection score", "OOF r", "OOF nominal p-value", "OOF nominal signal")
)
# DISPLAY-ONLY END
```

### 10.4 Totals and exploratory signals

The `r n_positive_oof_traits` traits with positive OOF correlation produce rankings, each containing `r n_selected_per_trait` of the `r n_available` individuals, for `r n_trait_candidate_occurrences` trait × candidate occurrences before recurrence is considered. Traits outside the positive-OOF set are: `r format_list_en(nonpositive_oof_traits)`. Among traits with positive OOF direction, those with nominal p-value `>= 0.05` are: `r format_list_en(positive_oof_nonsignificant_traits)`; these signals therefore remain explicitly exploratory.

Up to this point, each ranking has been treated separately within its trait. Recurrence changes only the level of summary: the `r n_trait_candidate_occurrences` trait × candidate occurrences from the positive-OOF rankings are grouped by individual to count in how many traits each candidate appears.

This step does not create a new score and does not rerank individuals by combining traits. A recurrent candidate is simply one present in at least two rankings in the scientific object consolidated in Module 04. The `N_traits >= 3` cutoff used below remains exclusively a presentation reduction for the main table. Thus, the sequence should be read as **within-trait selection → overlap counting across lists**, not as fitting a multi-trait model.

## 11. Recurrence across traits

### 11.1 Scientific contract and presentation filter

The recurrence classes shown here are the same ones defined in Module 04. The complete object retains candidates present in at least two traits; to keep the main table compact, this module displays only intermediate or high recurrence, that is, candidates present in three or more traits. The `N_traits >= 3` cutoff is a presentation choice only and does not redefine the stored candidate set.

### 11.2 Recurrence distribution

The first recurrence table describes the whole population according to the number of positive-OOF rankings in which each individual appears. The classes are presentation labels defined in Module 04 rather than significance thresholds.

The distribution below therefore describes the full recurrence pattern before the presentation subset used in Section 11.3 is applied.

```{r recurrence-distribution-table}
# DISPLAY-ONLY START
recurrence_class_labels <- c(
  high_4plus = "High (>=4)",
  intermediate_3 = "Intermediate (3)",
  broad_2 = "Broad (2)",
  specific_1 = "Specific (1)",
  none_0 = "None (0)"
)

recurrence_distribution_display <- recurrence_distribution |>
  mutate(
    Class = unname(
      recurrence_class_labels[Recurrence_class]
    )
  ) |>
  transmute(
    N_traits = as.integer(N_traits),
    Class,
    N_individuals = as.integer(N_individuals),
    Proportion_of_population = sprintf("%.3f", Proportion_of_population)
  )

format_table(
  recurrence_distribution_display,
  caption = "Distribution of individuals by recurrence class",
  col.names = c("Number of traits", "Class", "Individuals", "Population proportion")
)
# DISPLAY-ONLY END
```

### 11.3 Recurrent-candidate tables

To reduce width without removing information, the presentation is divided into two tables. The first summarizes identification, recurrence intensity, involved traits, and overall ranking position. The second separately shows the trait-specific rank profile, making it possible to inspect the lists in which each candidate appeared without overloading the main table.

Both tables use the presentation subset defined in Section 11.1.

```{r recurrent-candidates-table}
# DISPLAY-ONLY START

# Apply only the presentation filter used on this page.
main_multitrait_candidates <- multitrait_candidates |>
  filter(N_traits >= 3L) |>
  mutate(
    Class = unname(
      recurrence_class_labels[Recurrence_class]
    )
  )

# Main table: recurrence and overall ranking summary.
main_multitrait_summary <- main_multitrait_candidates |>
  select(
    IND,
    FAM,
    N_traits,
    Selection_frequency,
    Class,
    Traits,
    Best_rank,
    Mean_rank_present
  ) |>
  mutate(
    N_traits = as.integer(N_traits),
    Selection_frequency = sprintf("%.3f", Selection_frequency),
    Best_rank = as.integer(Best_rank),
    Mean_rank_present = sprintf("%.2f", Mean_rank_present)
  )

format_table(
  main_multitrait_summary,
  caption = "Summary of candidates with high or intermediate recurrence",
  col.names = c("IND", "FAM", "Number of traits", "Selection frequency", "Class", "Traits", "Best rank", "Mean rank when present")
)

# Complementary table: candidate positions in trait-specific rankings.
main_multitrait_rank_profile <- main_multitrait_candidates |>
  select(
    IND,
    ends_with("_rank")
  ) |>
  select(-Best_rank)

format_table(
  main_multitrait_rank_profile,
  caption = "Trait-specific rank profiles of recurrent candidates",
  col.names = c("IND", "MMCOM rank", "PRODCOM rank", "PRODEF rank", "CF rank", "DF rank", "DCI rank", "FFC rank", "FFP rank", "SS rank")
)

# DISPLAY-ONLY END
```

### 11.4 Interpretation of recurrence

Recurrence summarizes, across traits with positive OOF correlation, in how many lists each individual appears among the most favorable `r sprintf("%.0f%%", 100 * selection_proportion)`. `Selection_frequency` uses the number of positive-OOF traits as its denominator, not necessarily all `r n_traits` original traits. Recurrence supported by weak positive OOF signals remains exploratory. `N_traits`, frequency, and ranks are not combined into a multi-trait index; therefore, greater recurrence does not demonstrate joint superiority across multiple traits.

In the complete Module 04 object, `r n_recurrent_candidates` individuals recur in at least two traits. The main table on this page shows only the subset recurring in three or more traits; this presentation reduction does not modify the stored scientific set or create a multi-trait index.

## 12. Summary

The final reading should preserve the hierarchy established throughout the analysis: OOF performance drives predictive comparison; DIC, across-fold stability, and MCMC diagnostics provide complementary context; rankings are produced only when OOF direction is positive; and recurrence summarizes overlap among rankings without constructing a multi-trait index. All conclusions remain conditional on the represented breeding population, the available SSR panel, and the transductive validation design. For operational selection, OOF rankings should be treated as exploratory evidence rather than substitutes for final breeding values from a prediction stage designed for decision making.

<div style="display:flex; justify-content:space-between; gap:1rem; margin-top:2rem; padding-top:1rem; border-top:1px solid #ddd;">
<a href="04_results_en.html">← Module 04 — Comparison and selection</a>
<a href="index.html">Home →</a>
</div>
