---
title: "6. Genomic values and candidate rankings"
author: "Weverton Gomes da Costa"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
bibliography: ../REFERENCES.bib
link-citations: true
---

[Português](06_genomic_selection_pt.html) | [Previous](05_model_comparison_en.html) | [Home](index.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 8)
source(file.path("code", "00_config.R"))
source(file.path("code", "02_model_functions.R"))
source(file.path("code", "03_reporting_functions.R"))
configuracao <- obter_configuracao()
objeto <- carregar_resultado(configuracao$arquivo_preprocessamento)
resultado_bayesiano <- carregar_resultado(configuracao$arquivo_bayesiano)
resultado_rrblup <- carregar_resultado(configuracao$arquivo_rrblup)
resultado_gblup <- carregar_resultado(configuracao$arquivo_gblup)
comparacao <- carregar_resultado(configuracao$arquivo_comparacao)
```

## Selection scope

This page links predictive model choice to full-data genomic estimated breeding
values (GEBVs). For each trait, it extracts GEBVs from the model selected by
out-of-fold performance and joins block and family metadata. Two rankings are
retained: high genomic value and low genomic value.

These are **candidate rankings, not a final multi-trait recommendation**.
Production of defective fruit and internal-cavity diameter may have breeding
directions different from production, mass, or soluble solids. Fruit size and
firmness can also have target ranges instead of one-sided optima. The project
therefore does not silently assign directions or economic weights.

## Build ranking tables

The ranking object is built only after every fitted-model result and the common
comparison are available. It records the predictive model used for each trait
so the provenance of every GEBV is explicit.

```{r build-selection-tables}
resultados_disponiveis <- all(vapply(
  list(
    objeto, resultado_bayesiano, resultado_rrblup,
    resultado_gblup, comparacao
  ),
  Negate(is.null),
  logical(1)
))

if (resultados_disponiveis) {
  selecao <- construir_tabelas_selecao(
    objeto,
    resultado_bayesiano,
    resultado_rrblup,
    resultado_gblup,
    comparacao,
    configuracao
  )
} else {
  selecao <- carregar_resultado(configuracao$arquivo_selecao)
}
```

```{r selection-status, results='asis', echo=FALSE}
if (is.null(selecao)) {
  cat(mensagem_resultado_ausente(configuracao$arquivo_selecao, "en"))
}
```

## Model provenance

The table below identifies the predictive model that supplies the full-data
GEBVs for each trait. Its correlation and errors come only from held-out
predictions; the GEBV itself comes from the final fit using all individuals.

```{r selection-model-table, results='asis'}
if (!is.null(comparacao)) {
  modelos_selecao <- comparacao$best_models |>
    dplyr::select(Trait, Model, Correlation, RMSE, MAE, Bias, Calibration_slope)
  print(tabela_relatorio(modelos_selecao, "Model used for each trait-specific ranking"))
}
```

## High-value candidates

The first table reports the 15 highest GEBVs for every trait. Use it directly
only for traits whose breeding objective is explicitly to increase the trait.
For objectives requiring a decrease, use `Rank_low_value` in the complete
ranking or the high-and-low extremes exported by the pipeline.

```{r high-ranking-table, results='asis'}
if (!is.null(selecao)) {
  candidatos_altos <- selecao$top_15_by_trait |>
    dplyr::select(
      Trait, Rank_high_value, IND, BL, FAM, Model, GEBV, Fitted,
      Rank_low_value
    )
  print(tabela_relatorio(
    candidatos_altos,
    "Fifteen highest genomic values per trait",
    rolagem = TRUE,
    altura = "600px"
  ))
}
```

The plot provides a compact view of the ten highest GEBVs per trait and the
model from which they were obtained. GEBV magnitudes are not comparable across
traits because scales differ.

```{r high-ranking-plot, echo=FALSE, fig.height=12}
if (!is.null(selecao)) {
  top_10 <- selecao$rankings_by_trait |>
    dplyr::filter(Rank_high_value <= 10L) |>
    dplyr::mutate(IND = factor(IND, levels = unique(IND[order(GEBV)])))

  print(
    ggplot2::ggplot(top_10, ggplot2::aes(x = GEBV, y = IND, color = Model)) +
      ggplot2::geom_point(size = 2) +
      ggplot2::facet_wrap(~ Trait, scales = "free", ncol = 2) +
      ggplot2::labs(
        x = "GEBV",
        y = "Individual",
        color = "Selected model",
        title = "Ten highest genomic values per trait"
      ) +
      tema_relatorio()
  )
}
```

## High and low extremes

Both ends of each trait distribution are retained so the breeding program can
apply a documented direction later. The full CSV output contains all 150
individuals, both ranks, family, block, selected model, GEBV, and fitted value.

```{r extremes-table, results='asis'}
if (!is.null(selecao)) {
  extremos <- selecao$extreme_15_by_trait |>
    dplyr::select(
      Trait, Ranking_side, IND, BL, FAM, Model, GEBV,
      Rank_high_value, Rank_low_value
    )
  print(tabela_relatorio(
    extremos,
    "Fifteen candidates at each genomic-value extreme",
    rolagem = TRUE,
    altura = "600px"
  ))
}
```

## Required decision for a multi-trait index

A defensible selection index requires, at minimum, a direction or target for
each trait and either economic weights or desired gains. Once these are
provided, the current outputs can support standardized indices, restricted
indices, or independent-culling rules without rerunning marker preprocessing.

Until then, the reproducible deliverables are trait-specific rankings and
their validated model provenance.

[Back to home](index.html)
