---
title: "Genomic selection in papaya / Seleção genômica em mamoeiro"
author: "Weverton Gomes da Costa"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
bibliography: ../REFERENCES.bib
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source(file.path("code", "00_config.R"))
source(file.path("code", "01_data_functions.R"))
source(file.path("code", "03_reporting_functions.R"))
configuracao <- obter_configuracao()
objeto <- carregar_resultado(configuracao$arquivo_preprocessamento)
```

[English](#english) | [Português](#português)

## English

This reproducible `workflowr` project compares six Bayesian whole-genome
regression methods, RR-BLUP, and GBLUP for genomic prediction in papaya. The
design follows the Bayesian SSR framework of Silva et al. [-@silva2021brrguava]
and extends it with common out-of-fold prediction metrics for all eight models.

The data comprise 150 individuals, 35 multi-allelic SSR loci, and ten
production and fruit-quality traits. Individual identity is linked through the
`Genealex` sheet, preserving the alphanumeric identifier `144B`. Block and
family are fitted as categorical fixed effects.

```{r overview-en, results='asis'}
if (is.null(objeto)) {
  cat(mensagem_resultado_ausente(configuracao$arquivo_preprocessamento, "en"))
} else {
  resumo <- data.frame(
    Item = c("Individuals", "SSR loci", "Retained allele-dosage columns", "Traits", "CV folds"),
    Value = c(
      length(objeto$ids),
      ncol(objeto$codigos_ssr),
      ncol(objeto$dosagens_ssr),
      ncol(objeto$fenotipos),
      configuracao$n_folds
    )
  )
  print(tabela_relatorio(resumo, "Analysis inventory", digitos = 0))
}
```

Start with [data and preprocessing](01_preprocessing_en.html), then follow the
numbered pages in the navigation bar. The website can be built without running
the expensive MCMC models; computed results are loaded when available.

## Português

Este projeto reprodutível em `workflowr` compara seis regressões genômicas
Bayesianas, RR-BLUP e GBLUP para predição genômica em mamoeiro. O desenho segue
a abordagem Bayesiana com SSR de Silva et al. [-@silva2021brrguava] e a amplia
com métricas preditivas fora da amostra comuns aos oito modelos.

Os dados reúnem 150 indivíduos, 35 locos SSR multialélicos e dez
características de produção e qualidade de frutos. A identidade é vinculada
pela planilha `Genealex`, preservando o identificador alfanumérico `144B`.
Bloco e família são ajustados como efeitos fixos categóricos.

Comece por [dados e pré-processamento](01_preprocessing_pt.html) e siga as
páginas numeradas. O site pode ser construído sem executar as cadeias MCMC;
resultados previamente calculados são carregados quando disponíveis.

## Reproducibility

The repository versions the source workbooks, analysis code, tests, site
sources, and an `renv` dependency lockfile. The complete Bayesian profile uses
1,000,000 iterations, 200,000 burn-in iterations, and thinning of 4, matching
the main methods section of the reference article.

The primary comparison uses deterministic family-stratified eight-fold cross-
validation. Missing marker calls are imputed and centered using training data
inside each fold, preventing information leakage.

## Citation

Please cite the reference article and this repository when reusing the
workflow. See [About](about.html) for the methodological scope and
[License](license.html) for reuse terms.
